### 2021-05-05 proposal/proposal.Rmd
started proposal

---
output:
  pdf_document:
    includes:
      before_body: title.sty
    keep_tex: true
    citation_package: natbib
    number_sections: yes
# output:
#   bookdown::pdf_document2:
#     citation_package: natbib
#     number_citations: yes
# output: html_document
# output: rticles::rss_article
fontsize: 11pt
# geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
bibliography: proposal.bib
abstract: |
  asdf
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      # eval = FALSE,
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 3, 
                      fig.width = 5, 
                      fig.dpi = 300)

options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{indep}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{example}
\newtheorem*{example}{Example}

# Research Goal

Typical clustering methods often work by assuming that each cluster is represented by some central point around which the data belonging to that cluster lie Such methods, including Lloyd's algorithm for $k$-means clustering \cite{1056489} and Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}, involve iteratively updating the central points by averaging data belonging to each cluster and cluster memberships by comparing proximity to central points. The aim of our research is to explore clustering and community detection methods for which each cluster is represented not by a central point but by subspaces or manifolds. In addition, we aim to explore various applications of such methods and connect them to preexisting clustering and community detection problems.

# Literature Review

## Subspace Clustering

Subspace clustering, which assumes that points in $\mathbb{R}^d$ each lie on one of $K$ subspaces of $\mathbb{R}^d$, is an approach that has found a wide range of uses by the Statistics and Machine Learning communities, particularly within the field of Computer Vision \cite{5206547}.

Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample. Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$. The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$. If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces. Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.

In practice, SSC is performed by solving the LASSO problems:

\begin{equation} \label{eq:ssc}
c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
\end{equation}

for some sparsity parameter $\lambda > 0$. The $c_i$ vectors are then collected into $C$ and $B$ as described before. If $X$ is noisy in that each $x_i$ does not lie exactly on one of $K$ subspaces but near it, the choice of $\lambda$ becomes important in guaranteeing the Subspace Detection Property (SDP) \cite{jmlr-v28-wang13}.

\begin{definition}[Subspace Detection Property] 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
different subspaces, then $X$ obeys the Subspace Detection Property. 
\end{definition}

\begin{remark} 
In practice, a noisy sample $X$ often does not obey SDP. 
In such cases, $B$ is treated as an affinity matrix for a graph which 
is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
hand, if $X$ does obey the SDP, $B$ describes a graph 
with at least $K$ disconnected subgraphs. Ideally, when SDP holds, 
there are exactly $K$ subgraphs which map to each subspace, 
but it could be the case that some of the subspaces are represented by 
multiple disconnected subgraphs. SDP is contingent 
on choosing a sufficiently large sparsity parameter $\lambda$. 
\end{remark}

## (Generalized) Random Dot Product Graphs

It has been shown by \citet{jmlr-v28-wang13} that given sufficiently low noise in $X$ and sufficiently low affinity between pairs of subspaces as measured by the cosine of the angle between subspaces, SDP holds for a specific range of $\lambda$. This provides theoretical justification for applying SSC to Random Dot Product Graphs (RDPG) \cite{athreya2017statistical} and Generalized Random Dot Product Graphs (GRDPG) \cite{rubindelanchy2017statistical} for which the data in the latent space lie on subspaces.

\begin{definition}[(Generalized) Random Dot Product Graph]
Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y.\in [0, 1]$. $G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix $A$ is drawn such that $A_{ij} \sim \text{Bernoulli}(x_i^\top x_j)$ for $i < j$, with $A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$. If on the other hand $A_{ij} \sim \text{Bernoulli}(x_i^\top I_{p, q} x_j)$ where $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph. These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.

In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $x_1, ..., x_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$. 
\end{definition}

\begin{definition}[Adjacency Spectral Embedding]
Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$. Let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors. Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.

If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors. Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
\end{definition}

\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$, if $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,

\begin{equation}
\max_i \|(\hat{X}_n)_i - W_n (X_n)_i \| \stackrel{a.s.}{\to} 0
\end{equation}

Similarly, \citeauthor{rubindelanchy2017statistical} showed that for $(A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)$,

\begin{equation}
\max_i \|(\hat{X}_n)_i - Q_n (X_n)_i \| \stackrel{a.s.}{\to} 0
\end{equation}

where $Q_n$ is a sequence of matrices in $\mathbb{O}(p, q)$, the indefinite orthogonal group of order $p, q$.

Given sufficiently large sample size $n$, the scaled ASE of affinity matrix $A$ drawn from a RDPG or GRDPG will asymptotically approach the original latent positions $X$ with probability 1, up to a linear transformation (orthogonal transformation for the RDPG, a composition of an orthogonal transformation and scale transformation for the GRDPG). Thus if $X$ consists of points that lie on subspaces of $\mathbb{R}^d$, then both the scaled and unscaled ASE of $A \sim \text{RDPG}(X)$ or $A \sim \text{GRDPG}(X)$ will consist of points that lie near subspaces, with some noise that almost surely goes to $0$ as $n \to \infty$, motivating ASE followed by SSC as an asymptotically consistent method for community detection. The Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative graph model with underlying communities such that each community lies on a subspace. \citet{noroozi2019estimation} showed that SSC is able to recover the subspaces and therefore perform community detection for the PABM given $P = X I_{p, q} X^\top$, the edge probability matrix, rather than $A$, the adjacency matrix. Combining the results of \citeauthor{rubindelanchy2017statistical} and \citeauthor{jmlr-v28-wang13}, it should be possible to recover the communities using $A$ as well.

## Manifold Learning

In addition to subspaces, \citet{trosset2020learning} showed that the ASE of a RDPG can be used to recover one-dimensional manifolds. Suppose $f : [0, 1] \mapsto \mathcal{X}$ such that $f$ is smooth and $\mathcal{X}$ represents a curve or one-dimensional manifold in $\mathbb{R}^d$. If $t_1, ..., t_n \stackrel{iid}{\sim} F$ such that $F$ has support $[0, 1]$, the latent positions are $x_i = f(t_i)$ with $y_i$ is its corresponding point in the scaled ASE, and $d_{\epsilon}(\cdot, \cdot)$ is the shortest path distance of an $\epsilon$-neighborhood graph. Under certain mild conditions, the shortest path distances of the $\epsilon$-neighborhood graph of the ASE approaches the arc lengths along $f$:

\begin{equation}
d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{1 - f'(t)} dt
\end{equation}

This suggests that if the latent positions of a RDPG consists of $K$ disjoint manifolds, the $\epsilon$-neighborhood graph will consist of $K$ disjoint subgraphs as $\epsilon \to 0$ and $n \to \infty$.

# Proposed Research

## Subspace Clustering

As discussed in \S 2.2, if the latent positions of a RDPG or GRDPG model are such that they lie on a small number of subspaces, ASE followed by SSC may be able to identify whether vertices of the graph $v_i, v_j$ belong to the same subspace or to different subspaces, and one such model that is consistent with this construction is the PABM.

### Popularity Adjusted Block Model

We will first define the PABM.

\begin{definition}[Popularity Adjusted Block Model]
\label{pabm}
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl)}$ describes the edge probabilities between vertices in communities 
$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $\Bernoulli(P)$, i.e., 
$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

If each block $P^{(kl)}$ can be written as the outer product of two vectors:

\begin{equation} \label{eq:pabm}
  P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top
\end{equation}

for a set of $K^2$ fixed vectors $\{\lambda^{(st)}\}_{s, t = 1}^K$ where each 
$\lambda^{(st)}$ is a column vector 
of dimension $n_s$, then graph $G$ and its corresponding adjacency matrix $A$ 
is a realization of a popularity adjusted block model with parameters 
$\{\lambda^{(st)}\}_{s, t = 1}^K$. 
\end{definition}

We will use the notation $A \sim \text{PABM}(\{\lambda^{(kl)}\}_K)$ to denote a random 
adjacency matrix $A$ drawn from a PABM with parameters $\lambda^{(kl)}$ 
consisting of $K$ underlying communities.

It is trivial to show that the PABM, as well as all graphs such that the adjacency matrix is drawn such that $A_{ij} \sim Bernoulli(P_{ij})$, is a special case of the GRDPG. It can also be shown that the latent positions of the PABM under the GRDPG framework consists of $K$ $K$-dimensional subspaces in $\mathbb{R}^{K^2}$. While there is no unique latent configuration $X$ such that $X X^\top = P$, the edge probability $P$ for the PABM, they all have this subspace structure, and one in particular consists of *orthogonal* subspaces.

\begin{theorem}[Connecting the PABM to the GRDPG for $K = 2$]
\label{theorem1}  
Let

$$X = \begin{bmatrix} 
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\ 
0 & 0 & \lambda^{(21)} & \lambda^{(22)} 
\end{bmatrix}$$

$$U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix}$$

where each $\lambda^{(kl)}$ is a vector as in Definition 1. 
Then $A \sim GRDPG_{3, 1}(X U)$ and $A \sim PABM(\{(\lambda^{(kl)}\}_2)$ are 
equivalent.
\end{theorem}

\begin{theorem}[Generalization to $K > 2$] 
\label{theorem2}
There exists a block diagonal matrix 
$X \in \mathbb{R}^{n \times K^2}$ defined by PABM parameters 
$\{\lambda^{(kl)}\}_K$ and orthonormal matrix 
$U \in \mathbb{R}^{K^2 \times K^2}$ that is fixed 
for each $K$ such that $A \sim GRDPG_{K (K+1) / 2, K (K-1) / 2}(XU)$ and 
$A \sim PABM(\{(\lambda^{(kl)}\})_K)$ are equivalent.
\end{theorem}

\begin{proof}
Define the following matrices from $\{\lambda^{(kl)}\}_K$:

$$\Lambda^{(k)} = 
\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}$$

\begin{equation} \label{eq:xy}
X = \blockdiag(\Lambda^{(1)}, ..., \Lambda^{(K)}) \in \mathbb{R}^{n \times K^2}
\end{equation}

$$L^{(k)} = \blockdiag(\lambda^{(1k)}, ..., \lambda^{(Kk)}) \in 
\mathbb{R}^{n \times K}$$

$$Y = \begin{bmatrix} L^{(1)} & \cdots & L^{(K)} \end{bmatrix} \in 
\mathbb{R}^{n \times K^2}$$

Then $P = X Y^\top$.

Similar to the $K = 2$ case, we have $Y = X \Pi$ for a permutation matrix
$\Pi$, resulting in $P = X \Pi X^\top$.  
The permutation described by $\Pi$ has $K$ fixed points, which correspond to 
$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where 
$k = r (K + 1) + 1$ for $r = 0, ..., K - 1$. It also has 
$\binom{K}{2} = K (K - 1) / 2$ cycles of order $2$. Each cycle corresponds to 
a pair of eigenvalues $+1$ and $-1$ and a pair of eigenvectors 
$(e_s + e_t) / \sqrt{2}$ and $(e_s - e_t) / \sqrt{2}$.

Then $\Pi$ has $K (K + 1) / 2$ eigenvalues equal to $1$ and $K (K - 1) / 2$ 
eigenvalues equal to $-1$. $\Pi$ has the decomposed form

\begin{equation} \label{eq:permutation}
\Pi = U I_{K (K + 1) / 2, K (K - 1) / 2} U^\top
\end{equation}

The edge probability matrix then can be written as:

\begin{equation} \label{eq:pabm-grdpg}
P = X U I_{p, q} (X U)^\top
\end{equation}

\begin{equation} \label{eq:p}
p = K (K + 1) / 2
\end{equation}

\begin{equation} \label{eq:q}
q = K (K - 1) / 2
\end{equation}

and we can describe the PABM with $K$ communities as a GRDPG with latent 
positions $X U$ with signature $\big( K (K + 1) / 2, K (K - 1) / 2 \big)$.
\end{proof}

\begin{example}[$K = 3$] Using the same notation as in Theorem \ref{theorem2}:

$$X = \begin{bmatrix} 
\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
\end{bmatrix}$$

$$Y = \begin{bmatrix} 
\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
\end{bmatrix}$$

Then $P = X Y^\top$ and $Y = X \Pi$ where $\Pi$ is a permutation matrix 
consisting of $3$ fixed points and $3$ cycles of order 2:

$$\Pi = \begin{bmatrix} 
1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1
\end{bmatrix}$$

* Positions 1, 5, 9 are fixed.

* The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.

Therefore, we can decompose $\Pi = U I_{6, 3} U^\top$ where the first three 
columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the fixed 
positions $1$, $5$, and $9$, the next three columns consist of eigenvectors 
$(e_k + e_l) / \sqrt{2}$, and the last three columns consist of eigenvectors 
$(e_k - e_l) / \sqrt{2}$, where pairs $(k, l)$ correspond to the cycles of 
order 2 described above.

The latent positions are the rows of  
$$XU = \begin{bmatrix}
  \lambda^{(11)} & 0 & 0 & 
  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 & 
  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 \\
  0 & \lambda^{(22)} & 0 & 
  \lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} & 
  -\lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} \\
  0 & 0 & \lambda^{(33)} & 
  0 & \lambda^{(31)} / \sqrt{2} & \lambda^{(32)} / \sqrt{2} & 
  0 & -\lambda^{(31)} / \sqrt{2} & -\lambda^{(32)} / \sqrt{2}
\end{bmatrix}$$
\end{example}

This leads to the following theorem.

\begin{theorem}
\label{osc}
Let $P = V \Lambda V^\top$ be the spectral decomposition of the edge probability matrix of a PABM. Define $B = n V V^\top$. Then $B_{ij} = 0$ $\forall i, j$ in different communities. 
\end{theorem}

If $\hat{V}$ is the *unscaled* ASE of $A$, Theorem \ref{osc} and results from  \citeauthor{rubindelanchy2017statistical} together imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} 0$, leading to the following result:

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Community assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs  
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$.\;
  \caption{Orthogonal Spectral Clustering.}
\end{algorithm}

\begin{theorem}
\label{theorem4} 
Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,

\begin{equation} \label{eq:thm4}
\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
\end{equation}

This provides the result that $\forall i, j$ in different communities, 
$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
\end{theorem}

Since every ASE of the PABM consists of subspaces and as $n \to \infty$ each point of the ASE approaches its subspace almost surely, SSC should also work for PABM community detection. Combining this with the results by \citeauthor{jmlr-v28-wang13}, which state that if the points lie sufficiently close to their respective subspaces and the cosine of the angles between subspaces is sufficiently small, SDP will hold. We show that the unscaled ASE exhibits exactly these conditions for sufficiently large $n$.

\begin{theorem}
\label{theorem5}
Let $P_n$ describe the edge probability matrix of the PABM with 
$n$ vertices, and let $A_n \sim \Bernoulli(P_n)$.  Let $\hat{V}_n$ be the 
matrix of eigenvectors of $A_n$ corresponding to the $K (K + 1) / 2$ most
positive and $K (K - 1) / 2$ most negative eigenvalues. Then 
$\exists \lambda > 0$ and $N \in \mathbb{N}$ such that when $n > N$, 
$\sqrt{n} \hat{V}_n$ obeys the subspace detection property with probability 1.
\end{theorem}

## Manifold Clustering

We would like to extend subspace clustering to manifold clustering. The problem setup is as follows: Suppose that in the latent space $\mathcal{X} \subset \mathbb{R}^d$, sample $X$ of $n$ points lie on a union of $K$ disjoint manifolds with each manifold corresponding to a community. If $A \sim \text{RDPG}(X)$, we wish to recover the community labels (up to permutation) from $A$.

Similarly, suppose that probability distribution $F$ is described as follows:

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_i(t) \neq f_j(t)$ $\forall i \in [K]$ 
and $x \in [0, 1]$.
2. Sample labels
$z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
3. Sample $t_1, ..., t_n \stackrel{iid}{\sim} D$ where $D$ has support 
$[0, 1]$. 
4. Set latent positions $x_i = f_{z_i}(t_i)$ and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.

Then if $(A, X) \sim \text{RDPG}(F, n)$ and we observe $A$, we wish to recover hidden labels $z_1, ..., z_n$.

### Affine Subspaces

We will motivate an approach by the following example.

\begin{example}
Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ with order statistics 
$U_{(1)}, ..., U_{(n)}$. Then $\forall a \in (0, 1)$, $\exists N = N(\delta, a) < \infty$ such that $\forall n \geq N$, $P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq 1 - \delta / 2$.

This can be shown by the fact that $U_{(i+1)} - U_{(i)} \sim Beta(1, n)$. Then

\begin{equation}
P(U_{(i+1)} - U_{(i)} \leq a) = 1 - (1 - a)^n
\end{equation}

and

\begin{equation}
\label{eq:unsolvable}
P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq (P(U_{(i+1)} - U_{(i)} \leq a))^n = (1 - (1 - a)^n)^n
\end{equation}

This expression is monotone increasing $\forall n \geq N_1$ for some $N_1 < \infty$. 
Setting $(1 - (1 - a)^{N_2})^{N_2} \geq 1 - \delta / 2$, we can solve for a finite ${N_2}$. Then $N = \max(N_1, N_2)$.
\end{example}

If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, 1]$, then a sample of size $N(\delta, a)$ is sufficient to satisfy $P(\max_i \min_{i, j} X_{(i+1)} - X_{(i)} \leq \|X_i - Y_j\|) \geq 1 - \delta$ for $X_i$ in the first segment and $Y_j$ in the second segment and $X_{(i)}$ are order statistics in the first coordinate. If each segment corresponds to a community, this leads to the following two results: Single linkage clustering with $K = 2$ will perform perfect community detection with probability at least $1 - \delta$, and an $\epsilon$-neighborhood graph with $\epsilon \in (0, a)$ will consists of at least 2 disjoint subgraphs such that no subgraph consists of members of two different communities (which is analogous to the SDP), with probability at least $1 - \delta$.

### Mainfold Learning

If instead of sampling uniformly from line segments of unit length, we sample uniformly from a 1 dimensional manifolds of unit length, the above property still holds. Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_0^1 \sqrt{1 - f'(t)} dt = 1$. Then $U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} < \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$. If the shortest distance between the two manifolds is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.

# Summary

# Estimated Timeline of Completion

Literature review: August 2021

Complete proofs of main theorems: January 2022

Simulations and real data analyses: March 2022

Dissertation completion: April 2022

\newpage

# References

### 2021-05-05 proposal/proposal.tex
started proposal

--- 2021-05-05 proposal/proposal.Rmd
@@
----
-output:
-  pdf_document:
-    includes:
-      before_body: title.sty
-    keep_tex: true
-    citation_package: natbib
-    number_sections: yes
-# output:
-#   bookdown::pdf_document2:
-#     citation_package: natbib
-#     number_citations: yes
-# output: html_document
-# output: rticles::rss_article
-fontsize: 11pt
-# geometry: "left=1in,right=1in,top=1in,bottom=1in"
-urlcolor: blue
-header-includes:
-- \usepackage{setspace}
-- \usepackage{float}
-- \usepackage{mathtools}
-- \usepackage{natbib}
-- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
-- \setcitestyle{numbers,square,comma}
-- \usepackage{verbatim}
-- \usepackage{amsthm}
-- \usepackage{comment}
-bibliography: proposal.bib
-abstract: |
-  asdf
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+%
+\documentclass[
+  11pt,
+]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  colorlinks=true,
+  linkcolor=Maroon,
+  filecolor=Maroon,
+  citecolor=Blue,
+  urlcolor=blue,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+\usepackage{setspace}
+\usepackage{float}
+\usepackage{mathtools}
+\usepackage{natbib}
+\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
+\setcitestyle{numbers,square,comma}
+\usepackage{verbatim}
+\usepackage{amsthm}
+\usepackage{comment}
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
 
-```{r setup, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE, 
-                      # eval = FALSE,
-                      comment = NA, 
-                      warning = FALSE, 
-                      message = FALSE, 
-                      fig.pos = 'H', 
-                      fig.align = 'center', 
-                      fig.height = 3, 
-                      fig.width = 5, 
-                      fig.dpi = 300)
+\author{}
+\date{\vspace{-2.5em}}
 
-options(xtable.comment = FALSE, 
-        xtable.table.placement = 'H')
-```
+\begin{document}
+
+\pagenumbering{gobble}
+
+%\begin{titlepage}
+\begin{center}
+\LARGE{\textbf{Community Detection on Subspaces and Manifolds}}\\
+\vspace*{2\baselineskip}
+\normalsize{A dissertation proposal submitted in partial satisfaction of the requirements for the degree of \\}
+Doctor of Philosophy \\
+in \\
+Statistical Science \\
+\vspace*{2\baselineskip}
+\Large{John Koo}\\
+\vspace*{3\baselineskip}
+\Large{\textbf{Research Committee Members}}\\
+Dr. Michael Trosset \\
+Dr. Minh Tang \\
+Dr. Julia Fukuyama \\
+Dr. Roni Khardon \\
+Dr. Fangzheng Xie \\
+\vspace*{3\baselineskip}
+Date TBA \\
+\vspace*{1\baselineskip}
+Department of Statistics \\
+Indiana University \\
+Bloomington, Indiana \\
+\end{center}
+% \end{titlepage}
+
+\hypersetup{linkcolor = black}
+\newpage
+\pagenumbering{roman}
+\tableofcontents
+\addcontentsline{toc}{section}{\contentsname}
+
+\newpage
+
+\newpage
+\pagenumbering{arabic}
+\hypersetup{linkcolor = blue}
 
@@
 
-# Research Goal
+\hypertarget{research-goal}{%
+\section{Research Goal}\label{research-goal}}
 
-Typical clustering methods often work by assuming that each cluster is represented by some central point around which the data belonging to that cluster lie Such methods, including Lloyd's algorithm for $k$-means clustering \cite{1056489} and Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}, involve iteratively updating the central points by averaging data belonging to each cluster and cluster memberships by comparing proximity to central points.
-The aim of our research is to explore clustering and community detection methods for which each cluster is represented not by a central point but by subspaces or manifolds.
-In addition, we aim to explore various applications of such methods and connect them to preexisting clustering and community detection problems.
+Typical clustering methods often work by assuming that each cluster is
+represented by some central point around which the data belonging to
+that cluster lie Such methods, including Lloyd's algorithm for
+\(k\)-means clustering \cite{1056489} and Gaussian Mixture Models
+\cite{doi:10.1198/016214502760047131}, involve iteratively updating the
+central points by averaging data belonging to each cluster and cluster
+memberships by comparing proximity to central points. The aim of our
+research is to explore clustering and community detection methods for
+which each cluster is represented not by a central point but by
+subspaces or manifolds. In addition, we aim to explore various
+applications of such methods and connect them to preexisting clustering
+and community detection problems.
 
-# Literature Review
+\hypertarget{literature-review}{%
+\section{Literature Review}\label{literature-review}}
 
-## Subspace Clustering
+\hypertarget{subspace-clustering}{%
+\subsection{Subspace Clustering}\label{subspace-clustering}}
 
-Subspace clustering, which assumes that points in $\mathbb{R}^d$ each lie on one of $K$ subspaces of $\mathbb{R}^d$, is an approach that has found a wide range of uses by the Statistics and Machine Learning communities, particularly within the field of Computer Vision \cite{5206547}.
+Subspace clustering, which assumes that points in \(\mathbb{R}^d\) each
+lie on one of \(K\) subspaces of \(\mathbb{R}^d\), is an approach that
+has found a wide range of uses by the Statistics and Machine Learning
+communities, particularly within the field of Computer Vision
+\cite{5206547}.
 
-Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample.
-Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
-The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$.
-If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces.
-Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.
+Of particular interest is Sparse Subspace Clustering (SSC), which is
+performed by solving an optimization problem for each observed point in
+a sample. Given \(X \in \mathbb{R}^{n \times d}\) with vectors
+\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
+\(c_i = \min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
+\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
+collected into matrix
+\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
+construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
+perfectly on one of \(K\) subspaces, \(B\) is sparse such that
+\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
+Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
+and if the number of subgraphs is exactly \(K\), each subgraph maps onto
+a subspace.
 
@@
 
-for some sparsity parameter $\lambda > 0$.
-The $c_i$ vectors are then collected into $C$ and $B$ as described before.
-If $X$ is noisy in that each $x_i$ does not lie exactly on one of $K$ subspaces but near it, the choice of $\lambda$ becomes important in guaranteeing the Subspace Detection Property (SDP) \cite{jmlr-v28-wang13}.
+for some sparsity parameter \(\lambda > 0\). The \(c_i\) vectors are
+then collected into \(C\) and \(B\) as described before. If \(X\) is
+noisy in that each \(x_i\) does not lie exactly on one of \(K\)
+subspaces but near it, the choice of \(\lambda\) becomes important in
+guaranteeing the Subspace Detection Property (SDP)
+\cite{jmlr-v28-wang13}.
 
@@
 
-## (Generalized) Random Dot Product Graphs
+\hypertarget{generalized-random-dot-product-graphs}{%
+\subsection{(Generalized) Random Dot Product
+Graphs}\label{generalized-random-dot-product-graphs}}
 
-It has been shown by \citet{jmlr-v28-wang13} that given sufficiently low noise in $X$ and sufficiently low affinity between pairs of subspaces as measured by the cosine of the angle between subspaces, SDP holds for a specific range of $\lambda$.
-This provides theoretical justification for applying SSC to Random Dot Product Graphs (RDPG) \cite{athreya2017statistical} and Generalized Random Dot Product Graphs (GRDPG) \cite{rubindelanchy2017statistical} for which the data in the latent space lie on subspaces.
+It has been shown by \citet{jmlr-v28-wang13} that given sufficiently low
+noise in \(X\) and sufficiently low affinity between pairs of subspaces
+as measured by the cosine of the angle between subspaces, SDP holds for
+a specific range of \(\lambda\). This provides theoretical justification
+for applying SSC to Random Dot Product Graphs (RDPG)
+\cite{athreya2017statistical} and Generalized Random Dot Product Graphs
+(GRDPG) \cite{rubindelanchy2017statistical} for which the data in the
+latent space lie on subspaces.
 
@@
 
-\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$, if $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,
+\citeauthor{athreya2017statistical} showed that under mild conditions,
+if \((A_n, X_n) \sim \text{RDPG}(F, n)\), if \(\hat{X}_n\) is the scaled
+ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),
 
@@
 
-Similarly, \citeauthor{rubindelanchy2017statistical} showed that for $(A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)$,
+Similarly, \citeauthor{rubindelanchy2017statistical} showed that for
+\((A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)\),
 
@@
 
-where $Q_n$ is a sequence of matrices in $\mathbb{O}(p, q)$, the indefinite orthogonal group of order $p, q$.
+where \(Q_n\) is a sequence of matrices in \(\mathbb{O}(p, q)\), the
+indefinite orthogonal group of order \(p, q\).
 
-Given sufficiently large sample size $n$, the scaled ASE of affinity matrix $A$ drawn from a RDPG or GRDPG will asymptotically approach the original latent positions $X$ with probability 1, up to a linear transformation (orthogonal transformation for the RDPG, a composition of an orthogonal transformation and scale transformation for the GRDPG).
-Thus if $X$ consists of points that lie on subspaces of $\mathbb{R}^d$, then both the scaled and unscaled ASE of $A \sim \text{RDPG}(X)$ or $A \sim \text{GRDPG}(X)$ will consist of points that lie near subspaces, with some noise that almost surely goes to $0$ as $n \to \infty$, motivating ASE followed by SSC as an asymptotically consistent method for community detection.
-The Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative graph model with underlying communities such that each community lies on a subspace.
-\citet{noroozi2019estimation} showed that SSC is able to recover the subspaces and therefore perform community detection for the PABM given $P = X I_{p, q} X^\top$, the edge probability matrix, rather than $A$, the adjacency matrix.
-Combining the results of \citeauthor{rubindelanchy2017statistical} and \citeauthor{jmlr-v28-wang13}, it should be possible to recover the communities using $A$ as well.
+Given sufficiently large sample size \(n\), the scaled ASE of affinity
+matrix \(A\) drawn from a RDPG or GRDPG will asymptotically approach the
+original latent positions \(X\) with probability 1, up to a linear
+transformation (orthogonal transformation for the RDPG, a composition of
+an orthogonal transformation and scale transformation for the GRDPG).
+Thus if \(X\) consists of points that lie on subspaces of
+\(\mathbb{R}^d\), then both the scaled and unscaled ASE of
+\(A \sim \text{RDPG}(X)\) or \(A \sim \text{GRDPG}(X)\) will consist of
+points that lie near subspaces, with some noise that almost surely goes
+to \(0\) as \(n \to \infty\), motivating ASE followed by SSC as an
+asymptotically consistent method for community detection. The Popularity
+Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a
+generative graph model with underlying communities such that each
+community lies on a subspace. \citet{noroozi2019estimation} showed that
+SSC is able to recover the subspaces and therefore perform community
+detection for the PABM given \(P = X I_{p, q} X^\top\), the edge
+probability matrix, rather than \(A\), the adjacency matrix. Combining
+the results of \citeauthor{rubindelanchy2017statistical} and
+\citeauthor{jmlr-v28-wang13}, it should be possible to recover the
+communities using \(A\) as well.
 
-## Manifold Learning
+\hypertarget{manifold-learning}{%
+\subsection{Manifold Learning}\label{manifold-learning}}
 
-In addition to subspaces, \citet{trosset2020learning} showed that the ASE of a RDPG can be used to recover one-dimensional manifolds.
-Suppose $f : [0, 1] \mapsto \mathcal{X}$ such that $f$ is smooth and $\mathcal{X}$ represents a curve or one-dimensional manifold in $\mathbb{R}^d$.
-If $t_1, ..., t_n \stackrel{iid}{\sim} F$ such that $F$ has support $[0, 1]$, the latent positions are $x_i = f(t_i)$ with $y_i$ is its corresponding point in the scaled ASE, and $d_{\epsilon}(\cdot, \cdot)$ is the shortest path distance of an $\epsilon$-neighborhood graph.
-Under certain mild conditions, the shortest path distances of the $\epsilon$-neighborhood graph of the ASE approaches the arc lengths along $f$:
+In addition to subspaces, \citet{trosset2020learning} showed that the
+ASE of a RDPG can be used to recover one-dimensional manifolds. Suppose
+\(f : [0, 1] \mapsto \mathcal{X}\) such that \(f\) is smooth and
+\(\mathcal{X}\) represents a curve or one-dimensional manifold in
+\(\mathbb{R}^d\). If \(t_1, ..., t_n \stackrel{iid}{\sim} F\) such that
+\(F\) has support \([0, 1]\), the latent positions are \(x_i = f(t_i)\)
+with \(y_i\) is its corresponding point in the scaled ASE, and
+\(d_{\epsilon}(\cdot, \cdot)\) is the shortest path distance of an
+\(\epsilon\)-neighborhood graph. Under certain mild conditions, the
+shortest path distances of the \(\epsilon\)-neighborhood graph of the
+ASE approaches the arc lengths along \(f\):
 
@@
 
-This suggests that if the latent positions of a RDPG consists of $K$ disjoint manifolds, the $\epsilon$-neighborhood graph will consist of $K$ disjoint subgraphs as $\epsilon \to 0$ and $n \to \infty$.
+This suggests that if the latent positions of a RDPG consists of \(K\)
+disjoint manifolds, the \(\epsilon\)-neighborhood graph will consist of
+\(K\) disjoint subgraphs as \(\epsilon \to 0\) and \(n \to \infty\).
 
-# Proposed Research
+\hypertarget{proposed-research}{%
+\section{Proposed Research}\label{proposed-research}}
 
-## Subspace Clustering
+\hypertarget{subspace-clustering-1}{%
+\subsection{Subspace Clustering}\label{subspace-clustering-1}}
 
-As discussed in \S 2.2, if the latent positions of a RDPG or GRDPG model are such that they lie on a small number of subspaces, ASE followed by SSC may be able to identify whether vertices of the graph $v_i, v_j$ belong to the same subspace or to different subspaces, and one such model that is consistent with this construction is the PABM.
+As discussed in \S 2.2, if the latent positions of a RDPG or GRDPG model
+are such that they lie on a small number of subspaces, ASE followed by
+SSC may be able to identify whether vertices of the graph \(v_i, v_j\)
+belong to the same subspace or to different subspaces, and one such
+model that is consistent with this construction is the PABM.
 
-### Popularity Adjusted Block Model
+\hypertarget{popularity-adjusted-block-model}{%
+\subsubsection{Popularity Adjusted Block
+Model}\label{popularity-adjusted-block-model}}
 
@@
 that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
-realization of $\Bernoulli(P)$, i.e., 
-$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ 
+realization of $\text{Bernoulli}(P)$, i.e., 
+$A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ 
 ($A_{ij} = A_{ji}$ and $A_{ii} = 0$).
@@
 
-We will use the notation $A \sim \text{PABM}(\{\lambda^{(kl)}\}_K)$ to denote a random 
-adjacency matrix $A$ drawn from a PABM with parameters $\lambda^{(kl)}$ 
-consisting of $K$ underlying communities.
+We will use the notation \(A \sim \text{PABM}(\{\lambda^{(kl)}\}_K)\) to
+denote a random adjacency matrix \(A\) drawn from a PABM with parameters
+\(\lambda^{(kl)}\) consisting of \(K\) underlying communities.
 
-It is trivial to show that the PABM, as well as all graphs such that the adjacency matrix is drawn such that $A_{ij} \sim Bernoulli(P_{ij})$, is a special case of the GRDPG.
-It can also be shown that the latent positions of the PABM under the GRDPG framework consists of $K$ $K$-dimensional subspaces in $\mathbb{R}^{K^2}$.
-While there is no unique latent configuration $X$ such that $X X^\top = P$, the edge probability $P$ for the PABM, they all have this subspace structure, and one in particular consists of *orthogonal* subspaces.
+It is trivial to show that the PABM, as well as all graphs such that the
+adjacency matrix is drawn such that \(A_{ij} \sim Bernoulli(P_{ij})\),
+is a special case of the GRDPG. It can also be shown that the latent
+positions of the PABM under the GRDPG framework consists of \(K\)
+\(K\)-dimensional subspaces in \(\mathbb{R}^{K^2}\). While there is no
+unique latent configuration \(X\) such that \(X X^\top = P\), the edge
+probability \(P\) for the PABM, they all have this subspace structure,
+and one in particular consists of \emph{orthogonal} subspaces.
 
@@
 \begin{equation} \label{eq:xy}
-X = \blockdiag(\Lambda^{(1)}, ..., \Lambda^{(K)}) \in \mathbb{R}^{n \times K^2}
+X = \text{blockdiag}(\Lambda^{(1)}, ..., \Lambda^{(K)}) \in \mathbb{R}^{n \times K^2}
 \end{equation}
 
-$$L^{(k)} = \blockdiag(\lambda^{(1k)}, ..., \lambda^{(Kk)}) \in 
+$$L^{(k)} = \text{blockdiag}(\lambda^{(1k)}, ..., \lambda^{(Kk)}) \in 
 \mathbb{R}^{n \times K}$$
@@
 
-If $\hat{V}$ is the *unscaled* ASE of $A$, Theorem \ref{osc} and results from  \citeauthor{rubindelanchy2017statistical} together imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} 0$, leading to the following result:
+If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), Theorem \ref{osc}
+and results from \citeauthor{rubindelanchy2017statistical} together
+imply \(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} 0\), leading to the
+following result:
 
@@
 
-Since every ASE of the PABM consists of subspaces and as $n \to \infty$ each point of the ASE approaches its subspace almost surely, SSC should also work for PABM community detection.
-Combining this with the results by \citeauthor{jmlr-v28-wang13}, which state that if the points lie sufficiently close to their respective subspaces and the cosine of the angles between subspaces is sufficiently small, SDP will hold.
-We show that the unscaled ASE exhibits exactly these conditions for sufficiently large $n$.
+Since every ASE of the PABM consists of subspaces and as
+\(n \to \infty\) each point of the ASE approaches its subspace almost
+surely, SSC should also work for PABM community detection. Combining
+this with the results by \citeauthor{jmlr-v28-wang13}, which state that
+if the points lie sufficiently close to their respective subspaces and
+the cosine of the angles between subspaces is sufficiently small, SDP
+will hold. We show that the unscaled ASE exhibits exactly these
+conditions for sufficiently large \(n\).
 
@@
 Let $P_n$ describe the edge probability matrix of the PABM with 
-$n$ vertices, and let $A_n \sim \Bernoulli(P_n)$.  Let $\hat{V}_n$ be the 
+$n$ vertices, and let $A_n \sim \text{Bernoulli}(P_n)$.  Let $\hat{V}_n$ be the 
 matrix of eigenvectors of $A_n$ corresponding to the $K (K + 1) / 2$ most
@@
 
-## Manifold Clustering
+\hypertarget{manifold-clustering}{%
+\subsection{Manifold Clustering}\label{manifold-clustering}}
 
-We would like to extend subspace clustering to manifold clustering.
-The problem setup is as follows: Suppose that in the latent space $\mathcal{X} \subset \mathbb{R}^d$, sample $X$ of $n$ points lie on a union of $K$ disjoint manifolds with each manifold corresponding to a community.
-If $A \sim \text{RDPG}(X)$, we wish to recover the community labels (up to permutation) from $A$.
+We would like to extend subspace clustering to manifold clustering. The
+problem setup is as follows: Suppose that in the latent space
+\(\mathcal{X} \subset \mathbb{R}^d\), sample \(X\) of \(n\) points lie
+on a union of \(K\) disjoint manifolds with each manifold corresponding
+to a community. If \(A \sim \text{RDPG}(X)\), we wish to recover the
+community labels (up to permutation) from \(A\).
 
-Similarly, suppose that probability distribution $F$ is described as follows:
+Similarly, suppose that probability distribution \(F\) is described as
+follows:
 
-1. Define functions $f_1, ..., f_K$ such that 
-$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_i(t) \neq f_j(t)$ $\forall i \in [K]$ 
-and $x \in [0, 1]$.
-2. Sample labels
-$z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
-3. Sample $t_1, ..., t_n \stackrel{iid}{\sim} D$ where $D$ has support 
-$[0, 1]$. 
-4. Set latent positions $x_i = f_{z_i}(t_i)$ and 
-$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Define functions \(f_1, ..., f_K\) such that
+  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_i(t) \neq f_j(t)\)
+  \(\forall i \in [K]\) and \(x \in [0, 1]\).
+\item
+  Sample labels
+  \(z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)\).
+\item
+  Sample \(t_1, ..., t_n \stackrel{iid}{\sim} D\) where \(D\) has
+  support \([0, 1]\).
+\item
+  Set latent positions \(x_i = f_{z_i}(t_i)\) and
+  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
+\end{enumerate}
 
-Then if $(A, X) \sim \text{RDPG}(F, n)$ and we observe $A$, we wish to recover hidden labels $z_1, ..., z_n$.
+Then if \((A, X) \sim \text{RDPG}(F, n)\) and we observe \(A\), we wish
+to recover hidden labels \(z_1, ..., z_n\).
 
-### Affine Subspaces
+\hypertarget{affine-subspaces}{%
+\subsubsection{Affine Subspaces}\label{affine-subspaces}}
 
@@
 
-If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, 1]$, then a sample of size $N(\delta, a)$ is sufficient to satisfy $P(\max_i \min_{i, j} X_{(i+1)} - X_{(i)} \leq \|X_i - Y_j\|) \geq 1 - \delta$ for $X_i$ in the first segment and $Y_j$ in the second segment and $X_{(i)}$ are order statistics in the first coordinate.
-If each segment corresponds to a community, this leads to the following two results: Single linkage clustering with $K = 2$ will perform perfect community detection with probability at least $1 - \delta$, and an $\epsilon$-neighborhood graph with $\epsilon \in (0, a)$ will consists of at least 2 disjoint subgraphs such that no subgraph consists of members of two different communities (which is analogous to the SDP), with probability at least $1 - \delta$.
+If we extend this example such that \(n_1\) points are sampled uniformly
+from the segment \(f_1(t) = (t, 0)\) and \(n_2\) points are sampled
+uniformly from the segment \(f_2(t) = (t, a)\) for \(t \in [0, 1]\),
+then a sample of size \(N(\delta, a)\) is sufficient to satisfy
+\(P(\max_i \min_{i, j} X_{(i+1)} - X_{(i)} \leq \|X_i - Y_j\|) \geq 1 - \delta\)
+for \(X_i\) in the first segment and \(Y_j\) in the second segment and
+\(X_{(i)}\) are order statistics in the first coordinate. If each
+segment corresponds to a community, this leads to the following two
+results: Single linkage clustering with \(K = 2\) will perform perfect
+community detection with probability at least \(1 - \delta\), and an
+\(\epsilon\)-neighborhood graph with \(\epsilon \in (0, a)\) will
+consists of at least 2 disjoint subgraphs such that no subgraph consists
+of members of two different communities (which is analogous to the SDP),
+with probability at least \(1 - \delta\).
 
-### Mainfold Learning
+\hypertarget{mainfold-learning}{%
+\subsubsection{Mainfold Learning}\label{mainfold-learning}}
 
-If instead of sampling uniformly from line segments of unit length, we sample uniformly from a 1 dimensional manifolds of unit length, the above property still holds.
-Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_0^1 \sqrt{1 - f'(t)} dt = 1$.
-Then $U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} < \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.
-If the shortest distance between the two manifolds is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.
+If instead of sampling uniformly from line segments of unit length, we
+sample uniformly from a 1 dimensional manifolds of unit length, the
+above property still holds. Let
+\(U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)\) and
+\(f : [0, 1] \mapsto \mathbb{R}^d\) be a smooth function such that
+\(\int_0^1 \sqrt{1 - f'(t)} dt = 1\). Then
+\(U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
+\(P(U_{(i+1)} - U_{(i)} < \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
+If the shortest distance between the two manifolds is \(a\), then the
+same \(N\) as before is sufficient, although perhaps a more lenient
+lower bound can be derived based on the shape of \(f_i(\cdot)\).
 
-# Summary
+\hypertarget{summary}{%
+\section{Summary}\label{summary}}
 
-# Estimated Timeline of Completion
+\hypertarget{estimated-timeline-of-completion}{%
+\section{Estimated Timeline of
+Completion}\label{estimated-timeline-of-completion}}
 
@@
 
-# References
+\renewcommand\refname{References}
+  \bibliography{proposal.bib}
+
+\end{document}

### 2021-05-05 proposal/proposal.Rmd
typos

@@
 Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ with order statistics 
-$U_{(1)}, ..., U_{(n)}$. Then $\forall a \in (0, 1)$, $\exists N = N(\delta, a) < \infty$ such that $\forall n \geq N$, $P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq 1 - \delta / 2$.
+$U_{(1)}, ..., U_{(n)}$. Then $\forall a \in (0, 1)$ and $\delta \in (0, 1)$, $\exists N = N(\delta, a) < \infty$ such that $\forall n \geq N$,
 
-This can be shown by the fact that $U_{(i+1)} - U_{(i)} \sim Beta(1, n)$. Then
+\begin{equation}
+P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq 1 - \delta / 2
+\end{equation}
+
+Where $N(\delta, a)$ is monotone increasing w.r.t. $\delta$ and $a$. To prove this, we start with the fact that $U_{(i+1)} - U_{(i)} \sim Beta(1, n)$. Then
 
@@
 
-If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, 1]$, then a sample of size $N(\delta, a)$ is sufficient to satisfy $P(\max_i \min_{i, j} X_{(i+1)} - X_{(i)} \leq \|X_i - Y_j\|) \geq 1 - \delta$ for $X_i$ in the first segment and $Y_j$ in the second segment and $X_{(i)}$ are order statistics in the first coordinate.
-If each segment corresponds to a community, this leads to the following two results: Single linkage clustering with $K = 2$ will perform perfect community detection with probability at least $1 - \delta$, and an $\epsilon$-neighborhood graph with $\epsilon \in (0, a)$ will consists of at least 2 disjoint subgraphs such that no subgraph consists of members of two different communities (which is analogous to the SDP), with probability at least $1 - \delta$.
+If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, 1]$, then a sample of size $N(\delta, a)$ is sufficient to satisfy:
+
+\begin{equation}
+\begin{split}
+P(\max_i X_{(i+1)} - X_{(i)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta \\
+P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta
+\end{split}
+\end{equation}
+
+for $X_i$ in the first segment and $Y_j$ in the second segment and $X_{(i)}$, $Y_{(j)}$ are order statistics in the first coordinate.
+If each segment corresponds to a community, this leads to the following two results:
+
+1. Single linkage clustering with $K = 2$ will perform perfect community detection with probability at least $1 - \delta$.
+
+2.
+An $\epsilon$-neighborhood graph with $\epsilon \in (0, a)$ will consists of at least 2 disjoint subgraphs such that no subgraph consists of members of two different communities (analogous to the SDP), with probability at least $1 - \delta$.
+
+We can then further extend this to the case where points are drawn from unit segments with noise.
 
@@
 Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_0^1 \sqrt{1 - f'(t)} dt = 1$.
-Then $U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} < \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.
+Then $U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.
 If the shortest distance between the two manifolds is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.

### 2021-05-05 proposal/proposal.tex
typos

@@
 uniformly from the segment \(f_2(t) = (t, a)\) for \(t \in [0, 1]\),
-then a sample of size \(N(\delta, a)\) is sufficient to satisfy
-\(P(\max_i \min_{i, j} X_{(i+1)} - X_{(i)} \leq \|X_i - Y_j\|) \geq 1 - \delta\)
+then a sample of size \(N(\delta, a)\) is sufficient to satisfy:
+
+\begin{equation}
+\begin{split}
+P(\max_i X_{(i+1)} - X_{(i)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta \\
+P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta
+\end{split}
+\end{equation}
+
 for \(X_i\) in the first segment and \(Y_j\) in the second segment and
-\(X_{(i)}\) are order statistics in the first coordinate. If each
-segment corresponds to a community, this leads to the following two
-results: Single linkage clustering with \(K = 2\) will perform perfect
-community detection with probability at least \(1 - \delta\), and an
-\(\epsilon\)-neighborhood graph with \(\epsilon \in (0, a)\) will
-consists of at least 2 disjoint subgraphs such that no subgraph consists
-of members of two different communities (which is analogous to the SDP),
-with probability at least \(1 - \delta\).
+\(X_{(i)}\), \(Y_{(j)}\) are order statistics in the first coordinate.
+If each segment corresponds to a community, this leads to the following
+two results:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Single linkage clustering with \(K = 2\) will perform perfect
+  community detection with probability at least \(1 - \delta\).
+\item
+  An \(\epsilon\)-neighborhood graph with \(\epsilon \in (0, a)\) will
+  consists of at least 2 disjoint subgraphs such that no subgraph
+  consists of members of two different communities (analogous to the
+  SDP), with probability at least \(1 - \delta\).
+\end{enumerate}
+
+We can then further extend this to the case where points are drawn from
+unit segments with noise.
 
@@
 \(U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
-\(P(U_{(i+1)} - U_{(i)} < \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
+\(P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
 If the shortest distance between the two manifolds is \(a\), then the

### 2021-05-05 proposal/proposal.Rmd
typos

@@
 If instead of sampling uniformly from line segments of unit length, we sample uniformly from a 1 dimensional manifolds of unit length, the above property still holds.
-Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_0^1 \sqrt{1 - f'(t)} dt = 1$.
+Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_0^1 \sqrt{1 - f'(t)} dt = 1$ and $\int_u^v \sqrt{1 - f'(t)} dt = \|u - v\|$.
 Then $U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.

### 2021-05-05 proposal/proposal.Rmd
typos

--- 2021-05-05 proposal/proposal.Rmd
@@
 \begin{equation}
-d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{1 - f'(t)} dt
+d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{1  + (f'(t))^2} dt
 \end{equation}
@@
 If instead of sampling uniformly from line segments of unit length, we sample uniformly from a 1 dimensional manifolds of unit length, the above property still holds.
-Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_0^1 \sqrt{1 - f'(t)} dt = 1$.
+Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|$.
 Then $U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.
-If the shortest distance between the two manifolds is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.
+If the shortest distance between the two manifolds defined by $f_1$ and $f_2$ is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.
 

### 2021-05-05 proposal/proposal.tex
typos

@@
 \(f : [0, 1] \mapsto \mathbb{R}^d\) be a smooth function such that
-\(\int_0^1 \sqrt{1 - f'(t)} dt = 1\). Then
+\(\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|\). Then
 \(U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
 \(P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
-If the shortest distance between the two manifolds is \(a\), then the
-same \(N\) as before is sufficient, although perhaps a more lenient
-lower bound can be derived based on the shape of \(f_i(\cdot)\).
+If the shortest distance between the two manifolds defined by \(f_1\)
+and \(f_2\) is \(a\), then the same \(N\) as before is sufficient,
+although perhaps a more lenient lower bound can be derived based on the
+shape of \(f_i(\cdot)\).
 

### 2021-05-05 proposal/proposal.Rmd
typos

@@
 $f_k : [0, 1] \mapsto \mathcal{X}$ and $f_i(t) \neq f_j(t)$ $\forall i \in [K]$ 
-and $x \in [0, 1]$.
+and $t \in [0, 1]$.
 2. Sample labels
@@
 Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|$.
-Then $U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.
-If the shortest distance between the two manifolds defined by $f_1$ and $f_2$ is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.
+Then $U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.
+If the shortest distance between the two manifolds defined by $f_1$ and $f_2$ with the same restriction is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.
 

### 2021-05-05 proposal/proposal.tex
typos

@@
   \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_i(t) \neq f_j(t)\)
-  \(\forall i \in [K]\) and \(x \in [0, 1]\).
+  \(\forall i \in [K]\) and \(t \in [0, 1]\).
 \item
@@
 \(\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|\). Then
-\(U_{(i+1)} - U_{(i)} > \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
+\(U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
 \(P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
 If the shortest distance between the two manifolds defined by \(f_1\)
-and \(f_2\) is \(a\), then the same \(N\) as before is sufficient,
-although perhaps a more lenient lower bound can be derived based on the
-shape of \(f_i(\cdot)\).
+and \(f_2\) with the same restriction is \(a\), then the same \(N\) as
+before is sufficient, although perhaps a more lenient lower bound can be
+derived based on the shape of \(f_i(\cdot)\).
 

### 2021-05-10 proposal/proposal.Rmd
some more updates to proposal

@@
 
-# Research Goal
+# Introduction
 
-Typical clustering methods often work by assuming that each cluster is represented by some central point around which the data belonging to that cluster lie Such methods, including Lloyd's algorithm for $k$-means clustering \cite{1056489} and Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}, involve iteratively updating the central points by averaging data belonging to each cluster and cluster memberships by comparing proximity to central points.
-The aim of our research is to explore clustering and community detection methods for which each cluster is represented not by a central point but by subspaces or manifolds.
-In addition, we aim to explore various applications of such methods and connect them to preexisting clustering and community detection problems.
+## Research Goal
+
+Graph and network data have become increasingly more widespread in various fields including sociology, neuroscience, biostatistics, and computer science.
+This has resulted in certain challenges for researchers, as most traditional statistical and machine learning methods are incompatible with graphs and instead require the data to exist in a subset of Euclidean space.
+This includes clustering and community detection, which are the typical problems of interest when working with graphs.
+Common clustering methods often involve calculating some central or representative point around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}).
+Because these methods involve computing summary statistics within each cluster, they cannot be applied directly to graphs as community detection algorithms.
+
+Some proposed methods for unifying graph community detection with traditional clustering techniques involve embedding the graph into Euclidean space, of which the most popular is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}.
+The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models such that latent positions in Euclidean space are used to induce graphs.
+A community detection method motivated by this would involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.
+
+The aim of our research is to develop consistent community detection techniques under the RDPG and GRDPG frameworks.
+First, we explore existing generative graph models with underlying community structures that can be inferred by connecting the generative models to the RDPG or GRDPG.
+Then we explore other latent structures or mixture distributions in the latent space that induce graphs for which consistent community detection is possible.
+
+## Notation
+
+Let $G = (V, E)$ be an undirected, unweighted graph with no self-loops with $n$ vertices.
+Denote $A \in \{0, 1\}^{n \times n}$ as the adjacency matrix for $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $A_{ij} = 0$ otherwise.
+Because $G$ is symmetric and contains no self-loops, $A_{ij} = A_{ji}$ and $A_{ii} = 0$ for $i, j \in [n]$.
+We further restrict our analyses to independent Bernoulli graphs.
+Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities.
+Graph $G$ is sampled by $A_{ij} \indep P_{ij}$ for each $0 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$).
+Denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$, and denote $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.
 
 # Literature Review
+
+## Generative Graph Models and Community Detection
+
+Generative models for symmetric Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$.
+Community detection and recovery starts by restricting the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on the labels of the vertices.
+One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = B_{z_i z_j}$ where $B \in [0, 1]^{K \times K}$ with entries $B_{kl}$ denoting the edge probability of each vertex in community $k$ having an edge with each vertex in community $l$.
+The homogeneous SBM further restricts $P$ to two unique entries such that $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise.
+Multiple generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block Model (PABM).
+Like the SBM, these models involve edge probability matrix $P$ that is restricted to rank $d < n$ in part based on community labels.
+
+## (Generalized) Random Dot Product Graphs
+
+The RDPG starts with points in latent space $X \in \mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y \in [0, 1]$.
+$P$ is then constructed as $P = X X^\top$, and graph $G$ with adjacency matrix $A$ is drawn from $P$.
+We provide a more formal definition of the RDPG and GRDPG below.
+
+\begin{definition}[(Generalized) Random Dot Product Graph]
+Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y.\in [0, 1]$.
+$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix $A$ is drawn such that $A_{ij} \sim \text{Bernoulli}(x_i^\top x_j)$ for $i < j$, with $A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.
+If on the other hand $A_{ij} \sim \text{Bernoulli}(x_i^\top I_{p, q} x_j)$ where $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
+These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
+
+In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $x_1, ..., x_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
+\end{definition}
+
+The structure of the RDPG and GRDPG provides an intuitive method for recovery of the latent positions via spectral embedding.
+
+\begin{definition}[Adjacency Spectral Embedding]
+Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$.
+Let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
+Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.
+
+If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
+Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
+\end{definition}
+
+\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,
+
+\begin{equation}
+\max_i \|(\hat{X}_n)_i - W_n (X_n)_i \| \stackrel{a.s.}{\to} 0
+\end{equation}
+
+Similarly, \citet{rubindelanchy2017statistical} showed that for $(A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)$,
+
+\begin{equation}
+\max_i \|(\hat{X}_n)_i - Q_n (X_n)_i \| \stackrel{a.s.}{\to} 0
+\end{equation}
+
+where $Q_n$ is a sequence of matrices in $\mathbb{O}(p, q)$, the indefinite orthogonal group of order $p, q$.
+
+It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all graphs generated by $P$ are special cases of the GRDPG.
+This includes the SBM, DCBM, and PABM.
+
+\begin{example}[Connecting the SBM to the RDPG]
+Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the homogeneous SBM with two communities such that within-community edge probability $p$ and between-community edge probability $q$ where $p > q$.
+Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$.
+Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$.
+Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$.
+One RDPG representation of this SBM is:
+
+$$X = \begin{bmatrix} 
+\sqrt{p} & 0 \\
+\vdots & \vdots \\
+\sqrt{p} & 0 \\
+\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\ 
+\vdots & \vdots \\
+\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
+\end{bmatrix}
+\in \mathbb{R}^{n \times 2}$$
+
+where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$. Then it can be shown that
+
+$$P = X X^\top$$
+\end{example}
+
+The ASE of the assortative SBM consists of points in $\mathbb{R}^K$ that lie near one of $K$ centers, depending on the community label, leading to ASE followed by $K$-means clustering (or similar, e.g., GMM) as a consistent community detection algorithm \cite{lyzinski2014}.
+A similar result can be shown for the DCBM and PABM but with different structures in the ASE.
+
+Given sufficiently large sample size $n$, the scaled ASE of affinity matrix $A$ drawn from a RDPG or GRDPG will asymptotically approach the original latent positions $X$ with probability 1, up to a linear transformation (orthogonal transformation for the RDPG, a composition of an orthogonal and scale transformations for the GRDPG).
+Thus if $X$ consists of points that lie on subspaces of $\mathbb{R}^d$, then both the scaled and unscaled ASE of $A \sim \text{RDPG}(X)$ or $A \sim \text{GRDPG}(X)$ will consist of points that lie near subspaces, with some noise that almost surely goes to $0$ as $n \to \infty$, motivating ASE followed by SSC as an asymptotically consistent method for community detection.
+The Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative graph model with underlying communities such that each community lies on a subspace.
+\citet{noroozi2019estimation} showed that SSC is able to recover the subspaces and therefore perform community detection for the PABM given $P = X I_{p, q} X^\top$, the edge probability matrix, rather than $A$, the adjacency matrix.
+Combining the results of \citeauthor{rubindelanchy2017statistical} and \citeauthor{jmlr-v28-wang13}, it should be possible to recover the communities using $A$ as well.
 
@@
 
-## (Generalized) Random Dot Product Graphs
-
-It has been shown by \citet{jmlr-v28-wang13} that given sufficiently low noise in $X$ and sufficiently low affinity between pairs of subspaces as measured by the cosine of the angle between subspaces, SDP holds for a specific range of $\lambda$.
-This provides theoretical justification for applying SSC to Random Dot Product Graphs (RDPG) \cite{athreya2017statistical} and Generalized Random Dot Product Graphs (GRDPG) \cite{rubindelanchy2017statistical} for which the data in the latent space lie on subspaces.
-
-\begin{definition}[(Generalized) Random Dot Product Graph]
-Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y.\in [0, 1]$.
-$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix $A$ is drawn such that $A_{ij} \sim \text{Bernoulli}(x_i^\top x_j)$ for $i < j$, with $A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.
-If on the other hand $A_{ij} \sim \text{Bernoulli}(x_i^\top I_{p, q} x_j)$ where $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
-These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
-
-In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $x_1, ..., x_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
-If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
-\end{definition}
-
-\begin{definition}[Adjacency Spectral Embedding]
-Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$.
-Let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
-Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.
-
-If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
-Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
-\end{definition}
-
-\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$, if $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - W_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-Similarly, \citeauthor{rubindelanchy2017statistical} showed that for $(A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)$,
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - Q_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-where $Q_n$ is a sequence of matrices in $\mathbb{O}(p, q)$, the indefinite orthogonal group of order $p, q$.
-
-Given sufficiently large sample size $n$, the scaled ASE of affinity matrix $A$ drawn from a RDPG or GRDPG will asymptotically approach the original latent positions $X$ with probability 1, up to a linear transformation (orthogonal transformation for the RDPG, a composition of an orthogonal transformation and scale transformation for the GRDPG).
-Thus if $X$ consists of points that lie on subspaces of $\mathbb{R}^d$, then both the scaled and unscaled ASE of $A \sim \text{RDPG}(X)$ or $A \sim \text{GRDPG}(X)$ will consist of points that lie near subspaces, with some noise that almost surely goes to $0$ as $n \to \infty$, motivating ASE followed by SSC as an asymptotically consistent method for community detection.
-The Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative graph model with underlying communities such that each community lies on a subspace.
-\citet{noroozi2019estimation} showed that SSC is able to recover the subspaces and therefore perform community detection for the PABM given $P = X I_{p, q} X^\top$, the edge probability matrix, rather than $A$, the adjacency matrix.
-Combining the results of \citeauthor{rubindelanchy2017statistical} and \citeauthor{jmlr-v28-wang13}, it should be possible to recover the communities using $A$ as well.
-
 ## Manifold Learning
 
-In addition to subspaces, \citet{trosset2020learning} showed that the ASE of a RDPG can be used to recover one-dimensional manifolds.
+\citet{trosset2020learning} showed that the ASE of a RDPG can be used to recover one-dimensional manifolds.
 Suppose $f : [0, 1] \mapsto \mathcal{X}$ such that $f$ is smooth and $\mathcal{X}$ represents a curve or one-dimensional manifold in $\mathbb{R}^d$.
@@
 \begin{equation}
-d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{1  + (f'(t))^2} dt
+d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{\sum_r^d \Big( \frac{d f_r}{d t} \Big)^2} dt
 \end{equation}
 
-This suggests that if the latent positions of a RDPG consists of $K$ disjoint manifolds, the $\epsilon$-neighborhood graph will consist of $K$ disjoint subgraphs as $\epsilon \to 0$ and $n \to \infty$.
+\citet{athreya2020estimation} extended this further by generating a RDPG from a mixture of distributions on a curve.
+In their example, points were sampled from a mixture of two Beta distributions on the Hardy-Weinberg curve to construct the latent positions of a RDPG, with the goal of recovering the hidden mixture distribution from an observed graph.
 
@@
 
-## Subspace Clustering
+## Connecting Existing Generative Graph Models to the RDPG and GRDPG
+
+### Subspace Clustering
 
@@
 1. Define functions $f_1, ..., f_K$ such that 
-$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_i(t) \neq f_j(t)$ $\forall i \in [K]$ 
+$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_k(t) \neq f_l(t)$ $\forall k, l \in [K]$ 
 and $t \in [0, 1]$.

### 2021-05-10 proposal/proposal.tex
some more updates to proposal

@@
 \begin{center}
-\LARGE{\textbf{Community Detection on Subspaces and Manifolds}}\\
+\LARGE{\textbf{Community Detection Methods for Random Dot Product Graphs and Generalized Random Dot Product Graphs}}\\
 \vspace*{2\baselineskip}
@@
 
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
 \hypertarget{research-goal}{%
-\section{Research Goal}\label{research-goal}}
+\subsection{Research Goal}\label{research-goal}}
 
-Typical clustering methods often work by assuming that each cluster is
-represented by some central point around which the data belonging to
-that cluster lie Such methods, including Lloyd's algorithm for
-\(k\)-means clustering \cite{1056489} and Gaussian Mixture Models
-\cite{doi:10.1198/016214502760047131}, involve iteratively updating the
-central points by averaging data belonging to each cluster and cluster
-memberships by comparing proximity to central points. The aim of our
-research is to explore clustering and community detection methods for
-which each cluster is represented not by a central point but by
-subspaces or manifolds. In addition, we aim to explore various
-applications of such methods and connect them to preexisting clustering
-and community detection problems.
+Graph and network data have become increasingly more widespread in
+various fields including sociology, neuroscience, biostatistics, and
+computer science. This has resulted in certain challenges for
+researchers, as most traditional statistical and machine learning
+methods are incompatible with graphs and instead require the data to
+exist in a subset of Euclidean space. This includes clustering and
+community detection, which are the typical problems of interest when
+working with graphs. Common clustering methods often involve calculating
+some central or representative point around which the data belonging to
+that cluster lie (e.g., Lloyd's algorithm for \(K\)-means clustering
+\cite{1056489}, Gaussian Mixture Models
+\cite{doi:10.1198/016214502760047131}). Because these methods involve
+computing summary statistics within each cluster, they cannot be applied
+directly to graphs as community detection algorithms.
+
+Some proposed methods for unifying graph community detection with
+traditional clustering techniques involve embedding the graph into
+Euclidean space, of which the most popular is Spectral Clustering
+\cite{DBLP:journals/corr/abs-0711-0189}. The Random Dot Product Graph
+(RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot
+Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
+this further by explicitly constructing generative models such that
+latent positions in Euclidean space are used to induce graphs. A
+community detection method motivated by this would involve learning the
+latent positions given an observed graph and then learning the community
+labels given the latent positions.
+
+The aim of our research is to develop consistent community detection
+techniques under the RDPG and GRDPG frameworks. First, we explore
+existing generative graph models with underlying community structures
+that can be inferred by connecting the generative models to the RDPG or
+GRDPG. Then we explore other latent structures or mixture distributions
+in the latent space that induce graphs for which consistent community
+detection is possible.
+
+\hypertarget{notation}{%
+\subsection{Notation}\label{notation}}
+
+Let \(G = (V, E)\) be an undirected, unweighted graph with no self-loops
+with \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
+adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
+edge between vertices \(i\) and \(j\) and \(A_{ij} = 0\) otherwise.
+Because \(G\) is symmetric and contains no self-loops,
+\(A_{ij} = A_{ji}\) and \(A_{ii} = 0\) for \(i, j \in [n]\). We further
+restrict our analyses to independent Bernoulli graphs. Let
+\(P \in [0, 1]^{n \times n}\) be a symmetric matrix of edge
+probabilities. Graph \(G\) is sampled by
+\(A_{ij} \stackrel{\text{indep}}{\sim}P_{ij}\) for each
+\(0 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)). Denote
+\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
+as the sample \(x_1, ..., x_n \in \mathbb{R}^d\), and denote
+\(z_1, ..., z_n \in [K]\) as their corresponding (hidden) labels.
 
@@
 \section{Literature Review}\label{literature-review}}
+
+\hypertarget{generative-graph-models-and-community-detection}{%
+\subsection{Generative Graph Models and Community
+Detection}\label{generative-graph-models-and-community-detection}}
+
+Generative models for symmetric Bernoulli graphs involve defining the
+edge probability matrix \(P\) whose \(ij\)\textsuperscript{th} entry is
+the probability of an edge between vertices \(i\) and \(j\) for each
+\(i, j \in [n]\). Community detection and recovery starts by restricting
+the generative model such that for each pair of vertices, the
+probability of an edge between the vertices is conditioned on the labels
+of the vertices. One such model is the Stochastic Block Model (SBM)
+\cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
+each vertex belonging to one community, the SBM restricts \(P\) to
+\(K (K + 1) / 2\) unique entries such that \(P_{ij} = B_{z_i z_j}\)
+where \(B \in [0, 1]^{K \times K}\) with entries \(B_{kl}\) denoting the
+edge probability of each vertex in community \(k\) having an edge with
+each vertex in community \(l\). The homogeneous SBM further restricts
+\(P\) to two unique entries such that \(P_{ij} = p\) if \(z_i = z_j\)
+and \(P_{ij} = q\) otherwise. Multiple generalizations of the SBM have
+been introduced since, including the Degree Corrected Block Model (DCBM)
+and the Popularity Adjusted Block Model (PABM). Like the SBM, these
+models involve edge probability matrix \(P\) that is restricted to rank
+\(d < n\) in part based on community labels.
+
+\hypertarget{generalized-random-dot-product-graphs}{%
+\subsection{(Generalized) Random Dot Product
+Graphs}\label{generalized-random-dot-product-graphs}}
+
+The RDPG starts with points in latent space
+\(X \in \mathcal{X} \subset \mathbb{R}^d\) such that
+\(\forall x, y \in \mathcal{X}\), \(x^\top y \in [0, 1]\). \(P\) is then
+constructed as \(P = X X^\top\), and graph \(G\) with adjacency matrix
+\(A\) is drawn from \(P\). We provide a more formal definition of the
+RDPG and GRDPG below.
+
+\begin{definition}[(Generalized) Random Dot Product Graph]
+Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y.\in [0, 1]$.
+$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix $A$ is drawn such that $A_{ij} \sim \text{Bernoulli}(x_i^\top x_j)$ for $i < j$, with $A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.
+If on the other hand $A_{ij} \sim \text{Bernoulli}(x_i^\top I_{p, q} x_j)$ where $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
+These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
+
+In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $x_1, ..., x_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
+\end{definition}
+
+The structure of the RDPG and GRDPG provides an intuitive method for
+recovery of the latent positions via spectral embedding.
+
+\begin{definition}[Adjacency Spectral Embedding]
+Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$.
+Let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
+Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.
+
+If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
+Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
+\end{definition}
+
+\citeauthor{athreya2017statistical} showed that under mild conditions,
+if \((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
+ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),
+
+\begin{equation}
+\max_i \|(\hat{X}_n)_i - W_n (X_n)_i \| \stackrel{a.s.}{\to} 0
+\end{equation}
+
+Similarly, \citet{rubindelanchy2017statistical} showed that for
+\((A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)\),
+
+\begin{equation}
+\max_i \|(\hat{X}_n)_i - Q_n (X_n)_i \| \stackrel{a.s.}{\to} 0
+\end{equation}
+
+where \(Q_n\) is a sequence of matrices in \(\mathbb{O}(p, q)\), the
+indefinite orthogonal group of order \(p, q\).
+
+It is straightforward to show that all Bernoulli graphs with positive
+semidefinite \(P\) are special cases of the RDPG, which is a special
+case of the GRDPG, and all graphs generated by \(P\) are special cases
+of the GRDPG. This includes the SBM, DCBM, and PABM.
+
+\begin{example}[Connecting the SBM to the RDPG]
+Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the homogeneous SBM with two communities such that within-community edge probability $p$ and between-community edge probability $q$ where $p > q$.
+Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$.
+Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$.
+Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$.
+One RDPG representation of this SBM is:
+
+$$X = \begin{bmatrix} 
+\sqrt{p} & 0 \\
+\vdots & \vdots \\
+\sqrt{p} & 0 \\
+\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\ 
+\vdots & \vdots \\
+\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
+\end{bmatrix}
+\in \mathbb{R}^{n \times 2}$$
+
+where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$. Then it can be shown that
+
+$$P = X X^\top$$
+\end{example}
+
+The ASE of the assortative SBM consists of points in \(\mathbb{R}^K\)
+that lie near one of \(K\) centers, depending on the community label,
+leading to ASE followed by \(K\)-means clustering (or similar, e.g.,
+GMM) as a consistent community detection algorithm \cite{lyzinski2014}.
+A similar result can be shown for the DCBM and PABM but with different
+structures in the ASE.
+
+Given sufficiently large sample size \(n\), the scaled ASE of affinity
+matrix \(A\) drawn from a RDPG or GRDPG will asymptotically approach the
+original latent positions \(X\) with probability 1, up to a linear
+transformation (orthogonal transformation for the RDPG, a composition of
+an orthogonal and scale transformations for the GRDPG). Thus if \(X\)
+consists of points that lie on subspaces of \(\mathbb{R}^d\), then both
+the scaled and unscaled ASE of \(A \sim \text{RDPG}(X)\) or
+\(A \sim \text{GRDPG}(X)\) will consist of points that lie near
+subspaces, with some noise that almost surely goes to \(0\) as
+\(n \to \infty\), motivating ASE followed by SSC as an asymptotically
+consistent method for community detection. The Popularity Adjusted Block
+Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative
+graph model with underlying communities such that each community lies on
+a subspace. \citet{noroozi2019estimation} showed that SSC is able to
+recover the subspaces and therefore perform community detection for the
+PABM given \(P = X I_{p, q} X^\top\), the edge probability matrix,
+rather than \(A\), the adjacency matrix. Combining the results of
+\citeauthor{rubindelanchy2017statistical} and
+\citeauthor{jmlr-v28-wang13}, it should be possible to recover the
+communities using \(A\) as well.
 
@@
 
-\hypertarget{generalized-random-dot-product-graphs}{%
-\subsection{(Generalized) Random Dot Product
-Graphs}\label{generalized-random-dot-product-graphs}}
-
-It has been shown by \citet{jmlr-v28-wang13} that given sufficiently low
-noise in \(X\) and sufficiently low affinity between pairs of subspaces
-as measured by the cosine of the angle between subspaces, SDP holds for
-a specific range of \(\lambda\). This provides theoretical justification
-for applying SSC to Random Dot Product Graphs (RDPG)
-\cite{athreya2017statistical} and Generalized Random Dot Product Graphs
-(GRDPG) \cite{rubindelanchy2017statistical} for which the data in the
-latent space lie on subspaces.
-
-\begin{definition}[(Generalized) Random Dot Product Graph]
-Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y.\in [0, 1]$.
-$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix $A$ is drawn such that $A_{ij} \sim \text{Bernoulli}(x_i^\top x_j)$ for $i < j$, with $A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.
-If on the other hand $A_{ij} \sim \text{Bernoulli}(x_i^\top I_{p, q} x_j)$ where $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
-These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
-
-In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $x_1, ..., x_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
-If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
-\end{definition}
-
-\begin{definition}[Adjacency Spectral Embedding]
-Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$.
-Let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
-Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.
-
-If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
-Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
-\end{definition}
-
-\citeauthor{athreya2017statistical} showed that under mild conditions,
-if \((A_n, X_n) \sim \text{RDPG}(F, n)\), if \(\hat{X}_n\) is the scaled
-ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - W_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-Similarly, \citeauthor{rubindelanchy2017statistical} showed that for
-\((A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)\),
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - Q_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-where \(Q_n\) is a sequence of matrices in \(\mathbb{O}(p, q)\), the
-indefinite orthogonal group of order \(p, q\).
-
-Given sufficiently large sample size \(n\), the scaled ASE of affinity
-matrix \(A\) drawn from a RDPG or GRDPG will asymptotically approach the
-original latent positions \(X\) with probability 1, up to a linear
-transformation (orthogonal transformation for the RDPG, a composition of
-an orthogonal transformation and scale transformation for the GRDPG).
-Thus if \(X\) consists of points that lie on subspaces of
-\(\mathbb{R}^d\), then both the scaled and unscaled ASE of
-\(A \sim \text{RDPG}(X)\) or \(A \sim \text{GRDPG}(X)\) will consist of
-points that lie near subspaces, with some noise that almost surely goes
-to \(0\) as \(n \to \infty\), motivating ASE followed by SSC as an
-asymptotically consistent method for community detection. The Popularity
-Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a
-generative graph model with underlying communities such that each
-community lies on a subspace. \citet{noroozi2019estimation} showed that
-SSC is able to recover the subspaces and therefore perform community
-detection for the PABM given \(P = X I_{p, q} X^\top\), the edge
-probability matrix, rather than \(A\), the adjacency matrix. Combining
-the results of \citeauthor{rubindelanchy2017statistical} and
-\citeauthor{jmlr-v28-wang13}, it should be possible to recover the
-communities using \(A\) as well.
-
 \hypertarget{manifold-learning}{%
@@
 
-In addition to subspaces, \citet{trosset2020learning} showed that the
-ASE of a RDPG can be used to recover one-dimensional manifolds. Suppose
+\citet{trosset2020learning} showed that the ASE of a RDPG can be used to
+recover one-dimensional manifolds. Suppose
 \(f : [0, 1] \mapsto \mathcal{X}\) such that \(f\) is smooth and
@@
 \begin{equation}
-d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{1  + (f'(t))^2} dt
+d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{\sum_r^d \Big( \frac{d f_r}{d t} \Big)^2} dt
 \end{equation}
 
-This suggests that if the latent positions of a RDPG consists of \(K\)
-disjoint manifolds, the \(\epsilon\)-neighborhood graph will consist of
-\(K\) disjoint subgraphs as \(\epsilon \to 0\) and \(n \to \infty\).
+\citet{athreya2020estimation} extended this further by generating a RDPG
+from a mixture of distributions on a curve. In their example, points
+were sampled from a mixture of two Beta distributions on the
+Hardy-Weinberg curve to construct the latent positions of a RDPG, with
+the goal of recovering the hidden mixture distribution from an observed
+graph.
 
@@
 
+\hypertarget{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}{%
+\subsection{Connecting Existing Generative Graph Models to the RDPG and
+GRDPG}\label{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}}
+
 \hypertarget{subspace-clustering-1}{%
-\subsection{Subspace Clustering}\label{subspace-clustering-1}}
+\subsubsection{Subspace Clustering}\label{subspace-clustering-1}}
 
@@
   Define functions \(f_1, ..., f_K\) such that
-  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_i(t) \neq f_j(t)\)
-  \(\forall i \in [K]\) and \(t \in [0, 1]\).
+  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_k(t) \neq f_l(t)\)
+  \(\forall k, l \in [K]\) and \(t \in [0, 1]\).
 \item

### 2021-05-11 proposal/proposal.Rmd
updated proposal

@@
 
-Graph and network data have become increasingly more widespread in various fields including sociology, neuroscience, biostatistics, and computer science.
-This has resulted in certain challenges for researchers, as most traditional statistical and machine learning methods are incompatible with graphs and instead require the data to exist in a subset of Euclidean space.
-This includes clustering and community detection, which are the typical problems of interest when working with graphs.
-Common clustering methods often involve calculating some central or representative point around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}).
-Because these methods involve computing summary statistics within each cluster, they cannot be applied directly to graphs as community detection algorithms.
+Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science.
+This has resulted in various challenges for researchers who rely on traditional statistical and machine learning methods, many of which are incompatible with graph data and instead require the data to exist as feature vectors in Euclidean space.
+Such challenges often involve clustering and community detection.
+Common clustering methods typically involve calculating some central or representative point for each cluster around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}).
+Because these methods involve computing summary statistics within each cluster, such as the sample average, they cannot be applied directly to graphs, necessitating methods for transforming the graph data into feature vectors.
 
-Some proposed methods for unifying graph community detection with traditional clustering techniques involve embedding the graph into Euclidean space, of which the most popular is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}.
+One family of methods for unifying graph community detection with traditional clustering techniques is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the graph into Euclidean space, followed by applying a popular clustering algorithm such as $K$-means clustering.
 The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models such that latent positions in Euclidean space are used to induce graphs.
-A community detection method motivated by this would involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.
+A community detection algorithm motivated by this may involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.
 
@@
 Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities.
-Graph $G$ is sampled by $A_{ij} \indep P_{ij}$ for each $0 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$).
-Denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$, and denote $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.
+Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $0 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$).
+Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$, and denote $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.
 
@@
 Generative models for symmetric Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$.
-Community detection and recovery starts by restricting the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on the labels of the vertices.
-One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = B_{z_i z_j}$ where $B \in [0, 1]^{K \times K}$ with entries $B_{kl}$ denoting the edge probability of each vertex in community $k$ having an edge with each vertex in community $l$.
+In order to motivate community detection methods, we restrict the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on the labels of the vertices.
+One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = B_{z_i z_j}$ and $B \in [0, 1]^{K \times K}$ with entries $B_{kl}$ denoting the edge probability of each vertex in community $k$ having an edge with each vertex in community $l$.
 The homogeneous SBM further restricts $P$ to two unique entries such that $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise.
@@
 
+Like the SBM, DCBM, and PABM, the RDPG and GRDPG are generative models for graphs involving an edge probability matrix $P$.
 The RDPG starts with points in latent space $X \in \mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y \in [0, 1]$.
@@
 
-The structure of the RDPG and GRDPG provides an intuitive method for recovery of the latent positions via spectral embedding.
+The structure of the RDPG and GRDPG provides a straightforward method for recovery of the latent positions via spectral embedding.
 
@@
 
-\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,
+\citet{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,
 
@@
 This includes the SBM, DCBM, and PABM.
+In the following example, we show that the ASE of the SBM has a very particular form.
 
@@
 
-## Subspace Clustering
-
-Subspace clustering, which assumes that points in $\mathbb{R}^d$ each lie on one of $K$ subspaces of $\mathbb{R}^d$, is an approach that has found a wide range of uses by the Statistics and Machine Learning communities, particularly within the field of Computer Vision \cite{5206547}.
-
-Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample.
-Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
-The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$.
-If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces.
-Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.
-
-In practice, SSC is performed by solving the LASSO problems:
-
-\begin{equation} \label{eq:ssc}
-c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
-\end{equation}
-
-for some sparsity parameter $\lambda > 0$.
-The $c_i$ vectors are then collected into $C$ and $B$ as described before.
-If $X$ is noisy in that each $x_i$ does not lie exactly on one of $K$ subspaces but near it, the choice of $\lambda$ becomes important in guaranteeing the Subspace Detection Property (SDP) \cite{jmlr-v28-wang13}.
-
-\begin{definition}[Subspace Detection Property] 
-Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
-points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
-solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
-$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
-different subspaces, then $X$ obeys the Subspace Detection Property. 
-\end{definition}
-
-\begin{remark} 
-In practice, a noisy sample $X$ often does not obey SDP. 
-In such cases, $B$ is treated as an affinity matrix for a graph which 
-is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
-hand, if $X$ does obey the SDP, $B$ describes a graph 
-with at least $K$ disconnected subgraphs. Ideally, when SDP holds, 
-there are exactly $K$ subgraphs which map to each subspace, 
-but it could be the case that some of the subspaces are represented by 
-multiple disconnected subgraphs. SDP is contingent 
-on choosing a sufficiently large sparsity parameter $\lambda$. 
-\end{remark}
-
 ## Manifold Learning
@@
 
-### Subspace Clustering
-
-As discussed in \S 2.2, if the latent positions of a RDPG or GRDPG model are such that they lie on a small number of subspaces, ASE followed by SSC may be able to identify whether vertices of the graph $v_i, v_j$ belong to the same subspace or to different subspaces, and one such model that is consistent with this construction is the PABM.
+As discussed in \S 2.1, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space.
+A similar result has been shown for the DCBM.
+We will now extend this to the PABM.
 
@@
 
-We will first define the PABM.
+We first define the PABM.
 
@@
 
-We will use the notation $A \sim \text{PABM}(\{\lambda^{(kl)}\}_K)$ to denote a random 
-adjacency matrix $A$ drawn from a PABM with parameters $\lambda^{(kl)}$ 
-consisting of $K$ underlying communities.
+We will use the notation $A \sim \text{PABM}(\{\lambda^{(kl)}\}_K)$ 
+to denote a random adjacency matrix $A$ drawn from a PABM with parameters
+$\lambda^{(kl)}$ consisting of $K$ underlying communities.
 
@@
 
-This leads to the following theorem.
+In Theorem \ref{theorem2}, we showed that a possible latent configuration for a $K$-community PABM under the GRDPG framework consists of latent positions in $\mathbb{R}^{K^2}$ such that each community corresponds to a $K$-dimensional subspace such that each subspace is orthogonal to the others.
 
@@
 
-If $\hat{V}$ is the *unscaled* ASE of $A$, Theorem \ref{osc} and results from  \citeauthor{rubindelanchy2017statistical} together imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} 0$, leading to the following result:
+If $\hat{V}$ is the *unscaled* ASE of $A$, then results from  \citeauthor{rubindelanchy2017statistical} imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top$.
+Then for vertices $i$ and $j$ belonging to separate communities, the $ij$^th^ entry of $n \hat{V} \hat{V}^\top$ approaches 0 with probability 1:
+
+\begin{theorem}
+\label{theorem4} 
+Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
+(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
+and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
+
+\begin{equation} \label{eq:thm4}
+\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
+O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
+\end{equation}
+
+This provides the result that $\forall i, j$ in different communities, 
+$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
+\end{theorem}
 
@@
 
-\begin{theorem}
-\label{theorem4} 
-Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
-(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
-and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
+### Subspace Clustering
 
-\begin{equation} \label{eq:thm4}
-\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
-O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
+In addition to OSC, since the communities are subspaces in the latent space, we can leverage existing methods for subspace clustering.
+Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample.
+Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
+The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$.
+If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces.
+Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.
+
+In practice, SSC is performed by solving the LASSO problems:
+
+\begin{equation} \label{eq:ssc}
+c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
 \end{equation}
 
-This provides the result that $\forall i, j$ in different communities, 
-$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
-\end{theorem}
+for some sparsity parameter $\lambda > 0$.
+The $c_i$ vectors are then collected into $C$ and $B$ as described before.
+If $X$ is noisy in that each $x_i$ does not lie exactly on one of $K$ subspaces but near it, the choice of $\lambda$ becomes important in guaranteeing the Subspace Detection Property (SDP) \cite{jmlr-v28-wang13}.
+
+\begin{definition}[Subspace Detection Property] 
+Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
+points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
+solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
+$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
+different subspaces, then $X$ obeys the Subspace Detection Property. 
+\end{definition}
+
+\begin{remark} 
+In practice, a noisy sample $X$ often does not obey SDP. 
+In such cases, $B$ is treated as an affinity matrix for a graph which 
+is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
+hand, if $X$ does obey the SDP, $B$ describes a graph 
+with at least $K$ disconnected subgraphs. Ideally, when SDP holds, 
+there are exactly $K$ subgraphs which map to each subspace, 
+but it could be the case that some of the subspaces are represented by 
+multiple disconnected subgraphs. SDP is contingent 
+on choosing a sufficiently large sparsity parameter $\lambda$. 
+\end{remark}
 
@@
 
-## Manifold Clustering
+## Proposed Generative Graph Models
 
-We would like to extend subspace clustering to manifold clustering.
-The problem setup is as follows: Suppose that in the latent space $\mathcal{X} \subset \mathbb{R}^d$, sample $X$ of $n$ points lie on a union of $K$ disjoint manifolds with each manifold corresponding to a community.
+In the previous sections, we connected well-known and highly structured generative graph models to the RDPG and GRDPG to show that highly structured latent configurations generate graphs consistent with these models: The latent space for the SBM consists of $K$ point masses, the latent space for the DCBM consists of $K$ rays emanating from the origin, and the latent space for the PABM consists of $K$ $K$-dimensional subspaces.
+In the following sections, we explore additional structured latent configurations corresponding to community structure and develop methods for community detection based on the consistency of the ASE and the structural forms of the latent configurations.
+
+The general structure of interest can be described as follows: Suppose that in the latent space $\mathcal{X} \subset \mathbb{R}^d$, sample $X$ of $n$ points lie on a union of $K$ disjoint manifolds with each manifold corresponding to a community.
 If $A \sim \text{RDPG}(X)$, we wish to recover the community labels (up to permutation) from $A$.
 
-Similarly, suppose that probability distribution $F$ is described as follows:
+Equivalently, suppose that probability distribution $F$ is described as follows:
 
@@
 2. Sample labels
-$z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
-3. Sample $t_1, ..., t_n \stackrel{iid}{\sim} D$ where $D$ has support 
+$z_1, ..., z_n \iid Categorical(\pi_1, ..., \pi_K)$.
+3. Sample $t_1, ..., t_n \iid D$ where $D$ has support 
 $[0, 1]$. 
@@
 
-### Affine Subspaces
+### Affine Segments
 
@@
 If instead of sampling uniformly from line segments of unit length, we sample uniformly from a 1 dimensional manifolds of unit length, the above property still holds.
-Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|$.
+Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_u^v \sqrt{\sum_k (df_k / dt)^2} dt = \|u - v\|$.
 Then $U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$.
-If the shortest distance between the two manifolds defined by $f_1$ and $f_2$ with the same restriction is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.
+If the shortest distance between the two manifolds defined by $f_1$ and $f_2$ with the same restriction is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_k(\cdot)$.
 

### 2021-05-11 proposal/proposal.tex
updated proposal

@@
 
-Graph and network data have become increasingly more widespread in
-various fields including sociology, neuroscience, biostatistics, and
-computer science. This has resulted in certain challenges for
-researchers, as most traditional statistical and machine learning
-methods are incompatible with graphs and instead require the data to
-exist in a subset of Euclidean space. This includes clustering and
-community detection, which are the typical problems of interest when
-working with graphs. Common clustering methods often involve calculating
-some central or representative point around which the data belonging to
-that cluster lie (e.g., Lloyd's algorithm for \(K\)-means clustering
-\cite{1056489}, Gaussian Mixture Models
-\cite{doi:10.1198/016214502760047131}). Because these methods involve
-computing summary statistics within each cluster, they cannot be applied
-directly to graphs as community detection algorithms.
+Graph and network data have become increasingly widespread in various
+fields including sociology, neuroscience, biostatistics, and computer
+science. This has resulted in various challenges for researchers who
+rely on traditional statistical and machine learning methods, many of
+which are incompatible with graph data and instead require the data to
+exist as feature vectors in Euclidean space. Such challenges often
+involve clustering and community detection. Common clustering methods
+typically involve calculating some central or representative point for
+each cluster around which the data belonging to that cluster lie (e.g.,
+Lloyd's algorithm for \(K\)-means clustering \cite{1056489}, Gaussian
+Mixture Models \cite{doi:10.1198/016214502760047131}). Because these
+methods involve computing summary statistics within each cluster, such
+as the sample average, they cannot be applied directly to graphs,
+necessitating methods for transforming the graph data into feature
+vectors.
 
-Some proposed methods for unifying graph community detection with
-traditional clustering techniques involve embedding the graph into
-Euclidean space, of which the most popular is Spectral Clustering
-\cite{DBLP:journals/corr/abs-0711-0189}. The Random Dot Product Graph
+One family of methods for unifying graph community detection with
+traditional clustering techniques is Spectral Clustering
+\cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the
+graph into Euclidean space, followed by applying a popular clustering
+algorithm such as \(K\)-means clustering. The Random Dot Product Graph
 (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot
@@
 latent positions in Euclidean space are used to induce graphs. A
-community detection method motivated by this would involve learning the
+community detection algorithm motivated by this may involve learning the
 latent positions given an observed graph and then learning the community
@@
 \(P \in [0, 1]^{n \times n}\) be a symmetric matrix of edge
-probabilities. Graph \(G\) is sampled by
-\(A_{ij} \stackrel{\text{indep}}{\sim}P_{ij}\) for each
-\(0 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)). Denote
+probabilities. Graph \(G\) is sampled from \(P\) by drawing
+\(A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})\) for
+each \(0 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)).
+Finally, we denote
 \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
@@
 the probability of an edge between vertices \(i\) and \(j\) for each
-\(i, j \in [n]\). Community detection and recovery starts by restricting
-the generative model such that for each pair of vertices, the
+\(i, j \in [n]\). In order to motivate community detection methods, we
+restrict the generative model such that for each pair of vertices, the
 probability of an edge between the vertices is conditioned on the labels
@@
 each vertex belonging to one community, the SBM restricts \(P\) to
-\(K (K + 1) / 2\) unique entries such that \(P_{ij} = B_{z_i z_j}\)
-where \(B \in [0, 1]^{K \times K}\) with entries \(B_{kl}\) denoting the
-edge probability of each vertex in community \(k\) having an edge with
-each vertex in community \(l\). The homogeneous SBM further restricts
-\(P\) to two unique entries such that \(P_{ij} = p\) if \(z_i = z_j\)
-and \(P_{ij} = q\) otherwise. Multiple generalizations of the SBM have
-been introduced since, including the Degree Corrected Block Model (DCBM)
-and the Popularity Adjusted Block Model (PABM). Like the SBM, these
-models involve edge probability matrix \(P\) that is restricted to rank
+\(K (K + 1) / 2\) unique entries such that \(P_{ij} = B_{z_i z_j}\) and
+\(B \in [0, 1]^{K \times K}\) with entries \(B_{kl}\) denoting the edge
+probability of each vertex in community \(k\) having an edge with each
+vertex in community \(l\). The homogeneous SBM further restricts \(P\)
+to two unique entries such that \(P_{ij} = p\) if \(z_i = z_j\) and
+\(P_{ij} = q\) otherwise. Multiple generalizations of the SBM have been
+introduced since, including the Degree Corrected Block Model (DCBM) and
+the Popularity Adjusted Block Model (PABM). Like the SBM, these models
+involve edge probability matrix \(P\) that is restricted to rank
 \(d < n\) in part based on community labels.
@@
 
-The RDPG starts with points in latent space
-\(X \in \mathcal{X} \subset \mathbb{R}^d\) such that
-\(\forall x, y \in \mathcal{X}\), \(x^\top y \in [0, 1]\). \(P\) is then
-constructed as \(P = X X^\top\), and graph \(G\) with adjacency matrix
-\(A\) is drawn from \(P\). We provide a more formal definition of the
-RDPG and GRDPG below.
+Like the SBM, DCBM, and PABM, the RDPG and GRDPG are generative models
+for graphs involving an edge probability matrix \(P\). The RDPG starts
+with points in latent space \(X \in \mathcal{X} \subset \mathbb{R}^d\)
+such that \(\forall x, y \in \mathcal{X}\), \(x^\top y \in [0, 1]\).
+\(P\) is then constructed as \(P = X X^\top\), and graph \(G\) with
+adjacency matrix \(A\) is drawn from \(P\). We provide a more formal
+definition of the RDPG and GRDPG below.
 
@@
 
-The structure of the RDPG and GRDPG provides an intuitive method for
-recovery of the latent positions via spectral embedding.
+The structure of the RDPG and GRDPG provides a straightforward method
+for recovery of the latent positions via spectral embedding.
 
@@
 
-\citeauthor{athreya2017statistical} showed that under mild conditions,
-if \((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
+\citet{athreya2017statistical} showed that under mild conditions, if
+\((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
 ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),
@@
 case of the GRDPG, and all graphs generated by \(P\) are special cases
-of the GRDPG. This includes the SBM, DCBM, and PABM.
+of the GRDPG. This includes the SBM, DCBM, and PABM. In the following
+example, we show that the ASE of the SBM has a very particular form.
 
@@
 
-\hypertarget{subspace-clustering}{%
-\subsection{Subspace Clustering}\label{subspace-clustering}}
-
-Subspace clustering, which assumes that points in \(\mathbb{R}^d\) each
-lie on one of \(K\) subspaces of \(\mathbb{R}^d\), is an approach that
-has found a wide range of uses by the Statistics and Machine Learning
-communities, particularly within the field of Computer Vision
-\cite{5206547}.
-
-Of particular interest is Sparse Subspace Clustering (SSC), which is
-performed by solving an optimization problem for each observed point in
-a sample. Given \(X \in \mathbb{R}^{n \times d}\) with vectors
-\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
-\(c_i = \min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
-\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
-collected into matrix
-\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
-construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
-perfectly on one of \(K\) subspaces, \(B\) is sparse such that
-\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
-Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
-and if the number of subgraphs is exactly \(K\), each subgraph maps onto
-a subspace.
-
-In practice, SSC is performed by solving the LASSO problems:
-
-\begin{equation} \label{eq:ssc}
-c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
-\end{equation}
-
-for some sparsity parameter \(\lambda > 0\). The \(c_i\) vectors are
-then collected into \(C\) and \(B\) as described before. If \(X\) is
-noisy in that each \(x_i\) does not lie exactly on one of \(K\)
-subspaces but near it, the choice of \(\lambda\) becomes important in
-guaranteeing the Subspace Detection Property (SDP)
-\cite{jmlr-v28-wang13}.
-
-\begin{definition}[Subspace Detection Property] 
-Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
-points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
-solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
-$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
-different subspaces, then $X$ obeys the Subspace Detection Property. 
-\end{definition}
-
-\begin{remark} 
-In practice, a noisy sample $X$ often does not obey SDP. 
-In such cases, $B$ is treated as an affinity matrix for a graph which 
-is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
-hand, if $X$ does obey the SDP, $B$ describes a graph 
-with at least $K$ disconnected subgraphs. Ideally, when SDP holds, 
-there are exactly $K$ subgraphs which map to each subspace, 
-but it could be the case that some of the subspaces are represented by 
-multiple disconnected subgraphs. SDP is contingent 
-on choosing a sufficiently large sparsity parameter $\lambda$. 
-\end{remark}
-
 \hypertarget{manifold-learning}{%
@@
 
-\hypertarget{subspace-clustering-1}{%
-\subsubsection{Subspace Clustering}\label{subspace-clustering-1}}
-
-As discussed in \S 2.2, if the latent positions of a RDPG or GRDPG model
-are such that they lie on a small number of subspaces, ASE followed by
-SSC may be able to identify whether vertices of the graph \(v_i, v_j\)
-belong to the same subspace or to different subspaces, and one such
-model that is consistent with this construction is the PABM.
+As discussed in \S 2.1, all Bernoulli graph models can be expressed as a
+RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
+with a very specific structure in the latent space. A similar result has
+been shown for the DCBM. We will now extend this to the PABM.
 
@@
 
-This leads to the following theorem.
+In Theorem \ref{theorem2}, we showed that a possible latent
+configuration for a \(K\)-community PABM under the GRDPG framework
+consists of latent positions in \(\mathbb{R}^{K^2}\) such that each
+community corresponds to a \(K\)-dimensional subspace such that each
+subspace is orthogonal to the others.
 
@@
 
-If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), Theorem \ref{osc}
-and results from \citeauthor{rubindelanchy2017statistical} together
-imply \(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} 0\), leading to the
-following result:
+If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), then results from
+\citeauthor{rubindelanchy2017statistical} imply
+\(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top\). Then for
+vertices \(i\) and \(j\) belonging to separate communities, the
+\(ij\)\textsuperscript{th} entry of \(n \hat{V} \hat{V}^\top\)
+approaches 0 with probability 1:
+
+\begin{theorem}
+\label{theorem4} 
+Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
+(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
+and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
+
+\begin{equation} \label{eq:thm4}
+\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
+O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
+\end{equation}
+
+This provides the result that $\forall i, j$ in different communities, 
+$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
+\end{theorem}
 
@@
 
-\begin{theorem}
-\label{theorem4} 
-Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
-(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
-and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
+\hypertarget{subspace-clustering}{%
+\subsubsection{Subspace Clustering}\label{subspace-clustering}}
 
-\begin{equation} \label{eq:thm4}
-\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
-O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
+In addition to OSC, since the communities are subspaces in the latent
+space, we can leverage existing methods for subspace clustering. Of
+particular interest is Sparse Subspace Clustering (SSC), which is
+performed by solving an optimization problem for each observed point in
+a sample. Given \(X \in \mathbb{R}^{n \times d}\) with vectors
+\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
+\(c_i = \min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
+\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
+collected into matrix
+\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
+construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
+perfectly on one of \(K\) subspaces, \(B\) is sparse such that
+\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
+Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
+and if the number of subgraphs is exactly \(K\), each subgraph maps onto
+a subspace.
+
+In practice, SSC is performed by solving the LASSO problems:
+
+\begin{equation} \label{eq:ssc}
+c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
 \end{equation}
 
-This provides the result that $\forall i, j$ in different communities, 
-$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
-\end{theorem}
+for some sparsity parameter \(\lambda > 0\). The \(c_i\) vectors are
+then collected into \(C\) and \(B\) as described before. If \(X\) is
+noisy in that each \(x_i\) does not lie exactly on one of \(K\)
+subspaces but near it, the choice of \(\lambda\) becomes important in
+guaranteeing the Subspace Detection Property (SDP)
+\cite{jmlr-v28-wang13}.
+
+\begin{definition}[Subspace Detection Property] 
+Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
+points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
+solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
+$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
+different subspaces, then $X$ obeys the Subspace Detection Property. 
+\end{definition}
+
+\begin{remark} 
+In practice, a noisy sample $X$ often does not obey SDP. 
+In such cases, $B$ is treated as an affinity matrix for a graph which 
+is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
+hand, if $X$ does obey the SDP, $B$ describes a graph 
+with at least $K$ disconnected subgraphs. Ideally, when SDP holds, 
+there are exactly $K$ subgraphs which map to each subspace, 
+but it could be the case that some of the subspaces are represented by 
+multiple disconnected subgraphs. SDP is contingent 
+on choosing a sufficiently large sparsity parameter $\lambda$. 
+\end{remark}
 
@@
 
-\hypertarget{manifold-clustering}{%
-\subsection{Manifold Clustering}\label{manifold-clustering}}
+\hypertarget{proposed-generative-graph-models}{%
+\subsection{Proposed Generative Graph
+Models}\label{proposed-generative-graph-models}}
 
-We would like to extend subspace clustering to manifold clustering. The
-problem setup is as follows: Suppose that in the latent space
-\(\mathcal{X} \subset \mathbb{R}^d\), sample \(X\) of \(n\) points lie
-on a union of \(K\) disjoint manifolds with each manifold corresponding
-to a community. If \(A \sim \text{RDPG}(X)\), we wish to recover the
-community labels (up to permutation) from \(A\).
+In the previous sections, we connected well-known and highly structured
+generative graph models to the RDPG and GRDPG to show that highly
+structured latent configurations generate graphs consistent with these
+models: The latent space for the SBM consists of \(K\) point masses, the
+latent space for the DCBM consists of \(K\) rays emanating from the
+origin, and the latent space for the PABM consists of \(K\)
+\(K\)-dimensional subspaces. In the following sections, we explore
+additional structured latent configurations corresponding to community
+structure and develop methods for community detection based on the
+consistency of the ASE and the structural forms of the latent
+configurations.
 
-Similarly, suppose that probability distribution \(F\) is described as
-follows:
+The general structure of interest can be described as follows: Suppose
+that in the latent space \(\mathcal{X} \subset \mathbb{R}^d\), sample
+\(X\) of \(n\) points lie on a union of \(K\) disjoint manifolds with
+each manifold corresponding to a community. If
+\(A \sim \text{RDPG}(X)\), we wish to recover the community labels (up
+to permutation) from \(A\).
+
+Equivalently, suppose that probability distribution \(F\) is described
+as follows:
 
@@
   Sample labels
-  \(z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)\).
+  \(z_1, ..., z_n \stackrel{\text{iid}}{\sim}Categorical(\pi_1, ..., \pi_K)\).
 \item
-  Sample \(t_1, ..., t_n \stackrel{iid}{\sim} D\) where \(D\) has
+  Sample \(t_1, ..., t_n \stackrel{\text{iid}}{\sim}D\) where \(D\) has
   support \([0, 1]\).
@@
 
-\hypertarget{affine-subspaces}{%
-\subsubsection{Affine Subspaces}\label{affine-subspaces}}
+\hypertarget{affine-segments}{%
+\subsubsection{Affine Segments}\label{affine-segments}}
 
@@
 \(f : [0, 1] \mapsto \mathbb{R}^d\) be a smooth function such that
-\(\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|\). Then
+\(\int_u^v \sqrt{\sum_k (df_k / dt)^2} dt = \|u - v\|\). Then
 \(U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
@@
 before is sufficient, although perhaps a more lenient lower bound can be
-derived based on the shape of \(f_i(\cdot)\).
+derived based on the shape of \(f_k(\cdot)\).
 

### 2021-05-11 proposal/proposal.Rmd
updated proposal

@@
 
-### Subspace Clustering
-
 In addition to OSC, since the communities are subspaces in the latent space, we can leverage existing methods for subspace clustering.
@@
 The $c_i$ vectors are then collected into $C$ and $B$ as described before.
-If $X$ is noisy in that each $x_i$ does not lie exactly on one of $K$ subspaces but near it, the choice of $\lambda$ becomes important in guaranteeing the Subspace Detection Property (SDP) \cite{jmlr-v28-wang13}.
+If $X$ is noisy such that each $x_i$ does not lie exactly on one of $K$ subspaces but near it, the choice of $\lambda$ becomes important in guaranteeing the Subspace Detection Property (SDP) \cite{jmlr-v28-wang13}.
 

### 2021-05-11 proposal/proposal.tex
updated proposal

@@
 
-\hypertarget{subspace-clustering}{%
-\subsubsection{Subspace Clustering}\label{subspace-clustering}}
-
 In addition to OSC, since the communities are subspaces in the latent
@@
 then collected into \(C\) and \(B\) as described before. If \(X\) is
-noisy in that each \(x_i\) does not lie exactly on one of \(K\)
+noisy such that each \(x_i\) does not lie exactly on one of \(K\)
 subspaces but near it, the choice of \(\lambda\) becomes important in

### 2021-05-11 proposal/proposal.Rmd
updated proposal

@@
 
+In this section, we will consider latent positions sampled uniformly from parallel unit length segments.
+
+\begin{example}
+Let $U_1, ..., U_{n_1}, V_1, ..., V_{n_2} \iid Uniform(0, 1)$, $f_1(t) = (t, 0)$, and $f_2(t) = (t, a)$.
+$X_i = f_1(U_i)$ and $Y_j = f_2(V_j)$ $\forall i \in [n_1]$ and $j \in [n_2]$.
+If we observe $X_1, ..., X_{n_1}, Y_1, ..., Y_{n_2}$, what approach will allow us to group the observations by segment?
+\end{example}
+
 We will motivate an approach by the following example.
@@
 \begin{example}
-Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ with order statistics 
+Let $U_1, ..., U_n \iid Uniform(0, 1)$ with order statistics 
 $U_{(1)}, ..., U_{(n)}$. Then $\forall a \in (0, 1)$ and $\delta \in (0, 1)$, $\exists N = N(\delta, a) < \infty$ such that $\forall n \geq N$,
@@
 
-Where $N(\delta, a)$ is monotone increasing w.r.t. $\delta$ and $a$. To prove this, we start with the fact that $U_{(i+1)} - U_{(i)} \sim Beta(1, n)$. Then
+Where $N(\delta, a)$ is monotone increasing with respect to $\delta$ and $a$. To prove this, we start with the fact that $U_{(i+1)} - U_{(i)} \sim Beta(1, n)$. Then
 
@@
 \begin{equation}
-\begin{split}
-P(\max_i X_{(i+1)} - X_{(i)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta \\
-P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta
-\end{split}
+P(\max_i X_{(i+1)} - X_{(i)} \leq a \leq \min_{i, j} \|X_i - Y_j\|) 
+\geq 1 - \delta / 2
 \end{equation}
 
-for $X_i$ in the first segment and $Y_j$ in the second segment and $X_{(i)}$, $Y_{(j)}$ are order statistics in the first coordinate.
+and similar for $P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|)$, for $X_i$ in the first segment and $Y_j$ in the second segment and $X_{(i)}$, $Y_{(j)}$ are order statistics in the first coordinate.
 If each segment corresponds to a community, this leads to the following two results:

### 2021-05-11 proposal/proposal.tex
updated proposal

@@
 
+In this section, we will consider latent positions sampled uniformly
+from parallel unit length segments.
+
+\begin{example}
+Let $U_1, ..., U_{n_1}, V_1, ..., V_{n_2} \stackrel{\text{iid}}{\sim}Uniform(0, 1)$, $f_1(t) = (t, 0)$, and $f_2(t) = (t, a)$.
+$X_i = f_1(U_i)$ and $Y_j = f_2(V_j)$ $\forall i \in [n_1]$ and $j \in [n_2]$.
+If we observe $X_1, ..., X_{n_1}, Y_1, ..., Y_{n_2}$, what approach will allow us to group the observations by segment?
+\end{example}
+
 We will motivate an approach by the following example.
@@
 \begin{example}
-Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ with order statistics 
+Let $U_1, ..., U_n \stackrel{\text{iid}}{\sim}Uniform(0, 1)$ with order statistics 
 $U_{(1)}, ..., U_{(n)}$. Then $\forall a \in (0, 1)$ and $\delta \in (0, 1)$, $\exists N = N(\delta, a) < \infty$ such that $\forall n \geq N$,
@@
 \begin{equation}
-\begin{split}
-P(\max_i X_{(i+1)} - X_{(i)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta \\
-P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|) \geq 1 - \delta
-\end{split}
+P(\max_i X_{(i+1)} - X_{(i)} \leq a \leq \min_{i, j} \|X_i - Y_j\|) 
+\geq 1 - \delta / 2
 \end{equation}
 
-for \(X_i\) in the first segment and \(Y_j\) in the second segment and
+and similar for
+\(P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|)\), for
+\(X_i\) in the first segment and \(Y_j\) in the second segment and
 \(X_{(i)}\), \(Y_{(j)}\) are order statistics in the first coordinate.

### 2021-05-12 proposal/proposal.Rmd
updated proposal

@@
         xtable.table.placement = 'H')
+
+import::from(magrittr, `%>%`)
+library(ggplot2)
 ```
@@
 This has resulted in various challenges for researchers who rely on traditional statistical and machine learning methods, many of which are incompatible with graph data and instead require the data to exist as feature vectors in Euclidean space.
-Such challenges often involve clustering and community detection.
+Such challenges often involve clustering and community detection, which are often the goal of graph data analysis.
 Common clustering methods typically involve calculating some central or representative point for each cluster around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}).
@@
 The aim of our research is to develop consistent community detection techniques under the RDPG and GRDPG frameworks.
-First, we explore existing generative graph models with underlying community structures that can be inferred by connecting the generative models to the RDPG or GRDPG.
+First, we explore existing generative graph models with underlying community structures that can be inferred by connecting these models to the RDPG or GRDPG.
 Then we explore other latent structures or mixture distributions in the latent space that induce graphs for which consistent community detection is possible.
@@
 
-Let $G = (V, E)$ be an undirected, unweighted graph with no self-loops with $n$ vertices.
+Let $G = (V, E)$ be an undirected, unweighted graph with no self-loops and $n$ vertices.
 Denote $A \in \{0, 1\}^{n \times n}$ as the adjacency matrix for $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $A_{ij} = 0$ otherwise.
@@
 Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities.
-Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $0 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$).
-Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$, and denote $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.
+Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $1 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$).
+Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$ and $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.
 
@@
 
-Generative models for symmetric Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$.
-In order to motivate community detection methods, we restrict the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on the labels of the vertices.
-One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = B_{z_i z_j}$ and $B \in [0, 1]^{K \times K}$ with entries $B_{kl}$ denoting the edge probability of each vertex in community $k$ having an edge with each vertex in community $l$.
-The homogeneous SBM further restricts $P$ to two unique entries such that $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise.
-Multiple generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block Model (PABM).
-Like the SBM, these models involve edge probability matrix $P$ that is restricted to rank $d < n$ in part based on community labels.
+Generative models for independent Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$.
+In order to motivate community detection methods, we restrict the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on their labels.
+One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = p_{z_i z_j}$ where $p_{kl} = p_{lk}$ is the probability of an edge between each vertex in community $k$ to each vertex in community $l$.
+The homogeneous SBM further restricts $P$ to two unique entries: $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise.
+Several generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block Model (PABM).
+Like the SBM, these models involve imposing some restriction on $P$ based on community labels (and possibly other factors).
 
@@
 
-The structure of the RDPG and GRDPG provides a straightforward method for recovery of the latent positions via spectral embedding.
+The structure of the RDPG and GRDPG provides a straightforward method for recovery of the latent positions via spectral decomposition.
 
@@
 This includes the SBM, DCBM, and PABM.
-In the following example, we show that the ASE of the SBM has a very particular form.
+In the following example, we show that under the RDPG framework, the latent configuration that induces the SBM takes on a very particular form.
 
 \begin{example}[Connecting the SBM to the RDPG]
-Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the homogeneous SBM with two communities such that within-community edge probability $p$ and between-community edge probability $q$ where $p > q$.
+Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respective, and the between-community edge probability is $r$, with $p q > r^2$.
 Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$.
@@
 
-Given sufficiently large sample size $n$, the scaled ASE of affinity matrix $A$ drawn from a RDPG or GRDPG will asymptotically approach the original latent positions $X$ with probability 1, up to a linear transformation (orthogonal transformation for the RDPG, a composition of an orthogonal and scale transformations for the GRDPG).
-Thus if $X$ consists of points that lie on subspaces of $\mathbb{R}^d$, then both the scaled and unscaled ASE of $A \sim \text{RDPG}(X)$ or $A \sim \text{GRDPG}(X)$ will consist of points that lie near subspaces, with some noise that almost surely goes to $0$ as $n \to \infty$, motivating ASE followed by SSC as an asymptotically consistent method for community detection.
-The Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative graph model with underlying communities such that each community lies on a subspace.
-\citet{noroozi2019estimation} showed that SSC is able to recover the subspaces and therefore perform community detection for the PABM given $P = X I_{p, q} X^\top$, the edge probability matrix, rather than $A$, the adjacency matrix.
-Combining the results of \citeauthor{rubindelanchy2017statistical} and \citeauthor{jmlr-v28-wang13}, it should be possible to recover the communities using $A$ as well.
-
 ## Manifold Learning
@@
 Suppose $f : [0, 1] \mapsto \mathcal{X}$ such that $f$ is smooth and $\mathcal{X}$ represents a curve or one-dimensional manifold in $\mathbb{R}^d$.
-If $t_1, ..., t_n \stackrel{iid}{\sim} F$ such that $F$ has support $[0, 1]$, the latent positions are $x_i = f(t_i)$ with $y_i$ is its corresponding point in the scaled ASE, and $d_{\epsilon}(\cdot, \cdot)$ is the shortest path distance of an $\epsilon$-neighborhood graph.
+If $t_1, ..., t_n \iid F$ such that $F$ has support $[0, 1]$, the latent positions are $x_i = f(t_i)$ with $y_i$ is its corresponding point in the scaled ASE, and $d_{\epsilon}(\cdot, \cdot)$ is the shortest path distance of an $\epsilon$-neighborhood graph.
 Under certain mild conditions, the shortest path distances of the $\epsilon$-neighborhood graph of the ASE approaches the arc lengths along $f$:
@@
 As discussed in \S 2.1, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space.
-A similar result has been shown for the DCBM.
-We will now extend this to the PABM.
+A similar result has been shown for the DCBM \cite{lyzinski2014}.
+We have also extended this to the PABM.
+In \S 3.3.1, we show that the PABM is equivalent to the GRDPG model for which the communities correspond to subspaces in the latent configuration.
+We further show that while the latent configuration is not unique, there exists one in particular for which the subspaces are orthogonal.
+This leads to two straightforward methods for community detection, one involving computing inner products in the ASE and another which is a slight modification of an existing algorithm for subspace clustering.
+Finally, we will briefly explore alternative, non-spectral-based methods for community detection in the ASE involving Expectation Maximization, MCMC sampling, and Mean Field Variational Inference.
 
@@
 
-It is trivial to show that the PABM, as well as all graphs such that the adjacency matrix is drawn such that $A_{ij} \sim Bernoulli(P_{ij})$, is a special case of the GRDPG.
+It is straightforward to show that the PABM, as well as all graphs such that the adjacency matrix is drawn such that $A_{ij} \indep Bernoulli(P_{ij})$, is a special case of the GRDPG.
 It can also be shown that the latent positions of the PABM under the GRDPG framework consists of $K$ $K$-dimensional subspaces in $\mathbb{R}^{K^2}$.
@@
 
-In Theorem \ref{theorem2}, we showed that a possible latent configuration for a $K$-community PABM under the GRDPG framework consists of latent positions in $\mathbb{R}^{K^2}$ such that each community corresponds to a $K$-dimensional subspace such that each subspace is orthogonal to the others.
+In Theorem \ref{theorem2}, we showed that a possible latent configuration for a $K$-community PABM under the GRDPG framework consists of vectors in $\mathbb{R}^{K^2}$ such that each community corresponds to a $K$-dimensional subspace and each subspace is orthogonal to the others.
 
@@
 
-If $\hat{V}$ is the *unscaled* ASE of $A$, then results from  \citeauthor{rubindelanchy2017statistical} imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top$.
+If $\hat{V}$ is the *unscaled* ASE of $A$, then results from  \citet{rubindelanchy2017statistical} imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top$.
 Then for vertices $i$ and $j$ belonging to separate communities, the $ij$^th^ entry of $n \hat{V} \hat{V}^\top$ approaches 0 with probability 1:
@@
 
-In addition to OSC, since the communities are subspaces in the latent space, we can leverage existing methods for subspace clustering.
+Since the communities are subspaces in the latent space, we can also use existing methods for subspace clustering on the ASE.
 Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample.
@@
 but it could be the case that some of the subspaces are represented by 
-multiple disconnected subgraphs. SDP is contingent 
-on choosing a sufficiently large sparsity parameter $\lambda$. 
+multiple disconnected subgraphs.
 \end{remark}
 
-Since every ASE of the PABM consists of subspaces and as $n \to \infty$ each point of the ASE approaches its subspace almost surely, SSC should also work for PABM community detection.
-Combining this with the results by \citeauthor{jmlr-v28-wang13}, which state that if the points lie sufficiently close to their respective subspaces and the cosine of the angles between subspaces is sufficiently small, SDP will hold.
-We show that the unscaled ASE exhibits exactly these conditions for sufficiently large $n$.
+Since every ASE of the PABM consists of subspaces and as $n \to \infty$ each vector of the ASE approaches its subspace uniformly \cite{rubindelanchy2017statistical}, SSC is also able to perform community detection for the PABM.
+It has been shown by \citet{jmlr-v28-wang13} that if the points lie sufficiently close to their respective subspaces and the cosine of the angles between subspaces is sufficiently small, SDP will hold.
+We now state that the unscaled ASE of the PABM exhibits exactly these conditions for sufficiently large $n$.
 
@@
 positive and $K (K - 1) / 2$ most negative eigenvalues. Then 
-$\exists \lambda > 0$ and $N \in \mathbb{N}$ such that when $n > N$, 
+$\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
 $\sqrt{n} \hat{V}_n$ obeys the subspace detection property with probability 1.
@@
 
+#### Alternative Approaches to Community Detection
+
+So far, we connected the PABM to the GRDPG and used this to motivate community detection methods based on the ASE of the adjacency matrix sampled from the PABM.
+In this section, we will explore non-spectral methods for community detection.
+
+Denoting $\{\lambda_{ik}\}$ as the scalar popularity parameters for the PABM such that $\lambda^{(kl)}$ is the vector of $\lambda_{jl}$ terms for which vertex $j$ is in community $k$, we can write the likelihood for the PABM:
+
+$$p(A | z, \{\lambda_{ik}\}) = 
+\prod_{i < j} (\lambda_{i z_j} \lambda_{j z_i})^{A_{ij}} 
+(1 - \lambda_{i z_j} \lambda_{j z_i})^{1 - A_{ij}}$$
+
+\begin{equation}
+= \prod_{i < j} \prod_k^K \prod_l^K 
+(\lambda_{il} \lambda_{jk})^{A_{ij} z_{ik} z_{jl}} 
+(1 - \lambda_{il} \lambda_{jk})^{(1 - A_{ij}) z_{ik} z_{jl}}
+\end{equation}
+
+where $z_{ik} = I(Z_i = k)$.
+
+The log-likelihood is
+
+\begin{equation}
+\log p = \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
+(A_{ij} \log \lambda_{il} \lambda_{jk} + 
+(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk}))
+\end{equation}
+
+While direct maximization of the likelihood or log-likelihood is NP-complete, it is possible to apply an Expectation Maximization (EM) algorithm to this problem. To set this up,
+
 ## Proposed Generative Graph Models
 
-In the previous sections, we connected well-known and highly structured generative graph models to the RDPG and GRDPG to show that highly structured latent configurations generate graphs consistent with these models: The latent space for the SBM consists of $K$ point masses, the latent space for the DCBM consists of $K$ rays emanating from the origin, and the latent space for the PABM consists of $K$ $K$-dimensional subspaces.
+In the previous sections, we connected well-known and highly restricted generative graph models to the RDPG and GRDPG to show that highly structured latent configurations generate graphs consistent with these models: The latent space for the SBM consists of $K$ point masses, the latent space for the DCBM consists of $K$ rays emanating from the origin, and the latent space for the PABM consists of $K$ $K$-dimensional subspaces.
 In the following sections, we explore additional structured latent configurations corresponding to community structure and develop methods for community detection based on the consistency of the ASE and the structural forms of the latent configurations.
@@
 
+```{r, fig.cap = 'Points sampled uniformly on two parallel unit length line segments', fig.height = 1}
+set.seed(1)
+
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+
+a <- .1
+
+X <- cbind(runif(n1, 0, 1), rep(0, n1))
+Y <- cbind(runif(n2, 0, 1), rep(a, n2))
+
+data.matrix <- rbind(X, Y)
+z <- c(rep(1, n1), rep(2, n2))
+
+as.data.frame(data.matrix) %>% 
+  dplyr::mutate(label = factor(z)) %>% 
+  ggplot() + 
+  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0)) + 
+  geom_segment(aes(x = 0, xend = 1, y = a, yend = a)) + 
+  geom_point(aes(x = V1, y = V2, colour = label)) + 
+  labs(x = NULL, y = NULL, colour = NULL) + 
+  guides(colour = FALSE) + 
+  xlim(-.1, 1.1) + 
+  ylim(-a, 2 * a) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  coord_fixed() + 
+  ggthemes::theme_few() + 
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank())
+```
+
+```{r, eval = FALSE}
+source('~/dev/pabm-grdpg/functions.R')
+
+P <- data.matrix %*% t(data.matrix)
+A <- draw.graph(P)
+
+qgraph::qgraph(A, groups = factor(z), layout = 'spring')
+```
+
+```{r, eval = FALSE}
+eigen.A <- eigen(A)
+ind <- seq(2)
+U <- eigen.A$vectors[, ind]
+S <- diag(sqrt(abs(eigen.A$values[ind])))
+Z <- U %*% S
+pairs(Z, asp = 1, col = z)
+```
+
 We will motivate an approach by the following example.
@@
 
-If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, 1]$, then a sample of size $N(\delta, a)$ is sufficient to satisfy:
+If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, 1]$, then a sample of size $\min(n_1, n_2) \geq N(\delta, a)$ is sufficient to satisfy:
 

### 2021-05-12 proposal/proposal.tex
updated proposal

@@
 \begin{center}
-\LARGE{\textbf{Community Detection Methods for Random Dot Product Graphs and Generalized Random Dot Product Graphs}}\\
+\LARGE{\textbf{Community Detection Methods for the Generalized Random Dot Product Graph Model}}\\
 \vspace*{2\baselineskip}
@@
 exist as feature vectors in Euclidean space. Such challenges often
-involve clustering and community detection. Common clustering methods
-typically involve calculating some central or representative point for
-each cluster around which the data belonging to that cluster lie (e.g.,
-Lloyd's algorithm for \(K\)-means clustering \cite{1056489}, Gaussian
-Mixture Models \cite{doi:10.1198/016214502760047131}). Because these
-methods involve computing summary statistics within each cluster, such
-as the sample average, they cannot be applied directly to graphs,
-necessitating methods for transforming the graph data into feature
-vectors.
+involve clustering and community detection, which are often the goal of
+graph data analysis. Common clustering methods typically involve
+calculating some central or representative point for each cluster around
+which the data belonging to that cluster lie (e.g., Lloyd's algorithm
+for \(K\)-means clustering \cite{1056489}, Gaussian Mixture Models
+\cite{doi:10.1198/016214502760047131}). Because these methods involve
+computing summary statistics within each cluster, such as the sample
+average, they cannot be applied directly to graphs, necessitating
+methods for transforming the graph data into feature vectors.
 
@@
 existing generative graph models with underlying community structures
-that can be inferred by connecting the generative models to the RDPG or
-GRDPG. Then we explore other latent structures or mixture distributions
-in the latent space that induce graphs for which consistent community
-detection is possible.
+that can be inferred by connecting these models to the RDPG or GRDPG.
+Then we explore other latent structures or mixture distributions in the
+latent space that induce graphs for which consistent community detection
+is possible.
 
@@
 Let \(G = (V, E)\) be an undirected, unweighted graph with no self-loops
-with \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
+and \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
 adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
@@
 \(A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})\) for
-each \(0 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)).
+each \(1 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)).
 Finally, we denote
 \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
-as the sample \(x_1, ..., x_n \in \mathbb{R}^d\), and denote
+as the sample \(x_1, ..., x_n \in \mathbb{R}^d\) and
 \(z_1, ..., z_n \in [K]\) as their corresponding (hidden) labels.
@@
 
-Generative models for symmetric Bernoulli graphs involve defining the
+Generative models for independent Bernoulli graphs involve defining the
 edge probability matrix \(P\) whose \(ij\)\textsuperscript{th} entry is
@@
 restrict the generative model such that for each pair of vertices, the
-probability of an edge between the vertices is conditioned on the labels
-of the vertices. One such model is the Stochastic Block Model (SBM)
+probability of an edge between the vertices is conditioned on their
+labels. One such model is the Stochastic Block Model (SBM)
 \cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
 each vertex belonging to one community, the SBM restricts \(P\) to
-\(K (K + 1) / 2\) unique entries such that \(P_{ij} = B_{z_i z_j}\) and
-\(B \in [0, 1]^{K \times K}\) with entries \(B_{kl}\) denoting the edge
-probability of each vertex in community \(k\) having an edge with each
-vertex in community \(l\). The homogeneous SBM further restricts \(P\)
-to two unique entries such that \(P_{ij} = p\) if \(z_i = z_j\) and
-\(P_{ij} = q\) otherwise. Multiple generalizations of the SBM have been
-introduced since, including the Degree Corrected Block Model (DCBM) and
-the Popularity Adjusted Block Model (PABM). Like the SBM, these models
-involve edge probability matrix \(P\) that is restricted to rank
-\(d < n\) in part based on community labels.
+\(K (K + 1) / 2\) unique entries such that \(P_{ij} = p_{z_i z_j}\)
+where \(p_{kl} = p_{lk}\) is the probability of an edge between each
+vertex in community \(k\) to each vertex in community \(l\). The
+homogeneous SBM further restricts \(P\) to two unique entries:
+\(P_{ij} = p\) if \(z_i = z_j\) and \(P_{ij} = q\) otherwise. Several
+generalizations of the SBM have been introduced since, including the
+Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block
+Model (PABM). Like the SBM, these models involve imposing some
+restriction on \(P\) based on community labels (and possibly other
+factors).
 
@@
 The structure of the RDPG and GRDPG provides a straightforward method
-for recovery of the latent positions via spectral embedding.
+for recovery of the latent positions via spectral decomposition.
 
@@
 of the GRDPG. This includes the SBM, DCBM, and PABM. In the following
-example, we show that the ASE of the SBM has a very particular form.
+example, we show that under the RDPG framework, the latent configuration
+that induces the SBM takes on a very particular form.
 
 \begin{example}[Connecting the SBM to the RDPG]
-Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the homogeneous SBM with two communities such that within-community edge probability $p$ and between-community edge probability $q$ where $p > q$.
+Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respective, and the between-community edge probability is $r$, with $p q > r^2$.
 Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$.
@@
 
-Given sufficiently large sample size \(n\), the scaled ASE of affinity
-matrix \(A\) drawn from a RDPG or GRDPG will asymptotically approach the
-original latent positions \(X\) with probability 1, up to a linear
-transformation (orthogonal transformation for the RDPG, a composition of
-an orthogonal and scale transformations for the GRDPG). Thus if \(X\)
-consists of points that lie on subspaces of \(\mathbb{R}^d\), then both
-the scaled and unscaled ASE of \(A \sim \text{RDPG}(X)\) or
-\(A \sim \text{GRDPG}(X)\) will consist of points that lie near
-subspaces, with some noise that almost surely goes to \(0\) as
-\(n \to \infty\), motivating ASE followed by SSC as an asymptotically
-consistent method for community detection. The Popularity Adjusted Block
-Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1} is a generative
-graph model with underlying communities such that each community lies on
-a subspace. \citet{noroozi2019estimation} showed that SSC is able to
-recover the subspaces and therefore perform community detection for the
-PABM given \(P = X I_{p, q} X^\top\), the edge probability matrix,
-rather than \(A\), the adjacency matrix. Combining the results of
-\citeauthor{rubindelanchy2017statistical} and
-\citeauthor{jmlr-v28-wang13}, it should be possible to recover the
-communities using \(A\) as well.
-
 \hypertarget{manifold-learning}{%
@@
 \(\mathcal{X}\) represents a curve or one-dimensional manifold in
-\(\mathbb{R}^d\). If \(t_1, ..., t_n \stackrel{iid}{\sim} F\) such that
-\(F\) has support \([0, 1]\), the latent positions are \(x_i = f(t_i)\)
-with \(y_i\) is its corresponding point in the scaled ASE, and
-\(d_{\epsilon}(\cdot, \cdot)\) is the shortest path distance of an
-\(\epsilon\)-neighborhood graph. Under certain mild conditions, the
+\(\mathbb{R}^d\). If \(t_1, ..., t_n \stackrel{\text{iid}}{\sim}F\) such
+that \(F\) has support \([0, 1]\), the latent positions are
+\(x_i = f(t_i)\) with \(y_i\) is its corresponding point in the scaled
+ASE, and \(d_{\epsilon}(\cdot, \cdot)\) is the shortest path distance of
+an \(\epsilon\)-neighborhood graph. Under certain mild conditions, the
 shortest path distances of the \(\epsilon\)-neighborhood graph of the
@@
 with a very specific structure in the latent space. A similar result has
-been shown for the DCBM. We will now extend this to the PABM.
+been shown for the DCBM \cite{lyzinski2014}. We have also extended this
+to the PABM. In \S 3.3.1, we show that the PABM is equivalent to the
+GRDPG model for which the communities correspond to subspaces in the
+latent configuration. We further show that while the latent
+configuration is not unique, there exists one in particular for which
+the subspaces are orthogonal. This leads to two straightforward methods
+for community detection, one involving computing inner products in the
+ASE and another which is a slight modification of an existing algorithm
+for subspace clustering. Finally, we will briefly explore alternative,
+non-spectral-based methods for community detection in the ASE involving
+Expectation Maximization, MCMC sampling, and Mean Field Variational
+Inference.
 
@@
 
-It is trivial to show that the PABM, as well as all graphs such that the
-adjacency matrix is drawn such that \(A_{ij} \sim Bernoulli(P_{ij})\),
-is a special case of the GRDPG. It can also be shown that the latent
-positions of the PABM under the GRDPG framework consists of \(K\)
-\(K\)-dimensional subspaces in \(\mathbb{R}^{K^2}\). While there is no
-unique latent configuration \(X\) such that \(X X^\top = P\), the edge
-probability \(P\) for the PABM, they all have this subspace structure,
-and one in particular consists of \emph{orthogonal} subspaces.
+It is straightforward to show that the PABM, as well as all graphs such
+that the adjacency matrix is drawn such that
+\(A_{ij} \stackrel{\text{indep}}{\sim}Bernoulli(P_{ij})\), is a special
+case of the GRDPG. It can also be shown that the latent positions of the
+PABM under the GRDPG framework consists of \(K\) \(K\)-dimensional
+subspaces in \(\mathbb{R}^{K^2}\). While there is no unique latent
+configuration \(X\) such that \(X X^\top = P\), the edge probability
+\(P\) for the PABM, they all have this subspace structure, and one in
+particular consists of \emph{orthogonal} subspaces.
 
@@
 configuration for a \(K\)-community PABM under the GRDPG framework
-consists of latent positions in \(\mathbb{R}^{K^2}\) such that each
-community corresponds to a \(K\)-dimensional subspace such that each
-subspace is orthogonal to the others.
+consists of vectors in \(\mathbb{R}^{K^2}\) such that each community
+corresponds to a \(K\)-dimensional subspace and each subspace is
+orthogonal to the others.
 
@@
 If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), then results from
-\citeauthor{rubindelanchy2017statistical} imply
+\citet{rubindelanchy2017statistical} imply
 \(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top\). Then for
@@
 
-In addition to OSC, since the communities are subspaces in the latent
-space, we can leverage existing methods for subspace clustering. Of
-particular interest is Sparse Subspace Clustering (SSC), which is
-performed by solving an optimization problem for each observed point in
-a sample. Given \(X \in \mathbb{R}^{n \times d}\) with vectors
+Since the communities are subspaces in the latent space, we can also use
+existing methods for subspace clustering on the ASE. Of particular
+interest is Sparse Subspace Clustering (SSC), which is performed by
+solving an optimization problem for each observed point in a sample.
+Given \(X \in \mathbb{R}^{n \times d}\) with vectors
 \(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
@@
 but it could be the case that some of the subspaces are represented by 
-multiple disconnected subgraphs. SDP is contingent 
-on choosing a sufficiently large sparsity parameter $\lambda$. 
+multiple disconnected subgraphs.
 \end{remark}
@@
 Since every ASE of the PABM consists of subspaces and as
-\(n \to \infty\) each point of the ASE approaches its subspace almost
-surely, SSC should also work for PABM community detection. Combining
-this with the results by \citeauthor{jmlr-v28-wang13}, which state that
-if the points lie sufficiently close to their respective subspaces and
-the cosine of the angles between subspaces is sufficiently small, SDP
-will hold. We show that the unscaled ASE exhibits exactly these
-conditions for sufficiently large \(n\).
+\(n \to \infty\) each vector of the ASE approaches its subspace
+uniformly \cite{rubindelanchy2017statistical}, SSC is also able to
+perform community detection for the PABM. It has been shown by
+\citet{jmlr-v28-wang13} that if the points lie sufficiently close to
+their respective subspaces and the cosine of the angles between
+subspaces is sufficiently small, SDP will hold. We now state that the
+unscaled ASE of the PABM exhibits exactly these conditions for
+sufficiently large \(n\).
 
@@
 positive and $K (K - 1) / 2$ most negative eigenvalues. Then 
-$\exists \lambda > 0$ and $N \in \mathbb{N}$ such that when $n > N$, 
+$\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
 $\sqrt{n} \hat{V}_n$ obeys the subspace detection property with probability 1.
 \end{theorem}
+
+\hypertarget{alternative-approaches-to-community-detection}{%
+\paragraph{Alternative Approaches to Community
+Detection}\label{alternative-approaches-to-community-detection}}
+
+So far, we connected the PABM to the GRDPG and used this to motivate
+community detection methods based on the ASE of the adjacency matrix
+sampled from the PABM. In this section, we will explore non-spectral
+methods for community detection.
+
+Denoting \(\{\lambda_{ik}\}\) as the scalar popularity parameters for
+the PABM such that \(\lambda^{(kl)}\) is the vector of \(\lambda_{jl}\)
+terms for which vertex \(j\) is in community \(k\), we can write the
+likelihood for the PABM:
+
+\[p(A | z, \{\lambda_{ik}\}) = 
+\prod_{i < j} (\lambda_{i z_j} \lambda_{j z_i})^{A_{ij}} 
+(1 - \lambda_{i z_j} \lambda_{j z_i})^{1 - A_{ij}}\]
+
+\begin{equation}
+= \prod_{i < j} \prod_k^K \prod_l^K 
+(\lambda_{il} \lambda_{jk})^{A_{ij} z_{ik} z_{jl}} 
+(1 - \lambda_{il} \lambda_{jk})^{(1 - A_{ij}) z_{ik} z_{jl}}
+\end{equation}
+
+where \(z_{ik} = I(Z_i = k)\).
+
+The log-likelihood is
+
+\begin{equation}
+\log p = \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
+(A_{ij} \log \lambda_{il} \lambda_{jk} + 
+(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk}))
+\end{equation}
+
+While direct maximization of the likelihood or log-likelihood is
+NP-complete, it is possible to apply an Expectation Maximization (EM)
+algorithm to this problem.
 
@@
 
-In the previous sections, we connected well-known and highly structured
+In the previous sections, we connected well-known and highly restricted
 generative graph models to the RDPG and GRDPG to show that highly
@@
 
+\begin{figure}[H]
+
+{\centering \includegraphics{proposal_files/figure-latex/unnamed-chunk-1-1}
+
+}
+
+\caption{Points sampled uniformly on two parallel unit length line segments}\label{fig:unnamed-chunk-1}
+\end{figure}
+
 We will motivate an approach by the following example.
@@
 uniformly from the segment \(f_2(t) = (t, a)\) for \(t \in [0, 1]\),
-then a sample of size \(N(\delta, a)\) is sufficient to satisfy:
+then a sample of size \(\min(n_1, n_2) \geq N(\delta, a)\) is sufficient
+to satisfy:
 

### 2021-05-17 proposal/proposal.Rmd
uploading slides

@@
 
-# Literature Review
+# Preliminaries
 
-## Generative Graph Models and Community Detection
+## Block Models
 
@@
 
-# Proposed Research
-
-## Connecting Existing Generative Graph Models to the RDPG and GRDPG
+# The Popularity Adjusted Block Model
 
@@
 A similar result has been shown for the DCBM \cite{lyzinski2014}.
-We have also extended this to the PABM.
+We can also extend this to the PABM.
 In \S 3.3.1, we show that the PABM is equivalent to the GRDPG model for which the communities correspond to subspaces in the latent configuration.
@@
 
-### Popularity Adjusted Block Model
+## Connecting the PABM to the GRDPG
 
@@
 Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample.
-Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
+Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
 The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$.
@@
 
-#### Alternative Approaches to Community Detection
+## Alternative Approaches to Community Detection
 
@@
 
-## Proposed Generative Graph Models
+# Generalizations
 
@@
 
-### Affine Segments
+## Affine Segments
 
@@
 
-### Mainfold Learning
+## Manifolds
 
@@
 
-Simulations and real data analyses: March 2022
+Simulations, real data analyses, R package: March 2022
 

### 2021-05-17 proposal/proposal.tex
uploading slides

@@
 
-\hypertarget{literature-review}{%
-\section{Literature Review}\label{literature-review}}
+\hypertarget{preliminaries}{%
+\section{Preliminaries}\label{preliminaries}}
 
-\hypertarget{generative-graph-models-and-community-detection}{%
-\subsection{Generative Graph Models and Community
-Detection}\label{generative-graph-models-and-community-detection}}
+\hypertarget{block-models}{%
+\subsection{Block Models}\label{block-models}}
 
@@
 
-\hypertarget{proposed-research}{%
-\section{Proposed Research}\label{proposed-research}}
-
-\hypertarget{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}{%
-\subsection{Connecting Existing Generative Graph Models to the RDPG and
-GRDPG}\label{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}}
+\hypertarget{the-popularity-adjusted-block-model}{%
+\section{The Popularity Adjusted Block
+Model}\label{the-popularity-adjusted-block-model}}
 
@@
 with a very specific structure in the latent space. A similar result has
-been shown for the DCBM \cite{lyzinski2014}. We have also extended this
-to the PABM. In \S 3.3.1, we show that the PABM is equivalent to the
-GRDPG model for which the communities correspond to subspaces in the
-latent configuration. We further show that while the latent
-configuration is not unique, there exists one in particular for which
-the subspaces are orthogonal. This leads to two straightforward methods
-for community detection, one involving computing inner products in the
-ASE and another which is a slight modification of an existing algorithm
-for subspace clustering. Finally, we will briefly explore alternative,
+been shown for the DCBM \cite{lyzinski2014}. We can also extend this to
+the PABM. In \S 3.3.1, we show that the PABM is equivalent to the GRDPG
+model for which the communities correspond to subspaces in the latent
+configuration. We further show that while the latent configuration is
+not unique, there exists one in particular for which the subspaces are
+orthogonal. This leads to two straightforward methods for community
+detection, one involving computing inner products in the ASE and another
+which is a slight modification of an existing algorithm for subspace
+clustering. Finally, we will briefly explore alternative,
 non-spectral-based methods for community detection in the ASE involving
@@
 
-\hypertarget{popularity-adjusted-block-model}{%
-\subsubsection{Popularity Adjusted Block
-Model}\label{popularity-adjusted-block-model}}
+\hypertarget{connecting-the-pabm-to-the-grdpg}{%
+\subsection{Connecting the PABM to the
+GRDPG}\label{connecting-the-pabm-to-the-grdpg}}
 
@@
 \(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
-\(c_i = \min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
+\(c_i = \arg\min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
 \(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
@@
 \hypertarget{alternative-approaches-to-community-detection}{%
-\paragraph{Alternative Approaches to Community
+\subsection{Alternative Approaches to Community
 Detection}\label{alternative-approaches-to-community-detection}}
@@
 NP-complete, it is possible to apply an Expectation Maximization (EM)
-algorithm to this problem.
+algorithm to this problem. To set this up,
 
-\hypertarget{proposed-generative-graph-models}{%
-\subsection{Proposed Generative Graph
-Models}\label{proposed-generative-graph-models}}
+\hypertarget{generalizations}{%
+\section{Generalizations}\label{generalizations}}
 
@@
 \hypertarget{affine-segments}{%
-\subsubsection{Affine Segments}\label{affine-segments}}
+\subsection{Affine Segments}\label{affine-segments}}
 
@@
 
-\hypertarget{mainfold-learning}{%
-\subsubsection{Mainfold Learning}\label{mainfold-learning}}
+\hypertarget{manifolds}{%
+\subsection{Manifolds}\label{manifolds}}
 

### 2021-05-17 proposal/slides.Rmd
uploading slides

---
title: Community Detection Methods for the Generalized Random Dot Product Graph Model
subtitle: Dissertation Proposal Defense
author: John Koo
date: 'TBD'
output: 
  beamer_presentation:
    fig_crop: no
    theme: 'default'
    colortheme: 'beaver'
    includes:
      in_header: page_headers.tex
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      fig.lp = '')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
library(ggplot2)
import::from(magrittr, `%>%`)
theme_set(theme_bw())

source('~/dev/pabm-grdpg/functions.R')
set.seed(314159)
```

## Overview

* Preliminaries
  * Block Models and Community Detection
  * (Generalized) Random Dot Product Graphs
* Connecting Block Models to (G)RDPG Models
  * Popularity Adjusted Block Model
  * Subspace Clustering
* Community Detection in the (G)RDPG

# Preliminaries

## Bernoulli Graphs

<style type="text/css">
.caption {
    font-size: x-small;
}
</style>

$A \sim BernoulliGraph(P)$ iff:

* $G = (V, E)$
  * Undirected, unweighted, no self-loops
  * $|V| = n$
  * $|E| \leq n (n - 1) / 2$
* Adjacency matrix $A \in \{0, 1\}^{n \times n}$
  * $A_{ij} = \begin{cases} 
  1 & \exists \text{ edge between } i \text{ and } j \\
  0 & \text{else}
  \end{cases}$
* Edge probability matrix $P \in [0, 1]^{n \times n}$
* $A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ for $i < j$

```{r, fig.height = 1, fig.width = 2}
n <- 2 ** 5
p <- 1 / log(n)
P <- matrix(p, nrow = n, ncol = n)
A <- draw.graph(P)
qgraph::qgraph(A, vsize = 4)
```

## Block Models

* $A \sim BernoulliGraph(P)$
  * (Hidden) labels $z_1, ..., z_n \in [K]$
  * $P_{ij} = f(z_i, z_j, \cdot)$

* Example: Stochastic Block Model with two communities
    * $K = 2$
  * $P_{ij} = \begin{cases} 
  p & z_i = z_j = 1 \\
  q & z_i = z_j = 2 \\
  r & z_i \neq z_j
  \end{cases}$

```{r, fig.height = 1.5, fig.width = 4}
n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/4
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
A <- draw.graph(P)
qgraph::qgraph(A, vsize = 4, groups = factor(z))
```

## Block Models

* Erdos-Renyi Model (1959)
  * $P_{ij} = \theta$
  * Not a block model
* Stochastic Block Model (Lorrain and White, 1971)
  * $P_{ij} = \theta_{z_i z_j}$
  * $K (K + 1) / 2$ parameters $\theta_{kl}$
* Degree Corrected Block Model (Karrer and Newman, 2011)
  * $P_{ij} = \theta_{z_i z_j} \omega_i \omega_j$
  * $K (K + 1) / 2 + n$ parameters $\theta_{kl}$, $\omega_i$
* Popularity Adjusted Block Model (Sengupta and Chen, 2017)
  * $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
  * $K n$ parameters $\lambda_{ik}$

## Heirarchy of Block Models

* PABM $\to$ DCBM: $\lambda_{ik} = \sqrt{\theta_{z_i k}} \omega_i$
* DCBM $\to$ SBM: $\omega_i = 1$
* SBM $\to$ Erdos-Renyi: $\theta_{kl} = \theta$

\begin{center}
```{r, out.width = '150px'}
knitr::include_graphics('noroozi-pensky-heirarchy.png')
```

\tiny{Majid Noroozi and Marianna Pensky. "The Heirarchy of Block Models". March 2021.}
\end{center}

## Community Detection in Block Models

Likelihood

$$L = \prod_{i<j} \prod_{k, l}^K 
\big(p_{k, l, i, j}^{A_{ij}} 
(1 - p_{k, l, i, j})^{1 - A_{ij}} \big)^{z_{ik} z_{jl}}$$

* Example: DCBM ($p_{k, l, i, j} = \theta_{kl} \omega_i \omega_j$)

$$L = \prod_{i<j} \prod_{k, l}^K 
\big((\theta_{kl} \omega_i \omega_j)^{A_{ij}} 
(1 - \theta_{kl} \omega_i \omega_j)^{1 - A_{ij}} \big)^{z_{ik} z_{jl}}$$

* ML method for community detection: $\hat{\vec{z}} = \arg\max_{\vec{z}} L$

* NP-complete
  * Expectation-Maximization
  * Bayesian methods
  * **Spectral methods**

## (Generalized) Random Dot Product Graph Model

* Random Dot Product Graph $A \sim RDPG(X)$
  * Latent vectors $x_1, ..., x_n \in \mathcal{X} \subset \mathbb{R}^d$
    * $\mathcal{X} = \{x, y : 0 \leq x^\top y \leq 1\}$
  * Data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
  * Edge probability matrix $P = X X^\top$
  * Adjacency matrix $A \sim BernoulliGraph(P)$

* Generalized Random Dot Product Graph $A \sim GRDPG_{p, q}(X)$
  * Latent vectors $x_1, ..., x_n \in \mathcal{X} \subset \mathbb{R}^d$
    * $\mathcal{X} = \{x, y : 0 \leq x^\top I_{p, q} y \leq 1\}$
    * $I_{p, q} = blockdiag(I_p, -I_q)$
    * $p + q = d$
  * Data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
  * Edge probability matrix $P = X I_{pq} X^\top$
  * Adjacency matrix $A \sim BernoulliGraph(P)$

* If $X_1, ..., X_n \stackrel{iid}{\sim} F$, then $(A, X) \sim RDPG(F, n)$ or 
$(A, X) \sim GRDPG_{p, q}(F, n)$

## (Generalized) Random Dot Product Graph Model

### Recovery/Estimation

* Want to estimate $X$ given $A$
  * Alternatively, recover some property of $X$ given $A$ 
    * Interpoint distances
    * Inner products
    * Angles

* Adjacency Spectral Embedding
  * Given embedding dimension $d$, 
  $A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$
    * If RDPG, use $d$ greatest eigenvalues
    * If GRDPG, use $p$ most positive and $q$ most negative eigenvalues
    * $\hat{V} \in \mathbb{R}^{n \times d}$
    * $\hat{\Lambda} \in \mathbb{R}^{d \times d}$
  * RDPG: $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
  * GRDPG: $\hat{X} = \hat{V} |\hat{\Lambda}|^{1/2}$

* RDPG: $\max\limits_i \|\hat{X}_i - W_n X_i \| \stackrel{a.s.}{\to} 0$
(Athreya et al.)
* GRDPG: 
$\max\limits_i \|\hat{X}_i - Q_n X_i \| \stackrel{a.s.}{\to} 0$
(Rubin-Delanchy et al.)

## Connecting Block Models to the (G)RDPG Model

* All $G$ with $A \sim BernoulliGraph(P)$ are RDPG (if $P$ is 
positive semidefinite) or GRDPG
  * Includes all block models

* Example: SBM

$$X = \begin{bmatrix} 
\sqrt{p} & 0 \\
\vdots & \vdots \\
\sqrt{p} & 0 \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\
\vdots & \vdots \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
\end{bmatrix}$$

$$P = X X^\top$$

## Connecting Block Models to the (G)RDPG Model

Example: SBM (cont'd)

* $A \sim BernoulliGraph(X X^\top)$
* $A \approx \hat{X} \hat{X}^\top$
  * $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
* Apply clustering algorithm (e.g., $K$-means) on $\hat{X}$

```{r, fig.height = 8, fig.width = 8, out.width = '50%'}
A.eigen <- eigen(A)
X.hat <- A.eigen$vectors[, 1:2] %*% diag(A.eigen$values[1:2] ** .5)
plot(X.hat, asp = 1, col = z, xlab = NA, ylab = NA)
```

## Connecting Block Models to the (G)RDPG Model

```{r, out.width = '20%', fig.height = 2.5, fig.width = 2.5}
par(mar = rep(1.75, 4))
P.eigen = eigen(P)
X <- P.eigen$vectors[, 1:2] %*% diag(P.eigen$values[1:2] ** .5)
plot(X, asp = 1, col = z, xlab = NA, ylab = NA, main = 'SBM: Point masses')
```

```{r, out.width = '20%', fig.height = 2.5, fig.width = 2.5}
par(mar = rep(1.75, 4))
omega <- rbeta(n, 1, 1)
dc.matrix <- omega %*% t(omega)
P.dcbm <- P * dc.matrix
dcbm.eigen <- eigen(P.dcbm)
X.dcbm <- dcbm.eigen$vectors[, 1:2] %*% diag(dcbm.eigen$values[1:2] ** .5)
plot(X.dcbm, asp = 1, col = z, xlab = NA, ylab = NA, main = 'DCBM: Rays')

```

```{r, out.width = '33%', fig.height = 4, fig.width = 4}
par(mar = rep(1.75, 4))
Pz <- generate.P.beta(n)
P <- Pz$P
z <- Pz$clustering
X <- embedding(P)
pairs(X, col = z, asp = 1, pch = '.', main = 'PABM: Subspaces')
```

# Popularity Adjusted Block Model

## Popularity Adjusted Block Model (Reparameterization)

$A \sim PABM(\{\lambda^{(kl)}\}_K)$ iff

* w.l.o.g., organize $P$ such that each block 
$P^{(kl)} \in [0, 1]^{n_k \times n_l}$ contains edge 
probabilities between communities $k$ and $l$
* Popularity vectors $\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
are the popularity parameters of members of community $k$ to community $l$
* $\{\lambda^{(kl)}\}_K$ is the set of $K^2$ popularity vectors
* $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$
* $A \sim BernoulliGraph(P)$

## Connecting the PABM to the GRDPG ($K = 2$)

Theorem: $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
$A \sim GRDPG_{3, 1}(X U)$ for $X$ constructed from $\{\lambda^{(kl)}\}_2$ and 
$U \in \mathbb{O}(4)$

Proof:

$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix}$$

$$Y = \begin{bmatrix}
\lambda^{(11)} & 0 & \lambda^{(12)} & 0 \\
0 & \lambda^{(21)} & 0 & \lambda^{(22)}
\end{bmatrix}$$

$$P = X Y^\top$$

## Connecting the PABM to the GRDPG ($K = 2$)

Proof (cont'd):

$$Y = X \Pi$$

$$\Pi = \begin{bmatrix} 
1 & 0 & 0 & 0 \\ 
0 & 0 & 1 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix} = 
U I_{3, 1} U^\top$$

$$U = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & -1 / \sqrt{2} \\
0 & 1 & 0 & 0
\end{bmatrix}$$

$$P = (X U) I_{3, 1} (X U)^\top$$

## Connecting the PABM to the GRDPG

**Theorem**: $A \sim PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
$A \sim GRDPG_{p, q}(X U)$ such that

* $p = K (K + 1) / 2$
* $q = K (K - 1) / 2$
* $U$ is orthogonal
* $X$ is block diagonal

* Non-uniqueness: 
$A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ for 
$Q \in \mathbb{O}(p, q)$

## Orthogonal Spectral Clustering

**Theorem**: If $P = V \Lambda V^\top$ and $B = n V V^\top$, then $B_{ij} = 0$ 
$\forall i, j$ in different communities.

Orthogonalized Spectral Clustering algorithm:

* Input: Adjacency matrix $A$, number of communities $K$
* Output: Community assignments $1, ..., K$

1. Compute the eigenvectors of $A$ that correspond to the $K (K+1)/2$ most 
positive and $K (K-1) / 2$ most negative eigenvalues to construct $V$.
2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
3. Construct graph $G$ using $B$ as its similarity matrix.
4. Partition $G$ into $K$ disconnected subgraphs (e.g., using edge thresholding 
or spectral clustering).
5. Map each partition to the community labels $1, ..., K$.

## Orthogonal Spectral Clustering

**Theorem**: Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity 
matrix from OSC. 
Then $\forall$ pairs $(i, j)$ belonging to different communities 
and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,

$$
\max_{i, j} \hat{B}^{(ij)} = 
O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
$$

## Sparse Subspace Clustering

* $X$ is block diagonal and $U$ is orthogonal $\implies$ each community 
corresponds to a subspace in $\mathbb{R}^d$.
* Subspace property holds even with linear transformation 
$Q \in \mathbb{O}(p, q)$.
* If $P = V \Lambda V^\top$, then $V$ consists of *orthogonal* subspaces.

Sparse Subspace Clustering algorithm:

1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
subject to $x_i = X_{-i} c$ and $c_i^{(i)} = 0$.
2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$
3. Construct affinity matrix $B = |C| + C^\top|$

* If $X$ obeys the Subspace Detection Property, then $B$ is sparse such that 
$B_{ij} = 0$ if $i$ and $j$ belong to different communities and $\|c_i\| > 0$.
* Step (1) of SSC typically performed via LASSO: 
$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$

## Sparse Subspace Clustering

**Theorem**:

Let

* $P_n$ describe the edge probability matrix of the PABM with 
$n$ vertices
* $A_n \sim BernoulliGraph(P_n)$
* $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
$K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues.

Then

* $\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
$\sqrt{n} \hat{V}_n$ obeys the Subspace Detection Property with probability 1.

Remarks:

* For large $n$, we can identify $\lambda$ for SDP (Wang and Xu, 2016).
* SDP does not guarantee community detection.

# Generalizations to Community Detection for the (G)RDPG

## Generative Model

Let $(A, X) \sim RDPG(F, n)$ such that

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_k(t) \neq f_l(t)$ 
$\forall k, l \in [K]$. 
2. Sample labels
$Z_1, ..., Z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
3. Sample $T_1, ..., T_n \stackrel{iid}{\sim} D$ with support $[0, 1]$. 
4. Set latent positions $X_i = f_{Z_i}(T_i)$ and 
$X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
5. $A \sim BernoulliGraph(X X^\top)$

## Community Detection

* Athreya et al. and Rubin-Delanchy et al.: we can approximate properties of 
the latent configurations via ASE.
* General community detection method: Given $A$, $K$, and $d$ (or $p$ and $q$), 
  1. Use ASE to approximate the latent configuration.
  2. Use the appropriate clustering algorithm for the form of the latent
  configuration (manifolds).

## Parallel Segments

**Example**: 
Let $U_1, ..., U_{n_1}, U_{n_1 + 1}, ..., U_{n} \stackrel{iid}{\sim} Uniform(0, \cos \frac{\pi}{2} a)$, $f_1(t) = (t, 0)$, and $f_2(t) = (t, a)$. $X_i = f_1(U_i)$ for $i \leq n_1$ and $X_j = f_2(U_j)$ for $n_1 + 1 \leq j \leq n$. If we observe $X_1, ..., X_{n_1}, X_{n_1+1}, ..., X_{n}$, what approach will allow us to group the observations by segment?

```{r fig.height = 3, out.width = '50%'}
set.seed(1)

n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2

a <- .1

X <- cbind(runif(n1, 0, cos(pi / 2 * a)), rep(0, n1))
Y <- cbind(runif(n2, 0, cos(pi / 2 * a)), rep(a, n2))

data.matrix <- rbind(X, Y)
z <- c(rep(1, n1), rep(2, n2))

as.data.frame(data.matrix) %>% 
  dplyr::mutate(label = factor(z)) %>% 
  ggplot() + 
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .1) + 
  geom_segment(aes(x = 0, xend = 1, y = a, yend = a), size = .1) + 
  geom_point(aes(x = V1, y = V2, colour = label)) + 
  labs(x = NULL, y = NULL, colour = NULL) + 
  guides(colour = FALSE) + 
  xlim(-.1, 1.1) + 
  ylim(-a, 2 * a) + 
  scale_colour_brewer(palette = 'Set1') + 
  coord_fixed() + 
  # ggthemes::theme_few() + 
  theme_void() + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())
```

$\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
$\exists N(a, \delta, K) < \infty$ such that when $\min_k n_k \geq N$, 
with probability at least $1 - \delta$,

1. Single linkage clustering will produce perfect community detection.
2. An $\epsilon$-neighborhood graph with $\epsilon \in (0, a)$ will consist of 
at least $K$ disjoint subgraphs such that no subgraph contains members of two 
different communities.

## Noisy Parallel Segments and One-Dimensional Manifolds

**Example**: 
Starting with the parallel segments as before, suppose instead of observing $X_1, ..., X_n$, we have noisy observations $X_1 + \xi_1, ..., X_n + \xi_n$ such that 
$\max_i \|\xi_i\| = \xi \leq a / 3$.

```{r fig.height = 3, out.width = '50%'}
xi1 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
xi2 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))

as.data.frame(data.matrix) %>% 
  dplyr::mutate(V1 = V1 + xi1, V2 = V2 + xi2) %>% 
  dplyr::mutate(label = factor(z)) %>% 
  ggplot() + 
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .1) + 
  geom_segment(aes(x = 0, xend = 1, y = a, yend = a), size = .1) + 
  geom_point(aes(x = V1, y = V2, colour = label)) + 
  labs(x = NULL, y = NULL, colour = NULL) + 
  guides(colour = FALSE) + 
  xlim(-.1, 1.1) + 
  ylim(-a, 2 * a) + 
  scale_colour_brewer(palette = 'Set1') + 
  coord_fixed() + 
  # ggthemes::theme_few() + 
  theme_void() + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())
```

Then $\forall a \in (0, 1)$, $\delta \in (0, 1)$, $K \geq 2$, and $\xi \leq a / 3$, 
$\exists N(a, \delta, K, \xi) < \infty$ such that when $\min_k n_k \geq N$, 
with probability at least $1 - \delta$,

1. Single linkage will produce perfect community detection.
2. An $\epsilon$-neighborhood graph will consist of at least $K$ sub-graphs with no subgraph containing vertices of two communities.

This also holds for noisy points sampled on one-dimensional manifolds such that the manifolds are distance at least $a$ apart.

## Future Work

1. Show that the ASE of a random graph generated by these latent vectors 
produces the correct conditions for sufficiently large $n$.
2. Extend results to non-uniform distributions.
3. Extend results to multidimensional manifolds.
4. Relax condition for the minimum distance between manifolds.
5. Explore additional clustering techniques for these latent configurations.

# Additional Slides

## Expectation Maximization for the PABM

## MCMC Sampling for the PABM

## Variational Inference for the PABM

### 2021-05-17 proposal/slides.Rmd
updated slides

@@
   * (Generalized) Random Dot Product Graphs
-* Connecting Block Models to (G)RDPG Models
-  * Popularity Adjusted Block Model
+  * Connecting Block Models to the (G)RDPG
+* The Popularity Adjusted Block Model
+  * Connecting the PABM to the GRDPG
   * Subspace Clustering
-* Community Detection in the (G)RDPG
+* Community Detection for the (G)RDPG
 
@@
 
-# Generalizations to Community Detection for the (G)RDPG
+# General Community Detection for the (G)RDPG
 
@@
 
-$\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
+Then $\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
 $\exists N(a, \delta, K) < \infty$ such that when $\min_k n_k \geq N$, 
@@
 1. Single linkage clustering will produce perfect community detection.
-2. An $\epsilon$-neighborhood graph with $\epsilon \in (0, a)$ will consist of 
+2. An $\epsilon$-neighborhood graph with $\epsilon \leq a$ will consist of 
 at least $K$ disjoint subgraphs such that no subgraph contains members of two 
@@
 
-Then $\forall a \in (0, 1)$, $\delta \in (0, 1)$, $K \geq 2$, and $\xi \leq a / 3$, 
+Then $\forall a \in (0, 1)$, $\delta \in (0, 1)$, $K \geq 2$, $\xi \leq a / 3$, 
 $\exists N(a, \delta, K, \xi) < \infty$ such that when $\min_k n_k \geq N$, 
@@
 1. Single linkage will produce perfect community detection.
-2. An $\epsilon$-neighborhood graph will consist of at least $K$ sub-graphs with no subgraph containing vertices of two communities.
+2. An $\epsilon$-neighborhood graph will consist of at least $K$ sub-graphs with no subgraph containing vertices from multiple communities.
 
-This also holds for noisy points sampled on one-dimensional manifolds such that the manifolds are distance at least $a$ apart.
+This also holds for noisy points sampled from one-dimensional manifolds such that the manifolds are distance at least $a$ apart.
 

### 2021-05-18 proposal/slides.Rmd
updated slides

@@
 
-$A \sim BernoulliGraph(P)$ iff:
+::: columns
+
+:::: {.column width=67%}
+
+$A \sim BernoulliGraph(P)$ iff
 
@@
 
-```{r, fig.height = 1, fig.width = 2}
+::::
+
+:::: {.column width=33%}
+
+```{r, fig.height = 2, fig.width = 1.5}
 n <- 2 ** 5
@@
 ```
+
+::::
+
+:::
 
@@
 
-## Heirarchy of Block Models
+## Hierarchy of Block Models
 
@@
 ```{r, out.width = '150px'}
-knitr::include_graphics('noroozi-pensky-heirarchy.png')
+knitr::include_graphics('noroozi-pensky-hierarchy.png')
 ```
 
-\tiny{Majid Noroozi and Marianna Pensky. "The Heirarchy of Block Models". March 2021.}
+\tiny{Majid Noroozi and Marianna Pensky. "The Hierarchy of Block Models". March 2021.}
 \end{center}
@@
 X <- embedding(P)
-pairs(X, col = z, asp = 1, pch = '.', main = 'PABM: Subspaces')
+pairs(X, col = z, asp = 1, pch = '.', main = 'PABM: Orthogonal subspaces')
 ```
@@
 * $q = K (K - 1) / 2$
-* $U$ is orthogonal
-* $X$ is block diagonal
+* $U$ is orthogonal and predetermined for each $K$
+* $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$
+* $\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
+to different communities, $x_i^\top x_j = 0$.
 
-* Non-uniqueness: 
-$A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ for 
-$Q \in \mathbb{O}(p, q)$
+**Remark**: Non-uniqueness of the latent configuration  
+$A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ 
+$\forall Q \in \mathbb{O}(p, q)$
 
@@
 Orthogonalized Spectral Clustering algorithm:
-
-* Input: Adjacency matrix $A$, number of communities $K$
-* Output: Community assignments $1, ..., K$
 
@@
 2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$
-3. Construct affinity matrix $B = |C| + C^\top|$
+3. Construct affinity matrix $B = |C| + |C^\top|$
 
@@
 
-n1 <- 2 ** 5
-n2 <- 2 ** 5
+n1 <- 2 ** 6
+n2 <- 2 ** 6
 n <- n1 + n2
 
-a <- .1
+a <- .6
 
@@
   ggplot() + 
-  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .1) + 
-  geom_segment(aes(x = 0, xend = 1, y = a, yend = a), size = .1) + 
+  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = 0, yend = 0), size = .1) + 
+  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = a, yend = a), size = .1) + 
   geom_point(aes(x = V1, y = V2, colour = label)) + 
@@
   guides(colour = FALSE) + 
-  xlim(-.1, 1.1) + 
+  xlim(-.1, cos(pi / 2 * a) + .1) + 
   ylim(-a, 2 * a) + 
   scale_colour_brewer(palette = 'Set1') + 
-  coord_fixed() + 
+  # coord_fixed() + 
   # ggthemes::theme_few() + 
@@
 
-Then $\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
+$\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
 $\exists N(a, \delta, K) < \infty$ such that when $\min_k n_k \geq N$, 
@@
 1. Single linkage clustering will produce perfect community detection.
-2. An $\epsilon$-neighborhood graph with $\epsilon \leq a$ will consist of 
+2. Any $\epsilon$-neighborhood graph with $\epsilon \leq a$ will consist of 
 at least $K$ disjoint subgraphs such that no subgraph contains members of two 
@@
 ```{r fig.height = 3, out.width = '50%'}
-xi1 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
+# xi1 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
+xi1 <- rep(0, n)
 xi2 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
@@
 1. Single linkage will produce perfect community detection.
-2. An $\epsilon$-neighborhood graph will consist of at least $K$ sub-graphs with no subgraph containing vertices from multiple communities.
+2. Any $\epsilon$-neighborhood graph will consist of at least $K$ sub-graphs with no subgraph containing vertices from multiple communities.
 
 This also holds for noisy points sampled from one-dimensional manifolds such that the manifolds are distance at least $a$ apart.
+
+## Recovery from the Adjacency Matrix
+
+::: columns
+
+:::: {.column width=33%}
+
+```{r, fig.height = 2, fig.width = 3, out.width = '100%'}
+par(mar = rep(1, 4))
+plot(data.matrix, col = z * 2, asp = 1, xlab = NA, ylab = NA,
+     main = 'Latent configuration')
+```
+
+::::
+
+:::: {.column width=33%}
+
+```{r, fig.height = 2, fig.width = 3, out.width = '100%'}
+par(mar = rep(1, 4))
+P <- data.matrix %*% t(data.matrix)
+A <- draw.graph(P)
+qgraph::qgraph(A, groups = factor(z))
+title('Sampled graph')
+```
+
+::::
+
+:::: {.column width=33%}
+
+```{r, fig.height = 2, fig.width = 3, out.width = '100%'}
+par(mar = rep(1, 4))
+A.eigen <- eigen(A)
+Xhat <- A.eigen$vectors[, 1:2] %*% diag(sqrt(A.eigen$values[1:2]))
+plot(Xhat, col = z * 2, asp = 1, main = 'ASE')
+```
+
+::::
+
+:::
 
@@
 4. Relax condition for the minimum distance between manifolds.
-5. Explore additional clustering techniques for these latent configurations.
+5. Explore more robust clustering techniques for these latent configurations.
+6. Extend results to the GRDPG.
 
@@
 
+## Indefinite Orthogonal Group
+
+$$\mathbb{O}(p, q) = \{Q : Q I_{p, q} Q^\top = I_{p, q}\}$$
+
+* $Q^\top Q \neq I$
+* If $A \sim GRDPG_{p, q}(X)$, then $A \sim GRDPG_{p, q}(X Q)$
+* $(Q x)^\top (Q y) = x^\top Q^\top Q y \neq x^\top y$
+* $\|Q\| \neq 1$ $\implies$ $\|Q x - Q y\| \neq \|x - y\|$
+
 ## Expectation Maximization for the PABM
+
+Full data log-likelihood
+
+$$\begin{split}
+\log L = & 
+\sum_{i < j} \sum_{k, l} z_{ik} z_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} +
+(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk})) \\
++ & \sum_i \sum_k z_{ik} \log \pi_k
+\end{split}$$
+
+* E-step
+  * $\gamma_{ik} = P(Z_i = k \mid \{\pi_l\}, \{\lambda_{jl}\})$
+  * $\log \gamma_{ik} \propto 
+  \log \pi_k + 
+  \sum_{j \neq i} \sum_l \pi_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} + 
+  (1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))$
+* M-step
+  * $\pi_k = \frac{1}{n} \sum_i \gamma_{ik}$
+  * $\{\lambda_{ik}\} = \arg\max_{\{\lambda_{ik}\}} E_Z[\log L]$
 
@@
 
+Priors:
+
+* $Z_i \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$
+* $\lambda_{ik} \stackrel{indep}{\sim} Beta(a_{ik}, b_{ik})$
+
+Full joint distribution:
+
+$$\begin{split}
+\log p & = constant \\
+& + \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
+(A_{ij} \log \lambda_{il} \lambda_{jk} + 
+(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk})) \\
+& + \sum_k \sum_i z_{ik} \log \pi_k \\
+& + \sum_i \sum_k (a_{ik} - 1) \log \lambda_{ik} + (b_{ik} - 1) \log (1 - \lambda_{ik})
+\end{split}$$
+
 ## Variational Inference for the PABM
+
+Mean Field Variational Inference
+
+* Minimize $d_{KL}(p || q)$
+  * $p$ is the joint distribution
+  * $q$ is a density of some form
+* Restrict 
+$q(\vec{z}, \{\lambda_{ik}\}) = 
+\Big(\prod_i q_{z_i}(z_i) \Big) \Big(\prod_{i, k} q_{\lambda_{ik}}(\lambda_{ik})\Big)$
+* Iterative solution: 
+$q_{\theta_i}^{(t+1)} \propto \exp(E_{\theta_{-i}^{(t)}}[\log p])$
+
+* Approximate solution for the PABM
+  * $Z_i \mid \{a'_{ik}\}, \{b'_{ik}\} \sim Categorical(\pi'_1, ..., \pi'_K)$
+  * $\lambda_{ik} \mid 
+  \{a'_{-i, -k}\}, \{b'_{-i, -k}\}, \{\pi'_k\} \sim Beta(a'_{ik}, b'_{ik})$
+  * Iteratively update $\{\pi'_K\}, \{a'_{ik}\}, \{b'_{ik}\}$ until convergence

### 2021-05-18 proposal/slides.Rmd
updated slides

@@
 title: Community Detection Methods for the Generalized Random Dot Product Graph Model
-subtitle: Dissertation Proposal Defense
+subtitle: Dissertation Proposal
 author: John Koo
-date: 'TBD'
+date: 'June 2021'
 output: 
@@
 - \usepackage{comment}
+- \setbeamertemplate{itemize items}[circle]
 ---
@@
 
+# Introduction
+
+## Community Detection for Networks
+
+How can we cluster the nodes of a network?
+
+Statistical inference (parametric approach):
+
+1. Define a generative model 
+$G \mid z_1, ..., z_n, \vec{\theta} \sim P(\vec{z}, \vec{\theta})$.
+2. Develop a method for obtaining estimators $f(G) = \hat{z}_1, ..., \hat{z}_n$.
+3. Identify asymptotic properties of estimators and prove consistency.
+
 ## Overview
 
-* Preliminaries
-  * Block Models and Community Detection
+1. Probability Models for Networks
+    * Block Models and Community Structure
     * (Generalized) Random Dot Product Graphs
     * Connecting Block Models to the (G)RDPG
-* The Popularity Adjusted Block Model
+
+\vspace*{.5\baselineskip}
+
+2. Popularity Adjusted Block Model
     * Connecting the PABM to the GRDPG
-  * Subspace Clustering
-* Community Detection for the (G)RDPG
+    * Subspace Clustering for Community Detection
 
-# Preliminaries
+\vspace*{.5\baselineskip}
+
+3. Community Detection for the (G)RDPG
+    * Manifold Clustering
+    * Manfolds as (G)RDPG Latent Configurations
+
+# Probability Models for Networks
 
@@
 
-:::: {.column width=67%}
+:::: {.column width=62.5%}
+
+Let $G = (V, E)$ be an undirected and unweighted graph with $|V| = n$.
+
+$G$ is described by adjacency matrix $A$ such that
+$A_{ij} = \begin{cases} 
+1 & \exists \text{ edge between } i \text{ and } j \\
+0 & \text{else}
+\end{cases}$
 
@@
 
-* $G = (V, E)$
-  * Undirected, unweighted, no self-loops
-  * $|V| = n$
-  * $|E| \leq n (n - 1) / 2$
-* Adjacency matrix $A \in \{0, 1\}^{n \times n}$
-  * $A_{ij} = \begin{cases} 
-  1 & \exists \text{ edge between } i \text{ and } j \\
-  0 & \text{else}
-  \end{cases}$
-* Edge probability matrix $P \in [0, 1]^{n \times n}$
-* $A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ for $i < j$
+* $P \in [0, 1]^{n \times n}$ describes edge probabilities between pairs of 
+vertices.
+* $A_{ij} \stackrel{ind}{\sim} Bernoulli(P_{ij})$ for each $i < j$.
+* $A_{ji} = A_{ij}$ and $A_{ii} = 0$.
+
+Example: If $G$ is an Erdos-Renyi graph, then $P_{ij} = \theta$.
 
@@
 
-:::: {.column width=33%}
+:::: {.column width=37.5%}
 
-```{r, fig.height = 2, fig.width = 1.5}
+```{r, fig.height = 2, fig.width = 2}
 n <- 2 ** 5
@@
 
-* $A \sim BernoulliGraph(P)$
-  * (Hidden) labels $z_1, ..., z_n \in [K]$
-  * $P_{ij} = f(z_i, z_j, \cdot)$
+Suppose each vertex $v_1, ..., v_n$ has hidden labels $z_1, ..., z_n \in [K]$,  
+and each $P_{ij}$ depends on labels $z_i$ and $z_j$.  
+Then $A \sim BernoulliGraph(P)$ is a *block model*.
 
-* Example: Stochastic Block Model with two communities
-    * $K = 2$
+Example: Stochastic Block Model with two communities
+
+::: columns
+
+:::: column
+
+* $z_1, ..., z_n \in \{1, 2\}$
 * $P_{ij} = \begin{cases} 
@@
 
-```{r, fig.height = 1.5, fig.width = 4}
+* To make this an assortative SBM, set $p q > r^2$.
+* In this example, $p = 1/2$, $q = 1/4$, and $r = 1/8$.
+
+::::
+
+:::: column
+
+```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
 n1 <- 2 ** 5
@@
 
+::::
+
+:::
+
 ## Block Models
 
-* Erdos-Renyi Model (1959)
+Erdos-Renyi Model (1959)
+
 * $P_{ij} = \theta$
 * Not a block model
-* Stochastic Block Model (Lorrain and White, 1971)
+
+Stochastic Block Model (Lorrain and White, 1971)
+
 * $P_{ij} = \theta_{z_i z_j}$
 * $K (K + 1) / 2$ parameters $\theta_{kl}$
-* Degree Corrected Block Model (Karrer and Newman, 2011)
+
+Degree Corrected Block Model (Karrer and Newman, 2011)
+
 * $P_{ij} = \theta_{z_i z_j} \omega_i \omega_j$
 * $K (K + 1) / 2 + n$ parameters $\theta_{kl}$, $\omega_i$
-* Popularity Adjusted Block Model (Sengupta and Chen, 2017)
+
+Popularity Adjusted Block Model (Sengupta and Chen, 2017)
+
 * $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
@@
 
-* PABM $\to$ DCBM: $\lambda_{ik} = \sqrt{\theta_{z_i k}} \omega_i$
-* DCBM $\to$ SBM: $\omega_i = 1$
-* SBM $\to$ Erdos-Renyi: $\theta_{kl} = \theta$
+PABM $\to$ DCBM: $\lambda_{ik} = \sqrt{\theta_{z_i k}} \omega_i$
+
+DCBM $\to$ SBM: $\omega_i = 1$
+
+SBM $\to$ Erdos-Renyi: $\theta_{kl} = \theta$
 
 \begin{center}
-```{r, out.width = '150px'}
+```{r, out.width = '125px'}
 knitr::include_graphics('noroozi-pensky-hierarchy.png')
@@
 
-\tiny{Majid Noroozi and Marianna Pensky. "The Hierarchy of Block Models". March 2021.}
+\tiny{Majid Noroozi and Marianna Pensky, 2021}
 \end{center}
-
-## Community Detection in Block Models
-
-Likelihood
-
-$$L = \prod_{i<j} \prod_{k, l}^K 
-\big(p_{k, l, i, j}^{A_{ij}} 
-(1 - p_{k, l, i, j})^{1 - A_{ij}} \big)^{z_{ik} z_{jl}}$$
-
-* Example: DCBM ($p_{k, l, i, j} = \theta_{kl} \omega_i \omega_j$)
-
-$$L = \prod_{i<j} \prod_{k, l}^K 
-\big((\theta_{kl} \omega_i \omega_j)^{A_{ij}} 
-(1 - \theta_{kl} \omega_i \omega_j)^{1 - A_{ij}} \big)^{z_{ik} z_{jl}}$$
-
-* ML method for community detection: $\hat{\vec{z}} = \arg\max_{\vec{z}} L$
-
-* NP-complete
-  * Expectation-Maximization
-  * Bayesian methods
-  * **Spectral methods**
 
@@
 
-* Random Dot Product Graph $A \sim RDPG(X)$
-  * Latent vectors $x_1, ..., x_n \in \mathcal{X} \subset \mathbb{R}^d$
-    * $\mathcal{X} = \{x, y : 0 \leq x^\top y \leq 1\}$
-  * Data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
-  * Edge probability matrix $P = X X^\top$
-  * Adjacency matrix $A \sim BernoulliGraph(P)$
+Random Dot Product Graph $A \sim RDPG(X)$  
+(Young and Scheinerman, 2007)
 
-* Generalized Random Dot Product Graph $A \sim GRDPG_{p, q}(X)$
-  * Latent vectors $x_1, ..., x_n \in \mathcal{X} \subset \mathbb{R}^d$
-    * $\mathcal{X} = \{x, y : 0 \leq x^\top I_{p, q} y \leq 1\}$
-    * $I_{p, q} = blockdiag(I_p, -I_q)$
-    * $p + q = d$
-  * Data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
-  * Edge probability matrix $P = X I_{pq} X^\top$
-  * Adjacency matrix $A \sim BernoulliGraph(P)$
+* Latent vectors $x_1, ..., x_n \in \mathbb{R}^d$ such that 
+$x_i^\top x_j \in [0, 1]$
+* $A \sim BernoulliGraph(X X^\top)$ where 
+$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
 
-* If $X_1, ..., X_n \stackrel{iid}{\sim} F$, then $(A, X) \sim RDPG(F, n)$ or 
-$(A, X) \sim GRDPG_{p, q}(F, n)$
+Generalized Random Dot Product Graph $A \sim GRDPG_{p, q}(X)$  
+(Rubin-Delanchy, Cape, Tang, Priebe, 2020)
+
+* Latent vectors $x_1, ..., x_n \in \mathbb{R}^{p+q}$ such that 
+$x_i^\top I_{p, q} x_j \in [0, 1]$ and $I_{p, q} = blockdiag(I_p, -I_q)$
+* $A \sim BernoulliGraph(X I_{p, q} X^\top)$ where 
+$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
+
+If latent vectors $X_1, ..., X_n \stackrel{iid}{\sim} F$, then we write 
+$(A, X) \sim RDPG(F, n)$ or $(A, X) \sim GRDPG_{p, q}(F, n)$.
 
@@
 
-* Want to estimate $X$ given $A$
-  * Alternatively, recover some property of $X$ given $A$ 
-    * Interpoint distances
-    * Inner products
-    * Angles
+Want to estimate $X$ given $A$, or alternatively, 
+interpoint distances, inner products, or angles.
 
-* Adjacency Spectral Embedding
-  * Given embedding dimension $d$, 
-  $A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$
-    * If RDPG, use $d$ greatest eigenvalues
-    * If GRDPG, use $p$ most positive and $q$ most negative eigenvalues
-    * $\hat{V} \in \mathbb{R}^{n \times d}$
-    * $\hat{\Lambda} \in \mathbb{R}^{d \times d}$
-  * RDPG: $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
-  * GRDPG: $\hat{X} = \hat{V} |\hat{\Lambda}|^{1/2}$
+### Adjacency Spectral Embedding
 
-* RDPG: $\max\limits_i \|\hat{X}_i - W_n X_i \| \stackrel{a.s.}{\to} 0$
-(Athreya et al.)
-* GRDPG: 
+To embed in $\mathbb{R}^d$,
+
+1. Compute $A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$ 
+where $\hat{\Lambda} \in \mathbb{R}^{d \times d}$ and 
+$\hat{V} \in \mathbb{R}^{n \times d}$.  
+For RDPG, use $d$ greatest eigenvalues;  
+for GRDPG, use $p$ most positive and $q$ most negative eigenvalues.
+
+2. For RDPG, let $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$;  
+for GRDPG, let $\hat{X} = \hat{V} |\hat{\Lambda}|^{1/2}$.
+
+\vspace*{.5\baselineskip}
+
+RDPG: $\max\limits_i \|\hat{X}_i - W_n X_i \| \stackrel{a.s.}{\to} 0$
+(Athreya et al., 2018)  
+GRDPG: 
 $\max\limits_i \|\hat{X}_i - Q_n X_i \| \stackrel{a.s.}{\to} 0$
-(Rubin-Delanchy et al.)
+(Rubin-Delanchy et al., 2020)
 
@@
 
-* All $G$ with $A \sim BernoulliGraph(P)$ are RDPG (if $P$ is 
-positive semidefinite) or GRDPG
-  * Includes all block models
+All $G$ with $A \sim BernoulliGraph(P)$ are RDPG (if $P$ is 
+positive semidefinite) or GRDPG (includes all block models).
 
-* Example: SBM
+Example: Assortative SBM
 
@@
 
-```{r, out.width = '20%', fig.height = 2.5, fig.width = 2.5}
+::: columns
+
+:::: column
+
+```{r, out.width = '50%', fig.height = 2.5, fig.width = 2.5}
 par(mar = rep(1.75, 4))
@@
 
-```{r, out.width = '20%', fig.height = 2.5, fig.width = 2.5}
+\vspace*{.5\baselineskip}
+
+```{r, out.width = '50%', fig.height = 2.5, fig.width = 2.5}
 par(mar = rep(1.75, 4))
@@
 
-```{r, out.width = '33%', fig.height = 4, fig.width = 4}
+::::
+
+:::: column
+
+\vspace*{0\baselineskip}
+
+```{r, out.width = '100%', fig.height = 4, fig.width = 4}
 par(mar = rep(1.75, 4))
@@
 
+::::
+
+:::
+
 # Popularity Adjusted Block Model
 
-## Popularity Adjusted Block Model (Reparameterization)
+## Popularity Adjusted Block Model
+
+Definition based on Noroozi, Rimal, and Pensky (2020).
 
@@
 
-* w.l.o.g., organize $P$ such that each block 
+1. w.l.o.g., organize $P$ such that each block 
 $P^{(kl)} \in [0, 1]^{n_k \times n_l}$ contains edge 
-probabilities between communities $k$ and $l$
-* Popularity vectors $\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
-are the popularity parameters of members of community $k$ to community $l$
-* $\{\lambda^{(kl)}\}_K$ is the set of $K^2$ popularity vectors
-* $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$
-* $A \sim BernoulliGraph(P)$
+probabilities between communities $k$ and $l$.
+2. Organize parameters as vectors such that 
+$\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
+are the popularity parameters of members of community $k$ to community $l$.  
+$\{\lambda^{(kl)}\}_K$ is the set of $K^2$ popularity vectors.
+3. Then we can write each block of $P$ as 
+$P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$.
+4. Sample $A \sim BernoulliGraph(P)$.
 
@@
 
-Theorem: $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
+**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
 $A \sim GRDPG_{3, 1}(X U)$ for $X$ constructed from $\{\lambda^{(kl)}\}_2$ and 
@@
 
-**Theorem**: $A \sim PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
+**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
 $A \sim GRDPG_{p, q}(X U)$ such that
@@
 * $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$  
-* $\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
+$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
 to different communities, $x_i^\top x_j = 0$.
@@
 
-**Theorem**: If $P = V \Lambda V^\top$ and $B = n V V^\top$, then $B_{ij} = 0$ 
-$\forall i, j$ in different communities.
+**Theorem** (KTT): 
+If $P = V \Lambda V^\top$ and $B = n V V^\top$, 
+then $B_{ij} = 0$ $\forall i, j$ in different communities.
 
-Orthogonalized Spectral Clustering algorithm:
+Orthogonal Spectral Clustering algorithm:
 
@@
 4. Partition $G$ into $K$ disconnected subgraphs (e.g., using edge thresholding 
-or spectral clustering).
-5. Map each partition to the community labels $1, ..., K$.
+or spectral clustering). Map each partition to the community labels $1, ..., K$.
 
-## Orthogonal Spectral Clustering
-
-**Theorem**: Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity 
-matrix from OSC. 
+**Theorem** (KTT): 
+Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC. 
 Then $\forall$ pairs $(i, j)$ belonging to different communities 
@@
 
-**Theorem**:
+**Theorem** (KTT):
 
@@
 ```{r fig.height = 3, out.width = '50%'}
-set.seed(1)
+set.seed(123456)
 
-n1 <- 2 ** 6
-n2 <- 2 ** 6
+n1 <- 2 ** 7
+n2 <- 2 ** 7
 n <- n1 + n2
 
-a <- .6
+a <- .5
 
@@
 
-```{r, fig.height = 2, fig.width = 3, out.width = '100%'}
+```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
 par(mar = rep(1, 4))
@@
 qgraph::qgraph(A, groups = factor(z))
-title('Sampled graph')
+title('RDPG')
 ```
@@
 
+## Parameter Estimation
+
 ## Indefinite Orthogonal Group
@@
 * $\|Q\| \neq 1$ $\implies$ $\|Q x - Q y\| \neq \|x - y\|$
+
+## Community Detection in Block Models
+
+Likelihood
+
+$$L = \prod_{i<j} \prod_{k, l}^K 
+\big(p_{k, l, i, j}^{A_{ij}} 
+(1 - p_{k, l, i, j})^{1 - A_{ij}} \big)^{z_{ik} z_{jl}}$$
+
+Example: DCBM ($p_{k, l, i, j} = \theta_{kl} \omega_i \omega_j$)
+
+$$L = \prod_{i<j} \prod_{k, l}^K 
+\big((\theta_{kl} \omega_i \omega_j)^{A_{ij}} 
+(1 - \theta_{kl} \omega_i \omega_j)^{1 - A_{ij}} \big)^{z_{ik} z_{jl}}$$
+
+* ML method for community detection: $\hat{\vec{z}} = \arg\max_{\vec{z}} L$
+
+* NP-complete
+  * Expectation-Maximization
+  * Bayesian methods
+  * Spectral methods
 
@@
 
-* E-step
+E-step
+
 * $\gamma_{ik} = P(Z_i = k \mid \{\pi_l\}, \{\lambda_{jl}\})$
@@
 (1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))$
-* M-step
+
+M-step
+
 * $\pi_k = \frac{1}{n} \sum_i \gamma_{ik}$
@@
 * $Z_i \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$
-* $\lambda_{ik} \stackrel{indep}{\sim} Beta(a_{ik}, b_{ik})$
+* $\lambda_{ik} \stackrel{ind}{\sim} Beta(a_{ik}, b_{ik})$
 

### 2021-05-21 proposal/slides.Rmd
updated slides

@@
     * Subspace Clustering for Community Detection
+    * Orthogonal Spectral Clustering
 
@@
 
-$A \sim BernoulliGraph(P)$ iff
+$A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.
 
-* $P \in [0, 1]^{n \times n}$ describes edge probabilities between pairs of 
+$A \sim BernoulliGraph(P)$ iff:
+
+1. $P \in [0, 1]^{n \times n}$ describes edge probabilities between pairs of 
 vertices.
-* $A_{ij} \stackrel{ind}{\sim} Bernoulli(P_{ij})$ for each $i < j$.
-* $A_{ji} = A_{ij}$ and $A_{ii} = 0$.
+2. $A_{ij} \stackrel{ind}{\sim} Bernoulli(P_{ij})$ for each $i < j$.
 
@@
 **Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
-$A \sim GRDPG_{3, 1}(X U)$ for $X$ constructed from $\{\lambda^{(kl)}\}_2$ and 
-$U \in \mathbb{O}(4)$
+$A \sim GRDPG_{3, 1}(X U)$ for block diagonal $X$ 
+constructed from $\{\lambda^{(kl)}\}_2$ and 
+predetermined $U \in \mathbb{O}(4)$.
 
-Proof:
+Proof: Decompose $P$ as follows
 
@@
 
-$$P = X Y^\top$$
+$$P = X Y^\top = \begin{bmatrix}
+\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
+\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
+\end{bmatrix}$$
 
@@
 $\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
-to different communities, $x_i^\top x_j = 0$.
+to different communities, then $x_i^\top x_j = 0$.
 
-**Remark**: Non-uniqueness of the latent configuration  
+**Remark** (non-uniqueness of the latent configuration):  
 $A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ 
@@
 
-## Orthogonal Spectral Clustering
+## Sparse Subspace Clustering
+
+**Corollary**: $X$ is block diagonal by community and 
+$U$ is orthogonal $\implies$ 
+each community corresponds to a subspace in $\mathbb{R}^{K^2}$.
+
+Subspace property holds even with linear transformation 
+$Q \in \mathbb{O}(p, q)$.
+
+Sparse Subspace Clustering algorithm:
+
+1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
+subject to $x_i = X c$ and $c_i^{(i)} = 0$.
+2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
+3. Construct affinity matrix $B = |C| + |C^\top|$.
+
+If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
+$B_{ij} = 0$ if $i$ and $j$ belong to different communities and $\|c_i\| > 0$.
+
+Step (1) of SSC typically performed via LASSO: 
+$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$
+
+## Sparse Subspace Clustering
+
+Noroozi et al. observed that the rank of $P$ is $K^2$ 
+and the columns of $P$ belonging to each community has rank $K$ 
+to justify SSC for the PABM.
+
+$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } A_{\cdot, i} = A c_i
+\text{ and } c_i^{(i)} = 0$$
+
+They were able to show that this obeys SDP if we replace $A$ with $P$.
+
+GRDPG-based approach: Apply SSC to the ASE of $A$.
+
+$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } 
+\hat{x}_i = \hat{X} c_i \text{ and } c_i^{(i)} = 0$$
+$$A \approx \hat{X} I_{p, q} \hat{X}^\top$$
+
+## Sparse Subspace Clustering
+
+Wang and Xu (2016): If data matrix $X$ consists of points lying close to 
+low-dimensional subspaces such that:
+
+1. Each point's distance to its subspace is sufficiently small.
+2. The cosine of the angles between pairs of subspaces is sufficiently small.
+3. The points corresponding to each subspace cover a sufficient amount of 
+that subspace.
+
+Then SDP holds with probability 1.
 
@@
 then $B_{ij} = 0$ $\forall i, j$ in different communities.
-
-Orthogonal Spectral Clustering algorithm:
-
-1. Compute the eigenvectors of $A$ that correspond to the $K (K+1)/2$ most 
-positive and $K (K-1) / 2$ most negative eigenvalues to construct $V$.
-2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
-3. Construct graph $G$ using $B$ as its similarity matrix.
-4. Partition $G$ into $K$ disconnected subgraphs (e.g., using edge thresholding 
-or spectral clustering). Map each partition to the community labels $1, ..., K$.
-
-**Theorem** (KTT): 
-Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC. 
-Then $\forall$ pairs $(i, j)$ belonging to different communities 
-and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
-
-$$
-\max_{i, j} \hat{B}^{(ij)} = 
-O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
-$$
-
-## Sparse Subspace Clustering
-
-* $X$ is block diagonal and $U$ is orthogonal $\implies$ each community 
-corresponds to a subspace in $\mathbb{R}^d$.
-* Subspace property holds even with linear transformation 
-$Q \in \mathbb{O}(p, q)$.
-* If $P = V \Lambda V^\top$, then $V$ consists of *orthogonal* subspaces.
-
-Sparse Subspace Clustering algorithm:
-
-1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
-subject to $x_i = X_{-i} c$ and $c_i^{(i)} = 0$.
-2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$
-3. Construct affinity matrix $B = |C| + |C^\top|$
-
-* If $X$ obeys the Subspace Detection Property, then $B$ is sparse such that 
-$B_{ij} = 0$ if $i$ and $j$ belong to different communities and $\|c_i\| > 0$.
-* Step (1) of SSC typically performed via LASSO: 
-$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$
 
@@
 
-* $P_n$ describe the edge probability matrix of the PABM with 
-$n$ vertices
-* $A_n \sim BernoulliGraph(P_n)$
+* $P_n$ describe the edge probability matrix of the PABM with $n$ vertices, and 
+$A_n \sim BernoulliGraph(P_n)$.
 * $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
@@
 
+## Orthogonal Spectral Clustering
+
+**Theorem** (KTT): 
+If $P = V \Lambda V^\top$ and $B = n V V^\top$, 
+then $B_{ij} = 0$ $\forall i, j$ in different communities.
+
+Orthogonal Spectral Clustering algorithm:
+
+1. Let $V$ be the eigenvectors of $A$ corresponding to the $K (K+1)/2$ most 
+positive and $K (K-1) / 2$ most negative eigenvalues.
+2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
+3. Construct graph $G$ using $B$ as its similarity matrix.
+4. Partition $G$ into $K$ disconnected subgraphs.
+
+**Theorem** (KTT): 
+Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC. 
+Then $\forall$ pairs $(i, j)$ belonging to different communities 
+and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
+
+$$
+\max_{i, j} \hat{B}^{(ij)} = 
+O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
+$$
+
 # General Community Detection for the (G)RDPG
@@
 1. Define functions $f_1, ..., f_K$ such that 
-$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_k(t) \neq f_l(t)$ 
+$f_k : [0, 1]^r \mapsto \mathbb{R}^d$ and $f_k(t) \neq f_l(t)$ 
 $\forall k, l \in [K]$. 
@@
 $Z_1, ..., Z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
-3. Sample $T_1, ..., T_n \stackrel{iid}{\sim} D$ with support $[0, 1]$. 
+3. Sample $T_1, ..., T_n \stackrel{iid}{\sim} D$ with support $[0, 1]^r$. 
 4. Set latent positions $X_i = f_{Z_i}(T_i)$ and 
@@
 
-* Athreya et al. and Rubin-Delanchy et al.: we can approximate properties of 
+Athreya et al. and Rubin-Delanchy et al.: we can approximate properties of 
 the latent configurations via ASE.
-* General community detection method: Given $A$, $K$, and $d$ (or $p$ and $q$), 
+
+General community detection method:  
+Given $A$, $K$, and $d$ (or $p$ and $q$),
+
   1. Use ASE to approximate the latent configuration.
@@
 
-n1 <- 2 ** 7
-n2 <- 2 ** 7
+n1 <- 2 ** 5
+n2 <- 2 ** 5
 n <- n1 + n2
 
-a <- .5
+a <- .1
 
@@
 
+```{r}
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+u1 <- runif(n1)
+u2 <- runif(n2)
+x1 <- cos(pi / 3 * u1)
+y1 <- sin(pi / 3 * u1)
+x2 <- 1 + cos(pi / 3 * u2 + pi)
+y2 <- 1 + sin(pi / 3 * u2 + pi)
+data.matrix <- cbind(c(x1, x2), c(y1, y2))
+```
+
 ::: columns
@@
 A <- draw.graph(P)
-qgraph::qgraph(A, groups = factor(z))
+qgraph::qgraph(A, groups = factor(z), layout = 'spring')
 title('RDPG')

### 2021-05-24 proposal/slides.Rmd
updated slides

@@
 - \usepackage{comment}
-- \setbeamertemplate{itemize items}[circle]
 ---
@@
 ```
-
-# Introduction
 
@@
 
-Example: Stochastic Block Model with two communities
+Example 1: Stochastic Block Model with two communities
 
@@
 
-Want to estimate $X$ given $A$, or alternatively, 
+Want to estimate $X$ from $A$, or alternatively, 
 interpoint distances, inner products, or angles.
@@
 
-Example: Assortative SBM
+Example 1 (cont'd): Assortative SBM ($p q > r^2$) with $K = 2$
 
@@
 
-Example: SBM (cont'd)
+Example 1 (cont'd)
 
-* $A \sim BernoulliGraph(X X^\top)$
-* $A \approx \hat{X} \hat{X}^\top$
-  * $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
-* Apply clustering algorithm (e.g., $K$-means) on $\hat{X}$
+If we want to perform community detection,
 
-```{r, fig.height = 8, fig.width = 8, out.width = '50%'}
+1. Note that $A \sim BernoulliGraph(X X^\top)$ is a RDPG
+2. Compute the ASE $A \approx \hat{X} \hat{X}^\top$  
+$\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
+3. Apply clustering algorithm (e.g., $K$-means) to $\hat{X}$
+
+```{r, fig.height = 4, fig.width = 4, out.width = '50%'}
 A.eigen <- eigen(A)
 X.hat <- A.eigen$vectors[, 1:2] %*% diag(A.eigen$values[1:2] ** .5)
-plot(X.hat, asp = 1, col = z, xlab = NA, ylab = NA)
+plot(X.hat, asp = 1, col = z, xlab = NA, ylab = NA,
+     main = 'ASE of the adjacency matrix drawn from SBM')
 ```
@@
 $A \sim GRDPG_{3, 1}(X U)$ for block diagonal $X$ 
-constructed from $\{\lambda^{(kl)}\}_2$ and 
-predetermined $U \in \mathbb{O}(4)$.
+constructed from $\{\lambda^{(kl)}\}_2$ and $U \in \mathbb{O}(4)$.
 
@@
 **Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
-$A \sim GRDPG_{p, q}(X U)$ such that
+$A \sim GRDPG_{p, q}(X U)$ with
 
@@
 * $q = K (K - 1) / 2$
-* $U$ is orthogonal and predetermined for each $K$
+* $U \in \mathbb{O}(K^2)$
 * $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$ 
+with each block corresponding to a community  
 $\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
@@
 **Remark** (non-uniqueness of the latent configuration):  
-$A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ 
-$\forall Q \in \mathbb{O}(p, q)$
+If $A \sim GRDPG_{p, q}(X)$, then $A \sim GRDPG_{p, q}(X Q)$ for any $Q$ in 
+the indefinite orthogonal group with signature $p, q$.
 

### 2021-05-24 proposal/slides.Rmd
updated slides

@@
 
-Definition based on Noroozi, Rimal, and Pensky (2020).
+**Def** Popularity Adjusted Block Model (Sengupta and Chen, 2017):
 
-$A \sim PABM(\{\lambda^{(kl)}\}_K)$ iff
+Let each vertex $i \in [n]$ have $K$ popularity parameters
+$\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
+Then $A \sim PABM(P)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.
 
-1. w.l.o.g., organize $P$ such that each block 
-$P^{(kl)} \in [0, 1]^{n_k \times n_l}$ contains edge 
-probabilities between communities $k$ and $l$.
-2. Organize parameters as vectors such that 
+**Lemma** (Noroozi, Rimal, and Pensky, 2020):
+
+$A$ is sampled from a PABM if 
+$A \sim BernoulliGraph(P)$ such that:
+
+1. Let $P^{(kl)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
+between communities $k$ and $l$. 
+2. Organize popularity parameters as vectors 
 $\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
-are the popularity parameters of members of community $k$ to community $l$.  
-$\{\lambda^{(kl)}\}_K$ is the set of $K^2$ popularity vectors.
-3. Then we can write each block of $P$ as 
+such that $\lambda^{(kl)}_i$ is the popularity parameter 
+of the $i$^th^ vertex of community $k$ towards community $l$. 
+3. Each block can be decomposed as 
 $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$.
-4. Sample $A \sim BernoulliGraph(P)$.
+
+**Notation**: $A \sim PABM(\{\lambda^{(kl)}\}_K)$.
 
@@
 
-**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
-$A \sim GRDPG_{3, 1}(X U)$ for block diagonal $X$ 
-constructed from $\{\lambda^{(kl)}\}_2$ and $U \in \mathbb{O}(4)$.
+**Theorem** (KTT): 
+Given popularity vectors $\{\lambda^{(kl)}\}_2$, we can define block diagonal $X \in \mathbb{R}^{n \times 4}$ and $U \in \mathbb{O}(4)$ such that $A \sim PABM(\lambda^{(kl)}_2)$ is equivalent to $A \sim GRDPG(X U)$.
 
@@
 
-$$Y = \begin{bmatrix}
-\lambda^{(11)} & 0 & \lambda^{(12)} & 0 \\
-0 & \lambda^{(21)} & 0 & \lambda^{(22)}
-\end{bmatrix}$$
-
-$$P = X Y^\top = \begin{bmatrix}
-\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
-\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
-\end{bmatrix}$$
-
-## Connecting the PABM to the GRDPG ($K = 2$)
-
-Proof (cont'd):
-
-$$Y = X \Pi$$
-
-$$\Pi = \begin{bmatrix} 
-1 & 0 & 0 & 0 \\ 
-0 & 0 & 1 & 0 \\
-0 & 1 & 0 & 0 \\
-0 & 0 & 0 & 1
-\end{bmatrix} = 
-U I_{3, 1} U^\top$$
-
 $$U = \begin{bmatrix}
@@
 
-$$P = (X U) I_{3, 1} (X U)^\top$$
+$$P = (X U) I_{3, 1} (X U)^\top = 
+\begin{bmatrix}
+\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
+\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
+\end{bmatrix}$$
 
@@
 
-* $p = K (K + 1) / 2$
-* $q = K (K - 1) / 2$
+* $p = K (K + 1) / 2$, $q = K (K - 1) / 2$
 * $U \in \mathbb{O}(K^2)$
-* $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$ 
+* $X \in \mathbb{R}^{n \times K^2}$ is block diagonal and 
+composed of $\{\lambda^{(kl)}\}_K$ 
 with each block corresponding to a community  

### 2021-05-24 proposal/slides.Rmd
updated slides

@@
 
-All $G$ with $A \sim BernoulliGraph(P)$ are RDPG (if $P$ is 
-positive semidefinite) or GRDPG (includes all block models).
+All Bernoulli Graphs are RDPG (if $P$ is 
+positive semidefinite) or GRDPG.
 
 Example 1 (cont'd): Assortative SBM ($p q > r^2$) with $K = 2$
+
+::: columns
+
+:::: column
+
+$$P_{ij} = \begin{cases} 
+p & z_i = z_j = 1 \\
+q & z_i = z_j = 2 \\
+r & z_i \neq z_j
+\end{cases}$$
+
+```{r fig.height = 3, fig.width = 6, out.width = '100%'}
+qgraph::qgraph(A, vsize = 4, groups = factor(z))
+```
+
+::::
+
+:::: column
+
+$$P = 
+\begin{bmatrix} 
+P^{(11)} & P^{(12)} \\
+P^{(21)} & P^{(22)}
+\end{bmatrix} =
+X X^\top$$
 
@@
 
-$$P = X X^\top$$
+::::
+
+:::
 
@@
 ```{r, fig.height = 4, fig.width = 4, out.width = '50%'}
+P.eigen = eigen(P)
+X <- P.eigen$vectors[, 1:2] %*% diag(P.eigen$values[1:2] ** .5)
+
 A.eigen <- eigen(A)
 X.hat <- A.eigen$vectors[, 1:2] %*% diag(A.eigen$values[1:2] ** .5)
-plot(X.hat, asp = 1, col = z, xlab = NA, ylab = NA,
+plot(X.hat, asp = 1, col = z * 2, xlab = NA, ylab = NA, 
      main = 'ASE of the adjacency matrix drawn from SBM')
+points(X, pch = 16, col = z * 2)
 ```
@@
 par(mar = rep(1.75, 4))
-P.eigen = eigen(P)
-X <- P.eigen$vectors[, 1:2] %*% diag(P.eigen$values[1:2] ** .5)
 plot(X, asp = 1, col = z, xlab = NA, ylab = NA, main = 'SBM: Point masses')
@@
 par(mar = rep(1.75, 4))
-Pz <- generate.P.beta(n)
+Pz <- generate.P.beta(n * 10)
 P <- Pz$P
@@
 
-Wang and Xu (2016): If data matrix $X$ consists of points lying close to 
+**Theorem** (Wang and Xu, 2016): 
+If data matrix $X$ consists of points lying close to 
 low-dimensional subspaces such that:
@@
 
-Then SDP holds with probability 1.
+Then $X$ obeys SDP.
 

### 2021-05-24 proposal/slides.Rmd
updated slides

@@
 
-```{r, fig.height = 4, fig.width = 4, out.width = '50%'}
+```{r, fig.height = 4, fig.width = 5, out.width = '50%'}
 P.eigen = eigen(P)
@@
 
-$A$ is sampled from a PABM if 
-$A \sim BernoulliGraph(P)$ such that:
+$A$ is sampled from a PABM if $P$ can be described as:
 
-1. Let $P^{(kl)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
+1. Let each $P^{(kl)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
 between communities $k$ and $l$. 
@@
 $\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
-such that $\lambda^{(kl)}_i$ is the popularity parameter 
+such that $\lambda^{(kl)}_i = \lambda_{k_i l}$ is the popularity parameter 
 of the $i$^th^ vertex of community $k$ towards community $l$. 
@@
 
-**Corollary**: $X$ is block diagonal by community and 
-$U$ is orthogonal $\implies$ 
+**Corollary**: Since $X$ is block diagonal by community and $U$ is orthogonal,
 each community corresponds to a subspace in $\mathbb{R}^{K^2}$.
@@
 
-If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
-$B_{ij} = 0$ if $i$ and $j$ belong to different communities and $\|c_i\| > 0$.
-
 Step (1) of SSC typically performed via LASSO: 
 $c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$
+
+## Sparse Subspace Clustering
+
+If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
+$B_{ij} = 0$ if $x_i$ and $x_j$ belong to different subspaces
+
+::: columns
+
+:::: column
+
+```{r, fig.height = 5, fig.width = 5, out.width = '75%'}
+set.seed(314159)
+
+x1 <- runif(n, 1 / sqrt(50), 1)
+y1 <- rep(0, n)
+x2 <- rep(0, n)
+y2 <- x1
+z <- c(rep(1, n), rep(2, n))
+X <- cbind(c(x1, x2), c(y1, y2))
+X <- X + cbind(runif(n * 2, 0, .025),
+               runif(n * 2, 0, .025))
+X <- X %*% rbind(c(cos(pi / 9), sin(pi / 9)),
+                 c(-sin(pi/ 9), cos(pi / 9)))
+plot(X, col = z * 2, asp = 1, xlab = NA, ylab = NA)
+```
+
+::::
+
+:::: column
+
+```{r, fig.height = 5, fig.width = 5, out.width = '75%'}
+Y <- t(X)
+Y <- sweep(Y, 2, apply(Y, 2, function(y) sqrt(sum(y ** 2))), `/`)
+Y[is.nan(Y)] <- 0
+lambda <- .05
+N <- ncol(Y)
+B <- plyr::aaply(seq(N), 1, function(i) {
+  y <- Y[, i]
+  X <- Y[, -i]
+  betahat <- glmnet::glmnet(X, y, lambda = lambda, intercept = FALSE) %>% 
+    coef() %>% 
+    as.numeric()
+  if (i != N) {
+    betahat <- c(betahat[seq(i)], 0, betahat[seq(i + 1, N)])
+  } else {
+    betahat <- c(betahat, 0)
+  }
+  betahat <- betahat[-1]
+  return(betahat)
+}, .parallel = TRUE) %>% 
+  abs()
+B <- sweep(B, 2, apply(B, 2, max), `/`)
+B[is.nan(B)] <- 0
+W <- B + t(B)
+
+qgraph::qgraph(sign(W), layout = 'spring', groups = factor(z))
+```
+
+::::
+
+:::
 
@@
 
-$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } A_{\cdot, i} = A c_i
-\text{ and } c_i^{(i)} = 0$$
+$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } A_{\cdot, i} = A c
+\text{ and } c^{(i)} = 0$$
 
@@
 
-$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } 
-\hat{x}_i = \hat{X} c_i \text{ and } c_i^{(i)} = 0$$
+$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } 
+\hat{x}_i = \hat{X} c \text{ and } c^{(i)} = 0$$
 $$A \approx \hat{X} I_{p, q} \hat{X}^\top$$
@@
 1. Each point's distance to its subspace is sufficiently small.
-2. The cosine of the angles between pairs of subspaces is sufficiently small.
-3. The points corresponding to each subspace cover a sufficient amount of 
+2. The points corresponding to each subspace cover a sufficient amount of 
 that subspace.
+3. The cosine of the angles between pairs of subspaces is sufficiently small.
 

### 2021-05-25 proposal/proposal.Rmd
updated proposal

@@
 \newcommand{\blockdiag}{\text{blockdiag}}
-\newcommand{\indep}{\stackrel{\text{indep}}{\sim}}
+\newcommand{\indep}{\stackrel{\text{ind}}{\sim}}
 \newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
@@
 \newcommand{\Betadist}{\text{Beta}}
+\newcommand{\BG}{\text{BernoulliGraph}}
+\newcommand{\Cat}{\text{Categorical}}
 \newtheorem{definition}{Definition}
@@
 This has resulted in various challenges for researchers who rely on traditional statistical and machine learning methods, many of which are incompatible with graph data and instead require the data to exist as feature vectors in Euclidean space.
-Such challenges often involve clustering and community detection, which are often the goal of graph data analysis.
+This includes graph clustering and community detection, which are often the goal of data analysis on graphs and networks.
 Common clustering methods typically involve calculating some central or representative point for each cluster around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}).
@@
 One family of methods for unifying graph community detection with traditional clustering techniques is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the graph into Euclidean space, followed by applying a popular clustering algorithm such as $K$-means clustering.
-The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models such that latent positions in Euclidean space are used to induce graphs.
+The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models for graphs involving latent vectors in Euclidean space.
+In such a model, each vertex has a corresponding vector in latent space, and the edge probabilities between pairs of vertices is computed by operations on corresponding pairs of latent vectors.
 A community detection algorithm motivated by this may involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.
@@
 First, we explore existing generative graph models with underlying community structures that can be inferred by connecting these models to the RDPG or GRDPG.
+In particular, we use the connection between the Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1} and the GRDPG to motivate two community detection methods.
 Then we explore other latent structures or mixture distributions in the latent space that induce graphs for which consistent community detection is possible.
@@
 Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $1 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$).
-Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$ and $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.
+We denote $A \sim \BG(P)$ as graph with adjacency matrix $A$ sampled from edge probability matrix $P$ in this manner.
+If each vertex has a hidden label in $[K]$, they are denoted as $z_1, ..., z_n$.
+Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$.
 
@@
 Generative models for independent Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$.
-In order to motivate community detection methods, we restrict the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on their labels.
+In order to motivate community detection methods, we restrict the generative model such that for each pair of vertices $(i, j)$, the probability of an edge between the vertices $P_{ij}$ depends in some way on their labels $z_i$ and $z_j$.
+
 One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = p_{z_i z_j}$ where $p_{kl} = p_{lk}$ is the probability of an edge between each vertex in community $k$ to each vertex in community $l$.
 The homogeneous SBM further restricts $P$ to two unique entries: $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise.
-Several generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block Model (PABM).
+Several generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) \cite{Karrer_2011} and the Popularity Adjusted Block Model (PABM) \cite{307cbeb9b1be48299388437423d94bf1}.
 Like the SBM, these models involve imposing some restriction on $P$ based on community labels (and possibly other factors).
@@
 Like the SBM, DCBM, and PABM, the RDPG and GRDPG are generative models for graphs involving an edge probability matrix $P$.
-The RDPG starts with points in latent space $X \in \mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y \in [0, 1]$.
-$P$ is then constructed as $P = X X^\top$, and graph $G$ with adjacency matrix $A$ is drawn from $P$.
-We provide a more formal definition of the RDPG and GRDPG below.
+The RDPG starts with points in latent space $X \in \mathcal{X} \subset \mathbb{R}^d$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}$.
+$P$ is then constructed as $P = X X^\top$, and $A \sim \BG(P)$.
+We provide a more formal definition of the RDPG and GRDPG as follows:
 
 \begin{definition}[(Generalized) Random Dot Product Graph]
-Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y.\in [0, 1]$.
-$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix $A$ is drawn such that $A_{ij} \sim \text{Bernoulli}(x_i^\top x_j)$ for $i < j$, with $A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.
-If on the other hand $A_{ij} \sim \text{Bernoulli}(x_i^\top I_{p, q} x_j)$ where $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
+Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ where $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}$.
+$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix is drawn as $A \sim \BG(X X^\top)$.
+
+If on the other hand, $\mathcal{X} = \{x, y \in \mathbb{R}^{p+q} : x^\top I_{p, q} y \in [0, 1]\}$ and $A \sim \BG(X I_{p, q} X^\top)$ for $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
 These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
 
-In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $x_1, ..., x_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $X_1, ..., X_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
 If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
@@
 Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$.
-Let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
+Let $A \approx V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
 Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.
@@
 
-It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all graphs generated by $P$ are special cases of the GRDPG.
+It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all graphs generated by $P$ regardless of positive semidefiniteness are special cases of the GRDPG.
 This includes the SBM, DCBM, and PABM.
@@
 \begin{example}[Connecting the SBM to the RDPG]
-Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respective, and the between-community edge probability is $r$, with $p q > r^2$.
-Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$.
+Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respectively, and the between-community edge probability is $r$, with $p q > r^2$.
+Let community 1 have $n_1$ vertices, community 2 have $n_2$ vertices, and $n_1 + n_2 = n$.
 Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$.
 Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$.
-One RDPG representation of this SBM is:
+Then one RDPG representation of this SBM is:
 
@@
 
-where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$. Then it can be shown that
-
-$$P = X X^\top$$
+where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$.
+This is shown by reconstructing $P$ as $P = X X^\top$.
 \end{example}
@@
 
-As discussed in \S 2.1, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space.
+As discussed in \S 2.2, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space.
 A similar result has been shown for the DCBM \cite{lyzinski2014}.
 We can also extend this to the PABM.
-In \S 3.3.1, we show that the PABM is equivalent to the GRDPG model for which the communities correspond to subspaces in the latent configuration.
+In \S 3.1, we show that the PABM is equivalent to the GRDPG model for which the communities correspond to subspaces in the latent configuration.
 We further show that while the latent configuration is not unique, there exists one in particular for which the subspaces are orthogonal.
@@
 
-We first define the PABM.
+We first define the PABM using the construction provided by \citet{noroozi2019estimation}.
 
@@
 
-Similar to the $K = 2$ case, we have $Y = X \Pi$ for a permutation matrix
+Note that $Y = X \Pi$ for a permutation matrix
 $\Pi$, resulting in $P = X \Pi X^\top$.  
@@
 
-* Positions 1, 5, 9 are fixed.
+\item Positions 1, 5, 9 are fixed.
 
-* The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.
+\item The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.
 
@@
 
-While direct maximization of the likelihood or log-likelihood is NP-complete, it is possible to apply an Expectation Maximization (EM) algorithm to this problem. To set this up,
+### Expectation Maximization
 
-# Generalizations
+While direct maximization of the likelihood or log-likelihood is NP-complete, it is possible to apply an Expectation Maximization (EM) algorithm to this problem.
+To set this up, we write the complete data log likelihood as:
+
+$$Q = 
+\sum_{i < j} \sum_{k, l} z_{ik} z_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} +
+(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))
++ \sum_i \sum_k z_{ik} \log \pi_k$$
+
+Where $\pi_k$ are the *a priori* community label probabilities.
+
+Then the expectation step involves solving for $\gamma_{ik} = P(Z_i = k \mid \{\pi_l\}, \{\lambda_{jl}\})$, which is given by
+
+$$\log \gamma_{ik} \propto 
+\log \pi_k + 
+\sum_{j \neq i} \sum_l \pi_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} + 
+(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))$$
+
+For the maximization step, we can just set $\pi_k = \frac{1}{n} \sum_i \gamma_{ik}$, but the $\{\lambda_{ik}\}$ parameters cannot be maximized in closed form and require an optimization subroutine.
+Our proposed work involves finding an efficient optimization method for this subroutine.
+
+### Bayesian Methods
+
+The introduction of community label probabilities $\pi_1, ..., \pi_K$ suggests a hierarchical model in which the community labels are generated by a prior distribution.
+We can apply the same principle to the popularity parameters $\{\lambda_{ik}\}$ to make the model fully Bayesian.
+In this setup, we sample $z_i \iid \Cat(\pi_1, ..., \pi_K)$ and $\lambda_{ik} \indep \Betadist(a_{ik}, b_{ik})$ before sampling the observed data $A_{ij} \indep \lambda_{i z_j} \lambda_{j z_i}$.
+
+The full joint distribution for this model can be written as:
+
+$$\begin{split}
+\log p & = constant \\
+& + \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
+(A_{ij} \log \lambda_{il} \lambda_{jk} + 
+(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk})) \\
+& + \sum_k \sum_i z_{ik} \log \pi_k \\
+& + \sum_i \sum_k (a_{ik} - 1) \log \lambda_{ik} + (b_{ik} - 1) \log (1 - \lambda_{ik})
+\end{split}$$
+
+It is straightforward to obtain the conditional distribution for each $z_i$ but not each $\lambda_{ik}$, so a Gibbs sampler cannot be derived for this model.
+However, we can use a Metropolis-Hastings algorithm.
+In particular, by taking a first-order Taylor approximation, we can approximate the conditional distribution for each $\lambda_{ik}$ as a Beta distribution.
+Our proposed work involves investigating this method by simulating this model and implementing the Bayes sampler.
+
+# Generalizations to GRDPG-Based Community Detection
 
@@
 
-If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, 1]$, then a sample of size $\min(n_1, n_2) \geq N(\delta, a)$ is sufficient to satisfy:
+If we extend this example such that $n_1$ points are sampled uniformly from the segment $f_1(t) = (t, 0)$ and $n_2$ points are sampled uniformly from the segment $f_2(t) = (t, a)$ for $t \in [0, \cos \frac{\pi}{2} a]$, then a sample of size $\min(n_1, n_2) \geq N(\delta, a)$ is sufficient to satisfy:
 
@@
 We can then further extend this to the case where points are drawn from unit segments with noise.
+It is straightforward to show that if the maximum noise is less than $a / 3$, we can obtain a similar result as in the noiseless case.
+Thus if the distribution on the two segments is used to draw $(A, X) \sim \text{RDPG}(F, n)$, the same properties should hold for large $n$.
 
@@
 
+Our proposed work extends these results by the following.
+
+1. Show that the ASE of a (G)RDPG generated by this type of latent configuration produces the correct conditions as described above to apply the same probability statements.
+2. Extend these results by deriving similar probability statements for nonuniform distributions. 
+3. Relax the condition for the minimum distance between manifolds by investigating alternative clustering techniques on the latent space (e.g., spectral clustering).
+
 # Summary
+
+For this research, we propose using the connection between Block Models and the (Generalized) Random Dot Product Graph to motivate community detection methods.
+Previous research shows that the Stochastic Block Model and Degree Corrected Block Model have very structured latent configurations in the RDPG framework, and these latent structures can be used to develop community detection algorithms that result in perfect community detection for large $n$.
+In the same vein, we investigated the connection between the Popularity Adjusted Block Model and the GRDPG to motivate two community detection algorithms, the first of which is an application of an existing algorithm, Sparse Subspace Clustering, and the second is a new algorithm based on the particular latent form of the PABM, Orthogonal Spectral Clustering.
+Furthermore, we showed that for large $n$, our application of the SSC algorithm to the PABM upholds the Subspace Detection Property for large yet finite $n$, and the output of OSC results in perfect community detection almost surely as $n \to \infty$.
+We will extend these theoretical results with empirical simulations and real data examples.
+Finally, we will implement these two community detection algorithms in an R package.
+
+We aim to extend these results by defining more general structures in the latent space to construct RDPG or GRDPG models and show that community detection is possible via Adjacency Spectral Embedding.
+So far, we showed that if in the latent space the communities can be represented as points sampled uniformly on curves or one-dimensional manifolds separated by a nonzero distance, we can apply two existing clustering techniques to the ASE and show that they will perform perfect community detection with high probability.
+We will generalize these results to nonuniform distributions on higher dimensional manifolds as well as investigate alternative clustering techniques for the latent space that may perform better for smaller $n$.
 
@@
 
-\newpage
-
 # References

### 2021-05-25 proposal/proposal.tex
updated proposal

@@
 which are incompatible with graph data and instead require the data to
-exist as feature vectors in Euclidean space. Such challenges often
-involve clustering and community detection, which are often the goal of
-graph data analysis. Common clustering methods typically involve
-calculating some central or representative point for each cluster around
-which the data belonging to that cluster lie (e.g., Lloyd's algorithm
-for \(K\)-means clustering \cite{1056489}, Gaussian Mixture Models
-\cite{doi:10.1198/016214502760047131}). Because these methods involve
-computing summary statistics within each cluster, such as the sample
-average, they cannot be applied directly to graphs, necessitating
-methods for transforming the graph data into feature vectors.
+exist as feature vectors in Euclidean space. This includes graph
+clustering and community detection, which are often the goal of data
+analysis on graphs and networks. Common clustering methods typically
+involve calculating some central or representative point for each
+cluster around which the data belonging to that cluster lie (e.g.,
+Lloyd's algorithm for \(K\)-means clustering \cite{1056489}, Gaussian
+Mixture Models \cite{doi:10.1198/016214502760047131}). Because these
+methods involve computing summary statistics within each cluster, such
+as the sample average, they cannot be applied directly to graphs,
+necessitating methods for transforming the graph data into feature
+vectors.
 
@@
 Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
-this further by explicitly constructing generative models such that
-latent positions in Euclidean space are used to induce graphs. A
-community detection algorithm motivated by this may involve learning the
-latent positions given an observed graph and then learning the community
-labels given the latent positions.
+this further by explicitly constructing generative models for graphs
+involving latent vectors in Euclidean space. In such a model, each
+vertex has a corresponding vector in latent space, and the edge
+probabilities between pairs of vertices is computed by operations on
+corresponding pairs of latent vectors. A community detection algorithm
+motivated by this may involve learning the latent positions given an
+observed graph and then learning the community labels given the latent
+positions.
 
@@
 existing generative graph models with underlying community structures
-that can be inferred by connecting these models to the RDPG or GRDPG.
-Then we explore other latent structures or mixture distributions in the
-latent space that induce graphs for which consistent community detection
-is possible.
+that can be inferred by connecting these models to the RDPG or GRDPG. In
+particular, we use the connection between the Popularity Adjusted Block
+Model \cite{307cbeb9b1be48299388437423d94bf1} and the GRDPG to motivate
+two community detection methods. Then we explore other latent structures
+or mixture distributions in the latent space that induce graphs for
+which consistent community detection is possible.
 
@@
 probabilities. Graph \(G\) is sampled from \(P\) by drawing
-\(A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})\) for
-each \(1 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)).
-Finally, we denote
+\(A_{ij} \stackrel{\text{ind}}{\sim}\text{Bernoulli}(P_{ij})\) for each
+\(1 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)). We
+denote \(A \sim \text{BernoulliGraph}(P)\) as graph with adjacency
+matrix \(A\) sampled from edge probability matrix \(P\) in this manner.
+If each vertex has a hidden label in \([K]\), they are denoted as
+\(z_1, ..., z_n\). Finally, we denote
 \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
-as the sample \(x_1, ..., x_n \in \mathbb{R}^d\) and
-\(z_1, ..., z_n \in [K]\) as their corresponding (hidden) labels.
+as the sample \(x_1, ..., x_n \in \mathbb{R}^d\).
 
@@
 \(i, j \in [n]\). In order to motivate community detection methods, we
-restrict the generative model such that for each pair of vertices, the
-probability of an edge between the vertices is conditioned on their
-labels. One such model is the Stochastic Block Model (SBM)
+restrict the generative model such that for each pair of vertices
+\((i, j)\), the probability of an edge between the vertices \(P_{ij}\)
+depends in some way on their labels \(z_i\) and \(z_j\).
+
+One such model is the Stochastic Block Model (SBM)
 \cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
@@
 generalizations of the SBM have been introduced since, including the
-Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block
-Model (PABM). Like the SBM, these models involve imposing some
-restriction on \(P\) based on community labels (and possibly other
-factors).
+Degree Corrected Block Model (DCBM) \cite{Karrer_2011} and the
+Popularity Adjusted Block Model (PABM)
+\cite{307cbeb9b1be48299388437423d94bf1}. Like the SBM, these models
+involve imposing some restriction on \(P\) based on community labels
+(and possibly other factors).
 
@@
 with points in latent space \(X \in \mathcal{X} \subset \mathbb{R}^d\)
-such that \(\forall x, y \in \mathcal{X}\), \(x^\top y \in [0, 1]\).
-\(P\) is then constructed as \(P = X X^\top\), and graph \(G\) with
-adjacency matrix \(A\) is drawn from \(P\). We provide a more formal
-definition of the RDPG and GRDPG below.
+for \(\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}\).
+\(P\) is then constructed as \(P = X X^\top\), and
+\(A \sim \text{BernoulliGraph}(P)\). We provide a more formal definition
+of the RDPG and GRDPG as follows:
 
 \begin{definition}[(Generalized) Random Dot Product Graph]
-Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y.\in [0, 1]$.
-$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix $A$ is drawn such that $A_{ij} \sim \text{Bernoulli}(x_i^\top x_j)$ for $i < j$, with $A_{ji} = A_{ij}$ and $A_{ii} = 0$ $\forall i, j \in [n]$.
-If on the other hand $A_{ij} \sim \text{Bernoulli}(x_i^\top I_{p, q} x_j)$ where $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
+Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ where $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}$.
+$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix is drawn as $A \sim \text{BernoulliGraph}(X X^\top)$.
+
+If on the other hand, $\mathcal{X} = \{x, y \in \mathbb{R}^{p+q} : x^\top I_{p, q} y \in [0, 1]\}$ and $A \sim \text{BernoulliGraph}(X I_{p, q} X^\top)$ for $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
 These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
 
-In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $x_1, ..., x_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $X_1, ..., X_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
 If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
@@
 semidefinite \(P\) are special cases of the RDPG, which is a special
-case of the GRDPG, and all graphs generated by \(P\) are special cases
-of the GRDPG. This includes the SBM, DCBM, and PABM. In the following
-example, we show that under the RDPG framework, the latent configuration
-that induces the SBM takes on a very particular form.
+case of the GRDPG, and all graphs generated by \(P\) regardless of
+positive semidefiniteness are special cases of the GRDPG. This includes
+the SBM, DCBM, and PABM. In the following example, we show that under
+the RDPG framework, the latent configuration that induces the SBM takes
+on a very particular form.
 
 \begin{example}[Connecting the SBM to the RDPG]
-Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respective, and the between-community edge probability is $r$, with $p q > r^2$.
-Let community 1 have $n_1$ vertices and community 2 have $n_2$ vertices such that $n_1 + n_2 = n$.
+Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respectively, and the between-community edge probability is $r$, with $p q > r^2$.
+Let community 1 have $n_1$ vertices, community 2 have $n_2$ vertices, and $n_1 + n_2 = n$.
 Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$.
 Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$.
-One RDPG representation of this SBM is:
+Then one RDPG representation of this SBM is:
 
@@
 
-As discussed in \S 2.1, all Bernoulli graph models can be expressed as a
+As discussed in \S 2.2, all Bernoulli graph models can be expressed as a
 RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
@@
 been shown for the DCBM \cite{lyzinski2014}. We can also extend this to
-the PABM. In \S 3.3.1, we show that the PABM is equivalent to the GRDPG
+the PABM. In \S 3.1, we show that the PABM is equivalent to the GRDPG
 model for which the communities correspond to subspaces in the latent
@@
 
-We first define the PABM.
+We first define the PABM using the construction provided by
+\citet{noroozi2019estimation}.
 
@@
 realization of $\text{Bernoulli}(P)$, i.e., 
-$A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ 
+$A_{ij} \stackrel{\text{ind}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ 
 ($A_{ij} = A_{ji}$ and $A_{ii} = 0$).
@@
 that the adjacency matrix is drawn such that
-\(A_{ij} \stackrel{\text{indep}}{\sim}Bernoulli(P_{ij})\), is a special
+\(A_{ij} \stackrel{\text{ind}}{\sim}Bernoulli(P_{ij})\), is a special
 case of the GRDPG. It can also be shown that the latent positions of the
@@
 
+\hypertarget{expectation-maximization}{%
+\subsubsection{Expectation
+Maximization}\label{expectation-maximization}}
+
 While direct maximization of the likelihood or log-likelihood is
 NP-complete, it is possible to apply an Expectation Maximization (EM)
-algorithm to this problem. To set this up,
+algorithm to this problem. To set this up, we write the complete data
+log likelihood as:
 
-\hypertarget{generalizations}{%
-\section{Generalizations}\label{generalizations}}
+\[Q = 
+\sum_{i < j} \sum_{k, l} z_{ik} z_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} +
+(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))
++ \sum_i \sum_k z_{ik} \log \pi_k\]
+
+Where \(\pi_k\) are the \emph{a priori} community label probabilities.
+
+Then the expectation step involves solving for
+\(\gamma_{ik} = P(Z_i = k \mid \{\pi_l\}, \{\lambda_{jl}\})\), which is
+given by
+
+\[\log \gamma_{ik} \propto 
+\log \pi_k + 
+\sum_{j \neq i} \sum_l \pi_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} + 
+(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))\]
+
+For the maximization step, we can just set
+\(\pi_k = \frac{1}{n} \sum_i \gamma_{ik}\), but the \(\{\lambda_{ik}\}\)
+parameters cannot be maximized in closed form and require an
+optimization subroutine. Our proposed work involves finding an efficient
+optimization method for this subroutine.
+
+\hypertarget{bayesian-methods}{%
+\subsubsection{Bayesian Methods}\label{bayesian-methods}}
+
+The introduction of community label probabilities \(\pi_1, ..., \pi_K\)
+suggests a hierarchical model in which the community labels are
+generated by a prior distribution. We can apply the same principle to
+the popularity parameters \(\{\lambda_{ik}\}\) to make the model fully
+Bayesian. In this setup, we sample
+\(z_i \stackrel{\text{iid}}{\sim}\text{Categorical}(\pi_1, ..., \pi_K)\)
+and
+\(\lambda_{ik} \stackrel{\text{ind}}{\sim}\text{Beta}(a_{ik}, b_{ik})\)
+before sampling the observed data
+\(A_{ij} \stackrel{\text{ind}}{\sim}\lambda_{i z_j} \lambda_{j z_i}\).
+
+The full joint distribution for this model can be written as:
+
+\[\begin{split}
+\log p & = constant \\
+& + \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
+(A_{ij} \log \lambda_{il} \lambda_{jk} + 
+(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk})) \\
+& + \sum_k \sum_i z_{ik} \log \pi_k \\
+& + \sum_i \sum_k (a_{ik} - 1) \log \lambda_{ik} + (b_{ik} - 1) \log (1 - \lambda_{ik})
+\end{split}\]
+
+It is straightforward to obtain the conditional distribution for each
+\(z_i\) but not each \(\lambda_{ik}\), so a Gibbs sampler cannot be
+derived for this model. However, we can use a Metropolis-Hastings
+algorithm. In particular, by taking a first-order Taylor approximation,
+we can approximate the conditional distribution for each
+\(\lambda_{ik}\) as a Beta distribution. Our proposed work involves
+investigating this method by simulating this model and implementing the
+Bayes sampler.
+
+\hypertarget{generalizations-to-grdpg-based-community-detection}{%
+\section{Generalizations to GRDPG-Based Community
+Detection}\label{generalizations-to-grdpg-based-community-detection}}
 
@@
 from the segment \(f_1(t) = (t, 0)\) and \(n_2\) points are sampled
-uniformly from the segment \(f_2(t) = (t, a)\) for \(t \in [0, 1]\),
-then a sample of size \(\min(n_1, n_2) \geq N(\delta, a)\) is sufficient
-to satisfy:
+uniformly from the segment \(f_2(t) = (t, a)\) for
+\(t \in [0, \cos \frac{\pi}{2} a]\), then a sample of size
+\(\min(n_1, n_2) \geq N(\delta, a)\) is sufficient to satisfy:
 
@@
 We can then further extend this to the case where points are drawn from
-unit segments with noise.
+unit segments with noise. It is straightforward to show that if the
+maximum noise is less than \(a / 3\), we can obtain a similar result as
+in the noiseless case. Thus if the distribution on the two segments is
+used to draw \((A, X) \sim \text{RDPG}(F, n)\), the same properties
+should hold for large \(n\).
 
@@
 
+Our proposed work extends these results by the following.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Show that the ASE of a (G)RDPG generated by this type of latent
+  configuration produces the correct conditions as described above to
+  apply the same probability statements.
+\item
+  Extend these results by deriving similar probability statements for
+  nonuniform distributions.
+\item
+  Relax the condition for the minimum distance between manifolds by
+  investigating alternative clustering techniques on the latent space
+  (e.g., spectral clustering).
+\end{enumerate}
+
 \hypertarget{summary}{%
 \section{Summary}\label{summary}}
+
+For this research, we propose using the connection between Block Models
+and the (Generalized) Random Dot Product Graph to motivate community
+detection methods. Previous research shows that the Stochastic Block
+Model and Degree Corrected Block Model have very structured latent
+configurations in the RDPG framework, and these latent structures can be
+used to develop community detection algorithms that result in perfect
+community detection for large \(n\). In the same vein, we investigated
+the connection between the Popularity Adjusted Block Model and the GRDPG
+to motivate two community detection algorithms, the first of which is an
+application of an existing algorithm, Sparse Subspace Clustering, and
+the second is a new algorithm based on the particular latent form of the
+PABM, Orthogonal Spectral Clustering. Furthermore, we showed that for
+large \(n\), our application of the SSC algorithm to the PABM upholds
+the Subspace Detection Property for large yet finite \(n\), and the
+output of OSC results in perfect community detection almost surely as
+\(n \to \infty\). We will extend these theoretical results with
+empirical simulations and real data examples. Finally, we will implement
+these two community detection algorithms in an R package.
+
+We aim to extend these results by defining more general structures in
+the latent space to construct RDPG or GRDPG models and show that
+community detection is possible via Adjacency Spectral Embedding. So
+far, we showed that if in the latent space the communities can be
+represented as points sampled uniformly on curves or one-dimensional
+manifolds separated by a nonzero distance, we can apply two existing
+clustering techniques to the ASE and show that they will perform perfect
+community detection with high probability. We will generalize these
+results to nonuniform distributions on higher dimensional manifolds as
+well as investigate alternative clustering techniques for the latent
+space that may perform better for smaller \(n\).
 
@@
 
-\newpage
-
 \renewcommand\refname{References}

### 2021-05-26 proposal/proposal.tex
updated slides

@@
 \vspace*{3\baselineskip}
-Date TBA \\
+June 2, 2021 \\
 \vspace*{1\baselineskip}
@@
 \newpage
+
+\begin{center}
+\LARGE{Abstract}
+\end{center}
+
+\vspace*{2\baselineskip}
+
+\normalsize
+
+Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision.
+One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels).
+Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs.
+In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models, first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then by proposing clustering methods for these feature vector representations and deriving the asymptotic properties of these methods.
 
@@
 
-\hypertarget{alternative-approaches-to-community-detection}{%
-\subsection{Alternative Approaches to Community
-Detection}\label{alternative-approaches-to-community-detection}}
+\hypertarget{non-spectral-approaches-to-community-detection}{%
+\subsection{Non-Spectral Approaches to Community
+Detection}\label{non-spectral-approaches-to-community-detection}}
 

### 2021-05-26 proposal/slides.Rmd
updated slides

@@
 ---
-title: Community Detection Methods for the Generalized Random Dot Product Graph Model
+title: Community Detection Methods for Generalized Random Dot Product Graphs
 subtitle: Dissertation Proposal
@@
 
+\newcommand{\diag}{\text{diag}}
+\newcommand{\tr}{\text{Tr}}
+\newcommand{\blockdiag}{\text{blockdiag}}
+\newcommand{\indep}{\stackrel{\text{ind}}{\sim}}
+\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
+\newcommand{\Bernoulli}{\text{Bernoulli}}
+\newcommand{\Betadist}{\text{Beta}}
+\newcommand{\BG}{\text{BernoulliGraph}}
+\newcommand{\Cat}{\text{Categorical}}
+\newcommand{\Uniform}{\text{Uniform}}
+\newcommand{\RDPG}{\text{RDPG}}
+\newcommand{\GRDPG}{\text{GRDPG}}
+\newcommand{\PABM}{\text{PABM}}
+
+```{r out.width = '50%', fig.height = 3, fig.width = 4}
+Pz <- generate.P.beta(64, 4, 4, 1, 1, 4)
+P <- Pz$P
+z <- Pz$clustering
+A <- draw.graph(P)
+qgraph::qgraph(A, groups = factor(z), legend = FALSE)
+```
+
 How can we cluster the nodes of a network?
@@
 
-1. Define a generative model 
+1. Define a generative model for graph
 $G \mid z_1, ..., z_n, \vec{\theta} \sim P(\vec{z}, \vec{\theta})$.
 2. Develop a method for obtaining estimators $f(G) = \hat{z}_1, ..., \hat{z}_n$.
-3. Identify asymptotic properties of estimators and prove consistency.
+3. Identify asymptotic properties of estimators.
 
@@
 1. Probability Models for Networks
-    * Block Models and Community Structure
-    * (Generalized) Random Dot Product Graphs
-    * Connecting Block Models to the (G)RDPG
+    a. Block Models and Community Structure
+    b. (Generalized) Random Dot Product Graphs
+    c. Connecting Block Models to the (G)RDPG
 
@@
 2. Popularity Adjusted Block Model
-    * Connecting the PABM to the GRDPG
-    * Subspace Clustering for Community Detection
-    * Orthogonal Spectral Clustering
+    a. Connecting the PABM to the GRDPG
+    b. Subspace Clustering for Community Detection
+    c. Orthogonal Spectral Clustering
 
@@
 3. Community Detection for the (G)RDPG
-    * Manifold Clustering
-    * Manfolds as (G)RDPG Latent Configurations
+    a. Manifold Clustering
+    b. Manifolds as (G)RDPG Latent Configurations
 
@@
 
-$A \sim BernoulliGraph(P)$ iff:
+\vspace*{1\baselineskip}
+
+$A \sim \BG(P)$ iff:
 
@@
 vertices.
-2. $A_{ij} \stackrel{ind}{\sim} Bernoulli(P_{ij})$ for each $i < j$.
-
-Example: If $G$ is an Erdos-Renyi graph, then $P_{ij} = \theta$.
+2. $A_{ij} \indep \Bernoulli(P_{ij})$ for each $i < j$.
 
@@
 :::: {.column width=37.5%}
+
+**Example 1**: If every entry $P_{ij} = \theta$, then $A \sim \BG(P)$ is an Erdos-Renyi graph.
 
@@
 and each $P_{ij}$ depends on labels $z_i$ and $z_j$.  
-Then $A \sim BernoulliGraph(P)$ is a *block model*.
+Then $A \sim \BG(P)$ is a *block model*.
 
-Example 1: Stochastic Block Model with two communities
+**Example 2**: Stochastic Block Model with two communities
 
@@
 
-Random Dot Product Graph $A \sim RDPG(X)$  
+Random Dot Product Graph $A \sim \RDPG(X)$  
 (Young and Scheinerman, 2007)
@@
 $x_i^\top x_j \in [0, 1]$
-* $A \sim BernoulliGraph(X X^\top)$ where 
+* $A \sim \BG(X X^\top)$ where 
 $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
 
-Generalized Random Dot Product Graph $A \sim GRDPG_{p, q}(X)$  
+Generalized Random Dot Product Graph $A \sim \GRDPG_{p, q}(X)$  
 (Rubin-Delanchy, Cape, Tang, Priebe, 2020)
@@
 * Latent vectors $x_1, ..., x_n \in \mathbb{R}^{p+q}$ such that 
-$x_i^\top I_{p, q} x_j \in [0, 1]$ and $I_{p, q} = blockdiag(I_p, -I_q)$
-* $A \sim BernoulliGraph(X I_{p, q} X^\top)$ where 
+$x_i^\top I_{p, q} x_j \in [0, 1]$ and $I_{p, q} = \blockdiag(I_p, -I_q)$
+* $A \sim \BG(X I_{p, q} X^\top)$ where 
 $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
 
-If latent vectors $X_1, ..., X_n \stackrel{iid}{\sim} F$, then we write 
-$(A, X) \sim RDPG(F, n)$ or $(A, X) \sim GRDPG_{p, q}(F, n)$.
+If latent vectors $X_1, ..., X_n \iid F$, then we write 
+$(A, X) \sim \RDPG(F, n)$ or $(A, X) \sim \GRDPG_{p, q}(F, n)$.
 
@@
 All Bernoulli Graphs are RDPG (if $P$ is 
-positive semidefinite) or GRDPG.
+positive semidefinite)  
+or GRDPG (in general).
 
-Example 1 (cont'd): Assortative SBM ($p q > r^2$) with $K = 2$
+**Example 2** (cont'd): Assortative SBM ($p q > r^2$) with $K = 2$
 
@@
 ```{r fig.height = 3, fig.width = 6, out.width = '100%'}
-qgraph::qgraph(A, vsize = 4, groups = factor(z))
+qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
 ```
@@
 
-Example 1 (cont'd)
+**Example 2** (cont'd): If we want to perform community detection,
 
-If we want to perform community detection,
+1. Note that $A \sim \BG(P)$ is a RDPG since $P = X X^\top$.
+2. Compute the ASE $A \approx \hat{X} \hat{X}^\top$ with 
+$\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$.
+3. Apply clustering algorithm (e.g., $K$-means) to $\hat{X}$.
 
-1. Note that $A \sim BernoulliGraph(X X^\top)$ is a RDPG
-2. Compute the ASE $A \approx \hat{X} \hat{X}^\top$  
-$\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
-3. Apply clustering algorithm (e.g., $K$-means) to $\hat{X}$
-
-```{r, fig.height = 4, fig.width = 5, out.width = '50%'}
+```{r, fig.height = 5, fig.width = 5, out.width = '50%'}
 P.eigen = eigen(P)
@@
 $\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
-Then $A \sim PABM(P)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.
+Then $A \sim \PABM(P)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.
 
@@
 
-**Notation**: $A \sim PABM(\{\lambda^{(kl)}\}_K)$.
+**Notation**: $A \sim \PABM(\{\lambda^{(kl)}\}_K)$.
 
@@
 **Theorem** (KTT): 
-Given popularity vectors $\{\lambda^{(kl)}\}_2$, we can define block diagonal $X \in \mathbb{R}^{n \times 4}$ and $U \in \mathbb{O}(4)$ such that $A \sim PABM(\lambda^{(kl)}_2)$ is equivalent to $A \sim GRDPG(X U)$.
+Given popularity vectors $\{\lambda^{(kl)}\}_2$, we can define block diagonal $X \in \mathbb{R}^{n \times 4}$ and $U \in \mathbb{O}(4)$ such that $A \sim \PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to $A \sim \GRDPG_{3, 1}(X U)$.
 
@@
 
-**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
-$A \sim GRDPG_{p, q}(X U)$ with
+**Theorem** (KTT): $A \sim \PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
+$A \sim \GRDPG_{p, q}(X U)$ with
 
@@
 composed of $\{\lambda^{(kl)}\}_K$ 
-with each block corresponding to a community  
-$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
+with each block corresponding to a community.
+
+**Remark 1** (orthogonality of subspaces):
+If $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
 to different communities, then $x_i^\top x_j = 0$.
 
-**Remark** (non-uniqueness of the latent configuration):  
-If $A \sim GRDPG_{p, q}(X)$, then $A \sim GRDPG_{p, q}(X Q)$ for any $Q$ in 
+**Remark 2** (non-uniqueness of the latent configuration):  
+If $A \sim \GRDPG_{p, q}(Y)$, then $A \sim \GRDPG_{p, q}(Y Q)$ for any $Q$ in 
 the indefinite orthogonal group with signature $p, q$.
+
+**Remark 3**: Communities correspond to subspaces even with linear transformation $Q \in \mathbb{O}(p, q)$, but this may break the orthogonality property.
+
+## Orthogonal Spectral Clustering
+
+**Theorem** (KTT): 
+If $P = V \Lambda V^\top$ and $B = n V V^\top$,  
+then $B_{ij} = 0$ if $z_i \neq z_j$.
+
+**Algorithm**: Orthogonal Spectral Clustering:
+
+1. Let $V$ be the eigenvectors of $A$ corresponding to the $K (K+1)/2$ most 
+positive and $K (K-1) / 2$ most negative eigenvalues.
+2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
+3. Construct graph $G$ using $B$ as its similarity matrix.
+4. Partition $G$ into $K$ disconnected subgraphs.
+
+**Theorem** (KTT): 
+Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC. 
+Then $\forall$ pairs $(i, j)$ belonging to different communities 
+and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
+
+$$
+\max_{i, j} \hat{B}^{(ij)}_n = 
+O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
+$$
 
@@
 
-**Corollary**: Since $X$ is block diagonal by community and $U$ is orthogonal,
-each community corresponds to a subspace in $\mathbb{R}^{K^2}$.
+**Corollary**: The ASE of $A \sim \PABM(\{\lambda^{(kl)}\}_K)$ lies near a collection of $K$-dimensional subspaces in $K^2$ dimensions.
 
-Subspace property holds even with linear transformation 
-$Q \in \mathbb{O}(p, q)$.
-
-Sparse Subspace Clustering algorithm:
+**Algorithm**: Sparse Subspace Clustering:
 
 1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
-subject to $x_i = X c$ and $c_i^{(i)} = 0$.
+subject to $x_i = X c$ and $c^{(i)} = 0$.  
+This is typically performed via LASSO: 
+$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$
 2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
@@
 
-Step (1) of SSC typically performed via LASSO: 
-$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$
+## Sparse Subspace Clustering
+
+Noroozi et al. observed that the rank of $P$ is $K^2$ 
+and the columns of $P$ belonging to each community has rank $K$ 
+to justify SSC for the PABM.
+
+$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } A_{\cdot, i} = A c
+\text{ and } c^{(i)} = 0$$
+
+They were able to show that this obeys SDP if we replace $A$ with $P$.
+
+GRDPG-based approach: Apply SSC to the ASE of $A$.
+
+Stronger result: Apply SSC to the eigenvectors of $A$.
+
+$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } 
+\hat{v}_i = \hat{V} c \text{ and } c^{(i)} = 0$$
+$$A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$$
 
@@
 
+**Theorem** (KTT):
+
+Let
+
+* $P_n$ describe the edge probability matrix of the PABM with $n$ vertices, and 
+$A_n \sim \BG(P_n)$.
+* $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
+$K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues.
+
+Then
+
+* $\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
+$\sqrt{n} \hat{V}_n$ obeys the Subspace Detection Property with probability 1.
+
+Remarks:
+
+* For large $n$, we can identify $\lambda$ for SDP (Wang and Xu, 2016).
+* SDP does not guarantee community detection.
+
+## Simulation Study
+
+```{r clust_err_k, fig.width = 10, fig.height = 3, out.width = '100%'}
+setwd('~/dev/pabm-grdpg')
+clustering.df <- readr::read_csv('clustering-k.csv')
+ssc.df <- readr::read_csv('clustering-ssc-k.csv')
+clustering.df %>%
+  dplyr::group_by(n, K) %>%
+  dplyr::summarise(
+    med.err = median(error),
+    first.q = quantile(error, .25),
+    third.q = quantile(error, .75),
+    med.err.ssc = median(error.ssc),
+    first.q.ssc = quantile(error.ssc, .25),
+    third.q.ssc = quantile(error.ssc, .75),
+    med.err.ep = median(error.ep, na.rm = TRUE),
+    first.q.ep = quantile(error.ep, .25, na.rm = TRUE),
+    third.q.ep = quantile(error.ep, .75, na.rm = TRUE),
+    med.err.louvain = median(error.louvain),
+    first.q.louvain = quantile(error.louvain, .25),
+    third.q.louvain = quantile(error.louvain, .75)
+  ) %>% 
+  dplyr::ungroup() %>% 
+  dplyr::inner_join(
+    ssc.df %>% 
+      dplyr::group_by(n, K) %>% 
+      dplyr::summarise(med.err.ssc2 = median(error.ssc2),
+                       first.q.ssc2 = quantile(error.ssc2, .25),
+                       third.q.ssc2 = quantile(error.ssc2, .75)) %>% 
+      dplyr::ungroup()
+  ) %>% 
+  ggplot() +
+  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096)) +
+  # scale_x_continuous(breaks = c(128, 256, 512, 1024, 2048, 4096)) + 
+  scale_y_log10() +
+  labs(y = 'community detection error rate', 
+       colour = NULL) +
+  geom_line(aes(x = n, y = med.err,
+                colour = 'OSC')) +
+  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q,
+                    colour = 'OSC'), width = .1) + 
+  geom_line(aes(x = n, y = med.err.ssc,
+                colour = 'SSC-ASE')) + 
+  geom_errorbar(aes(x = n, ymin = first.q.ssc, ymax = third.q.ssc,
+                    colour = 'SSC-ASE'), width = .1) + 
+  # geom_line(aes(x = n, y = med.err.ep,
+  #               colour = 'MM-EP')) + 
+  # geom_errorbar(aes(x = n, ymin = first.q.ep, ymax = third.q.ep,
+  #                   colour = 'MM-EP'), width = .1) + 
+  geom_line(aes(x = n, y = med.err.ssc2,
+                colour = 'SSC-A')) + 
+  geom_errorbar(aes(x = n, ymin = first.q.ssc2, ymax = third.q.ssc2,
+                    colour = 'SSC-A'), width = .1) + 
+  geom_line(aes(x = n, y = med.err.louvain,
+                colour = 'MM-Louvain')) + 
+  geom_errorbar(aes(x = n, ymin = first.q.louvain, ymax = third.q.louvain,
+                    colour = 'MM-Louvain'), width = .1) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  facet_wrap(~ K, labeller = 'label_both')
+```
+
+# General Community Detection for the (G)RDPG
+
+## Recovery from the Adjacency Matrix
+
+```{r}
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+u1 <- runif(n1)
+u2 <- runif(n2)
+x1 <- cos(pi / 3 * u1)
+y1 <- sin(pi / 3 * u1)
+x2 <- 1 + cos(pi / 3 * u2 + pi)
+y2 <- 1 + sin(pi / 3 * u2 + pi)
+data.matrix <- cbind(c(x1, x2), c(y1, y2))
+```
+
+::: columns
+
+:::: {.column width=33%}
+
+```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
+par(mar = rep(1, 4))
+plot(data.matrix, col = z * 2, asp = 1, xlab = NA, ylab = NA,
+     main = 'Latent configuration')
+```
+
+::::
+
+:::: {.column width=33%}
+
+```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
+par(mar = rep(1, 4))
+P <- data.matrix %*% diag(c(1, 1)) %*% t(data.matrix)
+A <- draw.graph(P)
+qgraph::qgraph(A, groups = factor(z), layout = 'spring', legend = FALSE)
+title('(G)RDPG')
+```
+
+::::
+
+:::: {.column width=33%}
+
+```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
+par(mar = rep(1, 4))
+A.eigen <- eigen(A)
+Xhat <- A.eigen$vectors[, c(1, 2)] %*% diag(sqrt(abs(A.eigen$values[c(1, 2)])))
+plot(Xhat, col = z * 2, asp = 1, main = 'ASE')
+```
+
+::::
+
+:::
+
+## Generative Model
+
+Let $(A, X) \sim \GRDPG_{p, q}(F, n)$ as follows:
+
+1. Define functions $\gamma_1, ..., \gamma_K$ such that each 
+$\gamma_k : [0, 1]^r \mapsto \mathbb{R}^d$ and $\gamma_k(t) \neq \gamma_l(t)$ 
+when $k \neq l$.
+2. Sample labels
+$Z_1, ..., Z_n \iid \Cat(\pi_1, ..., \pi_K)$.
+3. Sample $T_1, ..., T_n \iid D$ with support $[0, 1]^r$. 
+4. Set latent positions $X_i = \gamma_{Z_i}(T_i)$ and 
+$X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
+5. $A \sim \BG(X I_{p, q} X^\top)$
+
+## Community Detection
+
+Recall that the ASE approximates the true latent positions:
+
+$$\max_i \|\hat{X}_i - Q_n X_i\| \stackrel{a.s.}{\to} 0$$
+
+This leads to a general community detection method:
+
+Given $A$, $K$, and $d$ (or $p$ and $q$),
+
+  1. Use ASE to approximate the latent configuration.
+
+  2. Use an appropriate clustering algorithm for the latent configuration.
+
+## Parallel Segments
+
+**Example 3**: 
+Let $U_1, ..., U_{n_1}, U_{n_1 + 1}, ..., U_{n} \iid \Uniform(0, \theta)$, $\gamma_1(t) = (t, 0)$, and $\gamma_2(t) = (t, a)$.
+$X_i = \gamma_1(U_i)$ for $i \leq n_1$ and $X_j = \gamma_2(U_j)$ for $n_1 + 1 \leq j \leq n$.
+If we observe $X_1, ..., X_{n_1}, X_{n_1+1}, ..., X_{n}$, what approach will allow us to group the observations by segment?
+
+```{r fig.height = 3, out.width = '50%'}
+set.seed(123456)
+
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+
+a <- .1
+
+X <- cbind(runif(n1, 0, cos(pi / 2 * a)), rep(0, n1))
+Y <- cbind(runif(n2, 0, cos(pi / 2 * a)), rep(a, n2))
+
+data.matrix <- rbind(X, Y)
+z <- c(rep(1, n1), rep(2, n2))
+
+as.data.frame(data.matrix) %>% 
+  dplyr::mutate(label = factor(z)) %>% 
+  ggplot() + 
+  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = 0, yend = 0), size = .1) + 
+  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = a, yend = a), size = .1) + 
+  geom_point(aes(x = V1, y = V2, colour = label)) + 
+  labs(x = NULL, y = NULL, colour = NULL) + 
+  guides(colour = FALSE) + 
+  xlim(-.1, cos(pi / 2 * a) + .1) + 
+  ylim(-a, 2 * a) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  # coord_fixed() + 
+  # ggthemes::theme_few() + 
+  theme_void() + 
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank())
+```
+
+$\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
+$\exists N(a, \delta, K) < \infty$ such that when $\min_k n_k \geq N$, 
+with probability at least $1 - \delta$,
+
+1. Single linkage clustering entails perfect community detection.
+
+2. Any $\epsilon$-neighborhood graph with $\epsilon \leq a$ will consist of 
+at least $K$ disjoint subgraphs such that no subgraph contains members of two 
+different communities.
+
+## Noisy Parallel Segments and One-Dimensional Manifolds
+
+**Example 4**: 
+Starting with the parallel segments as before, suppose instead of observing $X_1, ..., X_n$, we have noisy observations $X_1 + \xi_1, ..., X_n + \xi_n$ such that 
+$\max_i \|\xi_i\| = \xi \leq a / 3$.
+
+```{r fig.height = 3, out.width = '50%'}
+# xi1 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
+xi1 <- rep(0, n)
+xi2 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
+
+as.data.frame(data.matrix) %>% 
+  dplyr::mutate(V1 = V1 + xi1, V2 = V2 + xi2) %>% 
+  dplyr::mutate(label = factor(z)) %>% 
+  ggplot() + 
+  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = 0, yend = 0), size = .1) + 
+  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = a, yend = a), size = .1) + 
+  geom_point(aes(x = V1, y = V2, colour = label)) + 
+  labs(x = NULL, y = NULL, colour = NULL) + 
+  guides(colour = FALSE) + 
+  xlim(-.1, cos(pi / 2 * a) + .1) + 
+  ylim(-a, 2 * a) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  # coord_fixed() + 
+  # ggthemes::theme_few() + 
+  theme_void() + 
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank())
+```
+
+Then we can derive similar statements as in Example 3.
+
+This also holds for noisy points sampled uniformly on one-dimensional manifolds such that the manifolds are distance at least $a$ apart.
+
+Since the ASE of a RDPG generated from points on these segments/curves will reconstruct the original segments/curves with noise, we can apply this to the RDPG (if sufficient $n$).
+
+## Future Work
+
+1. Derive formal statements showing that the ASE of a random graph generated by these latent configurations produces the correct conditions for community detection.
+
+2. Extend results to non-uniform distributions.
+
+3. Extend results to multidimensional manifolds.
+
+4. Relax condition for the minimum distance between manifolds.
+
+5. Explore the performance of these techniques in the presence of outliers.
+
+6. Extend results to the GRDPG.
+
+# Additional Slides
+
+## Subspace Detection Property
+
 If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
-$B_{ij} = 0$ if $x_i$ and $x_j$ belong to different subspaces
+$B_{ij} = 0$ if $x_i$ and $x_j$ belong to different subspaces.
 
@@
 
-qgraph::qgraph(sign(W), layout = 'spring', groups = factor(z))
+qgraph::qgraph(sign(W), layout = 'spring', groups = factor(z), legend = FALSE)
 ```
@@
 :::
-
-## Sparse Subspace Clustering
-
-Noroozi et al. observed that the rank of $P$ is $K^2$ 
-and the columns of $P$ belonging to each community has rank $K$ 
-to justify SSC for the PABM.
-
-$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } A_{\cdot, i} = A c
-\text{ and } c^{(i)} = 0$$
-
-They were able to show that this obeys SDP if we replace $A$ with $P$.
-
-GRDPG-based approach: Apply SSC to the ASE of $A$.
-
-$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } 
-\hat{x}_i = \hat{X} c \text{ and } c^{(i)} = 0$$
-$$A \approx \hat{X} I_{p, q} \hat{X}^\top$$
-
-## Sparse Subspace Clustering
 
@@
 
-**Theorem** (KTT): 
-If $P = V \Lambda V^\top$ and $B = n V V^\top$, 
-then $B_{ij} = 0$ $\forall i, j$ in different communities.
-
-## Sparse Subspace Clustering
-
-**Theorem** (KTT):
-
-Let
-
-* $P_n$ describe the edge probability matrix of the PABM with $n$ vertices, and 
-$A_n \sim BernoulliGraph(P_n)$.
-* $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
-$K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues.
-
-Then
-
-* $\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
-$\sqrt{n} \hat{V}_n$ obeys the Subspace Detection Property with probability 1.
-
-Remarks:
-
-* For large $n$, we can identify $\lambda$ for SDP (Wang and Xu, 2016).
-* SDP does not guarantee community detection.
-
-## Orthogonal Spectral Clustering
-
-**Theorem** (KTT): 
-If $P = V \Lambda V^\top$ and $B = n V V^\top$, 
-then $B_{ij} = 0$ $\forall i, j$ in different communities.
-
-Orthogonal Spectral Clustering algorithm:
-
-1. Let $V$ be the eigenvectors of $A$ corresponding to the $K (K+1)/2$ most 
-positive and $K (K-1) / 2$ most negative eigenvalues.
-2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
-3. Construct graph $G$ using $B$ as its similarity matrix.
-4. Partition $G$ into $K$ disconnected subgraphs.
-
-**Theorem** (KTT): 
-Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC. 
-Then $\forall$ pairs $(i, j)$ belonging to different communities 
-and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
-
-$$
-\max_{i, j} \hat{B}^{(ij)} = 
-O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
-$$
-
-# General Community Detection for the (G)RDPG
-
-## Generative Model
-
-Let $(A, X) \sim RDPG(F, n)$ such that
-
-1. Define functions $f_1, ..., f_K$ such that 
-$f_k : [0, 1]^r \mapsto \mathbb{R}^d$ and $f_k(t) \neq f_l(t)$ 
-$\forall k, l \in [K]$. 
-2. Sample labels
-$Z_1, ..., Z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
-3. Sample $T_1, ..., T_n \stackrel{iid}{\sim} D$ with support $[0, 1]^r$. 
-4. Set latent positions $X_i = f_{Z_i}(T_i)$ and 
-$X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
-5. $A \sim BernoulliGraph(X X^\top)$
-
-## Community Detection
-
-Athreya et al. and Rubin-Delanchy et al.: we can approximate properties of 
-the latent configurations via ASE.
-
-General community detection method:  
-Given $A$, $K$, and $d$ (or $p$ and $q$),
-
-  1. Use ASE to approximate the latent configuration.
-  2. Use the appropriate clustering algorithm for the form of the latent
-  configuration (manifolds).
-
-## Parallel Segments
-
-**Example**: 
-Let $U_1, ..., U_{n_1}, U_{n_1 + 1}, ..., U_{n} \stackrel{iid}{\sim} Uniform(0, \cos \frac{\pi}{2} a)$, $f_1(t) = (t, 0)$, and $f_2(t) = (t, a)$.
-$X_i = f_1(U_i)$ for $i \leq n_1$ and $X_j = f_2(U_j)$ for $n_1 + 1 \leq j \leq n$.
-If we observe $X_1, ..., X_{n_1}, X_{n_1+1}, ..., X_{n}$, what approach will allow us to group the observations by segment?
-
-```{r fig.height = 3, out.width = '50%'}
-set.seed(123456)
-
-n1 <- 2 ** 5
-n2 <- 2 ** 5
-n <- n1 + n2
-
-a <- .1
-
-X <- cbind(runif(n1, 0, cos(pi / 2 * a)), rep(0, n1))
-Y <- cbind(runif(n2, 0, cos(pi / 2 * a)), rep(a, n2))
-
-data.matrix <- rbind(X, Y)
-z <- c(rep(1, n1), rep(2, n2))
-
-as.data.frame(data.matrix) %>% 
-  dplyr::mutate(label = factor(z)) %>% 
-  ggplot() + 
-  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = 0, yend = 0), size = .1) + 
-  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = a, yend = a), size = .1) + 
-  geom_point(aes(x = V1, y = V2, colour = label)) + 
-  labs(x = NULL, y = NULL, colour = NULL) + 
-  guides(colour = FALSE) + 
-  xlim(-.1, cos(pi / 2 * a) + .1) + 
-  ylim(-a, 2 * a) + 
-  scale_colour_brewer(palette = 'Set1') + 
-  # coord_fixed() + 
-  # ggthemes::theme_few() + 
-  theme_void() + 
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank())
-```
-
-$\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
-$\exists N(a, \delta, K) < \infty$ such that when $\min_k n_k \geq N$, 
-with probability at least $1 - \delta$,
-
-1. Single linkage clustering will produce perfect community detection.
-2. Any $\epsilon$-neighborhood graph with $\epsilon \leq a$ will consist of 
-at least $K$ disjoint subgraphs such that no subgraph contains members of two 
-different communities.
-
-## Noisy Parallel Segments and One-Dimensional Manifolds
-
-**Example**: 
-Starting with the parallel segments as before, suppose instead of observing $X_1, ..., X_n$, we have noisy observations $X_1 + \xi_1, ..., X_n + \xi_n$ such that 
-$\max_i \|\xi_i\| = \xi \leq a / 3$.
-
-```{r fig.height = 3, out.width = '50%'}
-# xi1 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
-xi1 <- rep(0, n)
-xi2 <- runif(n, -a / 6 / sqrt(2), a / 6 / sqrt(2))
-
-as.data.frame(data.matrix) %>% 
-  dplyr::mutate(V1 = V1 + xi1, V2 = V2 + xi2) %>% 
-  dplyr::mutate(label = factor(z)) %>% 
-  ggplot() + 
-  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = 0, yend = 0), size = .1) + 
-  geom_segment(aes(x = 0, xend = cos(pi / 2 * a), y = a, yend = a), size = .1) + 
-  geom_point(aes(x = V1, y = V2, colour = label)) + 
-  labs(x = NULL, y = NULL, colour = NULL) + 
-  guides(colour = FALSE) + 
-  xlim(-.1, cos(pi / 2 * a) + .1) + 
-  ylim(-a, 2 * a) + 
-  scale_colour_brewer(palette = 'Set1') + 
-  # coord_fixed() + 
-  # ggthemes::theme_few() + 
-  theme_void() + 
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank())
-```
-
-Then $\forall a \in (0, 1)$, $\delta \in (0, 1)$, $K \geq 2$, $\xi \leq a / 3$, 
-$\exists N(a, \delta, K, \xi) < \infty$ such that when $\min_k n_k \geq N$, 
-with probability at least $1 - \delta$,
-
-1. Single linkage will produce perfect community detection.
-2. Any $\epsilon$-neighborhood graph will consist of at least $K$ sub-graphs with no subgraph containing vertices from multiple communities.
-
-This also holds for noisy points sampled from one-dimensional manifolds such that the manifolds are distance at least $a$ apart.
-
-## Recovery from the Adjacency Matrix
-
-```{r}
-n1 <- 2 ** 5
-n2 <- 2 ** 5
-n <- n1 + n2
-z <- c(rep(1, n1), rep(2, n2))
-u1 <- runif(n1)
-u2 <- runif(n2)
-x1 <- cos(pi / 3 * u1)
-y1 <- sin(pi / 3 * u1)
-x2 <- 1 + cos(pi / 3 * u2 + pi)
-y2 <- 1 + sin(pi / 3 * u2 + pi)
-data.matrix <- cbind(c(x1, x2), c(y1, y2))
-```
-
-::: columns
-
-:::: {.column width=33%}
-
-```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
-par(mar = rep(1, 4))
-plot(data.matrix, col = z * 2, asp = 1, xlab = NA, ylab = NA,
-     main = 'Latent configuration')
-```
-
-::::
-
-:::: {.column width=33%}
-
-```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
-par(mar = rep(1, 4))
-P <- data.matrix %*% t(data.matrix)
-A <- draw.graph(P)
-qgraph::qgraph(A, groups = factor(z), layout = 'spring')
-title('RDPG')
-```
-
-::::
-
-:::: {.column width=33%}
-
-```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
-par(mar = rep(1, 4))
-A.eigen <- eigen(A)
-Xhat <- A.eigen$vectors[, 1:2] %*% diag(sqrt(A.eigen$values[1:2]))
-plot(Xhat, col = z * 2, asp = 1, main = 'ASE')
-```
-
-::::
-
-:::
-
-## Future Work
-
-1. Show that the ASE of a random graph generated by these latent vectors 
-produces the correct conditions for sufficiently large $n$.
-2. Extend results to non-uniform distributions.
-3. Extend results to multidimensional manifolds.
-4. Relax condition for the minimum distance between manifolds.
-5. Explore more robust clustering techniques for these latent configurations.
-6. Extend results to the GRDPG.
-
-# Additional Slides
-
 ## Parameter Estimation
@@
 * $Q^\top Q \neq I$
-* If $A \sim GRDPG_{p, q}(X)$, then $A \sim GRDPG_{p, q}(X Q)$
+* If $A \sim \GRDPG_{p, q}(X)$, then $A \sim \GRDPG_{p, q}(X Q)$
 * $(Q x)^\top (Q y) = x^\top Q^\top Q y \neq x^\top y$
@@
 
-## Community Detection in Block Models
+## Non-Spectral Community Detection for Block Models
 
@@
 
-* $Z_i \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$
-* $\lambda_{ik} \stackrel{ind}{\sim} Beta(a_{ik}, b_{ik})$
+* $Z_i \iid Categorical(\pi_1, ..., \pi_K)$
+* $\lambda_{ik} \indep Beta(a_{ik}, b_{ik})$
 

### 2021-05-28 proposal/slides.Rmd
added citation for ssc

@@
     a. Connecting the PABM to the GRDPG
-    b. Subspace Clustering for Community Detection
-    c. Orthogonal Spectral Clustering
+    b. Orthogonal Spectral Clustering
+    c. Sparse Subspace Clustering
 
@@
 **Example 1**: If every entry $P_{ij} = \theta$, then $A \sim \BG(P)$ is an Erdos-Renyi graph.  
+For this model, $A_{ij} \iid \Bernoulli(\theta)$.
 
@@
 A <- draw.graph(P)
-qgraph::qgraph(A, vsize = 4, groups = factor(z))
+qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
 ```
@@
 $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$.
-3. Apply clustering algorithm (e.g., $K$-means) to $\hat{X}$.
+3. Apply clustering algorithm (e.g., $K$-means) to $\hat{X}$,  
+noting that as $n \to \infty$, the ASE approaches point masses.
 
@@
 $\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
-Then $A \sim \PABM(P)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.
+Then $A \sim \PABM(P)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$,  
+e.g., if $z_i = k$ and $z_j = l$, $P_{ij} = \lambda_{il} \lambda_{jk}$.
 
@@
 
+$$X = \begin{bmatrix}
+\Lambda^{(1)} & \cdots & 0 \\
+0 & \ddots & 0 \\
+0 & \cdots & \Lambda^{(K)}
+\end{bmatrix}$$
+
+$$\Lambda^{(k)} = \begin{bmatrix} 
+\lambda^{(k1)} & \cdots & \lambda^{(kK)} 
+\end{bmatrix}$$
+
+## Connecting the PABM to the GRDPG
+
+::: columns
+
+:::: column
+
+$$X = \begin{bmatrix}
+\Lambda^{(1)} & \cdots & 0 \\
+0 & \ddots & 0 \\
+0 & \cdots & \Lambda^{(K)}
+\end{bmatrix}$$
+
+::::
+
+:::: column
+
+\vspace*{2\baselineskip}
+
+$$U \in \mathbb{O}(K^2)$$
+
+::::
+
+:::
+
+$$A \sim \PABM(\{\lambda^{(kl)}\}_K) \iff A \sim \GRDPG_{p, q}(X U)$$
+
 **Remark 1** (orthogonality of subspaces):
-If $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
-to different communities, then $x_i^\top x_j = 0$.
+If $y_i^\top$ and $y_j^\top$ are two rows of $X U$ corresponding 
+to different communities, then $y_i^\top y_j = 0$.
 
@@
 
-**Algorithm**: Sparse Subspace Clustering:
+**Algorithm**: Sparse Subspace Clustering (Elhamifar and Vidal, 2009):
 

### 2021-05-29 proposal/proposal.Rmd
updated slides

@@
 Common clustering methods typically involve calculating some central or representative point for each cluster around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}).
-Because these methods involve computing summary statistics within each cluster, such as the sample average, they cannot be applied directly to graphs, necessitating methods for transforming the graph data into feature vectors.
+Because these methods involve computing summary statistics within each cluster, such as the sample average, they cannot be applied directly to graphs, necessitating methods for transforming the graph data into feature vectors if they are to be used.
 
 One family of methods for unifying graph community detection with traditional clustering techniques is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the graph into Euclidean space, followed by applying a popular clustering algorithm such as $K$-means clustering.
-The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models for graphs involving latent vectors in Euclidean space.
-In such a model, each vertex has a corresponding vector in latent space, and the edge probabilities between pairs of vertices is computed by operations on corresponding pairs of latent vectors.
+The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models for graphs via latent vectors in Euclidean space.
+In such a model, each vertex has a corresponding vector in latent space, and the edge probabilities between pairs of vertices are computed by operations on the corresponding pairs of latent vectors.
 A community detection algorithm motivated by this may involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.
@@
 The aim of our research is to develop consistent community detection techniques under the RDPG and GRDPG frameworks.
-First, we explore existing generative graph models with underlying community structures that can be inferred by connecting these models to the RDPG or GRDPG.
+First, we explore existing generative graph models with underlying community structures (Block Models) that can be inferred by connecting these models to the RDPG or GRDPG.
 In particular, we use the connection between the Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1} and the GRDPG to motivate two community detection methods.
@@
 Let $G = (V, E)$ be an undirected, unweighted graph with no self-loops and $n$ vertices.
-Denote $A \in \{0, 1\}^{n \times n}$ as the adjacency matrix for $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $A_{ij} = 0$ otherwise.
+Denote $A \in \{0, 1\}^{n \times n}$ as the adjacency matrix of $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $A_{ij} = 0$ otherwise.
 Because $G$ is symmetric and contains no self-loops, $A_{ij} = A_{ji}$ and $A_{ii} = 0$ for $i, j \in [n]$.
@@
 In order to motivate community detection methods, we restrict the generative model such that for each pair of vertices $(i, j)$, the probability of an edge between the vertices $P_{ij}$ depends in some way on their labels $z_i$ and $z_j$.
+Such models are called Block Models.
 
-One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = p_{z_i z_j}$ where $p_{kl} = p_{lk}$ is the probability of an edge between each vertex in community $k$ to each vertex in community $l$.
+One type of Block Model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one of the $K$ communities, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = p_{z_i z_j}$ where $p_{kl} = p_{lk}$ is the probability of an edge between each vertex in community $k$ to each vertex in community $l$.
 The homogeneous SBM further restricts $P$ to two unique entries: $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise.
@@
 
-If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
+If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A \approx V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
 Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
@@
 
-\citet{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,
+\citet{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $\{W_n\}$,
 
@@
 
-where $Q_n$ is a sequence of matrices in $\mathbb{O}(p, q)$, the indefinite orthogonal group of order $p, q$.
+where $\{Q_n\}$ is a sequence of matrices in $\mathbb{O}(p, q)$, the indefinite orthogonal group of order $p, q$.
 
-It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all graphs generated by $P$ regardless of positive semidefiniteness are special cases of the GRDPG.
+It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all Bernoulli Graphs generated by $P$ regardless of positive semidefiniteness are special cases of the GRDPG.
 This includes the SBM, DCBM, and PABM.
@@
 This is shown by reconstructing $P$ as $P = X X^\top$.
+Thus the SBM is equivalent to a RDPG with the latent configuration of two point masses, one for each community.
 \end{example}
@@
 The ASE of the assortative SBM consists of points in $\mathbb{R}^K$ that lie near one of $K$ centers, depending on the community label, leading to ASE followed by $K$-means clustering (or similar, e.g., GMM) as a consistent community detection algorithm \cite{lyzinski2014}.
-A similar result can be shown for the DCBM and PABM but with different structures in the ASE.
+A similar result has been shown for the DCBM \cite{lyzinski2014}, and our work extends this to the PABM.
 
@@
 We further show that while the latent configuration is not unique, there exists one in particular for which the subspaces are orthogonal.
-This leads to two straightforward methods for community detection, one involving computing inner products in the ASE and another which is a slight modification of an existing algorithm for subspace clustering.
-Finally, we will briefly explore alternative, non-spectral-based methods for community detection in the ASE involving Expectation Maximization, MCMC sampling, and Mean Field Variational Inference.
+This leads to two straightforward methods for community detection, one involving computing inner products in the ASE and another which is a straightforward application of an existing algorithm for subspace clustering.
+Finally, we will briefly explore alternative, non-spectral-based methods for community detection in the ASE involving Expectation Maximization and Bayesian methods.
 
@@
 that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
-realization of $\Bernoulli(P)$, i.e., 
-$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ 
-($A_{ij} = A_{ji}$ and $A_{ii} = 0$).
+realization of $\BG(P)$.
 
@@
 
-It is straightforward to show that the PABM, as well as all graphs such that the adjacency matrix is drawn such that $A_{ij} \indep Bernoulli(P_{ij})$, is a special case of the GRDPG.
+It is straightforward to show that the PABM (as well as all Bernoulli Graphs) is a special case of the GRDPG.
 It can also be shown that the latent positions of the PABM under the GRDPG framework consists of $K$ $K$-dimensional subspaces in $\mathbb{R}^{K^2}$.
@@
 In Theorem \ref{theorem2}, we showed that a possible latent configuration for a $K$-community PABM under the GRDPG framework consists of vectors in $\mathbb{R}^{K^2}$ such that each community corresponds to a $K$-dimensional subspace and each subspace is orthogonal to the others.
+However, because latent configurations are not unique, we cannot apply this fact directly for community detection.
+
+\begin{remark}
+Let $A \sim \text{GRDPG}_{p, q}(Y)$. Then $A\sim \text{GRDPG}_{p, q}(Y Q)$ $\forall Q \in \mathbb{O}(p, q)$. This is because $Y Q I_{p, q} Q^\top Y^\top = Y I_{p, q} Y^\top$. 
+\end{remark}
+
+However, we can show that if we treat the rows of the eigenvectors of $P$ (ignoring the eigenvalues) as an embedding (unscaled ASE), then they always form orthogonal subspaces.
+This is stated more formally in Theorem \ref{osc}, and we call the resulting algorithm Orthogonal Spectral Clustering.
 
@@
 
-## Alternative Approaches to Community Detection
+## Non-Spectral Approaches to Community Detection
 

### 2021-05-29 proposal/proposal.tex
updated slides

@@
 Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs.
-In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models, first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then by proposing clustering methods for these feature vector representations and deriving the asymptotic properties of these methods.
+In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure), first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then applying clustering methods for these feature vector representations, and finally deriving the asymptotic properties of these methods.
 
@@
 necessitating methods for transforming the graph data into feature
-vectors.
+vectors if they are to be used.
 
@@
 Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
-this further by explicitly constructing generative models for graphs
-involving latent vectors in Euclidean space. In such a model, each
-vertex has a corresponding vector in latent space, and the edge
-probabilities between pairs of vertices is computed by operations on
-corresponding pairs of latent vectors. A community detection algorithm
-motivated by this may involve learning the latent positions given an
-observed graph and then learning the community labels given the latent
-positions.
+this further by explicitly constructing generative models for graphs via
+latent vectors in Euclidean space. In such a model, each vertex has a
+corresponding vector in latent space, and the edge probabilities between
+pairs of vertices are computed by operations on the corresponding pairs
+of latent vectors. A community detection algorithm motivated by this may
+involve learning the latent positions given an observed graph and then
+learning the community labels given the latent positions.
 
@@
 existing generative graph models with underlying community structures
-that can be inferred by connecting these models to the RDPG or GRDPG. In
-particular, we use the connection between the Popularity Adjusted Block
-Model \cite{307cbeb9b1be48299388437423d94bf1} and the GRDPG to motivate
-two community detection methods. Then we explore other latent structures
-or mixture distributions in the latent space that induce graphs for
-which consistent community detection is possible.
+(Block Models) that can be inferred by connecting these models to the
+RDPG or GRDPG. In particular, we use the connection between the
+Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1}
+and the GRDPG to motivate two community detection methods. Then we
+explore other latent structures or mixture distributions in the latent
+space that induce graphs for which consistent community detection is
+possible.
 
@@
 and \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
-adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
+adjacency matrix of \(G\) such that \(A_{ij} = 1\) if there exists an
 edge between vertices \(i\) and \(j\) and \(A_{ij} = 0\) otherwise.
@@
 \((i, j)\), the probability of an edge between the vertices \(P_{ij}\)
-depends in some way on their labels \(z_i\) and \(z_j\).
-
-One such model is the Stochastic Block Model (SBM)
+depends in some way on their labels \(z_i\) and \(z_j\). Such models are
+called Block Models.
+
+One type of Block Model is the Stochastic Block Model (SBM)
 \cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
-each vertex belonging to one community, the SBM restricts \(P\) to
-\(K (K + 1) / 2\) unique entries such that \(P_{ij} = p_{z_i z_j}\)
-where \(p_{kl} = p_{lk}\) is the probability of an edge between each
-vertex in community \(k\) to each vertex in community \(l\). The
-homogeneous SBM further restricts \(P\) to two unique entries:
-\(P_{ij} = p\) if \(z_i = z_j\) and \(P_{ij} = q\) otherwise. Several
-generalizations of the SBM have been introduced since, including the
-Degree Corrected Block Model (DCBM) \cite{Karrer_2011} and the
-Popularity Adjusted Block Model (PABM)
+each vertex belonging to one of the \(K\) communities, the SBM restricts
+\(P\) to \(K (K + 1) / 2\) unique entries such that
+\(P_{ij} = p_{z_i z_j}\) where \(p_{kl} = p_{lk}\) is the probability of
+an edge between each vertex in community \(k\) to each vertex in
+community \(l\). The homogeneous SBM further restricts \(P\) to two
+unique entries: \(P_{ij} = p\) if \(z_i = z_j\) and \(P_{ij} = q\)
+otherwise. Several generalizations of the SBM have been introduced
+since, including the Degree Corrected Block Model (DCBM)
+\cite{Karrer_2011} and the Popularity Adjusted Block Model (PABM)
 \cite{307cbeb9b1be48299388437423d94bf1}. Like the SBM, these models
@@
 \((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
-ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),
+ASE of \(A_n\), for some sequence of orthogonal matrices \(\{W_n\}\),
 
@@
 
-where \(Q_n\) is a sequence of matrices in \(\mathbb{O}(p, q)\), the
+where \(\{Q_n\}\) is a sequence of matrices in \(\mathbb{O}(p, q)\), the
 indefinite orthogonal group of order \(p, q\).
@@
 semidefinite \(P\) are special cases of the RDPG, which is a special
-case of the GRDPG, and all graphs generated by \(P\) regardless of
-positive semidefiniteness are special cases of the GRDPG. This includes
-the SBM, DCBM, and PABM. In the following example, we show that under
-the RDPG framework, the latent configuration that induces the SBM takes
-on a very particular form.
+case of the GRDPG, and all Bernoulli Graphs generated by \(P\)
+regardless of positive semidefiniteness are special cases of the GRDPG.
+This includes the SBM, DCBM, and PABM. In the following example, we show
+that under the RDPG framework, the latent configuration that induces the
+SBM takes on a very particular form.
 
@@
 GMM) as a consistent community detection algorithm \cite{lyzinski2014}.
-A similar result can be shown for the DCBM and PABM but with different
-structures in the ASE.
+A similar result has been shown for the DCBM \cite{lyzinski2014}, and
+our work extends this to the PABM.
 
@@
 detection, one involving computing inner products in the ASE and another
-which is a slight modification of an existing algorithm for subspace
-clustering. Finally, we will briefly explore alternative,
+which is a straightforward application of an existing algorithm for
+subspace clustering. Finally, we will briefly explore alternative,
 non-spectral-based methods for community detection in the ASE involving
-Expectation Maximization, MCMC sampling, and Mean Field Variational
-Inference.
+Expectation Maximization and Bayesian methods.
 
@@
 that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
-realization of $\text{Bernoulli}(P)$, i.e., 
-$A_{ij} \stackrel{\text{ind}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ 
-($A_{ij} = A_{ji}$ and $A_{ii} = 0$).
+realization of $\text{BernoulliGraph}(P)$.
 
@@
 
-It is straightforward to show that the PABM, as well as all graphs such
-that the adjacency matrix is drawn such that
-\(A_{ij} \stackrel{\text{ind}}{\sim}Bernoulli(P_{ij})\), is a special
-case of the GRDPG. It can also be shown that the latent positions of the
-PABM under the GRDPG framework consists of \(K\) \(K\)-dimensional
-subspaces in \(\mathbb{R}^{K^2}\). While there is no unique latent
-configuration \(X\) such that \(X X^\top = P\), the edge probability
-\(P\) for the PABM, they all have this subspace structure, and one in
-particular consists of \emph{orthogonal} subspaces.
+It is straightforward to show that the PABM (as well as all Bernoulli
+Graphs) is a special case of the GRDPG. It can also be shown that the
+latent positions of the PABM under the GRDPG framework consists of \(K\)
+\(K\)-dimensional subspaces in \(\mathbb{R}^{K^2}\). While there is no
+unique latent configuration \(X\) such that \(X X^\top = P\), the edge
+probability \(P\) for the PABM, they all have this subspace structure,
+and one in particular consists of \emph{orthogonal} subspaces.
 
@@
 corresponds to a \(K\)-dimensional subspace and each subspace is
-orthogonal to the others.
+orthogonal to the others. However, because latent configurations are not
+unique, we cannot apply this fact directly for community detection.
+
+\begin{remark}
+Let $A \sim \text{GRDPG}_{p, q}(Y)$. Then $A\sim \text{GRDPG}_{p, q}(Y Q)$ $\forall Q \in \mathbb{O}(p, q)$. This is because $Y Q I_{p, q} Q^\top Y^\top = Y I_{p, q} Y^\top$. 
+\end{remark}
+
+However, we can show that if we treat the rows of the eigenvectors of
+\(P\) (ignoring the eigenvalues) as an embedding (unscaled ASE), then
+they always form orthogonal subspaces. This is stated more formally in
+Theorem \ref{osc}, and we call the resulting algorithm Orthogonal
+Spectral Clustering.
 

### 2021-05-29 proposal/slides.Rmd
updated slides

@@
 subtitle: Dissertation Proposal
-author: John Koo
+author: 
+  - John Koo$^1$
+  - Michael Trosset$^1$
+  - Minh Tang$^2$
+  - Julia Fukuyama$^1$
+  - Roni Khardon$^3$
+  - Fangzheng Xie$^1$
+institute:
+  - $^1$Indiana University, Department of Statistics
+  - $^2$North Carolina State University, Department of Statistics
+  - $^3$Indiana University, Department of Computer Science
 date: 'June 2021'
@@
 
-**Algorithm**: Sparse Subspace Clustering (Elhamifar and Vidal, 2009):
+**Algorithm**: Sparse Subspace Clustering (Elhamifar & Vidal, 2009):
 
@@
                     colour = 'SSC-A'), width = .1) + 
-  geom_line(aes(x = n, y = med.err.louvain,
-                colour = 'MM-Louvain')) + 
-  geom_errorbar(aes(x = n, ymin = first.q.louvain, ymax = third.q.louvain,
-                    colour = 'MM-Louvain'), width = .1) + 
+  # geom_line(aes(x = n, y = med.err.louvain,
+  #               colour = 'MM-Louvain')) + 
+  # geom_errorbar(aes(x = n, ymin = first.q.louvain, ymax = third.q.louvain,
+  #                   colour = 'MM-Louvain'), width = .1) + 
   scale_colour_brewer(palette = 'Set1') + 
@@
 5. $A \sim \BG(X I_{p, q} X^\top)$
+
+Examples:
+
+1. $r = 1$ and each $\gamma_k$ is a point mass $\implies$ SBM
+2. $r = 1$ and the range of each $\gamma_k$ is a segment from the origin to point $y_k$ with $\|y_k\| \leq 1$ $\implies$ DCBM
+3. $r = K$ and the range of each $\gamma_k$ is the hyperplane of coordinates $x_{K (k-1) + 1}$ to $x_{k K}$ $\implies$ PABM
 
@@
 
-5. Explore the performance of these techniques in the presence of outliers.
+5. Extend results to the GRDPG.
 
-6. Extend results to the GRDPG.
+# Thank you
 
@@
 ## Parameter Estimation
+
+1. Let $A^{(kl)}$ be the block of edges between communities $k$ and $l$. 
+2. For each $k, l \in [K]$, do:
+    i. Compute the SVD $A^{(kl)} = U \Sigma V^\top$.
+    ii. Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
+    Assign $\sigma^{(kl)} \leftarrow \Sigma_{11}^{1/2}$. 
+    iii. Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
+    $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
+
+**Theorem** (KTT):
+
+$$\max_{k, l} \|\hat{\lambda}^{(kl)} - \lambda^{(kl)}\| =
+O_P \Big( \frac{(\log n_k)^c}{\sqrt{n_k}} \Big)$$
 

### 2021-05-29 proposal/slides.Rmd
updated slides

@@
 ---
-title: Community Detection Methods for Generalized Random Dot Product Graphs
-subtitle: Dissertation Proposal
-author: 
-  - John Koo$^1$
-  - Michael Trosset$^1$
-  - Minh Tang$^2$
-  - Julia Fukuyama$^1$
-  - Roni Khardon$^3$
-  - Fangzheng Xie$^1$
-institute:
-  - $^1$Indiana University, Department of Statistics
-  - $^2$North Carolina State University, Department of Statistics
-  - $^3$Indiana University, Department of Computer Science
-date: 'June 2021'
 output: 
@@
     includes:
-      in_header: page_headers.tex
+      in_header: 
+        - page_headers.tex
+        - slidestemplate.tex
 header-includes:
@@
 ```
+
+## {.plain}
+
+\center
+
+\LARGE
+
+\textcolor{darkred}{Community Detection Methods for the Generalized Random Dot Product Graph}
+
+\normalsize
+
+Dissertation Proposal
+
+June 2021
+
+\vspace*{1\baselineskip}
+
+::: columns
+
+:::: {.column width=33%}
+
+John Koo^1^  
+*PhD Candidate*
+
+::::
+
+:::: {.column width=33%}
+
+*Research Committee*  
+Dr. Michael Trosset^1^  
+Dr. Minh Tang^2^  
+Dr. Julia Fukuyama^1^  
+Dr. Roni Khardon^3^  
+Dr. Fangzheng Xie^1^
+
+::::
+
+:::
+
+\vspace*{1\baselineskip}
+
+\tiny
+
+^1^Department of Statistics, Indiana University  
+^2^Department of Statistics, North Carolina State University  
+^3^Department of Computer Science, Indiana University
+
+\normalsize
 

### 2021-06-01 proposal/proposal.Rmd
updated slides
fixed typo

@@
 Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample.
-Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
+Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i}^\top c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
 The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$.
@@
 \begin{equation} \label{eq:ssc}
-c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
+c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i}^\top c \|_2^2 + \lambda \|c\|_1
 \end{equation}
@@
 
-# Estimated Timeline of Completion
+## Estimated Timeline of Completion
 

### 2021-06-01 proposal/slides.Rmd
fixed typo

@@
 
-## Connecting the PABM to the GRDPG ($K = 2$)
-
-**Theorem** (KTT): 
-Given popularity vectors $\{\lambda^{(kl)}\}_2$, we can define block diagonal $X \in \mathbb{R}^{n \times 4}$ and $U \in \mathbb{O}(4)$ such that $A \sim \PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to $A \sim \GRDPG_{3, 1}(X U)$.
-
-Proof: Decompose $P$ as follows
-
-$$X = \begin{bmatrix}
-\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
-0 & 0 & \lambda^{(21)} & \lambda^{(22)}
-\end{bmatrix}$$
-
-$$U = \begin{bmatrix}
-1 & 0 & 0 & 0 \\
-0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
-0 & 0 & 1 / \sqrt{2} & -1 / \sqrt{2} \\
-0 & 1 & 0 & 0
-\end{bmatrix}$$
-
-$$P = (X U) I_{3, 1} (X U)^\top = 
-\begin{bmatrix}
-\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
-\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
-\end{bmatrix}$$
-
 ## Connecting the PABM to the GRDPG
@@
 0 & \cdots & \Lambda^{(K)}
-\end{bmatrix}$$
+\end{bmatrix} 
+\in \mathbb{R}^{n \times K^2}$$
 
@@
 \lambda^{(k1)} & \cdots & \lambda^{(kK)} 
-\end{bmatrix}$$
+\end{bmatrix} 
+\in \mathbb{R}^{n_k \times K}$$
 
@@
 1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
-subject to $x_i = X c$ and $c^{(i)} = 0$.  
+subject to $x_i = X^\top c$ and $c^{(i)} = 0$.  
 This is typically performed via LASSO: 
-$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$
+$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i}^\top c\|_2^2 + \lambda \|c\|_1$
 2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
@@
 If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
-$B_{ij} = 0$ if $x_i$ and $x_j$ belong to different subspaces.
+$B_{ij} = 0$ for each $x_i$ and $x_j$ belonging to different subspaces.
 
@@
 
-```{r, fig.height = 5, fig.width = 5, out.width = '75%'}
+```{r, fig.height = 3, fig.width = 4, out.width = '50%'}
 set.seed(314159)
+
+par(mar = rep(1, 4))
 
@@
 
-```{r, fig.height = 5, fig.width = 5, out.width = '75%'}
+```{r, fig.height = 3, fig.width = 5, out.width = '75%'}
 Y <- t(X)

### 2021-06-01 proposal/proposal.Rmd
fixed typo

@@
 Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample.
-Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i}^\top c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
+Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X^\top c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$.
 The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$.

### 2021-06-01 proposal/proposal.Rmd
found another typo

@@
 
-As discussed in \S 2.2, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space.
+As discussed in \S 2.2, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (assortative) SBM is a RDPG with a very specific structure in the latent space.
 A similar result has been shown for the DCBM \cite{lyzinski2014}.

### 2021-06-01 proposal/proposal.tex
updated slides
fixed typo
found another typo

@@
 As discussed in \S 2.2, all Bernoulli graph models can be expressed as a
-RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
+RDPG or GRDPG model, and in particular, the (assortative) SBM is a RDPG
 with a very specific structure in the latent space. A similar result has
@@
 \(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
-\(c_i = \arg\min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
+\(c_i = \arg\min\limits_{c} \|c\|_1\) subject to \(x_i = X^\top c\) and
 \(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
@@
 \hypertarget{estimated-timeline-of-completion}{%
-\section{Estimated Timeline of
+\subsection{Estimated Timeline of
 Completion}\label{estimated-timeline-of-completion}}

### 2021-06-01 proposal/slides.Rmd
updated slides

@@
 
-1. Note that $A \sim \BG(P)$ is a RDPG since $P = X X^\top$.
+1. Note that $A$ is a RDPG because $P = X X^\top$.
 2. Compute the ASE $A \approx \hat{X} \hat{X}^\top$ with 
@@
 $$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } 
-\hat{v}_i = \hat{V} c \text{ and } c^{(i)} = 0$$
+\hat{v}_i = \hat{V}^\top c \text{ and } c^{(i)} = 0$$
 $$A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$$
@@
 * $P_n$ describe the edge probability matrix of the PABM with $n$ vertices, and 
-$A_n \sim \BG(P_n)$.
+$A_n \sim \BG(P_n)$;
 * $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
@@
 
-* $\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
-$\sqrt{n} \hat{V}_n$ obeys the Subspace Detection Property with probability 1.
+* For some $\lambda > 0$ and $N < \infty$, $\sqrt{n} \hat{V}_n$ obeys the Subspace Detection Property with probability 1 when $n > N$.
 
@@
 
-This leads to a general community detection method:
+This suggests a general approach to community detection:
 
@@
 
-$\forall a \in (0, 1)$, $\delta \in (0, 1)$, and $K \geq 2$, 
-$\exists N(a, \delta, K) < \infty$ such that when $\min_k n_k \geq N$, 
+Then for any $a \in (0, 1)$, $\delta \in (0, 1)$, 
+we can find $N(a, \delta) < \infty$ such that when $\min_k n_k \geq N$, 
 with probability at least $1 - \delta$,
@@
 
-* $Z_i \iid Categorical(\pi_1, ..., \pi_K)$
-* $\lambda_{ik} \indep Beta(a_{ik}, b_{ik})$
+* $Z_i \iid \Cat(\pi_1, ..., \pi_K)$
+* $\lambda_{ik} \indep \Betadist(a_{ik}, b_{ik})$
 
@@
 * Approximate solution for the PABM
-  * $Z_i \mid \{a'_{ik}\}, \{b'_{ik}\} \sim Categorical(\pi'_1, ..., \pi'_K)$
+  * $Z_i \mid \{a'_{ik}\}, \{b'_{ik}\} \sim \Cat(\pi'_1, ..., \pi'_K)$
   * $\lambda_{ik} \mid 
-  \{a'_{-i, -k}\}, \{b'_{-i, -k}\}, \{\pi'_k\} \sim Beta(a'_{ik}, b'_{ik})$
+  \{a'_{-i, -k}\}, \{b'_{-i, -k}\}, \{\pi'_k\} \sim \Betadist(a'_{ik}, b'_{ik})$
   * Iteratively update $\{\pi'_K\}, \{a'_{ik}\}, \{b'_{ik}\}$ until convergence

### 2021-06-02 proposal/slides.Rmd
minor tweaks

@@
 
-* $P_{ij} = \theta$
-* Not a block model
+* $P_{ij} = \theta$ (not a block model)
+* 1 parameter $\theta$
 
@@
 Then $\forall$ pairs $(i, j)$ belonging to different communities 
-and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
+and sparsity factor satisfying $n \rho_n = \omega\big((\log n)^{4c}\big)$,
 
@@
 
-Examples:
-
-1. $r = 1$ and each $\gamma_k$ is a point mass $\implies$ SBM
-2. $r = 1$ and the range of each $\gamma_k$ is a segment from the origin to point $y_k$ with $\|y_k\| \leq 1$ $\implies$ DCBM
-3. $r = K$ and the range of each $\gamma_k$ is the hyperplane of coordinates $x_{K (k-1) + 1}$ to $x_{k K}$ $\implies$ PABM
-
 ## Community Detection

### 2021-06-03 proposal/prospectus.Rmd
added prospectus

---
title: |
  | Community Detection for the 
  | Generalized Random Dot Product Graph
subtitle: Dissertation Prospectus
author: John Koo
institution: Indiana University
output:
  pdf_document:
    keep_tex: true
    citation_package: natbib
    number_sections: yes
fontsize: 11pt
# geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
bibliography: proposal.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      # eval = FALSE,
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 3, 
                      fig.width = 5, 
                      fig.dpi = 300)

options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')

import::from(magrittr, `%>%`)
library(ggplot2)
```

\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}
\newcommand{\BG}{\text{BernoulliGraph}}
\newcommand{\Cat}{\text{Categorical}}
\newcommand{\GRDPG}{\text{GRDPG}}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{example}
\newtheorem*{example}{Example}

Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision. One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels). Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs. Common examples of such clustering techniques include $K$-means clustering \cite{1056489} and Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}. In addition, these techniques assume that the vectors are sampled from a specific type of distribution (whether implicitly in the case of $K$-means clustering or explicitly in the case of GMM). In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure). First, we demonstrate that under the Generalized Random Dot Product Graph \cite{rubindelanchy2017statistical} framework, all graphs generated by Block Models (generative graph models with inherent community structure) can be represented as collections of vectors in Euclidean space via the Adjacency Spectral Embedding \cite{athreya2017statistical, lyzinski2014}. Then, noting the particular structure or configuration that the vectors take, we select appropriate clustering techniques to apply on the vector representation of the graph. Finally we derive the asymptotic properties of these methods to show that our methods are consistent or achieve desirable properties with high probability. To illustrate this approach, we primarily focus on the Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1}.

Our work is essentially about connecting two well-known families of generative graph models, the Block Model and the (Generalized) Random Dot Product Graph \cite{rubindelanchy2017statistical} in order to perform community detection. It is straightforward to show that all Block Models are various cases of the GRDPG. This fact has been leveraged to develop community detection methods \cite{athreya2017statistical, lyzinski2014, rubindelanchy2017statistical} for the Stochastic Block Model \cite{doi:10.1080/0022250X.1971.9989788} and Degree Corrected Block Model \cite{Karrer_2011}. Our work begins by extending this to the Popularity Adjusted Block Model. We show that under the GRDPG framework, the latent configuration that generates the PABM consists of orthogonal subspaces. Then we show that two community detection algorithms arise naturally from this latent configuration: Orthogonal Spectral Clustering, which is an algorithm of our own design, and an existing algorithm, Sparse Subspace Clustering \cite{5206547}. We then extend this work to more general configurations in the latent space that imply community structure. The overall goal of our research is to develop a general framework or approach to community detection for a wide range of Block Models.

\newpage

The overall generative model for inducing community structure can be described by the following.

Let $(A, X) \sim \GRDPG_{p, q}(F, n)$ such that:

1. Define functions $\gamma_1, ..., \gamma_K$ such that each 
$\gamma_k : [0, 1]^r \mapsto \mathbb{R}^d$ and $\gamma_k(t) \neq \gamma_l(t)$ 
when $k \neq l$.
2. Sample labels
$Z_1, ..., Z_n \iid \Cat(\pi_1, ..., \pi_K)$.
3. Sample $T_1, ..., T_n \iid D$ with support $[0, 1]^r$. 
4. Set latent positions $X_i = \gamma_{Z_i}(T_i)$ and 
$X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
5. $A \sim \BG(X I_{p, q} X^\top)$

The SBM, DCBM, and PABM are all special cases of this generative model. Based on the structure of the $\gamma_k$ functions that correspond to each of the SBM, DCBM, and PABM, there are clear choices for clustering algorithms on the embeddings of each of these Block Models. Extensions of these results may consider which clustering algorithms result in consistent estimators given restrictions on each of the $\gamma_k$, and how to approach community detection when $\gamma_k$ are unknown altogether.

The estimated timeline of completion is as follows:

1. Literature review and complete proofs of main theorems: December 2022

2. Simulations, real data analyses, R package: March 2022

3. Dissertation Completion: April 2022

### 2021-06-03 proposal/prospectus.tex
added prospectus

--- 2021-05-26 proposal/proposal.tex
@@
 \hypersetup{
+  pdfauthor={John Koo},
   colorlinks=true,
@@
 
-\author{}
-\date{\vspace{-2.5em}}
+\title{Community Detection for the\\
+Generalized Random Dot Product Graph}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Dissertation Prospectus}
+\author{John Koo}
+\date{}
 
 \begin{document}
-
-\pagenumbering{gobble}
-
-%\begin{titlepage}
-\begin{center}
-\LARGE{\textbf{Community Detection Methods for the Generalized Random Dot Product Graph Model}}\\
-\vspace*{2\baselineskip}
-\normalsize{A dissertation proposal submitted in partial satisfaction of the requirements for the degree of \\}
-Doctor of Philosophy \\
-in \\
-Statistical Science \\
-\vspace*{2\baselineskip}
-\Large{John Koo}\\
-\vspace*{3\baselineskip}
-\Large{\textbf{Research Committee Members}}\\
-Dr. Michael Trosset \\
-Dr. Minh Tang \\
-Dr. Julia Fukuyama \\
-Dr. Roni Khardon \\
-Dr. Fangzheng Xie \\
-\vspace*{3\baselineskip}
-June 2, 2021 \\
-\vspace*{1\baselineskip}
-Department of Statistics \\
-Indiana University \\
-Bloomington, Indiana \\
-\end{center}
-% \end{titlepage}
-
-\hypersetup{linkcolor = black}
-\newpage
-\pagenumbering{roman}
-\tableofcontents
-\addcontentsline{toc}{section}{\contentsname}
-
-\newpage
-
-\begin{center}
-\LARGE{Abstract}
-\end{center}
-
-\vspace*{2\baselineskip}
-
-\normalsize
-
-Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision.
-One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels).
-Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs.
-In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models, first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then by proposing clustering methods for these feature vector representations and deriving the asymptotic properties of these methods.
-
-\newpage
-\pagenumbering{arabic}
-\hypersetup{linkcolor = blue}
+\maketitle
 
@@
 \newcommand{\Cat}{\text{Categorical}}
+\newcommand{\GRDPG}{\text{GRDPG}}
 \newtheorem{definition}{Definition}
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+Graph and network data, in which samples are represented not as a
+collection of feature vectors but as relationships between pairs of
+observations, are increasingly widespread in various fields ranging from
+sociology to computer vision. One common goal of analyzing graph data is
+community detection or graph clustering, in which the graph is
+partitioned into disconnected subgraphs in an unsupervised yet
+meaningful manner (e.g., by optimizing an objective function or
+recovering unobserved labels). Because traditional clustering techniques
+were developed for data that can be represented as vectors, they cannot
+be applied directly to graphs. Common examples of such clustering
+techniques include \(K\)-means clustering \cite{1056489} and Gaussian
+Mixture Models \cite{doi:10.1198/016214502760047131}. In addition, these
+techniques assume that the vectors are sampled from a specific type of
+distribution (whether implicitly in the case of \(K\)-means clustering
+or explicitly in the case of GMM). In this research, we investigate the
+use of a family of spectral decomposition based approaches for community
+detection in block models (random graph models with inherent community
+structure). First, we demonstrate that under the Generalized Random Dot
+Product Graph \cite{rubindelanchy2017statistical} framework, all graphs
+generated by Block Models (generative graph models with inherent
+community structure) can be represented as collections of vectors in
+Euclidean space via the Adjacency Spectral Embedding
+\cite{athreya2017statistical, lyzinski2014}. Then, noting the particular
+structure or configuration that the vectors take, we select appropriate
+clustering techniques to apply on the vector representation of the
+graph. Finally we derive the asymptotic properties of these methods to
+show that our methods are consistent or achieve desirable properties
+with high probability. To illustrate this approach, we primarily focus
+on the Popularity Adjusted Block Model
+\cite{307cbeb9b1be48299388437423d94bf1}.
 
-\hypertarget{research-goal}{%
-\subsection{Research Goal}\label{research-goal}}
+Our work is essentially about connecting two well-known families of
+generative graph models, the Block Model and the (Generalized) Random
+Dot Product Graph \cite{rubindelanchy2017statistical} in order to
+perform community detection. It is straightforward to show that all
+Block Models are various cases of the GRDPG. This fact has been
+leveraged to develop community detection methods
+\cite{athreya2017statistical, lyzinski2014, rubindelanchy2017statistical}
+for the Stochastic Block Model \cite{doi:10.1080/0022250X.1971.9989788}
+and Degree Corrected Block Model \cite{Karrer_2011}. Our work begins by
+extending this to the Popularity Adjusted Block Model. We show that
+under the GRDPG framework, the latent configuration that generates the
+PABM consists of orthogonal subspaces. Then we show that two community
+detection algorithms arise naturally from this latent configuration:
+Orthogonal Spectral Clustering, which is an algorithm of our own design,
+and an existing algorithm, Sparse Subspace Clustering \cite{5206547}. We
+then extend this work to more general configurations in the latent space
+that imply community structure. The overall goal of our research is to
+develop a general framework or approach to community detection for a
+wide range of Block Models.
 
-Graph and network data have become increasingly widespread in various
-fields including sociology, neuroscience, biostatistics, and computer
-science. This has resulted in various challenges for researchers who
-rely on traditional statistical and machine learning methods, many of
-which are incompatible with graph data and instead require the data to
-exist as feature vectors in Euclidean space. This includes graph
-clustering and community detection, which are often the goal of data
-analysis on graphs and networks. Common clustering methods typically
-involve calculating some central or representative point for each
-cluster around which the data belonging to that cluster lie (e.g.,
-Lloyd's algorithm for \(K\)-means clustering \cite{1056489}, Gaussian
-Mixture Models \cite{doi:10.1198/016214502760047131}). Because these
-methods involve computing summary statistics within each cluster, such
-as the sample average, they cannot be applied directly to graphs,
-necessitating methods for transforming the graph data into feature
-vectors.
+\newpage
 
-One family of methods for unifying graph community detection with
-traditional clustering techniques is Spectral Clustering
-\cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the
-graph into Euclidean space, followed by applying a popular clustering
-algorithm such as \(K\)-means clustering. The Random Dot Product Graph
-(RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot
-Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
-this further by explicitly constructing generative models for graphs
-involving latent vectors in Euclidean space. In such a model, each
-vertex has a corresponding vector in latent space, and the edge
-probabilities between pairs of vertices is computed by operations on
-corresponding pairs of latent vectors. A community detection algorithm
-motivated by this may involve learning the latent positions given an
-observed graph and then learning the community labels given the latent
-positions.
+The overall generative model for inducing community structure can be
+described by the following.
 
-The aim of our research is to develop consistent community detection
-techniques under the RDPG and GRDPG frameworks. First, we explore
-existing generative graph models with underlying community structures
-that can be inferred by connecting these models to the RDPG or GRDPG. In
-particular, we use the connection between the Popularity Adjusted Block
-Model \cite{307cbeb9b1be48299388437423d94bf1} and the GRDPG to motivate
-two community detection methods. Then we explore other latent structures
-or mixture distributions in the latent space that induce graphs for
-which consistent community detection is possible.
-
-\hypertarget{notation}{%
-\subsection{Notation}\label{notation}}
-
-Let \(G = (V, E)\) be an undirected, unweighted graph with no self-loops
-and \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
-adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
-edge between vertices \(i\) and \(j\) and \(A_{ij} = 0\) otherwise.
-Because \(G\) is symmetric and contains no self-loops,
-\(A_{ij} = A_{ji}\) and \(A_{ii} = 0\) for \(i, j \in [n]\). We further
-restrict our analyses to independent Bernoulli graphs. Let
-\(P \in [0, 1]^{n \times n}\) be a symmetric matrix of edge
-probabilities. Graph \(G\) is sampled from \(P\) by drawing
-\(A_{ij} \stackrel{\text{ind}}{\sim}\text{Bernoulli}(P_{ij})\) for each
-\(1 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)). We
-denote \(A \sim \text{BernoulliGraph}(P)\) as graph with adjacency
-matrix \(A\) sampled from edge probability matrix \(P\) in this manner.
-If each vertex has a hidden label in \([K]\), they are denoted as
-\(z_1, ..., z_n\). Finally, we denote
-\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
-as the sample \(x_1, ..., x_n \in \mathbb{R}^d\).
-
-\hypertarget{preliminaries}{%
-\section{Preliminaries}\label{preliminaries}}
-
-\hypertarget{block-models}{%
-\subsection{Block Models}\label{block-models}}
-
-Generative models for independent Bernoulli graphs involve defining the
-edge probability matrix \(P\) whose \(ij\)\textsuperscript{th} entry is
-the probability of an edge between vertices \(i\) and \(j\) for each
-\(i, j \in [n]\). In order to motivate community detection methods, we
-restrict the generative model such that for each pair of vertices
-\((i, j)\), the probability of an edge between the vertices \(P_{ij}\)
-depends in some way on their labels \(z_i\) and \(z_j\).
-
-One such model is the Stochastic Block Model (SBM)
-\cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
-each vertex belonging to one community, the SBM restricts \(P\) to
-\(K (K + 1) / 2\) unique entries such that \(P_{ij} = p_{z_i z_j}\)
-where \(p_{kl} = p_{lk}\) is the probability of an edge between each
-vertex in community \(k\) to each vertex in community \(l\). The
-homogeneous SBM further restricts \(P\) to two unique entries:
-\(P_{ij} = p\) if \(z_i = z_j\) and \(P_{ij} = q\) otherwise. Several
-generalizations of the SBM have been introduced since, including the
-Degree Corrected Block Model (DCBM) \cite{Karrer_2011} and the
-Popularity Adjusted Block Model (PABM)
-\cite{307cbeb9b1be48299388437423d94bf1}. Like the SBM, these models
-involve imposing some restriction on \(P\) based on community labels
-(and possibly other factors).
-
-\hypertarget{generalized-random-dot-product-graphs}{%
-\subsection{(Generalized) Random Dot Product
-Graphs}\label{generalized-random-dot-product-graphs}}
-
-Like the SBM, DCBM, and PABM, the RDPG and GRDPG are generative models
-for graphs involving an edge probability matrix \(P\). The RDPG starts
-with points in latent space \(X \in \mathcal{X} \subset \mathbb{R}^d\)
-for \(\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}\).
-\(P\) is then constructed as \(P = X X^\top\), and
-\(A \sim \text{BernoulliGraph}(P)\). We provide a more formal definition
-of the RDPG and GRDPG as follows:
-
-\begin{definition}[(Generalized) Random Dot Product Graph]
-Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ where $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}$.
-$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix is drawn as $A \sim \text{BernoulliGraph}(X X^\top)$.
-
-If on the other hand, $\mathcal{X} = \{x, y \in \mathbb{R}^{p+q} : x^\top I_{p, q} y \in [0, 1]\}$ and $A \sim \text{BernoulliGraph}(X I_{p, q} X^\top)$ for $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
-These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
-
-In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $X_1, ..., X_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
-If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
-\end{definition}
-
-The structure of the RDPG and GRDPG provides a straightforward method
-for recovery of the latent positions via spectral decomposition.
-
-\begin{definition}[Adjacency Spectral Embedding]
-Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$.
-Let $A \approx V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
-Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.
-
-If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
-Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
-\end{definition}
-
-\citet{athreya2017statistical} showed that under mild conditions, if
-\((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
-ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - W_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-Similarly, \citet{rubindelanchy2017statistical} showed that for
-\((A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)\),
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - Q_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-where \(Q_n\) is a sequence of matrices in \(\mathbb{O}(p, q)\), the
-indefinite orthogonal group of order \(p, q\).
-
-It is straightforward to show that all Bernoulli graphs with positive
-semidefinite \(P\) are special cases of the RDPG, which is a special
-case of the GRDPG, and all graphs generated by \(P\) regardless of
-positive semidefiniteness are special cases of the GRDPG. This includes
-the SBM, DCBM, and PABM. In the following example, we show that under
-the RDPG framework, the latent configuration that induces the SBM takes
-on a very particular form.
-
-\begin{example}[Connecting the SBM to the RDPG]
-Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respectively, and the between-community edge probability is $r$, with $p q > r^2$.
-Let community 1 have $n_1$ vertices, community 2 have $n_2$ vertices, and $n_1 + n_2 = n$.
-Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$.
-Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$.
-Then one RDPG representation of this SBM is:
-
-$$X = \begin{bmatrix} 
-\sqrt{p} & 0 \\
-\vdots & \vdots \\
-\sqrt{p} & 0 \\
-\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\ 
-\vdots & \vdots \\
-\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
-\end{bmatrix}
-\in \mathbb{R}^{n \times 2}$$
-
-where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$.
-This is shown by reconstructing $P$ as $P = X X^\top$.
-\end{example}
-
-The ASE of the assortative SBM consists of points in \(\mathbb{R}^K\)
-that lie near one of \(K\) centers, depending on the community label,
-leading to ASE followed by \(K\)-means clustering (or similar, e.g.,
-GMM) as a consistent community detection algorithm \cite{lyzinski2014}.
-A similar result can be shown for the DCBM and PABM but with different
-structures in the ASE.
-
-\hypertarget{manifold-learning}{%
-\subsection{Manifold Learning}\label{manifold-learning}}
-
-\citet{trosset2020learning} showed that the ASE of a RDPG can be used to
-recover one-dimensional manifolds. Suppose
-\(f : [0, 1] \mapsto \mathcal{X}\) such that \(f\) is smooth and
-\(\mathcal{X}\) represents a curve or one-dimensional manifold in
-\(\mathbb{R}^d\). If \(t_1, ..., t_n \stackrel{\text{iid}}{\sim}F\) such
-that \(F\) has support \([0, 1]\), the latent positions are
-\(x_i = f(t_i)\) with \(y_i\) is its corresponding point in the scaled
-ASE, and \(d_{\epsilon}(\cdot, \cdot)\) is the shortest path distance of
-an \(\epsilon\)-neighborhood graph. Under certain mild conditions, the
-shortest path distances of the \(\epsilon\)-neighborhood graph of the
-ASE approaches the arc lengths along \(f\):
-
-\begin{equation}
-d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{\sum_r^d \Big( \frac{d f_r}{d t} \Big)^2} dt
-\end{equation}
-
-\citet{athreya2020estimation} extended this further by generating a RDPG
-from a mixture of distributions on a curve. In their example, points
-were sampled from a mixture of two Beta distributions on the
-Hardy-Weinberg curve to construct the latent positions of a RDPG, with
-the goal of recovering the hidden mixture distribution from an observed
-graph.
-
-\hypertarget{the-popularity-adjusted-block-model}{%
-\section{The Popularity Adjusted Block
-Model}\label{the-popularity-adjusted-block-model}}
-
-As discussed in \S 2.2, all Bernoulli graph models can be expressed as a
-RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
-with a very specific structure in the latent space. A similar result has
-been shown for the DCBM \cite{lyzinski2014}. We can also extend this to
-the PABM. In \S 3.1, we show that the PABM is equivalent to the GRDPG
-model for which the communities correspond to subspaces in the latent
-configuration. We further show that while the latent configuration is
-not unique, there exists one in particular for which the subspaces are
-orthogonal. This leads to two straightforward methods for community
-detection, one involving computing inner products in the ASE and another
-which is a slight modification of an existing algorithm for subspace
-clustering. Finally, we will briefly explore alternative,
-non-spectral-based methods for community detection in the ASE involving
-Expectation Maximization, MCMC sampling, and Mean Field Variational
-Inference.
-
-\hypertarget{connecting-the-pabm-to-the-grdpg}{%
-\subsection{Connecting the PABM to the
-GRDPG}\label{connecting-the-pabm-to-the-grdpg}}
-
-We first define the PABM using the construction provided by
-\citet{noroozi2019estimation}.
-
-\begin{definition}[Popularity Adjusted Block Model]
-\label{pabm}
-Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
-set of $n$ 
-vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
-columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
-$P^{(kl)}$ describes the edge probabilities between vertices in communities 
-$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
-Let graph $G = (V, E)$ be an undirected, unweighted graph such 
-that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
-realization of $\text{Bernoulli}(P)$, i.e., 
-$A_{ij} \stackrel{\text{ind}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ 
-($A_{ij} = A_{ji}$ and $A_{ii} = 0$).
-
-If each block $P^{(kl)}$ can be written as the outer product of two vectors:
-
-\begin{equation} \label{eq:pabm}
-  P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top
-\end{equation}
-
-for a set of $K^2$ fixed vectors $\{\lambda^{(st)}\}_{s, t = 1}^K$ where each 
-$\lambda^{(st)}$ is a column vector 
-of dimension $n_s$, then graph $G$ and its corresponding adjacency matrix $A$ 
-is a realization of a popularity adjusted block model with parameters 
-$\{\lambda^{(st)}\}_{s, t = 1}^K$. 
-\end{definition}
-
-We will use the notation \(A \sim \text{PABM}(\{\lambda^{(kl)}\}_K)\) to
-denote a random adjacency matrix \(A\) drawn from a PABM with parameters
-\(\lambda^{(kl)}\) consisting of \(K\) underlying communities.
-
-It is straightforward to show that the PABM, as well as all graphs such
-that the adjacency matrix is drawn such that
-\(A_{ij} \stackrel{\text{ind}}{\sim}Bernoulli(P_{ij})\), is a special
-case of the GRDPG. It can also be shown that the latent positions of the
-PABM under the GRDPG framework consists of \(K\) \(K\)-dimensional
-subspaces in \(\mathbb{R}^{K^2}\). While there is no unique latent
-configuration \(X\) such that \(X X^\top = P\), the edge probability
-\(P\) for the PABM, they all have this subspace structure, and one in
-particular consists of \emph{orthogonal} subspaces.
-
-\begin{theorem}[Connecting the PABM to the GRDPG for $K = 2$]
-\label{theorem1}  
-Let
-
-$$X = \begin{bmatrix} 
-\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\ 
-0 & 0 & \lambda^{(21)} & \lambda^{(22)} 
-\end{bmatrix}$$
-
-$$U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
-0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
-0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
-0 & 1 & 0 & 0 \end{bmatrix}$$
-
-where each $\lambda^{(kl)}$ is a vector as in Definition 1. 
-Then $A \sim GRDPG_{3, 1}(X U)$ and $A \sim PABM(\{(\lambda^{(kl)}\}_2)$ are 
-equivalent.
-\end{theorem}
-
-\begin{theorem}[Generalization to $K > 2$] 
-\label{theorem2}
-There exists a block diagonal matrix 
-$X \in \mathbb{R}^{n \times K^2}$ defined by PABM parameters 
-$\{\lambda^{(kl)}\}_K$ and orthonormal matrix 
-$U \in \mathbb{R}^{K^2 \times K^2}$ that is fixed 
-for each $K$ such that $A \sim GRDPG_{K (K+1) / 2, K (K-1) / 2}(XU)$ and 
-$A \sim PABM(\{(\lambda^{(kl)}\})_K)$ are equivalent.
-\end{theorem}
-
-\begin{proof}
-Define the following matrices from $\{\lambda^{(kl)}\}_K$:
-
-$$\Lambda^{(k)} = 
-\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
-\in \mathbb{R}^{n_k \times K}$$
-
-\begin{equation} \label{eq:xy}
-X = \text{blockdiag}(\Lambda^{(1)}, ..., \Lambda^{(K)}) \in \mathbb{R}^{n \times K^2}
-\end{equation}
-
-$$L^{(k)} = \text{blockdiag}(\lambda^{(1k)}, ..., \lambda^{(Kk)}) \in 
-\mathbb{R}^{n \times K}$$
-
-$$Y = \begin{bmatrix} L^{(1)} & \cdots & L^{(K)} \end{bmatrix} \in 
-\mathbb{R}^{n \times K^2}$$
-
-Then $P = X Y^\top$.
-
-Note that $Y = X \Pi$ for a permutation matrix
-$\Pi$, resulting in $P = X \Pi X^\top$.  
-The permutation described by $\Pi$ has $K$ fixed points, which correspond to 
-$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where 
-$k = r (K + 1) + 1$ for $r = 0, ..., K - 1$. It also has 
-$\binom{K}{2} = K (K - 1) / 2$ cycles of order $2$. Each cycle corresponds to 
-a pair of eigenvalues $+1$ and $-1$ and a pair of eigenvectors 
-$(e_s + e_t) / \sqrt{2}$ and $(e_s - e_t) / \sqrt{2}$.
-
-Then $\Pi$ has $K (K + 1) / 2$ eigenvalues equal to $1$ and $K (K - 1) / 2$ 
-eigenvalues equal to $-1$. $\Pi$ has the decomposed form
-
-\begin{equation} \label{eq:permutation}
-\Pi = U I_{K (K + 1) / 2, K (K - 1) / 2} U^\top
-\end{equation}
-
-The edge probability matrix then can be written as:
-
-\begin{equation} \label{eq:pabm-grdpg}
-P = X U I_{p, q} (X U)^\top
-\end{equation}
-
-\begin{equation} \label{eq:p}
-p = K (K + 1) / 2
-\end{equation}
-
-\begin{equation} \label{eq:q}
-q = K (K - 1) / 2
-\end{equation}
-
-and we can describe the PABM with $K$ communities as a GRDPG with latent 
-positions $X U$ with signature $\big( K (K + 1) / 2, K (K - 1) / 2 \big)$.
-\end{proof}
-
-\begin{example}[$K = 3$] Using the same notation as in Theorem \ref{theorem2}:
-
-$$X = \begin{bmatrix} 
-\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
-\end{bmatrix}$$
-
-$$Y = \begin{bmatrix} 
-\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
-0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
-0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
-\end{bmatrix}$$
-
-Then $P = X Y^\top$ and $Y = X \Pi$ where $\Pi$ is a permutation matrix 
-consisting of $3$ fixed points and $3$ cycles of order 2:
-
-$$\Pi = \begin{bmatrix} 
-1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
-0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
-0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1
-\end{bmatrix}$$
-
-\item Positions 1, 5, 9 are fixed.
-
-\item The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.
-
-Therefore, we can decompose $\Pi = U I_{6, 3} U^\top$ where the first three 
-columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the fixed 
-positions $1$, $5$, and $9$, the next three columns consist of eigenvectors 
-$(e_k + e_l) / \sqrt{2}$, and the last three columns consist of eigenvectors 
-$(e_k - e_l) / \sqrt{2}$, where pairs $(k, l)$ correspond to the cycles of 
-order 2 described above.
-
-The latent positions are the rows of  
-$$XU = \begin{bmatrix}
-  \lambda^{(11)} & 0 & 0 & 
-  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 & 
-  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 \\
-  0 & \lambda^{(22)} & 0 & 
-  \lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} & 
-  -\lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} \\
-  0 & 0 & \lambda^{(33)} & 
-  0 & \lambda^{(31)} / \sqrt{2} & \lambda^{(32)} / \sqrt{2} & 
-  0 & -\lambda^{(31)} / \sqrt{2} & -\lambda^{(32)} / \sqrt{2}
-\end{bmatrix}$$
-\end{example}
-
-In Theorem \ref{theorem2}, we showed that a possible latent
-configuration for a \(K\)-community PABM under the GRDPG framework
-consists of vectors in \(\mathbb{R}^{K^2}\) such that each community
-corresponds to a \(K\)-dimensional subspace and each subspace is
-orthogonal to the others.
-
-\begin{theorem}
-\label{osc}
-Let $P = V \Lambda V^\top$ be the spectral decomposition of the edge probability matrix of a PABM. Define $B = n V V^\top$. Then $B_{ij} = 0$ $\forall i, j$ in different communities. 
-\end{theorem}
-
-If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), then results from
-\citet{rubindelanchy2017statistical} imply
-\(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top\). Then for
-vertices \(i\) and \(j\) belonging to separate communities, the
-\(ij\)\textsuperscript{th} entry of \(n \hat{V} \hat{V}^\top\)
-approaches 0 with probability 1:
-
-\begin{theorem}
-\label{theorem4} 
-Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
-(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
-and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
-
-\begin{equation} \label{eq:thm4}
-\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
-O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
-\end{equation}
-
-This provides the result that $\forall i, j$ in different communities, 
-$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
-\end{theorem}
-
-\begin{algorithm}[t]
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \KwData{Adjacency matrix $A$, number of communities $K$}
-  \KwResult{Community assignments $1, ..., K$}
-    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
-    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
-    $V$ using these eigenvectors as its columns.\;
-    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
-    Construct graph $G$ using $B$ as its similarity matrix.\;
-    Partition $G$ into $K$ disconnected subgraphs  
-    (e.g., using edge thresholding or spectral clustering).\;
-    Map each partition to the community labels $1, ..., K$.\;
-  \caption{Orthogonal Spectral Clustering.}
-\end{algorithm}
-
-Since the communities are subspaces in the latent space, we can also use
-existing methods for subspace clustering on the ASE. Of particular
-interest is Sparse Subspace Clustering (SSC), which is performed by
-solving an optimization problem for each observed point in a sample.
-Given \(X \in \mathbb{R}^{n \times d}\) with vectors
-\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
-\(c_i = \arg\min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
-\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
-collected into matrix
-\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
-construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
-perfectly on one of \(K\) subspaces, \(B\) is sparse such that
-\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
-Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
-and if the number of subgraphs is exactly \(K\), each subgraph maps onto
-a subspace.
-
-In practice, SSC is performed by solving the LASSO problems:
-
-\begin{equation} \label{eq:ssc}
-c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
-\end{equation}
-
-for some sparsity parameter \(\lambda > 0\). The \(c_i\) vectors are
-then collected into \(C\) and \(B\) as described before. If \(X\) is
-noisy such that each \(x_i\) does not lie exactly on one of \(K\)
-subspaces but near it, the choice of \(\lambda\) becomes important in
-guaranteeing the Subspace Detection Property (SDP)
-\cite{jmlr-v28-wang13}.
-
-\begin{definition}[Subspace Detection Property] 
-Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
-points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
-solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
-$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
-different subspaces, then $X$ obeys the Subspace Detection Property. 
-\end{definition}
-
-\begin{remark} 
-In practice, a noisy sample $X$ often does not obey SDP. 
-In such cases, $B$ is treated as an affinity matrix for a graph which 
-is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
-hand, if $X$ does obey the SDP, $B$ describes a graph 
-with at least $K$ disconnected subgraphs. Ideally, when SDP holds, 
-there are exactly $K$ subgraphs which map to each subspace, 
-but it could be the case that some of the subspaces are represented by 
-multiple disconnected subgraphs.
-\end{remark}
-
-Since every ASE of the PABM consists of subspaces and as
-\(n \to \infty\) each vector of the ASE approaches its subspace
-uniformly \cite{rubindelanchy2017statistical}, SSC is also able to
-perform community detection for the PABM. It has been shown by
-\citet{jmlr-v28-wang13} that if the points lie sufficiently close to
-their respective subspaces and the cosine of the angles between
-subspaces is sufficiently small, SDP will hold. We now state that the
-unscaled ASE of the PABM exhibits exactly these conditions for
-sufficiently large \(n\).
-
-\begin{theorem}
-\label{theorem5}
-Let $P_n$ describe the edge probability matrix of the PABM with 
-$n$ vertices, and let $A_n \sim \text{Bernoulli}(P_n)$.  Let $\hat{V}_n$ be the 
-matrix of eigenvectors of $A_n$ corresponding to the $K (K + 1) / 2$ most
-positive and $K (K - 1) / 2$ most negative eigenvalues. Then 
-$\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
-$\sqrt{n} \hat{V}_n$ obeys the subspace detection property with probability 1.
-\end{theorem}
-
-\hypertarget{non-spectral-approaches-to-community-detection}{%
-\subsection{Non-Spectral Approaches to Community
-Detection}\label{non-spectral-approaches-to-community-detection}}
-
-So far, we connected the PABM to the GRDPG and used this to motivate
-community detection methods based on the ASE of the adjacency matrix
-sampled from the PABM. In this section, we will explore non-spectral
-methods for community detection.
-
-Denoting \(\{\lambda_{ik}\}\) as the scalar popularity parameters for
-the PABM such that \(\lambda^{(kl)}\) is the vector of \(\lambda_{jl}\)
-terms for which vertex \(j\) is in community \(k\), we can write the
-likelihood for the PABM:
-
-\[p(A | z, \{\lambda_{ik}\}) = 
-\prod_{i < j} (\lambda_{i z_j} \lambda_{j z_i})^{A_{ij}} 
-(1 - \lambda_{i z_j} \lambda_{j z_i})^{1 - A_{ij}}\]
-
-\begin{equation}
-= \prod_{i < j} \prod_k^K \prod_l^K 
-(\lambda_{il} \lambda_{jk})^{A_{ij} z_{ik} z_{jl}} 
-(1 - \lambda_{il} \lambda_{jk})^{(1 - A_{ij}) z_{ik} z_{jl}}
-\end{equation}
-
-where \(z_{ik} = I(Z_i = k)\).
-
-The log-likelihood is
-
-\begin{equation}
-\log p = \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
-(A_{ij} \log \lambda_{il} \lambda_{jk} + 
-(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk}))
-\end{equation}
-
-\hypertarget{expectation-maximization}{%
-\subsubsection{Expectation
-Maximization}\label{expectation-maximization}}
-
-While direct maximization of the likelihood or log-likelihood is
-NP-complete, it is possible to apply an Expectation Maximization (EM)
-algorithm to this problem. To set this up, we write the complete data
-log likelihood as:
-
-\[Q = 
-\sum_{i < j} \sum_{k, l} z_{ik} z_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} +
-(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))
-+ \sum_i \sum_k z_{ik} \log \pi_k\]
-
-Where \(\pi_k\) are the \emph{a priori} community label probabilities.
-
-Then the expectation step involves solving for
-\(\gamma_{ik} = P(Z_i = k \mid \{\pi_l\}, \{\lambda_{jl}\})\), which is
-given by
-
-\[\log \gamma_{ik} \propto 
-\log \pi_k + 
-\sum_{j \neq i} \sum_l \pi_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} + 
-(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))\]
-
-For the maximization step, we can just set
-\(\pi_k = \frac{1}{n} \sum_i \gamma_{ik}\), but the \(\{\lambda_{ik}\}\)
-parameters cannot be maximized in closed form and require an
-optimization subroutine. Our proposed work involves finding an efficient
-optimization method for this subroutine.
-
-\hypertarget{bayesian-methods}{%
-\subsubsection{Bayesian Methods}\label{bayesian-methods}}
-
-The introduction of community label probabilities \(\pi_1, ..., \pi_K\)
-suggests a hierarchical model in which the community labels are
-generated by a prior distribution. We can apply the same principle to
-the popularity parameters \(\{\lambda_{ik}\}\) to make the model fully
-Bayesian. In this setup, we sample
-\(z_i \stackrel{\text{iid}}{\sim}\text{Categorical}(\pi_1, ..., \pi_K)\)
-and
-\(\lambda_{ik} \stackrel{\text{ind}}{\sim}\text{Beta}(a_{ik}, b_{ik})\)
-before sampling the observed data
-\(A_{ij} \stackrel{\text{ind}}{\sim}\lambda_{i z_j} \lambda_{j z_i}\).
-
-The full joint distribution for this model can be written as:
-
-\[\begin{split}
-\log p & = constant \\
-& + \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
-(A_{ij} \log \lambda_{il} \lambda_{jk} + 
-(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk})) \\
-& + \sum_k \sum_i z_{ik} \log \pi_k \\
-& + \sum_i \sum_k (a_{ik} - 1) \log \lambda_{ik} + (b_{ik} - 1) \log (1 - \lambda_{ik})
-\end{split}\]
-
-It is straightforward to obtain the conditional distribution for each
-\(z_i\) but not each \(\lambda_{ik}\), so a Gibbs sampler cannot be
-derived for this model. However, we can use a Metropolis-Hastings
-algorithm. In particular, by taking a first-order Taylor approximation,
-we can approximate the conditional distribution for each
-\(\lambda_{ik}\) as a Beta distribution. Our proposed work involves
-investigating this method by simulating this model and implementing the
-Bayes sampler.
-
-\hypertarget{generalizations-to-grdpg-based-community-detection}{%
-\section{Generalizations to GRDPG-Based Community
-Detection}\label{generalizations-to-grdpg-based-community-detection}}
-
-In the previous sections, we connected well-known and highly restricted
-generative graph models to the RDPG and GRDPG to show that highly
-structured latent configurations generate graphs consistent with these
-models: The latent space for the SBM consists of \(K\) point masses, the
-latent space for the DCBM consists of \(K\) rays emanating from the
-origin, and the latent space for the PABM consists of \(K\)
-\(K\)-dimensional subspaces. In the following sections, we explore
-additional structured latent configurations corresponding to community
-structure and develop methods for community detection based on the
-consistency of the ASE and the structural forms of the latent
-configurations.
-
-The general structure of interest can be described as follows: Suppose
-that in the latent space \(\mathcal{X} \subset \mathbb{R}^d\), sample
-\(X\) of \(n\) points lie on a union of \(K\) disjoint manifolds with
-each manifold corresponding to a community. If
-\(A \sim \text{RDPG}(X)\), we wish to recover the community labels (up
-to permutation) from \(A\).
-
-Equivalently, suppose that probability distribution \(F\) is described
-as follows:
+Let \((A, X) \sim \text{GRDPG}_{p, q}(F, n)\) such that:
 
@@
 \item
-  Define functions \(f_1, ..., f_K\) such that
-  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_k(t) \neq f_l(t)\)
-  \(\forall k, l \in [K]\) and \(t \in [0, 1]\).
+  Define functions \(\gamma_1, ..., \gamma_K\) such that each
+  \(\gamma_k : [0, 1]^r \mapsto \mathbb{R}^d\) and
+  \(\gamma_k(t) \neq \gamma_l(t)\) when \(k \neq l\).
 \item
   Sample labels
-  \(z_1, ..., z_n \stackrel{\text{iid}}{\sim}Categorical(\pi_1, ..., \pi_K)\).
+  \(Z_1, ..., Z_n \stackrel{\text{iid}}{\sim}\text{Categorical}(\pi_1, ..., \pi_K)\).
 \item
-  Sample \(t_1, ..., t_n \stackrel{\text{iid}}{\sim}D\) where \(D\) has
-  support \([0, 1]\).
+  Sample \(T_1, ..., T_n \stackrel{\text{iid}}{\sim}D\) with support
+  \([0, 1]^r\).
 \item
-  Set latent positions \(x_i = f_{z_i}(t_i)\) and
-  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
+  Set latent positions \(X_i = \gamma_{Z_i}(T_i)\) and
+  \(X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top\).
+\item
+  \(A \sim \text{BernoulliGraph}(X I_{p, q} X^\top)\)
 \end{enumerate}
 
-Then if \((A, X) \sim \text{RDPG}(F, n)\) and we observe \(A\), we wish
-to recover hidden labels \(z_1, ..., z_n\).
+The SBM, DCBM, and PABM are all special cases of this generative model.
+Based on the structure of the \(\gamma_k\) functions that correspond to
+each of the SBM, DCBM, and PABM, there are clear choices for clustering
+algorithms on the embeddings of each of these Block Models. Extensions
+of these results may consider which clustering algorithms result in
+consistent estimators given restrictions on each of the \(\gamma_k\),
+and how to approach community detection when \(\gamma_k\) are unknown
+altogether.
 
-\hypertarget{affine-segments}{%
-\subsection{Affine Segments}\label{affine-segments}}
-
-In this section, we will consider latent positions sampled uniformly
-from parallel unit length segments.
-
-\begin{example}
-Let $U_1, ..., U_{n_1}, V_1, ..., V_{n_2} \stackrel{\text{iid}}{\sim}Uniform(0, 1)$, $f_1(t) = (t, 0)$, and $f_2(t) = (t, a)$.
-$X_i = f_1(U_i)$ and $Y_j = f_2(V_j)$ $\forall i \in [n_1]$ and $j \in [n_2]$.
-If we observe $X_1, ..., X_{n_1}, Y_1, ..., Y_{n_2}$, what approach will allow us to group the observations by segment?
-\end{example}
-
-\begin{figure}[H]
-
-{\centering \includegraphics{proposal_files/figure-latex/unnamed-chunk-1-1}
-
-}
-
-\caption{Points sampled uniformly on two parallel unit length line segments}\label{fig:unnamed-chunk-1}
-\end{figure}
-
-We will motivate an approach by the following example.
-
-\begin{example}
-Let $U_1, ..., U_n \stackrel{\text{iid}}{\sim}Uniform(0, 1)$ with order statistics 
-$U_{(1)}, ..., U_{(n)}$. Then $\forall a \in (0, 1)$ and $\delta \in (0, 1)$, $\exists N = N(\delta, a) < \infty$ such that $\forall n \geq N$,
-
-\begin{equation}
-P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq 1 - \delta / 2
-\end{equation}
-
-Where $N(\delta, a)$ is monotone increasing with respect to $\delta$ and $a$. To prove this, we start with the fact that $U_{(i+1)} - U_{(i)} \sim Beta(1, n)$. Then
-
-\begin{equation}
-P(U_{(i+1)} - U_{(i)} \leq a) = 1 - (1 - a)^n
-\end{equation}
-
-and
-
-\begin{equation}
-\label{eq:unsolvable}
-P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq (P(U_{(i+1)} - U_{(i)} \leq a))^n = (1 - (1 - a)^n)^n
-\end{equation}
-
-This expression is monotone increasing $\forall n \geq N_1$ for some $N_1 < \infty$. 
-Setting $(1 - (1 - a)^{N_2})^{N_2} \geq 1 - \delta / 2$, we can solve for a finite ${N_2}$. Then $N = \max(N_1, N_2)$.
-\end{example}
-
-If we extend this example such that \(n_1\) points are sampled uniformly
-from the segment \(f_1(t) = (t, 0)\) and \(n_2\) points are sampled
-uniformly from the segment \(f_2(t) = (t, a)\) for
-\(t \in [0, \cos \frac{\pi}{2} a]\), then a sample of size
-\(\min(n_1, n_2) \geq N(\delta, a)\) is sufficient to satisfy:
-
-\begin{equation}
-P(\max_i X_{(i+1)} - X_{(i)} \leq a \leq \min_{i, j} \|X_i - Y_j\|) 
-\geq 1 - \delta / 2
-\end{equation}
-
-and similar for
-\(P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|)\), for
-\(X_i\) in the first segment and \(Y_j\) in the second segment and
-\(X_{(i)}\), \(Y_{(j)}\) are order statistics in the first coordinate.
-If each segment corresponds to a community, this leads to the following
-two results:
+The estimated timeline of completion is as follows:
 
@@
 \item
-  Single linkage clustering with \(K = 2\) will perform perfect
-  community detection with probability at least \(1 - \delta\).
+  Literature review and complete proofs of main theorems: December 2022
 \item
-  An \(\epsilon\)-neighborhood graph with \(\epsilon \in (0, a)\) will
-  consists of at least 2 disjoint subgraphs such that no subgraph
-  consists of members of two different communities (analogous to the
-  SDP), with probability at least \(1 - \delta\).
+  Simulations, real data analyses, R package: March 2022
+\item
+  Dissertation Completion: April 2022
 \end{enumerate}
 
-We can then further extend this to the case where points are drawn from
-unit segments with noise. It is straightforward to show that if the
-maximum noise is less than \(a / 3\), we can obtain a similar result as
-in the noiseless case. Thus if the distribution on the two segments is
-used to draw \((A, X) \sim \text{RDPG}(F, n)\), the same properties
-should hold for large \(n\).
-
-\hypertarget{manifolds}{%
-\subsection{Manifolds}\label{manifolds}}
-
-If instead of sampling uniformly from line segments of unit length, we
-sample uniformly from a 1 dimensional manifolds of unit length, the
-above property still holds. Let
-\(U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)\) and
-\(f : [0, 1] \mapsto \mathbb{R}^d\) be a smooth function such that
-\(\int_u^v \sqrt{\sum_k (df_k / dt)^2} dt = \|u - v\|\). Then
-\(U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
-\(P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
-If the shortest distance between the two manifolds defined by \(f_1\)
-and \(f_2\) with the same restriction is \(a\), then the same \(N\) as
-before is sufficient, although perhaps a more lenient lower bound can be
-derived based on the shape of \(f_k(\cdot)\).
-
-Our proposed work extends these results by the following.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Show that the ASE of a (G)RDPG generated by this type of latent
-  configuration produces the correct conditions as described above to
-  apply the same probability statements.
-\item
-  Extend these results by deriving similar probability statements for
-  nonuniform distributions.
-\item
-  Relax the condition for the minimum distance between manifolds by
-  investigating alternative clustering techniques on the latent space
-  (e.g., spectral clustering).
-\end{enumerate}
-
-\hypertarget{summary}{%
-\section{Summary}\label{summary}}
-
-For this research, we propose using the connection between Block Models
-and the (Generalized) Random Dot Product Graph to motivate community
-detection methods. Previous research shows that the Stochastic Block
-Model and Degree Corrected Block Model have very structured latent
-configurations in the RDPG framework, and these latent structures can be
-used to develop community detection algorithms that result in perfect
-community detection for large \(n\). In the same vein, we investigated
-the connection between the Popularity Adjusted Block Model and the GRDPG
-to motivate two community detection algorithms, the first of which is an
-application of an existing algorithm, Sparse Subspace Clustering, and
-the second is a new algorithm based on the particular latent form of the
-PABM, Orthogonal Spectral Clustering. Furthermore, we showed that for
-large \(n\), our application of the SSC algorithm to the PABM upholds
-the Subspace Detection Property for large yet finite \(n\), and the
-output of OSC results in perfect community detection almost surely as
-\(n \to \infty\). We will extend these theoretical results with
-empirical simulations and real data examples. Finally, we will implement
-these two community detection algorithms in an R package.
-
-We aim to extend these results by defining more general structures in
-the latent space to construct RDPG or GRDPG models and show that
-community detection is possible via Adjacency Spectral Embedding. So
-far, we showed that if in the latent space the communities can be
-represented as points sampled uniformly on curves or one-dimensional
-manifolds separated by a nonzero distance, we can apply two existing
-clustering techniques to the ASE and show that they will perform perfect
-community detection with high probability. We will generalize these
-results to nonuniform distributions on higher dimensional manifolds as
-well as investigate alternative clustering techniques for the latent
-space that may perform better for smaller \(n\).
-
-\hypertarget{estimated-timeline-of-completion}{%
-\section{Estimated Timeline of
-Completion}\label{estimated-timeline-of-completion}}
-
-Literature review: August 2021
-
-Complete proofs of main theorems: January 2022
-
-Simulations, real data analyses, R package: March 2022
-
-Dissertation completion: April 2022
-
-\renewcommand\refname{References}
   \bibliography{proposal.bib}

### 2021-06-14 proposal/prospectus.Rmd
minor fix
updated prospectus

@@
 Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision.
-One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels).
-Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs.
-Common examples of such clustering techniques include $K$-means clustering \cite{1056489} and Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}.
-In addition, these techniques assume that the vectors are sampled from a specific type of distribution (whether implicitly in the case of $K$-means clustering or explicitly in the case of GMM).
-In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure).
+Analyzing such data, however, can be a challenge because traditional statistical and machine learning techniques are often incompatible with graph data.
+For example, if we observe data sampled from a mixture of Gaussians, we can iteratively update the means and covariances by computing summary statistics for each mixture component and update community labels via Mahalanobis distance using the sample means and covariances \cite{doi:10.1198/016214502760047131}.
+A type of model that is analogous to mixture models for graph data is the Block Model, which is a family of generative models that have community structure.
+For such models, it is often impossible to compute analogous summary statistics (e.g., sample mean) for community detection or parameter estimation.
+In this research, we investigate the use of a family of spectral decomposition and embedding based approaches for community detection in block models (random graph models with inherent community structure).
 First, we demonstrate that under the Generalized Random Dot Product Graph \cite{rubindelanchy2017statistical} framework, all graphs generated by Block Models (generative graph models with inherent community structure) can be represented as collections of vectors in Euclidean space via the Adjacency Spectral Embedding \cite{athreya2017statistical, lyzinski2014}.
-Then, noting the particular structure or configuration that the vectors take, we select appropriate clustering techniques to apply on the vector representation of the graph.
+Then, noting the particular structure or configuration that the vectors take, we select appropriate clustering and parameter estimation techniques to apply on the vector representation of the graph.
 Finally we derive the asymptotic properties of these methods to show that our methods are consistent or achieve desirable properties with high probability.
@@
 The SBM, DCBM, and PABM are all special cases of this generative model.
-Based on the structure of the $\gamma_k$ functions that correspond to each of the SBM, DCBM, and PABM, there are clear choices for clustering algorithms on the embeddings of each of these Block Models.
+Based on the structure of the $\gamma_k$ functions that correspond to each of the SBM, DCBM, and PABM, there are natural choices for clustering algorithms on the embeddings of each of these Block Models, followed by straightforward parameter estimation techniques.
 Extensions of these results may consider which clustering algorithms result in consistent estimators given restrictions on each of the $\gamma_k$, and how to approach community detection when $\gamma_k$ are unknown altogether.
@@
 
-1. Literature review and complete proofs of main theorems: December 2022
+1. Literature review and proofs of main theorems: December 2022
 

### 2021-06-14 proposal/prospectus.tex
minor fix
updated prospectus

@@
 ]{article}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
@@
 \usepackage[margin=1in]{geometry}
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \usepackage{comment}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 \usepackage[]{natbib}
@@
 observations, are increasingly widespread in various fields ranging from
-sociology to computer vision. One common goal of analyzing graph data is
-community detection or graph clustering, in which the graph is
-partitioned into disconnected subgraphs in an unsupervised yet
-meaningful manner (e.g., by optimizing an objective function or
-recovering unobserved labels). Because traditional clustering techniques
-were developed for data that can be represented as vectors, they cannot
-be applied directly to graphs. Common examples of such clustering
-techniques include \(K\)-means clustering \cite{1056489} and Gaussian
-Mixture Models \cite{doi:10.1198/016214502760047131}. In addition, these
-techniques assume that the vectors are sampled from a specific type of
-distribution (whether implicitly in the case of \(K\)-means clustering
-or explicitly in the case of GMM). In this research, we investigate the
-use of a family of spectral decomposition based approaches for community
-detection in block models (random graph models with inherent community
-structure). First, we demonstrate that under the Generalized Random Dot
-Product Graph \cite{rubindelanchy2017statistical} framework, all graphs
-generated by Block Models (generative graph models with inherent
-community structure) can be represented as collections of vectors in
-Euclidean space via the Adjacency Spectral Embedding
+sociology to computer vision. Analyzing such data, however, can be a
+challenge because traditional statistical and machine learning
+techniques are often incompatible with graph data. For example, if we
+observe data sampled from a mixture of Gaussians, we can iteratively
+update the means and covariances by computing summary statistics for
+each mixture component and update community labels via Mahalanobis
+distance using the sample means and covariances
+\cite{doi:10.1198/016214502760047131}. A type of model that is analogous
+to mixture models for graph data is the Block Model, which is a family
+of generative models that have community structure. For such models, it
+is often impossible to compute analogous summary statistics (e.g.,
+sample mean) for community detection or parameter estimation. In this
+research, we investigate the use of a family of spectral decomposition
+and embedding based approaches for community detection in block models
+(random graph models with inherent community structure). First, we
+demonstrate that under the Generalized Random Dot Product Graph
+\cite{rubindelanchy2017statistical} framework, all graphs generated by
+Block Models (generative graph models with inherent community structure)
+can be represented as collections of vectors in Euclidean space via the
+Adjacency Spectral Embedding
 \cite{athreya2017statistical, lyzinski2014}. Then, noting the particular
 structure or configuration that the vectors take, we select appropriate
-clustering techniques to apply on the vector representation of the
-graph. Finally we derive the asymptotic properties of these methods to
-show that our methods are consistent or achieve desirable properties
-with high probability. To illustrate this approach, we primarily focus
-on the Popularity Adjusted Block Model
+clustering and parameter estimation techniques to apply on the vector
+representation of the graph. Finally we derive the asymptotic properties
+of these methods to show that our methods are consistent or achieve
+desirable properties with high probability. To illustrate this approach,
+we primarily focus on the Popularity Adjusted Block Model
 \cite{307cbeb9b1be48299388437423d94bf1}.
@@
 Based on the structure of the \(\gamma_k\) functions that correspond to
-each of the SBM, DCBM, and PABM, there are clear choices for clustering
-algorithms on the embeddings of each of these Block Models. Extensions
+each of the SBM, DCBM, and PABM, there are natural choices for
+clustering algorithms on the embeddings of each of these Block Models,
+followed by straightforward parameter estimation techniques. Extensions
 of these results may consider which clustering algorithms result in
@@
 \item
-  Literature review and complete proofs of main theorems: December 2022
+  Literature review and proofs of main theorems: December 2022
 \item

### 2021-06-14 proposal/prospectus.Rmd
updated prospectus

@@
 Analyzing such data, however, can be a challenge because traditional statistical and machine learning techniques are often incompatible with graph data.
-For example, if we observe data sampled from a mixture of Gaussians, we can iteratively update the means and covariances by computing summary statistics for each mixture component and update community labels via Mahalanobis distance using the sample means and covariances \cite{doi:10.1198/016214502760047131}.
+For example, if we observe data sampled from a mixture of Gaussians, we can iteratively update the means and covariances by computing summary statistics for each mixture component and update community labels via Mahalanobis distances using the sample means and covariances \cite{doi:10.1198/016214502760047131}.
 A type of model that is analogous to mixture models for graph data is the Block Model, which is a family of generative models that have community structure.
 For such models, it is often impossible to compute analogous summary statistics (e.g., sample mean) for community detection or parameter estimation.
-In this research, we investigate the use of a family of spectral decomposition and embedding based approaches for community detection in block models (random graph models with inherent community structure).
-First, we demonstrate that under the Generalized Random Dot Product Graph \cite{rubindelanchy2017statistical} framework, all graphs generated by Block Models (generative graph models with inherent community structure) can be represented as collections of vectors in Euclidean space via the Adjacency Spectral Embedding \cite{athreya2017statistical, lyzinski2014}.
+In this research, we investigate the use of a family of spectral decomposition and embedding based approaches for community detection and parameter estimation in block models (random graph models with inherent community structure).
+First, we demonstrate that under the Generalized Random Dot Product Graph \cite{rubindelanchy2017statistical} framework, all graphs generated by Block Models can be represented as collections of vectors in Euclidean space via the Adjacency Spectral Embedding \cite{athreya2017statistical, lyzinski2014}.
 Then, noting the particular structure or configuration that the vectors take, we select appropriate clustering and parameter estimation techniques to apply on the vector representation of the graph.
 Finally we derive the asymptotic properties of these methods to show that our methods are consistent or achieve desirable properties with high probability.
-To illustrate this approach, we primarily focus on the Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1}.
+To illustrate this approach, we primarily focus on a type of Block Model called the Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1}.
 
-Our work is essentially about connecting two well-known families of generative graph models, the Block Model and the (Generalized) Random Dot Product Graph \cite{rubindelanchy2017statistical} in order to perform community detection.
-It is straightforward to show that all Block Models are various cases of the GRDPG.
-This fact has been leveraged to develop community detection methods \cite{athreya2017statistical, lyzinski2014, rubindelanchy2017statistical} for the Stochastic Block Model \cite{doi:10.1080/0022250X.1971.9989788} and Degree Corrected Block Model \cite{Karrer_2011}.
+Our work is essentially about connecting two well-known families of generative graph models, the Block Model and the (Generalized) Random Dot Product Graph \cite{rubindelanchy2017statistical} in order to perform statistical inference.
+It is straightforward to show that all Block Models are special cases of the GRDPG.
+This fact has been leveraged to develop community detection and parameter estimation methods \cite{athreya2017statistical, lyzinski2014, rubindelanchy2017statistical} for the Stochastic Block Model \cite{doi:10.1080/0022250X.1971.9989788} and Degree Corrected Block Model \cite{Karrer_2011}.
 Our work begins by extending this to the Popularity Adjusted Block Model.
 We show that under the GRDPG framework, the latent configuration that generates the PABM consists of orthogonal subspaces.
+The Adjacency Spectral Embedding allows us to recover this latent configuration from an observed adjacency matrix.
 Then we show that two community detection algorithms arise naturally from this latent configuration: Orthogonal Spectral Clustering, which is an algorithm of our own design, and an existing algorithm, Sparse Subspace Clustering \cite{5206547}.
+Parameter estimation via spectral decomposition can be performed in a similar vein.
 We then extend this work to more general configurations in the latent space that imply community structure.
-The overall goal of our research is to develop a general framework or approach to community detection for a wide range of Block Models.
-
-\newpage
+The overall goal of our research is to develop a general framework or approach to statistical inference for a wide range of generative graph models with community structure by investigating their latent structure under the GRDPG framework.
 

### 2021-06-14 proposal/prospectus.tex
updated prospectus

@@
 each mixture component and update community labels via Mahalanobis
-distance using the sample means and covariances
+distances using the sample means and covariances
 \cite{doi:10.1198/016214502760047131}. A type of model that is analogous
@@
 research, we investigate the use of a family of spectral decomposition
-and embedding based approaches for community detection in block models
-(random graph models with inherent community structure). First, we
-demonstrate that under the Generalized Random Dot Product Graph
-\cite{rubindelanchy2017statistical} framework, all graphs generated by
-Block Models (generative graph models with inherent community structure)
-can be represented as collections of vectors in Euclidean space via the
-Adjacency Spectral Embedding
+and embedding based approaches for community detection and parameter
+estimation in block models (random graph models with inherent community
+structure). First, we demonstrate that under the Generalized Random Dot
+Product Graph \cite{rubindelanchy2017statistical} framework, all graphs
+generated by Block Models can be represented as collections of vectors
+in Euclidean space via the Adjacency Spectral Embedding
 \cite{athreya2017statistical, lyzinski2014}. Then, noting the particular
@@
 desirable properties with high probability. To illustrate this approach,
-we primarily focus on the Popularity Adjusted Block Model
-\cite{307cbeb9b1be48299388437423d94bf1}.
+we primarily focus on a type of Block Model called the Popularity
+Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1}.
 
@@
 Dot Product Graph \cite{rubindelanchy2017statistical} in order to
-perform community detection. It is straightforward to show that all
-Block Models are various cases of the GRDPG. This fact has been
-leveraged to develop community detection methods
+perform statistical inference. It is straightforward to show that all
+Block Models are special cases of the GRDPG. This fact has been
+leveraged to develop community detection and parameter estimation
+methods
 \cite{athreya2017statistical, lyzinski2014, rubindelanchy2017statistical}
@@
 under the GRDPG framework, the latent configuration that generates the
-PABM consists of orthogonal subspaces. Then we show that two community
-detection algorithms arise naturally from this latent configuration:
-Orthogonal Spectral Clustering, which is an algorithm of our own design,
-and an existing algorithm, Sparse Subspace Clustering \cite{5206547}. We
-then extend this work to more general configurations in the latent space
-that imply community structure. The overall goal of our research is to
-develop a general framework or approach to community detection for a
-wide range of Block Models.
-
-\newpage
+PABM consists of orthogonal subspaces. The Adjacency Spectral Embedding
+allows us to recover this latent configuration from an observed
+adjacency matrix. Then we show that two community detection algorithms
+arise naturally from this latent configuration: Orthogonal Spectral
+Clustering, which is an algorithm of our own design, and an existing
+algorithm, Sparse Subspace Clustering \cite{5206547}. Parameter
+estimation via spectral decomposition can be performed in a similar
+vein. We then extend this work to more general configurations in the
+latent space that imply community structure. The overall goal of our
+research is to develop a general framework or approach to statistical
+inference for a wide range of generative graph models with community
+structure by investigating their latent structure under the GRDPG
+framework.
 

### 2021-06-14 proposal/prospectus.Rmd
updated prospectus

@@
 
+The goal of statistical inference on data sampled from this model would be to estimate each $Z_i$ (community detection) and properties derived from each $\gamma_{Z_i}(T_i)$ (parameter estimation).
 The SBM, DCBM, and PABM are all special cases of this generative model.
 Based on the structure of the $\gamma_k$ functions that correspond to each of the SBM, DCBM, and PABM, there are natural choices for clustering algorithms on the embeddings of each of these Block Models, followed by straightforward parameter estimation techniques.
-Extensions of these results may consider which clustering algorithms result in consistent estimators given restrictions on each of the $\gamma_k$, and how to approach community detection when $\gamma_k$ are unknown altogether.
+For example, in the case of the SBM, we can use GMM on the ASE to estimate community labels, then use the sample means of each community to estimate the edge probabilities.
+Extensions of these results may consider which clustering algorithms result in consistent estimators for various types of $\gamma_k$, and how to approach this problem when $\gamma_k$ are unknown altogether.
 

### 2021-06-14 proposal/prospectus.tex
updated prospectus

@@
 
-The SBM, DCBM, and PABM are all special cases of this generative model.
-Based on the structure of the \(\gamma_k\) functions that correspond to
-each of the SBM, DCBM, and PABM, there are natural choices for
-clustering algorithms on the embeddings of each of these Block Models,
-followed by straightforward parameter estimation techniques. Extensions
-of these results may consider which clustering algorithms result in
-consistent estimators given restrictions on each of the \(\gamma_k\),
-and how to approach community detection when \(\gamma_k\) are unknown
-altogether.
+The goal of statistical inference on data sampled from this model would
+be to estimate each \(Z_i\) (community detection) and properties derived
+from each \(\gamma_{Z_i}(T_i)\) (parameter estimation). The SBM, DCBM,
+and PABM are all special cases of this generative model. Based on the
+structure of the \(\gamma_k\) functions that correspond to each of the
+SBM, DCBM, and PABM, there are natural choices for clustering algorithms
+on the embeddings of each of these Block Models, followed by
+straightforward parameter estimation techniques. For example, in the
+case of the SBM, we can use GMM on the ASE to estimate community labels,
+then use the sample means of each community to estimate the edge
+probabilities. Extensions of these results may consider which clustering
+algorithms result in consistent estimators for various types of
+\(\gamma_k\), and how to approach this problem when \(\gamma_k\) are
+unknown altogether.
 

### 2021-08-18 scratch/distributions-on-1d-manifolds.Rmd
started typing up stuff

---
title: |
  | Clustering of Distributions on 1-Dimensional Manifolds
output:
  pdf_document:
    keep_tex: true
    citation_package: natbib
fontsize: 11pt
# geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
bibliography: proposal.bib
---

## Setup

Suppose we have two manifolds $\mathcal{M}_1$ and $\mathcal{M}_2$ $\in \mathbb{R}^d$, each of length 1, defined by $f_1(t)$ and $f_2(t)$ respectively ($f_i : [0, 1] \mapsto \mathbb{R}^d$). Define $\delta$ as the minimum distance between the two manifolds, i.e., $\delta = \min_t \|f_1(t) - f_2(t)\|$, and let $\delta > 0$. For now, restrict each $f_i$ such that the distance along the manifold between $f_i(t)$ and $f_i(s)$ is equal to the difference between $t$ and $s$ (this also implies that each manifold is of length 1). 
We sample $T_1, ..., T_n \stackrel{iid}{\sim} F$ for continuous $F$ with support $[0, 1]$ and use $f_1$ to map the first $n_1$ points to $\mathcal{M}_1$ and $f_2$ to map the remaining $n_2 = n - n_1$ points to $\mathcal{M}_2$. Let $X_i = f_1(T_i)$ and $Y_j = f_2(T_j)$.
Without loss of generality, assume $n_1 \leq n_2$.

## Theory

Let $D_i = X_{(i)} - X_{(i-1)}$. Then if $\max_i D_i < \delta$, we have sufficient separation of points in $\mathcal{M}_1$.

### Uniform case

It can be shown that if $X_i \stackrel{iid}{\sim} Uniform(0, 1)$, then $D_i \sim Beta(1, n)$. Therefore, $P(\max_i D_i < \delta) \geq (P(D_i < \delta))^n = (1 - (1 - \delta)^n)^n$, which is a decreasing function for sufficiently large $n$. This gives us the result $\max_i D_i \stackrel{a.s.}{\to} 0$.

### Beta case

### General case

If $F$ is absolutely continuous, then it can be shown that

$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} (i-1) \binom{n}{i-1} f(x) (F(x))^{i-1} (1 - F(x + \delta))^{n-i} dx$$

Under certain conditions (namely $F(x + \delta) \geq F(x) + \delta$), we can make the substitution $u = F(x)$ and have the result

$$P(D_i \leq \delta) \leq 1- \int_0^{1 - \delta} (i-1) \binom{n}{i-1} u^{i-1} (1 - u - \delta)^{n-i} du$$

which gives the same result as in the uniform case.

## Computational Results

TBD

### 2021-08-18 scratch/distributions-on-1d-manifolds.tex
started typing up stuff

--- 2021-05-26 proposal/proposal.tex
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{setspace}
@@
 
+\title{Clustering of Distributions on 1-Dimensional Manifolds}
 \author{}
@@
 \begin{document}
+\maketitle
 
-\pagenumbering{gobble}
+\hypertarget{setup}{%
+\subsection{Setup}\label{setup}}
 
-%\begin{titlepage}
-\begin{center}
-\LARGE{\textbf{Community Detection Methods for the Generalized Random Dot Product Graph Model}}\\
-\vspace*{2\baselineskip}
-\normalsize{A dissertation proposal submitted in partial satisfaction of the requirements for the degree of \\}
-Doctor of Philosophy \\
-in \\
-Statistical Science \\
-\vspace*{2\baselineskip}
-\Large{John Koo}\\
-\vspace*{3\baselineskip}
-\Large{\textbf{Research Committee Members}}\\
-Dr. Michael Trosset \\
-Dr. Minh Tang \\
-Dr. Julia Fukuyama \\
-Dr. Roni Khardon \\
-Dr. Fangzheng Xie \\
-\vspace*{3\baselineskip}
-June 2, 2021 \\
-\vspace*{1\baselineskip}
-Department of Statistics \\
-Indiana University \\
-Bloomington, Indiana \\
-\end{center}
-% \end{titlepage}
+Suppose we have two manifolds \(\mathcal{M}_1\) and \(\mathcal{M}_2\)
+\(\in \mathbb{R}^d\), each of length 1, defined by \(f_1(t)\) and
+\(f_2(t)\) respectively (\(f_i : [0, 1] \mapsto \mathbb{R}^d\)). Define
+\(\delta\) as the minimum distance between the two manifolds, i.e.,
+\(\delta = \min_t \|f_1(t) - f_2(t)\|\), and let \(\delta > 0\). For
+now, restrict each \(f_i\) such that the distance along the manifold
+between \(f_i(t)\) and \(f_i(s)\) is equal to the difference between
+\(t\) and \(s\) (this also implies that each manifold is of length 1).
+We sample \(T_1, ..., T_n \stackrel{iid}{\sim} F\) for continuous \(F\)
+with support \([0, 1]\) and use \(f_1\) to map the first \(n_1\) points
+to \(\mathcal{M}_1\) and \(f_2\) to map the remaining \(n_2 = n - n_1\)
+points to \(\mathcal{M}_2\). Let \(X_i = f_1(T_i)\) and
+\(Y_j = f_2(T_j)\). Without loss of generality, assume \(n_1 \leq n_2\).
 
-\hypersetup{linkcolor = black}
-\newpage
-\pagenumbering{roman}
-\tableofcontents
-\addcontentsline{toc}{section}{\contentsname}
+\hypertarget{theory}{%
+\subsection{Theory}\label{theory}}
 
-\newpage
+Let \(D_i = X_{(i)} - X_{(i-1)}\). Then if \(\max_i D_i < \delta\), we
+have sufficient separation of points in \(\mathcal{M}_1\).
 
-\begin{center}
-\LARGE{Abstract}
-\end{center}
+\hypertarget{uniform-case}{%
+\subsubsection{Uniform case}\label{uniform-case}}
 
-\vspace*{2\baselineskip}
+It can be shown that if \(X_i \stackrel{iid}{\sim} Uniform(0, 1)\), then
+\(D_i \sim Beta(1, n)\). Therefore,
+\(P(\max_i D_i < \delta) \geq (P(D_i < \delta))^n = (1 - (1 - \delta)^n)^n\),
+which is a decreasing function for sufficiently large \(n\). This gives
+us the result \(\max_i D_i \stackrel{a.s.}{\to} 0\).
 
-\normalsize
+\hypertarget{beta-case}{%
+\subsubsection{Beta case}\label{beta-case}}
 
-Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision.
-One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels).
-Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs.
-In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models, first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then by proposing clustering methods for these feature vector representations and deriving the asymptotic properties of these methods.
+\hypertarget{general-case}{%
+\subsubsection{General case}\label{general-case}}
 
-\newpage
-\pagenumbering{arabic}
-\hypersetup{linkcolor = blue}
+If \(F\) is absolutely continuous, then it can be shown that
 
-\newcommand{\diag}{\text{diag}}
-\newcommand{\tr}{\text{Tr}}
-\newcommand{\blockdiag}{\text{blockdiag}}
-\newcommand{\indep}{\stackrel{\text{ind}}{\sim}}
-\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
-\newcommand{\Bernoulli}{\text{Bernoulli}}
-\newcommand{\Betadist}{\text{Beta}}
-\newcommand{\BG}{\text{BernoulliGraph}}
-\newcommand{\Cat}{\text{Categorical}}
-\newtheorem{definition}{Definition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-\theoremstyle{example}
-\newtheorem*{example}{Example}
+\[P(D_i \leq \delta) = 1 - \int_0^{1-\delta} (i-1) \binom{n}{i-1} f(x) (F(x))^{i-1} (1 - F(x + \delta))^{n-i} dx\]
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+Under certain conditions (namely \(F(x + \delta) \geq F(x) + \delta\)),
+we can make the substitution \(u = F(x)\) and have the result
 
-\hypertarget{research-goal}{%
-\subsection{Research Goal}\label{research-goal}}
+\[P(D_i \leq \delta) \leq 1- \int_0^{1 - \delta} (i-1) \binom{n}{i-1} u^{i-1} (1 - u - \delta)^{n-i} du\]
 
-Graph and network data have become increasingly widespread in various
-fields including sociology, neuroscience, biostatistics, and computer
-science. This has resulted in various challenges for researchers who
-rely on traditional statistical and machine learning methods, many of
-which are incompatible with graph data and instead require the data to
-exist as feature vectors in Euclidean space. This includes graph
-clustering and community detection, which are often the goal of data
-analysis on graphs and networks. Common clustering methods typically
-involve calculating some central or representative point for each
-cluster around which the data belonging to that cluster lie (e.g.,
-Lloyd's algorithm for \(K\)-means clustering \cite{1056489}, Gaussian
-Mixture Models \cite{doi:10.1198/016214502760047131}). Because these
-methods involve computing summary statistics within each cluster, such
-as the sample average, they cannot be applied directly to graphs,
-necessitating methods for transforming the graph data into feature
-vectors.
+which gives the same result as in the uniform case.
 
-One family of methods for unifying graph community detection with
-traditional clustering techniques is Spectral Clustering
-\cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the
-graph into Euclidean space, followed by applying a popular clustering
-algorithm such as \(K\)-means clustering. The Random Dot Product Graph
-(RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot
-Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
-this further by explicitly constructing generative models for graphs
-involving latent vectors in Euclidean space. In such a model, each
-vertex has a corresponding vector in latent space, and the edge
-probabilities between pairs of vertices is computed by operations on
-corresponding pairs of latent vectors. A community detection algorithm
-motivated by this may involve learning the latent positions given an
-observed graph and then learning the community labels given the latent
-positions.
+\hypertarget{computational-results}{%
+\subsection{Computational Results}\label{computational-results}}
 
-The aim of our research is to develop consistent community detection
-techniques under the RDPG and GRDPG frameworks. First, we explore
-existing generative graph models with underlying community structures
-that can be inferred by connecting these models to the RDPG or GRDPG. In
-particular, we use the connection between the Popularity Adjusted Block
-Model \cite{307cbeb9b1be48299388437423d94bf1} and the GRDPG to motivate
-two community detection methods. Then we explore other latent structures
-or mixture distributions in the latent space that induce graphs for
-which consistent community detection is possible.
+TBD
 
-\hypertarget{notation}{%
-\subsection{Notation}\label{notation}}
-
-Let \(G = (V, E)\) be an undirected, unweighted graph with no self-loops
-and \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
-adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
-edge between vertices \(i\) and \(j\) and \(A_{ij} = 0\) otherwise.
-Because \(G\) is symmetric and contains no self-loops,
-\(A_{ij} = A_{ji}\) and \(A_{ii} = 0\) for \(i, j \in [n]\). We further
-restrict our analyses to independent Bernoulli graphs. Let
-\(P \in [0, 1]^{n \times n}\) be a symmetric matrix of edge
-probabilities. Graph \(G\) is sampled from \(P\) by drawing
-\(A_{ij} \stackrel{\text{ind}}{\sim}\text{Bernoulli}(P_{ij})\) for each
-\(1 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)). We
-denote \(A \sim \text{BernoulliGraph}(P)\) as graph with adjacency
-matrix \(A\) sampled from edge probability matrix \(P\) in this manner.
-If each vertex has a hidden label in \([K]\), they are denoted as
-\(z_1, ..., z_n\). Finally, we denote
-\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
-as the sample \(x_1, ..., x_n \in \mathbb{R}^d\).
-
-\hypertarget{preliminaries}{%
-\section{Preliminaries}\label{preliminaries}}
-
-\hypertarget{block-models}{%
-\subsection{Block Models}\label{block-models}}
-
-Generative models for independent Bernoulli graphs involve defining the
-edge probability matrix \(P\) whose \(ij\)\textsuperscript{th} entry is
-the probability of an edge between vertices \(i\) and \(j\) for each
-\(i, j \in [n]\). In order to motivate community detection methods, we
-restrict the generative model such that for each pair of vertices
-\((i, j)\), the probability of an edge between the vertices \(P_{ij}\)
-depends in some way on their labels \(z_i\) and \(z_j\).
-
-One such model is the Stochastic Block Model (SBM)
-\cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
-each vertex belonging to one community, the SBM restricts \(P\) to
-\(K (K + 1) / 2\) unique entries such that \(P_{ij} = p_{z_i z_j}\)
-where \(p_{kl} = p_{lk}\) is the probability of an edge between each
-vertex in community \(k\) to each vertex in community \(l\). The
-homogeneous SBM further restricts \(P\) to two unique entries:
-\(P_{ij} = p\) if \(z_i = z_j\) and \(P_{ij} = q\) otherwise. Several
-generalizations of the SBM have been introduced since, including the
-Degree Corrected Block Model (DCBM) \cite{Karrer_2011} and the
-Popularity Adjusted Block Model (PABM)
-\cite{307cbeb9b1be48299388437423d94bf1}. Like the SBM, these models
-involve imposing some restriction on \(P\) based on community labels
-(and possibly other factors).
-
-\hypertarget{generalized-random-dot-product-graphs}{%
-\subsection{(Generalized) Random Dot Product
-Graphs}\label{generalized-random-dot-product-graphs}}
-
-Like the SBM, DCBM, and PABM, the RDPG and GRDPG are generative models
-for graphs involving an edge probability matrix \(P\). The RDPG starts
-with points in latent space \(X \in \mathcal{X} \subset \mathbb{R}^d\)
-for \(\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}\).
-\(P\) is then constructed as \(P = X X^\top\), and
-\(A \sim \text{BernoulliGraph}(P)\). We provide a more formal definition
-of the RDPG and GRDPG as follows:
-
-\begin{definition}[(Generalized) Random Dot Product Graph]
-Let $X \in \mathbb{R}^{n \times d}$ be a collection of $n$ points in $\mathcal{X} \subset \mathbb{R}^d$ where $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top y \in [0, 1]\}$.
-$G = (V, E)$ is a Random Dot Product Graph if its adjacency matrix is drawn as $A \sim \text{BernoulliGraph}(X X^\top)$.
-
-If on the other hand, $\mathcal{X} = \{x, y \in \mathbb{R}^{p+q} : x^\top I_{p, q} y \in [0, 1]\}$ and $A \sim \text{BernoulliGraph}(X I_{p, q} X^\top)$ for $I_{p, q} = \text{blockdiag}(I_p, -I_q)$ and $p + q = d$, then $A$ is the adjacency matrix of a Generalized Random Dot Product Graph.
-These are denoted by $A \sim \text{RDPG}(X)$ and $A \sim \text{GRDPG}_{p, q}(X)$ respectively.
-
-In addition, let $F$ be a probability distribution with support $\mathcal{X}$, and $X_1, ..., X_n \stackrel{iid}{\sim} F$ with $X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}^\top$.
-If $A$ is drawn from $X$ as before, then $(A, X) \sim \text{RDPG}(F, n)$ or $(A, X) \sim \text{GRDPG}_{p, q}(F, n)$.
-\end{definition}
-
-The structure of the RDPG and GRDPG provides a straightforward method
-for recovery of the latent positions via spectral decomposition.
-
-\begin{definition}[Adjacency Spectral Embedding]
-Let $A \sim \text{RDPG}(X)$ for $X \in \mathcal{X} \subset \mathbb{R}^{n \times d}$.
-Let $A \approx V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $d$ largest eigenvalues and their corresponding eigenvectors.
-Then the rows of $V \Lambda^{1/2}$ are the scaled Adjacency Spectral Embedding (ASE) of $A$, and the rows of $V$ are the unscaled ASE of $A$.
-
-If $A \sim \text{GRDPG}_{p, q}(X)$, then let $A = V \Lambda V^\top$ be the approximate spectral decomposition of $A$ corresponding to the $p$ most positive and $q$ most negative eigenvalues of $A$ and their corresponding eigenvectors.
-Then the rows of $V |\Lambda|^{1/2}$ and $V$ are the scaled and unscaled ASE of $A$ respectively.
-\end{definition}
-
-\citet{athreya2017statistical} showed that under mild conditions, if
-\((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
-ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - W_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-Similarly, \citet{rubindelanchy2017statistical} showed that for
-\((A_n, X_n) \sim \text{GRDPG}_{p, q}(F, n)\),
-
-\begin{equation}
-\max_i \|(\hat{X}_n)_i - Q_n (X_n)_i \| \stackrel{a.s.}{\to} 0
-\end{equation}
-
-where \(Q_n\) is a sequence of matrices in \(\mathbb{O}(p, q)\), the
-indefinite orthogonal group of order \(p, q\).
-
-It is straightforward to show that all Bernoulli graphs with positive
-semidefinite \(P\) are special cases of the RDPG, which is a special
-case of the GRDPG, and all graphs generated by \(P\) regardless of
-positive semidefiniteness are special cases of the GRDPG. This includes
-the SBM, DCBM, and PABM. In the following example, we show that under
-the RDPG framework, the latent configuration that induces the SBM takes
-on a very particular form.
-
-\begin{example}[Connecting the SBM to the RDPG]
-Let $G = (V, E)$ with adjacency matrix $A$ be sampled from the SBM with two communities for which the within-edge probabilities are $p$ and $q$ for communities 1 and 2 respectively, and the between-community edge probability is $r$, with $p q > r^2$.
-Let community 1 have $n_1$ vertices, community 2 have $n_2$ vertices, and $n_1 + n_2 = n$.
-Without loss of generality, organize $P$ and $A$ such that the $kl^{th}$ block represents edges between communities $k$ and $l$.
-Then $P = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} & P^{(22)} \end{bmatrix}$ where each block is a constant value, e.g., $P^{(11)}_{ij} = p$.
-Then one RDPG representation of this SBM is:
-
-$$X = \begin{bmatrix} 
-\sqrt{p} & 0 \\
-\vdots & \vdots \\
-\sqrt{p} & 0 \\
-\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\ 
-\vdots & \vdots \\
-\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
-\end{bmatrix}
-\in \mathbb{R}^{n \times 2}$$
-
-where the first $n_1$ rows are $\begin{bmatrix} \sqrt{p} & 0 \end{bmatrix}$ and the next $n_2$ rows are $\begin{bmatrix} \sqrt{r^2 / p} & \sqrt{q - r^2 / p} \end{bmatrix}$.
-This is shown by reconstructing $P$ as $P = X X^\top$.
-\end{example}
-
-The ASE of the assortative SBM consists of points in \(\mathbb{R}^K\)
-that lie near one of \(K\) centers, depending on the community label,
-leading to ASE followed by \(K\)-means clustering (or similar, e.g.,
-GMM) as a consistent community detection algorithm \cite{lyzinski2014}.
-A similar result can be shown for the DCBM and PABM but with different
-structures in the ASE.
-
-\hypertarget{manifold-learning}{%
-\subsection{Manifold Learning}\label{manifold-learning}}
-
-\citet{trosset2020learning} showed that the ASE of a RDPG can be used to
-recover one-dimensional manifolds. Suppose
-\(f : [0, 1] \mapsto \mathcal{X}\) such that \(f\) is smooth and
-\(\mathcal{X}\) represents a curve or one-dimensional manifold in
-\(\mathbb{R}^d\). If \(t_1, ..., t_n \stackrel{\text{iid}}{\sim}F\) such
-that \(F\) has support \([0, 1]\), the latent positions are
-\(x_i = f(t_i)\) with \(y_i\) is its corresponding point in the scaled
-ASE, and \(d_{\epsilon}(\cdot, \cdot)\) is the shortest path distance of
-an \(\epsilon\)-neighborhood graph. Under certain mild conditions, the
-shortest path distances of the \(\epsilon\)-neighborhood graph of the
-ASE approaches the arc lengths along \(f\):
-
-\begin{equation}
-d_{\epsilon}(y_i, y_j) \stackrel{p}{\to} \int_{t_i}^{t_j} \sqrt{\sum_r^d \Big( \frac{d f_r}{d t} \Big)^2} dt
-\end{equation}
-
-\citet{athreya2020estimation} extended this further by generating a RDPG
-from a mixture of distributions on a curve. In their example, points
-were sampled from a mixture of two Beta distributions on the
-Hardy-Weinberg curve to construct the latent positions of a RDPG, with
-the goal of recovering the hidden mixture distribution from an observed
-graph.
-
-\hypertarget{the-popularity-adjusted-block-model}{%
-\section{The Popularity Adjusted Block
-Model}\label{the-popularity-adjusted-block-model}}
-
-As discussed in \S 2.2, all Bernoulli graph models can be expressed as a
-RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
-with a very specific structure in the latent space. A similar result has
-been shown for the DCBM \cite{lyzinski2014}. We can also extend this to
-the PABM. In \S 3.1, we show that the PABM is equivalent to the GRDPG
-model for which the communities correspond to subspaces in the latent
-configuration. We further show that while the latent configuration is
-not unique, there exists one in particular for which the subspaces are
-orthogonal. This leads to two straightforward methods for community
-detection, one involving computing inner products in the ASE and another
-which is a slight modification of an existing algorithm for subspace
-clustering. Finally, we will briefly explore alternative,
-non-spectral-based methods for community detection in the ASE involving
-Expectation Maximization, MCMC sampling, and Mean Field Variational
-Inference.
-
-\hypertarget{connecting-the-pabm-to-the-grdpg}{%
-\subsection{Connecting the PABM to the
-GRDPG}\label{connecting-the-pabm-to-the-grdpg}}
-
-We first define the PABM using the construction provided by
-\citet{noroozi2019estimation}.
-
-\begin{definition}[Popularity Adjusted Block Model]
-\label{pabm}
-Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
-set of $n$ 
-vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
-columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
-$P^{(kl)}$ describes the edge probabilities between vertices in communities 
-$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
-Let graph $G = (V, E)$ be an undirected, unweighted graph such 
-that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
-realization of $\text{Bernoulli}(P)$, i.e., 
-$A_{ij} \stackrel{\text{ind}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ 
-($A_{ij} = A_{ji}$ and $A_{ii} = 0$).
-
-If each block $P^{(kl)}$ can be written as the outer product of two vectors:
-
-\begin{equation} \label{eq:pabm}
-  P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top
-\end{equation}
-
-for a set of $K^2$ fixed vectors $\{\lambda^{(st)}\}_{s, t = 1}^K$ where each 
-$\lambda^{(st)}$ is a column vector 
-of dimension $n_s$, then graph $G$ and its corresponding adjacency matrix $A$ 
-is a realization of a popularity adjusted block model with parameters 
-$\{\lambda^{(st)}\}_{s, t = 1}^K$. 
-\end{definition}
-
-We will use the notation \(A \sim \text{PABM}(\{\lambda^{(kl)}\}_K)\) to
-denote a random adjacency matrix \(A\) drawn from a PABM with parameters
-\(\lambda^{(kl)}\) consisting of \(K\) underlying communities.
-
-It is straightforward to show that the PABM, as well as all graphs such
-that the adjacency matrix is drawn such that
-\(A_{ij} \stackrel{\text{ind}}{\sim}Bernoulli(P_{ij})\), is a special
-case of the GRDPG. It can also be shown that the latent positions of the
-PABM under the GRDPG framework consists of \(K\) \(K\)-dimensional
-subspaces in \(\mathbb{R}^{K^2}\). While there is no unique latent
-configuration \(X\) such that \(X X^\top = P\), the edge probability
-\(P\) for the PABM, they all have this subspace structure, and one in
-particular consists of \emph{orthogonal} subspaces.
-
-\begin{theorem}[Connecting the PABM to the GRDPG for $K = 2$]
-\label{theorem1}  
-Let
-
-$$X = \begin{bmatrix} 
-\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\ 
-0 & 0 & \lambda^{(21)} & \lambda^{(22)} 
-\end{bmatrix}$$
-
-$$U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
-0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
-0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
-0 & 1 & 0 & 0 \end{bmatrix}$$
-
-where each $\lambda^{(kl)}$ is a vector as in Definition 1. 
-Then $A \sim GRDPG_{3, 1}(X U)$ and $A \sim PABM(\{(\lambda^{(kl)}\}_2)$ are 
-equivalent.
-\end{theorem}
-
-\begin{theorem}[Generalization to $K > 2$] 
-\label{theorem2}
-There exists a block diagonal matrix 
-$X \in \mathbb{R}^{n \times K^2}$ defined by PABM parameters 
-$\{\lambda^{(kl)}\}_K$ and orthonormal matrix 
-$U \in \mathbb{R}^{K^2 \times K^2}$ that is fixed 
-for each $K$ such that $A \sim GRDPG_{K (K+1) / 2, K (K-1) / 2}(XU)$ and 
-$A \sim PABM(\{(\lambda^{(kl)}\})_K)$ are equivalent.
-\end{theorem}
-
-\begin{proof}
-Define the following matrices from $\{\lambda^{(kl)}\}_K$:
-
-$$\Lambda^{(k)} = 
-\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
-\in \mathbb{R}^{n_k \times K}$$
-
-\begin{equation} \label{eq:xy}
-X = \text{blockdiag}(\Lambda^{(1)}, ..., \Lambda^{(K)}) \in \mathbb{R}^{n \times K^2}
-\end{equation}
-
-$$L^{(k)} = \text{blockdiag}(\lambda^{(1k)}, ..., \lambda^{(Kk)}) \in 
-\mathbb{R}^{n \times K}$$
-
-$$Y = \begin{bmatrix} L^{(1)} & \cdots & L^{(K)} \end{bmatrix} \in 
-\mathbb{R}^{n \times K^2}$$
-
-Then $P = X Y^\top$.
-
-Note that $Y = X \Pi$ for a permutation matrix
-$\Pi$, resulting in $P = X \Pi X^\top$.  
-The permutation described by $\Pi$ has $K$ fixed points, which correspond to 
-$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where 
-$k = r (K + 1) + 1$ for $r = 0, ..., K - 1$. It also has 
-$\binom{K}{2} = K (K - 1) / 2$ cycles of order $2$. Each cycle corresponds to 
-a pair of eigenvalues $+1$ and $-1$ and a pair of eigenvectors 
-$(e_s + e_t) / \sqrt{2}$ and $(e_s - e_t) / \sqrt{2}$.
-
-Then $\Pi$ has $K (K + 1) / 2$ eigenvalues equal to $1$ and $K (K - 1) / 2$ 
-eigenvalues equal to $-1$. $\Pi$ has the decomposed form
-
-\begin{equation} \label{eq:permutation}
-\Pi = U I_{K (K + 1) / 2, K (K - 1) / 2} U^\top
-\end{equation}
-
-The edge probability matrix then can be written as:
-
-\begin{equation} \label{eq:pabm-grdpg}
-P = X U I_{p, q} (X U)^\top
-\end{equation}
-
-\begin{equation} \label{eq:p}
-p = K (K + 1) / 2
-\end{equation}
-
-\begin{equation} \label{eq:q}
-q = K (K - 1) / 2
-\end{equation}
-
-and we can describe the PABM with $K$ communities as a GRDPG with latent 
-positions $X U$ with signature $\big( K (K + 1) / 2, K (K - 1) / 2 \big)$.
-\end{proof}
-
-\begin{example}[$K = 3$] Using the same notation as in Theorem \ref{theorem2}:
-
-$$X = \begin{bmatrix} 
-\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
-\end{bmatrix}$$
-
-$$Y = \begin{bmatrix} 
-\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
-0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
-0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
-\end{bmatrix}$$
-
-Then $P = X Y^\top$ and $Y = X \Pi$ where $\Pi$ is a permutation matrix 
-consisting of $3$ fixed points and $3$ cycles of order 2:
-
-$$\Pi = \begin{bmatrix} 
-1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
-0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
-0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1
-\end{bmatrix}$$
-
-\item Positions 1, 5, 9 are fixed.
-
-\item The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.
-
-Therefore, we can decompose $\Pi = U I_{6, 3} U^\top$ where the first three 
-columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the fixed 
-positions $1$, $5$, and $9$, the next three columns consist of eigenvectors 
-$(e_k + e_l) / \sqrt{2}$, and the last three columns consist of eigenvectors 
-$(e_k - e_l) / \sqrt{2}$, where pairs $(k, l)$ correspond to the cycles of 
-order 2 described above.
-
-The latent positions are the rows of  
-$$XU = \begin{bmatrix}
-  \lambda^{(11)} & 0 & 0 & 
-  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 & 
-  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 \\
-  0 & \lambda^{(22)} & 0 & 
-  \lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} & 
-  -\lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} \\
-  0 & 0 & \lambda^{(33)} & 
-  0 & \lambda^{(31)} / \sqrt{2} & \lambda^{(32)} / \sqrt{2} & 
-  0 & -\lambda^{(31)} / \sqrt{2} & -\lambda^{(32)} / \sqrt{2}
-\end{bmatrix}$$
-\end{example}
-
-In Theorem \ref{theorem2}, we showed that a possible latent
-configuration for a \(K\)-community PABM under the GRDPG framework
-consists of vectors in \(\mathbb{R}^{K^2}\) such that each community
-corresponds to a \(K\)-dimensional subspace and each subspace is
-orthogonal to the others.
-
-\begin{theorem}
-\label{osc}
-Let $P = V \Lambda V^\top$ be the spectral decomposition of the edge probability matrix of a PABM. Define $B = n V V^\top$. Then $B_{ij} = 0$ $\forall i, j$ in different communities. 
-\end{theorem}
-
-If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), then results from
-\citet{rubindelanchy2017statistical} imply
-\(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top\). Then for
-vertices \(i\) and \(j\) belonging to separate communities, the
-\(ij\)\textsuperscript{th} entry of \(n \hat{V} \hat{V}^\top\)
-approaches 0 with probability 1:
-
-\begin{theorem}
-\label{theorem4} 
-Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
-(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
-and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,
-
-\begin{equation} \label{eq:thm4}
-\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
-O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
-\end{equation}
-
-This provides the result that $\forall i, j$ in different communities, 
-$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
-\end{theorem}
-
-\begin{algorithm}[t]
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \KwData{Adjacency matrix $A$, number of communities $K$}
-  \KwResult{Community assignments $1, ..., K$}
-    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
-    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
-    $V$ using these eigenvectors as its columns.\;
-    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
-    Construct graph $G$ using $B$ as its similarity matrix.\;
-    Partition $G$ into $K$ disconnected subgraphs  
-    (e.g., using edge thresholding or spectral clustering).\;
-    Map each partition to the community labels $1, ..., K$.\;
-  \caption{Orthogonal Spectral Clustering.}
-\end{algorithm}
-
-Since the communities are subspaces in the latent space, we can also use
-existing methods for subspace clustering on the ASE. Of particular
-interest is Sparse Subspace Clustering (SSC), which is performed by
-solving an optimization problem for each observed point in a sample.
-Given \(X \in \mathbb{R}^{n \times d}\) with vectors
-\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
-\(c_i = \arg\min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
-\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
-collected into matrix
-\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
-construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
-perfectly on one of \(K\) subspaces, \(B\) is sparse such that
-\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
-Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
-and if the number of subgraphs is exactly \(K\), each subgraph maps onto
-a subspace.
-
-In practice, SSC is performed by solving the LASSO problems:
-
-\begin{equation} \label{eq:ssc}
-c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
-\end{equation}
-
-for some sparsity parameter \(\lambda > 0\). The \(c_i\) vectors are
-then collected into \(C\) and \(B\) as described before. If \(X\) is
-noisy such that each \(x_i\) does not lie exactly on one of \(K\)
-subspaces but near it, the choice of \(\lambda\) becomes important in
-guaranteeing the Subspace Detection Property (SDP)
-\cite{jmlr-v28-wang13}.
-
-\begin{definition}[Subspace Detection Property] 
-Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
-points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
-solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
-$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
-different subspaces, then $X$ obeys the Subspace Detection Property. 
-\end{definition}
-
-\begin{remark} 
-In practice, a noisy sample $X$ often does not obey SDP. 
-In such cases, $B$ is treated as an affinity matrix for a graph which 
-is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
-hand, if $X$ does obey the SDP, $B$ describes a graph 
-with at least $K$ disconnected subgraphs. Ideally, when SDP holds, 
-there are exactly $K$ subgraphs which map to each subspace, 
-but it could be the case that some of the subspaces are represented by 
-multiple disconnected subgraphs.
-\end{remark}
-
-Since every ASE of the PABM consists of subspaces and as
-\(n \to \infty\) each vector of the ASE approaches its subspace
-uniformly \cite{rubindelanchy2017statistical}, SSC is also able to
-perform community detection for the PABM. It has been shown by
-\citet{jmlr-v28-wang13} that if the points lie sufficiently close to
-their respective subspaces and the cosine of the angles between
-subspaces is sufficiently small, SDP will hold. We now state that the
-unscaled ASE of the PABM exhibits exactly these conditions for
-sufficiently large \(n\).
-
-\begin{theorem}
-\label{theorem5}
-Let $P_n$ describe the edge probability matrix of the PABM with 
-$n$ vertices, and let $A_n \sim \text{Bernoulli}(P_n)$.  Let $\hat{V}_n$ be the 
-matrix of eigenvectors of $A_n$ corresponding to the $K (K + 1) / 2$ most
-positive and $K (K - 1) / 2$ most negative eigenvalues. Then 
-$\exists \lambda > 0$ and $N < \infty$ such that when $n > N$, 
-$\sqrt{n} \hat{V}_n$ obeys the subspace detection property with probability 1.
-\end{theorem}
-
-\hypertarget{non-spectral-approaches-to-community-detection}{%
-\subsection{Non-Spectral Approaches to Community
-Detection}\label{non-spectral-approaches-to-community-detection}}
-
-So far, we connected the PABM to the GRDPG and used this to motivate
-community detection methods based on the ASE of the adjacency matrix
-sampled from the PABM. In this section, we will explore non-spectral
-methods for community detection.
-
-Denoting \(\{\lambda_{ik}\}\) as the scalar popularity parameters for
-the PABM such that \(\lambda^{(kl)}\) is the vector of \(\lambda_{jl}\)
-terms for which vertex \(j\) is in community \(k\), we can write the
-likelihood for the PABM:
-
-\[p(A | z, \{\lambda_{ik}\}) = 
-\prod_{i < j} (\lambda_{i z_j} \lambda_{j z_i})^{A_{ij}} 
-(1 - \lambda_{i z_j} \lambda_{j z_i})^{1 - A_{ij}}\]
-
-\begin{equation}
-= \prod_{i < j} \prod_k^K \prod_l^K 
-(\lambda_{il} \lambda_{jk})^{A_{ij} z_{ik} z_{jl}} 
-(1 - \lambda_{il} \lambda_{jk})^{(1 - A_{ij}) z_{ik} z_{jl}}
-\end{equation}
-
-where \(z_{ik} = I(Z_i = k)\).
-
-The log-likelihood is
-
-\begin{equation}
-\log p = \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
-(A_{ij} \log \lambda_{il} \lambda_{jk} + 
-(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk}))
-\end{equation}
-
-\hypertarget{expectation-maximization}{%
-\subsubsection{Expectation
-Maximization}\label{expectation-maximization}}
-
-While direct maximization of the likelihood or log-likelihood is
-NP-complete, it is possible to apply an Expectation Maximization (EM)
-algorithm to this problem. To set this up, we write the complete data
-log likelihood as:
-
-\[Q = 
-\sum_{i < j} \sum_{k, l} z_{ik} z_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} +
-(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))
-+ \sum_i \sum_k z_{ik} \log \pi_k\]
-
-Where \(\pi_k\) are the \emph{a priori} community label probabilities.
-
-Then the expectation step involves solving for
-\(\gamma_{ik} = P(Z_i = k \mid \{\pi_l\}, \{\lambda_{jl}\})\), which is
-given by
-
-\[\log \gamma_{ik} \propto 
-\log \pi_k + 
-\sum_{j \neq i} \sum_l \pi_{jl} (A_{ij} \log \lambda_{il} \lambda_{jk} + 
-(1 - A_{ij}) \log(1 - \lambda_{il} \lambda_{jk}))\]
-
-For the maximization step, we can just set
-\(\pi_k = \frac{1}{n} \sum_i \gamma_{ik}\), but the \(\{\lambda_{ik}\}\)
-parameters cannot be maximized in closed form and require an
-optimization subroutine. Our proposed work involves finding an efficient
-optimization method for this subroutine.
-
-\hypertarget{bayesian-methods}{%
-\subsubsection{Bayesian Methods}\label{bayesian-methods}}
-
-The introduction of community label probabilities \(\pi_1, ..., \pi_K\)
-suggests a hierarchical model in which the community labels are
-generated by a prior distribution. We can apply the same principle to
-the popularity parameters \(\{\lambda_{ik}\}\) to make the model fully
-Bayesian. In this setup, we sample
-\(z_i \stackrel{\text{iid}}{\sim}\text{Categorical}(\pi_1, ..., \pi_K)\)
-and
-\(\lambda_{ik} \stackrel{\text{ind}}{\sim}\text{Beta}(a_{ik}, b_{ik})\)
-before sampling the observed data
-\(A_{ij} \stackrel{\text{ind}}{\sim}\lambda_{i z_j} \lambda_{j z_i}\).
-
-The full joint distribution for this model can be written as:
-
-\[\begin{split}
-\log p & = constant \\
-& + \sum_{i < j} \sum_k \sum_l z_{ik} z_{jl} 
-(A_{ij} \log \lambda_{il} \lambda_{jk} + 
-(1 - A_{ij}) \log (1 - \lambda_{il} \lambda_{jk})) \\
-& + \sum_k \sum_i z_{ik} \log \pi_k \\
-& + \sum_i \sum_k (a_{ik} - 1) \log \lambda_{ik} + (b_{ik} - 1) \log (1 - \lambda_{ik})
-\end{split}\]
-
-It is straightforward to obtain the conditional distribution for each
-\(z_i\) but not each \(\lambda_{ik}\), so a Gibbs sampler cannot be
-derived for this model. However, we can use a Metropolis-Hastings
-algorithm. In particular, by taking a first-order Taylor approximation,
-we can approximate the conditional distribution for each
-\(\lambda_{ik}\) as a Beta distribution. Our proposed work involves
-investigating this method by simulating this model and implementing the
-Bayes sampler.
-
-\hypertarget{generalizations-to-grdpg-based-community-detection}{%
-\section{Generalizations to GRDPG-Based Community
-Detection}\label{generalizations-to-grdpg-based-community-detection}}
-
-In the previous sections, we connected well-known and highly restricted
-generative graph models to the RDPG and GRDPG to show that highly
-structured latent configurations generate graphs consistent with these
-models: The latent space for the SBM consists of \(K\) point masses, the
-latent space for the DCBM consists of \(K\) rays emanating from the
-origin, and the latent space for the PABM consists of \(K\)
-\(K\)-dimensional subspaces. In the following sections, we explore
-additional structured latent configurations corresponding to community
-structure and develop methods for community detection based on the
-consistency of the ASE and the structural forms of the latent
-configurations.
-
-The general structure of interest can be described as follows: Suppose
-that in the latent space \(\mathcal{X} \subset \mathbb{R}^d\), sample
-\(X\) of \(n\) points lie on a union of \(K\) disjoint manifolds with
-each manifold corresponding to a community. If
-\(A \sim \text{RDPG}(X)\), we wish to recover the community labels (up
-to permutation) from \(A\).
-
-Equivalently, suppose that probability distribution \(F\) is described
-as follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Define functions \(f_1, ..., f_K\) such that
-  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_k(t) \neq f_l(t)\)
-  \(\forall k, l \in [K]\) and \(t \in [0, 1]\).
-\item
-  Sample labels
-  \(z_1, ..., z_n \stackrel{\text{iid}}{\sim}Categorical(\pi_1, ..., \pi_K)\).
-\item
-  Sample \(t_1, ..., t_n \stackrel{\text{iid}}{\sim}D\) where \(D\) has
-  support \([0, 1]\).
-\item
-  Set latent positions \(x_i = f_{z_i}(t_i)\) and
-  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
-\end{enumerate}
-
-Then if \((A, X) \sim \text{RDPG}(F, n)\) and we observe \(A\), we wish
-to recover hidden labels \(z_1, ..., z_n\).
-
-\hypertarget{affine-segments}{%
-\subsection{Affine Segments}\label{affine-segments}}
-
-In this section, we will consider latent positions sampled uniformly
-from parallel unit length segments.
-
-\begin{example}
-Let $U_1, ..., U_{n_1}, V_1, ..., V_{n_2} \stackrel{\text{iid}}{\sim}Uniform(0, 1)$, $f_1(t) = (t, 0)$, and $f_2(t) = (t, a)$.
-$X_i = f_1(U_i)$ and $Y_j = f_2(V_j)$ $\forall i \in [n_1]$ and $j \in [n_2]$.
-If we observe $X_1, ..., X_{n_1}, Y_1, ..., Y_{n_2}$, what approach will allow us to group the observations by segment?
-\end{example}
-
-\begin{figure}[H]
-
-{\centering \includegraphics{proposal_files/figure-latex/unnamed-chunk-1-1}
-
-}
-
-\caption{Points sampled uniformly on two parallel unit length line segments}\label{fig:unnamed-chunk-1}
-\end{figure}
-
-We will motivate an approach by the following example.
-
-\begin{example}
-Let $U_1, ..., U_n \stackrel{\text{iid}}{\sim}Uniform(0, 1)$ with order statistics 
-$U_{(1)}, ..., U_{(n)}$. Then $\forall a \in (0, 1)$ and $\delta \in (0, 1)$, $\exists N = N(\delta, a) < \infty$ such that $\forall n \geq N$,
-
-\begin{equation}
-P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq 1 - \delta / 2
-\end{equation}
-
-Where $N(\delta, a)$ is monotone increasing with respect to $\delta$ and $a$. To prove this, we start with the fact that $U_{(i+1)} - U_{(i)} \sim Beta(1, n)$. Then
-
-\begin{equation}
-P(U_{(i+1)} - U_{(i)} \leq a) = 1 - (1 - a)^n
-\end{equation}
-
-and
-
-\begin{equation}
-\label{eq:unsolvable}
-P(\max_i U_{(i+1)} - U_{(i)} \leq a) \geq (P(U_{(i+1)} - U_{(i)} \leq a))^n = (1 - (1 - a)^n)^n
-\end{equation}
-
-This expression is monotone increasing $\forall n \geq N_1$ for some $N_1 < \infty$. 
-Setting $(1 - (1 - a)^{N_2})^{N_2} \geq 1 - \delta / 2$, we can solve for a finite ${N_2}$. Then $N = \max(N_1, N_2)$.
-\end{example}
-
-If we extend this example such that \(n_1\) points are sampled uniformly
-from the segment \(f_1(t) = (t, 0)\) and \(n_2\) points are sampled
-uniformly from the segment \(f_2(t) = (t, a)\) for
-\(t \in [0, \cos \frac{\pi}{2} a]\), then a sample of size
-\(\min(n_1, n_2) \geq N(\delta, a)\) is sufficient to satisfy:
-
-\begin{equation}
-P(\max_i X_{(i+1)} - X_{(i)} \leq a \leq \min_{i, j} \|X_i - Y_j\|) 
-\geq 1 - \delta / 2
-\end{equation}
-
-and similar for
-\(P(\max_j Y_{(j+1)} - Y_{(j)} \leq \min_{i, j} \|X_i - Y_j\|)\), for
-\(X_i\) in the first segment and \(Y_j\) in the second segment and
-\(X_{(i)}\), \(Y_{(j)}\) are order statistics in the first coordinate.
-If each segment corresponds to a community, this leads to the following
-two results:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Single linkage clustering with \(K = 2\) will perform perfect
-  community detection with probability at least \(1 - \delta\).
-\item
-  An \(\epsilon\)-neighborhood graph with \(\epsilon \in (0, a)\) will
-  consists of at least 2 disjoint subgraphs such that no subgraph
-  consists of members of two different communities (analogous to the
-  SDP), with probability at least \(1 - \delta\).
-\end{enumerate}
-
-We can then further extend this to the case where points are drawn from
-unit segments with noise. It is straightforward to show that if the
-maximum noise is less than \(a / 3\), we can obtain a similar result as
-in the noiseless case. Thus if the distribution on the two segments is
-used to draw \((A, X) \sim \text{RDPG}(F, n)\), the same properties
-should hold for large \(n\).
-
-\hypertarget{manifolds}{%
-\subsection{Manifolds}\label{manifolds}}
-
-If instead of sampling uniformly from line segments of unit length, we
-sample uniformly from a 1 dimensional manifolds of unit length, the
-above property still holds. Let
-\(U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)\) and
-\(f : [0, 1] \mapsto \mathbb{R}^d\) be a smooth function such that
-\(\int_u^v \sqrt{\sum_k (df_k / dt)^2} dt = \|u - v\|\). Then
-\(U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
-\(P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
-If the shortest distance between the two manifolds defined by \(f_1\)
-and \(f_2\) with the same restriction is \(a\), then the same \(N\) as
-before is sufficient, although perhaps a more lenient lower bound can be
-derived based on the shape of \(f_k(\cdot)\).
-
-Our proposed work extends these results by the following.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Show that the ASE of a (G)RDPG generated by this type of latent
-  configuration produces the correct conditions as described above to
-  apply the same probability statements.
-\item
-  Extend these results by deriving similar probability statements for
-  nonuniform distributions.
-\item
-  Relax the condition for the minimum distance between manifolds by
-  investigating alternative clustering techniques on the latent space
-  (e.g., spectral clustering).
-\end{enumerate}
-
-\hypertarget{summary}{%
-\section{Summary}\label{summary}}
-
-For this research, we propose using the connection between Block Models
-and the (Generalized) Random Dot Product Graph to motivate community
-detection methods. Previous research shows that the Stochastic Block
-Model and Degree Corrected Block Model have very structured latent
-configurations in the RDPG framework, and these latent structures can be
-used to develop community detection algorithms that result in perfect
-community detection for large \(n\). In the same vein, we investigated
-the connection between the Popularity Adjusted Block Model and the GRDPG
-to motivate two community detection algorithms, the first of which is an
-application of an existing algorithm, Sparse Subspace Clustering, and
-the second is a new algorithm based on the particular latent form of the
-PABM, Orthogonal Spectral Clustering. Furthermore, we showed that for
-large \(n\), our application of the SSC algorithm to the PABM upholds
-the Subspace Detection Property for large yet finite \(n\), and the
-output of OSC results in perfect community detection almost surely as
-\(n \to \infty\). We will extend these theoretical results with
-empirical simulations and real data examples. Finally, we will implement
-these two community detection algorithms in an R package.
-
-We aim to extend these results by defining more general structures in
-the latent space to construct RDPG or GRDPG models and show that
-community detection is possible via Adjacency Spectral Embedding. So
-far, we showed that if in the latent space the communities can be
-represented as points sampled uniformly on curves or one-dimensional
-manifolds separated by a nonzero distance, we can apply two existing
-clustering techniques to the ASE and show that they will perform perfect
-community detection with high probability. We will generalize these
-results to nonuniform distributions on higher dimensional manifolds as
-well as investigate alternative clustering techniques for the latent
-space that may perform better for smaller \(n\).
-
-\hypertarget{estimated-timeline-of-completion}{%
-\section{Estimated Timeline of
-Completion}\label{estimated-timeline-of-completion}}
-
-Literature review: August 2021
-
-Complete proofs of main theorems: January 2022
-
-Simulations, real data analyses, R package: March 2022
-
-Dissertation completion: April 2022
-
-\renewcommand\refname{References}
   \bibliography{proposal.bib}

### 2021-09-01 scratch/distributions-on-1d-manifolds.Rmd
proof sketch for general case

@@
 
-$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} (i-1) \binom{n}{i-1} f(x) (F(x))^{i-1} (1 - F(x + \delta))^{n-i} dx$$
+$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x + \delta))^{n-i} f(x) dx$$
 
-Under certain conditions (namely $F(x + \delta) \geq F(x) + \delta$), we can make the substitution $u = F(x)$ and have the result
+Making the approximation $F(x+\delta) \approx F(x) + \delta f(x)$ 
+and bounding $f(x) \geq a > 0$, we get:
 
-$$P(D_i \leq \delta) \leq 1- \int_0^{1 - \delta} (i-1) \binom{n}{i-1} u^{i-1} (1 - u - \delta)^{n-i} du$$
+$$P(D_i \leq \delta) \geq 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x) - a \delta)^{n-i} f(x) dx$$
 
-which gives the same result as in the uniform case.
+Then making the substitution $u = F(x) \implies du = f(x) dx$, we get
+
+$$1 - \int_0^{F(1-\delta)} \frac{n!}{(n-i+1)! (i-2)!} u^{i-2} (1 - u - a \delta)^{n-i} du$$
+
+Which becomes $1 - C P(n) e^{-n}$ for some $C > 0$ and $P(n)$ is a polynomial of $n$. This goes to 0 as $n \to \infty$, giving us the result $D_i \stackrel{p}{\to} 0$.
 

### 2021-09-01 scratch/distributions-on-1d-manifolds.tex
proof sketch for general case

@@
 
-\[P(D_i \leq \delta) = 1 - \int_0^{1-\delta} (i-1) \binom{n}{i-1} f(x) (F(x))^{i-1} (1 - F(x + \delta))^{n-i} dx\]
+\[P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x + \delta))^{n-i} f(x) dx\]
 
-Under certain conditions (namely \(F(x + \delta) \geq F(x) + \delta\)),
-we can make the substitution \(u = F(x)\) and have the result
+Making the approximation \(F(x+\delta) \approx F(x) + \delta f(x)\) and
+bounding \(f(x) \geq a > 0\), we get:
 
-\[P(D_i \leq \delta) \leq 1- \int_0^{1 - \delta} (i-1) \binom{n}{i-1} u^{i-1} (1 - u - \delta)^{n-i} du\]
+\[P(D_i \leq \delta) \geq 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x) - a \delta)^{n-i} f(x) dx\]
 
-which gives the same result as in the uniform case.
+Then making the substitution \(u = F(x) \implies du = f(x) dx\), we get
+
+\[1 - \int_0^{F(1-\delta)} \frac{n!}{(n-i+1)! (i-2)!} u^{i-2} (1 - u - a \delta)^{n-i} du\]
+
+Which becomes \(1 - C P(n) e^{-n}\) for some \(C > 0\) and \(P(n)\) is a
+polynomial of \(n\). This goes to 0 as \(n \to \infty\), giving us the
+result \(D_i \stackrel{p}{\to} 0\).
 

### 2021-09-01 scratch/distributions-on-1d-manifolds.tex
proof sketch for general case

@@
 
-\hypertarget{beta-case}{%
-\subsubsection{Beta case}\label{beta-case}}
-
 \hypertarget{general-case}{%
 \subsubsection{General case}\label{general-case}}
+
+\hypertarget{proof-sketch-that-pd_i-leq-delta-to-1}{%
+\paragraph{\texorpdfstring{Proof sketch that
+\(P(D_i \leq \delta) \to 1\)}{Proof sketch that P(D\_i \textbackslash leq \textbackslash delta) \textbackslash to 1}}\label{proof-sketch-that-pd_i-leq-delta-to-1}}
 
@@
 
-Which becomes \(1 - C P(n) e^{-n}\) for some \(C > 0\) and \(P(n)\) is a
-polynomial of \(n\). This goes to 0 as \(n \to \infty\), giving us the
-result \(D_i \stackrel{p}{\to} 0\).
+Which becomes \(1 - C P(n) \delta^{n}\) for some \(C > 0\) and \(P(n)\)
+is a polynomial of \(n\). This goes to 0 as \(n \to \infty\), giving us
+the result \(D_i \stackrel{p}{\to} 0\).
 

### 2021-09-01 scratch/distributions-on-1d-manifolds.Rmd
proof sketch for general case
typo

@@
 
-## Setup
+# Setup
 
@@
 
-## Theory
+# Theory
 
@@
 
-### Uniform case
+## Uniform case
 
@@
 
-### Beta case
+## General case
 
-### General case
+### Proof sketch that $P(D_i \leq \delta) \to 1$
 
@@
 
-Which becomes $1 - C P(n) e^{-n}$ for some $C > 0$ and $P(n)$ is a polynomial of $n$. This goes to 0 as $n \to \infty$, giving us the result $D_i \stackrel{p}{\to} 0$.
+Which becomes $1 - C P(n) \delta^{n}$ for some $C > 0$ and $P(n)$ is a polynomial of $n$, giving us the desired result $D_i \stackrel{p}{\to} 0$.
 
-## Computational Results
+# Computational Results
 

### 2021-09-01 scratch/distributions-on-1d-manifolds.tex
proof sketch for general case
typo

@@
 \hypertarget{setup}{%
-\subsection{Setup}\label{setup}}
+\section{Setup}\label{setup}}
 
@@
 \hypertarget{theory}{%
-\subsection{Theory}\label{theory}}
+\section{Theory}\label{theory}}
 
@@
 \hypertarget{uniform-case}{%
-\subsubsection{Uniform case}\label{uniform-case}}
+\subsection{Uniform case}\label{uniform-case}}
 
@@
 \hypertarget{general-case}{%
-\subsubsection{General case}\label{general-case}}
+\subsection{General case}\label{general-case}}
 
 \hypertarget{proof-sketch-that-pd_i-leq-delta-to-1}{%
-\paragraph{\texorpdfstring{Proof sketch that
+\subsubsection{\texorpdfstring{Proof sketch that
 \(P(D_i \leq \delta) \to 1\)}{Proof sketch that P(D\_i \textbackslash leq \textbackslash delta) \textbackslash to 1}}\label{proof-sketch-that-pd_i-leq-delta-to-1}}
@@
 Which becomes \(1 - C P(n) \delta^{n}\) for some \(C > 0\) and \(P(n)\)
-is a polynomial of \(n\). This goes to 0 as \(n \to \infty\), giving us
-the result \(D_i \stackrel{p}{\to} 0\).
+is a polynomial of \(n\), giving us the desired result
+\(D_i \stackrel{p}{\to} 0\).
 
 \hypertarget{computational-results}{%
-\subsection{Computational Results}\label{computational-results}}
+\section{Computational Results}\label{computational-results}}
 

### 2021-09-15 scratch/distributions-on-1d-manifolds.Rmd
cleaned up some stuff

@@
 
+\newcommand{\diag}{\mathrm{diag}}
+\newcommand{\tr}{\mathrm{Tr}}
+\newcommand{\blockdiag}{\mathrm{blockdiag}}
+\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
+\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
+\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
+\newcommand{\Betadist}{\mathrm{Beta}}
+\newcommand{\Uniform}{\mathrm{Uniform}}
+\newcommand{\BG}{\mathrm{BernoulliGraph}}
+\newcommand{\Categorical}{\mathrm{Categorical}}
+\newtheorem{definition}{Definition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\theoremstyle{remark}
+\newtheorem*{remark}{Remark}
+\theoremstyle{example}
+\newtheorem*{example}{Example}
+\newcommand{\dd}{\mathrm{d}}
+\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
+
 # Setup
@@
 
-# Theory
+# Preliminary Theory
 
-Let $D_i = X_{(i)} - X_{(i-1)}$. Then if $\max_i D_i < \delta$, we have sufficient separation of points in $\mathcal{M}_1$.
+## Distributions of differences of order statistics
 
-## Uniform case
+Let $D_i = X_{(i+1)} - X_{(i)}$. Then if $\max_i D_i < \delta$, we have sufficient separation of points in $\mathcal{M}_1$. 
+Then it is sufficient to quantify $P(\max_i D_i > \delta)$ as a function of $n$ and $\delta$ and show that this converges to zero as $n$ grows to $\infty$.
 
-It can be shown that if $X_i \stackrel{iid}{\sim} Uniform(0, 1)$, then $D_i \sim Beta(1, n)$.
-Therefore, $P(\max_i D_i < \delta) \geq (P(D_i < \delta))^n = (1 - (1 - \delta)^n)^n$, which is a decreasing function for sufficiently large $n$.
-This gives us the result $\max_i D_i \stackrel{a.s.}{\to} 0$.
+We denote $f(x)$ as the density of each $X_i$, $g_i(x)$ as the density of $X_{(i)}$, 
+$g_{ij}(x, y)$ as the joint density of $X_{(i)}, X_{(j)}$, and $h_i(d)$ as the density of $D_i$ (with corresponding capital letters for the cumulative distribution functions).
 
-## General case
+The following are taken as given[^wikipedia]:
 
-### Proof sketch that $P(D_i \leq \delta) \to 1$
+1. $g_i(x) = \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x))^{n-i} f(x)$.
+2. $g_{ij}(x) = \frac{n!}{(i-1)! (j-i-1)! (n-j)!} (F(x))^{i-1} (F(y) - F(x))^{j-i-1} (1 - F(y))^{n-j} f(x) f(y)$.
+3. By convolution, $h_i(d) = \int_0^{1} g_{i, i+1} (x, x + d) \dd x$.
+
+[^wikipedia]: https://en.wikipedia.org/wiki/Order_statistic
+
+\begin{lemma}[The probability density function of $D_i$]
+\begin{equation}
+\label{eq:pdf}
+h_i(d) = \int_0^{1-d} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} (1 - F(x+d))^{n-i-1} f(x) f(x+d) \dd x 
+\end{equation}
+\end{lemma}
+
+\begin{proof}
+This is just given by setting the integrand as $g_{i, i+1}(x, x+d)$.
+We briefly note that the integral can only be evaluated from 0 to $1-d$ in this case because of the $x+d$ terms.
+\end{proof}
+
+\begin{lemma}[The cumulative distribution function of $D_i$]
+
+\begin{equation}
+\label{eq:cdf}
+P(D_i < \delta) = H_i(\delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \dd x
+\end{equation}
+\end{lemma}
+
+\begin{proof}
+$$
+\begin{aligned}
+H_i(\delta) & = \int_x^{x+\delta} h_i(d) \dd d \\
+& = \int_x^{x+\delta} \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} ((F(x))^{i-1} (1 - F(x+d))^{n-i-1} f(x) f(x+d) \dd x \dd d \\
+& = \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} f(x) \int_x^{x+\delta} (1 - F(x+d))^{n-i-1} f(x+d) \dd d \dd x \\
+& = \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} f(x) \int_{F(x)}^{F(x+\delta)} (1 - u)^{n-i-1} \dd u \dd x \\
+& = \int_0^{1} \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} f(x) ((1 - F(x))^{n-i} - (1 - F(x + \delta))^{n-i}) \dd x \\ 
+& = \int_0^1 g_i(x) \dd x - \int_0^1 \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \dd x \\
+& = 1 - \int_0^1 \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \dd x \\
+\end{aligned}
+$$
+
+Because of the $x + \delta$ term, we can't actually evaluate this integral all the way up to $1$, and so we are left with 
+$$
+\begin{aligned}
+& = 1 - \int_0^{1 - \delta} \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \dd x.
+\end{aligned}
+$$
+\end{proof}
+
+Thus if we are able to show that Eq. (\ref{eq:cdf}) converges to 1, we are done.
+
+### Uniform case
+
+\begin{lemma}[Differences between order statistics of a uniform distribution]
+If $X_1, ..., X_n \iid \Uniform(0, 1)$, then each $D_i \sim \Betadist(1, n)$.
+\end{lemma}
+
+\begin{proof}
+We begin with Eq. (\ref{eq:pdf}), plugging in $f(x) = 1$ and $F(x) = x$:
+
+$$
+h_i(d) = \int_0^{1-d} \frac{n!}{(i-1)! (n-i-1)!} x^{i-1} (1-x-d)^{n-i-1} \dd x
+$$
+
+Then we proceed with integration by parts, setting 
+$u = x^{i-1} \implies du = (i-1) x^{i-2}$ and 
+$dv = (1-x-d)^{n-i-1} dx \implies v = -\frac{1}{n-i} (1-x-d)^{n-i-1}$. 
+Note that $u v |_0^{1-d} = 0$ in this case. This yields
+
+$$
+= \frac{n!}{(i-1)! (n-i-1)!} \int \frac{i-1}{n-i} x^{i-2} (1-x-d)^{n-i} \dd x
+$$
+
+Then applying integration by parts again until the $x^p$ term disappears, we get:
+
+$$
+\begin{aligned}
+& = \frac{n!}{(i-1)! (n-i-1)!} \frac{(i-1)!}{(n-i) \cdots (n-2)} \int_0^{1-d} (1-x-d)^{n-2} \dd x \\
+& = -\frac{n (n-1)}{n-1} (1-x-d)^{n-1} \Big|_0^{1-d} \\
+& = n (1 - d)^{n-1}
+\end{aligned}
+$$
+
+This the density function for $\Betadist(1, n)$, completing the proof.
+\end{proof}
+
+\begin{theorem}
+Let $X_1, ..., X_n \iid \Uniform(0, 1)$. 
+Then for any $\delta > 0$, $P(\max_i X_{(i+1)} - X_{(i)} < \delta) \to 1$ as 
+$n \to \infty$.
+\end{theorem}
+
+\begin{proof}[Proof (sketch)]
+Since $X_{(i+1)} - X_{(i)} = D_i \sim \Betadist(1, n)$, 
+$P(X_{(i+1)} - X_{(i)} < \delta) = 1 - (1 - \delta)^n $. This yields
+
+$$
+\begin{aligned}
+P(\max_i D_i < \delta) & \geq (P(D_i < \delta))^{n-1} \\
+& = (1 - (1 - \delta)^n)^{n-1} \\
+& \approx e^{-n \exp(-n \delta)}.
+\end{aligned}
+$$
+
+In the limit $n \to \infty$, this goes to 1.
+\end{proof}
+
+### General case
+
+\begin{theorem}
+Let $X_1, ..., X_n \iid F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
+Let $D_i = X_{(i+1)} - X_{(i)}$. 
+Then for any $\epsilon > 0$, there exists $N \in \mathbb{N}$ such that $P(\max_i D_i < \delta) > 1 - \epsilon$ when $n > N$.
+\end{theorem}
+
+\begin{proof}[Proof (sketch)]
+We start with Eq. (\ref{eq:cdf}):
+
+$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \dd x.$$
+
+Making the approximation $F(x+\delta) \approx F(x) + \delta f(x)$ 
+and bounding $f(x) \geq a$, we get:
+
+$$P(D_i \leq \delta) \geq 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x) - a \delta)^{n-i} f(x) \dd x.$$
+
+Then making the substitution $u = F(x) \implies du = f(x) \dd x$, we obtain
+
+$$1 - \int_0^{F(1-\delta)} \frac{n!}{(n-i)! (i-1)!} u^{i-1} (1 - u - a \delta)^{n-i} \dd u$$
+
+Evaluating the integral yields
+
+$$P(D_i < \delta) = 1 - (1 - a \delta)^n + (1 - F(1-\delta) - a \delta)^n.$$
+
+Then as before,
+
+$$
+\begin{aligned}
+P(\max_i D_i < \delta) & \geq (P(D_i < \delta))^{n-1} \\
+& \approx (1 - (1 - a \delta)^n + (1 - F(1-\delta) - a \delta)^n)^{n-1} \\
+& \to 1
+\end{aligned}
+$$
+
+\end{proof}
 
@@
 
-$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x + \delta))^{n-i} f(x) dx$$
+$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \dd x$$
 
@@
 
+# Algorithms
+
+TBD
+
 # Computational Results

### 2021-09-15 scratch/distributions-on-1d-manifolds.tex
cleaned up some stuff

@@
 
+\newcommand{\diag}{\mathrm{diag}}
+\newcommand{\tr}{\mathrm{Tr}}
+\newcommand{\blockdiag}{\mathrm{blockdiag}}
+\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
+\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
+\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
+\newcommand{\Betadist}{\mathrm{Beta}}
+\newcommand{\Uniform}{\mathrm{Uniform}}
+\newcommand{\BG}{\mathrm{BernoulliGraph}}
+\newcommand{\Categorical}{\mathrm{Categorical}}
+\newtheorem{definition}{Definition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\theoremstyle{remark}
+\newtheorem*{remark}{Remark}
+\theoremstyle{example}
+\newtheorem*{example}{Example}
+\newcommand{\dd}{\mathrm{d}}
+\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
+
 \hypertarget{setup}{%
@@
 
-\hypertarget{theory}{%
-\section{Theory}\label{theory}}
+\hypertarget{preliminary-theory}{%
+\section{Preliminary Theory}\label{preliminary-theory}}
 
-Let \(D_i = X_{(i)} - X_{(i-1)}\). Then if \(\max_i D_i < \delta\), we
-have sufficient separation of points in \(\mathcal{M}_1\).
+\hypertarget{distributions-of-differences-of-order-statistics}{%
+\subsection{Distributions of differences of order
+statistics}\label{distributions-of-differences-of-order-statistics}}
+
+Let \(D_i = X_{(i+1)} - X_{(i)}\). Then if \(\max_i D_i < \delta\), we
+have sufficient separation of points in \(\mathcal{M}_1\). Then it is
+sufficient to quantify \(P(\max_i D_i > \delta)\) as a function of \(n\)
+and \(\delta\) and show that this converges to zero as \(n\) grows to
+\(\infty\).
+
+We denote \(f(x)\) as the density of each \(X_i\), \(g_i(x)\) as the
+density of \(X_{(i)}\), \(g_{ij}(x, y)\) as the joint density of
+\(X_{(i)}, X_{(j)}\), and \(h_i(d)\) as the density of \(D_i\) (with
+corresponding capital letters for the cumulative distribution
+functions).
+
+The following are taken as given\footnote{\url{https://en.wikipedia.org/wiki/Order_statistic}}:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \(g_i(x) = \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x))^{n-i} f(x)\).
+\item
+  \(g_{ij}(x) = \frac{n!}{(i-1)! (j-i-1)! (n-j)!} (F(x))^{i-1} (F(y) - F(x))^{j-i-1} (1 - F(y))^{n-j} f(x) f(y)\).
+\item
+  By convolution,
+  \(h_i(d) = \int_0^{1} g_{i, i+1} (x, x + d) \mathrm{d}x\).
+\end{enumerate}
+
+\begin{lemma}[The probability density function of $D_i$]
+\begin{equation}
+\label{eq:pdf}
+h_i(d) = \int_0^{1-d} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} (1 - F(x+d))^{n-i-1} f(x) f(x+d) \mathrm{d}x 
+\end{equation}
+\end{lemma}
+
+\begin{proof}
+This is just given by setting the integrand as $g_{i, i+1}(x, x+d)$.
+We briefly note that the integral can only be evaluated from 0 to $1-d$ in this case because of the $x+d$ terms.
+\end{proof}
+
+\begin{lemma}[The cumulative distribution function of $D_i$]
+
+\begin{equation}
+\label{eq:cdf}
+P(D_i < \delta) = H_i(\delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x
+\end{equation}
+\end{lemma}
+
+\begin{proof}
+$$
+\begin{aligned}
+H_i(\delta) & = \int_x^{x+\delta} h_i(d) \mathrm{d}d \\
+& = \int_x^{x+\delta} \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} ((F(x))^{i-1} (1 - F(x+d))^{n-i-1} f(x) f(x+d) \mathrm{d}x \mathrm{d}d \\
+& = \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} f(x) \int_x^{x+\delta} (1 - F(x+d))^{n-i-1} f(x+d) \mathrm{d}d \mathrm{d}x \\
+& = \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} f(x) \int_{F(x)}^{F(x+\delta)} (1 - u)^{n-i-1} \mathrm{d}u \mathrm{d}x \\
+& = \int_0^{1} \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} f(x) ((1 - F(x))^{n-i} - (1 - F(x + \delta))^{n-i}) \mathrm{d}x \\ 
+& = \int_0^1 g_i(x) \mathrm{d}x - \int_0^1 \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x \\
+& = 1 - \int_0^1 \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x \\
+\end{aligned}
+$$
+
+Because of the $x + \delta$ term, we can't actually evaluate this integral all the way up to $1$, and so we are left with 
+$$
+\begin{aligned}
+& = 1 - \int_0^{1 - \delta} \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x.
+\end{aligned}
+$$
+\end{proof}
+
+Thus if we are able to show that Eq. (\ref{eq:cdf}) converges to 1, we
+are done.
 
 \hypertarget{uniform-case}{%
-\subsection{Uniform case}\label{uniform-case}}
+\subsubsection{Uniform case}\label{uniform-case}}
 
-It can be shown that if \(X_i \stackrel{iid}{\sim} Uniform(0, 1)\), then
-\(D_i \sim Beta(1, n)\). Therefore,
-\(P(\max_i D_i < \delta) \geq (P(D_i < \delta))^n = (1 - (1 - \delta)^n)^n\),
-which is a decreasing function for sufficiently large \(n\). This gives
-us the result \(\max_i D_i \stackrel{a.s.}{\to} 0\).
+\begin{lemma}[Differences between order statistics of a uniform distribution]
+If $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$, then each $D_i \sim \mathrm{Beta}(1, n)$.
+\end{lemma}
+
+\begin{proof}
+We begin with Eq. (\ref{eq:pdf}), plugging in $f(x) = 1$ and $F(x) = x$:
+
+$$
+h_i(d) = \int_0^{1-d} \frac{n!}{(i-1)! (n-i-1)!} x^{i-1} (1-x-d)^{n-i-1} \mathrm{d}x
+$$
+
+Then we proceed with integration by parts, setting 
+$u = x^{i-1} \implies du = (i-1) x^{i-2}$ and 
+$dv = (1-x-d)^{n-i-1} dx \implies v = -\frac{1}{n-i} (1-x-d)^{n-i-1}$. 
+Note that $u v |_0^{1-d} = 0$ in this case. This yields
+
+$$
+= \frac{n!}{(i-1)! (n-i-1)!} \int \frac{i-1}{n-i} x^{i-2} (1-x-d)^{n-i} \mathrm{d}x
+$$
+
+Then applying integration by parts again until the $x^p$ term disappears, we get:
+
+$$
+\begin{aligned}
+& = \frac{n!}{(i-1)! (n-i-1)!} \frac{(i-1)!}{(n-i) \cdots (n-2)} \int_0^{1-d} (1-x-d)^{n-2} \mathrm{d}x \\
+& = -\frac{n (n-1)}{n-1} (1-x-d)^{n-1} \Big|_0^{1-d} \\
+& = n (1 - d)^{n-1}
+\end{aligned}
+$$
+
+This the density function for $\mathrm{Beta}(1, n)$, completing the proof.
+\end{proof}
+
+\begin{theorem}
+Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$. 
+Then for any $\delta > 0$, $P(\max_i X_{(i+1)} - X_{(i)} < \delta) \to 1$ as 
+$n \to \infty$.
+\end{theorem}
+
+\begin{proof}[Proof (sketch)]
+Since $X_{(i+1)} - X_{(i)} = D_i \sim \mathrm{Beta}(1, n)$, 
+$P(X_{(i+1)} - X_{(i)} < \delta) = 1 - (1 - \delta)^n $. This yields
+
+$$
+\begin{aligned}
+P(\max_i D_i < \delta) & \geq (P(D_i < \delta))^{n-1} \\
+& = (1 - (1 - \delta)^n)^{n-1} \\
+& \approx e^{-n \exp(-n \delta)}.
+\end{aligned}
+$$
+
+In the limit $n \to \infty$, this goes to 1.
+\end{proof}
 
 \hypertarget{general-case}{%
-\subsection{General case}\label{general-case}}
+\subsubsection{General case}\label{general-case}}
 
-\hypertarget{proof-sketch-that-pd_i-leq-delta-to-1}{%
-\subsubsection{\texorpdfstring{Proof sketch that
-\(P(D_i \leq \delta) \to 1\)}{Proof sketch that P(D\_i \textbackslash leq \textbackslash delta) \textbackslash to 1}}\label{proof-sketch-that-pd_i-leq-delta-to-1}}
+\begin{theorem}
+Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
+Let $D_i = X_{(i+1)} - X_{(i)}$. 
+Then for any $\epsilon > 0$, there exists $N \in \mathbb{N}$ such that $P(\max_i D_i < \delta) > 1 - \epsilon$ when $n > N$.
+\end{theorem}
+
+\begin{proof}[Proof (sketch)]
+We start with Eq. (\ref{eq:cdf}):
+
+$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x.$$
+
+Making the approximation $F(x+\delta) \approx F(x) + \delta f(x)$ 
+and bounding $f(x) \geq a$, we get:
+
+$$P(D_i \leq \delta) \geq 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x) - a \delta)^{n-i} f(x) \mathrm{d}x.$$
+
+Then making the substitution $u = F(x) \implies du = f(x) \mathrm{d}x$, we obtain
+
+$$1 - \int_0^{F(1-\delta)} \frac{n!}{(n-i)! (i-1)!} u^{i-1} (1 - u - a \delta)^{n-i} \mathrm{d}u$$
+
+Evaluating the integral yields
+
+$$P(D_i < \delta) = 1 - (1 - a \delta)^n + (1 - F(1-\delta) - a \delta)^n.$$
+
+Then as before,
+
+$$
+\begin{aligned}
+P(\max_i D_i < \delta) & \geq (P(D_i < \delta))^{n-1} \\
+& \approx (1 - (1 - a \delta)^n + (1 - F(1-\delta) - a \delta)^n)^{n-1} \\
+& \to 1
+\end{aligned}
+$$
+
+\end{proof}
 
@@
 
-\[P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x + \delta))^{n-i} f(x) dx\]
+\[P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x\]
 
@@
 
+\hypertarget{algorithms}{%
+\section{Algorithms}\label{algorithms}}
+
+TBD
+
 \hypertarget{computational-results}{%

### 2021-09-15 scratch/distributions-on-1d-manifolds.Rmd
removed redundant stuff

@@
 
-If $F$ is absolutely continuous, then it can be shown that
-
-$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \dd x$$
-
-Making the approximation $F(x+\delta) \approx F(x) + \delta f(x)$ 
-and bounding $f(x) \geq a > 0$, we get:
-
-$$P(D_i \leq \delta) \geq 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x) - a \delta)^{n-i} f(x) dx$$
-
-Then making the substitution $u = F(x) \implies du = f(x) dx$, we get
-
-$$1 - \int_0^{F(1-\delta)} \frac{n!}{(n-i+1)! (i-2)!} u^{i-2} (1 - u - a \delta)^{n-i} du$$
-
-Which becomes $1 - C P(n) \delta^{n}$ for some $C > 0$ and $P(n)$ is a polynomial of $n$, giving us the desired result $D_i \stackrel{p}{\to} 0$.
-
 # Algorithms

### 2021-09-15 scratch/distributions-on-1d-manifolds.tex
removed redundant stuff

@@
 
-If \(F\) is absolutely continuous, then it can be shown that
-
-\[P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x\]
-
-Making the approximation \(F(x+\delta) \approx F(x) + \delta f(x)\) and
-bounding \(f(x) \geq a > 0\), we get:
-
-\[P(D_i \leq \delta) \geq 1 - \int_0^{1-\delta} \frac{n!}{(n-i+1)! (i-2)!} (F(x))^{i-2} (1 - F(x) - a \delta)^{n-i} f(x) dx\]
-
-Then making the substitution \(u = F(x) \implies du = f(x) dx\), we get
-
-\[1 - \int_0^{F(1-\delta)} \frac{n!}{(n-i+1)! (i-2)!} u^{i-2} (1 - u - a \delta)^{n-i} du\]
-
-Which becomes \(1 - C P(n) \delta^{n}\) for some \(C > 0\) and \(P(n)\)
-is a polynomial of \(n\), giving us the desired result
-\(D_i \stackrel{p}{\to} 0\).
-
 \hypertarget{algorithms}{%

### 2021-09-17 scratch/distributions-on-1d-manifolds.Rmd
fixed typo
syncing stuff

@@
 1. $g_i(x) = \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x))^{n-i} f(x)$.
-2. $g_{ij}(x) = \frac{n!}{(i-1)! (j-i-1)! (n-j)!} (F(x))^{i-1} (F(y) - F(x))^{j-i-1} (1 - F(y))^{n-j} f(x) f(y)$.
+2. $g_{ij}(x, y) = \frac{n!}{(i-1)! (j-i-1)! (n-j)!} (F(x))^{i-1} (F(y) - F(x))^{j-i-1} (1 - F(y))^{n-j} f(x) f(y)$.
 3. By convolution, $h_i(d) = \int_0^{1} g_{i, i+1} (x, x + d) \dd x$.
@@
 \begin{proof}
-This is just given by setting the integrand as $g_{i, i+1}(x, x+d)$.
-We briefly note that the integral can only be evaluated from 0 to $1-d$ in this case because of the $x+d$ terms.
+This is just a direct consequence of 2 and 3 under the given statements. 
+We also note that because the support of $X_i$ is $[0, 1]$, 
+the integral only needs to be evaluated from $0$ to $1 - d$ 
+because of the $f(x+d)$ and $1 - F(x+d)$ terms.
 \end{proof}

### 2021-09-17 scratch/distributions-on-1d-manifolds.tex
fixed typo
syncing stuff

@@
 \item
-  \(g_{ij}(x) = \frac{n!}{(i-1)! (j-i-1)! (n-j)!} (F(x))^{i-1} (F(y) - F(x))^{j-i-1} (1 - F(y))^{n-j} f(x) f(y)\).
+  \(g_{ij}(x, y) = \frac{n!}{(i-1)! (j-i-1)! (n-j)!} (F(x))^{i-1} (F(y) - F(x))^{j-i-1} (1 - F(y))^{n-j} f(x) f(y)\).
 \item

### 2021-09-29 scratch/distributions-on-1d-manifolds.Rmd
syncing stuff

@@
 - \usepackage{comment}
-bibliography: proposal.bib
+bibliography: misc.bib
 ---
@@
 \newcommand{\Categorical}{\mathrm{Categorical}}
+\newcommand{\Multinomial}{\mathrm{Multinomial}}
+\newcommand{\RDPG}{\mathrm{RDPG}}
+\newcommand{\GRDPG}{\mathrm{GRDPG}}
 \newtheorem{definition}{Definition}
@@
 
-Suppose we have two manifolds $\mathcal{M}_1$ and $\mathcal{M}_2$ $\in \mathbb{R}^d$, each of length 1, defined by $f_1(t)$ and $f_2(t)$ respectively ($f_i : [0, 1] \mapsto \mathbb{R}^d$).
-Define $\delta$ as the minimum distance between the two manifolds, i.e., $\delta = \min_t \|f_1(t) - f_2(t)\|$, and let $\delta > 0$.
-For now, restrict each $f_i$ such that the distance along the manifold between $f_i(t)$ and $f_i(s)$ is equal to the difference between $t$ and $s$ (this also implies that each manifold is of length 1).
-We sample $T_1, ..., T_n \stackrel{iid}{\sim} F$ for continuous $F$ with support $[0, 1]$ and use $f_1$ to map the first $n_1$ points to $\mathcal{M}_1$ and $f_2$ to map the remaining $n_2 = n - n_1$ points to $\mathcal{M}_2$.
-Let $X_i = f_1(T_i)$ and $Y_j = f_2(T_j)$.
-Without loss of generality, assume $n_1 \leq n_2$.
+Suppose we have two manifolds $\mathcal{M}_1$ and $\mathcal{M}_2$ $\in \mathbb{R}^d$, each of length 1, defined by $f_1(t)$ and $f_2(t)$ respectively ($f_k : [0, 1] \mapsto \mathbb{R}^d$). 
+Define $\delta$ as the minimum distance between any two manifolds, i.e., $\delta = \max_{k, \ell} \min_{s, t} \|f_k(s) - f_\ell(t)\|$, and let $\delta > 0$. 
+Restrict each $f_k$ such that the distance along the manifold between $f_k(t)$ and $f_k(s)$ is equal to the difference between $t$ and $s$, i.e., $f_k$ is the arclength parameterization of $\mathcal{M}_k$ (this also implies that each manifold is of length 1).
+Labels are sampled as $Z_1, ..., Z_n \iid \Multinomial(\alpha_1, \alpha_2)$, and we define $n_k$ as the number of labels with label $k$. 
+Sample $T_1, ..., T_n \iid F$ for some distribution $F$ with support $[0, 1]$. 
+Finally, let $X_i = f_{Z_i}(T_i)$ for each $i \in [n]$ be the latent vector of vertex $i$, and gather the latent vectors as $X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}$. 
+We observe $A \sim \RDPG(X)$ (or $A \sim \GRDPG_{p,q}(X)$) and wish to recover $Z_1, ..., Z_n$.
 
@@
 
-Thus if we are able to show that Eq. (\ref{eq:cdf}) converges to 1, we are done.
-
 ### Uniform case
@@
 Let $X_1, ..., X_n \iid \Uniform(0, 1)$. 
-Then for any $\delta > 0$, $P(\max_i X_{(i+1)} - X_{(i)} < \delta) \to 1$ as 
-$n \to \infty$.
+Then for any $\epsilon$ and $\delta > 0$, 
+there exists an $N = O \big(\frac{-\log \epsilon}{\delta} \big)$ such that 
+$P(\max_i X_{(i+1)} - X_{(i)} < \delta) \geq 1 - \epsilon$ when $n > N$.
 \end{theorem}
@@
 Let $D_i = X_{(i+1)} - X_{(i)}$. 
-Then for any $\epsilon > 0$, there exists $N \in \mathbb{N}$ such that $P(\max_i D_i < \delta) > 1 - \epsilon$ when $n > N$.
+Then for any $\epsilon > 0$, there exists $N \in \mathbb{N}$ such that $P(\max_i D_i < \delta) \geq 1 - \epsilon$ when $n > N$.
 \end{theorem}
@@
 \begin{aligned}
-P(\max_i D_i < \delta) & \geq (P(D_i < \delta))^{n-1} \\
-& \approx (1 - (1 - a \delta)^n + (1 - F(1-\delta) - a \delta)^n)^{n-1} \\
-& \to 1
+P(\max_i D_i < \delta) & = P(\text{all } D_i < \delta) \\
+& = 1 - P(\text{some } D_i > \delta) \\
+& \geq 1 - \sum_i^{n-1} P(D_i > \delta) \\
+& = 1 + (n - 1) (1 - a \delta)^n - (n - 1) (1 - F(1 - \delta) - a \delta)^n
 \end{aligned}
@@
 
+This converges to $1$ in the limit $n \to \infty$.
 \end{proof}
+
+## Extension to multidimensional manifolds
+
+Here, we extend the results on the line to the unit hypercube. The following theorem is a direct consequence of Lemma 2 of \citet{trosset2020rehabilitating}.
+
+\begin{theorem}
+Let $X_1, ..., X_n \iid F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support.
+Define $E_n$ as the event that every point on the support is within distance $\delta / 2$ to $X_i$ for $i \in [n]$.
+Then for any $\epsilon > 0$, there exists $N = O(-\log \epsilon)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
+\end{theorem}
 
@@
 TBD
+
+# References

### 2021-09-29 scratch/distributions-on-1d-manifolds.tex
syncing stuff

@@
 \(\in \mathbb{R}^d\), each of length 1, defined by \(f_1(t)\) and
-\(f_2(t)\) respectively (\(f_i : [0, 1] \mapsto \mathbb{R}^d\)). Define
-\(\delta\) as the minimum distance between the two manifolds, i.e.,
-\(\delta = \min_t \|f_1(t) - f_2(t)\|\), and let \(\delta > 0\). For
-now, restrict each \(f_i\) such that the distance along the manifold
-between \(f_i(t)\) and \(f_i(s)\) is equal to the difference between
-\(t\) and \(s\) (this also implies that each manifold is of length 1).
-We sample \(T_1, ..., T_n \stackrel{iid}{\sim} F\) for continuous \(F\)
-with support \([0, 1]\) and use \(f_1\) to map the first \(n_1\) points
-to \(\mathcal{M}_1\) and \(f_2\) to map the remaining \(n_2 = n - n_1\)
-points to \(\mathcal{M}_2\). Let \(X_i = f_1(T_i)\) and
-\(Y_j = f_2(T_j)\). Without loss of generality, assume \(n_1 \leq n_2\).
+\(f_2(t)\) respectively (\(f_k : [0, 1] \mapsto \mathbb{R}^d\)). Define
+\(\delta\) as the minimum distance between any two manifolds, i.e.,
+\(\delta = \max_{k, \ell} \min_{s, t} \|f_k(s) - f_\ell(t)\|\), and let
+\(\delta > 0\). Restrict each \(f_k\) such that the distance along the
+manifold between \(f_k(t)\) and \(f_k(s)\) is equal to the difference
+between \(t\) and \(s\), i.e., \(f_k\) is the arclength parameterization
+of \(\mathcal{M}_k\) (this also implies that each manifold is of length
+1). Labels are sampled as
+\(Z_1, ..., Z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, \alpha_2)\),
+and we define \(n_k\) as the number of labels with label \(k\). Sample
+\(T_1, ..., T_n \stackrel{\mathrm{iid}}{\sim}F\) for some distribution
+\(F\) with support \([0, 1]\). Finally, let \(X_i = f_{Z_i}(T_i)\) for
+each \(i \in [n]\) be the latent vector of vertex \(i\), and gather the
+latent vectors as
+\(X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}\). We observe
+\(A \sim \mathrm{RDPG}(X)\) (or \(A \sim \mathrm{GRDPG}_{p,q}(X)\)) and
+wish to recover \(Z_1, ..., Z_n\).
 
@@
 
-Thus if we are able to show that Eq. (\ref{eq:cdf}) converges to 1, we
-are done.
-
 \hypertarget{uniform-case}{%
@@
 Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$. 
-Then for any $\delta > 0$, $P(\max_i X_{(i+1)} - X_{(i)} < \delta) \to 1$ as 
-$n \to \infty$.
+Then for any $\epsilon$ and $\delta > 0$, 
+there exists an $N = O \big(\frac{-\log \epsilon}{\delta} \big)$ such that 
+$P(\max_i X_{(i+1)} - X_{(i)} < \delta) \geq 1 - \epsilon$ when $n > N$.
 \end{theorem}
@@
 
+This converges to $1$ in the limit $n \to \infty$.
 \end{proof}
+
+\hypertarget{extension-to-multidimensional-manifolds}{%
+\subsection{Extension to multidimensional
+manifolds}\label{extension-to-multidimensional-manifolds}}
+
+Here, we extend the results on the line to the unit hypercube. The
+following theorem is a direct consequence of Lemma 2 of
+\citet{trosset2020rehabilitating}.
+
+\begin{theorem}
+Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support.
+Define $E_n$ as the event that every point on the support is within distance $\delta / 2$ to $X_i$ for $i \in [n]$.
+Then for any $\epsilon > 0$, there exists $N = O(-\log \epsilon)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
+\end{theorem}
 
@@
 
-  \bibliography{proposal.bib}
+\renewcommand\refname{References}
+  \bibliography{misc.bib}
 

### 2021-09-29 scratch/distributions-on-1d-manifolds.Rmd
added multidimensional case

@@
 \begin{theorem}
+\ref{thm:generalized}
 Let $X_1, ..., X_n \iid F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
@@
 & \geq 1 - \sum_i^{n-1} P(D_i > \delta) \\
-& = 1 + (n - 1) (1 - a \delta)^n - (n - 1) (1 - F(1 - \delta) - a \delta)^n
+& = 1 - (n - 1) (1 - a \delta)^n + (n - 1) (1 - F(1 - \delta) - a \delta)^n
 \end{aligned}
@@
 This converges to $1$ in the limit $n \to \infty$.
+
+We can also approximate $F(1 - \delta) \approx 1 - a \delta$, which yields 
+$1 - (n - 1) (1 - a \delta)^n$. Setting this $\geq 1 - \epsilon$ ...
 \end{proof}
@@
 \begin{theorem}
+\ref{thm:multidim}
 Let $X_1, ..., X_n \iid F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 
-Define $E_n$ as the event that every point on the support is within distance $\delta / 2$ to $X_i$ for $i \in [n]$.
-Then for any $\epsilon > 0$, there exists $N = O(-\log \epsilon)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
+Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
+Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
 \end{theorem}
 
+\begin{proof}[Proof (sketch)]
+Divide the hypercube $[0, 1]^r$ into a grid of sub-hypercubes of side length at most $\eta / \sqrt{r}$. 
+$E_n$ is satisfied if each sub-hypercube contains at least one $X_i$ from the sample.
+
+$$
+\begin{aligned}
+P(E_n) & = 1 - P(\text{some cells don't contain } X_i) \\
+& \geq 1 - \sum_k^{\lceil \sqrt{r} / \eta \rceil} \prod_i^n P(X_i \text{ is not in the } k^{th} \text{ hypercube}) \\
+& \geq 1 - \lceil \sqrt{r} / \eta \rceil (1 - a \eta^r / r^{r/2})^n
+\end{aligned}
+$$
+
+Setting this $\geq 1 - \epsilon$ and solving for $n$ yields the desired rate. 
+\end{proof}
+
 # Algorithms
-
-TBD
 

### 2021-09-29 scratch/distributions-on-1d-manifolds.tex
added multidimensional case

@@
 \begin{theorem}
+\ref{thm:generalized}
 Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
@@
 \begin{theorem}
+\ref{thm:multidim}
 Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 
-Define $E_n$ as the event that every point on the support is within distance $\delta / 2$ to $X_i$ for $i \in [n]$.
-Then for any $\epsilon > 0$, there exists $N = O(-\log \epsilon)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
+Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
+Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
 \end{theorem}
+
+\begin{proof}[Proof (sketch)]
+Divide the hypercube $[0, 1]^r$ into a grid of sub-hypercubes of side length at most $\eta / \sqrt{r}$. 
+$E_n$ is satisfied if each sub-hypercube contains at least one $X_i$ from the sample.
+
+$$
+\begin{aligned}
+P(E_n) & = 1 - P(\text{some cells don't contain } X_i) \\
+& \geq 1 - \sum_k^{\lceil \sqrt{r} / \eta \rceil} \prod_i^n P(X_i \text{ is not in the } k^{th} \text{ hypercube}) \\
+& \geq 1 - \lceil \sqrt{r} / \eta \rceil (1 - a \eta^r / r^{r/2})^n
+\end{aligned}
+$$
+
+Setting this $\geq 1 - \epsilon$ and solving for $n$ yields the desired rate. 
+\end{proof}
 
@@
 \section{Algorithms}\label{algorithms}}
-
-TBD
 

### 2021-09-29 scratch/distributions-on-1d-manifolds.Rmd
typo

@@
 \begin{theorem}
-\ref{thm:generalized}
+\label{thm:generalized}
 Let $X_1, ..., X_n \iid F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
@@
 \begin{theorem}
-\ref{thm:multidim}
+\label{thm:multidim}
 Let $X_1, ..., X_n \iid F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 

### 2021-09-29 scratch/distributions-on-1d-manifolds.tex
typo

@@
 \begin{theorem}
-\ref{thm:generalized}
+\label{thm:generalized}
 Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
@@
 \begin{theorem}
-\ref{thm:multidim}
+\label{thm:multidim}
 Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 

### 2021-09-29 scratch/distributions-on-1d-manifolds.Rmd
typo

@@
 Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
-Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
+Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta^r / r^{r/2}}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
 \end{theorem}
@@
 P(E_n) & = 1 - P(\text{some cells don't contain } X_i) \\
-& \geq 1 - \sum_k^{\lceil \sqrt{r} / \eta \rceil} \prod_i^n P(X_i \text{ is not in the } k^{th} \text{ hypercube}) \\
-& \geq 1 - \lceil \sqrt{r} / \eta \rceil (1 - a \eta^r / r^{r/2})^n
+& \geq 1 - \sum_k^{\lceil \sqrt{r} / \eta \rceil^r} \prod_i^n P(X_i \text{ is not in the } k^{th} \text{ hypercube}) \\
+& \geq 1 - \lceil \sqrt{r} / \eta \rceil^r (1 - a \eta^r / r^{r/2})^n
 \end{aligned}
@@
 
+TBD
+
 # Computational Results

### 2021-09-29 scratch/distributions-on-1d-manifolds.tex
typo

@@
 \section{Algorithms}\label{algorithms}}
+
+TBD
 

### 2021-10-06 scratch/distributions-on-1d-manifolds.Rmd
updated writeup

@@
 
-### General case
+### Test for intersecting manifolds
+
+\begin{theorem}
+Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n_1$ and $n_2$ points drawn uniformly along them ($n = n_1 + n_2$).
+
+Suppose we want to perform a hypothesis test for whether the ends of the manifolds meet. 
+The $H_0$ is the case where $\delta = 0$ and $H_1$ is the case where $\delta > 0$. 
+Define the test statistic $T_n = \|X_{(n_1)} - Y_{(1)}\|$ (or similar, depending on which ends meet). 
+Then $T_n / 2 \stackrel{d}{\to} \Betadist(1, n)$ under $H_0$. 
+\end{theorem}
+
+## General case
 
@@
 
+### Community detection for manifolds that emanate from the origin
+
 ## Extension to multidimensional manifolds

### 2021-10-06 scratch/distributions-on-1d-manifolds.tex
updated writeup

@@
 
+\hypertarget{test-for-intersecting-manifolds}{%
+\subsubsection{Test for intersecting
+manifolds}\label{test-for-intersecting-manifolds}}
+
+\begin{theorem}
+Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n_1$ and $n_2$ points drawn uniformly along them ($n = n_1 + n_2$).
+
+Suppose we want to perform a hypothesis test for whether the ends of the manifolds meet. 
+The $H_0$ is the case where $\delta = 0$ and $H_1$ is the case where $\delta > 0$. 
+Define the test statistic $T_n = \|X_{(n_1)} - Y_{(1)}\|$ (or similar, depending on which ends meet). 
+Then $T_n / 2 \stackrel{d}{\to} \mathrm{Beta}(1, n)$ under $H_0$. 
+\end{theorem}
+
 \hypertarget{general-case}{%
-\subsubsection{General case}\label{general-case}}
+\subsection{General case}\label{general-case}}
 
@@
 \end{proof}
+
+\hypertarget{community-detection-for-manifolds-that-emanate-from-the-origin}{%
+\subsubsection{Community detection for manifolds that emanate from the
+origin}\label{community-detection-for-manifolds-that-emanate-from-the-origin}}
 

### 2021-10-13 scratch/distributions-on-1d-manifolds.Rmd
updated document

--- 2021-09-29 scratch/distributions-on-1d-manifolds.Rmd
@@
 ---
+
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(echo = FALSE, 
+                      comment = NA, 
+                      warning = FALSE, 
+                      message = FALSE, 
+                      fig.pos = 'H', 
+                      fig.align = 'center', 
+                      fig.height = 4, 
+                      fig.width = 4)
+options(xtable.comment = FALSE, 
+        xtable.table.placement = 'H')
+```
 
@@
 
-### General case
+### Test for intersecting manifolds
+
+\begin{theorem}
+Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n / 2$ points drawn uniformly along each manifold.
+
+Suppose we want to perform a hypothesis test for whether the ends of the manifolds meet. 
+The $H_0$ is the case where $\delta = 0$ and $H_1$ is the case where $\delta > 0$. 
+We define the test statistic $T_n$ as follows: 
+Construct an $\epsilon$-neighborhood graph using the smallest $\epsilon$ required to make the graph connected. 
+Let $\{d_i\}$ be the set of edge lengths of this graph. 
+Define $T_n = \max_i d_i$. 
+Then $T_n / 2 \stackrel{d}{\to} \Betadist(1, n)$ under $H_0$. 
+\end{theorem}
+
+This theorem isn't 100\% correct---in reality, we want to draw edges along the "super-manifold" that is created by joining the two manifolds (under $H_0$).
+But if $H_0$ is true, then these lengths should converge to the true distances along the super-manifold.
+
+### Community detection for "non-affine" one-dimensional manifolds
+
+For non-intersecting manifolds, distance-based graph techniques provide a straightforward method for community detection. 
+However, this is not the case in many models with community structure. 
+For example, the latent structure that induces the degree-corrected block model consists of line segments that emanate from the origin, so each community is a one-dimensional subspace, or linear manifold, and these manifolds meet at the origin.
+For the DCBM, a straightforward approach to community detection involves an ASE followed by clustering based on angles (e.g., cosine similarity). 
+As an extension to this case, we consider community detection for two one-dimensional manifolds that meet at the origin.
+
+The method that is proposed is as follows:
+
+1. For each point $x_i$, compute $\|x_i\|$.
+2. Construct an $\epsilon$-neighborhood graph for $X$, choosing $\epsilon$ large enough that the graph is connected. 
+Alternatively, construct a $K$-NN graph, again, choosing $K$ so that the graph is connected. 
+3. Remove the vertex $v_j$ that corresponds to the point $x_j = \min_i \|x_i\|$. 
+If this results in two disconnected subgraphs, we are done. 
+Otherwise, continue removing vertices in ascending order of magnitude until the graph is disconnected. 
+Then map each component to a community.
+4. Estimate curves for each community (e.g., using polynomial regression). 
+Map each removed point to a community based on its distance to the curve.
+
+In the following example, we draw $64$ points each uniformly along two manifolds in $\mathbb{R}^2$ that intersect at the origin:
+
+```{r}
+import::from(magrittr, `%>%`)
+
+source('~/dev/pabm-grdpg/functions.R')
+source('https://mtrosset.pages.iu.edu/Courses/675/graph.r')
+
+f1 <- function(t) {
+  x1 <- t / sqrt(.5 + sqrt(5) / 2)
+  x2 <- x1 ** 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- t / (.5 + sqrt(5) / 2)
+  x2 <- x1 ** .5
+  return(cbind(x1, x2))
+}
+
+set.seed(1)
+```
+
+```{r}
+n1 <- 2 ** 6
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+plot(X, col = z, xlim = c(0, 1), ylim = c(0, 1))
+
+r <- apply(X, 1, function(x) sum(x ** 2))
+order.r <- order(r)
+```
+
+The $5$-NN graph is as follows:
+
+```{r}
+W <- X %>% 
+  dist() %>% 
+  as.matrix() %>% 
+  graph.knn(5)
+W <- W + t(W)
+W <- sign(W)
+diag(W) <- 0
+
+qgraph::qgraph(W, groups = as.factor(z))
+```
+
+Removing the 10 points closest to the origin results in the following:
+
+```{r}
+remove <- which(order.r <= 10)
+```
+
+```{r}
+W <- W[-remove, -remove]
+qgraph::qgraph(W, groups = as.factor(z[-remove]))
+```
+
+## General case
 

### 2021-10-13 scratch/distributions-on-1d-manifolds.tex
updated document

@@
 \begin{theorem}
-Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n_1$ and $n_2$ points drawn uniformly along them ($n = n_1 + n_2$).
+Let $\mathcal{M}_1$ and $\mathcal{M}_2$ be two one-dimensional manifolds of unit length with $n / 2$ points drawn uniformly along each manifold.
 
@@
 The $H_0$ is the case where $\delta = 0$ and $H_1$ is the case where $\delta > 0$. 
-Define the test statistic $T_n = \|X_{(n_1)} - Y_{(1)}\|$ (or similar, depending on which ends meet). 
+We define the test statistic $T_n$ as follows: 
+Construct an $\epsilon$-neighborhood graph using the smallest $\epsilon$ required to make the graph connected. 
+Let $\{d_i\}$ be the set of edge lengths of this graph. 
+Define $T_n = \max_i d_i$. 
 Then $T_n / 2 \stackrel{d}{\to} \mathrm{Beta}(1, n)$ under $H_0$. 
 \end{theorem}
+
+This theorem isn't 100\% correct---in reality, we want to draw edges
+along the ``super-manifold'' that is created by joining the two
+manifolds (under \(H_0\)). But if \(H_0\) is true, then these lengths
+should converge to the true distances along the super-manifold.
+
+\hypertarget{community-detection-for-non-affine-one-dimensional-manifolds}{%
+\subsubsection{Community detection for ``non-affine'' one-dimensional
+manifolds}\label{community-detection-for-non-affine-one-dimensional-manifolds}}
+
+For non-intersecting manifolds, distance-based graph techniques provide
+a straightforward method for community detection. However, this is not
+the case in many models with community structure. For example, the
+latent structure that induces the degree-corrected block model consists
+of line segments that emanate from the origin, so each community is a
+one-dimensional subspace, or linear manifold, and these manifolds meet
+at the origin. For the DCBM, a straightforward approach to community
+detection involves an ASE followed by clustering based on angles (e.g.,
+cosine similarity). As an extension to this case, we consider community
+detection for two one-dimensional manifolds that meet at the origin.
+
+The method that is proposed is as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  For each point \(x_i\), compute \(\|x_i\|\).
+\item
+  Construct an \(\epsilon\)-neighborhood graph for \(X\), choosing
+  \(\epsilon\) large enough that the graph is connected. Alternatively,
+  construct a \(K\)-NN graph, again, choosing \(K\) so that the graph is
+  connected.
+\item
+  Remove the vertex \(v_j\) that corresponds to the point
+  \(x_j = \min_i \|x_i\|\). If this results in two disconnected
+  subgraphs, we are done. Otherwise, continue removing vertices in
+  ascending order of magnitude until the graph is disconnected. Then map
+  each component to a community.
+\item
+  Estimate curves for each community (e.g., using polynomial
+  regression). Map each removed point to a community based on its
+  distance to the curve.
+\end{enumerate}
+
+In the following example, we draw \(64\) points each uniformly along two
+manifolds in \(\mathbb{R}^2\) that intersect at the origin:
+
+\begin{center}\includegraphics{distributions-on-1d-manifolds_files/figure-latex/unnamed-chunk-2-1} \end{center}
+
+The \(5\)-NN graph is as follows:
+
+\begin{center}\includegraphics{distributions-on-1d-manifolds_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+Removing the 10 points closest to the origin results in the following:
+
+\begin{center}\includegraphics{distributions-on-1d-manifolds_files/figure-latex/unnamed-chunk-5-1} \end{center}
 
@@
 
-\hypertarget{community-detection-for-manifolds-that-emanate-from-the-origin}{%
-\subsubsection{Community detection for manifolds that emanate from the
-origin}\label{community-detection-for-manifolds-that-emanate-from-the-origin}}
-
 \hypertarget{extension-to-multidimensional-manifolds}{%

### 2021-10-20 curve-fitting-methods/curve-fitting-scratch.Rmd
tried a new method

---
title: |
  | Semi-Parametric Manifold Clustering
output:
  pdf_document:
    keep_tex: true
    citation_package: natbib
fontsize: 11pt
# geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
bibliography: misc.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 3, 
                      fig.width = 3)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')

library(ggplot2)
import::from(magrittr, `%>%`)
```

```{r}
source('~/dev/pabm-grdpg/functions.R')
source('https://mtrosset.pages.iu.edu/Courses/675/graph.r')

laplacian.eigenmap <- function(W, d = 2, normalized = TRUE) {
  n <- nrow(W)
  if (normalized) {
    L <- normalized.laplacian(W)
  } else {
    L <- graph.laplacian(W)
  }
  L.eigen <- eigen(L, symmetric = TRUE)
  sweep(L.eigen$vectors[, seq(n - 1, n - d)], 2, 
        sqrt(L.eigen$values[seq(n - 1, n - d)]), `/`)
}

spectral.clustering <- function(W, K = 2, d = K, normalized = TRUE) {
  Y <- laplacian.eigenmap(W, d, normalized)
  kmeans(Y, K)$cluster
}
```

\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Categorical}{\mathrm{Categorical}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{example}
\newtheorem{example}{Example}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}

# Estimating Polynomial Curves

## Problem Setup

Let:

* $T_1, ..., T_n \iid F$ with support $[0, 1]$. 
* $g(\cdot, \theta) : [0, 1] \mapsto \mathcal{X} \subset \mathbb{R}^d$. 
* $X_1, ..., X_n = g(T_1), ..., g(T_n)$

Assuming some parametric form of $g$ with parameters $\theta$, we want to find $\hat{\theta}$, some "reasonable" estimate for $\theta$. 
We observe $X_i$ but not $T_i$.

For now, we limit $d = 2$ and $g$ to quadratic functions.

```{r}
param.curve.quadr.2 <- function(t., 
                                beta10, beta11, beta12, 
                                beta20, beta21, beta22) {
  x1 <- beta10 + beta11 * t. + beta12 * t. ^ 2
  x2 <- beta20 + beta21 * t. + beta22 * t. ^ 2
  return(cbind(x1, x2))
}

estimate.one.t.quadr.2 <- function(x1, x2, 
                                   beta10, beta11, beta12, 
                                   beta20, beta21, beta22) {
  gamma0 <- beta10 * beta11 + beta20 * beta21 - beta11 * x1 - beta21 * x2
  gamma1 <- 2 * beta10 * beta12 + 2 * beta20 * beta22 + 
    beta11 ^ 2 + beta21 ^ 2 - 
    2 * beta12 * x1 - 2 * beta22 * x2
  gamma2 <- 3 * beta11 * beta12 + 3 * beta21 * beta22
  gamma3 <- 2 * (beta12 ^ 2 + beta22 ^ 2)
  roots <- polyroot(c(gamma0, gamma1, gamma2, gamma3))
  roots <- roots[abs(roots) == abs(Re(roots))]
  roots <- Re(roots)
  return(max(roots))
}

estimate.t.quadr.2 <- function(X, 
                               beta10, beta11, beta12, 
                               beta20, beta21, beta22) {
  apply(X, 1, function(x) {
    estimate.one.t.quadr.2(x[1], x[2], 
                           beta10, beta11, beta12, 
                           beta20, beta21, beta22)
  })
}

estimate.quadr.coefs.2 <- function(X, t.hat, intercept = TRUE) {
  sum.X <- colSums(X)
  sum.Xt <- colSums(sweep(X, 1, t.hat, `*`))
  sum.Xt2 <- colSums(sweep(X, 1, t.hat ^ 2, `*`))
  sum.t <- sum(t.hat)
  sum.t2 <- sum(t.hat ^ 2)
  sum.t3 <- sum(t.hat ^ 3)
  sum.t4 <- sum(t.hat ^ 4)

  if (intercept) {
    A.sub <- matrix(c(1, sum.t, sum.t2, 
                      sum.t, sum.t2, sum.t3, 
                      sum.t2, sum.t3, sum.t4), 
                    nrow = 3, ncol = 3)
    A <- Matrix::bdiag(A.sub, A.sub)

    b <- c(sum.X[1], sum.Xt[1], sum.Xt2[1], 
           sum.X[2], sum.Xt[2], sum.Xt2[2])
    return(solve(A, b))
  } else {
    A.sub <- matrix(c(sum.t2, sum.t3, sum.t3, sum.t4),
                    nrow = 2, ncol = 2)
    A <- Matrix::bdiag(A.sub, A.sub)
    b <- c(sum.Xt[1], sum.Xt2[1], sum.Xt[2], sum.Xt2[2])
    theta <- solve(A, b)
    return(c(0, theta[1:2], 0, theta[3:4]))
  }
}

estimate.quadr.curve.2 <- function(X, init.params, 
                                   intercept = TRUE,
                                   eps = 1e-6, maxit = 100) {
  if (missing(init.params)) {
    theta <- rnorm(6)
  } else {
    theta <- init.params
  }
  d.theta <- 1 / eps

  niter <- 0
  while (d.theta > eps) {
    theta.prev <- theta
    t.hat <- estimate.t.quadr.2(X,
                                theta[1], theta[2], theta[3],
                                theta[4], theta[5], theta[6])
    X.hat <- param.curve.quadr.2(t.hat, theta[1], theta[2], theta[3], 
                                 theta[4], theta[5], theta[6])
    theta <- estimate.quadr.coefs.2(X, t.hat, intercept)

    d.theta <- sum((theta - theta.prev) ^ 2) / sum(theta ^ 2)

    niter <- niter + 1
    if (niter >= maxit) {
      warning('failed to converge to a quadratic curve')
      break
    }
  }
  return(list(theta = theta, 
              X = X.hat,
              t = t.hat))
}
```

\begin{example}

Let $g(t) = (t^2, 2 t (1 - t) ) = (t^2, 2 t - 2 t^2)$. 
(This is the first two dimensions of the Hardy-Weinberg curve). 
Then $\theta = (0, 0, 1, 0, 2, -2)$.

```{r}
t. <- seq(0, 1, .01)
X <- param.curve.quadr.2(t., 0, 0, 1, 0, 2, -2)

ggplot() + 
  geom_line(aes(x = X[, 1], y = X[, 2])) + 
  theme_bw() + 
  labs(x = expression(x[1]), y = expression(x[2])) + 
  coord_fixed()
```

\end{example}

If we observe the $T_i$'s, then we can use a standard polynomial regression method to obtain $\hat{\theta}$. Since we do not observe them, the proposed iterative method is as follows:

1. Initialize $\hat{\theta}^{(0)}$ (e.g., by drawing from a probability distribution). 
2. Estimate each $\hat{t}_i^{(s)}$ by minimizing $L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2$.
3. Compute each $\hat{x}_i^{(s)} = g(\hat{t}_i^{(s)} | \hat{\theta}^{(s)})$
4. Estimate $\hat{\theta}^{(s+1)}$ by minimizing $L(\{\hat{t}_i^{(s)}\}, \theta | X) = \sum_i \|x_i - g(\hat{t}_i^{(s)} | \theta)\|^2$.
5. Repeat steps 2-4 until convergence.

If we restrict $g$ to be polynomials, then steps (2) and (4) have closed-form solutions.

\begin{example}

Write $g(t | \theta) = (g_1(t | \theta_1), ..., g_d(t | \theta_d))$ where $g_r(t | \theta_r)$ is the component of $g$ in the $r^{th}$ dimension and $\theta_r$ is the vector of parameters for the $r^{th}$ dimension. If $g_r$ are polynomials of degree $p$, then each $\theta_r$ contains up to $p + 1$ entries.

Given the observed points $x_1, ..., x_n \in \mathbb{R}^d$ and their corresponding index points $t_1, ..., t_n \in \mathbb{R}$, we can find each $\hat{\theta}_r$ individually by $\hat{\theta}_r = A^{-1} b$ where $b \in \mathbb{R}^{p+1}$ and $b_k = \sum_i x_i t_i^k$ and $A \in \mathbb{R}^{(p+1) \times (p+1)}$ and $A_{kl} = \sum_i t^{(k-1) (l-1)}$.

On the other hand, if we have parameters $\theta$ but not the index points $t_i$, we can minimize each $t_i$ individually by finding the roots of a $p+1$ polynomial with coefficients that depend on $x_1, ..., x_n$ and $\theta$.

In the following plot, we drew $n = 200$ points from the 2D H-W curve with $T_1, ..., T_n \iid Uniform(0, 1)$. 
The red line is the curve that was fit using the above method.

\end{example}

```{r}
set.seed(314159)
n <- 200
t. <- runif(n)
X <- param.curve.quadr.2(t., 0, 0, 1, 0, 2, -2)

out <- estimate.quadr.curve.2(X)
thetahat <- out$theta
t.hat <- sort(out$t)
Xhat <- param.curve.quadr.2(t.hat, 
                            thetahat[1], thetahat[2], thetahat[3],
                            thetahat[4], thetahat[5], thetahat[6])

ggplot() + 
  geom_point(aes(x = X[, 1], y = X[, 2])) + 
  theme_bw() + 
  labs(x = expression(x[1]), y = expression(x[2])) + 
  coord_fixed() + 
  geom_line(aes(x = Xhat[, 1], y = Xhat[, 2]),
            colour = 'red')
```

One problem with this method is the parameterization of the curve is not unique.

\begin{example}

In the next example, we draw $A \sim \RDPG(X)$ using the same H-W curve and sample size as above and estimate the true latent positions (up to rotation). 
In this example, we force the intercept terms to be zero.

\end{example}

```{r}
P <- X %*% t(X)
diag(P) <- 0
A <- draw.graph(P)
Y <- embedding(A, 2, 0)
```

```{r, cache = TRUE}
out <- estimate.quadr.curve.2(Y,
                              init.params = c(0, 0, 1, 0, 2, -2),
                              intercept = FALSE,
                              maxit = 200)
thetahat <- out$theta
t.hat <- sort(out$t)
Yhat <- param.curve.quadr.2(t.hat, 
                            thetahat[1], thetahat[2], thetahat[3],
                            thetahat[4], thetahat[5], thetahat[6])

ggplot() + 
  geom_point(aes(x = Y[, 1], y = Y[, 2])) + 
  theme_bw() + 
  labs(x = expression(x[1]), y = expression(x[2])) + 
  coord_fixed() + 
  geom_line(aes(x = Yhat[, 1], y = Yhat[, 2]),
            colour = 'red')
```

Next, suppose we have K curves parameterized by $g^{(k)}$, with points drawn along these curves. Then one possible clustering technique is as follows:

1. Assign an initial clustering (e.g., via spectral clustering).
2. Estimate the curve for each cluster.
3. Reassign the clusters by proximity to each cluster.
4. Repeat 2 and 3 until convergence.

\begin{example}

We again limit these to be quadratic functions in $\mathbb{R}^2$. 
Here, $K = 2$ and $n_1 = n_2 = 256$.

\end{example}

```{r}
compute.distances.quadr.2 <- function(X, theta) {
  t. <- estimate.t.quadr.2(X, 
                           theta[1], theta[2], theta[3], 
                           theta[4], theta[5], theta[6])
  X.hat <- param.curve.quadr.2(t., 
                               theta[1], theta[2], theta[3], 
                               theta[4], theta[5], theta[6])
  apply(X - X.hat, 1, function(x) sum(x ^ 2))
}

manifold.clustering.quadr.2 <- function(A, K = 2, 
                                        intercept = TRUE, 
                                        maxit = 100) {
  n <- nrow(A)

  z.hat <- spectral.clustering(A, K)
  X <- embedding(A, 2, 0)
  z.hat.prev <- sample(seq(K), n, replace = TRUE)

  niter <- 0

  while (!all(z.hat.prev == z.hat)) {
    z.hat.prev <- z.hat

    X1 <- X[z.hat == 1, ]
    X2 <- X[z.hat == 2, ]

    curve1 <- estimate.quadr.curve.2(X1, intercept = intercept)
    curve2 <- estimate.quadr.curve.2(X2, intercept = intercept)

    d1 <- compute.distances.quadr.2(X, curve1$theta)
    d2 <- compute.distances.quadr.2(X, curve2$theta)
    distances <- cbind(d1, d2)

    z.hat <- apply(distances, 1, which.min)

    niter <- niter + 1
    if (niter >= maxit) {
      warning('failed to converge to a clustering')
      break
    }
  }

  return(list(z = z.hat, 
              X = X,
              theta = list(curve1$theta, curve2$theta)))
}
```

```{r}
f1 <- function(t) {
  x1 <- t / sqrt(.5 + sqrt(5) / 2)
  x2 <- x1 ** 2
  return(cbind(x1, x2))
}

f2 <- function(t) {
  x2 <- t / sqrt(.5 + sqrt(5) / 2)
  x1 <- x2 ** 2
  return(cbind(x1, x2))
}

n1 <- n2 <- 2 ** 8
z <- c(rep(1, n1), rep(2, n2))
t1 <- sort(runif(n1))
t2 <- sort(runif(n2))
X1 <- f1(t1)
X2 <- f2(t2)
X <- rbind(X1, X2)

P <- X %*% t(X)
Xhat <- embedding(P, 2, 0)

manifold.out <- manifold.clustering.quadr.2(P, 2, intercept = FALSE)
zhat.manifold <- manifold.out$z
if (mean(z == zhat.manifold) < .5) {
  zhat.manifold <- 2 - zhat.manifold
}
t. <- seq(0, 1, .01)
theta1 <- manifold.out$theta[[1]]
theta2 <- manifold.out$theta[[2]]
Y1 <- param.curve.quadr.2(t., 
                          theta1[1], theta1[2], theta1[3], 
                          theta1[4], theta1[5], theta1[6])
Y2 <- param.curve.quadr.2(t.,
                          theta2[1], theta2[2], theta2[3],
                          theta2[4], theta2[5], theta2[6])
Y <- rbind(Y1, Y2)
z. <- c(rep(1, length(t.)),
        rep(2, length(t.)))

ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw() + 
  geom_line(aes(x = Y[, 1], y = Y[, 2],
                groups = factor(z.))) + 
  xlim(-1, 0) + ylim(-.5, .5) + 
  coord_fixed()
```

\begin{example}

Finally, we draw $A \sim \RDPG(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.

\end{example}

```{r cache = TRUE, fig.width = 4, fig.height = 4}
set.seed(12345)

A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)

manifold.out <- manifold.clustering.quadr.2(A, 2, intercept = FALSE)
zhat.manifold <- manifold.out$z
table(z, zhat.manifold)

plot(Xhat, col = z, asp = 1)
t. <- seq(-1, 1, .01)
theta1 <- manifold.out$theta[[1]]
theta2 <- manifold.out$theta[[2]]
Y1 <- param.curve.quadr.2(t., 
                          theta1[1], theta1[2], theta1[3], 
                          theta1[4], theta1[5], theta1[6])
Y2 <- param.curve.quadr.2(t.,
                          theta2[1], theta2[2], theta2[3],
                          theta2[4], theta2[5], theta2[6])
lines(Y1)
lines(Y2, col = 2)
```

### 2021-10-20 curve-fitting-methods/curve-fitting-scratch.tex
tried a new method

--- 2021-09-29 scratch/distributions-on-1d-manifolds.tex
@@
 
-\title{Clustering of Distributions on 1-Dimensional Manifolds}
+\title{Semi-Parametric Manifold Clustering}
 \author{}
@@
 \theoremstyle{example}
-\newtheorem*{example}{Example}
+\newtheorem{example}{Example}
 \newcommand{\dd}{\mathrm{d}}
@@
 
-\hypertarget{setup}{%
-\section{Setup}\label{setup}}
+\hypertarget{estimating-polynomial-curves}{%
+\section{Estimating Polynomial
+Curves}\label{estimating-polynomial-curves}}
 
-Suppose we have two manifolds \(\mathcal{M}_1\) and \(\mathcal{M}_2\)
-\(\in \mathbb{R}^d\), each of length 1, defined by \(f_1(t)\) and
-\(f_2(t)\) respectively (\(f_k : [0, 1] \mapsto \mathbb{R}^d\)). Define
-\(\delta\) as the minimum distance between any two manifolds, i.e.,
-\(\delta = \max_{k, \ell} \min_{s, t} \|f_k(s) - f_\ell(t)\|\), and let
-\(\delta > 0\). Restrict each \(f_k\) such that the distance along the
-manifold between \(f_k(t)\) and \(f_k(s)\) is equal to the difference
-between \(t\) and \(s\), i.e., \(f_k\) is the arclength parameterization
-of \(\mathcal{M}_k\) (this also implies that each manifold is of length
-1). Labels are sampled as
-\(Z_1, ..., Z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, \alpha_2)\),
-and we define \(n_k\) as the number of labels with label \(k\). Sample
-\(T_1, ..., T_n \stackrel{\mathrm{iid}}{\sim}F\) for some distribution
-\(F\) with support \([0, 1]\). Finally, let \(X_i = f_{Z_i}(T_i)\) for
-each \(i \in [n]\) be the latent vector of vertex \(i\), and gather the
-latent vectors as
-\(X = \begin{bmatrix} X_1 & \cdots & X_n \end{bmatrix}\). We observe
-\(A \sim \mathrm{RDPG}(X)\) (or \(A \sim \mathrm{GRDPG}_{p,q}(X)\)) and
-wish to recover \(Z_1, ..., Z_n\).
+\hypertarget{problem-setup}{%
+\subsection{Problem Setup}\label{problem-setup}}
 
-\hypertarget{preliminary-theory}{%
-\section{Preliminary Theory}\label{preliminary-theory}}
+Let:
 
-\hypertarget{distributions-of-differences-of-order-statistics}{%
-\subsection{Distributions of differences of order
-statistics}\label{distributions-of-differences-of-order-statistics}}
+\begin{itemize}
+\tightlist
+\item
+  \(T_1, ..., T_n \stackrel{\mathrm{iid}}{\sim}F\) with support
+  \([0, 1]\).
+\item
+  \(g(\cdot, \theta) : [0, 1] \mapsto \mathcal{X} \subset \mathbb{R}^d\).
+\item
+  \(X_1, ..., X_n = g(T_1), ..., g(T_n)\)
+\end{itemize}
 
-Let \(D_i = X_{(i+1)} - X_{(i)}\). Then if \(\max_i D_i < \delta\), we
-have sufficient separation of points in \(\mathcal{M}_1\). Then it is
-sufficient to quantify \(P(\max_i D_i > \delta)\) as a function of \(n\)
-and \(\delta\) and show that this converges to zero as \(n\) grows to
-\(\infty\).
+Assuming some parametric form of \(g\) with parameters \(\theta\), we
+want to find \(\hat{\theta}\), some ``reasonable'' estimate for
+\(\theta\). We observe \(X_i\) but not \(T_i\).
 
-We denote \(f(x)\) as the density of each \(X_i\), \(g_i(x)\) as the
-density of \(X_{(i)}\), \(g_{ij}(x, y)\) as the joint density of
-\(X_{(i)}, X_{(j)}\), and \(h_i(d)\) as the density of \(D_i\) (with
-corresponding capital letters for the cumulative distribution
-functions).
+For now, we limit \(d = 2\) and \(g\) to quadratic functions.
 
-The following are taken as given\footnote{\url{https://en.wikipedia.org/wiki/Order_statistic}}:
+\begin{example}
+
+Let $g(t) = (t^2, 2 t (1 - t) ) = (t^2, 2 t - 2 t^2)$. 
+(This is the first two dimensions of the Hardy-Weinberg curve). 
+Then $\theta = (0, 0, 1, 0, 2, -2)$.
+
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\end{example}
+
+If we observe the \(T_i\)'s, then we can use a standard polynomial
+regression method to obtain \(\hat{\theta}\). Since we do not observe
+them, the proposed iterative method is as follows:
 
@@
 \item
-  \(g_i(x) = \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x))^{n-i} f(x)\).
+  Initialize \(\hat{\theta}^{(0)}\) (e.g., by drawing from a probability
+  distribution).
 \item
-  \(g_{ij}(x, y) = \frac{n!}{(i-1)! (j-i-1)! (n-j)!} (F(x))^{i-1} (F(y) - F(x))^{j-i-1} (1 - F(y))^{n-j} f(x) f(y)\).
+  Estimate each \(\hat{t}_i^{(s)}\) by minimizing
+  \(L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2\).
 \item
-  By convolution,
-  \(h_i(d) = \int_0^{1} g_{i, i+1} (x, x + d) \mathrm{d}x\).
+  Compute each
+  \(\hat{x}_i^{(s)} = g(\hat{t}_i^{(s)} | \hat{\theta}^{(s)})\)
+\item
+  Estimate \(\hat{\theta}^{(s+1)}\) by minimizing
+  \(L(\{\hat{t}_i^{(s)}\}, \theta | X) = \sum_i \|x_i - g(\hat{t}_i^{(s)} | \theta)\|^2\).
+\item
+  Repeat steps 2-4 until convergence.
 \end{enumerate}
 
-\begin{lemma}[The probability density function of $D_i$]
-\begin{equation}
-\label{eq:pdf}
-h_i(d) = \int_0^{1-d} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} (1 - F(x+d))^{n-i-1} f(x) f(x+d) \mathrm{d}x 
-\end{equation}
-\end{lemma}
+If we restrict \(g\) to be polynomials, then steps (2) and (4) have
+closed-form solutions.
 
-\begin{proof}
-This is just a direct consequence of 2 and 3 under the given statements. 
-We also note that because the support of $X_i$ is $[0, 1]$, 
-the integral only needs to be evaluated from $0$ to $1 - d$ 
-because of the $f(x+d)$ and $1 - F(x+d)$ terms.
-\end{proof}
+\begin{example}
 
-\begin{lemma}[The cumulative distribution function of $D_i$]
+Write $g(t | \theta) = (g_1(t | \theta_1), ..., g_d(t | \theta_d))$ where $g_r(t | \theta_r)$ is the component of $g$ in the $r^{th}$ dimension and $\theta_r$ is the vector of parameters for the $r^{th}$ dimension.
+If $g_r$ are polynomials of degree $p$, then each $\theta_r$ contains up to $p + 1$ entries.
 
-\begin{equation}
-\label{eq:cdf}
-P(D_i < \delta) = H_i(\delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x
-\end{equation}
-\end{lemma}
+Given the observed points $x_1, ..., x_n \in \mathbb{R}^d$ and their corresponding index points $t_1, ..., t_n \in \mathbb{R}$, we can find each $\hat{\theta}_r$ individually by $\hat{\theta}_r = A^{-1} b$ where $b \in \mathbb{R}^{p+1}$ and $b_k = \sum_i x_i t_i^k$ and $A \in \mathbb{R}^{(p+1) \times (p+1)}$ and $A_{kl} = \sum_i t^{(k-1) (l-1)}$.
 
-\begin{proof}
-$$
-\begin{aligned}
-H_i(\delta) & = \int_x^{x+\delta} h_i(d) \mathrm{d}d \\
-& = \int_x^{x+\delta} \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} ((F(x))^{i-1} (1 - F(x+d))^{n-i-1} f(x) f(x+d) \mathrm{d}x \mathrm{d}d \\
-& = \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} f(x) \int_x^{x+\delta} (1 - F(x+d))^{n-i-1} f(x+d) \mathrm{d}d \mathrm{d}x \\
-& = \int_0^{1} \frac{n!}{(i-1)! (n-i-1)!} (F(x))^{i-1} f(x) \int_{F(x)}^{F(x+\delta)} (1 - u)^{n-i-1} \mathrm{d}u \mathrm{d}x \\
-& = \int_0^{1} \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} f(x) ((1 - F(x))^{n-i} - (1 - F(x + \delta))^{n-i}) \mathrm{d}x \\ 
-& = \int_0^1 g_i(x) \mathrm{d}x - \int_0^1 \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x \\
-& = 1 - \int_0^1 \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x \\
-\end{aligned}
-$$
+On the other hand, if we have parameters $\theta$ but not the index points $t_i$, we can minimize each $t_i$ individually by finding the roots of a $p+1$ polynomial with coefficients that depend on $x_1, ..., x_n$ and $\theta$.
 
-Because of the $x + \delta$ term, we can't actually evaluate this integral all the way up to $1$, and so we are left with 
-$$
-\begin{aligned}
-& = 1 - \int_0^{1 - \delta} \frac{n!}{(i-1)! (n-i)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x.
-\end{aligned}
-$$
-\end{proof}
+In the following plot, we drew $n = 200$ points from the 2D H-W curve with $T_1, ..., T_n \stackrel{\mathrm{iid}}{\sim}Uniform(0, 1)$. 
+The red line is the curve that was fit using the above method.
 
-\hypertarget{uniform-case}{%
-\subsubsection{Uniform case}\label{uniform-case}}
+\end{example}
 
-\begin{lemma}[Differences between order statistics of a uniform distribution]
-If $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$, then each $D_i \sim \mathrm{Beta}(1, n)$.
-\end{lemma}
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-4-1} \end{center}
 
-\begin{proof}
-We begin with Eq. (\ref{eq:pdf}), plugging in $f(x) = 1$ and $F(x) = x$:
+One problem with this method is the parameterization of the curve is not
+unique.
 
-$$
-h_i(d) = \int_0^{1-d} \frac{n!}{(i-1)! (n-i-1)!} x^{i-1} (1-x-d)^{n-i-1} \mathrm{d}x
-$$
+\begin{example}
 
-Then we proceed with integration by parts, setting 
-$u = x^{i-1} \implies du = (i-1) x^{i-2}$ and 
-$dv = (1-x-d)^{n-i-1} dx \implies v = -\frac{1}{n-i} (1-x-d)^{n-i-1}$. 
-Note that $u v |_0^{1-d} = 0$ in this case. This yields
+In the next example, we draw $A \sim \mathrm{RDPG}(X)$ using the same H-W curve and sample size as above and estimate the true latent positions (up to rotation). 
+In this example, we force the intercept terms to be zero.
 
-$$
-= \frac{n!}{(i-1)! (n-i-1)!} \int \frac{i-1}{n-i} x^{i-2} (1-x-d)^{n-i} \mathrm{d}x
-$$
+\end{example}
 
-Then applying integration by parts again until the $x^p$ term disappears, we get:
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-6-1} \end{center}
 
-$$
-\begin{aligned}
-& = \frac{n!}{(i-1)! (n-i-1)!} \frac{(i-1)!}{(n-i) \cdots (n-2)} \int_0^{1-d} (1-x-d)^{n-2} \mathrm{d}x \\
-& = -\frac{n (n-1)}{n-1} (1-x-d)^{n-1} \Big|_0^{1-d} \\
-& = n (1 - d)^{n-1}
-\end{aligned}
-$$
+Next, suppose we have K curves parameterized by \(g^{(k)}\), with points
+drawn along these curves. Then one possible clustering technique is as
+follows:
 
-This the density function for $\mathrm{Beta}(1, n)$, completing the proof.
-\end{proof}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Assign an initial clustering (e.g., via spectral clustering).
+\item
+  Estimate the curve for each cluster.
+\item
+  Reassign the clusters by proximity to each cluster.
+\item
+  Repeat 2 and 3 until convergence.
+\end{enumerate}
 
-\begin{theorem}
-Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$. 
-Then for any $\epsilon$ and $\delta > 0$, 
-there exists an $N = O \big(\frac{-\log \epsilon}{\delta} \big)$ such that 
-$P(\max_i X_{(i+1)} - X_{(i)} < \delta) \geq 1 - \epsilon$ when $n > N$.
-\end{theorem}
+\begin{example}
 
-\begin{proof}[Proof (sketch)]
-Since $X_{(i+1)} - X_{(i)} = D_i \sim \mathrm{Beta}(1, n)$, 
-$P(X_{(i+1)} - X_{(i)} < \delta) = 1 - (1 - \delta)^n $. This yields
+We again limit these to be quadratic functions in $\mathbb{R}^2$. 
+Here, $K = 2$ and $n_1 = n_2 = 256$.
 
-$$
-\begin{aligned}
-P(\max_i D_i < \delta) & \geq (P(D_i < \delta))^{n-1} \\
-& = (1 - (1 - \delta)^n)^{n-1} \\
-& \approx e^{-n \exp(-n \delta)}.
-\end{aligned}
-$$
+\end{example}
 
-In the limit $n \to \infty$, this goes to 1.
-\end{proof}
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-8-1} \end{center}
 
-\hypertarget{general-case}{%
-\subsubsection{General case}\label{general-case}}
+\begin{example}
 
-\begin{theorem}
-\label{thm:generalized}
-Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]$, and suppose $f(x)$ is continuous and $f(x) \geq a > 0$ everywhere on the support. 
-Let $D_i = X_{(i+1)} - X_{(i)}$. 
-Then for any $\epsilon > 0$, there exists $N \in \mathbb{N}$ such that $P(\max_i D_i < \delta) \geq 1 - \epsilon$ when $n > N$.
-\end{theorem}
+Finally, we draw $A \sim \mathrm{RDPG}(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
 
-\begin{proof}[Proof (sketch)]
-We start with Eq. (\ref{eq:cdf}):
+\end{example}
 
-$$P(D_i \leq \delta) = 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x + \delta))^{n-i} f(x) \mathrm{d}x.$$
+\begin{verbatim}
+   zhat.manifold
+z     1   2
+  1 187  69
+  2  87 169
+\end{verbatim}
 
-Making the approximation $F(x+\delta) \approx F(x) + \delta f(x)$ 
-and bounding $f(x) \geq a$, we get:
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-9-1} \end{center}
 
-$$P(D_i \leq \delta) \geq 1 - \int_0^{1-\delta} \frac{n!}{(n-i)! (i-1)!} (F(x))^{i-1} (1 - F(x) - a \delta)^{n-i} f(x) \mathrm{d}x.$$
-
-Then making the substitution $u = F(x) \implies du = f(x) \mathrm{d}x$, we obtain
-
-$$1 - \int_0^{F(1-\delta)} \frac{n!}{(n-i)! (i-1)!} u^{i-1} (1 - u - a \delta)^{n-i} \mathrm{d}u$$
-
-Evaluating the integral yields
-
-$$P(D_i < \delta) = 1 - (1 - a \delta)^n + (1 - F(1-\delta) - a \delta)^n.$$
-
-Then as before,
-
-$$
-\begin{aligned}
-P(\max_i D_i < \delta) & = P(\text{all } D_i < \delta) \\
-& = 1 - P(\text{some } D_i > \delta) \\
-& \geq 1 - \sum_i^{n-1} P(D_i > \delta) \\
-& = 1 - (n - 1) (1 - a \delta)^n + (n - 1) (1 - F(1 - \delta) - a \delta)^n
-\end{aligned}
-$$
-
-This converges to $1$ in the limit $n \to \infty$.
-
-We can also approximate $F(1 - \delta) \approx 1 - a \delta$, which yields 
-$1 - (n - 1) (1 - a \delta)^n$. Setting this $\geq 1 - \epsilon$ ...
-\end{proof}
-
-\hypertarget{extension-to-multidimensional-manifolds}{%
-\subsection{Extension to multidimensional
-manifolds}\label{extension-to-multidimensional-manifolds}}
-
-Here, we extend the results on the line to the unit hypercube. The
-following theorem is a direct consequence of Lemma 2 of
-\citet{trosset2020rehabilitating}.
-
-\begin{theorem}
-\label{thm:multidim}
-Let $X_1, ..., X_n \stackrel{\mathrm{iid}}{\sim}F$ with support $[0, 1]^r$, and $f(x) \geq a > 0$ everywhere on the support. 
-Define $E_n$ as the event that an $\eta$-neighborhood graph constructed from the sample is connected. 
-Then for any $\epsilon > 0$, there exists $N = O \bigg( \frac{\log \epsilon \eta^r / r^{r/2}}{\log (1 - \frac{a \eta^r}{r^{r / 2}})} \bigg)$ such that $P(E_n) > 1 - \epsilon$ when $n \geq N$.
-\end{theorem}
-
-\begin{proof}[Proof (sketch)]
-Divide the hypercube $[0, 1]^r$ into a grid of sub-hypercubes of side length at most $\eta / \sqrt{r}$. 
-$E_n$ is satisfied if each sub-hypercube contains at least one $X_i$ from the sample.
-
-$$
-\begin{aligned}
-P(E_n) & = 1 - P(\text{some cells don't contain } X_i) \\
-& \geq 1 - \sum_k^{\lceil \sqrt{r} / \eta \rceil^r} \prod_i^n P(X_i \text{ is not in the } k^{th} \text{ hypercube}) \\
-& \geq 1 - \lceil \sqrt{r} / \eta \rceil^r (1 - a \eta^r / r^{r/2})^n
-\end{aligned}
-$$
-
-Setting this $\geq 1 - \epsilon$ and solving for $n$ yields the desired rate. 
-\end{proof}
-
-\hypertarget{algorithms}{%
-\section{Algorithms}\label{algorithms}}
-
-TBD
-
-\hypertarget{computational-results}{%
-\section{Computational Results}\label{computational-results}}
-
-TBD
-
-\renewcommand\refname{References}
   \bibliography{misc.bib}

### 2021-10-27 curve-fitting-methods/curve-fitting-scratch.Rmd
added headers
typo
updated writeup so far

@@
 import::from(magrittr, `%>%`)
-```
-
-```{r}
-source('~/dev/pabm-grdpg/functions.R')
-source('https://mtrosset.pages.iu.edu/Courses/675/graph.r')
-
-laplacian.eigenmap <- function(W, d = 2, normalized = TRUE) {
-  n <- nrow(W)
-  if (normalized) {
-    L <- normalized.laplacian(W)
-  } else {
-    L <- graph.laplacian(W)
-  }
-  L.eigen <- eigen(L, symmetric = TRUE)
-  sweep(L.eigen$vectors[, seq(n - 1, n - d)], 2, 
-        sqrt(L.eigen$values[seq(n - 1, n - d)]), `/`)
-}
-
-spectral.clustering <- function(W, K = 2, d = K, normalized = TRUE) {
-  Y <- laplacian.eigenmap(W, d, normalized)
-  kmeans(Y, K)$cluster
-}
 ```
@@
 ```{r}
-param.curve.quadr.2 <- function(t., 
-                                beta10, beta11, beta12, 
-                                beta20, beta21, beta22) {
-  x1 <- beta10 + beta11 * t. + beta12 * t. ^ 2
-  x2 <- beta20 + beta21 * t. + beta22 * t. ^ 2
-  return(cbind(x1, x2))
-}
-
-estimate.one.t.quadr.2 <- function(x1, x2, 
-                                   beta10, beta11, beta12, 
-                                   beta20, beta21, beta22) {
-  gamma0 <- beta10 * beta11 + beta20 * beta21 - beta11 * x1 - beta21 * x2
-  gamma1 <- 2 * beta10 * beta12 + 2 * beta20 * beta22 + 
-    beta11 ^ 2 + beta21 ^ 2 - 
-    2 * beta12 * x1 - 2 * beta22 * x2
-  gamma2 <- 3 * beta11 * beta12 + 3 * beta21 * beta22
-  gamma3 <- 2 * (beta12 ^ 2 + beta22 ^ 2)
-  roots <- polyroot(c(gamma0, gamma1, gamma2, gamma3))
-  roots <- roots[abs(roots) == abs(Re(roots))]
-  roots <- Re(roots)
-  return(max(roots))
-}
-
-estimate.t.quadr.2 <- function(X, 
-                               beta10, beta11, beta12, 
-                               beta20, beta21, beta22) {
-  apply(X, 1, function(x) {
-    estimate.one.t.quadr.2(x[1], x[2], 
-                           beta10, beta11, beta12, 
-                           beta20, beta21, beta22)
-  })
-}
-
-estimate.quadr.coefs.2 <- function(X, t.hat, intercept = TRUE) {
-  sum.X <- colSums(X)
-  sum.Xt <- colSums(sweep(X, 1, t.hat, `*`))
-  sum.Xt2 <- colSums(sweep(X, 1, t.hat ^ 2, `*`))
-  sum.t <- sum(t.hat)
-  sum.t2 <- sum(t.hat ^ 2)
-  sum.t3 <- sum(t.hat ^ 3)
-  sum.t4 <- sum(t.hat ^ 4)
-
-  if (intercept) {
-    A.sub <- matrix(c(1, sum.t, sum.t2, 
-                      sum.t, sum.t2, sum.t3, 
-                      sum.t2, sum.t3, sum.t4), 
-                    nrow = 3, ncol = 3)
-    A <- Matrix::bdiag(A.sub, A.sub)
-
-    b <- c(sum.X[1], sum.Xt[1], sum.Xt2[1], 
-           sum.X[2], sum.Xt[2], sum.Xt2[2])
-    return(solve(A, b))
-  } else {
-    A.sub <- matrix(c(sum.t2, sum.t3, sum.t3, sum.t4),
-                    nrow = 2, ncol = 2)
-    A <- Matrix::bdiag(A.sub, A.sub)
-    b <- c(sum.Xt[1], sum.Xt2[1], sum.Xt[2], sum.Xt2[2])
-    theta <- solve(A, b)
-    return(c(0, theta[1:2], 0, theta[3:4]))
-  }
-}
-
-estimate.quadr.curve.2 <- function(X, init.params, 
-                                   intercept = TRUE,
-                                   eps = 1e-6, maxit = 100) {
-  if (missing(init.params)) {
-    theta <- rnorm(6)
-  } else {
-    theta <- init.params
-  }
-  d.theta <- 1 / eps
-
-  niter <- 0
-  while (d.theta > eps) {
-    theta.prev <- theta
-    t.hat <- estimate.t.quadr.2(X,
-                                theta[1], theta[2], theta[3],
-                                theta[4], theta[5], theta[6])
-    X.hat <- param.curve.quadr.2(t.hat, theta[1], theta[2], theta[3], 
-                                 theta[4], theta[5], theta[6])
-    theta <- estimate.quadr.coefs.2(X, t.hat, intercept)
-
-    d.theta <- sum((theta - theta.prev) ^ 2) / sum(theta ^ 2)
-
-    niter <- niter + 1
-    if (niter >= maxit) {
-      warning('failed to converge to a quadratic curve')
-      break
-    }
-  }
-  return(list(theta = theta, 
-              X = X.hat,
-              t = t.hat))
-}
+source('~/dev/pabm-grdpg/functions.R')
+source('~/dev/dissertation/curve-fitting-methods/functions.R')
 ```
@@
 
-Let $g(t) = (t^2, 2 t (1 - t) ) = (t^2, 2 t - 2 t^2)$. 
+Let $g(t) = (t^2, 2 t (1 - t) ) = (0 + 0 t + t^2, 0 + 2 t - 2 t^2)$. 
 (This is the first two dimensions of the Hardy-Weinberg curve). 
@@
 
-1. Initialize $\hat{\theta}^{(0)}$ (e.g., by drawing from a probability distribution). 
+1. Initialize $\hat{\theta}^{(0)}$ (e.g., randomly). 
 2. Estimate each $\hat{t}_i^{(s)}$ by minimizing $L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2$.
@@
 If we restrict $g$ to be polynomials, then steps (2) and (4) have closed-form solutions. 
+Alternatively, we can estimate $g$ using more general forms, e.g., splines, which may require approximation.
 
@@
 
+# Estimation with Noise
+
 \begin{example}
@@
 In the next example, we draw $A \sim \RDPG(X)$ using the same H-W curve and sample size as above and estimate the true latent positions (up to rotation).
-In this example, we force the intercept terms to be zero.
 
@@
 ```{r, cache = TRUE}
-out <- estimate.quadr.curve.2(Y,
+out1 <- estimate.quadr.curve.2(Y,
                               init.params = c(0, 0, 1, 0, 2, -2),
-                              intercept = FALSE,
+                              intercept = TRUE,
                               maxit = 200)
-thetahat <- out$theta
-t.hat <- sort(out$t)
+thetahat <- out1$theta
+t.hat <- sort(out1$t)
 Yhat <- param.curve.quadr.2(t.hat, 
@@
 
+# Clustering
+
 Next, suppose we have K curves parameterized by $g^{(k)}$, with points drawn along these curves. Then one possible clustering technique is as follows:
@@
 2. Estimate the curve for each cluster.
-3. Reassign the clusters by proximity to each cluster.
+3. Reassign the clusters by proximity to each curve.
 4. Repeat 2 and 3 until convergence.
@@
 Here, $K = 2$ and $n_1 = n_2 = 256$.
-
-\end{example}
-
-```{r}
-compute.distances.quadr.2 <- function(X, theta) {
-  t. <- estimate.t.quadr.2(X, 
-                           theta[1], theta[2], theta[3], 
-                           theta[4], theta[5], theta[6])
-  X.hat <- param.curve.quadr.2(t., 
-                               theta[1], theta[2], theta[3], 
-                               theta[4], theta[5], theta[6])
-  apply(X - X.hat, 1, function(x) sum(x ^ 2))
-}
-
-manifold.clustering.quadr.2 <- function(A, K = 2, 
-                                        intercept = TRUE, 
-                                        maxit = 100) {
-  n <- nrow(A)
-
-  z.hat <- spectral.clustering(A, K)
-  X <- embedding(A, 2, 0)
-  z.hat.prev <- sample(seq(K), n, replace = TRUE)
-
-  niter <- 0
-
-  while (!all(z.hat.prev == z.hat)) {
-    z.hat.prev <- z.hat
-
-    X1 <- X[z.hat == 1, ]
-    X2 <- X[z.hat == 2, ]
-
-    curve1 <- estimate.quadr.curve.2(X1, intercept = intercept)
-    curve2 <- estimate.quadr.curve.2(X2, intercept = intercept)
-
-    d1 <- compute.distances.quadr.2(X, curve1$theta)
-    d2 <- compute.distances.quadr.2(X, curve2$theta)
-    distances <- cbind(d1, d2)
-
-    z.hat <- apply(distances, 1, which.min)
-
-    niter <- niter + 1
-    if (niter >= maxit) {
-      warning('failed to converge to a clustering')
-      break
-    }
-  }
-
-  return(list(z = z.hat, 
-              X = X,
-              theta = list(curve1$theta, curve2$theta)))
-}
-```
 
@@
 
-manifold.out <- manifold.clustering.quadr.2(P, 2, intercept = FALSE)
+manifold.out <- manifold.clustering.quadr.2(P, 2, 
+                                            intercept = FALSE, 
+                                            method = 'polynomial')
 zhat.manifold <- manifold.out$z
@@
   theme_bw() + 
-  geom_line(aes(x = Y[, 1], y = Y[, 2],
-                groups = factor(z.))) + 
+  # geom_line(aes(x = Y[, 1], y = Y[, 2],
+  #               groups = factor(z.))) + 
   xlim(-1, 0) + ylim(-.5, .5) + 
@@
 
-\begin{example}
+Next, we draw $A \sim \RDPG(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
+The community detection error in this example is 30\%.
+In the following plot, the points are colored according to their true labels.
 
-Finally, we draw $A \sim \RDPG(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
-
-\end{example}
-
-```{r cache = TRUE, fig.width = 4, fig.height = 4}
+```{r cache = TRUE, fig.width = 3, fig.height = 3}
 set.seed(12345)
@@
 
-manifold.out <- manifold.clustering.quadr.2(A, 2, intercept = FALSE)
+manifold.out <- manifold.clustering.quadr.2(A, 2, 
+                                            intercept = FALSE,
+                                            method = 'polynomial')
 zhat.manifold <- manifold.out$z
-table(z, zhat.manifold)
+# table(z, zhat.manifold)
 
-plot(Xhat, col = z, asp = 1)
+# plot(Xhat, col = z, asp = 1)
 t. <- seq(-1, 1, .01)
@@
                           theta2[4], theta2[5], theta2[6])
-lines(Y1)
-lines(Y2, col = 2)
+Y <- rbind(Y1, Y2)
+z. <- c(rep(1, length(t.)),
+        rep(2, length(t.)))
+# lines(Y1)
+# lines(Y2, col = 2)
+
+ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  geom_path(aes(x = Y[, 1], y = Y[, 2],
+                groups = factor(z.))) + 
+  xlim(-.25, 1.25) + ylim(-.75, .75) + 
+  coord_fixed()
 ```
+
+\end{example}
+
+A possibly more robust modification to this is to use Bezier curves for $g$. 
+This is the same functional form as the polynomial curves we used before, but with different basis functions. 
+For the following examples, we consider the quadratic Bezier curve, which has the following basis form:
+
+$$
+g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
+$$
+
+Where as before, $g : [0, 1] \mapsto \mathbb{R}^d$ and $t \in [0, 1]$, and each $p_r \in \mathbb{R}^d$. Thus if we can fit each $p_r$, then the procedure is the same as before.
+
+Common methods for Bezier curve fitting do not require specification of $t_1, ..., t_n$. However, an ordering is required. The curve fitting method is then modified as follows:
+
+1. Initialize $\hat{\theta}^{(0)}$ (e.g., randomly). 
+2. Estimate each $\hat{t}_i^{(s)}$ by minimizing $L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2$.
+3. Reorder $X$ according to $t_1, ..., t_n$.
+4. Estimate $\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}$ for the reordered $X$ using Bezier curve fitting.
+5. Compute $\hat{\theta}^{(s+1)}$ from $\{\hat{p}_r^{(s)}\}$.
+6. Repeat steps 2-5 until convergence.
+
+\begin{example}
+
+```{r}
+f1 <- function(t) {
+  x1 <- t ^ 2
+  x2 <- 2 * t - 2 * t ^ 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- 2 * t - 2 * t ^ 2
+  x2 <- 1 - 2 * t + t ^ 2
+  return(cbind(x1, x2))
+}
+```
+
+We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$.
+$n_1 = n_2 = 128$ points are drawn uniformly from each.
+
+```{r}
+n1 <- 2 ** 7
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+T <- rbind(t1, t2)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+
+ggplot() + 
+  geom_point(aes(x = X[, 1], y = X[, 2], colour = factor(z))) + 
+  coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw()
+```
+
+We draw $A \sim \RDPG(X)$ and obtain the following ASE:
+
+```{r}
+P <- X %*% t(X)
+diag(P) <- 0
+A <- draw.graph(P)
+Xhat <- embedding(A, 2, 0)
+if (mean(Xhat[, 1]) < 0) Xhat[, 1] <- -Xhat[, 1]
+
+ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
+  coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw()
+```
+
+Fitting two quadratic Bezier curves to these data yields a community detection error rate of 10\%. 
+In the following plot, the points are labeled according to their estimated labels.
+
+```{r cache = TRUE}
+set.seed(314159265)
+manifold.out <- manifold.clustering.quadr.2(A, 2, 
+                                            # initialization = z,
+                                            # initialization = 'spectral',
+                                            initialization = 'random', 
+                                            method = 'bezier')
+zhat.manifold <- manifold.out$z
+
+t. <- seq(-1, 1, .01)
+theta1 <- manifold.out$theta[[1]]
+theta2 <- manifold.out$theta[[2]]
+Y1 <- param.curve.quadr.2(t., 
+                          theta1[1], theta1[2], theta1[3], 
+                          theta1[4], theta1[5], theta1[6])
+Y2 <- param.curve.quadr.2(t.,
+                          theta2[1], theta2[2], theta2[3],
+                          theta2[4], theta2[5], theta2[6])
+Y <- rbind(Y1, Y2)
+z. <- c(rep(1, length(t.)),
+        rep(2, length(t.)))
+
+ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(zhat.manifold))) + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  geom_path(aes(x = Y[, 1], y = Y[, 2],
+                groups = factor(z.))) + 
+  xlim(-.25, 1.25) + ylim(-.75, .75) + 
+  coord_fixed()
+```
+
+\end{example}

### 2021-10-27 curve-fitting-methods/curve-fitting-scratch.tex
added headers
typo
updated writeup so far

@@
 
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-2-1} \end{center}
 
@@
 \item
-  Initialize \(\hat{\theta}^{(0)}\) (e.g., by drawing from a probability
-  distribution).
+  Initialize \(\hat{\theta}^{(0)}\) (e.g., randomly).
 \item
@@
 If we restrict \(g\) to be polynomials, then steps (2) and (4) have
-closed-form solutions.
+closed-form solutions. Alternatively, we can estimate \(g\) using more
+general forms, e.g., splines, which may require approximation.
 
@@
 
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-4-1} \end{center}
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-3-1} \end{center}
 
@@
 
+\hypertarget{estimation-with-noise}{%
+\section{Estimation with Noise}\label{estimation-with-noise}}
+
 \begin{example}
@@
 In the next example, we draw $A \sim \mathrm{RDPG}(X)$ using the same H-W curve and sample size as above and estimate the true latent positions (up to rotation).
-In this example, we force the intercept terms to be zero.
 
@@
 
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-5-1} \end{center}
+
+\hypertarget{clustering}{%
+\section{Clustering}\label{clustering}}
 
@@
 \item
-  Reassign the clusters by proximity to each cluster.
+  Reassign the clusters by proximity to each curve.
 \item
@@
 
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-6-1} \end{center}
+
+Next, we draw $A \sim \mathrm{RDPG}(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
+The community detection error in this example is 30\%.
+In the following plot, the points are colored according to their true labels.
+
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-7-1} \end{center}
+
 \end{example}
 
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-8-1} \end{center}
+A possibly more robust modification to this is to use Bezier curves for
+\(g\). This is the same functional form as the polynomial curves we used
+before, but with different basis functions. For the following examples,
+we consider the quadratic Bezier curve, which has the following basis
+form:
+
+\[
+g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
+\]
+
+Where as before, \(g : [0, 1] \mapsto \mathbb{R}^d\) and
+\(t \in [0, 1]\), and each \(p_r \in \mathbb{R}^d\). Thus if we can fit
+each \(p_r\), then the procedure is the same as before.
+
+Common methods for Bezier curve fitting do not require specification of
+\(t_1, ..., t_n\). However, an ordering is required. The curve fitting
+method is then modified as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Initialize \(\hat{\theta}^{(0)}\) (e.g., randomly).
+\item
+  Estimate each \(\hat{t}_i^{(s)}\) by minimizing
+  \(L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2\).
+\item
+  Reorder \(X\) according to \(t_1, ..., t_n\).
+\item
+  Estimate \(\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}\) for the
+  reordered \(X\) using Bezier curve fitting.
+\item
+  Compute \(\hat{\theta}^{(s+1)}\) from \(\{\hat{p}_r^{(s)}\}\).
+\item
+  Repeat steps 2-5 until convergence.
+\end{enumerate}
 
@@
 
-Finally, we draw $A \sim \mathrm{RDPG}(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
+We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$.
+$n_1 = n_2 = 128$ points are drawn uniformly from each.
+
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-9-1} \end{center}
+
+We draw $A \sim \mathrm{RDPG}(X)$ and obtain the following ASE:
+
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-10-1} \end{center}
+
+Fitting two quadratic Bezier curves to these data yields a community detection error rate of 10\%. 
+In the following plot, the points are labeled according to their estimated labels.
+
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-11-1} \end{center}
 
 \end{example}
-
-\begin{verbatim}
-   zhat.manifold
-z     1   2
-  1 187  69
-  2  87 169
-\end{verbatim}
-
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-9-1} \end{center}
 

### 2021-10-28 curve-fitting-methods/souvenir-pless.R
adding some experimental/scratch stuff

@@ -1,0 +1,104 @@
+import::from(magrittr, `%>%`)
+library(ggplot2)
+
+source('https://mtrosset.pages.iu.edu/Courses/675/graph.r')
+source('https://mtrosset.pages.iu.edu/Courses/675/stress.r')
+source('~/dev/pabm-grdpg/functions.R')
+source('~/dev/dissertation/curve-fitting-methods/functions.R')
+
+weighted.mds <- function(D, d = 2, w = rep(1, ncol(D))) {
+  n <- ncol(D)
+  e <- rep(1, n)
+  H <- diag(n) - tcrossprod(e, w)
+  B <- -H %*% D %*% H / 2
+  eigen.B <- eigen(B, symmetric = TRUE)
+  Y <- sweep(eigen.B$vectors[, seq(d)], 2, sqrt(eigen.B$values[seq(d)]), `*`)
+  Y <- Y / (sum(abs(eigen.B$values)) / sum(eigen.B$values[seq(d)]))
+  return(Y)
+}
+
+compute.dist2.manifold <- function(Dg, De, w) {
+  w.sum <- sum(w)
+  (rowSums(sweep(Dg - De, 2, w, `*`)) / w.sum) ^ 2
+}
+
+update.weights <- function(D2, sigma2 = 1e-1) {
+  W <- exp(-D2 / sigma2) %>% 
+    sweep(., 1, rowSums(.), `/`) %>% 
+    return()
+}
+
+f1 <- function(t) {
+  x1 <- t ^ 2
+  x2 <- 2 * t - 2 * t ^ 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- 2 * t - 2 * t ^ 2
+  x2 <- 1 - 2 * t + t ^ 2
+  return(cbind(x1, x2))
+}
+
+n1 <- 2 ** 7
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+T <- rbind(t1, t2)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+plot(X, col = z, xlim = c(0, 1), ylim = c(0, 1), asp = 1)
+
+P <- X %*% t(X)
+diag(P) <- 0
+
[…]

### 2021-11-03 curve-fitting-methods/curve-fitting-scratch.Rmd
minor updates

@@
 $$
-g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
+g(t | p) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
 $$
 
-Where as before, $g : [0, 1] \mapsto \mathbb{R}^d$ and $t \in [0, 1]$, and each $p_r \in \mathbb{R}^d$. Thus if we can fit each $p_r$, then the procedure is the same as before.
+Where as before, $g : [0, 1] \mapsto \mathbb{R}^d$ and $t \in [0, 1]$, and each $p_r \in \mathbb{R}^d$. 
+Thus if we can fit each $p_r$, then the procedure is the same as before.
 
-Common methods for Bezier curve fitting do not require specification of $t_1, ..., t_n$. However, an ordering is required. The curve fitting method is then modified as follows:
+Common methods for Bezier curve fitting do not require specification of $t_1, ..., t_n$. 
+However, an ordering is required.
 
-1. Initialize $\hat{\theta}^{(0)}$ (e.g., randomly). 
-2. Estimate each $\hat{t}_i^{(s)}$ by minimizing $L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2$.
+The curve fitting method is then modified as follows:
+
+1. Initialize $\hat{p}^{(0)}$ (e.g., randomly). 
+2. Estimate each $\hat{t}_i^{(s)}$ by minimizing $L(t_i, \hat{p}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{p}^{(s)})\|^2$.
 3. Reorder $X$ according to $t_1, ..., t_n$.
-4. Estimate $\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}$ for the reordered $X$ using Bezier curve fitting.
-5. Compute $\hat{\theta}^{(s+1)}$ from $\{\hat{p}_r^{(s)}\}$.
-6. Repeat steps 2-5 until convergence.
+4. Estimate $\hat{p}^{(s)}$ for the reordered $X$ using Bezier curve fitting.
+5. Repeat steps 2-4 until convergence.
+
+Then this becomes step 2 of the clustering method outlined above.
+
+We can also estimate $\hat{p} = (T^\top T)^{-1} T^\top X$ where $T = \begin{bmatrix} t_1 & \cdots & t_R \end{bmatrix}$ and each $t_r = \binom{R}{r} (1-t)^{R-r} t^r$, where $R$ is the order of the Bezier curve, so using the method for estimating $t_1, ..., t_n$ as described before, we can iteratively solve for $\hat{p}$ in the same manner.
+If we use soft clustering (e.g., EM as opposed to $K$-means), we can also insert a diagonal matrix of weights $W_k$ to obtain
+
+$$\hat{p}_k = (T^\top W_k T)^{-1} T^\top W_k X$$
+
+where $X$ is the full data matrix, $T$ is the full model matrix, and $W_k = w_k$ with the entries of $w_k$ being the estimated probability of each $x_i$ being in cluster $k$.
 

### 2021-11-03 curve-fitting-methods/curve-fitting-scratch.tex
minor updates

@@
 \[
-g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
+g(t | p) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
 \]
@@
 Common methods for Bezier curve fitting do not require specification of
-\(t_1, ..., t_n\). However, an ordering is required. The curve fitting
-method is then modified as follows:
+\(t_1, ..., t_n\). However, an ordering is required.
+
+The curve fitting method is then modified as follows:
 
@@
 \item
-  Initialize \(\hat{\theta}^{(0)}\) (e.g., randomly).
+  Initialize \(\hat{p}^{(0)}\) (e.g., randomly).
 \item
   Estimate each \(\hat{t}_i^{(s)}\) by minimizing
-  \(L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2\).
+  \(L(t_i, \hat{p}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{p}^{(s)})\|^2\).
 \item
@@
 \item
-  Estimate \(\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}\) for the
-  reordered \(X\) using Bezier curve fitting.
+  Estimate \(\hat{p}^{(s)}\) for the reordered \(X\) using Bezier curve
+  fitting.
 \item
-  Compute \(\hat{\theta}^{(s+1)}\) from \(\{\hat{p}_r^{(s)}\}\).
-\item
-  Repeat steps 2-5 until convergence.
+  Repeat steps 2-4 until convergence.
 \end{enumerate}
+
+Then this becomes step 2 of the clustering method outlined above.
+
+We can also estimate \(\hat{p} = (T^\top T)^{-1} T^\top X\) where
+\(T = \begin{bmatrix} t_1 & \cdots & t_R \end{bmatrix}\) and each
+\(t_r = \binom{R}{r} (1-t)^{R-r} t^r\), where \(R\) is the order of the
+Bezier curve, so using the method for estimating \(t_1, ..., t_n\) as
+described before, we can iteratively solve for \(\hat{p}\) in the same
+manner. If we use soft clustering (e.g., EM as opposed to \(K\)-means),
+we can also insert a diagonal matrix of weights \(W_k\) to obtain
+
+\[\hat{p}_k = (T^\top W_k T)^{-1} T^\top W_k X\]
+
+where \(X\) is the full data matrix, \(T\) is the full model matrix, and
+\(W_k = w_k\) with the entries of \(w_k\) being the estimated
+probability of each \(x_i\) being in cluster \(k\).
 

### 2021-11-03 curve-fitting-methods/functions.R

@@ -1,0 +1,274 @@
+laplacian.eigenmap <- function(W, d = 2, normalized = TRUE) {
+  n <- nrow(W)
+  if (normalized) {
+    L <- normalized.laplacian(W)
+  } else {
+    L <- graph.laplacian(W)
+  }
+  L.eigen <- eigen(L, symmetric = TRUE)
+  sweep(L.eigen$vectors[, seq(n - 1, n - d)], 2, 
+        sqrt(L.eigen$values[seq(n - 1, n - d)]), `/`)
+}
+
+graph.laplacian <- function(W) {
+  return(diag(colSums(W)) - W)
+}
+
+spectral.clustering <- function(W, K = 2, d = K, normalized = TRUE) {
+  Y <- laplacian.eigenmap(W, d, normalized)
+  kmeans(Y, K)$cluster
+}
+
+param.curve.quadr.2 <- function(t., 
+                                beta10, beta11, beta12, 
+                                beta20, beta21, beta22) {
+  x1 <- beta10 + beta11 * t. + beta12 * t. ^ 2
+  x2 <- beta20 + beta21 * t. + beta22 * t. ^ 2
+  return(cbind(x1, x2))
+}
+
+estimate.one.t.quadr.2 <- function(x1, x2, 
+                                   beta10, beta11, beta12, 
+                                   beta20, beta21, beta22) {
+  gamma0 <- beta10 * beta11 + beta20 * beta21 - beta11 * x1 - beta21 * x2
+  gamma1 <- 2 * beta10 * beta12 + 2 * beta20 * beta22 + 
+    beta11 ^ 2 + beta21 ^ 2 - 
+    2 * beta12 * x1 - 2 * beta22 * x2
+  gamma2 <- 3 * beta11 * beta12 + 3 * beta21 * beta22
+  gamma3 <- 2 * (beta12 ^ 2 + beta22 ^ 2)
+  roots <- polyroot(c(gamma0, gamma1, gamma2, gamma3))
+  roots <- roots[abs(roots) == abs(Re(roots))]
+  roots <- Re(roots)
+  # if (length(roots) > 1) print(roots)
+  return(max(roots))
+}
+
+estimate.t.quadr.2 <- function(X, 
+                               beta10, beta11, beta12, 
+                               beta20, beta21, beta22) {
+  apply(X, 1, function(x) {
+    estimate.one.t.quadr.2(x[1], x[2], 
+                           beta10, beta11, beta12, 
+                           beta20, beta21, beta22)
+  })
+}
+
[…]

### 2021-11-03 curve-fitting-methods/souvenir-pless.R

@@ -66,7 +66,7 @@ plot(Xhat, col = z, asp = 1)
 D <- Xhat %>% 
   dist() %>% 
   as.matrix() %>% 
-  graph.eps(.1) %>% 
+  graph.eps(.2) %>% 
   graph.short()
 
 W <- runif(n * 2) %>% 

### 2021-11-10 curve-fitting-methods/curve-fitting-scratch.Rmd
updated notes

--- 2021-10-27 curve-fitting-methods/curve-fitting-scratch.Rmd
@@
 ```{r}
-set.seed(314159)
+set.seed(314)
 n <- 200
@@
 
-out <- estimate.quadr.curve.2(X)
+out <- estimate.quadr.curve.2(X, maxit = 100)
 thetahat <- out$theta
@@
 
-One problem with this method is the parameterization of the curve is not unique.
+Note: the parameterization of the curve is not unique.
 
@@
 out1 <- estimate.quadr.curve.2(Y,
-                              init.params = c(0, 0, 1, 0, 2, -2),
                               intercept = TRUE,
@@
 
+A modification to this that is possibly more robust is to use Bezier curves for $g$. 
+This is the same functional form as the polynomial curves used before, but with orthogonal bases:
+
+$$g(t | p) = \sum_{r = 0}^{R} p_r \binom{R}{r} (1 - t)^{R - r} t^r$$
+
+where $R$ is the order of the Bezier curve, each $p_s$ is a vector of length $d$, and, as before, $g : [0, 1] \mapsto \mathbb{R}^d$. 
+Thus, if we fit each $p_r$, then the procedure is the same as before.
+
+The least squares estimate for $p \in \mathbb{R}^{R \times d}$ is
+
+$$\hat{p} = (T^\top T)^{-1} T^\top X$$
+
+where $T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}$ and each $t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r$, and $X \in \mathbb{R}^{n \times d}$. 
+The same procedure for estimating $t_1, ..., t_n$ can be applied here.
+
+The parameterization for a given curve is not unique. 
+In particular, the above procedure will not necessarily provide $t_1, ..., t_n \in [0, 1]$. 
+One possible remedy for this is to, after estimating the $t_i$'s, normalize them to the unit interval. If we assume $t_1, ..., t_n \iid Uniform(a, b)$, then the UMVUE are
+
+$$\hat{a} = \frac{n t_{(1)} - t_{(n)}}{n - 1},
+\hat{b} = \frac{n t_{(n)} - t_{(1)}}{n - 1}$$
+
+which yields the normalization transformation $t \leftarrow (t - \hat{a}) / (\hat{b} - \hat{a})$.
+
+Alternatively, we can force the $t_i$'s to be approxiately uniform on the unit interval by the transformation $t \leftarrow \hat{F}(t)$, where $\hat{F}$ is the empirical CDF of $t_1, ..., t_n$.
+
+For initialization, we can use a one-dimensional Isomap embedding to estimate the $t_i$'s and use that to estimate $\hat{p}$.
+Experiments suggest that if the data are well-behaved (i.e., it looks like the curve we are trying to fit), this results in much faster convergence.
+
 # Clustering
@@
 1. Assign an initial clustering (e.g., via spectral clustering).
-2. Estimate the curve for each cluster.
+2. Estimate the curve for each cluster (using the same curve-fitting procedure as before).
 3. Reassign the clusters by proximity to each curve.
 4. Repeat 2 and 3 until convergence.
-
-\begin{example}
-
-We again limit these to be quadratic functions in $\mathbb{R}^2$. 
-Here, $K = 2$ and $n_1 = n_2 = 256$.
-
-```{r}
-f1 <- function(t) {
-  x1 <- t / sqrt(.5 + sqrt(5) / 2)
-  x2 <- x1 ** 2
-  return(cbind(x1, x2))
-}
-
-f2 <- function(t) {
-  x2 <- t / sqrt(.5 + sqrt(5) / 2)
-  x1 <- x2 ** 2
-  return(cbind(x1, x2))
-}
-
-n1 <- n2 <- 2 ** 8
-z <- c(rep(1, n1), rep(2, n2))
-t1 <- sort(runif(n1))
-t2 <- sort(runif(n2))
-X1 <- f1(t1)
-X2 <- f2(t2)
-X <- rbind(X1, X2)
-
-P <- X %*% t(X)
-Xhat <- embedding(P, 2, 0)
-
-manifold.out <- manifold.clustering.quadr.2(P, 2, 
-                                            intercept = FALSE, 
-                                            method = 'polynomial')
-zhat.manifold <- manifold.out$z
-if (mean(z == zhat.manifold) < .5) {
-  zhat.manifold <- 2 - zhat.manifold
-}
-t. <- seq(0, 1, .01)
-theta1 <- manifold.out$theta[[1]]
-theta2 <- manifold.out$theta[[2]]
-Y1 <- param.curve.quadr.2(t., 
-                          theta1[1], theta1[2], theta1[3], 
-                          theta1[4], theta1[5], theta1[6])
-Y2 <- param.curve.quadr.2(t.,
-                          theta2[1], theta2[2], theta2[3],
-                          theta2[4], theta2[5], theta2[6])
-Y <- rbind(Y1, Y2)
-z. <- c(rep(1, length(t.)),
-        rep(2, length(t.)))
-
-ggplot() + 
-  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
-  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
-  theme_bw() + 
-  # geom_line(aes(x = Y[, 1], y = Y[, 2],
-  #               groups = factor(z.))) + 
-  xlim(-1, 0) + ylim(-.5, .5) + 
-  coord_fixed()
-```
-
-Next, we draw $A \sim \RDPG(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
-The community detection error in this example is 30\%.
-In the following plot, the points are colored according to their true labels.
-
-```{r cache = TRUE, fig.width = 3, fig.height = 3}
-set.seed(12345)
-
-A <- draw.graph(P)
-Xhat <- embedding(A, 2, 0)
-
-manifold.out <- manifold.clustering.quadr.2(A, 2, 
-                                            intercept = FALSE,
-                                            method = 'polynomial')
-zhat.manifold <- manifold.out$z
-# table(z, zhat.manifold)
-
-# plot(Xhat, col = z, asp = 1)
-t. <- seq(-1, 1, .01)
-theta1 <- manifold.out$theta[[1]]
-theta2 <- manifold.out$theta[[2]]
-Y1 <- param.curve.quadr.2(t., 
-                          theta1[1], theta1[2], theta1[3], 
-                          theta1[4], theta1[5], theta1[6])
-Y2 <- param.curve.quadr.2(t.,
-                          theta2[1], theta2[2], theta2[3],
-                          theta2[4], theta2[5], theta2[6])
-Y <- rbind(Y1, Y2)
-z. <- c(rep(1, length(t.)),
-        rep(2, length(t.)))
-# lines(Y1)
-# lines(Y2, col = 2)
-
-ggplot() + 
-  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
-  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
-  theme_bw() + 
-  geom_path(aes(x = Y[, 1], y = Y[, 2],
-                groups = factor(z.))) + 
-  xlim(-.25, 1.25) + ylim(-.75, .75) + 
-  coord_fixed()
-```
-
-\end{example}
-
-A possibly more robust modification to this is to use Bezier curves for $g$. 
-This is the same functional form as the polynomial curves we used before, but with different basis functions. 
-For the following examples, we consider the quadratic Bezier curve, which has the following basis form:
-
-$$
-g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
-$$
-
-Where as before, $g : [0, 1] \mapsto \mathbb{R}^d$ and $t \in [0, 1]$, and each $p_r \in \mathbb{R}^d$. Thus if we can fit each $p_r$, then the procedure is the same as before.
-
-Common methods for Bezier curve fitting do not require specification of $t_1, ..., t_n$. However, an ordering is required. The curve fitting method is then modified as follows:
-
-1. Initialize $\hat{\theta}^{(0)}$ (e.g., randomly). 
-2. Estimate each $\hat{t}_i^{(s)}$ by minimizing $L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2$.
-3. Reorder $X$ according to $t_1, ..., t_n$.
-4. Estimate $\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}$ for the reordered $X$ using Bezier curve fitting.
-5. Compute $\hat{\theta}^{(s+1)}$ from $\{\hat{p}_r^{(s)}\}$.
-6. Repeat steps 2-5 until convergence.
 
@@
 We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$.
-$n_1 = n_2 = 128$ points are drawn uniformly from each.
+$n_1 = n_2 = 256$ points are drawn uniformly from each.
 
 ```{r}
-n1 <- 2 ** 7
+set.seed(314159265)
+n1 <- 2 ** 8
 n2 <- n1
@@
   theme_bw()
+
+P <- X %*% t(X)
+diag(P) <- 0
+A <- draw.graph(P)
+Xhat <- embedding(A, 2, 0)
+# if (mean(Xhat[, 1]) < 0) Xhat[, 1] <- -Xhat[, 1]
 ```
@@
 
-```{r}
-P <- X %*% t(X)
-diag(P) <- 0
-A <- draw.graph(P)
-Xhat <- embedding(A, 2, 0)
-if (mean(Xhat[, 1]) < 0) Xhat[, 1] <- -Xhat[, 1]
-
+```{r, fig.height = 3}
 ggplot() + 
@@
 
-```{r cache = TRUE}
-set.seed(314159265)
+```{r cache = TRUE, eval = TRUE}
+set.seed(314159)
 manifold.out <- manifold.clustering.quadr.2(A, 2, 
@@
                                             method = 'bezier')
+
 zhat.manifold <- manifold.out$z
@@
   geom_path(aes(x = Y[, 1], y = Y[, 2],
-                groups = factor(z.))) + 
-  xlim(-.25, 1.25) + ylim(-.75, .75) + 
+                group = factor(z.))) +
+  xlim(min(Xhat[, 1]), max(Xhat[, 1])) + ylim(min(Xhat[, 2]), max(Xhat[, 2])) + 
   coord_fixed()
@@
 \end{example}
+
+## Model-Based Clustering
+
+The CLT property of the ASE may allow us to approximate $W x - \hat{x} \sim \mathcal{N}(f_k(t), \Sigma_k)$.
+One possibly reasonable assumption for $\Sigma_k$ is $\Sigma_k = \Sigma_k(t) =  \diag(\sigma_{k1}^2(t), ..., \sigma_{kd}^2(t))$.
+A plot of the squared errors of cluster 1 in the previous example reveals the following:
+
+```{r, fig.height = 2, fig.width = 8}
+theta1 <- manifold.out$theta[[1]]
+theta2 <- manifold.out$theta[[2]]
+t. <- manifold.out$t
+X1 <- param.curve.quadr.2(t.[[1]], 
+                          theta1[1], theta1[2], theta1[3],
+                          theta1[4], theta1[5], theta1[6])
+X2 <- param.curve.quadr.2(t.[[2]], 
+                          theta2[1], theta2[2], theta2[3],
+                          theta2[4], theta2[5], theta2[6])
+# plot(Xhat[zhat.manifold == 1, 1] - X1[, 1])
+# plot(Xhat[zhat.manifold == 1, 2] - X1[, 2])
+
+p1 <- ggplot() + 
+  geom_point(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 1] - X1[, 1]) ^ 2)) + 
+  stat_smooth(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 1] - X1[, 1]) ^ 2)) + 
+  theme_bw() + 
+  labs(x = 't', y = expression((x[1] - hat(x)[1])^2))
+
+p2 <- ggplot() + 
+  geom_point(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 2] - X1[, 2]) ^ 2)) + 
+  stat_smooth(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 2] - X1[, 2]) ^ 2)) + 
+  theme_bw() + 
+  labs(x = 't', y = expression((x[2] - hat(x)[2])^2))
+
+gridExtra::grid.arrange(p1, p2, nrow = 1)
+```
+
+Fitting some curve to each $\sigma_{kr}^2(t)$ allows us to compute estimated probabilities of each point belonging to each cluster/manifold: $\hat{q}_{ik} = \mathcal{N}(x_i \mid g_k(\hat{t}_i), \Sigma_k(\hat{t}_i))$.
+Then we can use weighted least squares to estimate the parameters of $g$:
+
+$$\hat{p}_k = (T^\top Q_k T)^{-1} T^\top Q_k X$$
+
+where $Q_k = \diag(q_{1k}, ..., q_{nk})$.

### 2021-11-10 curve-fitting-methods/curve-fitting-scratch.tex
updated notes

--- 2021-10-27 curve-fitting-methods/curve-fitting-scratch.tex
@@
 
-One problem with this method is the parameterization of the curve is not
-unique.
+Note: the parameterization of the curve is not unique.
 
@@
 \begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-5-1} \end{center}
+
+A modification to this that is possibly more robust is to use Bezier
+curves for \(g\). This is the same functional form as the polynomial
+curves used before, but with orthogonal bases:
+
+\[g(t | p) = \sum_{r = 0}^{R} p_r \binom{R}{r} (1 - t)^{R - r} t^r\]
+
+where \(R\) is the order of the Bezier curve, each \(p_s\) is a vector
+of length \(d\), and, as before, \(g : [0, 1] \mapsto \mathbb{R}^d\).
+Thus, if we fit each \(p_r\), then the procedure is the same as before.
+
+The least squares estimate for \(p \in \mathbb{R}^{R \times d}\) is
+
+\[\hat{p} = (T^\top T)^{-1} T^\top X\]
+
+where \(T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}\)
+and each \(t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r\), and
+\(X \in \mathbb{R}^{n \times d}\). The same procedure for estimating
+\(t_1, ..., t_n\) can be applied here.
+
+The parameterization for a given curve is not unique. In particular, the
+above procedure will not necessarily provide
+\(t_1, ..., t_n \in [0, 1]\). One possible remedy for this is to, after
+estimating the \(t_i\)'s, normalize them to the unit interval. If we
+assume \(t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}Uniform(a, b)\),
+then the UMVUE are
+
+\[\hat{a} = \frac{n t_{(1)} - t_{(n)}}{n - 1},
+\hat{b} = \frac{n t_{(n)} - t_{(1)}}{n - 1}\]
+
+which yields the normalization transformation
+\(t \leftarrow (t - \hat{a}) / (\hat{b} - \hat{a})\).
+
+Alternatively, we can force the \(t_i\)'s to be approxiately uniform on
+the unit interval by the transformation \(t \leftarrow \hat{F}(t)\),
+where \(\hat{F}\) is the empirical CDF of \(t_1, ..., t_n\).
+
+For initialization, we can use a one-dimensional Isomap embedding to
+estimate the \(t_i\)'s and use that to estimate \(\hat{p}\). Experiments
+suggest that if the data are well-behaved (i.e., it looks like the curve
+we are trying to fit), this results in much faster convergence.
 
@@
 \item
-  Estimate the curve for each cluster.
+  Estimate the curve for each cluster (using the same curve-fitting
+  procedure as before).
 \item
@@
 
-We again limit these to be quadratic functions in $\mathbb{R}^2$. 
-Here, $K = 2$ and $n_1 = n_2 = 256$.
-
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-6-1} \end{center}
-
-Next, we draw $A \sim \mathrm{RDPG}(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
-The community detection error in this example is 30\%.
-In the following plot, the points are colored according to their true labels.
+We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$.
+$n_1 = n_2 = 256$ points are drawn uniformly from each.
 
@@
 
-\end{example}
-
-A possibly more robust modification to this is to use Bezier curves for
-\(g\). This is the same functional form as the polynomial curves we used
-before, but with different basis functions. For the following examples,
-we consider the quadratic Bezier curve, which has the following basis
-form:
-
-\[
-g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
-\]
-
-Where as before, \(g : [0, 1] \mapsto \mathbb{R}^d\) and
-\(t \in [0, 1]\), and each \(p_r \in \mathbb{R}^d\). Thus if we can fit
-each \(p_r\), then the procedure is the same as before.
-
-Common methods for Bezier curve fitting do not require specification of
-\(t_1, ..., t_n\). However, an ordering is required. The curve fitting
-method is then modified as follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Initialize \(\hat{\theta}^{(0)}\) (e.g., randomly).
-\item
-  Estimate each \(\hat{t}_i^{(s)}\) by minimizing
-  \(L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2\).
-\item
-  Reorder \(X\) according to \(t_1, ..., t_n\).
-\item
-  Estimate \(\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}\) for the
-  reordered \(X\) using Bezier curve fitting.
-\item
-  Compute \(\hat{\theta}^{(s+1)}\) from \(\{\hat{p}_r^{(s)}\}\).
-\item
-  Repeat steps 2-5 until convergence.
-\end{enumerate}
-
-\begin{example}
-
-We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$.
-$n_1 = n_2 = 128$ points are drawn uniformly from each.
-
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-9-1} \end{center}
-
 We draw $A \sim \mathrm{RDPG}(X)$ and obtain the following ASE:
 
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-10-1} \end{center}
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-8-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-11-1} \end{center}
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-9-1} \end{center}
 
 \end{example}
+
+\hypertarget{model-based-clustering}{%
+\subsection{Model-Based Clustering}\label{model-based-clustering}}
+
+The CLT property of the ASE may allow us to approximate
+\(W x - \hat{x} \sim \mathcal{N}(f_k(t), \Sigma_k)\). One possibly
+reasonable assumption for \(\Sigma_k\) is
+\(\Sigma_k = \Sigma_k(t) = \mathrm{diag}(\sigma_{k1}^2(t), ..., \sigma_{kd}^2(t))\).
+A plot of the squared errors of cluster 1 in the previous example
+reveals the following:
+
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-10-1} \end{center}
+
+Fitting some curve to each \(\sigma_{kr}^2(t)\) allows us to compute
+estimated probabilities of each point belonging to each
+cluster/manifold:
+\(\hat{q}_{ik} = \mathcal{N}(x_i \mid g_k(\hat{t}_i), \Sigma_k(\hat{t}_i))\).
+Then we can use weighted least squares to estimate the parameters of
+\(g\):
+
+\[\hat{p}_k = (T^\top Q_k T)^{-1} T^\top Q_k X\]
+
+where \(Q_k = \mathrm{diag}(q_{1k}, ..., q_{nk})\).
 

### 2021-11-10 curve-fitting-methods/functions.R

@@ -36,11 +36,40 @@ estimate.one.t.quadr.2 <- function(x1, x2,
     2 * beta12 * x1 - 2 * beta22 * x2
   gamma2 <- 3 * beta11 * beta12 + 3 * beta21 * beta22
   gamma3 <- 2 * (beta12 ^ 2 + beta22 ^ 2)
-  roots <- polyroot(c(gamma0, gamma1, gamma2, gamma3))
-  roots <- roots[abs(roots) == abs(Re(roots))]
-  roots <- Re(roots)
-  # if (length(roots) > 1) print(roots)
+  # roots <- polyroot(c(gamma0, gamma1, gamma2, gamma3))
+  # roots <- roots[abs(roots) == abs(Re(roots))]
+  # roots <- Re(roots)
+  roots <- optimize(
+    function(t) crossprod(c(x1 - beta10 - beta11 * t - beta12 * t^2, 
+                            x2 - beta20 - beta21 * t - beta22 * t^2)), 
+                    c(-1, 1)
+  )$minimum
+  if (length(roots) > 1) print(roots)
+  if (abs(roots) == 1) print(roots)
   return(max(roots))
+}
+
+estimate.one.t.bezier <- function(x, p) {
+  polynomial <- function(t, p) {
+    d <- ncol(p)
+    r <- nrow(p) - 1
+    T <- sapply(seq(0, r), function(s) {
+      choose(r, s) * (1 - t) ^ (r - s) * t ^ s
+    })
+    T <- matrix(rep(T, d), ncol = d, nrow = r + 1, byrow = FALSE)
+    colSums(T * p)
+  }
+  l <- function(t) crossprod(polynomial(t, p) - x)
+  minima <- optimize(l, c(-1, 1))$minimum
+  if (abs(minima) == 1) print(minima)
+  if (length(minima) > 1) print(minima)
+  return(max(minima))
+}
+
+estimate.t.bezier <- function(X, p) {
+  apply(X, 1, function(x) {
+    estimate.one.t.bezier(x, p)
+  })
 }
 
 estimate.t.quadr.2 <- function(X, 
@@ -92,12 +121,21 @@ estimate.quadr.curve.2 <- function(X, init.params,
   }
   d.theta <- 1 / eps
   
+  n <- nrow(X)
+  w <- n ^ 2 - 1
+  
   niter <- 0
   while (d.theta > eps) {
     theta.prev <- theta
     t.hat <- estimate.t.quadr.2(X,
[…]

### 2021-11-10 curve-fitting-methods/souvenir-pless.R

@@ -6,14 +6,34 @@ source('https://mtrosset.pages.iu.edu/Courses/675/stress.r')
 source('~/dev/pabm-grdpg/functions.R')
 source('~/dev/dissertation/curve-fitting-methods/functions.R')
 
-weighted.mds <- function(D, d = 2, w = rep(1, ncol(D))) {
+weighted.mds <- function(D, d = 2, w = rep(1, ncol(D)) / ncol(D)) {
   n <- ncol(D)
   e <- rep(1, n)
   H <- diag(n) - tcrossprod(e, w)
   B <- -H %*% D %*% H / 2
   eigen.B <- eigen(B, symmetric = TRUE)
-  Y <- sweep(eigen.B$vectors[, seq(d)], 2, sqrt(eigen.B$values[seq(d)]), `*`)
+  Y <- sweep(eigen.B$vectors[, seq(d), drop = FALSE], 2, sqrt(eigen.B$values[seq(d)]), `*`)
   Y <- Y / (sum(abs(eigen.B$values)) / sum(eigen.B$values[seq(d)]))
+  
+  eps <- 1e-3
+  W <- outer(w, w)
+  L <- graph.laplacian(W)
+  R <- chol(L + 1)
+  Linv <- chol2inv(R)
+  
+  stress <- sum(as.dist(W) * (dist(Y) - as.dist(D)) ^ 2)
+  stress.prev <- stress + 1 / eps
+  # stress.vec <- stress
+  
+  while (stress.prev - stress > eps) {
+    stress.prev <- stress
+    m <- as.dist(W) * as.dist(D) / dist(Y)
+    Y <- Linv %*% graph.laplacian(as.matrix(m)) %*% Y
+    stress <- sum(as.dist(W) * (dist(Y) - as.dist(D)) ^ 2)
+    # print(stress)
+    # stress.vec <- c(stress.vec, stress)
+  }
+  # plot(log(stress.vec))
   return(Y)
 }
 
@@ -69,22 +89,27 @@ D <- Xhat %>%
   graph.eps(.2) %>% 
   graph.short()
 
+d <- 1
+
 W <- runif(n * 2) %>% 
   matrix(nrow = n, ncol = 2) %>% 
   sweep(., 1, rowSums(.), `/`)
 
-Y1 <- weighted.mds(D, 2, W[, 1])
-Y2 <- weighted.mds(D, 2, W[, 2])
[…]

### 2021-11-17 curve-fitting-methods/curve-fitting-scratch.Rmd
still working on typing up new material

@@
 \newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
+\newcommand{\ind}{\stackrel{\mathrm{ind}}{\sim}}
 
@@
 
+## Theoretical Framework
+
+If we assume that the points are distributed as $X_i \ind \mathcal{N}(f(t_i | p), \Sigma(t_i | \phi))$, then we can write the (incomplete) log likelihood as:
+
+$$\ell(p, \phi) = -\frac{1}{2} \sum_i |\Sigma(t_i | \phi)| - \frac{1}{2} \sum_i (x_i - f(t_i | p))^\top (\Sigma(t_i | \phi))^{-1} (x_i - f(t_i | p))$$
+
+In the case $\Sigma(t_i | \phi) = \sigma^2 I$ and $f(t_i | p) = p^\top t_i$ where $t_i \in \mathbb{R}^{R+1}$ is the Bezier polynomial expansion of order $R$, then this becomes
+
+$$\ell(p, \sigma^2) = -\frac{n d}{2} \log \sigma^2 - \frac{1}{2 \sigma^2} \|X - T p\|^2_F$$
+
+Then given $t_1, ..., t_n$, the MLE of $p$ is again $(T^\top T)^{-1} T^\top X$, and the MLE of $\sigma^2$ is just $\frac{1}{n d} \|X - T \hat{p}\|^2_F$.
+And given $p$, we can estimate each $t_i$ in the same way as before to maximize $\ell$ ($\sigma^2$ is unnecessary here).
+
 # Clustering

### 2021-11-17 curve-fitting-methods/curve-fitting-scratch.tex
still working on typing up new material

@@
 
+\hypertarget{theoretical-framework}{%
+\subsection{Theoretical Framework}\label{theoretical-framework}}
+
+If we assume that the points are distributed as
+\(X_i \stackrel{\mathrm{ind}}{\sim}\mathcal{N}(f(t_i | p), \Sigma(t_i | \phi))\),
+then we can write the (incomplete) log likelihood as:
+
+\[\ell(p, \phi) = -\frac{1}{2} \sum_i |\Sigma(t_i | \phi)| - \frac{1}{2} \sum_i (x_i - f(t_i | p))^\top (\Sigma(t_i | \phi))^{-1} (x_i - f(t_i | p))\]
+
+In the case \(\Sigma(t_i | \phi) = \sigma^2 I\) and
+\(f(t_i | p) = p^\top t_i\) where \(t_i \in \mathbb{R}^{R+1}\) is the
+Bezier polynomial expansion of order \(R\), then this becomes
+
+\[\ell(p, \sigma^2) = -\frac{n d}{2} \log \sigma^2 - \frac{1}{2 \sigma^2} \|X - T p\|^2_F\]
+
+Then given \(t_1, ..., t_n\), the MLE of \(p\) is again
+\((T^\top T)^{-1} T^\top X\), and the MLE of \(\sigma^2\) is just
+\(\frac{1}{n d} \|X - T \hat{p}\|^2_F\). And given \(p\), we can
+estimate each \(t_i\) in the same way as before to maximize \(\ell\)
+(\(\sigma^2\) is unnecessary here).
+
 \hypertarget{clustering}{%

### 2021-11-17 curve-fitting-methods/functions.R

@@ -209,17 +209,18 @@ estimate.bezier.curve.2 <- function(X,
   while (d.mse > eps) {
     mse.prev <- mse
     t.hat <- estimate.t.bezier(X, p)
-    # min.t <- min(t.hat)
-    # max.t <- max(t.hat)
-    # a <- (n * (n + 1) * min.t - (n + 1) * max.t) / w
-    # b <- (-(n + 1) * min.t + n * (n + 1) * max.t) / w
-    # t.hat <- (t.hat - a) / (b - a)
-    t.hat <- ecdf(t.hat)(t.hat)
+    min.t <- min(t.hat)
+    max.t <- max(t.hat)
+    a <- (n * (n + 1) * min.t - (n + 1) * max.t) / w
+    b <- (-(n + 1) * min.t + n * (n + 1) * max.t) / w
+    t.hat <- (t.hat - a) / (b - a)
+    # t.hat <- ecdf(t.hat)(t.hat)
     T <- construct.bezier.model.matrix(t.hat, r)
     p <- solve(t(T) %*% W %*% T, t(T) %*% W %*% X)
     X.hat <- bezier.curve(t.hat, p)
-    # plot(X, asp = 1)
-    # lines(X.hat[order(t.hat), ])
+    print(bezier.mse(X, t.hat, p))
+    plot(X, asp = 1)
+    lines(X.hat[order(t.hat), ], col = 2)
     mse <- sum((X - X.hat) ^ 2)
     # print(mse)
     d.mse <- abs((mse - mse.prev) / mse)
@@ -382,4 +383,10 @@ manifold.clustering.quadr.2 <- function(A, K = 2, d = 2,
               t = list(curve1$t, curve2$t),
               theta = list(theta1, theta2),
               niter = niter))
+}
+
+bezier.mse <- function(X, t, p) {
+  r <- nrow(p) - 1L
+  T <- construct.bezier.model.matrix(t, r)
+  norm(X - T %*% p, 'F') ^ 2 / prod(dim(X))
 }

### 2021-11-17 curve-fitting-methods/souvenir-pless.R

@@ -39,7 +39,7 @@ weighted.mds <- function(D, d = 2, w = rep(1, ncol(D)) / ncol(D)) {
 
 compute.dist2.manifold <- function(Dg, De, w) {
   w.sum <- sum(w)
-  (rowSums(sweep(Dg - De, 2, w, `*`)) / w.sum) ^ 2
+  (rowSums(sweep((Dg - De) ^ 2, 2, w, `*`)) / w.sum)
 }
 
 update.weights <- function(D2, sigma2 = 1e-1) {
@@ -104,7 +104,7 @@ Y2 <- weighted.mds(D, d, W[, 2])
 d1 <- compute.dist2.manifold(D, as.matrix(dist(Y1)), W[, 1])
 d2 <- compute.dist2.manifold(D, as.matrix(dist(Y2)), W[, 2])
 
-W <- update.weights(cbind(d1, d2), 1e-1)
+W <- update.weights(cbind(d1, d2), 1e-2)
 
 zhat <- apply(W, 1, which.max)
 if (mean(z == zhat) < .5) zhat <- 3 - zhat

### 2021-12-08 curve-fitting-methods/curve-fitting-scratch.Rmd
updated notes

--- 2021-11-10 curve-fitting-methods/curve-fitting-scratch.Rmd
@@
 
-where $T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}$ and each $t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r$, and $X \in \mathbb{R}^{n \times d}$. 
+where $T = \begin{bmatrix} t_{\cdot, 1} & \cdots & t_{\cdot, R} \end{bmatrix}$ and each $t_{ir} = \binom{R}{r} (1 - t_i)^{R - r} t_i^r$, and $X \in \mathbb{R}^{n \times d}$. 
 The same procedure for estimating $t_1, ..., t_n$ can be applied here.
@@
 Experiments suggest that if the data are well-behaved (i.e., it looks like the curve we are trying to fit), this results in much faster convergence.
+
+## Theoretical Framework
+
+If we assume that the points are distributed as $X_i \ind \mathcal{N}(f(t_i | p), \Sigma(t_i | \phi))$, then we can write the (incomplete) log likelihood as:
+
+$$\ell(p, \phi) = -\frac{1}{2} \sum_i \log |\Sigma(t_i | \phi)| - \frac{1}{2} \sum_i (x_i - f(t_i | p))^\top (\Sigma(t_i | \phi))^{-1} (x_i - f(t_i | p))$$
+
+In the case $\Sigma(t_i | \phi) = \phi I$ and $f(t_i | p) = p^\top \tilde{t}_i$ where $\tilde{t}_i \in \mathbb{R}^{R+1}$ is the Bezier polynomial expansion of order $R$, then this becomes
+
+$$\ell(p, \phi) = -\frac{n d}{2} \log \phi - \frac{1}{2 \phi} \|X - T p\|^2_F$$
+
+Then given $t_1, ..., t_n$, the MLE of $p$ is again $(T^\top T)^{-1} T^\top X$, and the MLE of $\phi$ is just $\frac{1}{n d} \|X - T \hat{p}\|^2_F$.
+And given $p$, we can estimate each $t_i$ in the same way as before to maximize $\ell$ (the variance is unnecessary here).
+
+As noted before, the parameterization given by $T$ and $p$ is not unique. 
+We can arbitrarily scale the $t_i$'s and $p$ to obtain the same value of $\ell$. 
+Naively performing the proposed algorithm sometimes results in estimates for $p$ that diverge and estimates for $t_i$'s that converge to a single value. 
+To remedy this, we can either scale the $t_i$'s as before, 
+or we can scale $p$ by forcing $f$ to be an arclength parameterization.
+
+\begin{theorem}
+Let $x_1, ..., x_n \ind \mathcal{N}(p^\top \tilde{t}_i, \phi I_d)$,
+where each $x_i \in \mathbb{R}^d$, $\tilde{t}_i \in \mathbb{R}^{R+1}$ such that $\tilde{t}_{i,r} = \binom{R}{r} (1 - t_i)^{R-r} t_i^r$ for $i = 1, ..., n$ and $r = 0, ..., R$, 
+and $p \in \mathbb{R}^{(R+1) \times d}$.
+
+Then each iteration of the following decreases the negative log likelihood:
+
+\begin{enumerate}
+  \item $p^{(s)} \leftarrow (T^\top T)^{-1} T^\top X$, 
+where $T = \begin{bmatrix} \tilde{t}_1 & \cdots & \tilde{t}_n \end{bmatrix}^\top$ and 
+$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+  \item $t_i^{(s)} \leftarrow \arg\min_t \|x_i - p^\top \tilde{t}_i\|^2$
+\end{enumerate}
+\end{theorem}
+
+\begin{proof}[Proof (sketch)]
+
+For $\Sigma(t_i | \phi) = \phi I$, the negative log likelihood is, 
+up to some additive and multiplicative constants:
+
+$$l(p, \{t_i\}) = \|X - T p\|^2_F = X^\top X - 2 X^\top T p + p^\top T^\top T p$$
+
+To find the minimizer $\hat{p}$ given $t_1, ..., t_n$:
+
+$\nabla_p l = -2 T^\top X + 2 T^\top T p = 0$  
+$\implies \hat{p} = (T^\top T)^{-1} T^\top X$
+
+To find the minimizer $\hat{t}_i$ given $p$:
+
+We can rewrite the negative log likelihood as 
+$l(p, \{t_i\}) = \sum_i \|x_i - p^\top \tilde{t}_i\|^2$.
+
+Then each entry of $\nabla_{t} l$ depends only on $t_i$, so each $t_i$ can be optimized independently. 
+Furthermore, each $\frac{\partial}{\partial t_i} \|x_i - p^\top \tilde{t}_i\|^2$ is a polynomial in $t_i$, so it can be minimized by simply finding the roots of the polynomial.
+
+Thus this algorithm is a coordinate descent algorithm, which reduces the objective function with each iteration. 
+\end{proof}
+
+\begin{theorem}
+
+Let $A^{(n)} \sim \RDPG(F(p, \theta), n)$, and let $X^{(n)}$ be the ASE of $A^{(n)}$. 
+Let $\hat{p}, \{\hat{t}_i\}$ be the global minimizer of $l$ given $X^{(n)}$. 
+Then $l(\hat{p}, \{\hat{t}_i\}) \stackrel{a.s.}{\to} 0$.
+
+\end{theorem}
+
+### Covariance Structures
+
+While the CLT property of the ASE allows us to approximate $x_i \sim \mathcal{N} (W f(t_i | p), W \Sigma(t_i | \phi) W^\top)$, it is unclear what $\Sigma(t_i | \phi)$ looks like, and we cannot in general say $\Sigma(t_i | \phi) = \phi I_d$.
 

### 2021-12-08 curve-fitting-methods/curve-fitting-scratch.tex
updated notes

--- 2021-11-10 curve-fitting-methods/curve-fitting-scratch.tex
@@
 
-where \(T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}\)
-and each \(t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r\), and
+where
+\(T = \begin{bmatrix} t_{\cdot, 1} & \cdots & t_{\cdot, R} \end{bmatrix}\)
+and each \(t_{ir} = \binom{R}{r} (1 - t_i)^{R - r} t_i^r\), and
 \(X \in \mathbb{R}^{n \times d}\). The same procedure for estimating
@@
 we are trying to fit), this results in much faster convergence.
+
+\hypertarget{theoretical-framework}{%
+\subsection{Theoretical Framework}\label{theoretical-framework}}
+
+If we assume that the points are distributed as
+\(X_i \stackrel{\mathrm{ind}}{\sim}\mathcal{N}(f(t_i | p), \Sigma(t_i | \phi))\),
+then we can write the (incomplete) log likelihood as:
+
+\[\ell(p, \phi) = -\frac{1}{2} \sum_i \log |\Sigma(t_i | \phi)| - \frac{1}{2} \sum_i (x_i - f(t_i | p))^\top (\Sigma(t_i | \phi))^{-1} (x_i - f(t_i | p))\]
+
+In the case \(\Sigma(t_i | \phi) = \phi I\) and
+\(f(t_i | p) = p^\top \tilde{t}_i\) where
+\(\tilde{t}_i \in \mathbb{R}^{R+1}\) is the Bezier polynomial expansion
+of order \(R\), then this becomes
+
+\[\ell(p, \phi) = -\frac{n d}{2} \log \phi - \frac{1}{2 \phi} \|X - T p\|^2_F\]
+
+Then given \(t_1, ..., t_n\), the MLE of \(p\) is again
+\((T^\top T)^{-1} T^\top X\), and the MLE of \(\phi\) is just
+\(\frac{1}{n d} \|X - T \hat{p}\|^2_F\). And given \(p\), we can
+estimate each \(t_i\) in the same way as before to maximize \(\ell\)
+(the variance is unnecessary here).
+
+As noted before, the parameterization given by \(T\) and \(p\) is not
+unique. We can arbitrarily scale the \(t_i\)'s and \(p\) to obtain the
+same value of \(\ell\). Naively performing the proposed algorithm
+sometimes results in estimates for \(p\) that diverge and estimates for
+\(t_i\)'s that converge to a single value. To remedy this, we can either
+scale the \(t_i\)'s as before, or we can scale \(p\) by forcing \(f\) to
+be an arclength parameterization.
+
+\begin{theorem}
+Let $x_1, ..., x_n \stackrel{\mathrm{ind}}{\sim}\mathcal{N}(p^\top \tilde{t}_i, \phi I_d)$,
+where each $x_i \in \mathbb{R}^d$, $\tilde{t}_i \in \mathbb{R}^{R+1}$ such that $\tilde{t}_{i,r} = \binom{R}{r} (1 - t_i)^{R-r} t_i^r$ for $i = 1, ..., n$ and $r = 0, ..., R$, 
+and $p \in \mathbb{R}^{(R+1) \times d}$.
+
+Then each iteration of the following decreases the negative log likelihood:
+
+\begin{enumerate}
+  \item $p^{(s)} \leftarrow (T^\top T)^{-1} T^\top X$, 
+where $T = \begin{bmatrix} \tilde{t}_1 & \cdots & \tilde{t}_n \end{bmatrix}^\top$ and 
+$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+  \item $t_i^{(s)} \leftarrow \arg\min_t \|x_i - p^\top \tilde{t}_i\|^2$
+\end{enumerate}
+\end{theorem}
+
+\begin{proof}[Proof (sketch)]
+
+For $\Sigma(t_i | \phi) = \phi I$, the negative log likelihood is, 
+up to some additive and multiplicative constants:
+
+$$l(p, \{t_i\}) = \|X - T p\|^2_F = X^\top X - 2 X^\top T p + p^\top T^\top T p$$
+
+To find the minimizer $\hat{p}$ given $t_1, ..., t_n$:
+
+$\nabla_p l = -2 T^\top X + 2 T^\top T p = 0$  
+$\implies \hat{p} = (T^\top T)^{-1} T^\top X$
+
+To find the minimizer $\hat{t}_i$ given $p$:
+
+We can rewrite the negative log likelihood as 
+$l(p, \{t_i\}) = \sum_i \|x_i - p^\top \tilde{t}_i\|^2$.
+
+Then each entry of $\nabla_{t} l$ depends only on $t_i$, so each $t_i$ can be optimized independently. 
+Furthermore, each $\frac{\partial}{\partial t_i} \|x_i - p^\top \tilde{t}_i\|^2$ is a polynomial in $t_i$, so it can be minimized by simply finding the roots of the polynomial.
+
+Thus this algorithm is a coordinate descent algorithm, which reduces the objective function with each iteration. 
+\end{proof}
+
+\begin{theorem}
+
+Let $A^{(n)} \sim \mathrm{RDPG}(F(p, \theta), n)$, and let $X^{(n)}$ be the ASE of $A^{(n)}$. 
+Let $\hat{p}, \{\hat{t}_i\}$ be the global minimizer of $l$ given $X^{(n)}$. 
+Then $l(\hat{p}, \{\hat{t}_i\}) \stackrel{a.s.}{\to} 0$.
+
+\end{theorem}
+
+\hypertarget{covariance-structures}{%
+\subsubsection{Covariance Structures}\label{covariance-structures}}
+
+While the CLT property of the ASE allows us to approximate
+\(x_i \sim \mathcal{N} (W f(t_i | p), W \Sigma(t_i | \phi) W^\top)\), it
+is unclear what \(\Sigma(t_i | \phi)\) looks like, and we cannot in
+general say \(\Sigma(t_i | \phi) = \phi I_d\).
 

### 2021-12-08 curve-fitting-methods/functions.R

@@ -42,34 +42,53 @@ estimate.one.t.quadr.2 <- function(x1, x2,
   roots <- optimize(
     function(t) crossprod(c(x1 - beta10 - beta11 * t - beta12 * t^2, 
                             x2 - beta20 - beta21 * t - beta22 * t^2)), 
-                    c(-1, 1)
+    c(-1, 1)
   )$minimum
   if (length(roots) > 1) print(roots)
   if (abs(roots) == 1) print(roots)
   return(max(roots))
 }
 
-estimate.one.t.bezier <- function(x, p) {
+estimate.one.t.bezier <- function(x, p, min.t = -2, max.t = 2, tol = 1e-9) {
   polynomial <- function(t, p) {
-    d <- ncol(p)
+    # d <- ncol(p)
     r <- nrow(p) - 1
     T <- sapply(seq(0, r), function(s) {
       choose(r, s) * (1 - t) ^ (r - s) * t ^ s
     })
-    T <- matrix(rep(T, d), ncol = d, nrow = r + 1, byrow = FALSE)
-    colSums(T * p)
+    as.numeric(T %*% p)
   }
-  l <- function(t) crossprod(polynomial(t, p) - x)
-  minima <- optimize(l, c(-1, 1))$minimum
-  if (abs(minima) == 1) print(minima)
-  if (length(minima) > 1) print(minima)
+  l <- function(t) log(as.numeric(crossprod(polynomial(t, p) - x)))
+  minima <- optimize(l, c(min.t, max.t), tol = tol)$minimum
+  if (abs(minima) == max.t) warning(minima)
+  if (length(minima) > 1) warning(minima)
   return(max(minima))
 }
 
-estimate.t.bezier <- function(X, p) {
-  apply(X, 1, function(x) {
-    estimate.one.t.bezier(x, p)
-  })
+estimate.t.bezier <- function(X, p, t.prev, tol = 1e-9) {
+  if (missing(t.prev)) {
+    t.hat <- apply(X, 1, function(x) {
+      estimate.one.t.bezier(x, p, tol = tol)
+    })
+  } else {
+    t.hat <- sapply(seq_along(t.prev), function(i) {
+      estimate.one.t.bezier(X[i, ], p, t.prev[i] - 1, t.prev[i] + 1, tol = tol)
+    })
+    loss.new <- sapply(seq_along(t.hat),
+                       function(i) loss.one.t(X[i, ], t.hat[i], p))
[…]

### 2021-12-15 curve-fitting-methods/curve-fitting-scratch.Rmd
minor update

@@
 \begin{enumerate}
-  \item $p^{(s)} \leftarrow (T^\top T)^{-1} T^\top X$, 
-where $T = \begin{bmatrix} \tilde{t}_1 & \cdots & \tilde{t}_n \end{bmatrix}^\top$ and 
-$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
-  \item $t_i^{(s)} \leftarrow \arg\min_t \|x_i - p^\top \tilde{t}_i\|^2$
+  \item $p \leftarrow (T^\top T)^{-1} T^\top X$, 
+  where $T = \begin{bmatrix} \tilde{t}_1 & \cdots & \tilde{t}_n \end{bmatrix}^\top$ 
+  and $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+  \item $t_i \leftarrow \arg\min_t \|x_i - p^\top \tilde{t}_i\|^2$,
+  which can be solved by finding the roots of a polynomial of degree $2 R - 1$.
 \end{enumerate}
@@
 
+The objective function that this method aims to minimize is:
+
+$$\sum_{k=1}^K \sum_{i \in C_k} \|x_i - g_k(t_i | p_k)\|^2$$
+
+Where $g_k : [0, 1] \mapsto \mathbb{R}^d$ with parameters $p_k$ is the curve for community $k$.
+
+It can be shown that each iteration decreases the objective unless we are at a stationary point.
+
 ## Model-Based Clustering

### 2021-12-15 curve-fitting-methods/curve-fitting-scratch.tex
minor update

@@
 
+The objective function that this method aims to minimize is:
+
+\[\sum_{k=1}^K \sum_{i \in C_k} \|x_i - g_k(t_i | p_k)\|^2\]
+
+Where \(g_k : [0, 1] \mapsto \mathbb{R}^d\) with parameters \(p_k\) is
+the curve for community \(k\).
+
+It can be shown that each iteration decreases the objective unless we
+are at a stationary point.
+
 \hypertarget{model-based-clustering}{%

### 2022-01-12 curve-fitting-methods/curve-fitting-scratch.Rmd
added example

@@
 library(ggplot2)
-import::from(magrittr, `%>%`)
+import::from(magrittr, `%>%`, `%<>%`)
 ```
@@
 source('~/dev/dissertation/curve-fitting-methods/functions.R')
+source('https://mtrosset.pages.iu.edu/Courses/675/graph.r')
 ```
@@
 
-It can be shown that each iteration decreases the objective unless we are at a stationary point.
+**TBD**: It can be shown that each iteration decreases the objective unless we are at a stationary point.
 
-## Model-Based Clustering
+## Harry Potter Emnity Graph
 
-The CLT property of the ASE may allow us to approximate $W x - \hat{x} \sim \mathcal{N}(f_k(t), \Sigma_k)$.
-One possibly reasonable assumption for $\Sigma_k$ is $\Sigma_k = \Sigma_k(t) =  \diag(\sigma_{k1}^2(t), ..., \sigma_{kd}^2(t))$.
-A plot of the squared errors of cluster 1 in the previous example reveals the following:
+\begin{example}
+In this section, we analyze the Harry Potter emnity graph. 
+Rubin-Delanchy et al. suggested applying the GRDPG ASE to this graph with $p = 1$ and $q = 1$, revealing a DCBM-like structure in the latent space. 
+Passino and Heard instead fit Gaussian processes to the embedding. 
+Here, we consider what happens when we apply the Bezier curve clustering method. 
+\end{example}
 
-```{r, fig.height = 2, fig.width = 8}
+```{r, fig.width = 8}
+relations.df <- readr::read_csv('~/dev/potter-network/data/relations.csv') %>% 
+  dplyr::mutate(source = source + 1, target = target + 1)
+characters.df <- readr::read_csv('~/dev/potter-network/data/characters.csv') %>% 
+  dplyr::mutate(id = id + 1)
+n <- nrow(characters.df)
+
+A <- matrix(0, nrow = n, ncol = n)
+rownames(A) <- colnames(A) <- characters.df$name
+for (i in seq(nrow(relations.df))) {
+  if (relations.df$type[i] == '-') {
+    A[relations.df$source[i], relations.df$target[i]] %<>% magrittr::add(1)
+  } else {
+    A[relations.df$source[i], relations.df$target[i]] %<>% magrittr::subtract(1)
+  }
+}
+A[A < 0] <- 0
+A <- A + t(A)
+A <- sign(A)
+diag(A) <- 0
+deg.A <- colSums(A)
+
+A <- A[deg.A > 0, deg.A > 0]
+
+A.eigen <- eigen(A, symmetric = TRUE)
+Xhat <- embedding(A, p = 1, q = 1)
+
+# ggplot() + 
+#   geom_point(aes(x = Xhat[, 1], y = Xhat[, 2])) + 
+#   ggrepel::geom_text_repel(aes(x = Xhat[, 1], y = Xhat[, 2], 
+#                                label = characters.df$name[deg.A > 0]))
+
+z.init <- kmeans(sweep(Xhat, 1, apply(Xhat, 1, function(x) sqrt(sum(x ^ 2))), `/`), 2)$cluster
+```
+
+```{r, fig.width = 10, fig.height = 4, cache = TRUE}
+manifold.out <- manifold.clustering.quadr.2(A, 2, p = 1, q = 1, 
+                                            initialization = z.init, 
+                                            eps.isomap = .7,
+                                            method = 'bezier')
+
+zhat.manifold <- manifold.out$z
+
+t. <- seq(-1, 1, .01)
 theta1 <- manifold.out$theta[[1]]
 theta2 <- manifold.out$theta[[2]]
-t. <- manifold.out$t
-X1 <- param.curve.quadr.2(t.[[1]], 
+Y1 <- param.curve.quadr.2(t., 
                           theta1[1], theta1[2], theta1[3], 
                           theta1[4], theta1[5], theta1[6])
-X2 <- param.curve.quadr.2(t.[[2]], 
+Y2 <- param.curve.quadr.2(t.,
                           theta2[1], theta2[2], theta2[3],
                           theta2[4], theta2[5], theta2[6])
-# plot(Xhat[zhat.manifold == 1, 1] - X1[, 1])
-# plot(Xhat[zhat.manifold == 1, 2] - X1[, 2])
+Y <- rbind(Y1, Y2)
+z. <- c(rep(1, length(t.)),
+        rep(2, length(t.)))
 
-p1 <- ggplot() + 
-  geom_point(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 1] - X1[, 1]) ^ 2)) + 
-  stat_smooth(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 1] - X1[, 1]) ^ 2)) + 
+ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(zhat.manifold))) + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
   theme_bw() + 
-  labs(x = 't', y = expression((x[1] - hat(x)[1])^2))
-
-p2 <- ggplot() + 
-  geom_point(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 2] - X1[, 2]) ^ 2)) + 
-  stat_smooth(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 2] - X1[, 2]) ^ 2)) + 
-  theme_bw() + 
-  labs(x = 't', y = expression((x[2] - hat(x)[2])^2))
-
-gridExtra::grid.arrange(p1, p2, nrow = 1)
+  theme(legend.position = 'none') +
+  geom_path(aes(x = Y[, 1], y = Y[, 2],
+                group = factor(z.))) +
+  ggrepel::geom_text_repel(aes(x = Xhat[, 1], y = Xhat[, 2],
+                               label = characters.df$name[deg.A > 0])) + 
+  xlim(min(Xhat[, 1]), max(Xhat[, 1])) + ylim(min(Xhat[, 2]), max(Xhat[, 2]))
 ```
-
-Fitting some curve to each $\sigma_{kr}^2(t)$ allows us to compute estimated probabilities of each point belonging to each cluster/manifold: $\hat{q}_{ik} = \mathcal{N}(x_i \mid g_k(\hat{t}_i), \Sigma_k(\hat{t}_i))$.
-Then we can use weighted least squares to estimate the parameters of $g$:
-
-$$\hat{p}_k = (T^\top Q_k T)^{-1} T^\top Q_k X$$
-
-where $Q_k = \diag(q_{1k}, ..., q_{nk})$.

### 2022-01-12 curve-fitting-methods/curve-fitting-scratch.tex
added example

@@
 
-It can be shown that each iteration decreases the objective unless we
-are at a stationary point.
+\textbf{TBD}: It can be shown that each iteration decreases the
+objective unless we are at a stationary point.
 
-\hypertarget{model-based-clustering}{%
-\subsection{Model-Based Clustering}\label{model-based-clustering}}
+\hypertarget{harry-potter-emnity-graph}{%
+\subsection{Harry Potter Emnity Graph}\label{harry-potter-emnity-graph}}
 
-The CLT property of the ASE may allow us to approximate
-\(W x - \hat{x} \sim \mathcal{N}(f_k(t), \Sigma_k)\). One possibly
-reasonable assumption for \(\Sigma_k\) is
-\(\Sigma_k = \Sigma_k(t) = \mathrm{diag}(\sigma_{k1}^2(t), ..., \sigma_{kd}^2(t))\).
-A plot of the squared errors of cluster 1 in the previous example
-reveals the following:
+\begin{example}
+In this section, we analyze the Harry Potter emnity graph. 
+Rubin-Delanchy et al. suggested applying the GRDPG ASE to this graph with $p = 1$ and $q = 1$, revealing a DCBM-like structure in the latent space. 
+Passino and Heard instead fit Gaussian processes to the embedding. 
+Here, we consider what happens when we apply the Bezier curve clustering method. 
+\end{example}
 
-\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-10-1} \end{center}
-
-Fitting some curve to each \(\sigma_{kr}^2(t)\) allows us to compute
-estimated probabilities of each point belonging to each
-cluster/manifold:
-\(\hat{q}_{ik} = \mathcal{N}(x_i \mid g_k(\hat{t}_i), \Sigma_k(\hat{t}_i))\).
-Then we can use weighted least squares to estimate the parameters of
-\(g\):
-
-\[\hat{p}_k = (T^\top Q_k T)^{-1} T^\top Q_k X\]
-
-where \(Q_k = \mathrm{diag}(q_{1k}, ..., q_{nk})\).
+\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-11-1} \end{center}
 

### 2022-01-12 curve-fitting-methods/functions.R

@@ -202,9 +202,9 @@ estimate.bezier.curve.2 <- function(X,
                                     init.params, 
                                     weights,
                                     eps = 1e-3, maxit = 200,
-                                    initialization = 'random', 
-                                    normalize.t = FALSE, 
-                                    normalization.method = 'umvue',
+                                    initialization = 'isomap', 
+                                    normalize = TRUE, 
+                                    normalize.method = 'umvue',
                                     eps.isomap = 2e-1) {
   n <- nrow(X)
   d <- 2
@@ -373,11 +373,12 @@ extract.quad.params.from.bezier.fit <- function(p) {
     p2[1], 2 * (p2[2] - p2[1]), p2[1] - 2 * p2[2] + p2[3])
 }
 
-manifold.clustering.quadr.2 <- function(A, K = 2, d = 2, 
+manifold.clustering.quadr.2 <- function(A, K = 2, p = 2, q = 0, 
                                         method = 'bezier',
                                         initialization = 'random',
                                         intercept = TRUE, 
-                                        maxit = 200) {
+                                        maxit = 200,
+                                        eps.isomap = 1e-2) {
   n <- nrow(A)
   
   if (length(initialization) == n) {
@@ -389,11 +390,13 @@ manifold.clustering.quadr.2 <- function(A, K = 2, d = 2,
   } else {
     stop('choose a valid initialization')
   }
-  X <- embedding(A, d, 0)
+  X <- embedding(A, p, q)
   # if (mean(X[, 1]) < 0) X[, 1] <- -X[, 1]
   z.hat.prev <- sample(seq(K), n, replace = TRUE)
   
   niter <- 0
+  
+  loss <- c()
   
   while (!all(z.hat.prev == z.hat)) {
     z.hat.prev <- z.hat
@@ -410,13 +413,15 @@ manifold.clustering.quadr.2 <- function(A, K = 2, d = 2,
       d2 <- compute.distances.quadr.2(X, theta2)
       distances <- cbind(d1, d2)
     } else if (method == 'bezier') {
-      if (!exists('curve1')) {
-        curve1 <- estimate.bezier.curve.2(X1)
-        curve2 <- estimate.bezier.curve.2(X2)
-      } else {
[…]

### 2022-01-19 curve-fitting-methods/curve-fitting-scratch.Rmd
minor tweaks

@@
 
-```{r, fig.width = 8}
+```{r, fig.width = 4, fig.cap = 'Labels based on output of disassortative DCBM clustering'}
 relations.df <- readr::read_csv('~/dev/potter-network/data/relations.csv') %>% 
@@
 z.init <- kmeans(sweep(Xhat, 1, apply(Xhat, 1, function(x) sqrt(sum(x ^ 2))), `/`), 2)$cluster
+
+qgraph::qgraph(A, groups = factor(z.init))
 ```
@@
 manifold.out <- manifold.clustering.quadr.2(A, 2, p = 1, q = 1, 
-                                            initialization = z.init, 
+                                            # initialization = z.init, 
                                             eps.isomap = .7,
-                                            method = 'bezier')
+                                            intercept = FALSE)
 
@@
 
-t. <- seq(-1, 1, .01)
+t. <- seq(0, 1, .01)
 theta1 <- manifold.out$theta[[1]]
@@
         rep(2, length(t.)))
+Y <- rbind(manifold.out$X[[1]], manifold.out$X[[2]])
+z. <- sort(zhat.manifold)
 

### 2022-01-19 curve-fitting-methods/functions.R

@@ -49,13 +49,19 @@ estimate.one.t.quadr.2 <- function(x1, x2,
   return(max(roots))
 }
 
-estimate.one.t.bezier <- function(x, p, min.t = -2, max.t = 2, tol = 1e-9) {
+estimate.one.t.bezier <- function(x, p, min.t = 0, max.t = 2, intercept = TRUE, tol = 1e-9) {
   polynomial <- function(t, p) {
-    # d <- ncol(p)
-    r <- nrow(p) - 1
-    T <- sapply(seq(0, r), function(s) {
-      choose(r, s) * (1 - t) ^ (r - s) * t ^ s
-    })
+    if (intercept) {
+      r <- nrow(p) - 1
+      T <- sapply(seq(0, r), function(s) {
+        choose(r, s) * (1 - t) ^ (r - s) * t ^ s
+      })
+    } else {
+      r <- nrow(p)
+      T <- sapply(seq(r), function(s) {
+        choose(r, s) * (1 - t) ^ (r - s) * t ^ s
+      })
+    }
     as.numeric(T %*% p)
   }
   l <- function(t) log(as.numeric(crossprod(polynomial(t, p) - x)))
@@ -65,14 +71,15 @@ estimate.one.t.bezier <- function(x, p, min.t = -2, max.t = 2, tol = 1e-9) {
   return(max(minima))
 }
 
-estimate.t.bezier <- function(X, p, t.prev, tol = 1e-9) {
+estimate.t.bezier <- function(X, p, t.prev, intercept = TRUE, tol = 1e-9) {
   if (missing(t.prev)) {
     t.hat <- apply(X, 1, function(x) {
-      estimate.one.t.bezier(x, p, tol = tol)
+      estimate.one.t.bezier(x, p, tol = tol, intercept = intercept)
     })
   } else {
     t.hat <- sapply(seq_along(t.prev), function(i) {
-      estimate.one.t.bezier(X[i, ], p, t.prev[i] - 1, t.prev[i] + 1, tol = tol)
+      estimate.one.t.bezier(X[i, ], p, t.prev[i] - 1, t.prev[i] + 1, 
+                            tol = tol, intercept = intercept)
     })
     loss.new <- sapply(seq_along(t.hat),
                        function(i) loss.one.t(X[i, ], t.hat[i], p))
@@ -172,15 +179,25 @@ estimate.quadr.curve.2 <- function(X, init.params,
               t = t.hat))
 }
 
[…]

### 2022-05-12 dissertation/dissertation.Rmd
trying to move everything to nicer format

---
title: "Community Detection in the Setting of Generalized Random Dot Product Graphs"
author: "John Koo"
geometry: [top=1in, bottom=1in, right=1in, left=1.5in]
day: "1"
month: "December"
year: "2022"
degree: "Doctor of Philosophy"
field: "Statistical Science"
department: "Statistics"
uni: "Indiana University"
location: "Bloomington, IN"
chair: "Michael W. Trosset, Ph.D."
committee2: "Minh Tang, Ph.D."
committee3: "Julia Fukuyama, Ph.D."
committee4: "Roni Khardon, Ph.D."
committee5: "Fangzheng Xie, Ph.D."
bibliography: yourbibfile.bib
multipaper: true
output:
  dissertateUSU::dissertateUSU:
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args: [ "--csl", "apa7.csl" ]
---

```{r global_options, include=FALSE}
## This saves you from having to do this for every chunk
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
library(dissertateUSU)
```

<!-- Abstract -->
\newpage
\pagestyle{plain} <!-- can use \pagestyle{fancy} for a horizontal line in header -->
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{ABSTRACT}
\addcontentsline{toc}{section}{Abstract}

<!-- \abstracttitle -->
\doublespacing

Your abstract words go here. The line below puts the total number of pages at the end of your abstract (as required).

\begin{flushright}(\pageref{LastPage} pages)\end{flushright}

<!-- Dedication -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{DEDICATION}
\addcontentsline{toc}{section}{Dedication}

Dedicate it.

<!-- Acknowledgments -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\chapter*{ACKNOWLEDGMENTS}
\addcontentsline{toc}{section}{Acknowledgments}

Acknowledge those acknowledgable individuals and things.

<!-- Front Matter -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}
\tableofcontents

\newpage
\pagenumbering{arabic}

<!-- Chapter 1 -->
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Block Models for Community Detection}

```{r, child="Chapter1.Rmd"}
```

<!-- Chapter 2 -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Random Dot Product Graphs and Generalized Random Dot Product Graphs}

```{r, child="Chapter2.Rmd"}
```

<!-- Chapter 3 -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Popularity Adjusted Block Models are Generalized Random Dot Product Graphs}

```{r, child="Chapter3.Rmd"}
```

<!-- Chapter 4 -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter{Generalized Random Dot Product Graphs with Community Structure}

```{r, child="Chapter4.Rmd"}
```

<!-- References -->
\FloatBarrier
\newpage
\fancyhead[R]{\thepage}
\fancyfoot[C]{}

\chapter*{REFERENCES}

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.4in}
\setlength{\parskip}{6pt}
\noindent

### 2022-05-18 draft/draft.Rmd
starting over with new layout/formatting

---
output:
  pdf_document:
    includes:
      before_body: title.sty
    keep_tex: true
    citation_package: natbib
    number_sections: yes
# output:
#   bookdown::pdf_document2:
#     citation_package: natbib
#     number_citations: yes
# output: html_document
# output: rticles::rss_article
fontsize: 11pt
geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
bibliography: bibliography.bib
abstract: |
  asdf
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      # eval = FALSE,
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 3, 
                      fig.width = 5, 
                      fig.dpi = 300)

options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')

import::from(magrittr, `%>%`)
library(ggplot2)
```

```{=tex}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
```
# Introduction

## Graphs and Representations of Network Data

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
This has resulted in various challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors. 
To illustrate this challenge, consider the typical approach to building a statistical or machine learning model. 
Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is a vector of $p$ features and each column is a set of $n$ feature measurements. 
An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector. 
For instance, $\phi(x)$ might first compute the distance from $x$ to one of $K$ points in $\mathbb{R}^p$ and assign $x$ to the label of the nearest point. 
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm or Gaussian mixture models (in the case of unsupervised learning). 
However, this type of approach is incompatible with data that are represented as graphs, in which each observation is not a vector of numerical features but a set of relationships to other observations. 
Instead of feature vector $x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top$, we observe $a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}$ where each $a_{ij}$ is object $i$'s relationship to object $j$.

A more formal description of graph data is as follows: 
Suppose we observe a network of $n$ objects and pairwise relationships between them. 
This network is represented by a graph object $G = (V, E)$ with vertex set $V$, representing the objects, and edge set $E$, representing the pairwise relationships. 
The numeric representation of these data is in the form of *affinity matrix* $A \in \mathbb{R}^{n \times n}$ in which each $A_{ij}$ is object $i$'s relationship to object $j$. 
We assume that the entries of $A$ represent affinities or similarities, i.e., the higher the value of $A_{ij}$, the stronger the relationship $i$ has to $j$. 
If $A_{ij} = 0$, then $i$ has no direct relationship to $j$. 
$A$ is symmetric if it represents an undirected graph in which the relationship from $i$ to $j$ is the same as the relationship from $j$ to $i$. 
$A$ is binary, i.e., $A \in \{0, 1\}^{n \times n}$, if it represents an unweighted graph in which edges either exist or don't exist. 
If $A$ is binary, we call it an *adjacency matrix*.

## Probabilistic Models for Graphs

Given a sample or dataset, a typical analysis tasks are statistical inference, or the estimation of various parameters assuming the data come from a random distribution or process. 
These estimated parameters are often then used for making predictions or deriving insights about the population. 
For example, when fitting a Gaussian mixture model, the data are first assumed to come from a mixture of Gaussians. 
The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights. 
The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction for which mixture a new observation belongs to. 
In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.

Within the scope of this dissertation, we focus primarily on unweighted and undirected graphs without self-loops. 
The adjacency matrix that describes such a graph is binary, symmetric, and hollow. 
In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., $A_{ij} \indep P_{ij}$ for some $P_{ij} \in [0, 1]$ for each $i < j$ (setting $A_{ji} = A_{ij}$ since $A$ is symmetric, and $A_{ii} = 0$ since $A$ is hollow). 
Then similar to how the edges are compiled into an adjacency matrix $A$, the edge probabilities can be compiled into an edge probability matrix $P \in [0, 1]^{n \times n}$. 
In order to provide structure

## Block Models for Community Detection

### The Stochastic Block Model

### Generalizations of the Stochastic Block Model: the Degree Corrected Block Model and the Popularity Adjusted Block Model

### Likelihood Maximization Approaches to Block Models

\newpage

# Random Dot Product Graphs and Generalized Random Dot Product Graphs

## Connecting the Stochastic Block Model and the Degree Corrected Block Model to the Generalized Random Dot Product Graph

\newpage

# Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

\newpage

# Generalized Random Dot Product Graphs with Community Structure

\newpage

\section*{Appendix A}

\newpage

# References

### 2022-05-18 draft/draft.tex
starting over with new layout/formatting

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\doublespacing

\pagenumbering{gobble}

%\begin{titlepage}
\begin{center}
\LARGE{\textsc{Community Detection in the Setting of Generalized Random Dot Product Graphs}}\\
\vspace*{6\baselineskip}
\normalsize{John Koo}\\
\vspace*{10\baselineskip}
\singlespacing
\normalsize{Submitted to the faculty of the Univesity Graduate School \\
in partial fulfillment of the requirements for the degree \\
Doctor of Philosophy \\
in the Department of Statistics, \\
Indiana University \\
December 2022}
\vspace*{3\baselineskip}

\end{center}

\pagenumbering{roman}
\thispagestyle{empty}

\newpage

\singlespacing

Accepted by the Graduate Faculty, Indiana University, in partial fulfillment of the requirements for the degree of Doctor of Philosophy.

\vspace*{6\baselineskip}

\begin{tabular}{@{}p{1in}p{4in}@{}}
Approved: & \hrulefill \\
& Michael W. Trosset, Ph.D. \\
\\
\\
& \hrulefill \\
& Minh Tang, Ph.D. \\
\\
\\
& \hrulefill \\
& Julia Fukuyama, Ph.D. \\
\\
\\
& \hrulefill \\
& Roni Khardon, Ph.D. \\
\\
\\
& \hrulefill \\
& Fangzheng Xie, Ph.D. \\
\end{tabular}

\vspace*{16\baselineskip}

\raggedright

December 1, 2022

\newpage

\doublespacing

\begin{center}
\LARGE{\bf{Acknowledgements}}
\end{center}

\hypersetup{linkcolor = black}

\newpage

\begin{center}
\LARGE{Abstract}
\end{center}

\vspace*{2\baselineskip}

\normalsize

Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision. One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels). Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs. In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure), first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then applying clustering methods for these feature vector representations, and finally deriving the asymptotic properties of these methods.

\newpage
\tableofcontents
\addcontentsline{toc}{section}{\contentsname}

\newpage
\pagenumbering{arabic}
\hypersetup{linkcolor = blue}

\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}
\newcommand{\BG}{\text{BernoulliGraph}}
\newcommand{\Cat}{\text{Categorical}}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{example}
\newtheorem*{example}{Example}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{graphs-and-representations-of-network-data}{%
\subsection{Graphs and Representations of Network
Data}\label{graphs-and-representations-of-network-data}}

Graph and network data have become increasingly widespread in various
fields including sociology, neuroscience, biostatistics, and computer
science. This has resulted in various challenges for researchers who
rely on traditional statistical and machine learning methods that are
incompatible with graph data and instead assume that the data exist as
feature vectors. To illustrate this challenge, consider the typical
approach to building a statistical or machine learning model. Data are
often represented as an \(n \times p\) matrix
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which
each row \(x_i \in \mathbb{R}^p\) is a vector of \(p\) features and each
column is a set of \(n\) feature measurements. An analysis task for
these data might be to come up with a classification model
\(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical
values of each feature of a vector. For instance, \(\phi(x)\) might
first compute the distance from \(x\) to one of \(K\) points in
\(\mathbb{R}^p\) and assign \(x\) to the label of the nearest point.
Examples of this include linear discriminant analysis (in the case of
supervised learning) and Lloyd's algorithm or Gaussian mixture models
(in the case of unsupervised learning). However, this type of approach
is incompatible with data that are represented as graphs, in which each
observation is not a vector of numerical features but a set of
relationships to other observations. Instead of feature vector
\(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top\),
we observe
\(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}\) where
each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).

A more formal description of graph data is as follows: Suppose we
observe a network of \(n\) objects and pairwise relationships between
them. This network is represented by a graph object \(G = (V, E)\) with
vertex set \(V\), representing the objects, and edge set \(E\),
representing the pairwise relationships. The numeric representation of
these data is in the form of \emph{affinity matrix}
\(A \in \mathbb{R}^{n \times n}\) in which each \(A_{ij}\) is object
\(i\)'s relationship to object \(j\). We assume that the entries of
\(A\) represent affinities or similarities, i.e., the higher the value
of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\). If
\(A_{ij} = 0\), then \(i\) has no direct relationship to \(j\). \(A\) is
symmetric if it represents an undirected graph in which the relationship
from \(i\) to \(j\) is the same as the relationship from \(j\) to \(i\).
\(A\) is binary, i.e., \(A \in \{0, 1\}^{n \times n}\), if it represents
an unweighted graph in which edges either exist or don't exist. If \(A\)
is binary, we call it an \emph{adjacency matrix}.

\hypertarget{probabilistic-models-for-graphs}{%
\subsection{Probabilistic Models for
Graphs}\label{probabilistic-models-for-graphs}}

Given a sample or dataset, a typical analysis tasks are statistical
inference, or the estimation of various parameters assuming the data
come from a random distribution or process. These estimated parameters
are often then used for making predictions or deriving insights about
the population. For example, when fitting a Gaussian mixture model, the
data are first assumed to come from a mixture of Gaussians. The model
fitting process then involves estimating the means and standard
deviations of each Gaussian component, along with the mixture weights.
The resulting model provides insight into where each mixture component
is located, how disperse each component is, and how the data are
distributed between the components, as well as a prediction for which
mixture a new observation belongs to. In order to perform a similar type
of analysis for graphs, we must first define probability distributions
from which such data can be sampled.

Within the scope of this dissertation, we focus primarily on unweighted
and undirected graphs without self-loops. The adjacency matrix that
describes such a graph is binary, symmetric, and hollow. In this
setting, a plausible model is to sample each edge independently from a
Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
\(P_{ij \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
Then similar to how the edges are compiled into an adjacency matrix
\(A\), the edge probabilities can be compiled into an edge probability
matrix \(P \in [0, 1]^{n \times n}\). In order to provide structure

\hypertarget{block-models-for-community-detection}{%
\subsection{Block Models for Community
Detection}\label{block-models-for-community-detection}}

\hypertarget{the-stochastic-block-model}{%
\subsubsection{The Stochastic Block
Model}\label{the-stochastic-block-model}}

\hypertarget{generalizations-of-the-stochastic-block-model-the-degree-corrected-block-model-and-the-popularity-adjusted-block-model}{%
\subsubsection{Generalizations of the Stochastic Block Model: the Degree
Corrected Block Model and the Popularity Adjusted Block
Model}\label{generalizations-of-the-stochastic-block-model-the-degree-corrected-block-model-and-the-popularity-adjusted-block-model}}

\hypertarget{likelihood-maximization-approaches-to-block-models}{%
\subsubsection{Likelihood Maximization Approaches to Block
Models}\label{likelihood-maximization-approaches-to-block-models}}

\newpage

\hypertarget{random-dot-product-graphs-and-generalized-random-dot-product-graphs}{%
\section{Random Dot Product Graphs and Generalized Random Dot Product
Graphs}\label{random-dot-product-graphs-and-generalized-random-dot-product-graphs}}

\hypertarget{connecting-the-stochastic-block-model-and-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}{%
\subsection{Connecting the Stochastic Block Model and the Degree
Corrected Block Model to the Generalized Random Dot Product
Graph}\label{connecting-the-stochastic-block-model-and-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}}

\newpage

\hypertarget{popularity-adjusted-block-models-are-generalized-random-dot-product-graphs}{%
\section{Popularity Adjusted Block Models are Generalized Random Dot
Product
Graphs}\label{popularity-adjusted-block-models-are-generalized-random-dot-product-graphs}}

\newpage

\hypertarget{generalized-random-dot-product-graphs-with-community-structure}{%
\section{Generalized Random Dot Product Graphs with Community
Structure}\label{generalized-random-dot-product-graphs-with-community-structure}}

\newpage

\section*{Appendix A}

\newpage

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}

### 2022-05-18 draft/draft.Rmd
minor changes based on meeting

@@
 
-### Likelihood Maximization Approaches to Block Models
+### Review of Likelihood Maximization Approaches to Block Models
 
@@
 
-## Connecting the Stochastic Block Model and the Degree Corrected Block Model to the Generalized Random Dot Product Graph
+## Definitions
+
+## Connecting the Stochastic Block Model to the Generalized Random Dot Product Graph
+
+## Connecting the Degree Corrected Block Model to the Generalized Random Dot Product Graph
 

### 2022-05-18 draft/draft.tex
minor changes based on meeting

@@
 
-\newcommand{\diag}{\text{diag}}
-\newcommand{\tr}{\text{Tr}}
-\newcommand{\blockdiag}{\text{blockdiag}}
-\newcommand{\indep}{\stackrel{\text{ind}}{\sim}}
-\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
-\newcommand{\Bernoulli}{\text{Bernoulli}}
-\newcommand{\Betadist}{\text{Beta}}
-\newcommand{\BG}{\text{BernoulliGraph}}
-\newcommand{\Cat}{\text{Categorical}}
-\newtheorem{definition}{Definition}
+\newcommand{\diag}{\mathrm{diag}}
+\newcommand{\tr}{\mathrm{Tr}}
+\newcommand{\blockdiag}{\mathrm{blockdiag}}
+\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
+\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
+\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
+\newcommand{\Betadist}{\mathrm{Beta}}
+\newcommand{\BG}{\mathrm{BernoulliGraph}}
+\newcommand{\Uniform}{\mathrm{Uniform}}
+\newcommand{\PABM}{\mathrm{PABM}}
+\newcommand{\RDPG}{\mathrm{RDPG}}
+\newcommand{\GRDPG}{\mathrm{GRDPG}}
+\newcommand{\Multinomial}{\mathrm{Multinomial}}
 \newtheorem{theorem}{Theorem}
 \newtheorem{lemma}{Lemma}
+\newtheorem{proposition}{Proposition}
 \theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-\theoremstyle{example}
-\newtheorem*{example}{Example}
+\newtheorem{remark}{Remark}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}
+\newtheorem{example}{Example}
+\newcommand{\dd}{\mathrm{d}}
+\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
+\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
 
@@
 Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
-\(P_{ij \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
+\(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
 since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
@@
 
-\hypertarget{likelihood-maximization-approaches-to-block-models}{%
-\subsubsection{Likelihood Maximization Approaches to Block
-Models}\label{likelihood-maximization-approaches-to-block-models}}
+\hypertarget{review-of-likelihood-maximization-approaches-to-block-models}{%
+\subsubsection{Review of Likelihood Maximization Approaches to Block
+Models}\label{review-of-likelihood-maximization-approaches-to-block-models}}
 
@@
 
-\hypertarget{connecting-the-stochastic-block-model-and-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}{%
-\subsection{Connecting the Stochastic Block Model and the Degree
-Corrected Block Model to the Generalized Random Dot Product
-Graph}\label{connecting-the-stochastic-block-model-and-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}}
+\hypertarget{definitions}{%
+\subsection{Definitions}\label{definitions}}
+
+\hypertarget{connecting-the-stochastic-block-model-to-the-generalized-random-dot-product-graph}{%
+\subsection{Connecting the Stochastic Block Model to the Generalized
+Random Dot Product
+Graph}\label{connecting-the-stochastic-block-model-to-the-generalized-random-dot-product-graph}}
+
+\hypertarget{connecting-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}{%
+\subsection{Connecting the Degree Corrected Block Model to the
+Generalized Random Dot Product
+Graph}\label{connecting-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}}
 

### 2022-07-19 draft/draft.Rmd
reworking some of the outline/structure

@@
 ```{=tex}
+\setlength{\parindent}{10.0pt}
 \newcommand{\diag}{\mathrm{diag}}
@@
 Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
-This has resulted in various challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors.
-To illustrate this challenge, consider the typical approach to building a statistical or machine learning model. 
-Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is a vector of $p$ features and each column is a set of $n$ feature measurements.
+This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors.
+To illustrate this, consider the typical approach to building a statistical or machine learning model. 
+Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements.
 An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector. 
@@
 
-Given a sample or dataset, a typical analysis tasks are statistical inference, or the estimation of various parameters assuming the data come from a random distribution or process. 
+Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters assuming the data come from a random distribution or process. 
 These estimated parameters are often then used for making predictions or deriving insights about the population. 
@@
 The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights. 
-The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction for which mixture a new observation belongs to.
+The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction indicating to which mixture a new observation belongs.
 In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.
 
-Within the scope of this dissertation, we focus primarily on unweighted and undirected graphs without self-loops. 
-The adjacency matrix that describes such a graph is binary, symmetric, and hollow. 
+Within the scope of this dissertation, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs. 
+The adjacency matrix that describes these graphs is binary, symmetric, and hollow. 
 In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., $A_{ij} \indep P_{ij}$ for some $P_{ij} \in [0, 1]$ for each $i < j$ (setting $A_{ji} = A_{ij}$ since $A$ is symmetric, and $A_{ii} = 0$ since $A$ is hollow).
 Then similar to how the edges are compiled into an adjacency matrix $A$, the edge probabilities can be compiled into an edge probability matrix $P \in [0, 1]^{n \times n}$. 
-In order to provide structure
+This type of graph model is defined as a *Bernoulli graph*.
+
+\begin{definition}[Bernoulli graph]
+\end{definition}
+
+Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters is equal to the number of observations. 
+Additional structure must be introduced. 
+One such structured Bernoulli graph model is the $\ER$ graph, which is defined as follows.
+
+\begin{definition}[$\ER$ graph]
+\end{definition}
 
@@
 
-### Review of Likelihood Maximization Approaches to Block Models
+### The Hierarchy of Block Models
+
+### Review of Traditional Statistical Inference Approaches to Block Model Estimation
 

### 2022-07-19 draft/draft.tex
reworking some of the outline/structure

@@
 
+\setlength{\parindent}{10.0pt}
 \newcommand{\diag}{\mathrm{diag}}
@@
 fields including sociology, neuroscience, biostatistics, and computer
-science. This has resulted in various challenges for researchers who
-rely on traditional statistical and machine learning methods that are
+science. This has resulted in challenges for researchers who rely on
+traditional statistical and machine learning methods that are
 incompatible with graph data and instead assume that the data exist as
-feature vectors. To illustrate this challenge, consider the typical
-approach to building a statistical or machine learning model. Data are
-often represented as an \(n \times p\) matrix
+feature vectors. To illustrate this, consider the typical approach to
+building a statistical or machine learning model. Data are often
+represented as an \(n \times p\) matrix
 \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which
-each row \(x_i \in \mathbb{R}^p\) is a vector of \(p\) features and each
-column is a set of \(n\) feature measurements. An analysis task for
-these data might be to come up with a classification model
+each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features
+and each column is a set of \(n\) feature measurements. An analysis task
+for these data might be to come up with a classification model
 \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical
@@
 
-Given a sample or dataset, a typical analysis tasks are statistical
+Given a sample or dataset, a typical analysis task is statistical
 inference, or the estimation of various parameters assuming the data
@@
 is located, how disperse each component is, and how the data are
-distributed between the components, as well as a prediction for which
-mixture a new observation belongs to. In order to perform a similar type
-of analysis for graphs, we must first define probability distributions
-from which such data can be sampled.
+distributed between the components, as well as a prediction indicating
+to which mixture a new observation belongs. In order to perform a
+similar type of analysis for graphs, we must first define probability
+distributions from which such data can be sampled.
 
 Within the scope of this dissertation, we focus primarily on unweighted
-and undirected graphs without self-loops. The adjacency matrix that
-describes such a graph is binary, symmetric, and hollow. In this
-setting, a plausible model is to sample each edge independently from a
-Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
+and undirected graphs without self-loops, with a brief discussion on
+generalizing these methods to weighted or directed graphs. The adjacency
+matrix that describes these graphs is binary, symmetric, and hollow. In
+this setting, a plausible model is to sample each edge independently
+from a Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
 \(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
@@
 \(A\), the edge probabilities can be compiled into an edge probability
-matrix \(P \in [0, 1]^{n \times n}\). In order to provide structure
+matrix \(P \in [0, 1]^{n \times n}\). This type of graph model is
+defined as a \emph{Bernoulli graph}.
+
+\begin{definition}[Bernoulli graph]
+\end{definition}
+
+Statistical inference is not possible on a general Bernoulli graph with
+arbitrary edge probabilities since the number of parameters is equal to
+the number of observations. Additional structure must be introduced. One
+such structured Bernoulli graph model is the \(\ER\) graph, which is
+defined as follows.
+
+\begin{definition}[$\ER$ graph]
+\end{definition}
 
@@
 
-\hypertarget{review-of-likelihood-maximization-approaches-to-block-models}{%
-\subsubsection{Review of Likelihood Maximization Approaches to Block
-Models}\label{review-of-likelihood-maximization-approaches-to-block-models}}
+\hypertarget{the-hierarchy-of-block-models}{%
+\subsubsection{The Hierarchy of Block
+Models}\label{the-hierarchy-of-block-models}}
+
+\hypertarget{review-of-traditional-statistical-inference-approaches-to-block-model-estimation}{%
+\subsubsection{Review of Traditional Statistical Inference Approaches to
+Block Model
+Estimation}\label{review-of-traditional-statistical-inference-approaches-to-block-model-estimation}}
 

### 2022-08-22 draft/draft.Rmd
still writing

@@
 output:
-  pdf_document:
+  # pdf_document:
+  bookdown::pdf_document2:
     includes:
@@
 - \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
-- \setcitestyle{numbers,square,comma}
+- \setcitestyle{year,comma}
 - \usepackage{verbatim}
@@
 bibliography: bibliography.bib
+biblio-style: apalike
 abstract: |
@@
 library(ggplot2)
+library(ggraph)
+source('~/dev/pabm-grdpg/functions.R')
 ```
@@
 \newcommand{\Multinomial}{\mathrm{Multinomial}}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{proposition}{Proposition}
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
 \newcommand{\dd}{\mathrm{d}}
@@
 \newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
+\newcommand{\SBM}{\mathrm{SBM}}
+\newcommand{\DCBM}{\mathrm{DCBM}}
+\newcommand{\argmin}{\mathrm{argmin}}
 ```
+
 # Introduction
@@
 Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements.
-An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector. 
-For instance, $\phi(x)$ might first compute the distance from $x$ to one of $K$ points in $\mathbb{R}^p$ and assign $x$ to the label of the nearest point. 
-Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm or Gaussian mixture models (in the case of unsupervised learning). 
-However, this type of approach is incompatible with data that are represented as graphs, in which each observation is not a vector of numerical features but a set of relationships to other observations.
-Instead of feature vector $x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top$, we observe $a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}$ where each $a_{ij}$ is object $i$'s relationship to object $j$.
+An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$.
+For instance, $\phi(x)$ might first compute the distances from $x$ to $K$ points in $\mathbb{R}^p$ and then assign $x$ to the label of the nearest point. 
+Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning).
+However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
+Instead of feature vector $x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top \in \mathbb{R}^p$, we observe $a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
+For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
 
-A more formal description of graph data is as follows: 
+\begin{example}
+$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
+\begin{equation}
+\label{eq:kmeans}
+W(x_1, ..., x_n) = \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
+\end{equation}
+in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments we wish to optimize for, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
+A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration labels are updated according to which centroid they are closest to and centroids are updated via the sample mean of each cluster.
+
+If instead of vectors in $\mathbb{R}^p$ we have a graph, it is not obvious how to translate $K$-means clustering to these data. 
+In particular, there is no inherent notion of centroid or sample mean for graph data. 
+However, there are inherent notions of distances among vectors, such as shortest path distance, expected commute time, or resistance distance. 
+One way to adapt $K$-means to graph data is to use an equivalent objective function as Eq.~\ref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
+$$
+\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} d(x_i, x_j)^2.
+$$
+Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
+If the data live in a Euclidean vector space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
+or if the data are represented as a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
+While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
+An alternative algorithm is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.
+
+Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
+\end{example}
+
+We now provide a more formal description of graph data: 
 Suppose we observe a network of $n$ objects and pairwise relationships between them. 
-This network is represented by a graph object $G = (V, E)$ with vertex set $V$, representing the objects, and edge set $E$, representing the pairwise relationships. 
-The numeric representation of these data is in the form of *affinity matrix* $A \in \mathbb{R}^{n \times n}$ in which each $A_{ij}$ is object $i$'s relationship to object $j$. 
+This network is represented by a graph $G = (V, E)$ with vertex set $V = \{v_1, ..., v_n\}$, representing the $n$ objects, and edge set $E$, representing the up to $n (n - 1) / 2$ pairwise relationships (assuming that there are no self-loops).
+The numeric representation of these data is in the form of *affinity matrix* $A \in \mathbb{R}^{n \times n}$ in which each $A_{ij}$ represents object $i$'s relationship to object $j$. 
 We assume that the entries of $A$ represent affinities or similarities, i.e., the higher the value of $A_{ij}$, the stronger the relationship $i$ has to $j$. 
@@
 
+\begin{example}
+\label{ex:uk-faculty}
+\citet{PhysRevE.77.016107} constructed a friendship network among $81$ faculty from various schools at a university in the UK. 
+These data are represented as a graph in which each vertex is a faculty member and edges between pairs of vertices indicate whether the two faculty are friends. 
+The corresponding adjacency matrix $A \in \{0, 1\}^{81 \times 81}$ has zeros along the diagonal since there are no self-loops, and $A_{ij} = A_{ji} = 1$ if it is observed that the $i^{th}$ and $j^{th}$ faculty members are friends.
+The following is a visualization of this graph, with the vertices labeled by school affiliation.
+
+```{r friendship-network, fig.cap = 'Friendship network of 81 faculty at a UK university. The vertices are labeled by school affiliation.', cache = TRUE}
+data('UKfaculty', package = 'igraphdata')
+A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
+A <- sign(A + t(A))
+z <- factor(igraph::vertex_attr(UKfaculty)$Group)
+n <- length(z)
+
+ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = z), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
+\end{example}
+
 ## Probabilistic Models for Graphs
 
-Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters assuming the data come from a random distribution or process. 
+Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process. 
 These estimated parameters are often then used for making predictions or deriving insights about the population. 
@@
 
-Within the scope of this dissertation, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs. 
+Within the scope of this work, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs. 
 The adjacency matrix that describes these graphs is binary, symmetric, and hollow. 
-In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., $A_{ij} \indep P_{ij}$ for some $P_{ij} \in [0, 1]$ for each $i < j$ (setting $A_{ji} = A_{ij}$ since $A$ is symmetric, and $A_{ii} = 0$ since $A$ is hollow).
+In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., $A_{ij} \indep \Bernoulli(P_{ij})$ for some $P_{ij} \in [0, 1]$ for each $i < j$ (setting $A_{ji} = A_{ij}$ since $A$ is symmetric, and $A_{ii} = 0$ since $A$ is hollow).
 Then similar to how the edges are compiled into an adjacency matrix $A$, the edge probabilities can be compiled into an edge probability matrix $P \in [0, 1]^{n \times n}$. 
-This type of graph model is defined as a *Bernoulli graph*.
+This type of graph model is defined as a *Bernoulli graph* (also called an inhomogeneous $\ER$ graph). 
+If $A$ is the adjacency matrix of a Bernoulli graph with edge probability matrix $P$, we denote $A \sim \BG(P)$. 
+If the vertices and edge probabilities are sampled as a sequence of $n$ i.i.d. random variables from probability distribution $F$, then we denote $A \sim \BG(F, n)$.
 
-\begin{definition}[Bernoulli graph]
+Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
+Additional structure must be introduced. 
+One such structured Bernoulli graph model is the $\ER$ graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
+
+\begin{definition}[$\ER$ graph]
+\label{def:erdos-renyi}
+Let $P$ be an $n \times n$ matrix such that each $P_{ij} \equiv \theta \in [0, 1]$ is a constant. 
+Then $A \sim \BG(P)$ is an $\ER$ graph. 
 \end{definition}
 
-Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters is equal to the number of observations. 
-Additional structure must be introduced. 
-One such structured Bernoulli graph model is the $\ER$ graph, which is defined as follows.
+\begin{example}[Maximum likelihood estimator for the $\ER$ graph]
+One possible estimator for the lone parameter of an $\ER$ graph, $\theta$, is the maximum likelihood estimator, which is found by maximizing the log-likelihood function $\ell(\theta; A) = \log \theta \sum\limits_{i < j} A_{ij} + \log (1 - \theta) \sum\limits_{i < j} (1 - A_{ij})$.
+This can be solved directly by setting $\frac{d \ell}{d \theta} = 0$, which yields $\hat{\theta} = \frac{2 |E|}{n (n-1)}$.
+\end{example}
 
-\begin{definition}[$\ER$ graph]
+$\ER$ graphs lie on the opposite end of the spectrum in that they restrict the model to a single parameter, which does not result in a very interesting model or reflect many networks observed in real data.
+Much work has been done in developing various Bernoulli graph models that strike a balance between structure and flexibility. 
+Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called *block models*, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called *latent space models*.
+
+**[Maybe insert something here about the exponential family of random graph models and how, like mixture distributions, block models do not fit within this family?]**
+
+\begin{example}
+Consider a graph in which the vertices represent a collection of $n$ planets with intelligent life sending out radio waves, and the existence of an edge between vertices $i$ and $j$ represents whether planets $i$ and $j$ have established contact.
+Since radio waves decay according to the inverse-square law, a plausible model for the edge probability between two vertices is 
+$$P_{ij} = \frac{C \omega_i \omega_j}{d_{ij}^2}$$ 
+where $\omega_i$ is the $i^{th}$ planet's radio signal strength, $d_{ij}$ is the Euclidean distance between planets $i$ and $j$, and $C$ is a normalizing constant. 
+Let $P$ be the $n \times n$ matrix of these edge probabilities. 
+Then $A \sim \BG(P)$ represents the network of planets that have made contact.
+
+To estimate $\omega = \begin{bmatrix} \omega_1 & \cdots & \omega_n \end{bmatrix}^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
+For simplicity, set $C = 1$. 
+Then the log-likelihood function can be written as:
+$$
+\ell(\omega) = 
+\sum_{i < j} A_{ij} \log \bigg( \frac{\omega_i \omega_j}{d_{ij}^2 -
+\omega_i \omega_j} \bigg) + \log (d_{ij}^2 - \omega_i \omega_j) + const.
+$$
+The partial derivatives with respect to each $\omega_i$ are:
+$$
+\frac{\partial \ell}{\partial \omega_i} = \sum_{j \neq i} \frac{A_{ij}}{\omega_i} - \frac{\omega_j (1 - A_{ij})}{d_{ij}^2 - \omega_i \omega_j}.
+$$
+Setting the gradient to zero yields $n$ equations with $n$ unknowns, but it is not entirely obvious how to solve this system of equations. 
+Alternatively, we can use gradient ascent or some other numerical optimization method, since this particular log-likelihood function is concave. 
+\end{example}
+
+## Contributions of this work
+
+In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.
+The contributions of this work are as follows. 
+First, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
+Using estimators with well-established consistency properties for the generalized random dot product graph, we develop consistent community detection and parameter estimation algorithms for the popularity adjusted block model.
+
+\newpage
+
+# Block Models for Community Detection {#sec:blockmodel}
+
+Network analysis is often concerned with community detection. 
+This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices $i$ and $j$ depend on the communities to which vertices $v_i$ and $v_j$ belong.
+More formally, one assigns each vertex $v_i$ a community label $z_i$ and assumes a Bernoulli graph in which $P_{ij} = g(z_i, z_j, \theta_i, \theta_j)$ for some function $g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]$ and $\theta_i \in \mathbb{R}^d$ is a vector of real-valued parameters associated with each vertex.
+Within the context of this work, we will call such models *block models*. 
+This type of model frmaework allows us to frame the goal of community detection as a statistical inference problem: 
+identify the true community labels, up to permutation of the labels.
+
+In this section, we explore the three main types of block models: 
+the stochastic block model \citep{doi:10.1080/0022250X.1971.9989788}, the degree corrected block model \citep{Karrer_2011}, and the popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1}.
+The main focus is on how these models are connected, particularly in how they are nested models \citep{Noroozi2022}, as well as community detection and parameter estimation via likelihood maximization.
+We will later return to these block models and the relationship to another class of random graph models in sections \@ref(sec:grdpg) and \@ref(sec:pabm-grdpg).
+
+## The Stochastic Block Model {#sec:sbm}
+
+The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities has a constant edge probability $\theta_{k \ell}$.
+We now give the formal definition of the SBM within the context of Bernoulli graphs:
+
+\begin{definition}[Stochastic block model]
+\label{def:sbm}
+Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$). 
+Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
+If the edge probability matrix $P$ for this graph is such that 
+$P_{ij} = \theta_{z_i, z_j}$ and $A \sim \BG(P)$, then $G$ is a stochastic block model. 
 \end{definition}
 
-## Block Models for Community Detection
+There are a few conventions for exactly how a graph is sampled as an SBM. 
+Within the context of this work, we assume that for a particular SBM, the number of communities, $K$, and the community edge probabilities, $\theta_{k \ell}$, are fixed. 
+Then some possible ways of sampling from an SBM are as follows:
 
-### The Stochastic Block Model
+1. Assume the size of the graph $n$ is fixed. 
+Then each vertex $v_i$ has a label $z_i$, and the edge probability matrix $P$ is fixed.
+2. We sample the vertices and edges in sequence and suppose that each vertex $v_i$ comes with a label $z_i$, which determines its edge probabilities to the other vertices.
+3.
+The vertices are sampled in sequence, and each $z_i$ is sampled as $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, $\sum_k^K \alpha_k = 1$.
+The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}$
 
-### Generalizations of the Stochastic Block Model: the Degree Corrected Block Model and the Popularity Adjusted Block Model
+An extension to these sampling schema include Bayesian models, e.g., each $\theta_{k \ell}$ is sampled from a distribution with support $[0, 1]$.
+
+\begin{example}[Stochastic block model with $K = 2$ communities]
+\label{ex:assort-sbm}
+We construct an SBM with two communities in which $\theta_{11} = 1/2$, $\theta_{12} = 1/8$, and $\theta_{22} = 1/4$. 
+Note that in this example, the within-community edge probabilities are greater than the between-community edge probability. 
+A realization of this graph with $n_1 = n_2 = 32$ is illustrated by figure \ref{fig:assort-sbm}.
+
+```{r assort-sbm, fig.width = 3, , fig.height = 2, fig.cap = 'Two-community stochastic block model with $\\theta_{11}=1/2$, $\\theta_{22}=1/4$, and $\\theta_{12} = \\theta_{21} = 1/8$.', cache = TRUE}
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+p <- 1/2
+q <- 1/4
+r <- 1/8
+P <- matrix(r, nrow = n, ncol = n)
+P[seq(n1), seq(n1)] <- p
+P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
+A <- draw.graph(P)
+# qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
+ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
+\end{example}
+
+In most conceptions of graph models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}. 
+For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the graph appears to confirm this.
+However, SBMs are not restricted to this type of community structure. 
+We could just as easily construct an example in which $\theta_{11}$ and $\theta_{22}$ are less than $\theta_{12}$.
+Within the context of the SBM, we say that a graph is assortative if $P$ is positive semidefinite and disassortative otherwise, which roughly correspond to communities with higher within-community edge probabilities and higher between-community edge probabilities, respectively.
+For example, the SBM in example \ref{ex:assort-sbm} is assortative and has edge probability matrix $P$ that is rank 2 with nonzero eigenvalues $n (\frac{3}{8} \pm \frac{\sqrt{2}}{8}) > 0$.
+In the following example, we describe a network which can be modeled as a disassortative SBM.
+
+\begin{example}[Dating network as a disassortative stochastic block model]
+\label{ex:sbm-dating}
+Consider an undirected graph in which each vertex $v_i$ represents users of an online dating service, each label $z_i$ represents the user's gender, and each edge represents a successful match between pairs of users.
+For simplicity, we will restrict the labels to female and male (denoted as $1$ and $2$). 
+If we model this as an SBM, then $\theta_{11}$ is the probability of a match between two female users, $\theta_{22}$ is the probability of a match between two male users, and $\theta_{12} = \theta_{21}$ is the probability of a match between a female user and a male user.
+Based on trends in the United States \citep{gallup-lgbt},
+a plausible set of edge probabilities is $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$, which are used in the sample visualized in figure \ref{fig:dating-sbm}, along with $\alpha_1 = \alpha_2 = 1/2$ and $n = 64$.
+
+```{r dating-sbm, fig.width = 4, , fig.height = 2, fig.cap = 'Stochastic block model of a dating network. This model is disassortative.', cache = TRUE}
+n <- 2 ** 6
+z <- sample(c(1, 2), n, replace = TRUE)
+z <- sort(z)
+n1 <- sum(z == 1)
+n2 <- sum(z == 2)
+z[z == 1] <- 'F'
+z[z == 2] <- 'M'
+theta11 <- .02
+theta22 <- .02
+theta12 <- .2
+P <- matrix(theta12, nrow = n, ncol = n)
+P[seq(n1), seq(n1)] <- theta11
+P[seq(n1 + 1, n), seq(n1 + 1, n)] <- theta22
+A <- draw.graph(P)
+# qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
+ggraph(A, layout = 'auto') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = TRUE) + 
+  labs(colour = 'Gender') + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
+\end{example}
+
+\begin{example}[Assortativity and disassortativity of the two-community SBM]
+Let $P$ be the $n \times n$ probability matrix of a two-community SBM. 
+Then $rank(P) = 2$, and its two nonzero eigenvalues are:
+
+$$
+\lambda = \frac{n}{2} \bigg( 
+  \theta_{11} + \theta_{22} \pm 
+  \sqrt{(\theta_{11} - \theta_{22})^2 + 4 \theta_{12}^2} 
+\bigg).
+$$
+The first eigenvalue is always positive. 
+The second eigenvalue is positive if $\theta_{11} \theta_{22} > \theta_{12}^2$. 
+Thus, $P$ describes an assortative SBM if $\theta_{11} \theta_{22} > \theta_{12}^2$ or 
+a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
+\end{example}
+
+One approach to statistical inference for the SBM is likelihood maximization. 
+The log-likelihood function can be written as:
+\begin{equation}
+\label{eq:sbm-loglik}
+\ell(z, \theta; A) = \sum_{i < j} \sum_{k, \ell} z_{ik} z_{j \ell} \big( 
+A_{ij} \log \theta_{k \ell} - (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big),
+\end{equation}
+where $z_{ik} = I(z_i = k)$. If the labels are drawn from a multinomial distribution, then that can also be incorporated into the complete-data log-likelihood:
+\begin{equation}
+\label{eq:sbm-loglik-full}
+\ell(z, \theta, \alpha; A) = \sum_k \sum_i z_{ik} \log \alpha_k + 
+\sum_{i < j} \sum_{k, \ell} z_{ik} z_{j \ell} \big( 
+A_{ij} \log \theta_{k \ell} - (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big).
+\end{equation}
+
+The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard. 
+Because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the expectation maximization (EM) algorithm \citep{10.2307/2984875}.
+However, due to the $z_{ik} z_{j \ell}$ interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}. 
+While the labels are drawn independently, they are not necessarily conditionally independent. 
+Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form. 
+As an example, we write the EM algorithm for the likelihood function in equation \ref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
+
+\begin{algorithm}
+\label{algo:sbm-em}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}$}
+  \caption{Approximate EM algorithm for the SBM}
+  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}$.\;
+  \While{$\|\nabla \ell\| > \epsilon$} {
+    \For {$i = 1, ..., n$} {
+      \For {$k = 1, ..., K$} {
+        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \theta_{k \ell} + (1 - A_{ij}) \log (1 - \theta_{k \ell})) \bigg)$.\;
+        M-step: $\theta_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
+      }
+    }
+  }
+\end{algorithm}
+
+Applying algorithm \ref{algo:sbm-em} to example \ref{ex:assort-sbm} with initial guesses of $z_{ik} = 0.5$ for each $i, k$, $\hat{\theta}_{11} = \hat{\theta}_{22} = 0.9$, and $\hat{\theta}_{12} = 0.1$ results in a community detection error rate of 4.7\% and parameter estimates $\hat{\theta}_{11} = 0.503$, $\hat{\theta}_{22} = 0.241$, and $\hat{\theta}_{12} = 0.109$, compared to the true parameters $\theta_{11} = 0.5$, $\theta_{22} = 0.25$, and $\theta_{12} = 0.125$.
+The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of $z_{ik} = 0.5$, $\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1$, and $\hat{\theta}_{12} = 0.9$, results in a community detection error rate of 0\% and parameter estimates $\hat{\theta}_{11} = 0.034$, $\hat{\theta}_{22} = 0.018$, and $\hat{\theta}_{12} = 0.201$, compared to the true parameters $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$.
+
+### Random Graph Models and Sparsity
+
+## Generalizations of the Stochastic Block Model: The Degree Corrected Block Model and the Popularity Adjusted Block Model {#sec:dcbm-pabm}
+
+\begin{definition}[Degree corrected block model]
+\label{def:dcbm}
+Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM. 
+Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
+In addition, each vertex $v_i$ has a degree correction parameter $\omega_i \in [0, 1]$. 
+If the edge probability matrix $P$ for this graph is such that $P_{ij} = \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model.
+\end{definition}
+
+\begin{example}[Dating network as a disassortative DCBM]
+\label{ex:dcbm-dating}
+In example \ref{ex:sbm-dating}, we modeled an online dating network as an SBM in which the vertices correspond to individuals, the communities correspond to genders, and edges correspond to whether the online dating service successfully matched pairs of individuals.
+In this model, we assume that for each pair of genders, there is a constant probability of a match. 
+This does not take into account that some individuals are more likely to form connections than others.
+
+\end{example}
+
+\begin{definition}[Popularity adjusted block model]
+\label{def:pabm}
+Let $G = (V, E)$ be an undirected Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $K \geq 1$ be an integer that describes the number of communities, and let 
+$z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
+Suppose that each vertex $v_i$ has $K$ popularity parameters $\lambda_{i1}, ..., \lambda_{iK}$ for which each $\lambda_{ik}$ describes $v_i$'s affinity toward community $k$. 
+Then if the edge probability matrix $P$ for this graph is such that $P_{ij} = \lambda_{i, z_j} \lambda_{j, z_i}$ and $A \sim \BG(P)$, then $G$ is a popularity adjusted block model.
+\end{definition}
+
+\begin{example}[Dating network as a PABM]
+\label{ex:pabm-dating}
+\end{example}
 
@@
 
-### Review of Traditional Statistical Inference Approaches to Block Model Estimation
+\newpage
+
+# Random Dot Product Graphs and Generalized Random Dot Product Graphs {#sec:grdpg}
+
+## Definitions
+
+\begin{definition}[Generalized random dot product graph]
+\label{def:grdpg}
+Let $p \geq 1$ and $q \geq 0$ be integers. 
+Define $I_{p,q}$ as the block diagonal matrix $I_{p,q} = \Bigl[\begin{smallmatrix} I_{p} & 0 \\ 0 & - I_{q} \end{smallmatrix} \Bigr]$ where $I_p$ and $I_q$ are the identity matrices of dimensions $p \times p$ and $q \times q$ respectively.
+Denote $d = p + q$ and let $\mathcal{X}$ be the subset of $\mathbb{R}^d$ such that, for any $x, y \in \mathcal{X}$, we have $x^\top I_{p,q} y \in [0, 1]$. 
+Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
+A graph $G$ with adjacency matrix $A$ is said to be a generalized random dot product graph with latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$ if $A \sim \BG(P)$ where the edge probability matrix $P$ is given by $P = \rho_n X I_{p,q} X^\top$, i.e., the entries of $P$ are of the form $P_{ij} = \rho_n x_i^\top I_{p,q} x_j$.
+
+We use the notation $A \sim \GRDPG_{p,q}(X; \rho_n)$ to denote a random adjacency matrix $A$ drawn from latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$.
+If the $n$ vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F, n; \rho_n)$.
+
+If $q = 0$ and so $d = p$, (i.e., $P$ is positive semidefinite), then we call this a random dot product graph. 
+In this case, we use the notation $A \sim \RDPG(X; \rho_n)$ and $A \sim \RDPG(F, n; \rho_n)$. 
+\end{definition}
+
+\begin{definition}[Adjacency spectral embedding]
+\label{def:ase}
+\end{definition}
+
+\begin{definition}[Indefinite orthogonal group]
+\label{def:indef-ortho-group}
+\end{definition}
+
+\begin{remark}
+\label{rem:non_identifiable}
+The latent vectors that produce $X I_{p,q} X^\top = P$ are not unique
+\citep{rubindelanchy2017statistical}.
+More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, then for any $Q \in \mathbb{O}(p, q)$ we also have 
+$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j = x_i^\top I_{p, q} x_j = P_{ij}$. 
+Unlike in the RDPG case, transforming the latent positions via multiplication by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or distances.
+\end{remark}
+
+## Connecting the Stochastic Block Model and the Degree Corrected Block Model to the Generalized Random Dot Product Graph {#sec:sbm-dcbm-grdpg}
 
@@
 
-# Random Dot Product Graphs and Generalized Random Dot Product Graphs
+# Popularity Adjusted Block Models are Generalized Random Dot Product Graphs {#sec:pabm-grdpg}
 
-## Definitions
+In this section, we connect the PABM to the GRDPG and exploit that connection to develop algorithms for community detection and parameter estimation. 
+In order to make the explicit connection between the PABM and the GRDPG, we make use of an alternative but equivalent definition of the PABM which parameterizes the model in terms of popularity vectors, which are collections of popularity parameters.
 
-## Connecting the Stochastic Block Model to the Generalized Random Dot Product Graph
+\begin{remark}
+\label{rem:pabm_view2}
+In a PABM, each vertex $i$ has $K$ popularity parameters $\lambda_{i1}, \dots, \lambda_{iK}$, that describe its affinity toward each of the $K$ communities. 
+Another view of a PABM is as follows.
+Let $\tilde{P}$ be the matrix obtained by permuting the rows and columns of $P$ so that the vertices are reorganized by community memberships $z_i \in \{1,2,\dots,K\}$ in increasing order. 
+Denote by $\tilde{P}^{(k \ell)}$ the $n_k \times n_{\ell}$ submatrix of $\tilde{P}$ corresponding to the edge probabilities between vertices in communities $k$ and $\ell$. 
+Here $n_k = |\{ i \colon z_i = k\}|$ is the number of vertices assigned to community $k$, for $k = 1,2\dots,K$.
+Note that $\tilde{P}^{(k \ell)} = (\tilde{P}^{(\ell k)})^\top$. 
+Next let $\lambda^{(k \ell)} = \{\lambda_{i \ell} \colon z_i = k\} \in \mathbb{R}^{n_k}$; the elements of $\lambda^{(k \ell)}$ are the affinity parameters toward the $\ell$th community of all vertices in the $k^{th}$ community.
+Define $\lambda^{(\ell k)}$ analogously. 
+Then each block $\tilde{P}^{(k \ell)}$ can be written as the outer product of two vectors:
+\begin{equation} \label{eq:pabm}
+  \tilde{P}^{(k \ell)} = \rho_n \lambda^{(k \ell)} (\lambda^{(\ell k)})^{\top}.
+\end{equation}
 
-## Connecting the Degree Corrected Block Model to the Generalized Random Dot Product Graph
+We will henceforth use the notation \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K, \rho_n)\) to denote a random adjacency matrix \(A\) drawn from a PABM with $K$ communities, popularity parameters \(\{\lambda^{(k \ell)}\}\) and sparsity parameter $\rho_n$.
+\end{remark}
 
-\newpage
+## The Geometry of PABMs
 
-# Popularity Adjusted Block Models are Generalized Random Dot Product Graphs
+In section \@ref(sec:sbm-dcbm-grdpg), we showed how the SBM and DCBM are special cases of the GRDPG in which the latent vectors form a very particular geometry. 
+Similarly, we now  show the special geometry of the PABM when viewed as a GRDPG. 
+For ease of exposition, and without loss of generality, we drop the dependency on the sparsity parameter $\rho_n$ and assume $\rho_n \equiv 1$ throughout this subsection.
+
+\begin{theorem}[The latent configuration of the PABM]
+\label{thm:pabm-grdpg}
+Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_K)$ be an instance of a
+PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
+\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. 
+Then there exists a block diagonal matrix
+$X \in \mathbb{R}^{n \times K^2}$ defined by $\{\lambda^{(k \ell)}\}$ and a 
+$K^2 \times K^2$ {\em fixed} orthonormal matrix $U$ such 
+that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) /
+  2}(\tilde{\Pi}XU)$. Here $\tilde{\Pi}$ is the permutation matrix
+such that $P = \tilde{\Pi} \tilde{P} \tilde{\Pi}^{\top}$ where the
+rows and columns of $\tilde{P}$ are arranged according to increasing values of the
+community labels (see remark~\ref{rem:pabm_view2}). 
+\end{theorem}
+
+\begin{proof}
+We will prove this theorem in two parts. First, for demonstration
+purposes, we focus on the case when $K = 2$ to build intuition. 
+The general case of $K \geq 2$ is presented later.
+
+For the $K = 2$ case, the proof is straightforward. We will first work with
+the matrix $\tilde{P}$. Note that $\tilde{P}$ has the form
+
+$$\tilde{P} = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} &
+  P^{(22)} \end{bmatrix} = \begin{bmatrix} \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
+  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)}
+  (\lambda^{(22)})^\top \end{bmatrix}.$$
+  Now let
+$$X = \begin{bmatrix}
+\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
+0 & 0 & \lambda^{(21)} & \lambda^{(22)}
+\end{bmatrix} \quad \text{and} \quad
+U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
+0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
+0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
+0 & 1 & 0 & 0 \end{bmatrix}.$$
+Then by straightforward matrix multiplication, we obtain 
+$$X U I_{3, 1} U^\top X^\top =
+\begin{bmatrix}
+  \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
+  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
+\end{bmatrix} = \tilde{P}$$
+and hence $\tilde{P}$ also corresponds to the edge probability matrix of GRDPG
+with latent vectors described by $X U$. As $P = \tilde{\Pi} \tilde{P}
+\tilde{\Pi}^{\top}$ we conclude that $P$ has latent vectors described
+by $\tilde{\Pi} X U$.
+
+It is nevertheless instructive to look at a few intermediate steps. 
+More specifically, the product $U I_{3, 1} U^\top$ 
+yields a permutation matrix $\Pi$ with fixed points at positions $1$ and $4$ 
+and a cycle of order 2 swapping positions $2$ and $3$, i.e., 
+$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
+  0 & 0 & 1 & 0 \\
+  0 & 1 & 0 & 0 \\
+  0 & 0 & 0 & 1
+\end{bmatrix}.$$
+Furthermore, as $U$ is orthonormal and $I_{3, 1}$ is diagonal, 
+$U I_{3, 1} U^\top$ is also an eigendecomposition of $\Pi$ where the fixed
+points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
+while the cycles of order two are mapped to the eigenvectors  
+$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
+e_3)$; here $e_i$ denote the $i^\mathrm{th}$ basis vector in $\mathbb{R}^{4}$. 
+Thus, another way of decomposing the edge probability matrix is
+$\tilde{P} = X \Pi X^\top$ where the rows of $X$ lie in the union of
+two 2-dimensional orthogonal subspaces and $\Pi$ is a permutation matrix.
+
+For the general case, we can extend $\tilde{P} = X \Pi X^\top$ to larger $K$. 
+For a more concrete example of this, refer to Example~1. 
+We once again consider $\tilde{P}$ as defined in
+remark~\ref{rem:pabm_view2}.  
+We first define the following matrices
+
+\begin{gather}
+\label{eq:xy}
+\Lambda^{(k)} = \begin{bmatrix} \lambda^{(k1)} \mid \cdots \mid \lambda^{(kK)} \end{bmatrix}
+\in \mathbb{R}^{n_k \times K}, \quad
+X = \mathrm{blockdiag}(\Lambda^{(1)}, \dots, \Lambda^{(K)}) \in
+\mathbb{R}^{n \times K^2}, \\
+L^{(k)} = \mathrm{blockdiag}(\lambda^{(1k)}, \dots, \lambda^{(Kk)}) \in
+\mathbb{R}^{n \times K}, \quad
+Y = \begin{bmatrix} L^{(1)} \mid \cdots \mid L^{(K)} \end{bmatrix} \in
+\mathbb{R}^{n \times K^2}.
+\end{gather}
+
+It is then straightforward to verify that
+
+\begin{gather*}
+  XY^{\top} = \mathrm{blockdiag}(\Lambda^{(1)}, \dots,
+\Lambda^{(K)}) \begin{bmatrix} L_1^{\top} \\ \vdots \\
+  L_{K}^{\top} \end{bmatrix} = \begin{bmatrix} \Lambda^{(1)}
+  (L^{(1)})^{\top} \\ \vdots \\
+  \Lambda^{(K)} (L^{(K)})^{\top} \end{bmatrix}, \\
+\Lambda^{(k)} (L^{(k)})^{\top} = \begin{bmatrix} \lambda^{(k1)}
+  (\lambda^{(1k)})^{\top} \mid \dots \mid \lambda^{(kK)}
+  (\lambda^{(Kk)})^{\top} \end{bmatrix} = \begin{bmatrix} P^{(k1)}
+  \mid P^{(k2)} \mid \dots \mid P^{(kK)} \end{bmatrix}.
+\end{gather*}
+We therefore have $\tilde{P} = X Y^\top$. 
+Similar to the $K = 2$ case, we also have $Y = X \Pi$ for some permutation matrix
+$\Pi$ and hence $\tilde{P} = X \Pi X^\top$.
+The permutation described by $\Pi$ now has $K$ fixed points, which correspond to
+$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where
+$k = r (K + 1) + 1$ for $0 \leq r \leq K - 1$. It also has
+$\tbinom{K}{2}$ cycles of order $2$. Each cycle corresponds to
+a pair of eigenvalues $\{-1,+1\}$ and a pair of eigenvectors
+$\{(e_s + e_t)/\sqrt{2},(e_s - e_t)/ \sqrt{2}\}$.
+
+Let $p = K (K + 1) / 2$ and $q = K (K - 1) / 2$. 
+We therefore have
+\begin{equation} \label{eq:permutation}
+\Pi = U I_{p,q} U^\top
+\end{equation}
+where $U$ is a $K^2 \times K^2$ orthogonal matrix and hence
+\begin{equation} \label{eq:pabm-grdpg}
+\tilde{P} = X U I_{p, q} (X U)^\top.
+\end{equation}
+In summary we can describe the PABM with $K$ communities as a GRDPG with latent
+positions $\tilde{\Pi} X U$ and known signature $(p,q) = \bigl( \tfrac{1}{2} K (K + 1) ,
+\tfrac{1}{2} K (K - 1)\bigr)$.
+\end{proof}
+
+\begin{example} Let $A$ be a $3$ blocks PABM with latent vectors $\{\lambda^{(k \ell)} \colon 1 \leq k \leq 3, 1 \leq \ell \leq 3\}$. 
+Using the same notation as in theorem~\ref{thm:pabm-grdpg}, we can define
+\begin{gather*}
+X = \begin{bmatrix}
+\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
+0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
+0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
+\end{bmatrix}, \\
+Y = \begin{bmatrix}
+\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
+0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
+0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
+\end{bmatrix}.
+\end{gather*}
+Then $Y = X \Pi$ and $\tilde{P} = X Y^{\top}$ where $\Pi$ is a $9 \times 9$ 
+permutation matrix of the form
+$$\Pi = 
+\Bigl[e_1 \mid e_4 \mid e_7 \mid e_2 \mid e_5 \mid e_8 \mid e_3
+\mid e_6 \mid e_9 \Bigr].$$
+where $e_i$ denotes the $i^{th}$ basis vector in $\mathbb{R}^{9}$. 
+The matrix $\Pi$ corresponds to a permutation of $\{1,2,\dots,9\}$ with the following decomposition.
+\begin{enumerate}
+\item Positions 1, 5, 9 are fixed.
+\item There are three cycles of length 2, namely $(2, 4)$, $(3, 7)$, and $(6, 8)$.
+\end{enumerate}
+We can thus write $\Pi$ as $\Pi = U I_{6, 3} U^\top$ where the first three columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the fixed points, the next three columns are the eigenvectors $(e_k + e_{\ell}) / \sqrt{2}$, and the last three columns are the eigenvectors $(e_k - e_{\ell}) / \sqrt{2}$ for $(k, \ell) \in \{(2,4),(3,7),(6,8)\}$.
+
+The matrix $\tilde{P}$ is then the edge probabilities matrix for a 
+Generalized Random Dot Product Graph whose latent positions 
+are the rows of the matrix
+$$XU = \begin{bmatrix}
+  \lambda^{(11)} & 0 & 0 &
+  \frac{\lambda^{(12)}}{\sqrt{2}} & \frac{\lambda^{(13)}}{\sqrt{2}} & 0 &
+  \frac{\lambda^{(12)}}{\sqrt{2}} & \frac{\lambda^{(13)}}{\sqrt{2}} & 0 \\
+  0 & \lambda^{(22)} & 0 &
+  \frac{\lambda^{(21)}}{\sqrt{2}} & 0 & \frac{\lambda^{(23)}}{\sqrt{2}} &
+  -\frac{\lambda^{(21)}}{\sqrt{2}} & 0 & \frac{\lambda^{(23)}}{\sqrt{2}} \\
+  0 & 0 & \lambda^{(33)} &
+  0 & \frac{\lambda^{(31)}}{\sqrt{2}} & \frac{\lambda^{(32)}}{\sqrt{2}} &
+  0 & -\frac{\lambda^{(31)}}{\sqrt{2}} & -\frac{\lambda^{(32)}}{\sqrt{2}}
+\end{bmatrix}$$
+and the latent positions for $P$ is a permutation of the rows of
+$XU$. 
+\end{example}
+
+## Algorithms
+
+Two inference objectives arise from the PABM:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Community membership identification (up to permutation).
+\item
+  Parameter estimation (estimating $\lambda_{ik}$'s).
+\end{enumerate}
+
+In our methods, the data that are observed for estimation is the adjacency matrix, $A \sim \PABM(\{\lambda^{(k \ell)}\}_K, \rho_n)$, along with an assumed number of communities, $K$. 
+To motivate our methods, we first consider community detection and parameter estimation in the case where we know the edge probability matrix $P$ beforehand, noting that community memberships and popularity parameters are not immediately discernible from $P$ itself.
+After establishing methods for community detection and parameter estimation from $P$, we use the consistency property of the ASE \citep{doi:10.1080/01621459.2012.699795,rubindelanchy2017statistical} to demonstrate that the same methods work for $A$ almost surely as $n \to \infty$.
+
+### Previous Work
+
+\citet{307cbeb9b1be48299388437423d94bf1} 
+used Modularity Maximization (MM) and the Extreme Points (EP)
+algorithm \citep{le2016} for community detection and parameter
+estimation. They were able to show that as the sample size increases,
+the *proportion* of misclassified community labels (up to permutation)
+goes to 0.
+
+\cite{noroozi2019estimation} used Sparse Subspace Clustering (SSC) 
+\citep{5206547} for community detection in the PABM. 
+The SSC algorithm can be described as follows: 
+Given \(X \in \mathbb{R}^{n \times d}\) with vectors
+\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
+\(c_i = \argmin_{c} \|c\|_1\) subject to \(x_i = X^\top c\) and
+\(c^{(i)} = 0\), where $c^{(i)}$ is the $i^{th}$ entry of $c$, is solved for each \(i \in [n]\). 
+The solutions are collected into matrix
+\(C = \bigl[ c_1 \mid \cdots \mid c_n \bigr]^\top\) to
+construct an affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lies
+exactly on one of \(K\) subspaces, \(B\) describes an undirected graph
+consisting of *at least* \(K\) disjoint subgraphs, i.e., \(B_{ij} = 0\) if \(x_i, x_j\) lie on different subspaces. 
+The intuition here is that vectors that lie on the same subspace can be described as linear combinations of each other, assuming the number of vectors in the subspace is greater than the dimensionality of the subspace.
+Then once sparsity is enforced, for each $c_i$, its $j^{th}$ element $c_i^{(j)}$ is zero if $x_j$ belongs to a subspace that doesn't contain $x_i$, resulting in $B_{ij} = 0$. 
+If \(X\) instead represents points near \(K\) subspaces with some noise, 
+then this property may only hold approximately and a final graph partitioning step may be required 
+(e.g., edge thresholding or spectral clustering).
+
+In practice, due to presence of noise, SSC is often done by solving the LASSO problems
+\begin{equation} 
+\label{eq:ssc}
+c_i = \argmin_c \frac{1}{2} \|x_i - X_{-i}^\top c\|^2_2 + \vartheta \|c\|_1
+\end{equation}
+for some sparsity parameter \(\vartheta > 0\). 
+The \(c_i\) vectors are then collected into \(C\) and \(B\) as before.
+
+\begin{definition}[Subspace Detection Property]
+\label{def:subspace_detection}
+Let $X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top$ be noisy points sampled from $K$ subspaces, i.e., $x_i = y_i + z_i$ where the $y_i$ belongs to the union of $K$ subspaces and the $z_i$ are noise vectors.
+Let $\vartheta \geq 0$ be given and let $C$ and $B$ be constructed from the solutions of LASSO problems as described in equation~\ref{eq:ssc} with this given choice of $\vartheta$. 
+Then $X$ is said to satisfy the subspace detection property with sparsity parameter $\vartheta$ if each column of $C$ has nonzero $\ell_2$ norm and $B_{ij} = 0$ whenever $y_i$ and $y_j$ are from different subspaces.
+\end{definition}
+
+\begin{remark}
+\label{rem:sdp}
+One common approach to show that SSC works for a noisy sample $X$ is to show that $X$ satisfies the subspace detection property for some choice of $\vartheta$; 
+recall that $\vartheta$ is the sparsity parameter for
+the LASSO problems in equation~\ref{eq:ssc}. However, this is not sufficient to
+guarantee that SSC perfectly recovers the underlying subspaces.
+More specifically, if $X$ satisfies the 
+subspace detection property, 
+then $B$ describes a graph with {\em at least} $K$ disconnected subgraphs, 
+with the ideal case being that there are exactly $K$ subgraphs 
+which map onto each subspace. 
+Nevertheless it is also possible that the $K$ subspaces are represented by
+$K' > K$
+multiple disconnected subgraphs and we cannot, at least without a subsequent
+post-processing step, recover the $K$ subspaces directly from $B$;
+see \citet{sdp_sufficiency} and \citet{liu_ssc}
+for further discussions. Therefore in practice $B$ is usually treated as an
+affinity matrix and, as we allude to earlier, 
+the rows of $B$ are partitioned using some
+clustering algorithm to obtain the final clustering. 
+\end{remark}
+
+Theorem \ref{thm:pabm-grdpg} suggests that SSC is appropriate for community
+detection for the PABM, provided that we observe the edge probabilities
+matrix $P$. More precisely, given the matrix $\tilde{P}$ obtained by
+permuting the rows and columns of $P$ as described in
+remark \ref{rem:pabm_view2} we can recover $XU$ up
+to some non-identifiability indefinite orthogonal transformation $Q$. 
+Then using results from \citet{soltanolkotabi2012}, it can be easily shown
+that the subspace detection property holds for \(XU\). Indeed, the columns
+of \(XU\) from different communities correspond to mutually
+orthogonal subspaces. This then implies that the
+subspace detection property also holds for $XUQ$ for all invertible
+transformation $Q$ and hence the subspace detection property also holds for
+$\tilde{\Pi} X U Q$ for any $n \times n$ permutation matrix $\tilde{\Pi}$.
+
+However, because we do not observe $P$ but rather only the noisy
+adjacency matrix $A \sim \BG(P)$, the natural approach then is
+to perform SSC on the rows of the spectral embedding of $A$, 
+since the embedding of $P$ consists of $K$ subspaces (theore \ref{thm:pabm-grdpg}), 
+and so the embedding of $A$ will lie approximately on the $K$ subspaces. 
+We will show in theorem \ref{theorem5} that, with probability converging to
+one as $n \rightarrow \infty$, the rows of the ASE of $A$ also
+satisfy the subspace detection property. 
+Theorem \ref{theorem5} builds upon existing work by
+\citet{rubindelanchy2017statistical} who describe the convergence
+behavior of the ASE of \(A\) to that of \(\tilde{\Pi} XU\), and
+\citet{jmlr-v28-wang13} who show the necessary conditions for the
+subspace detection property to hold in noisy cases where the points lie
+near subspaces. Finally we emphasize that while
+\citet{noroozi2019estimation} also considered the use of SSC for
+community recovery in PABM, they instead applied SSC to the rows of
+\(A\) itself, foregoing
+the embedding step altogether. It is however much harder to show that
+the rows of $A$ satisfy the subspace detection property and thus, to
+the best of our knowledge, there is currently no consistency result
+regarding the application of SSC to the rows of $A$.
+
+We previously stated in theorem \ref{thm:pabm-grdpg} one possible set of latent positions that result in
+the edge probability matrix of a PABM, namely
+\(P = \tilde{\Pi} (XU) I_{p, q}
+(XU)^\top \tilde{\Pi}^{\top}\) where $X$ is block diagonal and
+$\tilde{\Pi}$ is a permutation matrix.  
+Furthermore, the explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
+such that points within each community lie on \(K\)-dimensional
+orthogonal subspaces, i.e. $\langle U^{\top} x_i, U^{\top} x_j \rangle = 0$ whenever $i$ and $j$ are in different communities. 
+Thus if we have (or can estimate) \(XU\) directly, then both the community
+detection and parameter identification problem are trivial because \(U\)
+is orthonormal and fixed for each value of \(K\). 
+However, direct
+identification or estimation of \(XU\) is possibly difficult 
+due to the non-identifiability of $XU$ (see remark \ref{rem:non_identifiable})
+when we are given only $P$. 
+More specifically, suppose we find a matrix $Y \in
+\mathbb{R}^{n \times K^2}$
+such that \(P = Y I_{p, q} Y^\top\). Then it is generally the case that
+\( Y = \tilde{\Pi} XU Q \) for some indefinite orthogonal matrix
+$Q \in \mathbb{O}(p,q)$. 
+However since \(Q\) is not necessarily an
+orthogonal matrix and hence, if $y_i$ denote the $i^{th}$ row of $Y$, 
+then $\langle U^{\top} x_i, U^{\top} x_j \rangle
+\neq \langle y_i, y_j \rangle$.
+This prevents us from transferring the orthogonality property of
+\(XU\) directly to $Y$.
+
+Nevertheless by using the special geometric structure of $X$ we can circumvent the
+non-identifiability of $Y$ and $XU$ by using instead the rows of the
+matrix $V$ of eigenvectors (corresponding to the non-zero eigenvalues) of $P$. In particular $V$ is identifiable
+up to orthogonal transformations and furthermore, due to the block
+diagonal structure of $X$, the rows of $V$ also lie on $K$ distinct orthogonal
+subspaces and hence $v_i^{\top} v_j = 0$ whenever $z_i \not = z_j$.
+
+\begin{theorem}
+\label{thm:osc-p}
+Let $P = V D V^\top$ be the spectral decomposition 
+of the edge probability matrix. 
+Let $B = n V V^\top$. 
+Assume $\lambda_{i z_i} > 0$ for each $i \in [n]$, 
+i.e., each vertex's popularity parameter to its own community is nonzero. 
+Then $B_{ij} = 0$ if and only if 
+vertices $i$ and $j$ are in different communities.
+\end{theorem}
+
+\begin{proof}
+We first show that $V V^\top =
+\tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top} $ where $X$ is 
+defined as in Eq.~\eqref{eq:xy}. Indeed, by Theorem 2, 
+\(P = \tilde{\Pi} X U I_{p, q} U^\top X^\top \tilde{\Pi}\) for $p = K(K+1)/2$ and $q = K(K-1)/2$. 
+The eigendecomposition \(P = V D V^\top\) also yields $P = V
+|D|^{1/2} I_{p, q} |D|^{1/2} V^\top$ where \(|\cdot|^{1/2}\) is
+applied entry-wise. Now let $Y = \tilde{\Pi} XU$ and $\tilde{Y} = V|D|^{1/2}$; note that
+$Y$ and $\tilde{Y}$ both have full column ranks. 
+Because $P = Y I_{p,q} Y^{\top} = \tilde{Y} I_{p,q} \tilde{Y}^{\top}$, we have
+$$Y = \tilde{Y} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}.$$
+Let $Q = I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}$ and note that
+$Y = \tilde{Y} Q$. We then have
+\begin{equation*}
+  \begin{split}
+  Q^{\top} I_{p,q} Q &= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} \tilde{Y} I_{p,q}
+I_{p,q} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} \\ 
+&= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} Y I_{p,q}
+Y^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} =  I_{p,q}
+\end{split}
+\end{equation*}
+and hence $Q$ is an indefinite orthogonal matrix.
+
+Let $R = U Q |D|^{-1/2}$ and note that $V = \tilde{\Pi} XR$. Because $R$ is
+invertible, we can write
+$$\tilde{\Pi} X (X^{\top} X)^{-1} X^{\top} \tilde{\Pi}^{\top} =
+\tilde{\Pi} X R (R^{\top} X^{\top} X R)^{-1}
+R^{\top} X^{\top} \tilde{\Pi}^{\top}.$$ 
+Furthermore, as $V$ has orthonormal columns, $R^{\top} X^{\top} X R =
+V^{\top} \tilde{\Pi} \tilde{\Pi}^{\top} V = V^{\top} V = I$. We thus conclude
+$$\tilde{\Pi} X (X^{\top} X)^{-1} X^{\top} \tilde{\Pi}^{\top} = V (V^{\top} V)^{-1} V^{\top} = V V^{\top}$$
+as desired.
+
+To complete the proof of theorem \ref{thm:osc-p}, recall that \(X\) 
+is block diagonal with each block corresponding to one community, 
+and hence \(X (X^\top X)^{-1} X^\top\) is also a
+block diagonal matrix with each block corresponding to a community. 
+As $B = n VV^{\top} = n \tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top}$, 
+we conclude that $B_{ij} = 0$ 
+whenever vertices $i$ and $j$ belong to different communities.  
+\end{proof}
+
+Theorem \ref{thm:osc-p} provides perfect community detection from \(P\).
+More specifically, let \(|B|\) be the affinity matrix for graph \(G'\), 
+where $|\cdot|$ is applied entry-wise. Then
+\(G'\) consists of exactly \(K\) disjoint subgraphs, 
+as $G'$ has no edges between communities. 
+All that is left to identify the communities is 
+to assign each subgraph a distinct community label. 
+In practice, we do not observe $P$ and instead only observe the noisy
+$A \sim \BG(P)$. A natural approach is then to use
+the affinity matrix $\hat{B} = n \hat{V} \hat{V}^{\top}$ where
+$\hat{V}$ is the matrix of eigenvectors (corresponding to the largest
+eigenvalues in modulus) of $A$. The resulting procedure, named
+Orthogonal Spectral Clustering, is presented in
+algorithm \ref{alg:osc}.
+The following result leverages existing theoretical properties
+of ASE for estimating of latent positions in a GRDPG \citep{rubindelanchy2017statistical} to show that
+$\hat{B}$ converges almost surely to $B$; in particular 
+ $\hat{B}_{ij} \stackrel{a.s.}{\to} 0$ 
+for each pair $(i, j)$ in different communities.
+
+\begin{algorithm}[tp]
+  \label{alg:osc}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Community assignments $1, ..., K$}
+    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most
+    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct
+    $V$ using these eigenvectors as its columns.\;
+    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
+    Construct graph $G$ using $B$ as its similarity matrix.\;
+    Partition $G$ into $K$ disconnected subgraphs
+    (e.g., using edge thresholding or spectral clustering).\;
+    Map each partition to the community labels $1, ..., K$.\;
+  \caption{Orthogonal Spectral Clustering.}
+\end{algorithm}
+
+\begin{theorem}
+\label{thm:osc}
+Assume the setting of theorem \ref{thm:osc-p}. 
+Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix
+obtained from OSC as described in algorithm~\ref{alg:osc}. 
+Then for $n \rho_n = \omega(\log^{4}{n})$, we have
+\begin{equation} 
+\label{eq:osc}
+\max_{i, j} |\hat{B}_{ij} - B_{ij}| = O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
+\end{equation}
+with high probability. In particular 
+$\hat{B}_{ij} -B_{ij} \overset{\mathrm{a.s.}}{\rightarrow} 0$ 
+where the convergence is
+uniform over all $i,j$. Hence for all pairs $(i,j)$ in different
+communities we have 
+$\hat{B}_{ij} \overset{\mathrm{a.s.}}{\rightarrow} 0$, 
+while for all pairs $(i, j)$ in the same community, 
+$\liminf_{n \to \infty} |\hat{B}_{ij}| > 0$ almost surely. 
+\end{theorem}
 
@@
 
-\section*{Appendix A}
+\section*{Appendix A: Proofs of Theorems}
+
+Let \(V_n\) and \(\hat{V}_n\)
+be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
+$K^2$ largest eigenvalues (in modulus), respectively. 
+We first state an important technical lemma for bounding the maximum
+$\ell_2$ norm difference between the rows of $\hat{V}_n$ and
+$V_n$. See \citet{cape_biometrika} and 
+\citet[Lemma~5]{rubindelanchy2017statistical} for a proof.
+
+\begin{lemma}
+\label{lem:technical}
+Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell \ell)}\}_{K})$ be a $K$-blocks
+PABM graph on $n$ vertices and let \(V\) and \(\hat{V}\)
+be the $n \times K^2$ matrices whose columns are the eigenvectors of 
+\(P\) and \(A\) corresponding to the
+$K^2$ largest eigenvalues in modulus, respectively.
+Let \(v_i^\top\) and \(\hat{v}_i^\top\) denote the $i$th 
+row of \(V\) and \(\hat{V}\), respectively. 
+Then there exists a constant $c > 1$ and an orthogonal matrix $W$ such
+that with high probability,
+$$\max_{i} \|W \hat{v}_i - v_i\|  = O\Big(\frac{\log^{c}n}{n \sqrt{\rho_n}} \Big).$$
+In particular we can take $c = 1 + \epsilon$ for any $\epsilon > 0$. 
+\end{lemma}
+
+\begin{proof}[Proof of theorem~\ref{thm:osc}]
+Recall the notations in lemma~\ref{lem:technical} and note that,
+under our assumption that the latent vectors $\lambda^{(k \ell)}$
+are all homogeneous, we have $\max_{i} \|v_i\| =
+O(n^{-1/2})$.
+
+Next recall theorem~\ref{thm:osc-p}; in particular $B_{ij} = nv_i^{\top}
+v_j$. 
+We therefore have
+\[\begin{split}
+ \max_{ij} |\hat{B}_{ij} - B_{ij}| &= \max_{ij} n |\hat{v}_i^\top \hat{v}_j -
+v_i^\top v_j| \\
+& \leq n \max_{ij} |\hat{v}_i^\top W W^\top \hat{v}_j -
+v_i^\top v_j| \\
+& \leq n \max_{i,j} \Bigl(\|W^{\top} \hat{v}_i - v_i\| \times \|\hat{v}_j\|
++ \|W^{\top} \hat{v}_j - v_j\| \times \|v_i\|\Bigr) \\
+& \leq n \Bigl(\max_{ij}  \|W \hat{v}_i  - v_i \|^2 +  \|W
+\hat{v}_i  - v_i \| \times \|v_j\| +  \|W \hat{v}_j  - v_j \| \times \|v_i\|\Bigr) 
+\\ &
+\leq n \max_{i} \|W \hat{v}_i  - v_i \|^2 + 2n \max_{i}
+\|W \hat{v}_i  - v_i \| \times \max_{j} \|v_j\|
+\\
+& = O \Big( \frac{\log^{c}{n}}{n^{1/2} \rho_n^{1/2}} \Big)
+\end{split}\]
+with high probability.
+Theorem~\ref{thm:osc} follows from the above bound together with the
+conclusion in theorem~\ref{thm:osc-p} that $B_{ij} = 0$ whenever vertices $i$ and $j$
+belongs to different communities. 
+\end{proof}
 

### 2022-08-22 draft/draft.tex
still writing

@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  colorlinks=true,
-  linkcolor={Maroon},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 \usepackage[linesnumbered,ruled,vlined]{algorithm2e}
-\setcitestyle{numbers,square,comma}
+\setcitestyle{year,comma}
 \usepackage{verbatim}
@@
 \usepackage[]{natbib}
-\bibliographystyle{plainnat}
+\bibliographystyle{apalike}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  colorlinks=true,
+  linkcolor={Maroon},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={blue},
+  pdfcreator={LaTeX via pandoc}}
 
@@
 
+\usepackage{amsthm}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}{Lemma}[section]
+\newtheorem{corollary}{Corollary}[section]
+\newtheorem{proposition}{Proposition}[section]
+\newtheorem{conjecture}{Conjecture}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
+\newtheorem{exercise}{Exercise}[section]
+\theoremstyle{definition}
+\newtheorem{hypothesis}{Hypothesis}[section]
+\theoremstyle{remark}
+\newtheorem*{remark}{Remark}
+\newtheorem*{solution}{Solution}
 \begin{document}
@@
 
+{
+\hypersetup{linkcolor=}
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
 \setlength{\parindent}{10.0pt}
@@
 \newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
+\newcommand{\SBM}{\mathrm{SBM}}
+\newcommand{\DCBM}{\mathrm{DCBM}}
+\newcommand{\argmin}{\mathrm{argmin}}
 
@@
 \hypertarget{graphs-and-representations-of-network-data}{%
-\subsection{Graphs and Representations of Network
-Data}\label{graphs-and-representations-of-network-data}}
+\subsection{Graphs and Representations of Network Data}\label{graphs-and-representations-of-network-data}}
 
-Graph and network data have become increasingly widespread in various
-fields including sociology, neuroscience, biostatistics, and computer
-science. This has resulted in challenges for researchers who rely on
-traditional statistical and machine learning methods that are
-incompatible with graph data and instead assume that the data exist as
-feature vectors. To illustrate this, consider the typical approach to
-building a statistical or machine learning model. Data are often
-represented as an \(n \times p\) matrix
-\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which
-each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features
-and each column is a set of \(n\) feature measurements. An analysis task
-for these data might be to come up with a classification model
-\(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical
-values of each feature of a vector. For instance, \(\phi(x)\) might
-first compute the distance from \(x\) to one of \(K\) points in
-\(\mathbb{R}^p\) and assign \(x\) to the label of the nearest point.
-Examples of this include linear discriminant analysis (in the case of
-supervised learning) and Lloyd's algorithm or Gaussian mixture models
-(in the case of unsupervised learning). However, this type of approach
-is incompatible with data that are represented as graphs, in which each
-observation is not a vector of numerical features but a set of
-relationships to other observations. Instead of feature vector
-\(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top\),
-we observe
-\(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}\) where
-each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
+Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science.
+This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors.
+To illustrate this, consider the typical approach to building a statistical or machine learning model.
+Data are often represented as an \(n \times p\) matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
+An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
+For instance, \(\phi(x)\) might first compute the distances from \(x\) to \(K\) points in \(\mathbb{R}^p\) and then assign \(x\) to the label of the nearest point.
+Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning).
+However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
+Instead of feature vector \(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top \in \mathbb{R}^p\), we observe \(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
+For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
 
-A more formal description of graph data is as follows: Suppose we
-observe a network of \(n\) objects and pairwise relationships between
-them. This network is represented by a graph object \(G = (V, E)\) with
-vertex set \(V\), representing the objects, and edge set \(E\),
-representing the pairwise relationships. The numeric representation of
-these data is in the form of \emph{affinity matrix}
-\(A \in \mathbb{R}^{n \times n}\) in which each \(A_{ij}\) is object
-\(i\)'s relationship to object \(j\). We assume that the entries of
-\(A\) represent affinities or similarities, i.e., the higher the value
-of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\). If
-\(A_{ij} = 0\), then \(i\) has no direct relationship to \(j\). \(A\) is
-symmetric if it represents an undirected graph in which the relationship
-from \(i\) to \(j\) is the same as the relationship from \(j\) to \(i\).
-\(A\) is binary, i.e., \(A \in \{0, 1\}^{n \times n}\), if it represents
-an unweighted graph in which edges either exist or don't exist. If \(A\)
-is binary, we call it an \emph{adjacency matrix}.
+\begin{example}
+$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
+\begin{equation}
+\label{eq:kmeans}
+W(x_1, ..., x_n) = \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
+\end{equation}
+in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments we wish to optimize for, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
+A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration labels are updated according to which centroid they are closest to and centroids are updated via the sample mean of each cluster.
+
+If instead of vectors in $\mathbb{R}^p$ we have a graph, it is not obvious how to translate $K$-means clustering to these data. 
+In particular, there is no inherent notion of centroid or sample mean for graph data. 
+However, there are inherent notions of distances among vectors, such as shortest path distance, expected commute time, or resistance distance. 
+One way to adapt $K$-means to graph data is to use an equivalent objective function as Eq.~\ref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
+$$
+\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} d(x_i, x_j)^2.
+$$
+Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
+If the data live in a Euclidean vector space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
+or if the data are represented as a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
+While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
+An alternative algorithm is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.
+
+Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
+\end{example}
+
+We now provide a more formal description of graph data:
+Suppose we observe a network of \(n\) objects and pairwise relationships between them.
+This network is represented by a graph \(G = (V, E)\) with vertex set \(V = \{v_1, ..., v_n\}\), representing the \(n\) objects, and edge set \(E\), representing the up to \(n (n - 1) / 2\) pairwise relationships (assuming that there are no self-loops).
+The numeric representation of these data is in the form of \emph{affinity matrix} \(A \in \mathbb{R}^{n \times n}\) in which each \(A_{ij}\) represents object \(i\)'s relationship to object \(j\).
+We assume that the entries of \(A\) represent affinities or similarities, i.e., the higher the value of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\).
+If \(A_{ij} = 0\), then \(i\) has no direct relationship to \(j\).
+\(A\) is symmetric if it represents an undirected graph in which the relationship from \(i\) to \(j\) is the same as the relationship from \(j\) to \(i\).
+\(A\) is binary, i.e., \(A \in \{0, 1\}^{n \times n}\), if it represents an unweighted graph in which edges either exist or don't exist.
+If \(A\) is binary, we call it an \emph{adjacency matrix}.
+
+\begin{example}
+\label{ex:uk-faculty}
+\citet{PhysRevE.77.016107} constructed a friendship network among $81$ faculty from various schools at a university in the UK. 
+These data are represented as a graph in which each vertex is a faculty member and edges between pairs of vertices indicate whether the two faculty are friends. 
+The corresponding adjacency matrix $A \in \{0, 1\}^{81 \times 81}$ has zeros along the diagonal since there are no self-loops, and $A_{ij} = A_{ji} = 1$ if it is observed that the $i^{th}$ and $j^{th}$ faculty members are friends.
+The following is a visualization of this graph, with the vertices labeled by school affiliation.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/friendship-network-1}
+
+}
+
+\caption{Friendship network of 81 faculty at a UK university. The vertices are labeled by school affiliation.}\label{fig:friendship-network}
+\end{figure}
+\end{example}
 
 \hypertarget{probabilistic-models-for-graphs}{%
-\subsection{Probabilistic Models for
-Graphs}\label{probabilistic-models-for-graphs}}
+\subsection{Probabilistic Models for Graphs}\label{probabilistic-models-for-graphs}}
 
-Given a sample or dataset, a typical analysis task is statistical
-inference, or the estimation of various parameters assuming the data
-come from a random distribution or process. These estimated parameters
-are often then used for making predictions or deriving insights about
-the population. For example, when fitting a Gaussian mixture model, the
-data are first assumed to come from a mixture of Gaussians. The model
-fitting process then involves estimating the means and standard
-deviations of each Gaussian component, along with the mixture weights.
-The resulting model provides insight into where each mixture component
-is located, how disperse each component is, and how the data are
-distributed between the components, as well as a prediction indicating
-to which mixture a new observation belongs. In order to perform a
-similar type of analysis for graphs, we must first define probability
-distributions from which such data can be sampled.
+Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process.
+These estimated parameters are often then used for making predictions or deriving insights about the population.
+For example, when fitting a Gaussian mixture model, the data are first assumed to come from a mixture of Gaussians.
+The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights.
+The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction indicating to which mixture a new observation belongs.
+In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.
 
-Within the scope of this dissertation, we focus primarily on unweighted
-and undirected graphs without self-loops, with a brief discussion on
-generalizing these methods to weighted or directed graphs. The adjacency
-matrix that describes these graphs is binary, symmetric, and hollow. In
-this setting, a plausible model is to sample each edge independently
-from a Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
-\(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
-since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
-Then similar to how the edges are compiled into an adjacency matrix
-\(A\), the edge probabilities can be compiled into an edge probability
-matrix \(P \in [0, 1]^{n \times n}\). This type of graph model is
-defined as a \emph{Bernoulli graph}.
+Within the scope of this work, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs.
+The adjacency matrix that describes these graphs is binary, symmetric, and hollow.
+In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., \(A_{ij} \indep \Bernoulli(P_{ij})\) for some \(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\) since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
+Then similar to how the edges are compiled into an adjacency matrix \(A\), the edge probabilities can be compiled into an edge probability matrix \(P \in [0, 1]^{n \times n}\).
+This type of graph model is defined as a \emph{Bernoulli graph} (also called an inhomogeneous \(\ER\) graph).
+If \(A\) is the adjacency matrix of a Bernoulli graph with edge probability matrix \(P\), we denote \(A \sim \BG(P)\).
+If the vertices and edge probabilities are sampled as a sequence of \(n\) i.i.d. random variables from probability distribution \(F\), then we denote \(A \sim \BG(F, n)\).
 
-\begin{definition}[Bernoulli graph]
+Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
+Additional structure must be introduced.
+One such structured Bernoulli graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
+
+\begin{definition}[$\ER$ graph]
+\label{def:erdos-renyi}
+Let $P$ be an $n \times n$ matrix such that each $P_{ij} \equiv \theta \in [0, 1]$ is a constant. 
+Then $A \sim \BG(P)$ is an $\ER$ graph. 
 \end{definition}
 
-Statistical inference is not possible on a general Bernoulli graph with
-arbitrary edge probabilities since the number of parameters is equal to
-the number of observations. Additional structure must be introduced. One
-such structured Bernoulli graph model is the \(\ER\) graph, which is
-defined as follows.
+\begin{example}[Maximum likelihood estimator for the $\ER$ graph]
+One possible estimator for the lone parameter of an $\ER$ graph, $\theta$, is the maximum likelihood estimator, which is found by maximizing the log-likelihood function $\ell(\theta; A) = \log \theta \sum\limits_{i < j} A_{ij} + \log (1 - \theta) \sum\limits_{i < j} (1 - A_{ij})$.
+This can be solved directly by setting $\frac{d \ell}{d \theta} = 0$, which yields $\hat{\theta} = \frac{2 |E|}{n (n-1)}$.
+\end{example}
 
-\begin{definition}[$\ER$ graph]
-\end{definition}
+\(\ER\) graphs lie on the opposite end of the spectrum in that they restrict the model to a single parameter, which does not result in a very interesting model or reflect many networks observed in real data.
+Much work has been done in developing various Bernoulli graph models that strike a balance between structure and flexibility.
+Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called \emph{block models}, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called \emph{latent space models}.
 
-\hypertarget{block-models-for-community-detection}{%
-\subsection{Block Models for Community
-Detection}\label{block-models-for-community-detection}}
+\textbf{{[}Maybe insert something here about the exponential family of random graph models and how, like mixture distributions, block models do not fit within this family?{]}}
 
-\hypertarget{the-stochastic-block-model}{%
-\subsubsection{The Stochastic Block
-Model}\label{the-stochastic-block-model}}
+\begin{example}
+Consider a graph in which the vertices represent a collection of $n$ planets with intelligent life sending out radio waves, and the existence of an edge between vertices $i$ and $j$ represents whether planets $i$ and $j$ have established contact.
+Since radio waves decay according to the inverse-square law, a plausible model for the edge probability between two vertices is 
+$$P_{ij} = \frac{C \omega_i \omega_j}{d_{ij}^2}$$ 
+where $\omega_i$ is the $i^{th}$ planet's radio signal strength, $d_{ij}$ is the Euclidean distance between planets $i$ and $j$, and $C$ is a normalizing constant. 
+Let $P$ be the $n \times n$ matrix of these edge probabilities. 
+Then $A \sim \BG(P)$ represents the network of planets that have made contact.
 
-\hypertarget{generalizations-of-the-stochastic-block-model-the-degree-corrected-block-model-and-the-popularity-adjusted-block-model}{%
-\subsubsection{Generalizations of the Stochastic Block Model: the Degree
-Corrected Block Model and the Popularity Adjusted Block
-Model}\label{generalizations-of-the-stochastic-block-model-the-degree-corrected-block-model-and-the-popularity-adjusted-block-model}}
+To estimate $\omega = \begin{bmatrix} \omega_1 & \cdots & \omega_n \end{bmatrix}^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
+For simplicity, set $C = 1$. 
+Then the log-likelihood function can be written as:
+$$
+\ell(\omega) = 
+\sum_{i < j} A_{ij} \log \bigg( \frac{\omega_i \omega_j}{d_{ij}^2 -
+\omega_i \omega_j} \bigg) + \log (d_{ij}^2 - \omega_i \omega_j) + const.
+$$
+The partial derivatives with respect to each $\omega_i$ are:
+$$
+\frac{\partial \ell}{\partial \omega_i} = \sum_{j \neq i} \frac{A_{ij}}{\omega_i} - \frac{\omega_j (1 - A_{ij})}{d_{ij}^2 - \omega_i \omega_j}.
+$$
+Setting the gradient to zero yields $n$ equations with $n$ unknowns, but it is not entirely obvious how to solve this system of equations. 
+Alternatively, we can use gradient ascent or some other numerical optimization method, since this particular log-likelihood function is concave. 
+\end{example}
 
-\hypertarget{the-hierarchy-of-block-models}{%
-\subsubsection{The Hierarchy of Block
-Models}\label{the-hierarchy-of-block-models}}
+\hypertarget{contributions-of-this-work}{%
+\subsection{Contributions of this work}\label{contributions-of-this-work}}
 
-\hypertarget{review-of-traditional-statistical-inference-approaches-to-block-model-estimation}{%
-\subsubsection{Review of Traditional Statistical Inference Approaches to
-Block Model
-Estimation}\label{review-of-traditional-statistical-inference-approaches-to-block-model-estimation}}
+In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.
+The contributions of this work are as follows.
+First, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
+Using estimators with well-established consistency properties for the generalized random dot product graph, we develop consistent community detection and parameter estimation algorithms for the popularity adjusted block model.
 
@@
 
-\hypertarget{random-dot-product-graphs-and-generalized-random-dot-product-graphs}{%
-\section{Random Dot Product Graphs and Generalized Random Dot Product
-Graphs}\label{random-dot-product-graphs-and-generalized-random-dot-product-graphs}}
+\hypertarget{sec:blockmodel}{%
+\section{Block Models for Community Detection}\label{sec:blockmodel}}
+
+Network analysis is often concerned with community detection.
+This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices \(i\) and \(j\) depend on the communities to which vertices \(v_i\) and \(v_j\) belong.
+More formally, one assigns each vertex \(v_i\) a community label \(z_i\) and assumes a Bernoulli graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
+Within the context of this work, we will call such models \emph{block models}.
+This type of model frmaework allows us to frame the goal of community detection as a statistical inference problem:
+identify the true community labels, up to permutation of the labels.
+
+In this section, we explore the three main types of block models:
+the stochastic block model \citep{doi:10.1080/0022250X.1971.9989788}, the degree corrected block model \citep{Karrer_2011}, and the popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1}.
+The main focus is on how these models are connected, particularly in how they are nested models \citep{Noroozi2022}, as well as community detection and parameter estimation via likelihood maximization.
+We will later return to these block models and the relationship to another class of random graph models in sections \ref{sec:grdpg} and \ref{sec:pabm-grdpg}.
+
+\hypertarget{sec:sbm}{%
+\subsection{The Stochastic Block Model}\label{sec:sbm}}
+
+The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities has a constant edge probability \(\theta_{k \ell}\).
+We now give the formal definition of the SBM within the context of Bernoulli graphs:
+
+\begin{definition}[Stochastic block model]
+\label{def:sbm}
+Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$). 
+Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
+If the edge probability matrix $P$ for this graph is such that 
+$P_{ij} = \theta_{z_i, z_j}$ and $A \sim \BG(P)$, then $G$ is a stochastic block model. 
+\end{definition}
+
+There are a few conventions for exactly how a graph is sampled as an SBM.
+Within the context of this work, we assume that for a particular SBM, the number of communities, \(K\), and the community edge probabilities, \(\theta_{k \ell}\), are fixed.
+Then some possible ways of sampling from an SBM are as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Assume the size of the graph \(n\) is fixed.
+  Then each vertex \(v_i\) has a label \(z_i\), and the edge probability matrix \(P\) is fixed.
+\item
+  We sample the vertices and edges in sequence and suppose that each vertex \(v_i\) comes with a label \(z_i\), which determines its edge probabilities to the other vertices.
+\item
+  The vertices are sampled in sequence, and each \(z_i\) is sampled as \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), \(\sum_k^K \alpha_k = 1\).
+The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}\)
+\end{enumerate}
+
+An extension to these sampling schema include Bayesian models, e.g., each \(\theta_{k \ell}\) is sampled from a distribution with support \([0, 1]\).
+
+\begin{example}[Stochastic block model with $K = 2$ communities]
+\label{ex:assort-sbm}
+We construct an SBM with two communities in which $\theta_{11} = 1/2$, $\theta_{12} = 1/8$, and $\theta_{22} = 1/4$. 
+Note that in this example, the within-community edge probabilities are greater than the between-community edge probability. 
+A realization of this graph with $n_1 = n_2 = 32$ is illustrated by figure \ref{fig:assort-sbm}.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/assort-sbm-1}
+
+}
+
+\caption{Two-community stochastic block model with $\theta_{11}=1/2$, $\theta_{22}=1/4$, and $\theta_{12} = \theta_{21} = 1/8$.}\label{fig:assort-sbm}
+\end{figure}
+\end{example}
+
+In most conceptions of graph models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}.
+For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the graph appears to confirm this.
+However, SBMs are not restricted to this type of community structure.
+We could just as easily construct an example in which \(\theta_{11}\) and \(\theta_{22}\) are less than \(\theta_{12}\).
+Within the context of the SBM, we say that a graph is assortative if \(P\) is positive semidefinite and disassortative otherwise, which roughly correspond to communities with higher within-community edge probabilities and higher between-community edge probabilities, respectively.
+For example, the SBM in example \ref{ex:assort-sbm} is assortative and has edge probability matrix \(P\) that is rank 2 with nonzero eigenvalues \(n (\frac{3}{8} \pm \frac{\sqrt{2}}{8}) > 0\).
+In the following example, we describe a network which can be modeled as a disassortative SBM.
+
+\begin{example}[Dating network as a disassortative stochastic block model]
+\label{ex:sbm-dating}
+Consider an undirected graph in which each vertex $v_i$ represents users of an online dating service, each label $z_i$ represents the user's gender, and each edge represents a successful match between pairs of users.
+For simplicity, we will restrict the labels to female and male (denoted as $1$ and $2$). 
+If we model this as an SBM, then $\theta_{11}$ is the probability of a match between two female users, $\theta_{22}$ is the probability of a match between two male users, and $\theta_{12} = \theta_{21}$ is the probability of a match between a female user and a male user.
+Based on trends in the United States \citep{gallup-lgbt},
+a plausible set of edge probabilities is $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$, which are used in the sample visualized in figure \ref{fig:dating-sbm}, along with $\alpha_1 = \alpha_2 = 1/2$ and $n = 64$.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/dating-sbm-1}
+
+}
+
+\caption{Stochastic block model of a dating network. This model is disassortative.}\label{fig:dating-sbm}
+\end{figure}
+\end{example}
+
+\begin{example}[Assortativity and disassortativity of the two-community SBM]
+Let $P$ be the $n \times n$ probability matrix of a two-community SBM. 
+Then $rank(P) = 2$, and its two nonzero eigenvalues are:
+
+$$
+\lambda = \frac{n}{2} \bigg( 
+  \theta_{11} + \theta_{22} \pm 
+  \sqrt{(\theta_{11} - \theta_{22})^2 + 4 \theta_{12}^2} 
+\bigg).
+$$
+The first eigenvalue is always positive. 
+The second eigenvalue is positive if $\theta_{11} \theta_{22} > \theta_{12}^2$. 
+Thus, $P$ describes an assortative SBM if $\theta_{11} \theta_{22} > \theta_{12}^2$ or 
+a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
+\end{example}
+
+One approach to statistical inference for the SBM is likelihood maximization.
+The log-likelihood function can be written as:
+\begin{equation}
+\label{eq:sbm-loglik}
+\ell(z, \theta; A) = \sum_{i < j} \sum_{k, \ell} z_{ik} z_{j \ell} \big( 
+A_{ij} \log \theta_{k \ell} - (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big),
+\end{equation}
+where \(z_{ik} = I(z_i = k)\). If the labels are drawn from a multinomial distribution, then that can also be incorporated into the complete-data log-likelihood:
+\begin{equation}
+\label{eq:sbm-loglik-full}
+\ell(z, \theta, \alpha; A) = \sum_k \sum_i z_{ik} \log \alpha_k + 
+\sum_{i < j} \sum_{k, \ell} z_{ik} z_{j \ell} \big( 
+A_{ij} \log \theta_{k \ell} - (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big).
+\end{equation}
+
+The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard.
+Because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the expectation maximization (EM) algorithm \citep{10.2307/2984875}.
+However, due to the \(z_{ik} z_{j \ell}\) interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}.
+While the labels are drawn independently, they are not necessarily conditionally independent.
+Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form.
+As an example, we write the EM algorithm for the likelihood function in equation \ref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
+
+\begin{algorithm}
+\label{algo:sbm-em}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}$}
+  \caption{Approximate EM algorithm for the SBM}
+  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}$.\;
+  \While{$\|\nabla \ell\| > \epsilon$} {
+    \For {$i = 1, ..., n$} {
+      \For {$k = 1, ..., K$} {
+        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \theta_{k \ell} + (1 - A_{ij}) \log (1 - \theta_{k \ell})) \bigg)$.\;
+        M-step: $\theta_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
+      }
+    }
+  }
+\end{algorithm}
+
+Applying algorithm \ref{algo:sbm-em} to example \ref{ex:assort-sbm} with initial guesses of \(z_{ik} = 0.5\) for each \(i, k\), \(\hat{\theta}_{11} = \hat{\theta}_{22} = 0.9\), and \(\hat{\theta}_{12} = 0.1\) results in a community detection error rate of 4.7\% and parameter estimates \(\hat{\theta}_{11} = 0.503\), \(\hat{\theta}_{22} = 0.241\), and \(\hat{\theta}_{12} = 0.109\), compared to the true parameters \(\theta_{11} = 0.5\), \(\theta_{22} = 0.25\), and \(\theta_{12} = 0.125\).
+The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of \(z_{ik} = 0.5\), \(\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1\), and \(\hat{\theta}_{12} = 0.9\), results in a community detection error rate of 0\% and parameter estimates \(\hat{\theta}_{11} = 0.034\), \(\hat{\theta}_{22} = 0.018\), and \(\hat{\theta}_{12} = 0.201\), compared to the true parameters \(\theta_{11} = \theta_{22} = 0.02\) and \(\theta_{12} = 0.2\).
+
+\hypertarget{random-graph-models-and-sparsity}{%
+\subsubsection{Random Graph Models and Sparsity}\label{random-graph-models-and-sparsity}}
+
+\hypertarget{sec:dcbm-pabm}{%
+\subsection{Generalizations of the Stochastic Block Model: The Degree Corrected Block Model and the Popularity Adjusted Block Model}\label{sec:dcbm-pabm}}
+
+\begin{definition}[Degree corrected block model]
+\label{def:dcbm}
+Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM. 
+Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
+In addition, each vertex $v_i$ has a degree correction parameter $\omega_i \in [0, 1]$. 
+If the edge probability matrix $P$ for this graph is such that $P_{ij} = \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model.
+\end{definition}
+
+\begin{example}[Dating network as a disassortative DCBM]
+\label{ex:dcbm-dating}
+In example \ref{ex:sbm-dating}, we modeled an online dating network as an SBM in which the vertices correspond to individuals, the communities correspond to genders, and edges correspond to whether the online dating service successfully matched pairs of individuals.
+In this model, we assume that for each pair of genders, there is a constant probability of a match. 
+This does not take into account that some individuals are more likely to form connections than others.
+
+\end{example}
+
+\begin{definition}[Popularity adjusted block model]
+\label{def:pabm}
+Let $G = (V, E)$ be an undirected Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $K \geq 1$ be an integer that describes the number of communities, and let 
+$z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
+Suppose that each vertex $v_i$ has $K$ popularity parameters $\lambda_{i1}, ..., \lambda_{iK}$ for which each $\lambda_{ik}$ describes $v_i$'s affinity toward community $k$. 
+Then if the edge probability matrix $P$ for this graph is such that $P_{ij} = \lambda_{i, z_j} \lambda_{j, z_i}$ and $A \sim \BG(P)$, then $G$ is a popularity adjusted block model.
+\end{definition}
+
+\begin{example}[Dating network as a PABM]
+\label{ex:pabm-dating}
+\end{example}
+
+\hypertarget{the-hierarchy-of-block-models}{%
+\subsubsection{The Hierarchy of Block Models}\label{the-hierarchy-of-block-models}}
+
+\newpage
+
+\hypertarget{sec:grdpg}{%
+\section{Random Dot Product Graphs and Generalized Random Dot Product Graphs}\label{sec:grdpg}}
 
@@
 
-\hypertarget{connecting-the-stochastic-block-model-to-the-generalized-random-dot-product-graph}{%
-\subsection{Connecting the Stochastic Block Model to the Generalized
-Random Dot Product
-Graph}\label{connecting-the-stochastic-block-model-to-the-generalized-random-dot-product-graph}}
+\begin{definition}[Generalized random dot product graph]
+\label{def:grdpg}
+Let $p \geq 1$ and $q \geq 0$ be integers. 
+Define $I_{p,q}$ as the block diagonal matrix $I_{p,q} = \Bigl[\begin{smallmatrix} I_{p} & 0 \\ 0 & - I_{q} \end{smallmatrix} \Bigr]$ where $I_p$ and $I_q$ are the identity matrices of dimensions $p \times p$ and $q \times q$ respectively.
+Denote $d = p + q$ and let $\mathcal{X}$ be the subset of $\mathbb{R}^d$ such that, for any $x, y \in \mathcal{X}$, we have $x^\top I_{p,q} y \in [0, 1]$. 
+Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
+A graph $G$ with adjacency matrix $A$ is said to be a generalized random dot product graph with latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$ if $A \sim \BG(P)$ where the edge probability matrix $P$ is given by $P = \rho_n X I_{p,q} X^\top$, i.e., the entries of $P$ are of the form $P_{ij} = \rho_n x_i^\top I_{p,q} x_j$.
 
-\hypertarget{connecting-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}{%
-\subsection{Connecting the Degree Corrected Block Model to the
-Generalized Random Dot Product
-Graph}\label{connecting-the-degree-corrected-block-model-to-the-generalized-random-dot-product-graph}}
+We use the notation $A \sim \GRDPG_{p,q}(X; \rho_n)$ to denote a random adjacency matrix $A$ drawn from latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$.
+If the $n$ vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F, n; \rho_n)$.
+
+If $q = 0$ and so $d = p$, (i.e., $P$ is positive semidefinite), then we call this a random dot product graph. 
+In this case, we use the notation $A \sim \RDPG(X; \rho_n)$ and $A \sim \RDPG(F, n; \rho_n)$. 
+\end{definition}
+
+\begin{definition}[Adjacency spectral embedding]
+\label{def:ase}
+\end{definition}
+
+\begin{definition}[Indefinite orthogonal group]
+\label{def:indef-ortho-group}
+\end{definition}
+
+\begin{remark}
+\label{rem:non_identifiable}
+The latent vectors that produce $X I_{p,q} X^\top = P$ are not unique
+\citep{rubindelanchy2017statistical}.
+More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, then for any $Q \in \mathbb{O}(p, q)$ we also have 
+$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j = x_i^\top I_{p, q} x_j = P_{ij}$. 
+Unlike in the RDPG case, transforming the latent positions via multiplication by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or distances.
+\end{remark}
+
+\hypertarget{sec:sbm-dcbm-grdpg}{%
+\subsection{Connecting the Stochastic Block Model and the Degree Corrected Block Model to the Generalized Random Dot Product Graph}\label{sec:sbm-dcbm-grdpg}}
 
@@
 
-\hypertarget{popularity-adjusted-block-models-are-generalized-random-dot-product-graphs}{%
-\section{Popularity Adjusted Block Models are Generalized Random Dot
-Product
-Graphs}\label{popularity-adjusted-block-models-are-generalized-random-dot-product-graphs}}
+\hypertarget{sec:pabm-grdpg}{%
+\section{Popularity Adjusted Block Models are Generalized Random Dot Product Graphs}\label{sec:pabm-grdpg}}
+
+In this section, we connect the PABM to the GRDPG and exploit that connection to develop algorithms for community detection and parameter estimation.
+In order to make the explicit connection between the PABM and the GRDPG, we make use of an alternative but equivalent definition of the PABM which parameterizes the model in terms of popularity vectors, which are collections of popularity parameters.
+
+\begin{remark}
+\label{rem:pabm_view2}
+In a PABM, each vertex $i$ has $K$ popularity parameters $\lambda_{i1}, \dots, \lambda_{iK}$, that describe its affinity toward each of the $K$ communities. 
+Another view of a PABM is as follows.
+Let $\tilde{P}$ be the matrix obtained by permuting the rows and columns of $P$ so that the vertices are reorganized by community memberships $z_i \in \{1,2,\dots,K\}$ in increasing order. 
+Denote by $\tilde{P}^{(k \ell)}$ the $n_k \times n_{\ell}$ submatrix of $\tilde{P}$ corresponding to the edge probabilities between vertices in communities $k$ and $\ell$. 
+Here $n_k = |\{ i \colon z_i = k\}|$ is the number of vertices assigned to community $k$, for $k = 1,2\dots,K$.
+Note that $\tilde{P}^{(k \ell)} = (\tilde{P}^{(\ell k)})^\top$. 
+Next let $\lambda^{(k \ell)} = \{\lambda_{i \ell} \colon z_i = k\} \in \mathbb{R}^{n_k}$; the elements of $\lambda^{(k \ell)}$ are the affinity parameters toward the $\ell$th community of all vertices in the $k^{th}$ community.
+Define $\lambda^{(\ell k)}$ analogously. 
+Then each block $\tilde{P}^{(k \ell)}$ can be written as the outer product of two vectors:
+\begin{equation} \label{eq:pabm}
+  \tilde{P}^{(k \ell)} = \rho_n \lambda^{(k \ell)} (\lambda^{(\ell k)})^{\top}.
+\end{equation}
+
+We will henceforth use the notation \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K, \rho_n)\) to denote a random adjacency matrix \(A\) drawn from a PABM with $K$ communities, popularity parameters \(\{\lambda^{(k \ell)}\}\) and sparsity parameter $\rho_n$.
+\end{remark}
+
+\hypertarget{the-geometry-of-pabms}{%
+\subsection{The Geometry of PABMs}\label{the-geometry-of-pabms}}
+
+In section \ref{sec:sbm-dcbm-grdpg}, we showed how the SBM and DCBM are special cases of the GRDPG in which the latent vectors form a very particular geometry.
+Similarly, we now show the special geometry of the PABM when viewed as a GRDPG.
+For ease of exposition, and without loss of generality, we drop the dependency on the sparsity parameter \(\rho_n\) and assume \(\rho_n \equiv 1\) throughout this subsection.
+
+\begin{theorem}[The latent configuration of the PABM]
+\label{thm:pabm-grdpg}
+Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_K)$ be an instance of a
+PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
+\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. 
+Then there exists a block diagonal matrix
+$X \in \mathbb{R}^{n \times K^2}$ defined by $\{\lambda^{(k \ell)}\}$ and a 
+$K^2 \times K^2$ {\em fixed} orthonormal matrix $U$ such 
+that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) /
+  2}(\tilde{\Pi}XU)$. Here $\tilde{\Pi}$ is the permutation matrix
+such that $P = \tilde{\Pi} \tilde{P} \tilde{\Pi}^{\top}$ where the
+rows and columns of $\tilde{P}$ are arranged according to increasing values of the
+community labels (see remark~\ref{rem:pabm_view2}). 
+\end{theorem}
+
+\begin{proof}
+We will prove this theorem in two parts. First, for demonstration
+purposes, we focus on the case when $K = 2$ to build intuition. 
+The general case of $K \geq 2$ is presented later.
+
+For the $K = 2$ case, the proof is straightforward. We will first work with
+the matrix $\tilde{P}$. Note that $\tilde{P}$ has the form
+
+$$\tilde{P} = \begin{bmatrix} P^{(11)} & P^{(12)} \\ P^{(21)} &
+  P^{(22)} \end{bmatrix} = \begin{bmatrix} \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
+  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)}
+  (\lambda^{(22)})^\top \end{bmatrix}.$$
+  Now let
+$$X = \begin{bmatrix}
+\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
+0 & 0 & \lambda^{(21)} & \lambda^{(22)}
+\end{bmatrix} \quad \text{and} \quad
+U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
+0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
+0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
+0 & 1 & 0 & 0 \end{bmatrix}.$$
+Then by straightforward matrix multiplication, we obtain 
+$$X U I_{3, 1} U^\top X^\top =
+\begin{bmatrix}
+  \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
+  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
+\end{bmatrix} = \tilde{P}$$
+and hence $\tilde{P}$ also corresponds to the edge probability matrix of GRDPG
+with latent vectors described by $X U$. As $P = \tilde{\Pi} \tilde{P}
+\tilde{\Pi}^{\top}$ we conclude that $P$ has latent vectors described
+by $\tilde{\Pi} X U$.
+
+It is nevertheless instructive to look at a few intermediate steps. 
+More specifically, the product $U I_{3, 1} U^\top$ 
+yields a permutation matrix $\Pi$ with fixed points at positions $1$ and $4$ 
+and a cycle of order 2 swapping positions $2$ and $3$, i.e., 
+$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
+  0 & 0 & 1 & 0 \\
+  0 & 1 & 0 & 0 \\
+  0 & 0 & 0 & 1
+\end{bmatrix}.$$
+Furthermore, as $U$ is orthonormal and $I_{3, 1}$ is diagonal, 
+$U I_{3, 1} U^\top$ is also an eigendecomposition of $\Pi$ where the fixed
+points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
+while the cycles of order two are mapped to the eigenvectors  
+$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
+e_3)$; here $e_i$ denote the $i^\mathrm{th}$ basis vector in $\mathbb{R}^{4}$. 
+Thus, another way of decomposing the edge probability matrix is
+$\tilde{P} = X \Pi X^\top$ where the rows of $X$ lie in the union of
+two 2-dimensional orthogonal subspaces and $\Pi$ is a permutation matrix.
+
+For the general case, we can extend $\tilde{P} = X \Pi X^\top$ to larger $K$. 
+For a more concrete example of this, refer to Example~1. 
+We once again consider $\tilde{P}$ as defined in
+remark~\ref{rem:pabm_view2}.  
+We first define the following matrices
+
+\begin{gather}
+\label{eq:xy}
+\Lambda^{(k)} = \begin{bmatrix} \lambda^{(k1)} \mid \cdots \mid \lambda^{(kK)} \end{bmatrix}
+\in \mathbb{R}^{n_k \times K}, \quad
+X = \mathrm{blockdiag}(\Lambda^{(1)}, \dots, \Lambda^{(K)}) \in
+\mathbb{R}^{n \times K^2}, \\
+L^{(k)} = \mathrm{blockdiag}(\lambda^{(1k)}, \dots, \lambda^{(Kk)}) \in
+\mathbb{R}^{n \times K}, \quad
+Y = \begin{bmatrix} L^{(1)} \mid \cdots \mid L^{(K)} \end{bmatrix} \in
+\mathbb{R}^{n \times K^2}.
+\end{gather}
+
+It is then straightforward to verify that
+
+\begin{gather*}
+  XY^{\top} = \mathrm{blockdiag}(\Lambda^{(1)}, \dots,
+\Lambda^{(K)}) \begin{bmatrix} L_1^{\top} \\ \vdots \\
+  L_{K}^{\top} \end{bmatrix} = \begin{bmatrix} \Lambda^{(1)}
+  (L^{(1)})^{\top} \\ \vdots \\
+  \Lambda^{(K)} (L^{(K)})^{\top} \end{bmatrix}, \\
+\Lambda^{(k)} (L^{(k)})^{\top} = \begin{bmatrix} \lambda^{(k1)}
+  (\lambda^{(1k)})^{\top} \mid \dots \mid \lambda^{(kK)}
+  (\lambda^{(Kk)})^{\top} \end{bmatrix} = \begin{bmatrix} P^{(k1)}
+  \mid P^{(k2)} \mid \dots \mid P^{(kK)} \end{bmatrix}.
+\end{gather*}
+We therefore have $\tilde{P} = X Y^\top$. 
+Similar to the $K = 2$ case, we also have $Y = X \Pi$ for some permutation matrix
+$\Pi$ and hence $\tilde{P} = X \Pi X^\top$.
+The permutation described by $\Pi$ now has $K$ fixed points, which correspond to
+$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where
+$k = r (K + 1) + 1$ for $0 \leq r \leq K - 1$. It also has
+$\tbinom{K}{2}$ cycles of order $2$. Each cycle corresponds to
+a pair of eigenvalues $\{-1,+1\}$ and a pair of eigenvectors
+$\{(e_s + e_t)/\sqrt{2},(e_s - e_t)/ \sqrt{2}\}$.
+
+Let $p = K (K + 1) / 2$ and $q = K (K - 1) / 2$. 
+We therefore have
+\begin{equation} \label{eq:permutation}
+\Pi = U I_{p,q} U^\top
+\end{equation}
+where $U$ is a $K^2 \times K^2$ orthogonal matrix and hence
+\begin{equation} \label{eq:pabm-grdpg}
+\tilde{P} = X U I_{p, q} (X U)^\top.
+\end{equation}
+In summary we can describe the PABM with $K$ communities as a GRDPG with latent
+positions $\tilde{\Pi} X U$ and known signature $(p,q) = \bigl( \tfrac{1}{2} K (K + 1) ,
+\tfrac{1}{2} K (K - 1)\bigr)$.
+\end{proof}
+
+\begin{example} Let $A$ be a $3$ blocks PABM with latent vectors $\{\lambda^{(k \ell)} \colon 1 \leq k \leq 3, 1 \leq \ell \leq 3\}$. 
+Using the same notation as in theorem~\ref{thm:pabm-grdpg}, we can define
+\begin{gather*}
+X = \begin{bmatrix}
+\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
+0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
+0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
+\end{bmatrix}, \\
+Y = \begin{bmatrix}
+\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
+0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
+0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
+\end{bmatrix}.
+\end{gather*}
+Then $Y = X \Pi$ and $\tilde{P} = X Y^{\top}$ where $\Pi$ is a $9 \times 9$ 
+permutation matrix of the form
+$$\Pi = 
+\Bigl[e_1 \mid e_4 \mid e_7 \mid e_2 \mid e_5 \mid e_8 \mid e_3
+\mid e_6 \mid e_9 \Bigr].$$
+where $e_i$ denotes the $i^{th}$ basis vector in $\mathbb{R}^{9}$. 
+The matrix $\Pi$ corresponds to a permutation of $\{1,2,\dots,9\}$ with the following decomposition.
+\begin{enumerate}
+\item Positions 1, 5, 9 are fixed.
+\item There are three cycles of length 2, namely $(2, 4)$, $(3, 7)$, and $(6, 8)$.
+\end{enumerate}
+We can thus write $\Pi$ as $\Pi = U I_{6, 3} U^\top$ where the first three columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the fixed points, the next three columns are the eigenvectors $(e_k + e_{\ell}) / \sqrt{2}$, and the last three columns are the eigenvectors $(e_k - e_{\ell}) / \sqrt{2}$ for $(k, \ell) \in \{(2,4),(3,7),(6,8)\}$.
+
+The matrix $\tilde{P}$ is then the edge probabilities matrix for a 
+Generalized Random Dot Product Graph whose latent positions 
+are the rows of the matrix
+$$XU = \begin{bmatrix}
+  \lambda^{(11)} & 0 & 0 &
+  \frac{\lambda^{(12)}}{\sqrt{2}} & \frac{\lambda^{(13)}}{\sqrt{2}} & 0 &
+  \frac{\lambda^{(12)}}{\sqrt{2}} & \frac{\lambda^{(13)}}{\sqrt{2}} & 0 \\
+  0 & \lambda^{(22)} & 0 &
+  \frac{\lambda^{(21)}}{\sqrt{2}} & 0 & \frac{\lambda^{(23)}}{\sqrt{2}} &
+  -\frac{\lambda^{(21)}}{\sqrt{2}} & 0 & \frac{\lambda^{(23)}}{\sqrt{2}} \\
+  0 & 0 & \lambda^{(33)} &
+  0 & \frac{\lambda^{(31)}}{\sqrt{2}} & \frac{\lambda^{(32)}}{\sqrt{2}} &
+  0 & -\frac{\lambda^{(31)}}{\sqrt{2}} & -\frac{\lambda^{(32)}}{\sqrt{2}}
+\end{bmatrix}$$
+and the latent positions for $P$ is a permutation of the rows of
+$XU$. 
+\end{example}
+
+\hypertarget{algorithms}{%
+\subsection{Algorithms}\label{algorithms}}
+
+Two inference objectives arise from the PABM:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Community membership identification (up to permutation).
+\item
+  Parameter estimation (estimating $\lambda_{ik}$'s).
+\end{enumerate}
+
+In our methods, the data that are observed for estimation is the adjacency matrix, \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K, \rho_n)\), along with an assumed number of communities, \(K\).
+To motivate our methods, we first consider community detection and parameter estimation in the case where we know the edge probability matrix \(P\) beforehand, noting that community memberships and popularity parameters are not immediately discernible from \(P\) itself.
+After establishing methods for community detection and parameter estimation from \(P\), we use the consistency property of the ASE \citep{doi:10.1080/01621459.2012.699795,rubindelanchy2017statistical} to demonstrate that the same methods work for \(A\) almost surely as \(n \to \infty\).
+
+\hypertarget{previous-work}{%
+\subsubsection{Previous Work}\label{previous-work}}
+
+\citet{307cbeb9b1be48299388437423d94bf1}
+used Modularity Maximization (MM) and the Extreme Points (EP)
+algorithm \citep{le2016} for community detection and parameter
+estimation. They were able to show that as the sample size increases,
+the \emph{proportion} of misclassified community labels (up to permutation)
+goes to 0.
+
+\cite{noroozi2019estimation} used Sparse Subspace Clustering (SSC)
+\citep{5206547} for community detection in the PABM.
+The SSC algorithm can be described as follows:
+Given \(X \in \mathbb{R}^{n \times d}\) with vectors
+\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
+\(c_i = \argmin_{c} \|c\|_1\) subject to \(x_i = X^\top c\) and
+\(c^{(i)} = 0\), where \(c^{(i)}\) is the \(i^{th}\) entry of \(c\), is solved for each \(i \in [n]\).
+The solutions are collected into matrix
+\(C = \bigl[ c_1 \mid \cdots \mid c_n \bigr]^\top\) to
+construct an affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lies
+exactly on one of \(K\) subspaces, \(B\) describes an undirected graph
+consisting of \emph{at least} \(K\) disjoint subgraphs, i.e., \(B_{ij} = 0\) if \(x_i, x_j\) lie on different subspaces.
+The intuition here is that vectors that lie on the same subspace can be described as linear combinations of each other, assuming the number of vectors in the subspace is greater than the dimensionality of the subspace.
+Then once sparsity is enforced, for each \(c_i\), its \(j^{th}\) element \(c_i^{(j)}\) is zero if \(x_j\) belongs to a subspace that doesn't contain \(x_i\), resulting in \(B_{ij} = 0\).
+If \(X\) instead represents points near \(K\) subspaces with some noise,
+then this property may only hold approximately and a final graph partitioning step may be required
+(e.g., edge thresholding or spectral clustering).
+
+In practice, due to presence of noise, SSC is often done by solving the LASSO problems
+\begin{equation} 
+\label{eq:ssc}
+c_i = \argmin_c \frac{1}{2} \|x_i - X_{-i}^\top c\|^2_2 + \vartheta \|c\|_1
+\end{equation}
+for some sparsity parameter \(\vartheta > 0\).
+The \(c_i\) vectors are then collected into \(C\) and \(B\) as before.
+
+\begin{definition}[Subspace Detection Property]
+\label{def:subspace_detection}
+Let $X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top$ be noisy points sampled from $K$ subspaces, i.e., $x_i = y_i + z_i$ where the $y_i$ belongs to the union of $K$ subspaces and the $z_i$ are noise vectors.
+Let $\vartheta \geq 0$ be given and let $C$ and $B$ be constructed from the solutions of LASSO problems as described in equation~\ref{eq:ssc} with this given choice of $\vartheta$. 
+Then $X$ is said to satisfy the subspace detection property with sparsity parameter $\vartheta$ if each column of $C$ has nonzero $\ell_2$ norm and $B_{ij} = 0$ whenever $y_i$ and $y_j$ are from different subspaces.
+\end{definition}
+
+\begin{remark}
+\label{rem:sdp}
+One common approach to show that SSC works for a noisy sample $X$ is to show that $X$ satisfies the subspace detection property for some choice of $\vartheta$; 
+recall that $\vartheta$ is the sparsity parameter for
+the LASSO problems in equation~\ref{eq:ssc}. However, this is not sufficient to
+guarantee that SSC perfectly recovers the underlying subspaces.
+More specifically, if $X$ satisfies the 
+subspace detection property, 
+then $B$ describes a graph with {\em at least} $K$ disconnected subgraphs, 
+with the ideal case being that there are exactly $K$ subgraphs 
+which map onto each subspace. 
+Nevertheless it is also possible that the $K$ subspaces are represented by
+$K' > K$
+multiple disconnected subgraphs and we cannot, at least without a subsequent
+post-processing step, recover the $K$ subspaces directly from $B$;
+see \citet{sdp_sufficiency} and \citet{liu_ssc}
+for further discussions. Therefore in practice $B$ is usually treated as an
+affinity matrix and, as we allude to earlier, 
+the rows of $B$ are partitioned using some
+clustering algorithm to obtain the final clustering. 
+\end{remark}
+
+Theorem \ref{thm:pabm-grdpg} suggests that SSC is appropriate for community
+detection for the PABM, provided that we observe the edge probabilities
+matrix \(P\). More precisely, given the matrix \(\tilde{P}\) obtained by
+permuting the rows and columns of \(P\) as described in
+remark \ref{rem:pabm_view2} we can recover \(XU\) up
+to some non-identifiability indefinite orthogonal transformation \(Q\).
+Then using results from \citet{soltanolkotabi2012}, it can be easily shown
+that the subspace detection property holds for \(XU\). Indeed, the columns
+of \(XU\) from different communities correspond to mutually
+orthogonal subspaces. This then implies that the
+subspace detection property also holds for \(XUQ\) for all invertible
+transformation \(Q\) and hence the subspace detection property also holds for
+\(\tilde{\Pi} X U Q\) for any \(n \times n\) permutation matrix \(\tilde{\Pi}\).
+
+However, because we do not observe \(P\) but rather only the noisy
+adjacency matrix \(A \sim \BG(P)\), the natural approach then is
+to perform SSC on the rows of the spectral embedding of \(A\),
+since the embedding of \(P\) consists of \(K\) subspaces (theore \ref{thm:pabm-grdpg}),
+and so the embedding of \(A\) will lie approximately on the \(K\) subspaces.
+We will show in theorem \ref{theorem5} that, with probability converging to
+one as \(n \rightarrow \infty\), the rows of the ASE of \(A\) also
+satisfy the subspace detection property.
+Theorem \ref{theorem5} builds upon existing work by
+\citet{rubindelanchy2017statistical} who describe the convergence
+behavior of the ASE of \(A\) to that of \(\tilde{\Pi} XU\), and
+\citet{jmlr-v28-wang13} who show the necessary conditions for the
+subspace detection property to hold in noisy cases where the points lie
+near subspaces. Finally we emphasize that while
+\citet{noroozi2019estimation} also considered the use of SSC for
+community recovery in PABM, they instead applied SSC to the rows of
+\(A\) itself, foregoing
+the embedding step altogether. It is however much harder to show that
+the rows of \(A\) satisfy the subspace detection property and thus, to
+the best of our knowledge, there is currently no consistency result
+regarding the application of SSC to the rows of \(A\).
+
+We previously stated in theorem \ref{thm:pabm-grdpg} one possible set of latent positions that result in
+the edge probability matrix of a PABM, namely
+\(P = \tilde{\Pi} (XU) I_{p, q} (XU)^\top \tilde{\Pi}^{\top}\) where \(X\) is block diagonal and
+\(\tilde{\Pi}\) is a permutation matrix.\\
+Furthermore, the explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
+such that points within each community lie on \(K\)-dimensional
+orthogonal subspaces, i.e.~\(\langle U^{\top} x_i, U^{\top} x_j \rangle = 0\) whenever \(i\) and \(j\) are in different communities.
+Thus if we have (or can estimate) \(XU\) directly, then both the community
+detection and parameter identification problem are trivial because \(U\)
+is orthonormal and fixed for each value of \(K\).
+However, direct
+identification or estimation of \(XU\) is possibly difficult
+due to the non-identifiability of \(XU\) (see remark \ref{rem:non_identifiable})
+when we are given only \(P\).
+More specifically, suppose we find a matrix \(Y \in \mathbb{R}^{n \times K^2}\)
+such that \(P = Y I_{p, q} Y^\top\). Then it is generally the case that
+\(Y = \tilde{\Pi} XU Q\) for some indefinite orthogonal matrix
+\(Q \in \mathbb{O}(p,q)\).
+However since \(Q\) is not necessarily an
+orthogonal matrix and hence, if \(y_i\) denote the \(i^{th}\) row of \(Y\),
+then \(\langle U^{\top} x_i, U^{\top} x_j \rangle \neq \langle y_i, y_j \rangle\).
+This prevents us from transferring the orthogonality property of
+\(XU\) directly to \(Y\).
+
+Nevertheless by using the special geometric structure of \(X\) we can circumvent the
+non-identifiability of \(Y\) and \(XU\) by using instead the rows of the
+matrix \(V\) of eigenvectors (corresponding to the non-zero eigenvalues) of \(P\). In particular \(V\) is identifiable
+up to orthogonal transformations and furthermore, due to the block
+diagonal structure of \(X\), the rows of \(V\) also lie on \(K\) distinct orthogonal
+subspaces and hence \(v_i^{\top} v_j = 0\) whenever \(z_i \not = z_j\).
+
+\begin{theorem}
+\label{thm:osc-p}
+Let $P = V D V^\top$ be the spectral decomposition 
+of the edge probability matrix. 
+Let $B = n V V^\top$. 
+Assume $\lambda_{i z_i} > 0$ for each $i \in [n]$, 
+i.e., each vertex's popularity parameter to its own community is nonzero. 
+Then $B_{ij} = 0$ if and only if 
+vertices $i$ and $j$ are in different communities.
+\end{theorem}
+
+\begin{proof}
+We first show that $V V^\top =
+\tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top} $ where $X$ is 
+defined as in Eq.~\eqref{eq:xy}. Indeed, by Theorem 2, 
+\(P = \tilde{\Pi} X U I_{p, q} U^\top X^\top \tilde{\Pi}\) for $p = K(K+1)/2$ and $q = K(K-1)/2$. 
+The eigendecomposition \(P = V D V^\top\) also yields $P = V
+|D|^{1/2} I_{p, q} |D|^{1/2} V^\top$ where \(|\cdot|^{1/2}\) is
+applied entry-wise. Now let $Y = \tilde{\Pi} XU$ and $\tilde{Y} = V|D|^{1/2}$; note that
+$Y$ and $\tilde{Y}$ both have full column ranks. 
+Because $P = Y I_{p,q} Y^{\top} = \tilde{Y} I_{p,q} \tilde{Y}^{\top}$, we have
+$$Y = \tilde{Y} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}.$$
+Let $Q = I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q}$ and note that
+$Y = \tilde{Y} Q$. We then have
+\begin{equation*}
+  \begin{split}
+  Q^{\top} I_{p,q} Q &= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} \tilde{Y} I_{p,q}
+I_{p,q} I_{p,q} \tilde{Y}^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} \\ 
+&= I_{p,q} (Y^{\top} Y)^{-1} Y^{\top} Y I_{p,q}
+Y^{\top} Y (Y^{\top} Y)^{-1} I_{p,q} =  I_{p,q}
+\end{split}
+\end{equation*}
+and hence $Q$ is an indefinite orthogonal matrix.
+
+Let $R = U Q |D|^{-1/2}$ and note that $V = \tilde{\Pi} XR$. Because $R$ is
+invertible, we can write
+$$\tilde{\Pi} X (X^{\top} X)^{-1} X^{\top} \tilde{\Pi}^{\top} =
+\tilde{\Pi} X R (R^{\top} X^{\top} X R)^{-1}
+R^{\top} X^{\top} \tilde{\Pi}^{\top}.$$ 
+Furthermore, as $V$ has orthonormal columns, $R^{\top} X^{\top} X R =
+V^{\top} \tilde{\Pi} \tilde{\Pi}^{\top} V = V^{\top} V = I$. We thus conclude
+$$\tilde{\Pi} X (X^{\top} X)^{-1} X^{\top} \tilde{\Pi}^{\top} = V (V^{\top} V)^{-1} V^{\top} = V V^{\top}$$
+as desired.
+
+To complete the proof of theorem \ref{thm:osc-p}, recall that \(X\) 
+is block diagonal with each block corresponding to one community, 
+and hence \(X (X^\top X)^{-1} X^\top\) is also a
+block diagonal matrix with each block corresponding to a community. 
+As $B = n VV^{\top} = n \tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top}$, 
+we conclude that $B_{ij} = 0$ 
+whenever vertices $i$ and $j$ belong to different communities.  
+\end{proof}
+
+Theorem \ref{thm:osc-p} provides perfect community detection from \(P\).
+More specifically, let \(|B|\) be the affinity matrix for graph \(G'\),
+where \(|\cdot|\) is applied entry-wise. Then
+\(G'\) consists of exactly \(K\) disjoint subgraphs,
+as \(G'\) has no edges between communities.
+All that is left to identify the communities is
+to assign each subgraph a distinct community label.
+In practice, we do not observe \(P\) and instead only observe the noisy
+\(A \sim \BG(P)\). A natural approach is then to use
+the affinity matrix \(\hat{B} = n \hat{V} \hat{V}^{\top}\) where
+\(\hat{V}\) is the matrix of eigenvectors (corresponding to the largest
+eigenvalues in modulus) of \(A\). The resulting procedure, named
+Orthogonal Spectral Clustering, is presented in
+algorithm \ref{alg:osc}.
+The following result leverages existing theoretical properties
+of ASE for estimating of latent positions in a GRDPG \citep{rubindelanchy2017statistical} to show that
+\(\hat{B}\) converges almost surely to \(B\); in particular
+\(\hat{B}_{ij} \stackrel{a.s.}{\to} 0\)
+for each pair \((i, j)\) in different communities.
+
+\begin{algorithm}[tp]
+  \label{alg:osc}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Community assignments $1, ..., K$}
+    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most
+    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct
+    $V$ using these eigenvectors as its columns.\;
+    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
+    Construct graph $G$ using $B$ as its similarity matrix.\;
+    Partition $G$ into $K$ disconnected subgraphs
+    (e.g., using edge thresholding or spectral clustering).\;
+    Map each partition to the community labels $1, ..., K$.\;
+  \caption{Orthogonal Spectral Clustering.}
+\end{algorithm}
+
+\begin{theorem}
+\label{thm:osc}
+Assume the setting of theorem \ref{thm:osc-p}. 
+Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix
+obtained from OSC as described in algorithm~\ref{alg:osc}. 
+Then for $n \rho_n = \omega(\log^{4}{n})$, we have
+\begin{equation} 
+\label{eq:osc}
+\max_{i, j} |\hat{B}_{ij} - B_{ij}| = O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
+\end{equation}
+with high probability. In particular 
+$\hat{B}_{ij} -B_{ij} \overset{\mathrm{a.s.}}{\rightarrow} 0$ 
+where the convergence is
+uniform over all $i,j$. Hence for all pairs $(i,j)$ in different
+communities we have 
+$\hat{B}_{ij} \overset{\mathrm{a.s.}}{\rightarrow} 0$, 
+while for all pairs $(i, j)$ in the same community, 
+$\liminf_{n \to \infty} |\hat{B}_{ij}| > 0$ almost surely. 
+\end{theorem}
 
@@
 \hypertarget{generalized-random-dot-product-graphs-with-community-structure}{%
-\section{Generalized Random Dot Product Graphs with Community
-Structure}\label{generalized-random-dot-product-graphs-with-community-structure}}
+\section{Generalized Random Dot Product Graphs with Community Structure}\label{generalized-random-dot-product-graphs-with-community-structure}}
 
@@
 
-\section*{Appendix A}
+\section*{Appendix A: Proofs of Theorems}
+
+Let \(V_n\) and \(\hat{V}_n\)
+be the \(n \times K^2\) matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
+\(K^2\) largest eigenvalues (in modulus), respectively.
+We first state an important technical lemma for bounding the maximum
+\(\ell_2\) norm difference between the rows of \(\hat{V}_n\) and
+\(V_n\). See \citet{cape_biometrika} and
+\citet[Lemma~5]{rubindelanchy2017statistical} for a proof.
+
+\begin{lemma}
+\label{lem:technical}
+Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell \ell)}\}_{K})$ be a $K$-blocks
+PABM graph on $n$ vertices and let \(V\) and \(\hat{V}\)
+be the $n \times K^2$ matrices whose columns are the eigenvectors of 
+\(P\) and \(A\) corresponding to the
+$K^2$ largest eigenvalues in modulus, respectively.
+Let \(v_i^\top\) and \(\hat{v}_i^\top\) denote the $i$th 
+row of \(V\) and \(\hat{V}\), respectively. 
+Then there exists a constant $c > 1$ and an orthogonal matrix $W$ such
+that with high probability,
+$$\max_{i} \|W \hat{v}_i - v_i\|  = O\Big(\frac{\log^{c}n}{n \sqrt{\rho_n}} \Big).$$
+In particular we can take $c = 1 + \epsilon$ for any $\epsilon > 0$. 
+\end{lemma}
+
+\begin{proof}[Proof of theorem~\ref{thm:osc}]
+Recall the notations in lemma~\ref{lem:technical} and note that,
+under our assumption that the latent vectors $\lambda^{(k \ell)}$
+are all homogeneous, we have $\max_{i} \|v_i\| =
+O(n^{-1/2})$.
+
+Next recall theorem~\ref{thm:osc-p}; in particular $B_{ij} = nv_i^{\top}
+v_j$. 
+We therefore have
+\[\begin{split}
+ \max_{ij} |\hat{B}_{ij} - B_{ij}| &= \max_{ij} n |\hat{v}_i^\top \hat{v}_j -
+v_i^\top v_j| \\
+& \leq n \max_{ij} |\hat{v}_i^\top W W^\top \hat{v}_j -
+v_i^\top v_j| \\
+& \leq n \max_{i,j} \Bigl(\|W^{\top} \hat{v}_i - v_i\| \times \|\hat{v}_j\|
++ \|W^{\top} \hat{v}_j - v_j\| \times \|v_i\|\Bigr) \\
+& \leq n \Bigl(\max_{ij}  \|W \hat{v}_i  - v_i \|^2 +  \|W
+\hat{v}_i  - v_i \| \times \|v_j\| +  \|W \hat{v}_j  - v_j \| \times \|v_i\|\Bigr) 
+\\ &
+\leq n \max_{i} \|W \hat{v}_i  - v_i \|^2 + 2n \max_{i}
+\|W \hat{v}_i  - v_i \| \times \max_{j} \|v_j\|
+\\
+& = O \Big( \frac{\log^{c}{n}}{n^{1/2} \rho_n^{1/2}} \Big)
+\end{split}\]
+with high probability.
+Theorem~\ref{thm:osc} follows from the above bound together with the
+conclusion in theorem~\ref{thm:osc-p} that $B_{ij} = 0$ whenever vertices $i$ and $j$
+belongs to different communities. 
+\end{proof}
 

### 2022-08-23 draft/draft.Rmd
still writing

@@
         xtable.table.placement = 'H')
-
+```
+
+```{r}
 import::from(magrittr, `%>%`)
@@
 ```{=tex}
-\setlength{\parindent}{10.0pt}
+\setlength{\parindent}{20.0pt}
 \newcommand{\diag}{\mathrm{diag}}
@@
 
+\newpage
+
 # Introduction
@@
 However, there are inherent notions of distances among vectors, such as shortest path distance, expected commute time, or resistance distance. 
-One way to adapt $K$-means to graph data is to use an equivalent objective function as Eq.~\ref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
+One way to adapt $K$-means to graph data is to use an equivalent objective function as Eq.~\eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
 $$
@@
 $$
-\ell(\omega) = 
+\ell(\omega; A) = 
 \sum_{i < j} A_{ij} \log \bigg( \frac{\omega_i \omega_j}{d_{ij}^2 -
@@
 
-## Contributions of this work
+## Contributions of This Work
 
@@
 Within the context of this work, we will call such models *block models*. 
-This type of model frmaework allows us to frame the goal of community detection as a statistical inference problem: 
+This type of model framework allows us to frame the goal of community detection as a statistical inference problem: 
 identify the true community labels, up to permutation of the labels.
@@
 Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form. 
-As an example, we write the EM algorithm for the likelihood function in equation \ref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
+As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
 
@@
 \label{algo:sbm-em}
+  \setstretch{1.5}
   \DontPrintSemicolon
@@
 
-### Random Graph Models and Sparsity
+## Random Graph Models and Sparsity
 
@@
 This does not take into account that some individuals are more likely to form connections than others. 
+Modeling this as a DCBM allows us to take each individual's popularity into account.
 
@@
 
-### The Hierarchy of Block Models
+## The Hierarchy of Block Models
 
@@
 
-## Connecting the Stochastic Block Model and the Degree Corrected Block Model to the Generalized Random Dot Product Graph {#sec:sbm-dcbm-grdpg}
+## Connecting the SBM and DCBM to the GRDPG {#sec:sbm-dcbm-grdpg}
 
@@
 Let $X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top$ be noisy points sampled from $K$ subspaces, i.e., $x_i = y_i + z_i$ where the $y_i$ belongs to the union of $K$ subspaces and the $z_i$ are noise vectors.
-Let $\vartheta \geq 0$ be given and let $C$ and $B$ be constructed from the solutions of LASSO problems as described in equation~\ref{eq:ssc} with this given choice of $\vartheta$. 
+Let $\vartheta \geq 0$ be given and let $C$ and $B$ be constructed from the solutions of LASSO problems as described in equation~\eqref{eq:ssc} with this given choice of $\vartheta$. 
 Then $X$ is said to satisfy the subspace detection property with sparsity parameter $\vartheta$ if each column of $C$ has nonzero $\ell_2$ norm and $B_{ij} = 0$ whenever $y_i$ and $y_j$ are from different subspaces.
@@
 recall that $\vartheta$ is the sparsity parameter for
-the LASSO problems in equation~\ref{eq:ssc}. However, this is not sufficient to
+the LASSO problems in equation~\eqref{eq:ssc}. However, this is not sufficient to
 guarantee that SSC perfectly recovers the underlying subspaces.
@@
 and so the embedding of $A$ will lie approximately on the $K$ subspaces. 
-We will show in theorem \ref{theorem5} that, with probability converging to
+We will show in theorem \ref{thm:ssc} that, with probability converging to
 one as $n \rightarrow \infty$, the rows of the ASE of $A$ also
 satisfy the subspace detection property. 
-Theorem \ref{theorem5} builds upon existing work by
+Theorem \ref{thm:ssc} builds upon existing work by
 \citet{rubindelanchy2017statistical} who describe the convergence
@@
 regarding the application of SSC to the rows of $A$.
+
+### Algorithms for Community Detection
 
@@
   \label{alg:osc}
+  \setstretch{1.5}
   \DontPrintSemicolon
@@
   \KwData{Adjacency matrix $A$, number of communities $K$}
-  \KwResult{Community assignments $1, ..., K$}
+  \KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$}
     Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most
@@
 
+Theorem \ref{thm:osc} guarantees that for any $\epsilon > 0$, 
+the number of edges of $\hat{B}$ between vertices of different communities that are larger than $\epsilon$ converges to zero with probability converging to one as $n$ increases. 
+We can always find an $\epsilon > 0$ such that $\hat{B}_{ij} > \epsilon$ with probability converging to one as $n$ increases. 
+Thus, by using $\hat{B}$, we can perfectly recover all the latent community assignments $z_1, z_2, \dots, z_n$, i.e., the number of misclustered vertices is zero asymptotically almost surely. 
+We note that theorem \ref{thm:osc} is stronger than existing results in the literature; 
+in particular theorem 1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
+originally introduces the PABM model) only guarantees that the *proportion* of misclustered vertices converges to $0$ as $n \rightarrow \infty$. 
+Furthermore theorem 1 of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity parameter $\rho_n$ to satisfies $n \rho_n^2 = \omega(\log^2{n})$ which is a considerably stronger assumption than the assumption $n \rho_n =\omega(\log^{4}{n})$ used in theorem \ref{thm:osc}.
+Indeed, $n\rho_n^2 = \omega(\log^{2}{n})$ implies $n \rho_n = \omega(n^{1/2})$. 
+We emphasize that the assumption $n \rho_n = \omega(\log^{c}{n})$ for some constant $c > 1$ is commonly used in the context of graph estimation using spectral methods.
+
+Theorems \ref{thm:pabm-grdpg}, \ref{thm:osc-p}, and \ref{thm:osc} also provide
+a natural path toward using SSC for community detection. 
+In particular we established in theorem \ref{thm:pabm-grdpg} that an ASE of the edge
+probability matrix \(P\) can be constructed from a latent vector configuration 
+consisting of orthogonal subspaces. Theorem \ref{thm:osc-p} shows how 
+this property can also be recovered from the eigenvectors of \(P\). 
+Then theorem \ref{thm:osc} shows that, by replacing $P$ with $A$, the
+rows of $\hat{V}$ also lie on asymptotically orthogonal subspaces.
+Motivated by theorem \ref{thm:osc}, theorem \ref{thm:ssc} below
+shows that the subspace detection property also holds for the rows of
+$\sqrt{n} \hat{V}$.
+
+\begin{algorithm}[t]
+  \label{alg:ssc}
+  \setstretch{1.5}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \caption{Sparse Subspace Clustering using LASSO.}
+  \KwData{Adjacency matrix $A$, number of communities $K$,
+  hyperparameter $\lambda$}
+  \KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$}
+    Find $V$, the matrix of eigenvectors of $A$
+    corresponding to the $K (K + 1) / 2$ most positive
+    and the $K (K - 1) / 2$ most negative eigenvalues.\;
+    Normalize $V \leftarrow \sqrt{n} V$.\;
+    \For {$i = 1, ..., n$} {
+      Assign $v_i^\top$ as the $i^{th}$ row of $V$.
+      Assign $V_{-i} = \bigl[
+      v_1 \mid \cdots \mid v_{i-1} \mid v_{i+1} \mid \cdots \mid v_n \bigr]^\top$.\;
+      Solve the LASSO problem
+      $c_i = \arg\min_{\beta}
+      \frac{1}{2} \|v_i - V_{-i} \beta\|_2^2 + \lambda \|\beta\|_1$.\;
+      Assign $\tilde{c}_i = (c_i^{(1)}, \dots, c_i^{(i-1)}, 0, c_i^{(i)}, \dots, c_i^{(n-1)})^\top$ such that the superscript is the index of
+      $\tilde{c}_i$.\;
+    }
+    Assign
+    $C = \bigr[ \tilde{c}_1 \mid \cdots \mid \tilde{c}_n \bigr]$.\;
+    Compute the affinity matrix $B = |C| + |C^\top|$.\;
+    Construct graph $G$ using $B$ as its similarity matrix.\;
+    Partition $G$ into $K$ disconnected subgraphs (e.g., using edge
+    thresholding or spectral clustering).\;
+    Map each partition to the community labels $1, ..., K$.
+\end{algorithm}
+
+\begin{theorem}
+\label{thm:ssc}
+Let $P$ describe the edge probability matrix of the PABM with $n$ vertices, and let $A \sim \Bernoulli(P)$. 
+Let $\hat{V}$ be the matrix of eigenvectors of $A$ corresponding to the $K^2$ largest eigenvalues in modulus. 
+Then for any $\epsilon > 0$, there exists a choice of $\vartheta > 0$ and $N \in \mathbb{N}$ such that for all $n \geq N$, $\sqrt{n} \hat{V}$ obeys the subspace detection property with probability at least $1 - \epsilon$.
+\end{theorem}
+
+### Algorithm for Parameter Estimation
+
+For ease of exposition we now assume in this subsection that the edge probability matrix \(P\) for the PABM had been arranged so that the rows and columns are organized by community so that $\tilde{P} = P$ (see remark \ref{rem:pabm_view2}).
+Then the \(k\ell\)\textsuperscript{th} block is an outer product of two vectors, i.e., \(P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top\). 
+Therefore, given \(P^{(k \ell)}\), \(\lambda^{(k \ell)}\) and \(\lambda^{(\ell k)}\) are solvable up to multiplicative constant using singular value decomposition. 
+More specifically, let \(P^{(k \ell)} = (\sigma^{(k \ell)})^2 u^{(k \ell)} (v^{(k \ell)})^\top\) be the singular value decomposition of \(P^{(k \ell)}\) where \(u^{(k \ell)} \in \mathbb{R}^{n_k}\) and \(v^{(k \ell)} \in \mathbb{R}^{n_\ell}\) are vectors and \(\sigma^{(k \ell)}\) is a scalar.
+Then \(\rho_n^{1/2} \lambda^{(k \ell)} = s_1 u^{(k \ell)}\) and \(\rho_n^{1/2} \lambda^{(\ell k)} = s_2 v^{(k \ell)}\) for unidentifiable $s_1 s_2 = (\sigma^{(k \ell)})^2$.
+Because each $\lambda^{(k \ell)}$ is not strictly identifiable, we instead estimate each $\tilde{\lambda}^{(k \ell)} = \sigma^{(k \ell)} u^{(k \ell)}$. 
+Given the adjacency matrix \(A\) instead of edge probability matrix \(P\), we can simply use plug-in estimators by taking the SVD of each $A^{(k \ell)}$ to obtain $\hat{\lambda}^{(k \ell)} = \hat{\sigma}^{(k \ell)} \hat{u}^{(k \ell)}$ using the largest singular value of $A$ and its corresponding singular vectors.
+
+\begin{algorithm}[tp]
+  \label{alg:param_est}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \setstretch{1.5}
+  \caption{PABM parameter estimation.}
+  \KwData{Adjacency matrix $A$, community assignments $1, ..., K$}
+  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(k \ell)}\}_K$.}
+  Arrange the rows and columns of $A$ by community such that each 
+  $A^{(k \ell)}$ block consists of estimated edge probabilities between 
+  communities $k$ and $l$.\;
+  \For {$k, \ell = 1, ..., K$, $k \leq \ell$} {
+    Compute $A^{(k \ell)} = U \Sigma V^\top$, the SVD of the $k\ell^{th}$ block.\;
+    Assign $u^{(k \ell)}$ and $v^{(k \ell)}$ as the first columns of $U$ and $V$. 
+    Assign $(\sigma^{(k \ell)})^2 \leftarrow \Sigma_{11}$.\;
+    Assign $\hat{\lambda}^{(k \ell)} \leftarrow \pm \sigma^{(k \ell)} u^{(k \ell)}$ and $\hat{\lambda}^{(\ell k)} \leftarrow \pm \sigma^{(k \ell)} v^{(k \ell)}$.
+  }
+\end{algorithm}
+
+\begin{theorem}
+\label{thm:param-est}
+Let each $\tilde{\lambda}^{(k \ell)}$ be the popularity vector derived from its corresponding $P^{(k \ell)}$ and let $\hat{\lambda}^{(k \ell)}$ be its estimate obtained from $A^{(k \ell)}$ using algorithm~\ref{alg:param_est}.
+Then if $n \rho_n = \omega( \log^{4}{n})$,
+\begin{equation} 
+\label{eq:param-est}
+\max_{k, \ell \in \{1, ..., K\}} 
+\|\hat{\lambda}^{(k \ell)} - \tilde{\lambda}^{(k \ell)}\|_{\infty} = 
+O\bigg(\frac{\log n_k}{\sqrt{n_k}} \bigg)
+\end{equation}
+with high probability. 
+Here $\|\cdot\|_{\infty}$ denotes the $\ell_\infty$ norm of a vector. Let $\hat{\Lambda}$ be the matrix
+$$\hat{\Lambda} = \begin{bmatrix} \hat{\lambda}^{(11)} &
+  \hat{\lambda}^{(12)} & \cdots & \hat{\lambda}^{(1 K)} \\
+  \hat{\lambda}^{(21)} &
+  \hat{\lambda}^{(22)} & \cdots & \hat{\lambda}^{(2 K)} \\
+  \vdots & \vdots & \cdots & \vdots \\
+   \hat{\lambda}^{(K1)} &
+  \hat{\lambda}^{(K2)} & \cdots & \hat{\lambda}^{(K K)} 
+\end{bmatrix}$$
+and let $\hat{P} = \hat{X} U I_{p,q} U^\top \hat{X}^{\top}$, where $\hat{X}$ is defined from $\hat{\Lambda}$ and $U$ is defined from $K$ as in theorem~\ref{thm:pabm-grdpg}. 
+Equation~\eqref{eq:param-est} then implies
+\begin{equation}
+\label{eq:conv_prob}
+\frac{1}{n} \|\rho_n^{-1} \hat{P} - \rho_n^{-1} P\|_{F} = O((n
+\rho_n)^{-1/2}), \quad \max_{ij} |\rho_n^{-1} \hat{P}_{ij} -
+\rho_n^{-1} P_{ij}| = O((n \rho_n)^{-1/2})
+\end{equation}
+with high probability. 
+\end{theorem}
+
+Equation \eqref{eq:param-est} guarantees that $n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} - \Lambda\|_{F} = O((n \rho_n)^{-1/2})$. 
+Equation \eqref{eq:conv_prob} then guarantees that the mean square error for $\rho_n^{-1} (\hat{P} - P)$ converges to $0$ almost surely and furthermore the entries of $\rho_n^{-1} \hat{P}$ converge uniformly to the entries of $\rho_n^{-1} P$;
+recall that $\rho_n^{-1} P_{ij} = \lambda_{iz_j} \lambda_{j z_i}$. 
+We note that these results are stronger than existing results in \citet{307cbeb9b1be48299388437423d94bf1}; 
+for example theorem 2 in \citet{307cbeb9b1be48299388437423d94bf1} only guarantees 
+$n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} - \Lambda\|_{F} = o(1)$ as $n \rightarrow \infty$.
+
+## Simulation Study
+
+## Applications
+
 \newpage
 
-# Generalized Random Dot Product Graphs with Community Structure
+# Generalized Random Dot Product Graphs with Nonlinear Community Structure
+
+## The Manifold Block Model
+
+## Algorithms for Nonintersecting Manifolds
+
+## Algorithms for Intersecting Manifolds
+
+## Examples and Simulations
+
+## Applications
+
+## Conclusions
 
@@
 
+We now provide a proof of theorem \ref{thm:ssc}.
+Our proof is based on verifying the sufficient conditions given in theorem 6 of \citet{jmlr-v28-wang13} under which sparse subspace clustering based on solving the optimization problem in equation \eqref{eq:ssc} yields an affinity matrix $B = |C| + |C^{\top}|$ satisfying the subspace detection property of definition \ref{def:subspace_detection}.
+We first recall a few definitions used in \citet{soltanolkotabi2012} and \citet{jmlr-v28-wang13}; 
+for ease of exposition, these definitions are stated using the notations of the current paper and we will drop the explicit dependency on $n$ from our eigenvectors $\hat{V}$ of $A$ and $V$ of $P$.
+
+\begin{definition}[Inradius]
+\label{def:inradius}
+The inradius of a convex body $\mathcal{P}$, denoted by $r(\mathcal{P})$, is
+defined as the radius of the largest Euclidean ball inscribed in $\mathcal{P}$.
+Let $X$ be a $n \times d$ matrix with rows $x_1, x_2, \dots,
+x_n$. We then define, with a slight abuse of notation, $r(X)$ as the
+inradius of the convex hull formed by $\{\pm x_1, \pm x_2, \dots, \pm x_n\}$. 
+\end{definition}
+
+\begin{definition}[Subspace incoherence]
+\label{def:subspace_incoherence}
+Let $\hat{V}$ be the eigenvectors of $A$
+corresponding to the $K^2$ largest eigenvalues in modulus. Let
+$\hat{V}^{(k)}$ denote the matrix formed by keeping only the rows of
+$\hat{V}$ corresponding to the $k^{th}$
+community and let $\hat{V}^{(-k)}$ denote the matrix formed by
+omitting the rows of $\hat{V}$
+corresponding to the $k^{th}$ community. Let $(\hat{v}_i^{(k)})^\top$ denote
+the $i$th row of $\hat{V}^{(k)}$ and $\hat{V}_{-i}^{(k)}$ be $\hat{V}^{(k)}$ with
+the $i^{th}$ row omitted. Let $V$, $V^{(k)}$, $V^{(-k)}$, and
+$v_i^{(k)}$ be defined similarly using the eigenvectors $V$ of
+$P$. Finally let $\mathcal{S}^{(k)}$ be the vector space spanned by the
+rows of $V^{(k)}$.
+
+Now define $\nu_{i}^{(k)}$ for $k = 1,2,\dots,K$ and $i =
+1,2,\dots,n_{k}$ as the solution of the following optimization problem
+$$\nu_{i}^{(k)} = \max_\eta (\hat{v}_i^{(k)})^\top \eta - \frac{1}{2
+  \lambda} \eta^\top \eta, \quad \text{subject to $\|V_{-i}^{(k)}
+  \eta\|_\infty \leq 1$.}$$
+Given $\nu_i^{(k)}$, let $\mathbb{P}_{\mathcal{S}^{(k)}}(\nu_i^{(k)})$
+be the vector in $\mathbb{R}^{K^2}$ corresponding to the orthogonal projection of $\nu_i^{(k)}$ onto the vector space
+$\mathcal{S}^{(k)}$ and define the projected dual direction $w_{i}^{(k)}$
+as
+$$w_i^{(k)} =
+\frac{\mathbb{P}_{\mathcal{S}^{(k)}}(\nu_i^{(k)})}{\|\mathbb{P}_{\mathcal{S}^{(k)}}(\nu_i^{(k)})\|}.$$
+Now let $W^{(k)} = \bigl[ w_1^{(k)} \mid \cdots \mid w_{n_k}^{(k)} \bigr]^\top$
+and define the subspace incoherence for $\hat{V}^{(k)}$ by
+$$\mu^{(k)} = \mu(\hat{V}^{(k)}) = \max\limits_{v \in V^{(-k)}} \|W^{(k)} v\|_\infty.$$
+\end{definition}
+
+\begin{proof}[Proof of theorem \ref{thm:ssc}]
+For a given $k = 1,2\dots,K$, let $r^{(k)} = \min_{i}r(V_{-i}^{(k)})$ be inradius of the convex hull formed by
+the rows of $V_{-i}^{(k)}$ and let $r_* = \min_{k} r^{(k)}$. Then Theorem~6 in
+\citet{jmlr-v28-wang13} states that there exists a $\lambda > 0$
+such that $\sqrt{n} \hat{V}$ satisfies the subspace detection property in definition \ref{def:subspace_detection} whenever the following two conditions are satisfied: 
+\begin{gather}
+  \label{eq:cond1}
+  \mu^{(k)} < r^{(k)} \quad \text{for all $k = 1,2,\dots,K$}, \\
+  \label{eq:cond2}
+  \max_{i} \|W \hat{v}_{i} - v_{i}\| \leq \min_{k} \frac{r_*(r^{(k)} -
+    \mu^{(k)})}{2 + 7 r^{(k)}}.
+\end{gather}
+We now verify that for sufficiently large n, equation \eqref{eq:cond1} and Eq.~\eqref{eq:cond2}
+holds with high probability.
+
+{\bf Verifying equation \eqref{eq:cond1}}. If $n$ is sufficiently large then
+there are enough vertices in each community $k$ so that
+$\mathrm{span}(V_{-i}^{(k)}) = \mathcal{S}^{(k)}$ for all $i$ and hence
+\(r^{(k)} = \min_{i} r(V_{-i}^{(k)}) > 0\) for
+all $k = 1,2,\dots,K$.
+
+Next, by theorem \ref{thm:osc-p} we have that the subspaces
+$\{\mathcal{S}^{(1)}, \dots, \mathcal{S}^{(K)}\}$
+are mutually orthogonal, i.e., $v^{\top} w = 0$ for all $v \in
+\mathcal{S}^{(k)}$ and $w \in \mathcal{S}^{(\ell)}$ with $k \not =
+\ell$. Now let $z \in \mathbb{R}^{K^2}$ be arbitrary and let
+$\tilde{z} = \mathbb{P}_{\mathcal{S}^{(k)}} z$ be the projection of
+$z$ onto $\mathcal{S}^{(k)}$. We then have $v^{\top} \tilde{z} =
+0$ for all $v \in V^{(-k)}$. Because $z$ is arbitrary, this implies 
+$\|W^{(k)} v\|_{\infty} = 0$ for all $v
+\in V^{(-k)}$ and hence $\mu^{(k)} = 0$ for all $k
+=1,2,\dots,K$. Therefore $\mu^{(k)} < r^{(k)}$ for all $k =
+1=2,\dots,K$ as desired.
+
+{\bf Verifying equation \eqref{eq:cond2}}.
+Let $\delta = \max_{i} \sqrt{n} \|W \hat{v}_{i} - v_{i}\|$. 
+Then from lemma \ref{lem:technical}, we have \(\delta \stackrel{a.s.}{\to} 0\) and hence
+$$\delta < \min_{k} \frac{r_* (r^{(k)} - \mu^{(k)})}{2 + 7 r^{(k)}}$$
+asymptotically almost surely.
+
+In summary $\sqrt{n} \hat{V}$ satisfies the subspace detection property with probability converging to $1$ as \(n \to \infty\).
+\end{proof}
+
+\begin{remark}
+Theorem 6 of \citet{jmlr-v28-wang13} assumes that each row $v_i$
+of $V$ has unit norm, i.e., $\|v_{i}\| = 1$  for all $i$. 
+This assumption has the effect of scaling the $r^{(k)}$ so that $r^{(k)} \leq 1$ for all $k = 1,2,\dots,K$. 
+We emphasize that this assumption has no effect on the proof of Theorem \ref{thm:ssc}. 
+Indeed, because $\mu^{(k)} = 0$ for all $k$, as long as the rows of $V^{(k)}$ spans the subspace $\mathcal{S}^{(k)}$, then $a r^{(k)} > \mu^{(k)}$ for any scalar $a > 0$. 
+\end{remark}
+
+\begin{proof}[Proof of Theorem \ref{thm:param-est}] 
+Let \(P\) be organized by community such that \(P^{(k \ell)}\) denote the $n_k \times n_{\ell}$ matrix obtained by keeping only the rows of $P$ corresponding to vertices in community $k$ and the columns of $P$ corresponding to vertices in community $\ell$.
+We define $A^{(k \ell)}$ analogously. 
+Recall that $P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^{\top}$ for all $k, \ell$. 
+We now consider estimation of $P^{(k \ell)}$ for the cases when $k = \ell$ versus when $k \not = \ell$.
+
+\emph{Case \(k = l\)}. 
+Let $P^{(kk)} = \sigma_{kk}^2 u^{(kk)} (u^{(kk)})^\top$ be the singular value decomposition of $P^{(kk)}$. 
+We can then define $\tilde{\lambda}^{(kk)} = \sigma_{kk} u^{(kk)}$. 
+Now let $\hat{U}^{(kk)} \hat{\Sigma}^{(kk)} (\hat{U}^{(kk)})^\top$ be the singular value decomposition of \(A^{(kk)}\), and let $\hat{\sigma}_{kk}^2 \hat{u}^{(kk)} (\hat{u}^{(kk)})^\top$ be the best rank-one approximation of $A^{(kk)}$.
+Define \(\hat{\lambda}^{(kk)} = \hat{\sigma}_{kk} \hat{u}^{(kk)}\). 
+Then \(\hat{\lambda}^{(kk)}\) is the adjacency spectral embedding approximation of \(\lambda^{(kk)}\), and by Theorem 5 of \citet{rubindelanchy2017statistical}, we have
+$$\|\hat{\lambda}^{(kk)} - \lambda^{(kk)}\|_{\infty} = O\Bigl(\frac{\log n_k}{\sqrt{n_k}}\Bigr)$$
+with high probability. 
+Here $\|\cdot\|_{\infty}$ denote the $\ell_{\infty}$ norm of a vector.
+
+\emph{Case \(k \neq l\)}. 
+Let \(P^{(k \ell)} = \sigma_{k \ell}^2 u^{(k \ell)} (v^{(k \ell)})^\top\) and \(P^{(\ell k)} = \sigma_{kl}^2 u^{(\ell k)} (v^{(\ell k)})^\top\) be the singular value decompositions and note that \(\sigma_{k \ell} = \sigma_{\ell k}\), \(u^{(k \ell)} = v^{(\ell k)}\), and \(v^{(k \ell)} = u^{(\ell k)}\).
+Now define \(\lambda^{(k \ell)} = \sigma_{k \ell} u^{(k \ell)}\) and \(\lambda^{(\ell k)} = \sigma_{k \ell} v^{(k \ell)}\).
+
+Next consider the Hermitian dilation
+\[
+M^{(k \ell)} = 2 \begin{bmatrix} 0 & P^{(k \ell)} \\ P^{(\ell k)} & 0 \end{bmatrix}
+\]
+which is a symmetric \((n_k + n_\ell) \times (n_k + n_\ell)\) matrix. 
+The eigendecomposition of \(M^{(k \ell)}\) is then
+
+\[M^{(k \ell)} = 
+\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix} \times 
+\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
+\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix}^\top\]
+Thus treating \(M^{(k \ell)}\) as the edge probability matrix of a GRDPG, we have latent positions in \(\mathbb{R}^2\) given by the $(n_k + n_{\ell}) \times 2$ matrix
+\[\Lambda^{(k \ell)} = \begin{bmatrix} 
+  \sigma_{k \ell} u^{(k \ell)} & \sigma_{k \ell} u^{(k \ell)} \\ 
+  \sigma_{k \ell} v^{(k \ell)} & -\sigma_{k \ell} v^{(k \ell)} 
+\end{bmatrix} = 
+\begin{bmatrix} 
+  \lambda^{(k \ell)} & \lambda^{(k \ell)} \\ 
+  \lambda^{( \ell k)} & -\lambda^{( \ell k)} 
+\end{bmatrix}.\]
+Now consider
+\[\hat{M}^{(k \ell)} = \begin{bmatrix} 0 & A^{(k \ell)} \\ A^{(\ell k)} & 0 \end{bmatrix}\]
+
+We can then view \(\hat{M}^{(k \ell)}\) as an adjacency matrix drawn from
+the edge probabilities matrix \(M^{(k \ell)}\). Now suppose that the adjacency spectral
+embedding of $\hat{M}^{(k \ell)}$ is represented as the $(n_k +
+n_{\ell}) \times 2$ matrix
+\[\hat{\Lambda}^{(k \ell)} = \begin{bmatrix} 
+  \hat{\lambda}^{(k \ell)} & \hat{\lambda}^{(k \ell)} \\ 
+  \hat{\lambda}^{(\ell k)} & -\hat{\lambda}^{(\ell k)} 
+\end{bmatrix}\]
+where each \(\hat{\lambda}^{(k \ell)}\) is defined as in Algorithm 3. Then
+by Theorem 5 of \citet{rubindelanchy2017statistical}, there
+exists an indefinite orthogonal transformation $W^{*}$ such that,  with
+high probability,
+$$\max_{i} |W^{*} \hat{\Lambda}_{i}^{(k \ell)} - \Lambda_{i}^{(k \ell)} \| =
+O\Bigl(\frac{\log (n_k + n_{\ell})}{\sqrt{n_k + n_{\ell}}}\Bigr)$$
+with high probability. Here $\Lambda_{i}^{(k \ell)}$ and
+$\hat{\Lambda}_i^{(k \ell)}$ denote the $i$th rows of $\Lambda^{(k
+  \ell)}$ and $\hat{\Lambda}^{(k \ell)}$, respectively.
+
+Furthermore, by looking at the proof of Theorem~5 in
+\citep{rubindelanchy2017statistical}, we see that $W^{*}$ is also
+blocks diagonal with $2$ blocks where the positive eigenvalues of $M^{(k \ell)}$
+forming a block and the negative eigenvalues of $M^{(k \ell)}$ forming
+the remaining block. 
+Because $M^{(k \ell)}$ has one positive eigenvalue and one negative
+eigenvalue, we see that $W^{*}$ is necessarily of the form $W^{*}
+= \Bigl[\begin{smallmatrix} 1 & 0 \\ 0 & - 1\end{smallmatrix}\Bigr]$
+Using this form for $W^{*}$, we obtain
+$$\max\{\|\hat{\lambda}^{(k \ell)} - \lambda^{(k \ell)}\|_{\infty},
+\|\hat{\lambda}^{(\ell k)} - \lambda^{(\ell k)}\|_{\infty}\}  =
+O\Bigl(\frac{\log(n_k + n_{\ell})}{\sqrt{n_k + n_{\ell}}}\Bigr)$$
+with high probability. Combining this bound with the bound for
+$\|\hat{\lambda}^{(kk)} - \lambda^{(kk)}\|_{\infty}$ given above
+yields equation~\eqref{eq:param-est} in theorem~\ref{thm:param-est}. 
+\end{proof}
+
 \newpage

### 2022-08-23 draft/draft.tex
still writing

@@
 \newpage
-\tableofcontents
+
+% \tableofcontents
+
 \addcontentsline{toc}{section}{\contentsname}
@@
 }
-\setlength{\parindent}{10.0pt}
+\setlength{\parindent}{20.0pt}
 \newcommand{\diag}{\mathrm{diag}}
@@
 
+\newpage
+
 \hypertarget{introduction}{%
@@
 \hypertarget{contributions-of-this-work}{%
-\subsection{Contributions of this work}\label{contributions-of-this-work}}
+\subsection{Contributions of This Work}\label{contributions-of-this-work}}
 
@@
 Within the context of this work, we will call such models \emph{block models}.
-This type of model frmaework allows us to frame the goal of community detection as a statistical inference problem:
+This type of model framework allows us to frame the goal of community detection as a statistical inference problem:
 identify the true community labels, up to permutation of the labels.
@@
 Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form.
-As an example, we write the EM algorithm for the likelihood function in equation \ref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
+As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
 
@@
 \hypertarget{random-graph-models-and-sparsity}{%
-\subsubsection{Random Graph Models and Sparsity}\label{random-graph-models-and-sparsity}}
+\subsection{Random Graph Models and Sparsity}\label{random-graph-models-and-sparsity}}
 
@@
 \hypertarget{the-hierarchy-of-block-models}{%
-\subsubsection{The Hierarchy of Block Models}\label{the-hierarchy-of-block-models}}
+\subsection{The Hierarchy of Block Models}\label{the-hierarchy-of-block-models}}
 
@@
 \hypertarget{sec:sbm-dcbm-grdpg}{%
-\subsection{Connecting the Stochastic Block Model and the Degree Corrected Block Model to the Generalized Random Dot Product Graph}\label{sec:sbm-dcbm-grdpg}}
+\subsection{Connecting the SBM and DCBM to the GRDPG}\label{sec:sbm-dcbm-grdpg}}
 
@@
 and so the embedding of \(A\) will lie approximately on the \(K\) subspaces.
-We will show in theorem \ref{theorem5} that, with probability converging to
+We will show in theorem \ref{thm:ssc} that, with probability converging to
 one as \(n \rightarrow \infty\), the rows of the ASE of \(A\) also
 satisfy the subspace detection property.
-Theorem \ref{theorem5} builds upon existing work by
+Theorem \ref{thm:ssc} builds upon existing work by
 \citet{rubindelanchy2017statistical} who describe the convergence
@@
 regarding the application of SSC to the rows of \(A\).
+
+\hypertarget{algorithms-for-community-detection}{%
+\subsubsection{Algorithms for Community Detection}\label{algorithms-for-community-detection}}
 
@@
 
+Theorem \ref{thm:osc} guarantees that for any \(\epsilon > 0\),
+the number of edges of \(\hat{B}\) between vertices of different communities that are larger than \(\epsilon\) converges to zero with probability converging to one as \(n\) increases.
+We can always find an \(\epsilon > 0\) such that \(\hat{B}_{ij} > \epsilon\) with probability converging to one as \(n\) increases.
+Thus, by using \(\hat{B}\), we can perfectly recover all the latent community assignments \(z_1, z_2, \dots, z_n\), i.e., the number of misclustered vertices is zero asymptotically almost surely.
+We note that theorem \ref{thm:osc} is stronger than existing results in the literature;
+in particular theorem 1 of \citet{307cbeb9b1be48299388437423d94bf1} (the paper that
+originally introduces the PABM model) only guarantees that the \emph{proportion} of misclustered vertices converges to \(0\) as \(n \rightarrow \infty\).
+Furthermore theorem 1 of \citet{307cbeb9b1be48299388437423d94bf1} also requires the sparsity parameter \(\rho_n\) to satisfies \(n \rho_n^2 = \omega(\log^2{n})\) which is a considerably stronger assumption than the assumption \(n \rho_n =\omega(\log^{4}{n})\) used in theorem \ref{thm:osc}.
+Indeed, \(n\rho_n^2 = \omega(\log^{2}{n})\) implies \(n \rho_n = \omega(n^{1/2})\).
+We emphasize that the assumption \(n \rho_n = \omega(\log^{c}{n})\) for some constant \(c > 1\) is commonly used in the context of graph estimation using spectral methods.
+
+Theorems \ref{thm:pabm-grdpg}, \ref{thm:osc-p}, and \ref{thm:osc} also provide
+a natural path toward using SSC for community detection.
+In particular we established in theorem \ref{thm:pabm-grdpg} that an ASE of the edge
+probability matrix \(P\) can be constructed from a latent vector configuration
+consisting of orthogonal subspaces. Theorem \ref{thm:osc-p} shows how
+this property can also be recovered from the eigenvectors of \(P\).
+Then theorem \ref{thm:osc} shows that, by replacing \(P\) with \(A\), the
+rows of \(\hat{V}\) also lie on asymptotically orthogonal subspaces.
+Motivated by theorem \ref{thm:osc}, theorem \ref{thm:ssc} below
+shows that the subspace detection property also holds for the rows of
+\(\sqrt{n} \hat{V}\).
+
+\begin{algorithm}[t]
+  \label{alg:ssc}
+  \setstretch{1.5}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \caption{Sparse Subspace Clustering using LASSO.}
+  \KwData{Adjacency matrix $A$, number of communities $K$,
+  hyperparameter $\lambda$}
+  \KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$}
+    Find $V$, the matrix of eigenvectors of $A$
+    corresponding to the $K (K + 1) / 2$ most positive
+    and the $K (K - 1) / 2$ most negative eigenvalues.\;
+    Normalize $V \leftarrow \sqrt{n} V$.\;
+    \For {$i = 1, ..., n$} {
+      Assign $v_i^\top$ as the $i^{th}$ row of $V$.
+      Assign $V_{-i} = \bigl[
+      v_1 \mid \cdots \mid v_{i-1} \mid v_{i+1} \mid \cdots \mid v_n \bigr]^\top$.\;
+      Solve the LASSO problem
+      $c_i = \arg\min_{\beta}
+      \frac{1}{2} \|v_i - V_{-i} \beta\|_2^2 + \lambda \|\beta\|_1$.\;
+      Assign $\tilde{c}_i = (c_i^{(1)}, \dots, c_i^{(i-1)}, 0, c_i^{(i)}, \dots, c_i^{(n-1)})^\top$ such that the superscript is the index of
+      $\tilde{c}_i$.\;
+    }
+    Assign
+    $C = \bigr[ \tilde{c}_1 \mid \cdots \mid \tilde{c}_n \bigr]$.\;
+    Compute the affinity matrix $B = |C| + |C^\top|$.\;
+    Construct graph $G$ using $B$ as its similarity matrix.\;
+    Partition $G$ into $K$ disconnected subgraphs (e.g., using edge
+    thresholding or spectral clustering).\;
+    Map each partition to the community labels $1, ..., K$.
+\end{algorithm}
+
+\begin{theorem}
+\label{thm:ssc}
+Let $P$ describe the edge probability matrix of the PABM with $n$ vertices, and let $A \sim \Bernoulli(P)$. 
+Let $\hat{V}$ be the matrix of eigenvectors of $A$ corresponding to the $K^2$ largest eigenvalues in modulus. 
+Then for any $\epsilon > 0$, there exists a choice of $\vartheta > 0$ and $N \in \mathbb{N}$ such that for all $n \geq N$, $\sqrt{n} \hat{V}$ obeys the subspace detection property with probability at least $1 - \epsilon$.
+\end{theorem}
+
+\hypertarget{algorithm-for-parameter-estimation}{%
+\subsubsection{Algorithm for Parameter Estimation}\label{algorithm-for-parameter-estimation}}
+
+For ease of exposition we now assume in this subsection that the edge probability matrix \(P\) for the PABM had been arranged so that the rows and columns are organized by community so that \(\tilde{P} = P\) (see remark \ref{rem:pabm_view2}).
+Then the \(k\ell\)\textsuperscript{th} block is an outer product of two vectors, i.e., \(P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top\).
+Therefore, given \(P^{(k \ell)}\), \(\lambda^{(k \ell)}\) and \(\lambda^{(\ell k)}\) are solvable up to multiplicative constant using singular value decomposition.
+More specifically, let \(P^{(k \ell)} = (\sigma^{(k \ell)})^2 u^{(k \ell)} (v^{(k \ell)})^\top\) be the singular value decomposition of \(P^{(k \ell)}\) where \(u^{(k \ell)} \in \mathbb{R}^{n_k}\) and \(v^{(k \ell)} \in \mathbb{R}^{n_\ell}\) are vectors and \(\sigma^{(k \ell)}\) is a scalar.
+Then \(\rho_n^{1/2} \lambda^{(k \ell)} = s_1 u^{(k \ell)}\) and \(\rho_n^{1/2} \lambda^{(\ell k)} = s_2 v^{(k \ell)}\) for unidentifiable \(s_1 s_2 = (\sigma^{(k \ell)})^2\).
+Because each \(\lambda^{(k \ell)}\) is not strictly identifiable, we instead estimate each \(\tilde{\lambda}^{(k \ell)} = \sigma^{(k \ell)} u^{(k \ell)}\).
+Given the adjacency matrix \(A\) instead of edge probability matrix \(P\), we can simply use plug-in estimators by taking the SVD of each \(A^{(k \ell)}\) to obtain \(\hat{\lambda}^{(k \ell)} = \hat{\sigma}^{(k \ell)} \hat{u}^{(k \ell)}\) using the largest singular value of \(A\) and its corresponding singular vectors.
+
+\begin{algorithm}[tp]
+  \label{alg:param_est}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \setstretch{1.5}
+  \caption{PABM parameter estimation.}
+  \KwData{Adjacency matrix $A$, community assignments $1, ..., K$}
+  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(k \ell)}\}_K$.}
+  Arrange the rows and columns of $A$ by community such that each 
+  $A^{(k \ell)}$ block consists of estimated edge probabilities between 
+  communities $k$ and $l$.\;
+  \For {$k, \ell = 1, ..., K$, $k \leq \ell$} {
+    Compute $A^{(k \ell)} = U \Sigma V^\top$, the SVD of the $k\ell^{th}$ block.\;
+    Assign $u^{(k \ell)}$ and $v^{(k \ell)}$ as the first columns of $U$ and $V$. 
+    Assign $(\sigma^{(k \ell)})^2 \leftarrow \Sigma_{11}$.\;
+    Assign $\hat{\lambda}^{(k \ell)} \leftarrow \pm \sigma^{(k \ell)} u^{(k \ell)}$ and $\hat{\lambda}^{(\ell k)} \leftarrow \pm \sigma^{(k \ell)} v^{(k \ell)}$.
+  }
+\end{algorithm}
+
+\begin{theorem}
+\label{thm:param-est}
+Let each $\tilde{\lambda}^{(k \ell)}$ be the popularity vector derived from its corresponding $P^{(k \ell)}$ and let $\hat{\lambda}^{(k \ell)}$ be its estimate obtained from $A^{(k \ell)}$ using algorithm~\ref{alg:param_est}.
+Then if $n \rho_n = \omega( \log^{4}{n})$,
+\begin{equation} 
+\label{eq:param-est}
+\max_{k, \ell \in \{1, ..., K\}} 
+\|\hat{\lambda}^{(k \ell)} - \tilde{\lambda}^{(k \ell)}\|_{\infty} = 
+O\bigg(\frac{\log n_k}{\sqrt{n_k}} \bigg)
+\end{equation}
+with high probability. 
+Here $\|\cdot\|_{\infty}$ denotes the $\ell_\infty$ norm of a vector. Let $\hat{\Lambda}$ be the matrix
+$$\hat{\Lambda} = \begin{bmatrix} \hat{\lambda}^{(11)} &
+  \hat{\lambda}^{(12)} & \cdots & \hat{\lambda}^{(1 K)} \\
+  \hat{\lambda}^{(21)} &
+  \hat{\lambda}^{(22)} & \cdots & \hat{\lambda}^{(2 K)} \\
+  \vdots & \vdots & \cdots & \vdots \\
+   \hat{\lambda}^{(K1)} &
+  \hat{\lambda}^{(K2)} & \cdots & \hat{\lambda}^{(K K)} 
+\end{bmatrix}$$
+and let $\hat{P} = \hat{X} U I_{p,q} U^\top \hat{X}^{\top}$, where $\hat{X}$ is defined from $\hat{\Lambda}$ and $U$ is defined from $K$ as in theorem~\ref{thm:pabm-grdpg}. 
+Equation~\eqref{eq:param-est} then implies
+\begin{equation}
+\label{eq:conv_prob}
+\frac{1}{n} \|\rho_n^{-1} \hat{P} - \rho_n^{-1} P\|_{F} = O((n
+\rho_n)^{-1/2}), \quad \max_{ij} |\rho_n^{-1} \hat{P}_{ij} -
+\rho_n^{-1} P_{ij}| = O((n \rho_n)^{-1/2})
+\end{equation}
+with high probability. 
+\end{theorem}
+
+Equation \eqref{eq:param-est} guarantees that \(n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} - \Lambda\|_{F} = O((n \rho_n)^{-1/2})\).
+Equation \eqref{eq:conv_prob} then guarantees that the mean square error for \(\rho_n^{-1} (\hat{P} - P)\) converges to \(0\) almost surely and furthermore the entries of \(\rho_n^{-1} \hat{P}\) converge uniformly to the entries of \(\rho_n^{-1} P\);
+recall that \(\rho_n^{-1} P_{ij} = \lambda_{iz_j} \lambda_{j z_i}\).
+We note that these results are stronger than existing results in \citet{307cbeb9b1be48299388437423d94bf1};
+for example theorem 2 in \citet{307cbeb9b1be48299388437423d94bf1} only guarantees
+\(n^{-1/2} \|\rho_n^{-1/2} \hat{\Lambda} - \Lambda\|_{F} = o(1)\) as \(n \rightarrow \infty\).
+
+\hypertarget{simulation-study}{%
+\subsection{Simulation Study}\label{simulation-study}}
+
+\hypertarget{applications}{%
+\subsection{Applications}\label{applications}}
+
 \newpage
 
-\hypertarget{generalized-random-dot-product-graphs-with-community-structure}{%
-\section{Generalized Random Dot Product Graphs with Community Structure}\label{generalized-random-dot-product-graphs-with-community-structure}}
+\hypertarget{generalized-random-dot-product-graphs-with-nonlinear-community-structure}{%
+\section{Generalized Random Dot Product Graphs with Nonlinear Community Structure}\label{generalized-random-dot-product-graphs-with-nonlinear-community-structure}}
+
+\hypertarget{the-manifold-block-model}{%
+\subsection{The Manifold Block Model}\label{the-manifold-block-model}}
+
+\hypertarget{algorithms-for-nonintersecting-manifolds}{%
+\subsection{Algorithms for Nonintersecting Manifolds}\label{algorithms-for-nonintersecting-manifolds}}
+
+\hypertarget{algorithms-for-intersecting-manifolds}{%
+\subsection{Algorithms for Intersecting Manifolds}\label{algorithms-for-intersecting-manifolds}}
+
+\hypertarget{examples-and-simulations}{%
+\subsection{Examples and Simulations}\label{examples-and-simulations}}
+
+\hypertarget{applications-1}{%
+\subsection{Applications}\label{applications-1}}
+
+\hypertarget{conclusions}{%
+\subsection{Conclusions}\label{conclusions}}
 
@@
 
+We now provide a proof of theorem \ref{thm:ssc}.
+Our proof is based on verifying the sufficient conditions given in theorem 6 of \citet{jmlr-v28-wang13} under which sparse subspace clustering based on solving the optimization problem in equation \eqref{eq:ssc} yields an affinity matrix \(B = |C| + |C^{\top}|\) satisfying the subspace detection property of definition \ref{def:subspace_detection}.
+We first recall a few definitions used in \citet{soltanolkotabi2012} and \citet{jmlr-v28-wang13};
+for ease of exposition, these definitions are stated using the notations of the current paper and we will drop the explicit dependency on \(n\) from our eigenvectors \(\hat{V}\) of \(A\) and \(V\) of \(P\).
+
+\begin{definition}[Inradius]
+\label{def:inradius}
+The inradius of a convex body $\mathcal{P}$, denoted by $r(\mathcal{P})$, is
+defined as the radius of the largest Euclidean ball inscribed in $\mathcal{P}$.
+Let $X$ be a $n \times d$ matrix with rows $x_1, x_2, \dots,
+x_n$. We then define, with a slight abuse of notation, $r(X)$ as the
+inradius of the convex hull formed by $\{\pm x_1, \pm x_2, \dots, \pm x_n\}$. 
+\end{definition}
+
+\begin{definition}[Subspace incoherence]
+\label{def:subspace_incoherence}
+Let $\hat{V}$ be the eigenvectors of $A$
+corresponding to the $K^2$ largest eigenvalues in modulus. Let
+$\hat{V}^{(k)}$ denote the matrix formed by keeping only the rows of
+$\hat{V}$ corresponding to the $k^{th}$
+community and let $\hat{V}^{(-k)}$ denote the matrix formed by
+omitting the rows of $\hat{V}$
+corresponding to the $k^{th}$ community. Let $(\hat{v}_i^{(k)})^\top$ denote
+the $i$th row of $\hat{V}^{(k)}$ and $\hat{V}_{-i}^{(k)}$ be $\hat{V}^{(k)}$ with
+the $i^{th}$ row omitted. Let $V$, $V^{(k)}$, $V^{(-k)}$, and
+$v_i^{(k)}$ be defined similarly using the eigenvectors $V$ of
+$P$. Finally let $\mathcal{S}^{(k)}$ be the vector space spanned by the
+rows of $V^{(k)}$.
+
+Now define $\nu_{i}^{(k)}$ for $k = 1,2,\dots,K$ and $i =
+1,2,\dots,n_{k}$ as the solution of the following optimization problem
+$$\nu_{i}^{(k)} = \max_\eta (\hat{v}_i^{(k)})^\top \eta - \frac{1}{2
+  \lambda} \eta^\top \eta, \quad \text{subject to $\|V_{-i}^{(k)}
+  \eta\|_\infty \leq 1$.}$$
+Given $\nu_i^{(k)}$, let $\mathbb{P}_{\mathcal{S}^{(k)}}(\nu_i^{(k)})$
+be the vector in $\mathbb{R}^{K^2}$ corresponding to the orthogonal projection of $\nu_i^{(k)}$ onto the vector space
+$\mathcal{S}^{(k)}$ and define the projected dual direction $w_{i}^{(k)}$
+as
+$$w_i^{(k)} =
+\frac{\mathbb{P}_{\mathcal{S}^{(k)}}(\nu_i^{(k)})}{\|\mathbb{P}_{\mathcal{S}^{(k)}}(\nu_i^{(k)})\|}.$$
+Now let $W^{(k)} = \bigl[ w_1^{(k)} \mid \cdots \mid w_{n_k}^{(k)} \bigr]^\top$
+and define the subspace incoherence for $\hat{V}^{(k)}$ by
+$$\mu^{(k)} = \mu(\hat{V}^{(k)}) = \max\limits_{v \in V^{(-k)}} \|W^{(k)} v\|_\infty.$$
+\end{definition}
+
+\begin{proof}[Proof of theorem \ref{thm:ssc}]
+For a given $k = 1,2\dots,K$, let $r^{(k)} = \min_{i}r(V_{-i}^{(k)})$ be inradius of the convex hull formed by
+the rows of $V_{-i}^{(k)}$ and let $r_* = \min_{k} r^{(k)}$. Then Theorem~6 in
+\citet{jmlr-v28-wang13} states that there exists a $\lambda > 0$
+such that $\sqrt{n} \hat{V}$ satisfies the subspace detection property in definition \ref{def:subspace_detection} whenever the following two conditions are satisfied: 
+\begin{gather}
+  \label{eq:cond1}
+  \mu^{(k)} < r^{(k)} \quad \text{for all $k = 1,2,\dots,K$}, \\
+  \label{eq:cond2}
+  \max_{i} \|W \hat{v}_{i} - v_{i}\| \leq \min_{k} \frac{r_*(r^{(k)} -
+    \mu^{(k)})}{2 + 7 r^{(k)}}.
+\end{gather}
+We now verify that for sufficiently large n, equation \eqref{eq:cond1} and Eq.~\eqref{eq:cond2}
+holds with high probability.
+
+{\bf Verifying equation \eqref{eq:cond1}}. If $n$ is sufficiently large then
+there are enough vertices in each community $k$ so that
+$\mathrm{span}(V_{-i}^{(k)}) = \mathcal{S}^{(k)}$ for all $i$ and hence
+\(r^{(k)} = \min_{i} r(V_{-i}^{(k)}) > 0\) for
+all $k = 1,2,\dots,K$.
+
+Next, by theorem \ref{thm:osc-p} we have that the subspaces
+$\{\mathcal{S}^{(1)}, \dots, \mathcal{S}^{(K)}\}$
+are mutually orthogonal, i.e., $v^{\top} w = 0$ for all $v \in
+\mathcal{S}^{(k)}$ and $w \in \mathcal{S}^{(\ell)}$ with $k \not =
+\ell$. Now let $z \in \mathbb{R}^{K^2}$ be arbitrary and let
+$\tilde{z} = \mathbb{P}_{\mathcal{S}^{(k)}} z$ be the projection of
+$z$ onto $\mathcal{S}^{(k)}$. We then have $v^{\top} \tilde{z} =
+0$ for all $v \in V^{(-k)}$. Because $z$ is arbitrary, this implies 
+$\|W^{(k)} v\|_{\infty} = 0$ for all $v
+\in V^{(-k)}$ and hence $\mu^{(k)} = 0$ for all $k
+=1,2,\dots,K$. Therefore $\mu^{(k)} < r^{(k)}$ for all $k =
+1=2,\dots,K$ as desired.
+
+{\bf Verifying equation \eqref{eq:cond2}}.
+Let $\delta = \max_{i} \sqrt{n} \|W \hat{v}_{i} - v_{i}\|$. 
+Then from lemma \ref{lem:technical}, we have \(\delta \stackrel{a.s.}{\to} 0\) and hence
+$$\delta < \min_{k} \frac{r_* (r^{(k)} - \mu^{(k)})}{2 + 7 r^{(k)}}$$
+asymptotically almost surely.
+
+In summary $\sqrt{n} \hat{V}$ satisfies the subspace detection property with probability converging to $1$ as \(n \to \infty\).
+\end{proof}
+
+\begin{remark}
+Theorem 6 of \citet{jmlr-v28-wang13} assumes that each row $v_i$
+of $V$ has unit norm, i.e., $\|v_{i}\| = 1$  for all $i$. 
+This assumption has the effect of scaling the $r^{(k)}$ so that $r^{(k)} \leq 1$ for all $k = 1,2,\dots,K$. 
+We emphasize that this assumption has no effect on the proof of Theorem \ref{thm:ssc}. 
+Indeed, because $\mu^{(k)} = 0$ for all $k$, as long as the rows of $V^{(k)}$ spans the subspace $\mathcal{S}^{(k)}$, then $a r^{(k)} > \mu^{(k)}$ for any scalar $a > 0$. 
+\end{remark}
+
+\begin{proof}[Proof of Theorem \ref{thm:param-est}] 
+Let \(P\) be organized by community such that \(P^{(k \ell)}\) denote the $n_k \times n_{\ell}$ matrix obtained by keeping only the rows of $P$ corresponding to vertices in community $k$ and the columns of $P$ corresponding to vertices in community $\ell$.
+We define $A^{(k \ell)}$ analogously. 
+Recall that $P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^{\top}$ for all $k, \ell$. 
+We now consider estimation of $P^{(k \ell)}$ for the cases when $k = \ell$ versus when $k \not = \ell$.
+
+\emph{Case \(k = l\)}. 
+Let $P^{(kk)} = \sigma_{kk}^2 u^{(kk)} (u^{(kk)})^\top$ be the singular value decomposition of $P^{(kk)}$. 
+We can then define $\tilde{\lambda}^{(kk)} = \sigma_{kk} u^{(kk)}$. 
+Now let $\hat{U}^{(kk)} \hat{\Sigma}^{(kk)} (\hat{U}^{(kk)})^\top$ be the singular value decomposition of \(A^{(kk)}\), and let $\hat{\sigma}_{kk}^2 \hat{u}^{(kk)} (\hat{u}^{(kk)})^\top$ be the best rank-one approximation of $A^{(kk)}$.
+Define \(\hat{\lambda}^{(kk)} = \hat{\sigma}_{kk} \hat{u}^{(kk)}\). 
+Then \(\hat{\lambda}^{(kk)}\) is the adjacency spectral embedding approximation of \(\lambda^{(kk)}\), and by Theorem 5 of \citet{rubindelanchy2017statistical}, we have
+$$\|\hat{\lambda}^{(kk)} - \lambda^{(kk)}\|_{\infty} = O\Bigl(\frac{\log n_k}{\sqrt{n_k}}\Bigr)$$
+with high probability. 
+Here $\|\cdot\|_{\infty}$ denote the $\ell_{\infty}$ norm of a vector.
+
+\emph{Case \(k \neq l\)}. 
+Let \(P^{(k \ell)} = \sigma_{k \ell}^2 u^{(k \ell)} (v^{(k \ell)})^\top\) and \(P^{(\ell k)} = \sigma_{kl}^2 u^{(\ell k)} (v^{(\ell k)})^\top\) be the singular value decompositions and note that \(\sigma_{k \ell} = \sigma_{\ell k}\), \(u^{(k \ell)} = v^{(\ell k)}\), and \(v^{(k \ell)} = u^{(\ell k)}\).
+Now define \(\lambda^{(k \ell)} = \sigma_{k \ell} u^{(k \ell)}\) and \(\lambda^{(\ell k)} = \sigma_{k \ell} v^{(k \ell)}\).
+
+Next consider the Hermitian dilation
+\[
+M^{(k \ell)} = 2 \begin{bmatrix} 0 & P^{(k \ell)} \\ P^{(\ell k)} & 0 \end{bmatrix}
+\]
+which is a symmetric \((n_k + n_\ell) \times (n_k + n_\ell)\) matrix. 
+The eigendecomposition of \(M^{(k \ell)}\) is then
+
+\[M^{(k \ell)} = 
+\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix} \times 
+\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
+\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix}^\top\]
+Thus treating \(M^{(k \ell)}\) as the edge probability matrix of a GRDPG, we have latent positions in \(\mathbb{R}^2\) given by the $(n_k + n_{\ell}) \times 2$ matrix
+\[\Lambda^{(k \ell)} = \begin{bmatrix} 
+  \sigma_{k \ell} u^{(k \ell)} & \sigma_{k \ell} u^{(k \ell)} \\ 
+  \sigma_{k \ell} v^{(k \ell)} & -\sigma_{k \ell} v^{(k \ell)} 
+\end{bmatrix} = 
+\begin{bmatrix} 
+  \lambda^{(k \ell)} & \lambda^{(k \ell)} \\ 
+  \lambda^{( \ell k)} & -\lambda^{( \ell k)} 
+\end{bmatrix}.\]
+Now consider
+\[\hat{M}^{(k \ell)} = \begin{bmatrix} 0 & A^{(k \ell)} \\ A^{(\ell k)} & 0 \end{bmatrix}\]
+
+We can then view \(\hat{M}^{(k \ell)}\) as an adjacency matrix drawn from
+the edge probabilities matrix \(M^{(k \ell)}\). Now suppose that the adjacency spectral
+embedding of $\hat{M}^{(k \ell)}$ is represented as the $(n_k +
+n_{\ell}) \times 2$ matrix
+\[\hat{\Lambda}^{(k \ell)} = \begin{bmatrix} 
+  \hat{\lambda}^{(k \ell)} & \hat{\lambda}^{(k \ell)} \\ 
+  \hat{\lambda}^{(\ell k)} & -\hat{\lambda}^{(\ell k)} 
+\end{bmatrix}\]
+where each \(\hat{\lambda}^{(k \ell)}\) is defined as in Algorithm 3. Then
+by Theorem 5 of \citet{rubindelanchy2017statistical}, there
+exists an indefinite orthogonal transformation $W^{*}$ such that,  with
+high probability,
+$$\max_{i} |W^{*} \hat{\Lambda}_{i}^{(k \ell)} - \Lambda_{i}^{(k \ell)} \| =
+O\Bigl(\frac{\log (n_k + n_{\ell})}{\sqrt{n_k + n_{\ell}}}\Bigr)$$
+with high probability. Here $\Lambda_{i}^{(k \ell)}$ and
+$\hat{\Lambda}_i^{(k \ell)}$ denote the $i$th rows of $\Lambda^{(k
+  \ell)}$ and $\hat{\Lambda}^{(k \ell)}$, respectively.
+
+Furthermore, by looking at the proof of Theorem~5 in
+\citep{rubindelanchy2017statistical}, we see that $W^{*}$ is also
+blocks diagonal with $2$ blocks where the positive eigenvalues of $M^{(k \ell)}$
+forming a block and the negative eigenvalues of $M^{(k \ell)}$ forming
+the remaining block. 
+Because $M^{(k \ell)}$ has one positive eigenvalue and one negative
+eigenvalue, we see that $W^{*}$ is necessarily of the form $W^{*}
+= \Bigl[\begin{smallmatrix} 1 & 0 \\ 0 & - 1\end{smallmatrix}\Bigr]$
+Using this form for $W^{*}$, we obtain
+$$\max\{\|\hat{\lambda}^{(k \ell)} - \lambda^{(k \ell)}\|_{\infty},
+\|\hat{\lambda}^{(\ell k)} - \lambda^{(\ell k)}\|_{\infty}\}  =
+O\Bigl(\frac{\log(n_k + n_{\ell})}{\sqrt{n_k + n_{\ell}}}\Bigr)$$
+with high probability. Combining this bound with the bound for
+$\|\hat{\lambda}^{(kk)} - \lambda^{(kk)}\|_{\infty}$ given above
+yields equation~\eqref{eq:param-est} in theorem~\ref{thm:param-est}. 
+\end{proof}
+
 \newpage

### 2022-08-29 draft/draft.Rmd
still writing

@@
 
-\begin{example}
+\begin{example}[Graph clustering]
 $K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
@@
 In particular, there is no inherent notion of centroid or sample mean for graph data. 
-However, there are inherent notions of distances among vectors, such as shortest path distance, expected commute time, or resistance distance. 
+However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
 One way to adapt $K$-means to graph data is to use an equivalent objective function as Eq.~\eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
 $$
-\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} d(x_i, x_j)^2.
+\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
 $$
 Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
-If the data live in a Euclidean vector space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
-or if the data are represented as a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
+If they are vectors in Euclidean space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
+or if they are vertices on a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
 While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
@@
 Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
+One way to combine the graph distance approach and the graph embedding approach is to apply Lloyd's algorithm on an embedding that approximates graph distances \citep{vonLuxburg2007}. 
 \end{example}
@@
 If $A$ is binary, we call it an *adjacency matrix*. 
+In this work, we focus primarily on undirected and unweighted graphs without self loops.
 
@@
 
+\begin{definition}[Degree of a vertex]
+\label{def:degree}
+Let $G = (V, E)$ be an undirected graph with vertex set $V = \{v_1, ..., v_n\}$. 
+$d_i$, the degree of vertex $v_i$ is the sum of the weights of edges that connect to $v_i$. 
+If $G$ is unweighted, then $v_i$ is equivalently the number of edges that connect to $v_i$. 
+If $G$ is represented by affinity or adjacency matrix $A$, then $d_i = \sum_j A_{ij}$.
+
+If $G$ is a Bernoulli graph with edge probability matrix $P$, then the expected degree of vertex $v_i$ is $E[d_i] = \sum_j P_{ij}$. 
+\end{definition}
+
 ## Contributions of This Work
@@
 
-The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities has a constant edge probability $\theta_{k \ell}$.
+The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities $(k, \ell)$ has a constant edge probability $\theta_{k \ell}$.
 We now give the formal definition of the SBM within the context of Bernoulli graphs:
@@
 
-1. Assume the size of the graph $n$ is fixed. 
-Then each vertex $v_i$ has a label $z_i$, and the edge probability matrix $P$ is fixed.
-2. We sample the vertices and edges in sequence and suppose that each vertex $v_i$ comes with a label $z_i$, which determines its edge probabilities to the other vertices.
-3.
-The vertices are sampled in sequence, and each $z_i$ is sampled as $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, $\sum_k^K \alpha_k = 1$.
-The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}$
-
-An extension to these sampling schema include Bayesian models, e.g., each $\theta_{k \ell}$ is sampled from a distribution with support $[0, 1]$.
+1. Let the size of the graph, $n$, be fixed. 
+Then each vertex $v_i$, and its corresponding label, $z_i$, are also fixed, in addition to the pairwise community edge probabilities $\{\theta_{k \ell}\}_K$, which are all treated as model parameters.
+Under this sampling scheme, there is no notion of increasing $n$. 
+2. The vertices and edges are sampled in sequence, and we suppose that each vertex $v_i$ comes with a corresponding label, $z_i$, which determines its edge probabilities to the other vertices. 
+Under this sampling scheme, there is no assumed distribution on the labels, which are treated as fixed parameters rather than random variables, along with $\{\theta_{k \ell}\}_K$, but there is a notion of increasing sample size $n$, which allows for asymptotics.
+3. The vertices are sampled in sequence, and each label, $z_i$, is sampled as $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, $\sum_k^K \alpha_k = 1$. 
+The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}_K$.
+Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \begin{bmatrix} \alpha_1 & \cdots & \alpha_K \end{bmatrix}^\top$ and $\{\theta_{k \ell}\}_K$.
+
+We use the notation $A \sim \SBM(z, \{\theta_{k \ell}\}_K)$ for the first two sampling schemes, and we use $A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K)$ for the third.
+
+An extension to these include Bayesian models, e.g., each $\theta_{k \ell}$ is sampled from a distribution with support $[0, 1]$.
 
@@
 Note that in this example, the within-community edge probabilities are greater than the between-community edge probability. 
-A realization of this graph with $n_1 = n_2 = 32$ is illustrated by figure \ref{fig:assort-sbm}.
+A realization of this graph with $n_1 = n_2 = 32$ (where $n_k = |\{v_i : z_i = k\}|$) is illustrated in figure \ref{fig:assort-sbm}.
 
@@
 
-In most conceptions of graph models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}. 
-For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the graph appears to confirm this.
-However, SBMs are not restricted to this type of community structure. 
+In most conceptions of block models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}. 
+For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the network appears to confirm this.
+However, SBMs are not necessarily restricted to this type of community structure. 
 We could just as easily construct an example in which $\theta_{11}$ and $\theta_{22}$ are less than $\theta_{12}$. 
-Within the context of the SBM, we say that a graph is assortative if $P$ is positive semidefinite and disassortative otherwise, which roughly correspond to communities with higher within-community edge probabilities and higher between-community edge probabilities, respectively.
+Within the context of block models, we say that a graph is *assortative* if $P$ is positive semidefinite and *disassortative* otherwise, which roughly correspond to communities with higher within-community edge probabilities and lower between-community edge probabilities, respectively.
 For example, the SBM in example \ref{ex:assort-sbm} is assortative and has edge probability matrix $P$ that is rank 2 with nonzero eigenvalues $n (\frac{3}{8} \pm \frac{\sqrt{2}}{8}) > 0$.
@@
   \KwData{Adjacency matrix $A$, number of communities $K$}
-  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}$}
+  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$}
   \caption{Approximate EM algorithm for the SBM}
@@
       \For {$k = 1, ..., K$} {
-        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \theta_{k \ell} + (1 - A_{ij}) \log (1 - \theta_{k \ell})) \bigg)$.\;
-        M-step: $\theta_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
+        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \hat{\theta}_{k \ell} + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell})) \bigg)$.\;
+        M-step: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
       }
@@
 
-Applying algorithm \ref{algo:sbm-em} to example \ref{ex:assort-sbm} with initial guesses of $z_{ik} = 0.5$ for each $i, k$, $\hat{\theta}_{11} = \hat{\theta}_{22} = 0.9$, and $\hat{\theta}_{12} = 0.1$ results in a community detection error rate of 4.7\% and parameter estimates $\hat{\theta}_{11} = 0.503$, $\hat{\theta}_{22} = 0.241$, and $\hat{\theta}_{12} = 0.109$, compared to the true parameters $\theta_{11} = 0.5$, $\theta_{22} = 0.25$, and $\theta_{12} = 0.125$.
-The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of $z_{ik} = 0.5$, $\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1$, and $\hat{\theta}_{12} = 0.9$, results in a community detection error rate of 0\% and parameter estimates $\hat{\theta}_{11} = 0.034$, $\hat{\theta}_{22} = 0.018$, and $\hat{\theta}_{12} = 0.201$, compared to the true parameters $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$.
-
-## Random Graph Models and Sparsity
+Applying algorithm \ref{algo:sbm-em} to example \ref{ex:assort-sbm} with initial guesses of $\pi_{ik} = 0.5$ for each $i, k$, $\hat{\theta}_{11} = \hat{\theta}_{22} = 0.9$, and $\hat{\theta}_{12} = 0.1$ results in a community detection error rate of 4.7\% and parameter estimates $\hat{\theta}_{11} = 0.503$, $\hat{\theta}_{22} = 0.241$, and $\hat{\theta}_{12} = 0.109$, compared to the true parameters $\theta_{11} = 0.5$, $\theta_{22} = 0.25$, and $\theta_{12} = 0.125$.
+The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of $\pi_{ik} = 0.5$, $\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1$, and $\hat{\theta}_{12} = 0.9$, results in a community detection error rate of 0\% and parameter estimates $\hat{\theta}_{11} = 0.034$, $\hat{\theta}_{22} = 0.018$, and $\hat{\theta}_{12} = 0.201$, compared to the true parameters $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$.
+
+## Random Graph Models and Sparsity {#sec:sparsity}
+
+So far, we have described *dense* Bernoulli graphs. 
+For such models, as the sample size $n$ increases, the expected degree of each node grows linearly. 
+For example, consider the SBM with the third sampling setup described in section \@ref(sec:sbm).
+Then the expected degree of vertex $v_i$ is 
+$$
+\begin{split}
+E[d_i] & = \sum_j P_{ij} \\
+& = \sum_{j \neq i} \theta_{z_i, z_j} \\ 
+& = \Big( \sum_k n_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
+& = n \Big( \sum_k \alpha_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
+& = O(n).
+\end{split}
+$$
+
+In many real networks, the degree of each vertex often does not grow proportionally with the size of the network. 
+To account for this, a sparsity factor $\rho_n \in (0, 1]$ is introduced in the edge probabilities, i.e., $P_{ij} \leftarrow \rho_n P_{ij}$, for some sequence $\{\rho_n\}$ such that $\lim\limits_{n \to \infty} \rho_n = 0$.
+For example, a sparse SBM has edge probabilities $P_{ij} = \rho_n \theta_{z_i, z_j}$, for which we use the notation $A \sim \SBM(z, \{\theta_{k \ell}\}_K; \rho_n)$ or $A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K; \rho_n)$, depending on whether we treat the labels as random or fixed.
+Then the expected degree grows as $O(n \rho_n)$ instead of linearly as $O(n)$. 
+For the sake of unifying the sparse and dense regimes, we also allow for the special case $\rho_n = 1$ and include the sparsity factor throughout, unless otherwise stated. 
+Finally, we also note that while $\rho_n$ limits the rate of growth of the expected degree, our theoretical results still require $n \rho_n$ to diverge to infinity, albeit at a slower rate than $O(n)$ (see \cite{JMLR:v18:16-480} and \cite{rubindelanchy2017statistical} for further discussion).
 
@@
 In addition, each vertex $v_i$ has a degree correction parameter $\omega_i \in [0, 1]$. 
-If the edge probability matrix $P$ for this graph is such that $P_{ij} = \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model.
+If the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model. 
+We use the notation $A \sim \DCBM(z, \{\theta_{k \ell}\}_K, \omega; \rho_n)$ to denote a random adjacency matrix drawn from the DCBM with labels $z$, base edge probabilities $\{\theta_{k \ell}\}_K$, and degree correction factors $\omega$.
+Under the sampling scheme in which the labels are drawn from a multivariate distribution with class probabilities $\alpha$, we use the notation $A \sim \DCBM(\alpha, \{\theta_{k \ell}\}_K, \omega; \rho_n)$.
 \end{definition}
@@
 In this model, we assume that for each pair of genders, there is a constant probability of a match. 
-This does not take into account that some individuals are more likely to form connections than others. 
-Modeling this as a DCBM allows us to take each individual's popularity into account.
-
+This does not take into account that some individuals are more likely to form connections than others, based on each individual's activity on the online service or popularity with other users. 
+Modeling this as a DCBM allows us to take this into account by adjusting each user's degree by $\omega_i$. 
+The following is a visualization of this network with parameters $\theta_{11} = \theta_{22} = 0.04$, $\theta_{12} = 0.4$, and each $\omega_i$ drawn uniformly between $0$ and $1$.
+
+```{r dating-dcbm, fig.width = 4, , fig.height = 2, fig.cap = 'A dating network modeled as a DCBM. This model is disassortative.', cache = TRUE}
+n <- 2 ** 6
+z <- sample(c(1, 2), n, replace = TRUE)
+z <- sort(z)
+n1 <- sum(z == 1)
+n2 <- sum(z == 2)
+z[z == 1] <- 'F'
+z[z == 2] <- 'M'
+theta11 <- .02
+theta22 <- .02
+theta12 <- .2
+omega <- runif(n)
+Omega <- omega %*% t(omega)
+P <- matrix(theta12, nrow = n, ncol = n)
+P[seq(n1), seq(n1)] <- theta11
+P[seq(n1 + 1, n), seq(n1 + 1, n)] <- theta22
+P <- P * Omega
+A <- draw.graph(P)
+# qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
+ggraph(A, layout = 'auto') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = TRUE) + 
+  labs(colour = 'Gender') + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
 \end{example}
+
+\begin{remark}
+Unlike the SBM (when excluding the sparsity factor), each DCBM is not unique and therefore non-identifiable. 
+\end{remark}
 
@@
 Suppose that each vertex $v_i$ has $K$ popularity parameters $\lambda_{i1}, ..., \lambda_{iK}$ for which each $\lambda_{ik}$ describes $v_i$'s affinity toward community $k$. 
-Then if the edge probability matrix $P$ for this graph is such that $P_{ij} = \lambda_{i, z_j} \lambda_{j, z_i}$ and $A \sim \BG(P)$, then $G$ is a popularity adjusted block model.
+Then if the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \lambda_{i, z_j} \lambda_{j, z_i}$ and $A \sim \BG(P)$, then $G$ is a popularity adjusted block model.
 \end{definition}
@@
 Denote by $\tilde{P}^{(k \ell)}$ the $n_k \times n_{\ell}$ submatrix of $\tilde{P}$ corresponding to the edge probabilities between vertices in communities $k$ and $\ell$. 
-Here $n_k = |\{ i \colon z_i = k\}|$ is the number of vertices assigned to community $k$, for $k = 1,2\dots,K$.
 Note that $\tilde{P}^{(k \ell)} = (\tilde{P}^{(\ell k)})^\top$. 
@@
 
+## Community Detection as Clustering in the Latent Space
+
+As established in section \@ref(sec:grdpg), all Bernoulli random graphs are generalized random dot product graphs. 
+Whether this is useful for community detection in block models depends entirely on the configuration of the latent vectors. 
+In the case of the $\ER$ model, SBM, DCBM, and PABM, each community forms a linear structure, i.e., a subspace, and this rigid structure can be exploited for community detection. 
+In this section, we explore community-wise nonlinear latent structures.
+
 ## The Manifold Block Model
+
+\begin{definition}[Manifold block model]
+Let $p, q \geq 0$, $d = p + q \geq 1$, $1 \leq r < d$, $K \geq 2$, and $n > K$ be integers. 
+Define manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \in \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1]\}$ each by continuous function $g_k : [0, 1]^r \to \mathcal{X}$.
+Define probability distributions $F_1, ..., F_K$ each with support $[0, 1]^r$. 
+Then the following mixture model is a {\em manifold block model}.
+
+\begin{enumerate}
+\item Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$.
+\item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
+\item Compile the latent vectors into data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
+\end{enumerate}
+\end{definition}
+
+\begin{example}
+\end{example}
 

### 2022-08-29 draft/draft.tex
still writing

@@
 If \(A\) is binary, we call it an \emph{adjacency matrix}.
+In this work, we focus primarily on undirected and unweighted graphs without self loops.
 
@@
 
+\begin{definition}[Degree of a vertex]
+\label{def:degree}
+Let $G = (V, E)$ be an undirected graph with vertex set $V = \{v_1, ..., v_n\}$. 
+$d_i$, the degree of vertex $v_i$ is the sum of the weights of edges that connect to $v_i$. 
+If $G$ is unweighted, then $v_i$ is equivalently the number of edges that connect to $v_i$. 
+If $G$ is represented by affinity or adjacency matrix $A$, then $d_i = \sum_j A_{ij}$.
+
+If $G$ is a Bernoulli graph with edge probability matrix $P$, then the expected degree of vertex $v_i$ is $E[d_i] = \sum_j P_{ij}$. 
+\end{definition}
+
 \hypertarget{contributions-of-this-work}{%
@@
 
-The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities has a constant edge probability \(\theta_{k \ell}\).
+The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities \((k, \ell)\) has a constant edge probability \(\theta_{k \ell}\).
 We now give the formal definition of the SBM within the context of Bernoulli graphs:
@@
 \item
-  Assume the size of the graph \(n\) is fixed.
-  Then each vertex \(v_i\) has a label \(z_i\), and the edge probability matrix \(P\) is fixed.
+  Let the size of the graph, \(n\), be fixed.
+  Then each vertex \(v_i\), and its corresponding label, \(z_i\), are also fixed, in addition to the pairwise community edge probabilities \(\{\theta_{k \ell}\}_K\), which are all treated as model parameters.
+  Under this sampling scheme, there is no notion of increasing \(n\).
 \item
-  We sample the vertices and edges in sequence and suppose that each vertex \(v_i\) comes with a label \(z_i\), which determines its edge probabilities to the other vertices.
+  The vertices and edges are sampled in sequence, and we suppose that each vertex \(v_i\) comes with a corresponding label, \(z_i\), which determines its edge probabilities to the other vertices.
+  Under this sampling scheme, there is no assumed distribution on the labels, which are treated as fixed parameters rather than random variables, along with \(\{\theta_{k \ell}\}_K\), but there is a notion of increasing sample size \(n\), which allows for asymptotics.
 \item
-  The vertices are sampled in sequence, and each \(z_i\) is sampled as \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), \(\sum_k^K \alpha_k = 1\).
-The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}\)
+  The vertices are sampled in sequence, and each label, \(z_i\), is sampled as \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), \(\sum_k^K \alpha_k = 1\).
+  The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}_K\).
+  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \begin{bmatrix} \alpha_1 & \cdots & \alpha_K \end{bmatrix}^\top\) and \(\{\theta_{k \ell}\}_K\).
 \end{enumerate}
 
-An extension to these sampling schema include Bayesian models, e.g., each \(\theta_{k \ell}\) is sampled from a distribution with support \([0, 1]\).
+We use the notation \(A \sim \SBM(z, \{\theta_{k \ell}\}_K)\) for the first two sampling schemes, and we use \(A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K)\) for the third.
+
+An extension to these include Bayesian models, e.g., each \(\theta_{k \ell}\) is sampled from a distribution with support \([0, 1]\).
 
@@
 
-In most conceptions of graph models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}.
-For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the graph appears to confirm this.
-However, SBMs are not restricted to this type of community structure.
+In most conceptions of block models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}.
+For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the network appears to confirm this.
+However, SBMs are not necessarily restricted to this type of community structure.
 We could just as easily construct an example in which \(\theta_{11}\) and \(\theta_{22}\) are less than \(\theta_{12}\).
-Within the context of the SBM, we say that a graph is assortative if \(P\) is positive semidefinite and disassortative otherwise, which roughly correspond to communities with higher within-community edge probabilities and higher between-community edge probabilities, respectively.
+Within the context of block models, we say that a graph is \emph{assortative} if \(P\) is positive semidefinite and \emph{disassortative} otherwise, which roughly correspond to communities with higher within-community edge probabilities and lower between-community edge probabilities, respectively.
 For example, the SBM in example \ref{ex:assort-sbm} is assortative and has edge probability matrix \(P\) that is rank 2 with nonzero eigenvalues \(n (\frac{3}{8} \pm \frac{\sqrt{2}}{8}) > 0\).
@@
 
-Applying algorithm \ref{algo:sbm-em} to example \ref{ex:assort-sbm} with initial guesses of \(z_{ik} = 0.5\) for each \(i, k\), \(\hat{\theta}_{11} = \hat{\theta}_{22} = 0.9\), and \(\hat{\theta}_{12} = 0.1\) results in a community detection error rate of 4.7\% and parameter estimates \(\hat{\theta}_{11} = 0.503\), \(\hat{\theta}_{22} = 0.241\), and \(\hat{\theta}_{12} = 0.109\), compared to the true parameters \(\theta_{11} = 0.5\), \(\theta_{22} = 0.25\), and \(\theta_{12} = 0.125\).
-The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of \(z_{ik} = 0.5\), \(\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1\), and \(\hat{\theta}_{12} = 0.9\), results in a community detection error rate of 0\% and parameter estimates \(\hat{\theta}_{11} = 0.034\), \(\hat{\theta}_{22} = 0.018\), and \(\hat{\theta}_{12} = 0.201\), compared to the true parameters \(\theta_{11} = \theta_{22} = 0.02\) and \(\theta_{12} = 0.2\).
-
-\hypertarget{random-graph-models-and-sparsity}{%
-\subsection{Random Graph Models and Sparsity}\label{random-graph-models-and-sparsity}}
+Applying algorithm \ref{algo:sbm-em} to example \ref{ex:assort-sbm} with initial guesses of \(\pi_{ik} = 0.5\) for each \(i, k\), \(\hat{\theta}_{11} = \hat{\theta}_{22} = 0.9\), and \(\hat{\theta}_{12} = 0.1\) results in a community detection error rate of 4.7\% and parameter estimates \(\hat{\theta}_{11} = 0.503\), \(\hat{\theta}_{22} = 0.241\), and \(\hat{\theta}_{12} = 0.109\), compared to the true parameters \(\theta_{11} = 0.5\), \(\theta_{22} = 0.25\), and \(\theta_{12} = 0.125\).
+The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of \(\pi_{ik} = 0.5\), \(\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1\), and \(\hat{\theta}_{12} = 0.9\), results in a community detection error rate of 0\% and parameter estimates \(\hat{\theta}_{11} = 0.034\), \(\hat{\theta}_{22} = 0.018\), and \(\hat{\theta}_{12} = 0.201\), compared to the true parameters \(\theta_{11} = \theta_{22} = 0.02\) and \(\theta_{12} = 0.2\).
+
+\hypertarget{sec:sparsity}{%
+\subsection{Random Graph Models and Sparsity}\label{sec:sparsity}}
+
+So far, we have described \emph{dense} Bernoulli graphs.
+For such models, as the sample size \(n\) increases, the expected degree of each node grows linearly.
+For example, consider the SBM with the third sampling setup described in section \ref{sec:sbm}.
+Then the expected degree of vertex \(v_i\) is
+\[
+\begin{split}
+E[d_i] & = \sum_j P_{ij} \\
+& = \sum_{j \neq i} \theta_{z_i, z_j} \\ 
+& = \Big( \sum_k n_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
+& = n \Big( \sum_k \alpha_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
+& = O(n).
+\end{split}
+\]
+
+In many real networks, the degree of each vertex often does not grow proportionally with the size of the network.
+To account for this, a sparsity factor \(\rho_n \in (0, 1]\) is introduced in the edge probabilities, i.e., \(P_{ij} \leftarrow \rho_n P_{ij}\), for some sequence \(\{\rho_n\}\) such that \(\lim\limits_{n \to \infty} \rho_n = 0\).\\
+For example, a sparse SBM has edge probabilities \(P_{ij} = \rho_n \theta_{z_i, z_j}\), for which we use the notation \(A \sim \SBM(z, \{\theta_{k \ell}\}_K; \rho_n)\) or \(A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K; \rho_n)\), depending on whether we treat the labels as random or fixed.
+Then the expected degree grows as \(O(n \rho_n)\) instead of linearly as \(O(n)\).
+For the sake of unifying the sparse and dense regimes, we also allow for the special case \(\rho_n = 1\) and include the sparsity factor throughout, unless otherwise stated.
+Finally, we also note that while \(\rho_n\) limits the rate of growth of the expected degree, our theoretical results still require \(n \rho_n\) to diverge to infinity, albeit at a slower rate than \(O(n)\) (see \cite{JMLR:v18:16-480} and \cite{rubindelanchy2017statistical} for further discussion).
 
@@
 In this model, we assume that for each pair of genders, there is a constant probability of a match. 
-This does not take into account that some individuals are more likely to form connections than others. 
-Modeling this as a DCBM allows us to take each individual's popularity into account.
-
+This does not take into account that some individuals are more likely to form connections than others, based on each individual's activity on the online service or popularity with other users. 
+Modeling this as a DCBM allows us to take this into account by adjusting each user's degree by $\omega_i$. 
+The following is a visualization of this network with parameters $\theta_{11} = \theta_{22} = 0.04$, $\theta_{12} = 0.4$, and each $\omega_i$ drawn uniformly between $0$ and $1$.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/dating-dcbm-1}
+
+}
+
+\caption{A dating network modeled as a DCBM. This model is disassortative.}\label{fig:dating-dcbm}
+\end{figure}
 \end{example}
+
+\begin{remark}
+Unlike the SBM (when excluding the sparsity factor), each DCBM is not unique and therefore non-identifiable. 
+\end{remark}
 
@@
 
+\hypertarget{community-detection-as-clustering-in-the-latent-space}{%
+\subsection{Community Detection as Clustering in the Latent Space}\label{community-detection-as-clustering-in-the-latent-space}}
+
+As established in section \ref{sec:grdpg}, all Bernoulli random graphs are generalized random dot product graphs.
+Whether this is useful for community detection in block models depends entirely on the configuration of the latent vectors.
+In the case of the \(\ER\) model, SBM, DCBM, and PABM, each community forms a linear structure, i.e., a subspace, and this rigid structure can be exploited for community detection.
+In this section, we explore community-wise nonlinear latent structures.
+
 \hypertarget{the-manifold-block-model}{%
 \subsection{The Manifold Block Model}\label{the-manifold-block-model}}
+
+\begin{definition}[Manifold block model]
+Let $p, q \geq 0$, $d = p + q \geq 1$, $1 \leq r < d$, $K \geq 2$, and $n > K$ be integers. 
+Define manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \in \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1]\}$ each by continuous function $g_k : [0, 1]^r \to \mathcal{X}$.
+Define probability distributions $F_1, ..., F_K$ each with support $[0, 1]^r$. 
+Then the following mixture model is a {\em manifold block model}.
+
+\begin{enumerate}
+\item Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$.
+\item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
+\item Compile the latent vectors into data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
+\end{enumerate}
+\end{definition}
+
+\begin{example}
+\end{example}
 

### 2022-08-29 draft/draft.Rmd
still writing

@@
 Modeling this as a DCBM allows us to take this into account by adjusting each user's degree by $\omega_i$. 
-The following is a visualization of this network with parameters $\theta_{11} = \theta_{22} = 0.04$, $\theta_{12} = 0.4$, and each $\omega_i$ drawn uniformly between $0$ and $1$.
+The following is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n \equiv 1$.
 
 ```{r dating-dcbm, fig.width = 4, , fig.height = 2, fig.cap = 'A dating network modeled as a DCBM. This model is disassortative.', cache = TRUE}
+set.seed(12345)
+
 n <- 2 ** 6
@@
 z[z == 2] <- 'M'
-theta11 <- .02
-theta22 <- .02
-theta12 <- .2
-omega <- runif(n)
+theta11 <- .03
+theta22 <- .03
+theta12 <- .3
+omega <- runif(n, 1/3, 1)
 Omega <- omega %*% t(omega)
@@
 A <- draw.graph(P)
-# qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
 ggraph(A, layout = 'auto') + 
@@
 
-\begin{remark}
-Unlike the SBM (when excluding the sparsity factor), each DCBM is not unique and therefore non-identifiable. 
-\end{remark}
-
 \begin{definition}[Popularity adjusted block model]
@@
 Then if the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \lambda_{i, z_j} \lambda_{j, z_i}$ and $A \sim \BG(P)$, then $G$ is a popularity adjusted block model.
+We use the notation $A \sim \PABM(z, \{\lambda_{i, k}\}_K; \rho_n)$ to denote a random adjacency matrix drawn from the PABM with labels $z$ and popularity parameters $\{\lambda_{i, k}\}_K$. 
+Under the sampling scheme in which the labels are drawn from a multivariate distribution with class probabilities $\alpha$, we use the notation $A \sim \PABM(\alpha, \{\lambda_{i, k}\}_K; \rho_n)$. 
 \end{definition}
+
+\begin{remark}
+Unlike the SBM (when excluding the sparsity factor), each DCBM is not unique and therefore non-identifiable. 
+For example, doubling each $\theta_{k \ell}$ and dividing each $\omega_i$ by $\sqrt{2}$ results in the same edge probability matrix. 
+Nevertheless, a naive EM type algorithm similar to the one in section \ref{sec:sbm} results in high community detection accuracy for most networks, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees.
+Likewise, it is not clear 
+\end{remark}
 

### 2022-08-29 draft/draft.tex
still writing

@@
 Modeling this as a DCBM allows us to take this into account by adjusting each user's degree by $\omega_i$. 
-The following is a visualization of this network with parameters $\theta_{11} = \theta_{22} = 0.04$, $\theta_{12} = 0.4$, and each $\omega_i$ drawn uniformly between $0$ and $1$.
+The following is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n \equiv 1$.
 
@@
 \end{example}
-
-\begin{remark}
-Unlike the SBM (when excluding the sparsity factor), each DCBM is not unique and therefore non-identifiable. 
-\end{remark}
 

### 2022-09-12 draft/draft.Rmd
still writing

@@
 
-If $G$ is a Bernoulli graph with edge probability matrix $P$, then the expected degree of vertex $v_i$ is $E[d_i] = \sum_j P_{ij}$. 
+If $G$ is a Bernoulli random graph with edge probability matrix $P$, then the degree of each vertex $v_i$ is a random variable, and its expected value (also called expected degree) is $E[d_i] = \sum_j P_{ij}$.
+For example, the expected degree of each vertex of an $\ER$ graph with $n$ vertices and constant edge probability $p$ is $(n - 1) p$. 
 \end{definition}
@@
 
-# Block Models for Community Detection {#sec:blockmodel}
+# Preliminaries
+
+## Block Models for Community Detection {#sec:blockmodel}
 
@@
 For the sake of unifying the sparse and dense regimes, we also allow for the special case $\rho_n = 1$ and include the sparsity factor throughout, unless otherwise stated. 
-Finally, we also note that while $\rho_n$ limits the rate of growth of the expected degree, our theoretical results still require $n \rho_n$ to diverge to infinity, albeit at a slower rate than $O(n)$ (see \cite{JMLR:v18:16-480} and \cite{rubindelanchy2017statistical} for further discussion).
+Finally, we also note that while $\rho_n$ limits the rate of growth of the expected degree, our theoretical results still require $n \rho_n$ to diverge to infinity, albeit at a slower rate than $O(n)$.
+In general, for consistency in estimation and inference, most results on Bernoulli random graphs require $n \rho_n = \omega \big( (\log n)^c \big)$ for some $c > 1$. 
+(See \cite{JMLR:v18:16-480}, \cite{https://doi.org/10.48550/arxiv.2106.09840}, and \cite{rubindelanchy2017statistical} for further discussion.)
 
 ## Generalizations of the Stochastic Block Model: The Degree Corrected Block Model and the Popularity Adjusted Block Model {#sec:dcbm-pabm}
+
+The SBM assumes homogeneity within each community. 
+In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label. 
+The Degree Corrected Block Model (DCBM) and Popularity Adjusted Block Model (PABM) are generalizations of the SBM that address these limitations. 
+In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label.
+We also show that the DCBM is a special case of the PABM and that the SBM, DCBM, and PABM are nested models.
 
@@
 \begin{remark}
-Unlike the SBM (when excluding the sparsity factor), each DCBM is not unique and therefore non-identifiable. 
-For example, doubling each $\theta_{k \ell}$ and dividing each $\omega_i$ by $\sqrt{2}$ results in the same edge probability matrix. 
+Unlike the SBM (when excluding the sparsity factor), each DCBM and PABM is not unique and therefore non-identifiable. 
+For example, in the case of the DCBM, doubling each $\theta_{k \ell}$ and dividing each $\omega_i$ by $\sqrt{2}$ results in the same edge probability matrix. 
 Nevertheless, a naive EM type algorithm similar to the one in section \ref{sec:sbm} results in high community detection accuracy for most networks, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees.
-Likewise, it is not clear 
+Likewise, it is not immediately obvious how to approach the estimation of the $\theta_{k \ell}$'s, $\omega_i$'s, or $\lambda_{ik}$'s due to their non-uniqueness. 
+Applying an EM algorithm to example \ref{ex:dcbm-dating} results in 84\% accuracy but fails to capture the original (which are not necessarily ``true'') parameters. 
 \end{remark}
@@
 
-# Random Dot Product Graphs and Generalized Random Dot Product Graphs {#sec:grdpg}
-
-## Definitions
+## Random Dot Product Graphs and Generalized Random Dot Product Graphs {#sec:grdpg}
 
@@
 We use the notation $A \sim \GRDPG_{p,q}(X; \rho_n)$ to denote a random adjacency matrix $A$ drawn from latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$.
-If the $n$ vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F, n; \rho_n)$.
+If the $n$ vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F; \rho_n)$.
 
 If $q = 0$ and so $d = p$, (i.e., $P$ is positive semidefinite), then we call this a random dot product graph. 
-In this case, we use the notation $A \sim \RDPG(X; \rho_n)$ and $A \sim \RDPG(F, n; \rho_n)$. 
+In this case, we use the notation $A \sim \RDPG(X; \rho_n)$ and $A \sim \RDPG(F; \rho_n)$. 
 \end{definition}

### 2022-09-12 draft/draft.tex
still writing

@@
 
+\hypertarget{preliminaries}{%
+\section{Preliminaries}\label{preliminaries}}
+
 \hypertarget{sec:blockmodel}{%
-\section{Block Models for Community Detection}\label{sec:blockmodel}}
+\subsection{Block Models for Community Detection}\label{sec:blockmodel}}
 
@@
 For the sake of unifying the sparse and dense regimes, we also allow for the special case \(\rho_n = 1\) and include the sparsity factor throughout, unless otherwise stated.
-Finally, we also note that while \(\rho_n\) limits the rate of growth of the expected degree, our theoretical results still require \(n \rho_n\) to diverge to infinity, albeit at a slower rate than \(O(n)\) (see \cite{JMLR:v18:16-480} and \cite{rubindelanchy2017statistical} for further discussion).
+Finally, we also note that while \(\rho_n\) limits the rate of growth of the expected degree, our theoretical results still require \(n \rho_n\) to diverge to infinity, albeit at a slower rate than \(O(n)\).
+In general, for consistency in estimation and inference, most results on Bernoulli random graphs require \(n \rho_n = \omega \big( (\log n)^c \big)\) for some \(c > 1\).
+(See \cite{JMLR:v18:16-480}, \cite{https://doi.org/10.48550/arxiv.2106.09840}, and \cite{rubindelanchy2017statistical} for further discussion.)
 
@@
 \subsection{Generalizations of the Stochastic Block Model: The Degree Corrected Block Model and the Popularity Adjusted Block Model}\label{sec:dcbm-pabm}}
+
+The SBM assumes homogeneity within each community.
+In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label.
+The Degree Corrected Block Model (DCBM) and Popularity Adjusted Block Model (PABM) are generalizations of the SBM that address these limitations.
+In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label.
+We also show that the DCBM is a special case of the PABM and that the SBM, DCBM, and PABM are nested models.
 
@@
 \hypertarget{sec:grdpg}{%
-\section{Random Dot Product Graphs and Generalized Random Dot Product Graphs}\label{sec:grdpg}}
-
-\hypertarget{definitions}{%
-\subsection{Definitions}\label{definitions}}
+\subsection{Random Dot Product Graphs and Generalized Random Dot Product Graphs}\label{sec:grdpg}}
 

### 2022-09-19 draft/draft.Rmd
still writing

@@
 However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
-One way to adapt $K$-means to graph data is to use an equivalent objective function as Eq.~\eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
+One way to adapt $K$-means to graph data is to use an equivalent objective function as equation~\eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
 $$
@@
 \end{definition}
+
+Showing explicitly that the SBM is a special case of the DCBM only requires us to set $\omega$ as a constant vector, e.g., $\omega = \begin{bmatrix} 1 & \cdots & 1 \end{bmatrix}^\top$.
 
@@
 \tilde{\Pi} X (X^\top X)^{-1} X^\top \tilde{\Pi}^{\top} $ where $X$ is 
-defined as in Eq.~\eqref{eq:xy}. Indeed, by Theorem 2, 
+defined as in equation~\eqref{eq:xy}. Indeed, by Theorem 2, 
 \(P = \tilde{\Pi} X U I_{p, q} U^\top X^\top \tilde{\Pi}\) for $p = K(K+1)/2$ and $q = K(K-1)/2$. 
@@
 In the case of the $\ER$ model, SBM, DCBM, and PABM, each community forms a linear structure, i.e., a subspace, and this rigid structure can be exploited for community detection. 
-In this section, we explore community-wise nonlinear latent structures.
+In this section, we explore GRDPGs with community-wise parameterized nonlinear latent structures.
+
+To motivate this model, consider a generalization of the $\ER$ model in which each edge probability isn't constant but comes from a simple distribution. 
+As with all Bernoulli random graphs, this is also a GRDPG, but by using a specific type of distribution, it is possible to construct specific latent structures on which the latent vectors lie. 
+Or, in reverse, by constructing specific latent structures, the edge probabilities of the Bernoulli random graph follow specific distributions.
+
+\begin{example}
+In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
+This corresponds to a one-dimensional RDPG with a latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$. 
+In another view, the RDPG with latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
+
+Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \begin{bmatrix} \cos \frac{\pi}{2} t & \sin \frac{\pi}{2} t \end{bmatrix}^\top$ for $t \in [0, 1]$.
+For simplicity, we set $\rho_n \equiv 1$. 
+Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$.
+\end{example}
+
+By changing the latent structure from a point mass to a curve, we are able to come up with more flexible Bernoulli random graph models in which the edge probabilities follow more general probability distributions.
+Community structure can then be added by sampling latent vectors from multiple curves. 
+The goal of this section is to explore GRDPGs with such structures and develop algorithms for community detection and parameter estimation.
 
@@
 
+We begin by defining the model:
+
 \begin{definition}[Manifold block model]
+\label{def:mbm}
 Let $p, q \geq 0$, $d = p + q \geq 1$, $1 \leq r < d$, $K \geq 2$, and $n > K$ be integers. 
@@
 
+Definition \ref{def:mbm}
+
 \begin{example}
+Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
+Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+Finally, let $A \sim \RDPG(X)$.
+Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.
+Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.
+
+```{r two-curves, fig.cap = 'Manifold block model described in Example 1.
+The latent configuration is on the left, a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.', fig.height = 2, fig.width = 7}
+set.seed(123456)
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+u1 <- runif(n1)
+u2 <- runif(n2)
+x1 <- cos(pi / 3 * u1)
+y1 <- sin(pi / 3 * u1)
+x2 <- 1 + cos(pi / 3 * u2 + pi)
+y2 <- 1 + sin(pi / 3 * u2 + pi)
+data.matrix <- cbind(c(x1, x2), c(y1, y2))
+A <- draw.graph(data.matrix %*% t(data.matrix))
+Xhat <- embedding(A, 2, 0)
+
+par(mfrow = c(1, 3))
+# par(fig = c(0, .7, 0, 1))
+plot(data.matrix, col = z * 2, asp = 1,
+     xlab = NA, ylab = NA, 
+     # xaxt = 'n', yaxt = 'n',
+     xlim = c(-.1, 1.1), ylim = c(0, 1))
+# ggplot() +
+#   geom_point(aes(x = data.matrix[, 1], y = data.matrix[, 2], colour = factor(z))) +
+#   labs(x = NULL, y = NULL) +
+#   scale_colour_brewer(palette = 'Set1') +
+#   theme(legend.position = 'none') +
+#   coord_fixed()
+# par(fig = c(.3, 1, 0, 1))
+qgraph::qgraph(A, groups = factor(z), legend = FALSE)
+plot(Xhat, col = z * 2, asp = 1, 
+     # xaxt = 'n', yaxt = 'n', 
+     xlab = NA, ylab = NA)
+```
 \end{example}
+
+In the case of the SBM, a GRDPG-motivated community detection algorithm is to construct the ASE of the graph and then apply a point cloud based clustering algorithm such as $K$-means or GMM on the embedding.
+We saw that in the case of the DCBM and PABM, because the latent structures are not based on point masses, while the ASE does reveal the communities, the final clustering step requires other clustering algorithms that are tailored to the latent community structure.
+Likewise, in the following sections, we develop algorithms specific to the structure of the manifold block model (MBM), one in the case of disjoint manifolds and one in the general case in which the latent manifolds can intersect.
+Both algorithms have theoretical consistency properties.
 
@@
 \end{gather}
-We now verify that for sufficiently large n, equation \eqref{eq:cond1} and Eq.~\eqref{eq:cond2}
+We now verify that for sufficiently large n, equation \eqref{eq:cond1} and equation~\eqref{eq:cond2}
 holds with high probability.

### 2022-09-19 draft/draft.tex
still writing

@@
 \end{definition}
+
+Showing explicitly that the SBM is a special case of the DCBM only requires us to set \(\omega\) as a constant vector, e.g., \(\omega = \begin{bmatrix} 1 & \cdots & 1 \end{bmatrix}^\top\).
 
@@
 In the case of the \(\ER\) model, SBM, DCBM, and PABM, each community forms a linear structure, i.e., a subspace, and this rigid structure can be exploited for community detection.
-In this section, we explore community-wise nonlinear latent structures.
+In this section, we explore GRDPGs with community-wise parameterized nonlinear latent structures.
+
+To motivate this model, consider a generalization of the \(\ER\) model in which each edge probability isn't constant but comes from a simple distribution.
+As with all Bernoulli random graphs, this is also a GRDPG, but by using a specific type of distribution, it is possible to construct specific latent structures on which the latent vectors lie.
+Or, in reverse, by constructing specific latent structures, the edge probabilities of the Bernoulli random graph follow specific distributions.
+
+\begin{example}
+In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
+This corresponds to a one-dimensional RDPG with a latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$. 
+In another view, the RDPG with latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
+
+Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \begin{bmatrix} \cos \frac{\pi}{2} t & \sin \frac{\pi}{2} t \end{bmatrix}^\top$ for $t \in [0, 1]$.
+For simplicity, we set $\rho_n \equiv 1$. 
+Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$.
+\end{example}
+
+By changing the latent structure from a point mass to a curve, we are able to come up with more flexible Bernoulli random graph models in which the edge probabilities follow more general probability distributions.
+Community structure can then be added by sampling latent vectors from multiple curves.
+The goal of this section is to explore GRDPGs with such structures and develop algorithms for community detection and parameter estimation.
 
@@
 
+Definition \ref{def:mbm}
+
 \begin{example}
+Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
+Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+Finally, let $A \sim \RDPG(X)$.
+Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.
+Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/two-curves-1}
+
+}
+
+\caption{Manifold block model described in Example 1.
+The latent configuration is on the left, a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.}\label{fig:two-curves}
+\end{figure}
 \end{example}
+
+In the case of the SBM, a GRDPG-motivated community detection algorithm is to construct the ASE of the graph and then apply a point cloud based clustering algorithm such as \(K\)-means or GMM on the embedding.
+We saw that in the case of the DCBM and PABM, because the latent structures are not based on point masses, while the ASE does reveal the communities, the final clustering step requires other clustering algorithms that are tailored to the latent community structure.
+Likewise, in the following sections, we develop algorithms specific to the structure of the manifold block model (MBM), one in the case of disjoint manifolds and one in the general case in which the latent manifolds can intersect.
+Both algorithms have theoretical consistency properties.
 

### 2022-09-26 draft/draft.Rmd
still writing

@@
 
-# Generalized Random Dot Product Graphs with Nonlinear Community Structure
+# Generalized Random Dot Product Graphs with Nonlinear Community Structure {#sec:mbm}
 
@@
 
-Definition \ref{def:mbm}
+Definition \ref{def:mbm} describes the most general version of the manifold block model (MBM). 
+Some plausible constraints may be to fix one distribution for each underlying distribution $F_{z_i} \equiv F$ or to restrict the manifolds to be one-dimensional, i.e., $r \equiv 1$. 
+We use these constraints to construct the following example:
 
@@
 We saw that in the case of the DCBM and PABM, because the latent structures are not based on point masses, while the ASE does reveal the communities, the final clustering step requires other clustering algorithms that are tailored to the latent community structure.
-Likewise, in the following sections, we develop algorithms specific to the structure of the manifold block model (MBM), one in the case of disjoint manifolds and one in the general case in which the latent manifolds can intersect.
-Both algorithms have theoretical consistency properties.
-
-## Algorithms for Nonintersecting Manifolds
-
-## Algorithms for Intersecting Manifolds
+Likewise, in the following sections, we develop algorithms specific to the structure of the MBM, one in the case of disjoint manifolds (section \@ref(sec:nonintersect)) and one in the general case in which the latent manifolds can intersect (section \@ref(sec:intersect)).
+We also derive theoretical consistency properties for both settings.
+
+## Algorithms for Nonintersecting Manifolds {#sec:nonintersect}
+
+In this section, we consider the following setting: Suppose that each community $C_k$ is represented by a closed manifold $\mathcal{M}_k$ in the latent space of a RDPG or GRDPG. 
+Define $\delta = \min\limits_{k \neq \ell} \min\limits_{x \in \mathcal{M}_k, y \in \mathcal{M}_\ell} \|x - y\|$, the minimum distance between two manifolds. 
+Furthermore, assume that $\delta > 0$, i.e., the manifolds do not intersect.
+
+In the original latent space, if the vectors on each manifold is sufficiently dense, it is possible to construct for each manifold $\mathcal{M}_k$ an $\eta_k$-neighborhood graph for some $\eta_k > 0$.
+
+## Algorithms for Intersecting Manifolds {#sec:intersect}
 

### 2022-09-26 draft/draft.tex
still writing

@@
 
-\hypertarget{generalized-random-dot-product-graphs-with-nonlinear-community-structure}{%
-\section{Generalized Random Dot Product Graphs with Nonlinear Community Structure}\label{generalized-random-dot-product-graphs-with-nonlinear-community-structure}}
+\hypertarget{sec:mbm}{%
+\section{Generalized Random Dot Product Graphs with Nonlinear Community Structure}\label{sec:mbm}}
 
@@
 
-Definition \ref{def:mbm}
+Definition \ref{def:mbm} describes the most general version of the manifold block model (MBM).
+Some plausible constraints may be to fix one distribution for each underlying distribution \(F_{z_i} \equiv F\) or to restrict the manifolds to be one-dimensional, i.e., \(r \equiv 1\).
+We use these constraints to construct the following example:
 
@@
 We saw that in the case of the DCBM and PABM, because the latent structures are not based on point masses, while the ASE does reveal the communities, the final clustering step requires other clustering algorithms that are tailored to the latent community structure.
-Likewise, in the following sections, we develop algorithms specific to the structure of the manifold block model (MBM), one in the case of disjoint manifolds and one in the general case in which the latent manifolds can intersect.
-Both algorithms have theoretical consistency properties.
-
-\hypertarget{algorithms-for-nonintersecting-manifolds}{%
-\subsection{Algorithms for Nonintersecting Manifolds}\label{algorithms-for-nonintersecting-manifolds}}
-
-\hypertarget{algorithms-for-intersecting-manifolds}{%
-\subsection{Algorithms for Intersecting Manifolds}\label{algorithms-for-intersecting-manifolds}}
+Likewise, in the following sections, we develop algorithms specific to the structure of the MBM, one in the case of disjoint manifolds (section \ref{sec:nonintersect}) and one in the general case in which the latent manifolds can intersect (section \ref{sec:intersect}).
+We also derive theoretical consistency properties for both settings.
+
+\hypertarget{sec:nonintersect}{%
+\subsection{Algorithms for Nonintersecting Manifolds}\label{sec:nonintersect}}
+
+In this section, we consider the following setting: Suppose that each community \(C_k\) is represented by a closed manifold \(\mathcal{M}_k\) in the latent space of a RDPG or GRDPG.
+Define \(\delta = \min\limits_{k \neq \ell} \min\limits_{x \in \mathcal{M}_k, y \in \mathcal{M}_\ell} \|x - y\|\), the minimum distance between two manifolds.
+Furthermore, assume that \(\delta > 0\), i.e., the manifolds do not intersect.
+
+In the original latent space, if the vectors on each manifold is sufficiently dense, it is possible to construct for each manifold \(\mathcal{M}_k\) an \(\eta_k\)-neighborhood graph for some \(\eta_k > 0\).
+
+\hypertarget{sec:intersect}{%
+\subsection{Algorithms for Intersecting Manifolds}\label{sec:intersect}}
 

### 2022-10-05 draft/draft.Rmd
still writing

@@
 
+In the Bernoulli random graph, the edge probability matrix $P$ is of size $n \times n$, but when $K \ll n$, the structure of $P$ is much simpler than its dimensions imply. 
+In particular, $P$ is rank $K$ in the SBM and DCBM and rank $K^2$ in the PABM. 
+This property lends these block models to be analyzed as another family of Bernoulli random graphs with low rank called the random dot product graph (RDPG) and generalized random dot product graph (GRDPG).
+In this model, $P$ is decomposed into an outer product of two low-rank data matrices.
+
 \begin{definition}[Generalized random dot product graph]
@@
 Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \begin{bmatrix} \cos \frac{\pi}{2} t & \sin \frac{\pi}{2} t \end{bmatrix}^\top$ for $t \in [0, 1]$.
-For simplicity, we set $\rho_n \equiv 1$. 
 Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$.
@@
 
-In the original latent space, if the vectors on each manifold is sufficiently dense, it is possible to construct for each manifold $\mathcal{M}_k$ an $\eta_k$-neighborhood graph for some $\eta_k > 0$.
+In the original latent space, it is possible to construct for each manifold $\mathcal{M}_k$ a connected $\eta_k$-neighborhood graph for some $\eta_k > 0$. 
+If the subsample of vectors on each manifold is sufficiently dense such that $\max_k \eta_k = \eta < \delta$, then an $\eta$-neighborhood graph on the entire sample results in a graph with $K$ disconnected subgraphs that map onto each manifold exactly.
+This is equivalent to applying single linkage clustering on the ASE and cutting the dendrogram at $K$ clusters (see algorithm \ref{alg:ase-singlelink}).
+The remainder of this section explores under which conditions these criteria are met for the latent configuration, in which the latent vectors lie exactly on manifolds, as well as for the ASE of the GRDPG sampled from these latent configurations, in which the vectors lie approximately on manifolds with some noise that approaches zero asymptotically.
+
+\begin{algorithm}[h]
+  \label{alg:ase-singlelink}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \setstretch{1.5}
+  \KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $(p, q)$.}
+  \KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$.}
+  Compute $\hat{X}$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
+  Apply single linkage clustering with $K$ communities on $\hat{X}$.\;
+  \caption{ASE clustering for nonintersecting communities.}
+\end{algorithm}
+
+\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting manifolds]
+\label{theorem:nonintersect-grdpg}
+Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$.
+Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
+Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
+Then for some $C > 0$ and any $\eta \in (0, C \delta)$, $\lim\limits_{n \to \infty} P(A_n(\eta)) = 1$. 
+\end{theorem}
+
+If the manifolds are one-dimensional, then a more precise rate of convergence can be derived.
+
+\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting curves]
+\label{theorem:nonintersect-curves}
+\end{theorem}
 
@@
 
+\begin{lemma}
+\label{lemma:no-noise}
+Let $x_1, ..., x_n$ be drawn from $K$ compact, connected manifolds $\mathcal{M}_1$, ..., $\mathcal{M}_K$ each with probability measures $F_1$, ..., $F_K$, and the manifolds are separated by distance at least $\delta > 0$.
+Suppose that for any $\epsilon > 0$ and $x$ drawn from each $F_k$, on $\mathcal{M}_k$, $F(B(x, \epsilon)) > 0$ where $B(x, \epsilon)$ is the open ball of radius $\epsilon$ centered at $x$. 
+Let $E_n(\eta)$ denote the event that an $\eta$-neighborhood graph constructed from $x_1, ..., x_n$ is comprised of exactly $K$ disjoint subgraphs that map to each of the $K$ manifolds. 
+Then if each $n_k \to \infty$ as $n \to \infty$, $\lim\limits_{n \to \infty} P(E_n(\eta)) = 1$ for each $\eta \in (0, \delta)$. 
+\end{lemma}
+
+\begin{proof}
+It is clear that if $\eta \in (0, \delta)$, an $\eta$-neighborhood graph constructed from the sample will always consist of at least $K$ disjoint subgraphs for which no subgraph contains vertices belonging to points from two different manifolds.
+Then it is sufficient to show that for a sufficiently large $n$, any $\eta$-neighborhood graph (where $\eta \in (0, \delta)$) will achieve $E_n$.
+
+Define each $E_{n_k}^{(k)}(\eta)$ as the event that if a sub-sample of size $n_k$ drawn from manifold $\mathcal{M}_k$, every $x \in \mathcal{M}_k$ is within distance $\eta$ of some $x_j$ of the sub-sample.
+Then if $E_{n_k}^{(k)}(\eta)$ is true, the $\eta$-neighborhood graph results in a connected subgraph for points within the $k^{th}$ manifold. 
+By lemma 2 of \citet{trosset2020rehabilitating}, $P((E_{n_k}^{(k)}(\eta))^c) \leq \ell_k (1 - b_k)^{n_k}$ for some $\ell_k \in \mathbb{N}$ and $b_k \in (0, 1]$. 
+If each $E_{n_k}^{(k)}(\eta)$ is true, then $E_n$ is achieved, so $E_n(\eta) = \bigcap_k E_{n_k}^{(k)}(\eta))$. 
+$\bigcap_k E_{n_k}^{(k)}(\eta) = \Big( \bigcup_k (E_{n_k}^{(k)}(\eta))^c \Big)^c$, so it is sufficient to show $\lim\limits_{n \to \infty} P \big( \bigcup_k (E_{n_k}^{(k)}(\eta))^c \big) \to 0$.
+
+$$
+\begin{aligned}
+P \big( \bigcup_k(E_{n_k}^{(k)})^c \big) & \leq \sum_k P \big((E_{n_k}^{(k)})^c \big) \\
+& \leq \sum_k \ell_k (1 - b_k)^{n_k} \\
+& \leq K \ell_{\max} (1 - b_{\min})^{n_{\min}},
+\end{aligned}
+$$
+
+which tends to $0$ as $n \to \infty$. 
+\end{proof}
+
+\begin{proof}[Proof of theorem \ref{theorem:nonintersect-grdpg}]
+Define $E_n(\eta)$ as in lemma \ref{lemma:no-noise} for manifolds $Q_n(\mathcal{M}_1), ..., Q_n(\mathcal{M}_K)$ and each $e_i = \hat{x}_i - Q_n x_i$ where $\hat{x}_i$ is the $i^{th}$ embedding vector and $Q_n$ is some indefinite orthogonal transformation as in \citet{https://doi.org/10.1111/rssb.12509}.
+Since $Q_n$ is a linear map, for any $\eta \in (0, \|Q_n\| \delta)$, $P(E_n(\eta)) \to 1$ as $n \to \infty$. 
+Let $\epsilon_i = \|e_i\|$, $\epsilon = \max_i \epsilon_i$, and $C_n = \|Q_n\|$.
+
+$A_n(\eta)$ is true if $\eta < \min\limits_{k, \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \|\hat{x}_i - \hat{x}_j\|$, which is defined as event $D_n(\eta)$, and $\eta \geq \max\limits_k \max\limits_{x_i, x_j \in \mathcal{M}_k} \|\hat{x}_i - \hat{x}_j\|$, which is defined as event $\hat{E}_n(\eta)$.
+
+For any $x_i, x_j$ from different manifolds, $\|\hat{x}_i - \hat{x}_j\| \geq C_n \delta - 2 \epsilon$ if $2 \epsilon \leq C_n \delta$. 
+By theorem 3 of \citet{rubindelanchy2017statistical}, for some finite $M > 0$, $P \big(\epsilon < M \frac{\log^c n}{\sqrt{n}} \big) \to 1$ as $n \to \infty$, so $P(C_n \delta < 2 \epsilon) \leq P(C_n \delta < 2 M n^{1/2} \log^c n) \to 0$ since $C_n \delta > 0$.
+Then since $P(C_n \delta - 2 \epsilon > 0) \to 1$, there is an $\epsilon \in (0, C_n \delta - 2 \epsilon)$ with probability 1. 
+Thus, $P(D_n(\eta)) \to 1$.
+
+Then to show that $P(A_n) \to 1$:
+
+$$
+\begin{aligned}
+P(A_n) & = P(\hat{E}_n(\eta) \cap D_n) \\
+& = P((\hat{E}_n^c(\eta) \cup D_n^c)^c) \\
+& = 1 - P(\hat{E}_n^c(\eta) \cup D_n^c) \\
+& \geq 1 - P(\hat{E}_n^c(\eta)) - P(D_n^c) \\ 
+& = P(\hat{E}_n(\eta)) + P(D_n) - 1,
+\end{aligned}
+$$
+which tends toward $1$ as $n \to \infty$ since both $P(E_n(\eta))$ and $P(D_n)$ tend toward $1$ as $n \to \infty$. 
+\end{proof}
+
 \newpage

### 2022-10-05 draft/draft.tex
still writing

@@
 
+In the Bernoulli random graph, the edge probability matrix \(P\) is of size \(n \times n\), but when \(K \ll n\), the structure of \(P\) is much simpler than its dimensions imply.
+In particular, \(P\) is rank \(K\) in the SBM and DCBM and rank \(K^2\) in the PABM.
+This property lends these block models to be analyzed as another family of Bernoulli random graphs with low rank called the random dot product graph (RDPG) and generalized random dot product graph (GRDPG).
+In this model, \(P\) is decomposed into an outer product of two low-rank data matrices.
+
 \begin{definition}[Generalized random dot product graph]
@@
 
-In the original latent space, if the vectors on each manifold is sufficiently dense, it is possible to construct for each manifold \(\mathcal{M}_k\) an \(\eta_k\)-neighborhood graph for some \(\eta_k > 0\).
+In the original latent space, it is possible to construct for each manifold \(\mathcal{M}_k\) a connected \(\eta_k\)-neighborhood graph for some \(\eta_k > 0\).
+If the subsample of vectors on each manifold is sufficiently dense such that \(\max_k \eta_k = \eta < \delta\), then an \(\eta\)-neighborhood graph on the entire sample results in a graph with \(K\) disconnected subgraphs that map onto each manifold exactly.
+This is equivalent to applying single linkage clustering on the ASE and cutting the dendrogram at \(K\) clusters (see algorithm \ref{alg:ase-singlelink}).
+The remainder of this section explores under which conditions these criteria are met for the latent configuration, in which the latent vectors lie exactly on manifolds, as well as for the ASE of the GRDPG sampled from these latent configurations, in which the vectors lie approximately on manifolds with some noise that approaches zero asymptotically.
+
+\begin{algorithm}[h]
+  \label{alg:ase-singlelink}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \setstretch{1.5}
+  \KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $(p, q)$.}
+  \KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$.}
+  Compute $\hat{X}$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
+  Apply single linkage clustering with $K$ communities on $\hat{X}$.\;
+  \caption{ASE clustering for nonintersecting communities.}
+\end{algorithm}
+
+\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting manifolds]
+\label{theorem:nonintersect-grdpg}
+Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$.
+Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
+Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
+Then for some $C > 0$ and any $\eta \in (0, C \delta)$, $\lim\limits_{n \to \infty} P(A_n(\eta)) = 1$. 
+\end{theorem}
+
+If the manifolds are one-dimensional, then a more precise rate of convergence can be derived.
+
+\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting curves]
+\label{theorem:nonintersect-curves}
+\end{theorem}
 

### 2022-10-05 draft/draft.Rmd
fixed a broken reference link

@@
 
-## The Hierarchy of Block Models
+## A Hierarchy of Block Models
 
@@
 \begin{proof}[Proof of theorem \ref{theorem:nonintersect-grdpg}]
-Define $E_n(\eta)$ as in lemma \ref{lemma:no-noise} for manifolds $Q_n(\mathcal{M}_1), ..., Q_n(\mathcal{M}_K)$ and each $e_i = \hat{x}_i - Q_n x_i$ where $\hat{x}_i$ is the $i^{th}$ embedding vector and $Q_n$ is some indefinite orthogonal transformation as in \citet{https://doi.org/10.1111/rssb.12509}.
+Define $E_n(\eta)$ as in lemma \ref{lemma:no-noise} for manifolds $Q_n(\mathcal{M}_1), ..., Q_n(\mathcal{M}_K)$ and each $e_i = \hat{x}_i - Q_n x_i$ where $\hat{x}_i$ is the $i^{th}$ embedding vector and $Q_n$ is some indefinite orthogonal transformation as in \citet{rubindelanchy2017statistical}.
 Since $Q_n$ is a linear map, for any $\eta \in (0, \|Q_n\| \delta)$, $P(E_n(\eta)) \to 1$ as $n \to \infty$. 
@@
 
+\section*{Appendix B: Details on Fitting Bezier Curves to Noisy Data}
+
+\newpage
+
+\section*{Appendix C: Supplementary Materials}
+
+\newpage
+
 # References

### 2022-10-05 draft/draft.tex
fixed a broken reference link

@@
 
-\hypertarget{the-hierarchy-of-block-models}{%
-\subsection{The Hierarchy of Block Models}\label{the-hierarchy-of-block-models}}
+\hypertarget{a-hierarchy-of-block-models}{%
+\subsection{A Hierarchy of Block Models}\label{a-hierarchy-of-block-models}}
 
@@
 
+\section*{Appendix B: Details on Fitting Bezier Curves to Noisy Data}
+
+\newpage
+
+\section*{Appendix C: Supplementary Materials}
+
+\newpage
+
 \renewcommand\refname{References}

### 2022-10-12 draft/draft.Rmd
still writing

@@
 - \usepackage{comment}
+- \DeclareMathOperator*{\argmax}{arg\,max}
+- \DeclareMathOperator*{\argmin}{arg\,min}
 bibliography: bibliography.bib
@@
 \newcommand{\DCBM}{\mathrm{DCBM}}
-\newcommand{\argmin}{\mathrm{argmin}}
 ```
@@
 \label{eq:kmeans}
-W(x_1, ..., x_n) = \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
+\argmin_{z_1, ..., z_n} ~ W(z_1, ..., z_n; x_1, ..., x_n) = 
+\argmin_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
 \end{equation}
-in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments we wish to optimize for, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
-A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration labels are updated according to which centroid they are closest to and centroids are updated via the sample mean of each cluster.
-
-If instead of vectors in $\mathbb{R}^p$ we have a graph, it is not obvious how to translate $K$-means clustering to these data. 
+in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments which we wish to optimize, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
+A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.
+
+If we have a graph instead of vectors in $\mathbb{R}^p$, it is not obvious how to translate $K$-means clustering to these data. 
 In particular, there is no inherent notion of centroid or sample mean for graph data. 
 However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
-One way to adapt $K$-means to graph data is to use an equivalent objective function as equation~\eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
+One way to adapt $K$-means to graph data is to use an equivalent objective function as equation \eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
 $$
-\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
+\argmin_{z_1, ..., z_n} ~ \tilde{W}(z_1, ..., z_n; x_1, ..., x_n) =
+\argmin_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
 $$
 Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
-If they are vectors in Euclidean space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
+If they are vectors in Euclidean space, then we can use $d(x_i, x_j) = \|x_i - x_j\|_2$, 
 or if they are vertices on a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
@@
 Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
-One way to combine the graph distance approach and the graph embedding approach is to apply Lloyd's algorithm on an embedding that approximates graph distances \citep{vonLuxburg2007}. 
+In this context, a natural choice of embedding is one in which pairwise Euclidean distances between embedding vectors are (approximately) equal to pairwise graph distances between corresponding vertices.
+For example, the combinatorial Laplacian eigenmap is an embedding in which Euclidean distances between pairs of vectors in the embedding are equal to expected commute times between corresponding pairs of vertices in the graph \citep{vonLuxburg2007}.
 \end{example}
@@
 This network is represented by a graph $G = (V, E)$ with vertex set $V = \{v_1, ..., v_n\}$, representing the $n$ objects, and edge set $E$, representing the up to $n (n - 1) / 2$ pairwise relationships (assuming that there are no self-loops).
-The numeric representation of these data is in the form of *affinity matrix* $A \in \mathbb{R}^{n \times n}$ in which each $A_{ij}$ represents object $i$'s relationship to object $j$. 
+The numeric representation of these data is in the form of *affinity matrix* $A \in \mathbb{R}_+^{n \times n}$ in which each $A_{ij}$ represents object $i$'s relationship to object $j$. 
 We assume that the entries of $A$ represent affinities or similarities, i.e., the higher the value of $A_{ij}$, the stronger the relationship $i$ has to $j$. 
@@
 Then similar to how the edges are compiled into an adjacency matrix $A$, the edge probabilities can be compiled into an edge probability matrix $P \in [0, 1]^{n \times n}$. 
-This type of graph model is defined as a *Bernoulli graph* (also called an inhomogeneous $\ER$ graph). 
-If $A$ is the adjacency matrix of a Bernoulli graph with edge probability matrix $P$, we denote $A \sim \BG(P)$. 
+This type of graph model is defined as a *Bernoulli random graph* (also called an inhomogeneous $\ER$ graph). 
+If $A$ is the adjacency matrix of a Bernoulli random graph with edge probability matrix $P$, we denote $A \sim \BG(P)$. 
 If the vertices and edge probabilities are sampled as a sequence of $n$ i.i.d. random variables from probability distribution $F$, then we denote $A \sim \BG(F, n)$.
 
-Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
+Statistical inference is not possible on a general Bernoulli random graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
 Additional structure must be introduced. 
-One such structured Bernoulli graph model is the $\ER$ graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
+One such structured Bernoulli random graph model is the $\ER$ graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
 
@@
 $\ER$ graphs lie on the opposite end of the spectrum in that they restrict the model to a single parameter, which does not result in a very interesting model or reflect many networks observed in real data.
-Much work has been done in developing various Bernoulli graph models that strike a balance between structure and flexibility. 
+Much work has been done in developing various Bernoulli random graph models that strike a balance between structure and flexibility. 
 Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called *block models*, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called *latent space models*.
-
-**[Maybe insert something here about the exponential family of random graph models and how, like mixture distributions, block models do not fit within this family?]**
 
@@
 Network analysis is often concerned with community detection. 
-This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices $i$ and $j$ depend on the communities to which vertices $v_i$ and $v_j$ belong.
-More formally, one assigns each vertex $v_i$ a community label $z_i$ and assumes a Bernoulli graph in which $P_{ij} = g(z_i, z_j, \theta_i, \theta_j)$ for some function $g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]$ and $\theta_i \in \mathbb{R}^d$ is a vector of real-valued parameters associated with each vertex.
+This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices $v_i$ and $v_j$ depend on the community assignments $z_i$ and $z_j$.
+More formally, one assigns each vertex $v_i$ a community label $z_i \in \{1, ..., K\}$ and assumes a Bernoulli random graph in which $P_{ij} = g(z_i, z_j, \theta_i, \theta_j)$ for some function $g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]$ and $\theta_i \in \mathbb{R}^d$ is a vector of real-valued parameters associated with each vertex.
 Within the context of this work, we will call such models *block models*. 
 This type of model framework allows us to frame the goal of community detection as a statistical inference problem: 
-identify the true community labels, up to permutation of the labels.
+come up with estimators for the true community labels, up to permutation.
 
@@
 The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities $(k, \ell)$ has a constant edge probability $\theta_{k \ell}$.
-We now give the formal definition of the SBM within the context of Bernoulli graphs:
+We now give the formal definition of the SBM within the context of Bernoulli random graphs:
 
@@
 \label{def:sbm}
-Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
 Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$). 
@@
 
-An extension to these include Bayesian models, e.g., each $\theta_{k \ell}$ is sampled from a distribution with support $[0, 1]$.
+An extension to these include Bayesian models in which there are prior distributions for the parameters, e.g., each $\theta_{k \ell}$ has a prior distribution with support $[0, 1]$, such as $\theta_{k \ell} \sim \Betadist(a_{k \ell}, b_{k \ell})$, and the mixture parameters $\alpha$ has a prior distribution on the $K-1$ simplex, such as $\alpha \sim \mathrm{Dirichlet}(\beta_1, ..., \beta_K)$.
 
@@
 In most conceptions of block models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}. 
+This reflects the general intuition that pairs within the same community are more likely to be connected than pairs between two different communities. 
 For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the network appears to confirm this.
@@
 a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
+This corresponds to the high-level understanding of assortativity and disassortativity based on whether within or between community edge probabilities are greater. 
 \end{example}
@@
 The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard. 
-Because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the expectation maximization (EM) algorithm \citep{10.2307/2984875}.
+But because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the widely studied expectation maximization (EM) algorithm \citep{10.2307/2984875}.
 However, due to the $z_{ik} z_{j \ell}$ interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}. 
@@
 
-So far, we have described *dense* Bernoulli graphs. 
+So far, we have described *dense* Bernoulli random graphs. 
 For such models, as the sample size $n$ increases, the expected degree of each node grows linearly. 
@@
 In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label. 
-The Degree Corrected Block Model (DCBM) and Popularity Adjusted Block Model (PABM) are generalizations of the SBM that address these limitations. 
+The degree corrected block model (DCBM) and popularity adjusted block model (PABM) are generalizations of the SBM that address these limitations. 
 In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label.
@@
 \label{def:dcbm}
-Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
 Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM. 
@@
 \end{definition}
-
-Showing explicitly that the SBM is a special case of the DCBM only requires us to set $\omega$ as a constant vector, e.g., $\omega = \begin{bmatrix} 1 & \cdots & 1 \end{bmatrix}^\top$.
 
@@
 
+The DCBM allows for each vertex to have its own expected degree while still maintaining community structure. 
+However, it still assumes that each vertex within a community has the same relative propensity to connect to each community. 
+For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that everyone is ten times more likely to connect with a member of the opposite sex than to a member of the same sex.
+The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community. 
+In the dating network example, the PABM allows us to model each individual's preferences instead of using population level averages.
+
 \begin{definition}[Popularity adjusted block model]
 \label{def:pabm}
-Let $G = (V, E)$ be an undirected Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $G = (V, E)$ be an undirected Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
 Let $K \geq 1$ be an integer that describes the number of communities, and let 
@@
 
+\begin{example}[Dating network as a PABM]
+\label{ex:pabm-dating}
+In examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, we modeled an online dating network as a SBM and a DCBM, in which the vertices correspond to individuals, the communities correspond to genders, and the edges represent whether the online dating service successfully matched pairs of individuals.
+In these models, it is assumed that all individuals have the same sexual orientation. 
+Using the parameters specified in these examples, we assume that each individual is 10 times more likely to match with a member of the opposite sex compared to a member of the same sex. 
+The PABM allows us to model each individual's sexual orientation. 
+Figure \ref{fig:dating-pabm} is a visualization of the online dating network modeled as a PABM in which the popularity parameters are drawn as $\lambda_{ik} \sim \Betadist(1, 6)$ if $z_i = k$ and $\lambda_{ik} \sim \Betadist(2, 2)$ if $z_i \neq k$.
+Like in the SBM and DCBM examples, the average edge probability between communities is approximately ten times greater than the average edge probability within communities.
+
+```{r dating-pabm, fig.width = 4, , fig.height = 2, fig.cap = 'A dating network modeled as a PABM. This model is disassortative.', cache = TRUE}
+set.seed(12345)
+n <- 2 ** 6
+Az <- draw.pabm.beta.2(n, 1, 6, 2, 2, .5)
+A <- Az$A
+z <- Az$z
+z[z == 1] <- 'F'
+z[z == 2] <- 'M'
+
+ggraph(A, layout = 'auto') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = TRUE) + 
+  labs(colour = 'Gender') + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
+\end{example}
+
 \begin{remark}
@@
 Likewise, it is not immediately obvious how to approach the estimation of the $\theta_{k \ell}$'s, $\omega_i$'s, or $\lambda_{ik}$'s due to their non-uniqueness. 
-Applying an EM algorithm to example \ref{ex:dcbm-dating} results in 84\% accuracy but fails to capture the original (which are not necessarily ``true'') parameters. 
+Applying an EM algorithm to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying a similar EM algorithm to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
 \end{remark}
-
-\begin{example}[Dating network as a PABM]
-\label{ex:pabm-dating}
-\end{example}
 
@@
 This property lends these block models to be analyzed as another family of Bernoulli random graphs with low rank called the random dot product graph (RDPG) and generalized random dot product graph (GRDPG).
-In this model, $P$ is decomposed into an outer product of two low-rank data matrices.
+In these models, $P$ is decomposed into an outer product of two low-rank data matrices.
 

### 2022-10-12 draft/draft.tex
still writing

@@
 \usepackage{comment}
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
 \ifLuaTeX
@@
 \newcommand{\DCBM}{\mathrm{DCBM}}
-\newcommand{\argmin}{\mathrm{argmin}}
 
@@
 \label{eq:kmeans}
-W(x_1, ..., x_n) = \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
+\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ W(z_1, ..., z_n; x_1, ..., x_n) = 
+\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
 \end{equation}
-in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments we wish to optimize for, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
-A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration labels are updated according to which centroid they are closest to and centroids are updated via the sample mean of each cluster.
-
-If instead of vectors in $\mathbb{R}^p$ we have a graph, it is not obvious how to translate $K$-means clustering to these data. 
+in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments which we wish to optimize, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
+A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.
+
+If we have a graph instead of vectors in $\mathbb{R}^p$, it is not obvious how to translate $K$-means clustering to these data. 
 In particular, there is no inherent notion of centroid or sample mean for graph data. 
 However, there are inherent notions of distances among vertices, such as shortest path distance, expected commute time, or resistance distance. 
-One way to adapt $K$-means to graph data is to use an equivalent objective function as equation~\eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
+One way to adapt $K$-means to graph data is to use an equivalent objective function as equation \eqref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:
 $$
-\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
+\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \tilde{W}(z_1, ..., z_n; x_1, ..., x_n) =
+\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
 $$
 Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
-If they are vectors in Euclidean space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
+If they are vectors in Euclidean space, then we can use $d(x_i, x_j) = \|x_i - x_j\|_2$, 
 or if they are vertices on a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
@@
 This network is represented by a graph \(G = (V, E)\) with vertex set \(V = \{v_1, ..., v_n\}\), representing the \(n\) objects, and edge set \(E\), representing the up to \(n (n - 1) / 2\) pairwise relationships (assuming that there are no self-loops).
-The numeric representation of these data is in the form of \emph{affinity matrix} \(A \in \mathbb{R}^{n \times n}\) in which each \(A_{ij}\) represents object \(i\)'s relationship to object \(j\).
+The numeric representation of these data is in the form of \emph{affinity matrix} \(A \in \mathbb{R}_+^{n \times n}\) in which each \(A_{ij}\) represents object \(i\)'s relationship to object \(j\).
 We assume that the entries of \(A\) represent affinities or similarities, i.e., the higher the value of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\).
@@
 Then similar to how the edges are compiled into an adjacency matrix \(A\), the edge probabilities can be compiled into an edge probability matrix \(P \in [0, 1]^{n \times n}\).
-This type of graph model is defined as a \emph{Bernoulli graph} (also called an inhomogeneous \(\ER\) graph).
-If \(A\) is the adjacency matrix of a Bernoulli graph with edge probability matrix \(P\), we denote \(A \sim \BG(P)\).
+This type of graph model is defined as a \emph{Bernoulli random graph} (also called an inhomogeneous \(\ER\) graph).
+If \(A\) is the adjacency matrix of a Bernoulli random graph with edge probability matrix \(P\), we denote \(A \sim \BG(P)\).
 If the vertices and edge probabilities are sampled as a sequence of \(n\) i.i.d. random variables from probability distribution \(F\), then we denote \(A \sim \BG(F, n)\).
 
-Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
+Statistical inference is not possible on a general Bernoulli random graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
 Additional structure must be introduced.
-One such structured Bernoulli graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
+One such structured Bernoulli random graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
 
@@
 \(\ER\) graphs lie on the opposite end of the spectrum in that they restrict the model to a single parameter, which does not result in a very interesting model or reflect many networks observed in real data.
-Much work has been done in developing various Bernoulli graph models that strike a balance between structure and flexibility.
+Much work has been done in developing various Bernoulli random graph models that strike a balance between structure and flexibility.
 Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called \emph{block models}, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called \emph{latent space models}.
-
-\textbf{{[}Maybe insert something here about the exponential family of random graph models and how, like mixture distributions, block models do not fit within this family?{]}}
 
@@
 Network analysis is often concerned with community detection.
-This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices \(i\) and \(j\) depend on the communities to which vertices \(v_i\) and \(v_j\) belong.
-More formally, one assigns each vertex \(v_i\) a community label \(z_i\) and assumes a Bernoulli graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
+This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices \(v_i\) and \(v_j\) depend on the community assignments \(z_i\) and \(z_j\).
+More formally, one assigns each vertex \(v_i\) a community label \(z_i \in \{1, ..., K\}\) and assumes a Bernoulli random graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
 Within the context of this work, we will call such models \emph{block models}.
 This type of model framework allows us to frame the goal of community detection as a statistical inference problem:
-identify the true community labels, up to permutation of the labels.
+come up with estimators for the true community labels, up to permutation.
 
@@
 The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities \((k, \ell)\) has a constant edge probability \(\theta_{k \ell}\).
-We now give the formal definition of the SBM within the context of Bernoulli graphs:
+We now give the formal definition of the SBM within the context of Bernoulli random graphs:
 
@@
 
-An extension to these include Bayesian models, e.g., each \(\theta_{k \ell}\) is sampled from a distribution with support \([0, 1]\).
+An extension to these include Bayesian models in which there are prior distributions for the parameters, e.g., each \(\theta_{k \ell}\) has a prior distribution with support \([0, 1]\), such as \(\theta_{k \ell} \sim \Betadist(a_{k \ell}, b_{k \ell})\), and the mixture parameters \(\alpha\) has a prior distribution on the \(K-1\) simplex, such as \(\alpha \sim \mathrm{Dirichlet}(\beta_1, ..., \beta_K)\).
 
@@
 In most conceptions of block models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}.
+This reflects the general intuition that pairs within the same community are more likely to be connected than pairs between two different communities.
 For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the network appears to confirm this.
@@
 The SBM is a type of mixture model, and like most classical mixture models and clustering problems, likelihood maximization is NP-hard.
-Because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the expectation maximization (EM) algorithm \citep{10.2307/2984875}.
+But because the SBM is a mixture model, a candidate algorithm for finding a (local) maximum of the likelihood is the widely studied expectation maximization (EM) algorithm \citep{10.2307/2984875}.
 However, due to the \(z_{ik} z_{j \ell}\) interaction terms, the expectation step cannot be solved in closed form \citep{kolaczyk2014statistical}.
@@
 
-So far, we have described \emph{dense} Bernoulli graphs.
+So far, we have described \emph{dense} Bernoulli random graphs.
 For such models, as the sample size \(n\) increases, the expected degree of each node grows linearly.
@@
 In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label.
-The Degree Corrected Block Model (DCBM) and Popularity Adjusted Block Model (PABM) are generalizations of the SBM that address these limitations.
+The degree corrected block model (DCBM) and popularity adjusted block model (PABM) are generalizations of the SBM that address these limitations.
 In particular, the DCBM allows each vertex to have an expected degree independent of its label, and the PABM allows each vertex to have edge probabilities to each community independent of its own label.
@@
 
-Showing explicitly that the SBM is a special case of the DCBM only requires us to set \(\omega\) as a constant vector, e.g., \(\omega = \begin{bmatrix} 1 & \cdots & 1 \end{bmatrix}^\top\).
-
 \begin{example}[Dating network as a disassortative DCBM]
@@
 
+The DCBM allows for each vertex to have its own expected degree while still maintaining community structure.
+However, it still assumes that each vertex within a community has the same relative propensity to connect to each community.
+For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that everyone is ten times more likely to connect with a member of the opposite sex than to a member of the same sex.
+The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community.
+In the dating network example, the PABM allows us to model each individual's preferences instead of using population level averages.
+
 \begin{definition}[Popularity adjusted block model]
 \label{def:pabm}
-Let $G = (V, E)$ be an undirected Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
+Let $G = (V, E)$ be an undirected Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
 Let $K \geq 1$ be an integer that describes the number of communities, and let 
@@
 
+\begin{example}[Dating network as a PABM]
+\label{ex:pabm-dating}
+In examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, we modeled an online dating network as a SBM and a DCBM, in which the vertices correspond to individuals, the communities correspond to genders, and the edges represent whether the online dating service successfully matched pairs of individuals.
+In these models, it is assumed that all individuals have the same sexual orientation. 
+Using the parameters specified in these examples, we assume that each individual is 10 times more likely to match with a member of the opposite sex compared to a member of the same sex. 
+The PABM allows us to model each individual's sexual orientation. 
+Figure \ref{fig:dating-pabm} is a visualization of the online dating network modeled as a PABM in which the popularity parameters are drawn as $\lambda_{ik} \sim \Betadist(1, 6)$ if $z_i = k$ and $\lambda_{ik} \sim \Betadist(2, 2)$ if $z_i \neq k$.
+Like in the SBM and DCBM examples, the average edge probability between communities is approximately ten times greater than the average edge probability within communities.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/dating-pabm-1}
+
+}
+
+\caption{A dating network modeled as a PABM. This model is disassortative.}\label{fig:dating-pabm}
+\end{figure}
+\end{example}
+
 \begin{remark}
@@
 This property lends these block models to be analyzed as another family of Bernoulli random graphs with low rank called the random dot product graph (RDPG) and generalized random dot product graph (GRDPG).
-In this model, \(P\) is decomposed into an outer product of two low-rank data matrices.
+In these models, \(P\) is decomposed into an outer product of two low-rank data matrices.
 
@@
 \(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
-\(c_i = \argmin_{c} \|c\|_1\) subject to \(x_i = X^\top c\) and
+\(c_i = \mathop{\mathrm{arg\,min}}_{c} \|c\|_1\) subject to \(x_i = X^\top c\) and
 \(c^{(i)} = 0\), where \(c^{(i)}\) is the \(i^{th}\) entry of \(c\), is solved for each \(i \in [n]\).
@@
 \label{eq:ssc}
-c_i = \argmin_c \frac{1}{2} \|x_i - X_{-i}^\top c\|^2_2 + \vartheta \|c\|_1
+c_i = \mathop{\mathrm{arg\,min}}_c \frac{1}{2} \|x_i - X_{-i}^\top c\|^2_2 + \vartheta \|c\|_1
 \end{equation}

### 2022-10-19 draft/draft.Rmd
almost done

@@
 urlcolor: blue
+language:
+  label:
+    rem: 'Remark '
 header-includes:
@@
 ```{r}
-import::from(magrittr, `%>%`)
+import::from(magrittr, `%>%`, `%<>%`)
+import::from(foreach, foreach, `%dopar%`)
 library(ggplot2)
@@
 source('~/dev/pabm-grdpg/functions.R')
+source('~/dev/manifold-block-models/functions.R')
+
+doMC::registerDoMC(parallel::detectCores())
 ```
@@
 \newcommand{\DCBM}{\mathrm{DCBM}}
+\newcommand{\rank}{\mathrm{rank}}
+\newcommand{\MBM}{\mathrm{MBM}}
 ```
@@
 However, it still assumes that each vertex within a community has the same relative propensity to connect to each community. 
-For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that everyone is ten times more likely to connect with a member of the opposite sex than to a member of the same sex.
+For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex.
 The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community. 
@@
 Figure \ref{fig:dating-pabm} is a visualization of the online dating network modeled as a PABM in which the popularity parameters are drawn as $\lambda_{ik} \sim \Betadist(1, 6)$ if $z_i = k$ and $\lambda_{ik} \sim \Betadist(2, 2)$ if $z_i \neq k$.
-Like in the SBM and DCBM examples, the average edge probability between communities is approximately ten times greater than the average edge probability within communities.
+Like in the SBM and DCBM examples, the average edge probability between genders is approximately ten times that of the average edge probability within genders, but unlike in the SBM and DCBM examples, each individual has its own affinity toward each gender.
 
@@
 For example, in the case of the DCBM, doubling each $\theta_{k \ell}$ and dividing each $\omega_i$ by $\sqrt{2}$ results in the same edge probability matrix. 
-Nevertheless, a naive EM type algorithm similar to the one in section \ref{sec:sbm} results in high community detection accuracy for most networks, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees.
+\citet{Karrer_2011} suggest imposing the constraint $\sum_{i : z_i = k} \omega_i = 1$ for each $k \in \{1, ..., K\}$, but other similar constraints can be used to enforce uniqueness, such as $\max_i \omega_i = 1$ and $\min_i \omega_i = 0$.
+Despite the lack of uniqueness without constraints, a naive EM type algorithm similar to the one in section \ref{sec:sbm} results in high community detection accuracy for most realizations of the DCBM and PABM, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees.
 Likewise, it is not immediately obvious how to approach the estimation of the $\theta_{k \ell}$'s, $\omega_i$'s, or $\lambda_{ik}$'s due to their non-uniqueness. 
-Applying an EM algorithm to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying a similar EM algorithm to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
+\citet{noroozi2019estimation} suggest focusing estimation on the edge probability matrix $P$ instead of the individual parameters to avoid this ambiguity. 
+Applying an EM algorithm similar to algorithm \ref{algo:sbm-em} adapted for the DCBM to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying a similar EM algorithm adapted for the PABM to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
 \end{remark}
@@
 
+The PABM, DCBM, and SBM are nested models, that is, the SBM is a special case of the DCBM which in turn is a special case of the SBM. 
+In this section, these connections are made explicit, and we explore the implications for modeling. 
+For ease of notation, throughout this section the rows and columns of $P$ are assumed to be organized by community in increasing order, i.e., the first $n_1$ rows and columns of $P$ correspond to community $1$, the next $n_2$ rows and columns correspond to community $2$, etc.
+We use the notation $P^{(k \ell)} = (P^{(\ell k)})^\top$ to denote the block of $P$ that corresponds to the edges between communities $k$ and $\ell$. 
+In section \@ref(sec:pabm-grdpg), we will deal with this notation and convention more carefully. 
+For additional details on the connections among the SBM, DCBM, and PABM, refer to \citet{Noroozi2022}.
+
+To see that the SBM is a special case of the DCBM, we can simply restrict the degree parameters $\omega_i$ to all be a constant value, e.g., $\omega_i \equiv 1$. 
+But in order to better understand the relationship between these two models, we will break down the models further.
+
+Define $B$ as the $K \times K$ matrix of edge probabilities between pairs of communities in the SBM. 
+Then $B_{k \ell} = \theta_{k \ell}$ and $P_{ij} = \theta_{z_i, z_j} = B_{z_i, z_j}$. 
+We can further show that block $P^{(k \ell)}$ can be written as $P^{(k \ell)} = e^{(k)} B_{k \ell} (e^{(\ell)})^\top$ where $e^{(k)}$ is the $1$-vector of dimension $k$. 
+Since $B_{k \ell}$ is a scalar, $P^{(k \ell)} = e^{(k)} B_{k \ell} (e^{(\ell)})^\top$ is a singular value decomposition of a rank 1 matrix and thus each block $P^{(k \ell)}$ is rank 1. 
+Then we can see that the entire edge probability matrix can be written as 
+$$P = Z B Z^\top$$ 
+and $Z$ is an $n \times K$ matrix in which $Z_{ik} = I(z_i = k) = z_{ik}$. Thus, if $B$ is full rank, i.e., $\rank(B) = K$, then $P$ is also rank $K$.
+
+To extend the SBM to the DCBM, we replace $e^{(k)}$ with the vector $\omega^{(k)}$ which contains the degree parameters of vertices in community $k$. 
+Then for the DCBM, $P^{(k \ell)} = \omega^{(k)} B_{k \ell} (\omega^{(\ell)})^\top$. 
+Similar to the SBM, this is a singular value decomposition of a rank 1 matrix. 
+The full edge probability matrix can be written as 
+$$P = (\Omega Z) B (\Omega Z)^\top$$ 
+for $\Omega = \diag(\omega)$. 
+Since $\Omega$ is diagonal, we again have that $\rank(P) = K$ (again, assuming that $B$ is full rank). 
+Thus, the dimensionality of the DCBM is the same as the SBM, and to restrict the DCBM to the PABM, let $\Omega = I$.
+
+To extend the DCBM to the PABM, we note that the $k \ell$^th^ block of the edge probability matrix can be written as $P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top$ where $\lambda^{(k \ell)} \in \mathbb{R}^{n_k}$ is the vector of popularity parameters belonging to vertices in community $k$ that point to community $\ell$, i.e., $\lambda^{(k \ell)}_i = \lambda_{i + \sum_{s=1}^{k-1} n_s, \ell}$.
+Similar to how $\omega^{(k)}$ is a sub-vector of $\omega$ 
+To match this decomposition with the one for the DCBM, we can fix each $B_{k \ell} \equiv 1$ and write $P^{(k \ell)} = \lambda^{(k \ell)} B_{k \ell} (\lambda^{(\ell k)})^\top$. 
+Restricting each $\lambda^{(k \ell)} = \omega^{(k)}$ reduces the PABM to the DCBM. 
+To write out the full edge probability matrix of the PABM, let $\Lambda^{(k)} =\begin{bmatrix} \lambda^{(k1)} & \cdots & \lambda^{(kK)} \end{bmatrix}$ and $X = \blockdiag(\Lambda^{(1)}, ..., \Lambda^{(K)})$.
+Then for some $K^2 \times K^2$ permutation matrix $\Pi$, 
+$$P = X \Pi X^\top.$$ 
+From this decomposition, we can see that unlike the SBM and DCBM, $\rank(P) = \rank(X) = K^2$. 
+We will explore this decomposition of the edge probability matrix of the PABM in more detail in section \@ref(sec:pabm-grdpg).
+
+The full hierarchy of nested block models is summarized in table \@ref(tab:hierarchy).
+
+```{r hierarchy, results = 'asis'}
+blockmodel.df <- dplyr::tibble(
+  Model = c('SBM', 'DCBM', 'PABM'),
+  Parameters = c('$K (K-1) / 2$', '$K (K-1) / 2 + n$', '$n K$'),
+  Dimensions = c('$K$', '$K$', '$K^2$'),
+  `Matrix Decomposition` = c('$P = Z B Z^\\top$', '$P = (\\Omega Z) B (\\Omega Z)^\\top$', '$P = X \\Pi X^\\top$'),
+  `Reduction to simpler model` = c('', '$\\omega = e$', '$\\lambda_{ik} = \\omega_i \\sqrt{\\theta_{z_i, k}}$')
+)
+knitr::kable(blockmodel.df, 'latex', 
+             escape = FALSE,
+             caption = 'Summary of an hierarchical or nested view of block momdels.')
+```
+
 \newpage
 
-## Random Dot Product Graphs and Generalized Random Dot Product Graphs {#sec:grdpg}
+## The Generalized Random Dot Product Graph {#sec:grdpg}
 
@@
 
-\begin{definition}[Adjacency spectral embedding]
-\label{def:ase}
-\end{definition}
-
 \begin{definition}[Indefinite orthogonal group]
-\label{def:indef-ortho-group}
+\label{def:indefinite}
+The indefinite orthogonal group with signature $(p, q)$ is
+the set $\{Q \in \mathbb{R}^{d \times d} \colon Q I_{p, q} Q^{\top} = I_{p, q}\}$,
+denoted as $\mathbb{O}(p, q)$. Here $d = p + q$. 
 \end{definition}
@@
 \end{remark}
+
+Now that we defined the GRDPG as a generative model for Bernoulli random graphs, 
+we need a method for estimation. 
+We can use Adjacency Spectral Embedding (ASE) \citep{doi:10.1080/01621459.2012.699795} to recover the latent vectors of a GRDPG. 
+This procedure consists of taking the spectral decomposition of $A$ and keeping the $p + q$ largest eigenvalues (in modulus) and corresponding eigenvectors of $A$.
+
+\begin{definition}[Adjacency spectral embedding]
+\label{def:ase}
+Let $A$ be an $n
+\times n$ adjacency matrix and denote the
+eigendecomposition of $A$ as
+$$A = \sum_{i} \hat{\lambda}_i \hat{v}_i \hat{v}_i^{\top}, \quad
+\text{where} \,\,
+|\hat{\lambda}|_1 \geq |\hat{\lambda}_2| \geq \dots \geq
+|\hat{\lambda}_n|.$$
+Let $\hat{D}$ be a diagonal matrix whose diagonal entries are
+the eigenvalues $\{\hat{\lambda}_i\}_{i=1}^{d}$ arranged in decreasing values (not in
+decreasing modulus) and let $\hat{V}$ be the $n \times d$ matrix whose
+columns are the corresponding eigenvectors $\{\hat{v}_i\}_{i=1}^{d}$
+arranged in the same order as the diagonal entries of $\hat{D}$. Now
+define $\hat{X} = \hat{V} |\hat{D}|^{1/2}$ where the $|\cdot|$
+operation is applied elementwise to the entries of $\hat{D}$. Then
+$\hat{X}$ serves as an estimate of $X$, up to some non-identifiable
+orthogonal transformation $Q$ as described in
+remark~\ref{rem:non_identifiable} and definition~\ref{def:indefinite};
+see \citet{rubindelanchy2017statistical} for further details.
+\end{definition}
+
+\citet{doi:10.1080/01621459.2012.699795} and \citet{rubindelanchy2017statistical} showed that the ASE is consistent in estimating the latent vectors, up to an indefinite orthogonal transformation. 
+More specifically, for sparsity parameters such that $n \rho_n = \omega (\log^{4 c} n)$ for some constant $c > 1$ and indefinite orthogonal transformation $Q \in \mathbb{O}(p, q)$, 
+$$\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)$$
+where $x^\top$ and $\hat{x}^\top$ are the rows of $\hat{X}$ (from the ASE) and $X$, respectively.
+
+\begin{example}
+Suppose that latent vectors for an RDPG are drawn from the curve 
+$x(t) = \begin{bmatrix} t^2 & 2 t (t-1) \end{bmatrix}^\top$, $t \in [0, 1]$. 
+Let $x_i = x(t_i)$. 
+Then for each $i < j$, the edges are drawn as $A_{ij} \indep \Bernoulli(x_i^\top x_j)$. 
+If after observing $A$, we wish to estimate the curve, the ASE provides an estimator in which the estimated latent vectors converge to the curve, up to an orthogonal transformation (figure \ref{fig:ase-example}).
+
+```{r ase-example, fig.length = 8, fig.height = 2, fig.cap = '200 points drawn from a curve (left), and the ASE of an adjacency matrix drawn from this curve as an RDPG (right).
+The red line is the original curve rotated to match the ASE.'}
+n <- 200
+t. <- runif(n)
+x <- t. ^ 2
+y <- 2 * t. * (1 - t.)
+latent.vec.plot <- ggplot() + 
+  geom_point(aes(x = x, y = y)) + 
+  coord_fixed() + 
+  theme_bw() + 
+  labs(x = NULL, y = NULL)
+
+X <- cbind(x, y)
+P <- X %*% t(X)
+A <- draw.graph(P)
+Y <- embedding(P, 2, 0)
+Xhat <- embedding(A, 2, 0)
+ase.plot <- ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2])) + 
+  geom_line(aes(x = Y[, 1], y = Y[, 2]), colour = 'red') + 
+  coord_fixed() + 
+  theme_bw() + 
+  labs(x = NULL, y = NULL)
+
+gridExtra::grid.arrange(latent.vec.plot, ase.plot, nrow = 1)
+```
+
+\end{example}
 
@@
 
+For each simulation, community labels are drawn from a multinomial
+distribution, the popularity vectors \(\{\lambda^{(k \ell)}\}_K\) are drawn
+from two types of joint distributions depending on whether \(k =
+\ell\) or $k \not = \ell$. The edge probability matrix \(P\) is constructed using the popularity
+vectors and finally the adjacency matrix \(A\)
+is drawn \(A \sim \mathrm{Bernoulli}(P)\). OSC (algorithm \ref{alg:osc}) is then used for community detection, and this
+method is compared against (1) SSC using the spectral embedding $\hat{V}$
+(algorithm \ref{alg:ssc}), (2) SSC using the rows of the
+observed adjacency matrix $A$ as is done in \citet{noroozi2019estimation}
+and (3) modularity maximization (MM) as is done in
+\citet{307cbeb9b1be48299388437423d94bf1}. We denote the two SSC
+implementations using the rows of $A$ and using the spectral embedding
+of $A$ as SSC-A and SSC-ASE, respectively. 
+The parameters \(\vartheta\)
+that controls the sparsity for SSC-A and SSC-ASE were chosen via a preliminary cross-validation experiment. 
+In practice, $\vartheta$ that guarantee SDP (if it is possible for the particular simulated data) often result in more than $K$ disconnected subgraphs, so a smaller $\vartheta$ that does not guarantee SDP was chosen,
+and the final clustering step of SSC-A and SSC-ASE was done
+by fitting a Gaussian mixture model to the normalized Laplacian
+eigenmap embeddings \citep{belkin03:_laplac} of the affinity matrix \(B\).
+We also estimate the latent popularity vectors $\{\lambda^{(k \ell)}\}$
+by assuming that the true community labels are known and then apply
+algorithm \ref{alg:param_est}, and we compare this estimation method against an
+MLE-based estimator as described in \citet{noroozi2019estimation} and
+\citet{307cbeb9b1be48299388437423d94bf1}.
+
+Modularity maximization is NP-hard, so
+\citet{307cbeb9b1be48299388437423d94bf1} used the extreme points
+(EP) algorithm (\cite{le2016}) as a greedy
+relaxation of the optimization problem; the EP algorithm has a running
+time of $O(n^{K-1})$ where $n$ is the number of vertices in the graph
+and $K$ is the number of communities.
+For these simulations we instead replace the EP algorithm with the
+Louvain algorithm for modularity maximization,
+as the implementation of the EP algorithm in
+\citet{307cbeb9b1be48299388437423d94bf1} is too computationally expensive for \(K > 2\). For \(K = 2\), it
+was verified that the Louvain algorithm produces comparable results
+to EP-MM.
+
+For comparing methods, we define the community detection error as:
+\[L_c(\hat{\sigma}, \sigma; V) = \min_\pi \sum_i I(\pi \circ \hat{\sigma}(v_i) = \sigma(v_i))\]
+where \(\sigma(v_i)\) is the true community label of vertex \(v_i\), \(\hat{\sigma}(v_i)\) is the predicted label of \(v_i\), and \(\pi\) is a permutation operator.
+This is effectively the "misclustering count" of clustering function \(\hat{\sigma}\).
+
+For parameter estimation, because the popularity parameters $\{\lambda_{ik}\}$ are unidentifiable, we instead estimate the edge probabilities $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$ via the quantities $\hat{P}_{ij} = \hat{\lambda}_{iz_j} \hat{\lambda}_{jz_i}$.
+The parameter estimation error is then given by the normalized Frobenius norm of $P$ divided by the number of vertices, i.e.,
+$$\mathrm{RMSE}(\hat{P}, P) = \frac{1}{n} \|\hat{P} - P\|_F.$$
+
+We also note that unlike the MLE-based method \citep{307cbeb9b1be48299388437423d94bf1}, the ASE method in algorithm \ref{alg:param_est} can be trivially modified so as to not require the community labels if we are only interested in estimating $P$.
+More specifically we first compute the ASE $\hat{Z}$ of $A$ (see definition \ref{def:ase}) and then compute $\hat{P} = \hat{Z} I_{p,q} \hat{Z}^{\top}$. 
+The resulting estimate $\hat{P}$ will have the same convergence rate as that given in equation (\ref{eq:conv_prob}).
+
+### Balanced Communities {#sec:balanced-communities}
+
+In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
+multinomial distribution with mixture parameters
+\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(k \ell)}\}_K\) according
+to the drawn community labels, \(P\) was constructed using the drawn
+\(\{\lambda^{(k \ell)}\}_K\), and \(A\) was drawn from \(P\).
+
+For these examples, we set the following parameters:
+
+\begin{itemize}
+\tightlist
+\item
+  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
+\item
+  Number of underlying communities \(K = 2, 3, 4\)
+\item
+  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\), (i.e.,
+  each community label has an equal probability of being drawn)
+\item
+  Community labels
+  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
+\item
+  Within-group popularities
+  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
+\item
+  Between-group popularities
+  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
+  \(k \neq \ell\)
+\end{itemize}
+Fifty simulations were performed for each combination of $n$ and $K$. 
+The results for community recovery and parameter estimations are presented in figures \ref{fig:clust_err_ct_sim} and \ref{fig:p_block_est}, respectively.
+
+Figure \ref{fig:clust_err_ct_sim} shows that OSC recovers the community
+perfectly as $n$ increases, i.e., the number of mislabeled vertices
+goes to $0$. 
+The performance of SSC-ASE is comparable to OSC for $K \geq 3$ but is noticeably worse when $K = 2$. 
+Similarly, SSC on both the embedding and on the adjacency matrix produces similar trends for \(K > 2\). 
+The difference in performance between SSC-A and SSC-ASE for \(K = 2\) can be attributed to the final spectral clustering step of the affinity matrix.
+While the subspace detection property is guaranteed for large $n$, in our simulations, setting the
+sparsity parameter $\vartheta$ to the required value usually resulted
+in more than $K$ disconnected subgraphs in the affinity matrix
+$\hat{B}$. 
+We instead chose a smaller sparsity parameter, 
+necessitating a final clustering step. 
+A GMM was fit to the normalized Laplacian eigenmap of $\hat{B}$,
+but visual inspection suggests that the communities are not
+distributed as a mixture of Gaussians in the eigenmap. 
+A different choice of mixture distribution may result in better performance.
+
+\begin{figure}[tp]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_ct_sim-1}
+}
+\caption{Median and IQR of community detection error. Communities are approximately balanced. Simulations were repeated 50 times for each sample size.}
+\label{fig:clust_err_ct_sim}
+\end{figure}
+
+Given ground truth community labels, figure \ref{fig:p_block_est} shows
+that algorithm \ref{alg:param_est} and the MLE-based
+plug-in estimators perform comparably, with root mean square
+error decaying at rate approximately \(n^{-1/2}\) as $n$ increases.
+
+\begin{figure}[tp]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_p-1}}
+\caption{Median and IQR RMSE for edge probability matrices
+  reconstructed from the outputs of algorithm~\ref{alg:param_est}
+  (red) compared against outputs of an MLE-based method (blue) proposed in \citet{307cbeb9b1be48299388437423d94bf1}.
+  Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}
+\label{fig:p_block_est}
+\end{figure}
+
+### Imbalanced Communities {#sec:imbalanced-communities}
+
+Simulations performed in this section are the same as those in the
+previous section with the exception of the mixture parameters
+\(\{\alpha_1, ..., \alpha_K\}\) used to draw community labels from the
+multinomial distribution. For these examples, we set the following
+parameters:
+
+\begin{itemize}
+\tightlist
+\item
+  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
+\item
+  Number of underlying communities \(K = 2, 3, 4\)
+\item
+  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{\ell=1}^K \ell^{-1}}\)
+  for \(k = 1, ..., K\)
+\item
+  Community labels
+  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
+\item
+  Within-group popularities
+  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
+\item
+  Between-group popularities
+  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
+  \(k \neq \ell\)
+\end{itemize}
+
+Fifty simulations were performed for each combination of $n$ and $K$. 
+The results for community recovery and parameter estimations are presented in figures \ref{fig:clust_err_ct_sim_imba} and \ref{fig:lambda_est_p_imba}, respectively.
+
+From figure \ref{fig:clust_err_ct_sim_imba} we once again see that the
+number of mislabeled vertices trending to $0$ for OSC. 
+The performance of SSC-ASE is comparable to that of OSC for $K > 2$ but is worse when $K = 2$. 
+Figure \ref{fig:lambda_est_p_imba} indicates that the parameter estimation error also decays at rate $n^{-1/2}$ similar to that in the balanced communities setting.
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_ct_sim_imba-1}
+}
+\caption{Median and IQR of community detection error. Communities are imbalanced. 
+Simulations were repeated 50 times for each sample size.}\label{fig:clust_err_ct_sim_imba}
+\end{figure}
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_p_imba-1}
+}
+\caption{Median and IQR RMSE of edge probabilities derived from the
+  outputs of Algorithm \ref{alg:param_est} (red) compared against an
+  MLE-based method (blue) described in
+  \cite{307cbeb9b1be48299388437423d94bf1}.
+  Simulations were repeated 50 times for each sample size. Communities were drawn to be imbalanced.}
+\label{fig:lambda_est_p_imba}
+\end{figure}
+
+### Disassortative Models {#sec:disassortative-models}
+
+\citet{rubindelanchy2017statistical} demonstrated the power of applying GRDPG-based approaches to disassortative block models. 
+Likewise, demonstrate that GRDPG-based algorithms OSC and SSC-ASE perform well on disassortative PABMs. 
+Simulations performed in this section are the same as those in section \@ref(sec:balanced-communities) with the exception of the distributions from which within and between-group popularity parameters are drawn.
+Here, we draw these parameters such that the expected value is $1/3$ for the within-group popularity parameters and $2/3$ for the between-group popularity parameters:
+
+\begin{itemize}
+\tightlist
+\item
+  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
+\item
+  Number of underlying communities \(K = 2, 3, 4\)
+\item
+  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{\ell=1}^K \ell^{-1}}\)
+  for \(k = 1, ..., K\)
+\item
+  Community labels
+  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
+\item
+  Within-group popularities
+  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\)
+\item
+  Between-group popularities
+  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\) for
+  \(k \neq \ell\)
+\end{itemize}
+
+OSC, SSC-ASE, and SSC-A perform similarly on disassortative PABMs compared to the simulations in sections \@ref(sec:balanced-communities) and \@ref(sec:imbalanced-communities) 
+(figures \ref{fig:clust_k_disassortative} and \ref{fig:lambda_est_disassortative}).
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_k_disassortative}
+}
+\caption{Median and IQR of community detection error. 
+Communities are approximately balanced. 
+Edge probabilities were drawn to be disassortative. 
+Simulations were repeated 50 times for each sample size.}
+\label{fig:clust_k_disassortative}
+\end{figure}
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_disassortative}
+}
+\caption{Median and IQR RMSE of edge probabilities derived from the
+  outputs of algorithm \ref{alg:param_est} (red) compared against an
+  MLE-based method (blue) described in
+  \cite{307cbeb9b1be48299388437423d94bf1}.
+  Simulations were repeated 50 times for each sample size. 
+  Communities were drawn to be approximately balanced.
+  Edge probabilities were drawn to be disassortative.}
+\label{fig:lambda_est_disassortative}
+\end{figure}
+
+\begin{remark}
+While we call the models in this section ``disassortative'', it is our view that the assortative/disassortative distinction is not applicable to PABMs due to its flexibility. 
+Unlike the SBM and DCBM, in the PABM, each vertex is free to have a higher affinity to its own community or to other communities, independent of each other. 
+In other words, within the same community, $v_i$ may have a larger popularity parameter to its own community whereas $v_j$ may have a larger popularity parameter to a different community. 
+Furthermore, when viewed as GRDPGs, the assortativity or disassortativity of SBMs and DCBMs affects whether $P$ is positive semidefinite, which affects ASE-based approaches to analyzing the graph \citep{rubindelanchy2017statistical},
+but in the full rank PABM, $P$ is never positive semidefinite and will always have $K (K + 1) / 2$ positive eigenvalues and $K (K - 1) / 2$ negative eigenvalues. 
+\end{remark}
+
 ## Applications
+
+In the first example, we applied OSC (algorithm \ref{alg:osc}) to the Leeds Butterfly
+dataset \citep{Wang_2018} consisting of visual similarity measurements
+among 832 butterflies across 10 species. The graph was modified to match
+the example from \citet{noroozi2019estimation}: Only the $K=4$ most
+frequent species were considered, and the similarities were discretized
+to \(\{0, 1\}\) via thresholding. 
+Figure \ref{fig:butterfly} shows a sorted adjacency matrix sorted by the resultant clustering.
+
+Comparing against the ground truth species labels, \citet{noroozi2019estimation} 
+report that SSC on the adjacency matrix achieves an adjusted Rand index of 73\% in their implementation, whereas OSC achieves 92\% and SSC on the ASE achieves 96\%.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/butterfly-1}
+
+}
+
+\caption{Adjacency matrix of the Leeds Butterfly dataset 
+after sorting by the clustering outputted by OSC.}\label{fig:butterfly}
+\end{figure}
+
+In the second example (figure \ref{fig:mp} and table \ref{tab:unnamed-chunk-6}), we applied OSC to the British MPs Twitter network
+\citep{greene2013producing}, the Political Blogs network
+\citep{10.1145/1134271.1134277}, and the DBLP network
+\citep{NIPS2009_3855, 10.1007/978-3-642-15880-3_42}. 
+For this data analysis, we subsetted the data as described in
+\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the
+same networks. Our methods slightly underperformed compared to modularity
+maximization, although performance is comparable. 
+The run time of OSC is however much smaller than that of modularity maximization.
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/mp-1}
+}
+\caption{Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by OSC.}\label{fig:mp}
+\end{figure}
+
+\begin{table}[H]
+\centering
+\begin{tabular}[t]{l|r|r|r}
+\hline
+Network & MM & SSC-ASE & OSC\\
+\hline
+British MPs & 0.003 & 0.012 & 0.006\\
+\hline
+Political Blogs & 0.050 & 0.187 & 0.062\\
+\hline
+DBLP & 0.028 & 0.072 & 0.059\\
+\hline
+\end{tabular}
+\caption{\label{tab:unnamed-chunk-6}Community detection error rates on the British MPs Twitter, Political Blogs, and DBLP networks using modularity maximization, sparse subspace clustering, and OSC.}
+\end{table}
+
+In the third example (figure \ref{fig:households-figure} and 
+table \ref{tab:households-table}), 
+we analyzed the Karantaka villages data studied by
+\citet{DVN/U3BIHX_2013}. We chose the \texttt{visitgo}
+networks from villages 12, 31, and 46 at the household level. 
+In these networks, each node is a household and each edge is 
+an interaction between members of pairs of households. 
+The label of interest is the religious affiliation. 
+The networks were truncated to religions "1" and "2", 
+and vertices of degree 0 were removed. 
+The villages were chosen based on there being an adequate number of nodes 
+between households within each religion.
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/unnamed-chunk-7-1}
+}
+\caption{Adjacency matrix of the Karnataka villages data, arranged by the clustering produced by OSC (left).
+The villages studied here are, from left to right, 12, 31, and 46.}\label{fig:households-figure}
+\end{figure}
+
+\begin{table}[H]
+\centering
+\begin{tabular}[t]{l|r|r|r}
+\hline
+Network & MM & SSC-ASE & OSC\\
+\hline
+Village 12 & 0.270 & 0.291 & 0.227\\
+\hline
+Village 31 & 0.125 & 0.059 & 0.051\\
+\hline
+Village 46 & 0.052 & 0.069 & 0.056\\
+\hline
+\end{tabular}
+\caption{\label{tab:households-table}Community detection error rates for identifying household religion.}
+\end{table}
 
@@
 \begin{example}
+\label{ex:separated-curves}
 Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
@@
 Finally, let $A \sim \RDPG(X)$.
-Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.
+Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the ASE of the graph.
 Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.
 
-```{r two-curves, fig.cap = 'Manifold block model described in Example 1.
+```{r two-curves, fig.cap = 'Manifold block model from two disjoint curves.
 The latent configuration is on the left, a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.', fig.height = 2, fig.width = 7}
@@
 
-## Algorithms for Nonintersecting Manifolds {#sec:nonintersect}
+## Algorithm for Nonintersecting Manifolds {#sec:nonintersect}
 
@@
 
-\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting curves]
-\label{theorem:nonintersect-curves}
+\begin{example}
+\label{ex:separated-curves-2}
+Refer back to the RDPG in example \ref{ex:separated-curves} in which the latent vectors are drawn uniformly from two nonintersecting curves. 
+Figure \ref{fig:two-curves} shows that while the ASE introduces noise to the original latent vectors, there is still sufficient separation between the two communities. 
+Applying single-linkage clustering on the ASE results in recovering the original community labels, up to permutation (figure \ref{fig:two-curves-2}).
+
+```{r two-curves-2, fig.cap = 'Single-linkage clustering applied to the ASE of an MBM with two nonintersecting curves.
+The ASE is on the left and the dendrogram from single-linkage clustering, colored by the original community labels, is on the right.', fig.height = 3, fig.width = 7}
+par(mfrow = c(1, 2))
+plot(Xhat, col = z * 2, asp = 1, 
+     # xaxt = 'n', yaxt = 'n', 
+     xlab = NA, ylab = NA,
+     xaxt = 'n', yaxt = 'n')
+
+singlelink <- hclust(dist(Xhat), method = 'single')
+singlelink %>% 
+  as.dendrogram() %>% 
+  dendextend::color_branches(clusters = z, col = c(2, 4)) %>% 
+  plot(axes = FALSE, labels = FALSE, leaflab = 'none')
+```
+\end{example}
+
+## Algorithm for Intersecting Manifolds {#sec:intersect}
+
+The consistency of the ASE provides asymptotic guarantees for perfect community label recovery for the nonintersecting MBM. 
+This relies on the fact that single-linkage clustering and $\eta$-neighborhood clustering graphs result in disconnected components if clusters of points are far away from each other. 
+However, many graphs do not behave so nicely. 
+For example, the DCBM and PABM can both be viewed as MBMs in which the latent vectors come from (linear) manifolds that intersect at the origin (see section \@ref(sec:sbm-dcbm-grdpg)). 
+Nonlinear intersecting manifolds have also been observed in ASEs of real graphs \citep{SannaPassino2022, https://doi.org/10.48550/arxiv.1709.05454}. 
+While the DCBM and PABM cases can be handled by algorithms specific to linear structures such as SSC and OSC, nonlinear intersecting manifolds require a more flexible approach. 
+For the intersecting case, we restrict the MBM to the case in which each $F_{z_i}$ is identical and univariate on the unit interval $[0, 1]$. 
+The full mixture model is described as follows:
+
+1. Draw $t_1, ..., t_n \iid F$ for probability distribution $F$ with support $[0, 1]$.
+2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
+3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+4. Draw $A \sim \GRDPG_{p,q}(X; \rho_n)$.
+
+\begin{example}
+\label{ex:intersect-curves}
+Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$.
+This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
+For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
+$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
+Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
+The community structure is also not clear from the adjacency matrix. 
+However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).
+
+```{r intersect-curves-example, fig.height = 2, fig.width = 8, fig.cap = 'Latent vectors on intersecting curves (left), along with an RDPG drawn from this configuration (center) and its ASE (right).', cache = TRUE}
+f1 <- function(t) {
+  x1 <- t ^ 2
+  x2 <- 2 * t - 2 * t ^ 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- 2 * t - 2 * t ^ 2
+  x2 <- 1 - 2 * t + t ^ 2
+  return(cbind(x1, x2))
+}
+
+set.seed(314159)
+
+n1 <- 2 ** 7
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+T <- rbind(t1, t2)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+
+latent.vecs <- ggplot() + 
+  geom_point(aes(x = X[, 1], y = X[, 2], colour = factor(z)), size = .5) + 
+  coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  scale_colour_brewer(palette = 'Set1')
+
+P <- X %*% t(X)
+diag(P) <- 0
+A <- draw.graph(P)
+
+rdpg <- ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+
+Xhat <- embedding(A, 2, 0)
+
+ase <- ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z)), size = .5) + 
+  coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  scale_colour_brewer(palette = 'Set1')
+
+gridExtra::grid.arrange(latent.vecs, rdpg, ase, nrow = 1)
+```
+\end{example}
+
+In this setting, if we know the labels $z_1, ..., z_n$, then for each $k$^th^ subsample for which $z_i = k$, we can fit the curve $\hat{g}_k$ (up to an indefinite orthogonal transformation). 
+On the other hand, if we know the curves $g_1, ..., g_K$, then we can assign each embedding vector $\hat{x}_i$ the predicted label $\hat{z}_i$ based on which curve it is closest to. 
+This lends itself to an alternating coordinate descent algorithm in which we alternate between fitting curves and assigning labels at each iteration (algorithm \ref{alg:kcurves}). 
+We call this algorithm $K$-curves clustering. 
+In this algorithm, the loss function that we wish to minimize is
+\begin{equation}
+\label{eq:kcurves-loss}
+\sum_{k=1}^K \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2
+\end{equation}
+
+\begin{algorithm}[h]
+\label{alg:kcurves}
+\DontPrintSemicolon
+\SetAlgoLined
+\setstretch{1.5}
+\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$, $q$, stopping criterion $\epsilon$}
+\KwResult{Community assignments $1, ..., K$, curves $g_1, ..., g_K$}
+Compute $X$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
+Initialize community labels $z_1, ..., z_n$.\;
+\Repeat {the change in $\sum_k \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2$ is less than $\epsilon$} {
+\For {$k = 1, ..., K$} {
+Define $X_k$ as the rows of $X$ for which $z_i = k$.\;
+Fit curve $g_k$ and positions $t_{k_i}$ to $X_k$ by minimizing $\sum_{k_i} \|x_{k_i} - g_k(t_{k_i})\|^2$.\;
+}
+\For {$k = 1, ..., K$} {
+Assign $z_i \leftarrow \arg\min_\ell \|x_i - g_\ell(t_i)\|^2$.\;
+}
+}
+\caption{$K$-curves clustering.}
+\end{algorithm}
+
+$K$-curves clustering either requires knowledge of the functional form of each $g_k$ or a function that is flexible enough to approximate it. 
+The choice of $g_k$ affects the difficulty of the algorithm. 
+As a balance between flexibility and ease of estimation, we consider the case where each $g_k$ is a Bezier polynomial of degree $R$ with coefficients $p_k$. 
+Then we have $g_k(t) = g(t; p_k) = \sum_{r=0}^R p_k^{(r)} \binom{R}{r} (1-t)^{R-r} t^r$.
+
+\begin{example}
+\label{ex:intersect-curves-fit}
+Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
+In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \begin{bmatrix} 0 & 0 \end{bmatrix}^\top$, i.e., the curves meet at the origin. 
+Here, an accuracy of 87.5\% is achieved.
+
+```{r intersect-curves-2, fig.height = 2, fig.length = 2, cache = TRUE, fig.cap = '$K$-curves clustering applied to example \\ref{ex:intersect-curves}.
+The embedding vectors are labeled by predicted community.'}
+f1 <- function(t) {
+  x1 <- t ^ 2
+  x2 <- 2 * t - 2 * t ^ 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- 2 * t - 2 * t ^ 2
+  x2 <- 1 - 2 * t + t ^ 2
+  return(cbind(x1, x2))
+}
+
+set.seed(314159)
+
+n1 <- 2 ** 7
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+T <- rbind(t1, t2)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+
+P <- X %*% t(X)
+diag(P) <- 0
+A <- draw.graph(P)
+
+Xhat <- embedding(A, 2, 0)
+
+set.seed(1)
+out <- manifold.clustering(Xhat, 2, 
+                           parallel = TRUE,
+                           intercept = FALSE,
+                           maxit = 10, 
+                           normalize = TRUE,
+                           eps = 1e-3,
+                           verbose = FALSE,
+                           animate = FALSE)
+
+plot.estimated.curves(Xhat, out) + 
+  scale_colour_brewer(palette = 'Set1')
+```
+\end{example}
+
+While example \ref{ex:intersect-curves-fit} achieved high accuracy, algorithm \ref{alg:kcurves} often results in poor performance when it is paired with poor initialization. 
+In example \ref{ex:intersect-curves-fit}, it just so happened that we randomly chose initial curves that iterated to good fits, but this is not guaranteed. 
+We also restricted the curves to meet at the origin, which appears to have helped guide the curves to good fits. 
+In figures \ref{fig:loss} and \ref{fig:init-strategies}, we show the results of various initializations on fitting curves to example \ref{ex:intersect-curves}, such as random initialization and spectral clustering.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-7-1}
+
+}
+
+\caption{ASE labeled by estimated community labels for each initialization strategy.
+Upper right: random initialization assuming intersection at the origin.
+Upper middle: random initialization without assuming intersection at the origin.
+Upper right: initialization via spectral clustering assuming intersection at the origin.
+Lower left: initialization via spectral clustering without assuming intersection at the origin.
+Lower middle: initialization with ground truth labels, assuming intersection at the origin.
+Lower right: initialization with ground truth labels, without assuming intersection at the origin.}\label{fig:init-strategies}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-6-1}
+
+}
+
+\caption{Clustering loss vs. iteration for each run of $K$-curve clustering.}\label{fig:loss}
+\end{figure}
+
+Alternatively, if we turn this problem into a semi-supervised problem in which we know a few of the community labels, then the convergence of the ASE guarantees high accuracy.
+
+\begin{theorem}
+\label{thm:kcurves-semisupervised}
+Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$ and its density function is such that $f(t) > a > 0$ for every $t \in [0, 1]$, and each $g_k$ is is a Bezier curve of order $R$.
+Suppose that for each community $k$, we have labels for $R + 1$ vertices. 
+Define the community detection error rate as $\frac{1}{n} \min_\pi \sum_i I(z_i = \pi(\hat{z}_i))$, where $\pi$ is a permutation function and $\hat{z}_i$ are the predicted labels from $K$-curves clustering.
+Then as $n \to \infty$, $K$-curves clustering produces zero community detection error rate, with high probability. 
 \end{theorem}
 
-## Algorithms for Intersecting Manifolds {#sec:intersect}
-
-## Examples and Simulations
+## Simulation Study
+
+For a simulation study, we again turn to examples.  \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}. 
+The exact simulation setup is as follows:
+
+1. Draw $z_1, ..., z_n \iid \Multinomial(1/2, 1/2)$.
+2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
+3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$. 
+Collect the latent vectors into matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
+4. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
+
+For this simulation study, we set $n = 128, 256, 512, 1024, 2048$, and we start with completely random initialization as well as initializations in which we know 4 or 8 labels from each community. 
+For each setting, we simulate 50 replicates. 
+Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing $p_1^{(0)} = p_2^{(0)} = (0, 0)$. 
+Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting.
+
+```{r sim-curves, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates.'}
+n.vec <- 2 ^ c(7, 8, 9, 10, 11)
+nsamp.vec <- c(0, 4, 8)
+
+clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced.csv')
+
+clust.summary.df <- clustering.df %>% 
+  dplyr::group_by(n, nsamp) %>% 
+  dplyr::summarise(
+    med.count = median(error.count),
+    first.q.count = quantile(error.count, .25),
+    third.q.count = quantile(error.count, .75),
+    med.rate = median(error.rate),
+    first.q.rate = quantile(error.rate, .25),
+    third.q.rate = quantile(error.rate, .75)
+  ) %>% 
+  dplyr::ungroup()
+
+ggplot(clust.summary.df) + 
+  theme_bw() + 
+  theme(text = element_text(size = 10)) + 
+  scale_y_log10() +
+  scale_x_log10(breaks = n.vec) +
+  labs(y = 'community detection error rate',
+       x = 'n',
+       colour = 'known labels\nper community', 
+       shape = NULL) + 
+  geom_line(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
+  geom_point(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
+  geom_errorbar(aes(x = n, 
+                    ymin = first.q.rate, ymax = third.q.rate, 
+                    colour = factor(nsamp)),
+                width = .125) + 
+  scale_colour_brewer(palette = 'Set1')
+```
+
+We also explored the behavior of random initialization by setting $n = 512$ and compiling the loss values outputted by $K$-curves clustering for each replicate. 
+Here, we repeated the simulations $256$ times. 
+Figure \ref{fig:repeat-losses} suggests that in this particular setting, about half the time, random initialization of $K$-curves clustering results in the desired curves, while the rest of the time, $K$-curves clustering settled on a local minimum.
+Plotting each of the $256 \times 2$ fitted curves confirms this conclusion (figure \ref{fig:repeat-curves}).
+
+```{r repeat-losses, fig.height = 3, fig.width = 8, fig.cap = 'Histogram of loss values outputted by $K$-curves clustering over 256 replicates.'}
+repeat.df <- readr::read_csv('~/dev/manifold-block-models/simulations/repeat-study/repeat-study.csv')
+ggplot(repeat.df) + 
+  geom_histogram(aes(x = loss),
+                 colour = 'black', fill = 'white') + 
+  labs(x = 'loss values') + 
+  theme_bw()
+```
+
+```{r repeat-curves, fig.height = 3, fig.width = 3, fig.cap = 'Fitted curves over 256 replicates. The curves are colored by the loss values.'}
+plot.repeat.study <- function(repeat.df) {
+  repeat.df <- na.omit(repeat.df)
+  min.t <- 0
+  max.t <- 1
+  length.out <- 1e2
+  t. <- seq(min.t, max.t, length.out = length.out)
+  niter <- nrow(repeat.df)
+  p.df <- repeat.df[c('p121', 'p122', 'p131', 'p132', 
+                      'p221', 'p222', 'p231', 'p232')]
+  p.clust <- kmeans(p.df, 4)$cluster
+  loss.clust <- kmeans(repeat.df$loss, 2)$cluster
+  curves.df <- foreach::foreach(i = seq(niter), .combine = dplyr::bind_rows) %dopar% {
+    p1 <- matrix(c(repeat.df$p121[i], repeat.df$p122[i], 
+                   repeat.df$p131[i], repeat.df$p132[i]),
+                 nrow = 2, byrow = TRUE)
+    p1 <- rbind(0, p1)
+    p2 <- matrix(c(repeat.df$p221[i], repeat.df$p222[i],
+                   repeat.df$p231[i], repeat.df$p232[i]),
+                 nrow = 2, byrow = TRUE)
+    p2 <- rbind(0, p2)
+    p <- list(p1, p2)
+    plyr::ldply(seq_along(p), function(k) {
+      curve <- bezier.curve(t., p[[k]])
+      if (mean(curve[, 1]) < 0) {
+        curve[, 1] <- -curve[, 1]
+      }
+      dplyr::tibble(t = t.,
+                    x = curve[, 1],
+                    y = curve[, 2],
+                    z = k,
+                    iter = i,
+                    p.clust = p.clust[i],
+                    loss.clust = loss.clust[i],
+                    loss = repeat.df$loss[i])
+    })
+  }
+  ggplot(curves.df) + 
+    geom_path(aes(x = x, y = y, group = interaction(z, iter),
+                  colour = loss),
+              alpha = 1 / log(niter)) + 
+    theme_bw() + 
+    coord_fixed() + 
+    labs(colour = 'loss',
+         x = expression(x[1]),
+         y = expression(x[2])) + 
+    viridis::scale_color_viridis()
+}
+
+plot.repeat.study(repeat.df)
+```
 
 ## Applications
+
+In the first example, we applied $K$-curves clustering (algorithm \ref{alg:kcurves}) to the Macaque visuotactile brain areas and connections network \citep{https://doi.org/10.1111/j.1460-9568.2006.04678.x}, which is visualized in figure \ref{fig:macaque-graph}.
+In this graph, there are 45 vertices which each correspond to a brain area, and the communities represent each area's tactile function. 
+Applying an ASE with $p = 2$ and $q = 0$ to this graph results in an embedding that appears to fall approximately on two curves (figure \ref{fig:macaque-ase}) which intersect at the origin. 
+Figure \ref{fig:macaque-fitted} shows the result of $K$-curves clustering on this graph, 
+which yields a 4.4\% error rate.
+
+```{r macaque-graph, fig.cap = 'Macaque visuotactile brain areas and connections network. The vertices represent brain areas.', fig.height = 2, fig.width = 3}
+data(macaque, package = 'igraphdata')
+A <- as.matrix(igraph::as_adjacency_matrix(macaque, type = 'both'))
+A <- sign(A + t(A))
+z <- as.numeric(factor(igraph::vertex_attr(macaque)$shape))
+
+ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
+
+```{r macaque-ase, fig.cap = 'ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.', fig.width = 4, fig.height = 2}
+Xhat <- embedding(A, 2, 0)
+ggplot() + 
+  geom_point(aes(x = Xhat[, 1],
+                 y = Xhat[, 2],
+                 colour = factor(z))) + 
+  labs(colour = NULL, y = expression(x[2]), x = expression(x[1])) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  coord_fixed() + 
+  theme_bw()
+```
+
+```{r macaque-fitted, fig.cap = '$K$-means clustering on the ASE of the Macaque brain areas network.
+The embedding vectors are colored by predicted label.', fig.width = 4, fig.height = 2, cache = TRUE}
+set.seed(123)
+z.spectral <- spectral.clustering(A)
+macaque.out <- manifold.clustering(Xhat, 2, 
+                                   # initialization = 'random',
+                                   initialization = z.spectral,
+                                   # initialization = z,
+                                   intercept = FALSE, 
+                                   parallel = TRUE, 
+                                   verbose = FALSE, 
+                                   animate = FALSE, 
+                                   curve.init = 'x')
+
+plot.estimated.curves(Xhat, macaque.out) + 
+  scale_colour_brewer(palette = 'Set1')
+```
+
+In the second example, we consider the Harry Potter enmity graph (figure \ref{fig:hp}).
+
+```{r hp, fig.cap = 'Harry Potter enmity graph.
+The vertices represent characters, the edges represent whether pairs of characters are enemies, and the communities represent whether the characters are on the side of the protagonist or the antagonist.', fig.width = 4, fig.height = 2}
+relations.df <- readr::read_csv('~/dev/potter-network/data/relations.csv') %>% 
+  dplyr::mutate(source = source + 1, target = target + 1)
+characters.df <- readr::read_csv('~/dev/potter-network/data/characters.csv') %>% 
+  dplyr::mutate(id = id + 1)
+n <- nrow(characters.df)
+
+A <- matrix(0, nrow = n, ncol = n)
+rownames(A) <- colnames(A) <- characters.df$name
+for (i in seq(nrow(relations.df))) {
+  if (relations.df$type[i] == '-') {
+    A[relations.df$source[i], relations.df$target[i]] %<>% magrittr::add(1)
+  } else {
+    A[relations.df$source[i], relations.df$target[i]] %<>% magrittr::subtract(1)
+  }
+}
+A[A < 0] <- 0
+A <- A + t(A)
+A <- sign(A)
+diag(A) <- 0
+deg.A <- colSums(A)
+
+A <- A[deg.A > 0, deg.A > 0]
+
+ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
 

### 2022-10-19 draft/draft.tex
almost done

@@
 \newcommand{\DCBM}{\mathrm{DCBM}}
+\newcommand{\rank}{\mathrm{rank}}
+\newcommand{\MBM}{\mathrm{MBM}}
 
@@
 However, it still assumes that each vertex within a community has the same relative propensity to connect to each community.
-For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that everyone is ten times more likely to connect with a member of the opposite sex than to a member of the same sex.
+For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex.
 The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community.
@@
 
+The PABM, DCBM, and SBM are nested models, that is, the SBM is a special case of the DCBM which in turn is a special case of the SBM.
+In this section, these connections are made explicit, and we explore the implications for modeling.
+For ease of notation, throughout this section the rows and columns of \(P\) are assumed to be organized by community in increasing order, i.e., the first \(n_1\) rows and columns of \(P\) correspond to community \(1\), the next \(n_2\) rows and columns correspond to community \(2\), etc.
+We use the notation \(P^{(k \ell)} = (P^{(\ell k)})^\top\) to denote the block of \(P\) that corresponds to the edges between communities \(k\) and \(\ell\).
+In section \ref{sec:pabm-grdpg}, we will deal with this notation and convention more carefully.
+For additional details on the connections among the SBM, DCBM, and PABM, refer to \citet{Noroozi2022}.
+
+To see that the SBM is a special case of the DCBM, we can simply restrict the degree parameters \(\omega_i\) to all be a constant value, e.g., \(\omega_i \equiv 1\).
+But in order to better understand the relationship between these two models, we will break down the models further.
+
+Define \(B\) as the \(K \times K\) matrix of edge probabilities between pairs of communities in the SBM.
+Then \(B_{k \ell} = \theta_{k \ell}\) and \(P_{ij} = \theta_{z_i, z_j} = B_{z_i, z_j}\).
+We can further show that block \(P^{(k \ell)}\) can be written as \(P^{(k \ell)} = e^{(k)} B_{k \ell} (e^{(\ell)})^\top\) where \(e^{(k)}\) is the \(1\)-vector of dimension \(k\).
+Since \(B_{k \ell}\) is a scalar, \(P^{(k \ell)} = e^{(k)} B_{k \ell} (e^{(\ell)})^\top\) is a singular value decomposition of a rank 1 matrix and thus each block \(P^{(k \ell)}\) is rank 1.
+Then we can see that the entire edge probability matrix can be written as
+\[P = Z B Z^\top\]
+and \(Z\) is an \(n \times K\) matrix in which \(Z_{ik} = I(z_i = k) = z_{ik}\). Thus, if \(B\) is full rank, i.e., \(\rank(B) = K\), then \(P\) is also rank \(K\).
+
+To extend the SBM to the DCBM, we replace \(e^{(k)}\) with the vector \(\omega^{(k)}\) which contains the degree parameters of vertices in community \(k\).
+Then for the DCBM, \(P^{(k \ell)} = \omega^{(k)} B_{k \ell} (\omega^{(\ell)})^\top\).
+Similar to the SBM, this is a singular value decomposition of a rank 1 matrix.
+The full edge probability matrix can be written as
+\[P = (\Omega Z) B (\Omega Z)^\top\]
+for \(\Omega = \diag(\omega)\).
+Since \(\Omega\) is diagonal, we again have that \(\rank(P) = K\) (again, assuming that \(B\) is full rank).
+Thus, the dimensionality of the DCBM is the same as the SBM, and to restrict the DCBM to the PABM, let \(\Omega = I\).
+
+To extend the DCBM to the PABM, we note that the \(k \ell\)\textsuperscript{th} block of the edge probability matrix can be written as \(P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top\) where \(\lambda^{(k \ell)} \in \mathbb{R}^{n_k}\) is the vector of popularity parameters belonging to vertices in community \(k\) that point to community \(\ell\), i.e., \(\lambda^{(k \ell)}_i = \lambda_{i + \sum_{s=1}^{k-1} n_s, \ell}\).
+Similar to how \(\omega^{(k)}\) is a sub-vector of \(\omega\)
+To match this decomposition with the one for the DCBM, we can fix each \(B_{k \ell} \equiv 1\) and write \(P^{(k \ell)} = \lambda^{(k \ell)} B_{k \ell} (\lambda^{(\ell k)})^\top\).
+Restricting each \(\lambda^{(k \ell)} = \omega^{(k)}\) reduces the PABM to the DCBM.
+To write out the full edge probability matrix of the PABM, let \(\Lambda^{(k)} =\begin{bmatrix} \lambda^{(k1)} & \cdots & \lambda^{(kK)} \end{bmatrix}\) and \(X = \blockdiag(\Lambda^{(1)}, ..., \Lambda^{(K)})\).
+Then for some \(K^2 \times K^2\) permutation matrix \(\Pi\),
+\[P = X \Pi X^\top.\]
+From this decomposition, we can see that unlike the SBM and DCBM, \(\rank(P) = \rank(X) = K^2\).
+We will explore this decomposition of the edge probability matrix of the PABM in more detail in section \ref{sec:pabm-grdpg}.
+
+The full hierarchy of nested block models is summarized in table \ref{tab:hierarchy}.
+
+\begin{table}
+
+\caption{\label{tab:hierarchy}Summary of an hierarchical or nested view of block momdels.}
+\centering
+\begin{tabular}[t]{l|l|l|l|l}
+\hline
+Model & Parameters & Dimensions & Matrix Decomposition & Reduction to simpler model\\
+\hline
+SBM & $K (K-1) / 2$ & $K$ & $P = Z B Z^\top$ & \\
+\hline
+DCBM & $K (K-1) / 2 + n$ & $K$ & $P = (\Omega Z) B (\Omega Z)^\top$ & $\omega = e$\\
+\hline
+PABM & $n K$ & $K^2$ & $P = X \Pi X^\top$ & $\lambda_{ik} = \omega_i \sqrt{\theta_{z_i, k}}$\\
+\hline
+\end{tabular}
+\end{table}
+
 \newpage
@@
 \hypertarget{sec:grdpg}{%
-\subsection{Random Dot Product Graphs and Generalized Random Dot Product Graphs}\label{sec:grdpg}}
+\subsection{The Generalized Random Dot Product Graph}\label{sec:grdpg}}
 
@@
 \end{remark}
+
+Now that we defined the GRDPG as a generative model for Bernoulli random graphs,
+we need a method for estimation.
+We can use Adjacency Spectral Embedding (ASE) \citep{doi:10.1080/01621459.2012.699795} to recover the latent vectors of a GRDPG.
+This procedure consists of taking the spectral decomposition of \(A\) and keeping the \(p + q\) largest eigenvalues (in modulus) and corresponding eigenvectors of \(A\).
+
+\begin{definition}[Adjacency spectral embedding]
+\label{def:ase}
+Let $A$ be an $n
+\times n$ adjacency matrix and denote the
+eigendecomposition of $A$ as
+$$A = \sum_{i} \hat{\lambda}_i \hat{v}_i \hat{v}_i^{\top}, \quad
+\text{where} \,\,
+|\hat{\lambda}|_1 \geq |\hat{\lambda}_2| \geq \dots \geq
+|\hat{\lambda}_n|.$$
+Let $\hat{D}$ be a diagonal matrix whose diagonal entries are
+the eigenvalues $\{\hat{\lambda}_i\}_{i=1}^{d}$ arranged in decreasing values (not in
+decreasing modulus) and let $\hat{V}$ be the $n \times d$ matrix whose
+columns are the corresponding eigenvectors $\{\hat{v}_i\}_{i=1}^{d}$
+arranged in the same order as the diagonal entries of $\hat{D}$. Now
+define $\hat{X} = \hat{V} |\hat{D}|^{1/2}$ where the $|\cdot|$
+operation is applied elementwise to the entries of $\hat{D}$. Then
+$\hat{X}$ serves as an estimate of $X$, up to some non-identifiable
+orthogonal transformation $Q$ as described in
+remark~\ref{rem:non_identifiable} and definition~\ref{def:indefinite};
+see \citet{rubindelanchy2017statistical} for further details.
+\end{definition}
+
+\citet{doi:10.1080/01621459.2012.699795} and \citet{rubindelanchy2017statistical} showed that the ASE is consistent in estimating the latent vectors, up to an indefinite orthogonal transformation.
+More specifically, for sparsity parameters such that \(n \rho_n = \omega (\log^{4 c} n)\) for some constant \(c > 1\) and indefinite orthogonal transformation \(Q \in \mathbb{O}(p, q)\),
+\[\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)\]
+where \(x^\top\) and \(\hat{x}^\top\) are the rows of \(\hat{X}\) (from the ASE) and \(X\), respectively.
+
+\begin{example}
+Suppose that latent vectors for an RDPG are drawn from the curve 
+$x(t) = \begin{bmatrix} t^2 & 2 t (t-1) \end{bmatrix}^\top$, $t \in [0, 1]$. 
+Let $x_i = x(t_i)$. 
+Then for each $i < j$, the edges are drawn as $A_{ij} \indep \Bernoulli(x_i^\top x_j)$. 
+If after observing $A$, we wish to estimate the curve, the ASE provides an estimator in which the estimated latent vectors converge to the curve, up to an orthogonal transformation (figure \ref{fig:ase-example}).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/ase-example-1}
+
+}
+
+\caption{200 points drawn from a curve (left), and the ASE of an adjacency matrix drawn from this curve as an RDPG (right).
+The red line is the original curve rotated to match the ASE.}\label{fig:ase-example}
+\end{figure}
+
+\end{example}
 
@@
 
+For each simulation, community labels are drawn from a multinomial
+distribution, the popularity vectors \(\{\lambda^{(k \ell)}\}_K\) are drawn
+from two types of joint distributions depending on whether \(k = \ell\) or \(k \not = \ell\). The edge probability matrix \(P\) is constructed using the popularity
+vectors and finally the adjacency matrix \(A\)
+is drawn \(A \sim \mathrm{Bernoulli}(P)\). OSC (algorithm \ref{alg:osc}) is then used for community detection, and this
+method is compared against (1) SSC using the spectral embedding \(\hat{V}\)
+(algorithm \ref{alg:ssc}), (2) SSC using the rows of the
+observed adjacency matrix \(A\) as is done in \citet{noroozi2019estimation}
+and (3) modularity maximization (MM) as is done in
+\citet{307cbeb9b1be48299388437423d94bf1}. We denote the two SSC
+implementations using the rows of \(A\) and using the spectral embedding
+of \(A\) as SSC-A and SSC-ASE, respectively.
+The parameters \(\vartheta\)
+that controls the sparsity for SSC-A and SSC-ASE were chosen via a preliminary cross-validation experiment.
+In practice, \(\vartheta\) that guarantee SDP (if it is possible for the particular simulated data) often result in more than \(K\) disconnected subgraphs, so a smaller \(\vartheta\) that does not guarantee SDP was chosen,
+and the final clustering step of SSC-A and SSC-ASE was done
+by fitting a Gaussian mixture model to the normalized Laplacian
+eigenmap embeddings \citep{belkin03:_laplac} of the affinity matrix \(B\).
+We also estimate the latent popularity vectors \(\{\lambda^{(k \ell)}\}\)
+by assuming that the true community labels are known and then apply
+algorithm \ref{alg:param_est}, and we compare this estimation method against an
+MLE-based estimator as described in \citet{noroozi2019estimation} and
+\citet{307cbeb9b1be48299388437423d94bf1}.
+
+Modularity maximization is NP-hard, so
+\citet{307cbeb9b1be48299388437423d94bf1} used the extreme points
+(EP) algorithm (\cite{le2016}) as a greedy
+relaxation of the optimization problem; the EP algorithm has a running
+time of \(O(n^{K-1})\) where \(n\) is the number of vertices in the graph
+and \(K\) is the number of communities.
+For these simulations we instead replace the EP algorithm with the
+Louvain algorithm for modularity maximization,
+as the implementation of the EP algorithm in
+\citet{307cbeb9b1be48299388437423d94bf1} is too computationally expensive for \(K > 2\). For \(K = 2\), it
+was verified that the Louvain algorithm produces comparable results
+to EP-MM.
+
+For comparing methods, we define the community detection error as:
+\[L_c(\hat{\sigma}, \sigma; V) = \min_\pi \sum_i I(\pi \circ \hat{\sigma}(v_i) = \sigma(v_i))\]
+where \(\sigma(v_i)\) is the true community label of vertex \(v_i\), \(\hat{\sigma}(v_i)\) is the predicted label of \(v_i\), and \(\pi\) is a permutation operator.
+This is effectively the ``misclustering count'' of clustering function \(\hat{\sigma}\).
+
+For parameter estimation, because the popularity parameters \(\{\lambda_{ik}\}\) are unidentifiable, we instead estimate the edge probabilities \(P_{ij} = \lambda_{i z_j} \lambda_{j z_i}\) via the quantities \(\hat{P}_{ij} = \hat{\lambda}_{iz_j} \hat{\lambda}_{jz_i}\).
+The parameter estimation error is then given by the normalized Frobenius norm of \(P\) divided by the number of vertices, i.e.,
+\[\mathrm{RMSE}(\hat{P}, P) = \frac{1}{n} \|\hat{P} - P\|_F.\]
+
+We also note that unlike the MLE-based method \citep{307cbeb9b1be48299388437423d94bf1}, the ASE method in algorithm \ref{alg:param_est} can be trivially modified so as to not require the community labels if we are only interested in estimating \(P\).
+More specifically we first compute the ASE \(\hat{Z}\) of \(A\) (see definition \ref{def:ase}) and then compute \(\hat{P} = \hat{Z} I_{p,q} \hat{Z}^{\top}\).
+The resulting estimate \(\hat{P}\) will have the same convergence rate as that given in equation (\ref{eq:conv_prob}).
+
+\hypertarget{sec:balanced-communities}{%
+\subsubsection{Balanced Communities}\label{sec:balanced-communities}}
+
+In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
+multinomial distribution with mixture parameters
+\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(k \ell)}\}_K\) according
+to the drawn community labels, \(P\) was constructed using the drawn
+\(\{\lambda^{(k \ell)}\}_K\), and \(A\) was drawn from \(P\).
+
+For these examples, we set the following parameters:
+
+\begin{itemize}
+\tightlist
+\item
+  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
+\item
+  Number of underlying communities \(K = 2, 3, 4\)
+\item
+  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\), (i.e.,
+  each community label has an equal probability of being drawn)
+\item
+  Community labels
+  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
+\item
+  Within-group popularities
+  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
+\item
+  Between-group popularities
+  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
+  \(k \neq \ell\)
+\end{itemize}
+
+Fifty simulations were performed for each combination of \(n\) and \(K\).
+The results for community recovery and parameter estimations are presented in figures \ref{fig:clust_err_ct_sim} and \ref{fig:p_block_est}, respectively.
+
+Figure \ref{fig:clust_err_ct_sim} shows that OSC recovers the community
+perfectly as \(n\) increases, i.e., the number of mislabeled vertices
+goes to \(0\).
+The performance of SSC-ASE is comparable to OSC for \(K \geq 3\) but is noticeably worse when \(K = 2\).
+Similarly, SSC on both the embedding and on the adjacency matrix produces similar trends for \(K > 2\).
+The difference in performance between SSC-A and SSC-ASE for \(K = 2\) can be attributed to the final spectral clustering step of the affinity matrix.
+While the subspace detection property is guaranteed for large \(n\), in our simulations, setting the
+sparsity parameter \(\vartheta\) to the required value usually resulted
+in more than \(K\) disconnected subgraphs in the affinity matrix
+\(\hat{B}\).
+We instead chose a smaller sparsity parameter,
+necessitating a final clustering step.
+A GMM was fit to the normalized Laplacian eigenmap of \(\hat{B}\),
+but visual inspection suggests that the communities are not
+distributed as a mixture of Gaussians in the eigenmap.
+A different choice of mixture distribution may result in better performance.
+
+\begin{figure}[tp]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_ct_sim-1}
+}
+\caption{Median and IQR of community detection error. Communities are approximately balanced. Simulations were repeated 50 times for each sample size.}
+\label{fig:clust_err_ct_sim}
+\end{figure}
+
+Given ground truth community labels, figure \ref{fig:p_block_est} shows
+that algorithm \ref{alg:param_est} and the MLE-based
+plug-in estimators perform comparably, with root mean square
+error decaying at rate approximately \(n^{-1/2}\) as \(n\) increases.
+
+\begin{figure}[tp]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_p-1}}
+\caption{Median and IQR RMSE for edge probability matrices
+  reconstructed from the outputs of algorithm~\ref{alg:param_est}
+  (red) compared against outputs of an MLE-based method (blue) proposed in \citet{307cbeb9b1be48299388437423d94bf1}.
+  Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}
+\label{fig:p_block_est}
+\end{figure}
+
+\hypertarget{sec:imbalanced-communities}{%
+\subsubsection{Imbalanced Communities}\label{sec:imbalanced-communities}}
+
+Simulations performed in this section are the same as those in the
+previous section with the exception of the mixture parameters
+\(\{\alpha_1, ..., \alpha_K\}\) used to draw community labels from the
+multinomial distribution. For these examples, we set the following
+parameters:
+
+\begin{itemize}
+\tightlist
+\item
+  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
+\item
+  Number of underlying communities \(K = 2, 3, 4\)
+\item
+  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{\ell=1}^K \ell^{-1}}\)
+  for \(k = 1, ..., K\)
+\item
+  Community labels
+  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
+\item
+  Within-group popularities
+  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
+\item
+  Between-group popularities
+  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
+  \(k \neq \ell\)
+\end{itemize}
+
+Fifty simulations were performed for each combination of \(n\) and \(K\).
+The results for community recovery and parameter estimations are presented in figures \ref{fig:clust_err_ct_sim_imba} and \ref{fig:lambda_est_p_imba}, respectively.
+
+From figure \ref{fig:clust_err_ct_sim_imba} we once again see that the
+number of mislabeled vertices trending to \(0\) for OSC.
+The performance of SSC-ASE is comparable to that of OSC for \(K > 2\) but is worse when \(K = 2\).
+Figure \ref{fig:lambda_est_p_imba} indicates that the parameter estimation error also decays at rate \(n^{-1/2}\) similar to that in the balanced communities setting.
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_ct_sim_imba-1}
+}
+\caption{Median and IQR of community detection error. Communities are imbalanced. 
+Simulations were repeated 50 times for each sample size.}\label{fig:clust_err_ct_sim_imba}
+\end{figure}
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_p_imba-1}
+}
+\caption{Median and IQR RMSE of edge probabilities derived from the
+  outputs of Algorithm \ref{alg:param_est} (red) compared against an
+  MLE-based method (blue) described in
+  \cite{307cbeb9b1be48299388437423d94bf1}.
+  Simulations were repeated 50 times for each sample size. Communities were drawn to be imbalanced.}
+\label{fig:lambda_est_p_imba}
+\end{figure}
+
+\hypertarget{sec:disassortative-models}{%
+\subsubsection{Disassortative Models}\label{sec:disassortative-models}}
+
+\citet{rubindelanchy2017statistical} demonstrated the power of applying GRDPG-based approaches to disassortative block models.
+Likewise, demonstrate that GRDPG-based algorithms OSC and SSC-ASE perform well on disassortative PABMs.
+Simulations performed in this section are the same as those in section \ref{sec:balanced-communities} with the exception of the distributions from which within and between-group popularity parameters are drawn.
+Here, we draw these parameters such that the expected value is \(1/3\) for the within-group popularity parameters and \(2/3\) for the between-group popularity parameters:
+
+\begin{itemize}
+\tightlist
+\item
+  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
+\item
+  Number of underlying communities \(K = 2, 3, 4\)
+\item
+  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{\ell=1}^K \ell^{-1}}\)
+  for \(k = 1, ..., K\)
+\item
+  Community labels
+  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
+\item
+  Within-group popularities
+  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\)
+\item
+  Between-group popularities
+  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\) for
+  \(k \neq \ell\)
+\end{itemize}
+
+OSC, SSC-ASE, and SSC-A perform similarly on disassortative PABMs compared to the simulations in sections \ref{sec:balanced-communities} and \ref{sec:imbalanced-communities}
+(figures \ref{fig:clust_k_disassortative} and \ref{fig:lambda_est_disassortative}).
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_k_disassortative}
+}
+\caption{Median and IQR of community detection error. 
+Communities are approximately balanced. 
+Edge probabilities were drawn to be disassortative. 
+Simulations were repeated 50 times for each sample size.}
+\label{fig:clust_k_disassortative}
+\end{figure}
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_disassortative}
+}
+\caption{Median and IQR RMSE of edge probabilities derived from the
+  outputs of algorithm \ref{alg:param_est} (red) compared against an
+  MLE-based method (blue) described in
+  \cite{307cbeb9b1be48299388437423d94bf1}.
+  Simulations were repeated 50 times for each sample size. 
+  Communities were drawn to be approximately balanced.
+  Edge probabilities were drawn to be disassortative.}
+\label{fig:lambda_est_disassortative}
+\end{figure}
+
+\begin{remark}
+While we call the models in this section ``disassortative'', it is our view that the assortative/disassortative distinction is not applicable to PABMs due to its flexibility. 
+Unlike the SBM and DCBM, in the PABM, each vertex is free to have a higher affinity to its own community or to other communities, independent of each other. 
+In other words, within the same community, $v_i$ may have a larger popularity parameter to its own community whereas $v_j$ may have a larger popularity parameter to a different community. 
+Furthermore, when viewed as GRDPGs, the assortativity or disassortativity of SBMs and DCBMs affects whether $P$ is positive semidefinite, which affects ASE-based approaches to analyzing the graph \citep{rubindelanchy2017statistical},
+but in the full rank PABM, $P$ is never positive semidefinite and will always have $K (K + 1) / 2$ positive eigenvalues and $K (K - 1) / 2$ negative eigenvalues. 
+\end{remark}
+
 \hypertarget{applications}{%
 \subsection{Applications}\label{applications}}
+
+In the first example, we applied OSC (algorithm \ref{alg:osc}) to the Leeds Butterfly
+dataset \citep{Wang_2018} consisting of visual similarity measurements
+among 832 butterflies across 10 species. The graph was modified to match
+the example from \citet{noroozi2019estimation}: Only the \(K=4\) most
+frequent species were considered, and the similarities were discretized
+to \(\{0, 1\}\) via thresholding.
+Figure \ref{fig:butterfly} shows a sorted adjacency matrix sorted by the resultant clustering.
+
+Comparing against the ground truth species labels, \citet{noroozi2019estimation}
+report that SSC on the adjacency matrix achieves an adjusted Rand index of 73\% in their implementation, whereas OSC achieves 92\% and SSC on the ASE achieves 96\%.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/butterfly-1}
+
+}
+
+\caption{Adjacency matrix of the Leeds Butterfly dataset 
+after sorting by the clustering outputted by OSC.}\label{fig:butterfly}
+\end{figure}
+
+In the second example (figure \ref{fig:mp} and table \ref{tab:unnamed-chunk-6}), we applied OSC to the British MPs Twitter network
+\citep{greene2013producing}, the Political Blogs network
+\citep{10.1145/1134271.1134277}, and the DBLP network
+\citep{NIPS2009_3855, 10.1007/978-3-642-15880-3_42}.
+For this data analysis, we subsetted the data as described in
+\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the
+same networks. Our methods slightly underperformed compared to modularity
+maximization, although performance is comparable.
+The run time of OSC is however much smaller than that of modularity maximization.
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/mp-1}
+}
+\caption{Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by OSC.}\label{fig:mp}
+\end{figure}
+
+\begin{table}[H]
+\centering
+\begin{tabular}[t]{l|r|r|r}
+\hline
+Network & MM & SSC-ASE & OSC\\
+\hline
+British MPs & 0.003 & 0.012 & 0.006\\
+\hline
+Political Blogs & 0.050 & 0.187 & 0.062\\
+\hline
+DBLP & 0.028 & 0.072 & 0.059\\
+\hline
+\end{tabular}
+\caption{\label{tab:unnamed-chunk-6}Community detection error rates on the British MPs Twitter, Political Blogs, and DBLP networks using modularity maximization, sparse subspace clustering, and OSC.}
+\end{table}
+
+In the third example (figure \ref{fig:households-figure} and
+table \ref{tab:households-table}),
+we analyzed the Karantaka villages data studied by
+\citet{DVN/U3BIHX_2013}. We chose the \texttt{visitgo}
+networks from villages 12, 31, and 46 at the household level.
+In these networks, each node is a household and each edge is
+an interaction between members of pairs of households.
+The label of interest is the religious affiliation.
+The networks were truncated to religions ``1'' and ``2'',
+and vertices of degree 0 were removed.
+The villages were chosen based on there being an adequate number of nodes
+between households within each religion.
+
+\begin{figure}[H]
+{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/unnamed-chunk-7-1}
+}
+\caption{Adjacency matrix of the Karnataka villages data, arranged by the clustering produced by OSC (left).
+The villages studied here are, from left to right, 12, 31, and 46.}\label{fig:households-figure}
+\end{figure}
+
+\begin{table}[H]
+\centering
+\begin{tabular}[t]{l|r|r|r}
+\hline
+Network & MM & SSC-ASE & OSC\\
+\hline
+Village 12 & 0.270 & 0.291 & 0.227\\
+\hline
+Village 31 & 0.125 & 0.059 & 0.051\\
+\hline
+Village 46 & 0.052 & 0.069 & 0.056\\
+\hline
+\end{tabular}
+\caption{\label{tab:households-table}Community detection error rates for identifying household religion.}
+\end{table}
 
@@
 Finally, let $A \sim \RDPG(X)$.
-Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.
+Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the ASE of the graph.
 Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.
@@
 
-\caption{Manifold block model described in Example 1.
+\caption{Manifold block model from two disjoint curves.
 The latent configuration is on the left, a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.}\label{fig:two-curves}
@@
 \hypertarget{sec:nonintersect}{%
-\subsection{Algorithms for Nonintersecting Manifolds}\label{sec:nonintersect}}
+\subsection{Algorithm for Nonintersecting Manifolds}\label{sec:nonintersect}}
 
@@
 
-\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting curves]
-\label{theorem:nonintersect-curves}
+\begin{example}
+\label{ex:separated-curves-2}
+Refer back to the RDPG in example \ref{ex:separated-curves} in which the latent vectors are drawn uniformly from two nonintersecting curves. 
+Figure \ref{fig:two-curves} shows that while the ASE introduces noise to the original latent vectors, there is still sufficient separation between the two communities. 
+Applying single-linkage clustering on the ASE results in recovering the original community labels, up to permutation (figure \ref{fig:two-curves-2}).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/two-curves-2-1}
+
+}
+
+\caption{Single-linkage clustering applied to the ASE of an MBM with two nonintersecting curves.
+The ASE is on the left and the dendrogram from single-linkage clustering, colored by the original community labels, is on the right.}\label{fig:two-curves-2}
+\end{figure}
+\end{example}
+
+\hypertarget{sec:intersect}{%
+\subsection{Algorithm for Intersecting Manifolds}\label{sec:intersect}}
+
+The consistency of the ASE provides asymptotic guarantees for perfect community label recovery for the nonintersecting MBM.
+This relies on the fact that single-linkage clustering and \(\eta\)-neighborhood clustering graphs result in disconnected components if clusters of points are far away from each other.
+However, many graphs do not behave so nicely.
+For example, the DCBM and PABM can both be viewed as MBMs in which the latent vectors come from (linear) manifolds that intersect at the origin (see section \ref{sec:sbm-dcbm-grdpg}).
+Nonlinear intersecting manifolds have also been observed in ASEs of real graphs \citep{SannaPassino2022, https://doi.org/10.48550/arxiv.1709.05454}.
+While the DCBM and PABM cases can be handled by algorithms specific to linear structures such as SSC and OSC, nonlinear intersecting manifolds require a more flexible approach.
+For the intersecting case, we restrict the MBM to the case in which each \(F_{z_i}\) is identical and univariate on the unit interval \([0, 1]\).
+The full mixture model is described as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Draw \(t_1, ..., t_n \iid F\) for probability distribution \(F\) with support \([0, 1]\).
+\item
+  Draw \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), the community labels.
+\item
+  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
+\item
+  Draw \(A \sim \GRDPG_{p,q}(X; \rho_n)\).
+\end{enumerate}
+
+\begin{example}
+\label{ex:intersect-curves}
+Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$.
+This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
+For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
+$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
+Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
+The community structure is also not clear from the adjacency matrix. 
+However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/intersect-curves-example-1}
+
+}
+
+\caption{Latent vectors on intersecting curves (left), along with an RDPG drawn from this configuration (center) and its ASE (right).}\label{fig:intersect-curves-example}
+\end{figure}
+\end{example}
+
+In this setting, if we know the labels \(z_1, ..., z_n\), then for each \(k\)\textsuperscript{th} subsample for which \(z_i = k\), we can fit the curve \(\hat{g}_k\) (up to an indefinite orthogonal transformation).
+On the other hand, if we know the curves \(g_1, ..., g_K\), then we can assign each embedding vector \(\hat{x}_i\) the predicted label \(\hat{z}_i\) based on which curve it is closest to.
+This lends itself to an alternating coordinate descent algorithm in which we alternate between fitting curves and assigning labels at each iteration (algorithm \ref{alg:kcurves}).
+We call this algorithm \(K\)-curves clustering.
+In this algorithm, the loss function that we wish to minimize is
+\begin{equation}
+\label{eq:kcurves-loss}
+\sum_{k=1}^K \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2
+\end{equation}
+
+\begin{algorithm}[h]
+\label{alg:kcurves}
+\DontPrintSemicolon
+\SetAlgoLined
+\setstretch{1.5}
+\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$, $q$, stopping criterion $\epsilon$}
+\KwResult{Community assignments $1, ..., K$, curves $g_1, ..., g_K$}
+Compute $X$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
+Initialize community labels $z_1, ..., z_n$.\;
+\Repeat {the change in $\sum_k \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2$ is less than $\epsilon$} {
+\For {$k = 1, ..., K$} {
+Define $X_k$ as the rows of $X$ for which $z_i = k$.\;
+Fit curve $g_k$ and positions $t_{k_i}$ to $X_k$ by minimizing $\sum_{k_i} \|x_{k_i} - g_k(t_{k_i})\|^2$.\;
+}
+\For {$k = 1, ..., K$} {
+Assign $z_i \leftarrow \arg\min_\ell \|x_i - g_\ell(t_i)\|^2$.\;
+}
+}
+\caption{$K$-curves clustering.}
+\end{algorithm}
+
+\(K\)-curves clustering either requires knowledge of the functional form of each \(g_k\) or a function that is flexible enough to approximate it.
+The choice of \(g_k\) affects the difficulty of the algorithm.
+As a balance between flexibility and ease of estimation, we consider the case where each \(g_k\) is a Bezier polynomial of degree \(R\) with coefficients \(p_k\).
+Then we have \(g_k(t) = g(t; p_k) = \sum_{r=0}^R p_k^{(r)} \binom{R}{r} (1-t)^{R-r} t^r\).
+
+\begin{example}
+\label{ex:intersect-curves-fit}
+Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
+In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \begin{bmatrix} 0 & 0 \end{bmatrix}^\top$, i.e., the curves meet at the origin. 
+Here, an accuracy of 87.5\% is achieved.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/intersect-curves-2-1}
+
+}
+
+\caption{$K$-curves clustering applied to example \ref{ex:intersect-curves}. The embedding vectors are labeled by predicted community.}\label{fig:intersect-curves-2}
+\end{figure}
+\end{example}
+
+While example \ref{ex:intersect-curves-fit} achieved high accuracy, algorithm \ref{alg:kcurves} often results in poor performance when it is paired with poor initialization.
+In example \ref{ex:intersect-curves-fit}, it just so happened that we randomly chose initial curves that iterated to good fits, but this is not guaranteed.
+We also restricted the curves to meet at the origin, which appears to have helped guide the curves to good fits.
+In figures \ref{fig:loss} and \ref{fig:init-strategies}, we show the results of various initializations on fitting curves to example \ref{ex:intersect-curves}, such as random initialization and spectral clustering.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-7-1}
+
+}
+
+\caption{ASE labeled by estimated community labels for each initialization strategy.
+Upper right: random initialization assuming intersection at the origin.
+Upper middle: random initialization without assuming intersection at the origin.
+Upper right: initialization via spectral clustering assuming intersection at the origin.
+Lower left: initialization via spectral clustering without assuming intersection at the origin.
+Lower middle: initialization with ground truth labels, assuming intersection at the origin.
+Lower right: initialization with ground truth labels, without assuming intersection at the origin.}\label{fig:init-strategies}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-6-1}
+
+}
+
+\caption{Clustering loss vs. iteration for each run of $K$-curve clustering.}\label{fig:loss}
+\end{figure}
+
+Alternatively, if we turn this problem into a semi-supervised problem in which we know a few of the community labels, then the convergence of the ASE guarantees high accuracy.
+
+\begin{theorem}
+\label{thm:kcurves-semisupervised}
+Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$ and its density function is such that $f(t) > a > 0$ for every $t \in [0, 1]$, and each $g_k$ is is a Bezier curve of order $R$.
+Suppose that for each community $k$, we have labels for $R + 1$ vertices. 
+Define the community detection error rate as $\frac{1}{n} \min_\pi \sum_i I(z_i = \pi(\hat{z}_i))$, where $\pi$ is a permutation function and $\hat{z}_i$ are the predicted labels from $K$-curves clustering.
+Then as $n \to \infty$, $K$-curves clustering produces zero community detection error rate, with high probability. 
 \end{theorem}
 
-\hypertarget{sec:intersect}{%
-\subsection{Algorithms for Intersecting Manifolds}\label{sec:intersect}}
-
-\hypertarget{examples-and-simulations}{%
-\subsection{Examples and Simulations}\label{examples-and-simulations}}
+\hypertarget{simulation-study-1}{%
+\subsection{Simulation Study}\label{simulation-study-1}}
+
+For a simulation study, we again turn to examples. \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}.
+The exact simulation setup is as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Draw \(z_1, ..., z_n \iid \Multinomial(1/2, 1/2)\).
+\item
+  Draw \(t_1, ..., t_n \iid \Uniform(0, 1)\).
+\item
+  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top\) and \(g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top\).
+  Collect the latent vectors into matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
+\item
+  Draw \(A \sim \RDPG(X; \rho_n)\) (for these simulations we fix \(\rho_n \equiv 1\)).
+\end{enumerate}
+
+For this simulation study, we set \(n = 128, 256, 512, 1024, 2048\), and we start with completely random initialization as well as initializations in which we know 4 or 8 labels from each community.
+For each setting, we simulate 50 replicates.
+Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing \(p_1^{(0)} = p_2^{(0)} = (0, 0)\).
+Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/sim-curves-1}
+
+}
+
+\caption{Median and IQR error rates over 50 replicates.}\label{fig:sim-curves}
+\end{figure}
+
+We also explored the behavior of random initialization by setting \(n = 512\) and compiling the loss values outputted by \(K\)-curves clustering for each replicate.
+Here, we repeated the simulations \(256\) times.
+Figure \ref{fig:repeat-losses} suggests that in this particular setting, about half the time, random initialization of \(K\)-curves clustering results in the desired curves, while the rest of the time, \(K\)-curves clustering settled on a local minimum.
+Plotting each of the \(256 \times 2\) fitted curves confirms this conclusion (figure \ref{fig:repeat-curves}).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/repeat-losses-1}
+
+}
+
+\caption{Histogram of loss values outputted by $K$-curves clustering over 256 replicates.}\label{fig:repeat-losses}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/repeat-curves-1}
+
+}
+
+\caption{Fitted curves over 256 replicates. The curves are colored by the loss values.}\label{fig:repeat-curves}
+\end{figure}
 
@@
 \subsection{Applications}\label{applications-1}}
+
+In the first example, we applied \(K\)-curves clustering (algorithm \ref{alg:kcurves}) to the Macaque visuotactile brain areas and connections network \citep{https://doi.org/10.1111/j.1460-9568.2006.04678.x}, which is visualized in figure \ref{fig:macaque-graph}.
+In this graph, there are 45 vertices which each correspond to a brain area, and the communities represent each area's tactile function.
+Applying an ASE with \(p = 2\) and \(q = 0\) to this graph results in an embedding that appears to fall approximately on two curves (figure \ref{fig:macaque-ase}) which intersect at the origin.
+Figure \ref{fig:macaque-fitted} shows the result of \(K\)-curves clustering on this graph,
+which yields a 4.4\% error rate.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-graph-1}
+
+}
+
+\caption{Macaque visuotactile brain areas and connections network. The vertices represent brain areas.}\label{fig:macaque-graph}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-ase-1}
+
+}
+
+\caption{ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.}\label{fig:macaque-ase}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-fitted-1}
+
+}
+
+\caption{$K$-means clustering on the ASE of the Macaque brain areas network. The embedding vectors are colored by predicted label.}\label{fig:macaque-fitted}
+\end{figure}
 

### 2022-10-19 draft/draft.Rmd
almost done

@@
 library(ggraph)
+library(mclust)
 source('~/dev/pabm-grdpg/functions.R')
@@
 \begin{example}[Assortativity and disassortativity of the two-community SBM]
+\label{ex:sbm-assortativity}
 Let $P$ be the $n \times n$ probability matrix of a two-community SBM. 
@@
 
-## A Hierarchy of Block Models
+## A Hierarchy of Block Models {#sec:hierarchy}
 
@@
   Parameters = c('$K (K-1) / 2$', '$K (K-1) / 2 + n$', '$n K$'),
-  Dimensions = c('$K$', '$K$', '$K^2$'),
+  Dimension = c('$K$', '$K$', '$K^2$'),
   `Matrix Decomposition` = c('$P = Z B Z^\\top$', '$P = (\\Omega Z) B (\\Omega Z)^\\top$', '$P = X \\Pi X^\\top$'),
@@
 ## Connecting the SBM and DCBM to the GRDPG {#sec:sbm-dcbm-grdpg}
+
+In this section, we connect the SBM and DCBM to the GRDPG and show how this connection is useful for community detection and parameter estimation. 
+We begin by showing that *all* Bernoulli random graphs are GRDPGs.
+
+\begin{proposition}
+Let $A \sim \BG(P; \rho_n)$. 
+Then there exists a $d \leq n$ and $p + q = d$ such that for some $X \in \mathbb{R}^d$, $A \sim \GRDPG_{p, q}(X; \rho_n)$. 
+If $P$ is positive semidefinite, then $A \sim \RDPG(X; \rho_n)$. 
+\end{proposition}
+
+\begin{proof}
+$P$ is the edge probability matrix of a GRDPG if and only if there exists an $X \in \mathbb{R}^{n \times d}$ such that $P = X I_{p,q} X^\top$. 
+$P$ is a square and symmetric matrix, so it has an eigendecomposition $P = V \Lambda V^\top$ where $V \in \mathbb{R}^{n \times d}$, $\Lambda$ is a $d \times d$ diagonal matrix, and $d = \rank(P)$. 
+Let $p$ and $q$ respectively be the number of positive and negative eigenvalues of $P$. 
+Then we can equivalently write $P = V |\Lambda|^{1/2} I_{p,q} |\Lambda|^{1/2} V^\top$ where $|\cdot|$ is applied element-wise. 
+This is exactly the form of the GRDPG with latent vectors $V |\Lambda|^{1/2}$,
+i.e., if $A \sim \BG(P; \rho_n)$, then $A \sim \GRDPG_{p,q}(V |\Lambda|^{1/2}; \rho_n)$. 
+If $P$ is positive semidefinite, then $q = 0$ and we have $P = V \Lambda^{1/2} (V \Lambda^{1/2})^\top$, so $A \sim \RDPG(V \Lambda^{1/2}; \rho_n)$. 
+\end{proof}
+
+The SBM and DCBM are both Bernoulli random graphs, so they are also GRDPG. 
+Recall that the SBM and DCBM are rank $K$. 
+Therefore, we can find $X \in \mathbb{R}^{n \times K}$ that are the latent vectors of the SBM and DCBM. 
+If we analyze the SBM and DCBM as GRDPGs, then we observe community-wise structures in the latent space that are useful for estimation. 
+The latent vectors that generate the SBM consist of $K$ point masses in $\mathbb{R}^K$, each of which correspond to a community, and similarly, the latent vectors that generate the DCBM consist of $K$ line segments emitting from the origin in $\mathbb{R}^K$, each of which correspond to a community, and this property is observed in both assortative and disassortative models \citep{rubindelanchy2017statistical}.
+The ASE recovers these structures with noise, so clustering applied to the ASE can recover the original community labels. 
+\citet{rubindelanchy2017statistical} showed that this approach to community detection for the SBM and DCBM is asymptotically consistent. 
+Note that for a given $P$, $X$ is not unique, i.e., for any $Q \in \mathbb{O}(p, q)$, $X Q$ results in the same $P$ (see remark \ref{rem:non_identifiable}). 
+However, this does not change the general structure of the SBM and DCBM in the latent space. 
+Multiplication by $Q \in \mathbb{O}(p,q)$ does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, because $Q$ is a linear transformation.
+
+\begin{example}
+Consider a two-community assortative SBM. 
+Recall that assortativity in SBMs is characterized by positive definite edge probability matrices. 
+Therefore, this model is also an RDPG. 
+To show this explicitly, suppose that $P$ is organized by community such that the $P^{(k \ell)}$ block corresponds to edges from community $k$ to $\ell$, and consider 
+$$X = \begin{bmatrix}
+\sqrt{\theta_{11}} & 0 \\ 
+\vdots & \vdots \\ 
+\sqrt{\theta_{11}} & 0 \\ 
+\sqrt{\theta_{12}^2 / \theta_{11}} & \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\vdots & \vdots \\
+\sqrt{\theta_{12}^2 / \theta_{11}} & \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\end{bmatrix}.$$
+This describes two point masses in $\mathbb{R}^2$, one at $(\sqrt{\theta_{11}}, 0)$ and another at $(\sqrt{\theta_{12}^2 / \theta_{11}}, \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}})$. 
+Taking the dot product of two vectors in the first point mass results in $\theta_{11}$, taking the dot product of two vectors in the second point mass results in $\theta_{22}$, and taking the dot product of a vector from the first point mass with a vector from the second point mass results in $\theta_{12}$, which correspond exactly to the edge probabilities between communities of the SBM.
+Therefore, $(X X^\top)_{ij} = \theta_{z_i, z_j}$. 
+Note that this latent configuration implicitly assumes that $\theta_{11} \theta_{22} > \theta_{12}^2$, which is the condition we saw for the assortative two-community SBM in example \ref{ex:sbm-assortativity}.
+
+If instead, we have a DCBM, consider the latent configuration $\Omega X$ where $\Omega = \diag(\omega_1, ..., \omega_n)$ is the diagonal matrix of degree parameters. 
+Then $P = \Omega X (\Omega X)^\top$ is the edge probability of a DCBM. 
+Writing out $\Omega X$ explicitly yields:
+$$X = \begin{bmatrix}
+\omega_1 \sqrt{\theta_{11}} & 0 \\ 
+\vdots & \vdots \\ 
+\omega_{n_1} \sqrt{\theta_{11}} & 0 \\ 
+\omega_{n_1 + 1} \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_{n_1 + 1} \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\vdots & \vdots \\
+\omega_n \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_n \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\end{bmatrix}.$$
+The resulting rows of $X$ are point masses that have been stretched out, turning into rays. 
+\end{example}
+
+\begin{example}
+Recall the UK faculty network in \ref{ex:uk-faculty}. 
+If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
+We will assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
+Because the fourth community only contains two members, we omit it from this analysis. 
+Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM. 
+Fitting three line segments to these data and clustering the embedding vectors by which line segment is closest results in a community detection accuracy of 83.5\%.
+
+```{r ukfaculty-ase, fig.cap = 'The ASE of the UK faculty network. The vertices are labeled by the school to which each faculty member belongs.', fig.height = 3, fig.width = 4, cache = TRUE }
+data(UKfaculty)
+A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
+A <- sign(A + t(A))
+z <- factor(igraph::vertex_attr(UKfaculty)$Group)
+A <- A[z != 4, z != 4]
+z <- z[z != 4]
+Xhat <- embedding(A, 3, 0)
+pairs(Xhat, col = z, asp = 1)
+```
+\end{example}
+
+The PABM is also a Bernoulli random graph, so it is also a GRDPG. 
+We explore this in the next section (see section \@ref(sec:pabm-grdpg)).
 
@@
 
+### Previous Work
+
 ## Algorithm for Nonintersecting Manifolds {#sec:nonintersect}
@@
 Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting. 
-
-```{r sim-curves, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates.'}
+The simulation results suggest that for large $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.
+
+```{r sim-curves, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates.', cache = TRUE}
 n.vec <- 2 ^ c(7, 8, 9, 10, 11)
@@
 
-```{r repeat-losses, fig.height = 3, fig.width = 8, fig.cap = 'Histogram of loss values outputted by $K$-curves clustering over 256 replicates.'}
+```{r repeat-losses, fig.height = 3, fig.width = 8, fig.cap = 'Histogram of loss values outputted by $K$-curves clustering over 256 replicates.', cache = TRUE}
 repeat.df <- readr::read_csv('~/dev/manifold-block-models/simulations/repeat-study/repeat-study.csv')
@@
 
-```{r repeat-curves, fig.height = 3, fig.width = 3, fig.cap = 'Fitted curves over 256 replicates. The curves are colored by the loss values.'}
+```{r repeat-curves, fig.height = 3, fig.width = 3, fig.cap = 'Fitted curves over 256 replicates. The curves are colored by the loss values.', cache = TRUE}
 plot.repeat.study <- function(repeat.df) {
@@
 
-In the second example, we consider the Harry Potter enmity graph (figure \ref{fig:hp}).
-
-```{r hp, fig.cap = 'Harry Potter enmity graph.
-The vertices represent characters, the edges represent whether pairs of characters are enemies, and the communities represent whether the characters are on the side of the protagonist or the antagonist.', fig.width = 4, fig.height = 2}
+In the second example, we considered the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}. 
+This graph does not come with ground truth labels. 
+\citet{rubindelanchy2017statistical} and \citet{SannaPassino2022} analyzed this graph as a GRDPG, and the ASE suggests that an appropriate model for this graph is a disassortative DCBM with $K = 2$ communities.
+Indeed, if we construct an ASE with $p = 1$ and $q = 1$ and label the points by character name, we see embedding vectors lie neatly along two rays emitting from the origin, with one ray populated by protagonists and another ray populated by antagonists (figure \ref{fig:hp-ase}).
+Nevertheless, we tried fitting an MBM on this graph via $K$-curves clustering. 
+Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM. 
+We initialized $K$-curves clustering by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known *a priori*. 
+The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.
+
+```{r hp, fig.cap = 'Harry Potter enmity graph. The vertices represent characters and the edges represent whether pairs of characters are enemies.', fig.width = 4, fig.height = 2}
 relations.df <- readr::read_csv('~/dev/potter-network/data/relations.csv') %>% 
@@
 
-ggraph(A, layout = 'stress') + 
-  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
-  geom_node_point(aes(colour = factor(z)), 
-                  show.legend = FALSE) + 
-  labs(colour = NULL) + 
-  theme_void() + 
-  scale_colour_brewer(palette = 'Set1')
+qgraph::qgraph(A, layout = 'spring')
 ```
+
+```{r hp-ase, fig.cap = 'ASE of the Harry Potter emnity graph.
+The way in which the embedding vectors are organized suggest that a DCBM is an appropriate model for this graph.', fig.height = 3, fig.width = 10}
+Xhat <- embedding(A, p = 1, q = 1)
+
+ggplot() + 
+  geom_point(aes(x = Xhat[, 2], y = Xhat[, 1])) + 
+  ggrepel::geom_text_repel(aes(x = Xhat[, 2], y = Xhat[, 1], 
+                               label = characters.df$name[deg.A > 0])) + 
+  # coord_fixed() + 
+  theme_bw() + 
+  labs(x = expression(x[2]), y = expression(x[1]))
+```
+
+```{r hp-kcurves, fig.cap = 'ASE of the Harry Potter emnity graph with quadratic Bezier curves fitted via $K$-curves clustering.
+The embedding vectors are colored by predicted label.', fig.height = 3, fig.width = 10, cache = TRUE}
+z.init <- rep(NA, nrow(A))
+z.init[rownames(A) == 'Harry Potter'] <- 1
+z.init[rownames(A) == 'Hermione Granger'] <- 1
+z.init[rownames(A) == 'Lord Voldemort'] <- 2
+z.init[rownames(A) == 'Peter Pettigrew'] <- 2
+
+hp.out <- manifold.clustering(Xhat[, c(2, 1)], 2, 
+                              initialization = z.init,
+                              intercept = FALSE, 
+                              parallel = TRUE, 
+                              verbose = FALSE, 
+                              animate = FALSE, 
+                              curve.init = 'x')
+
+plot.estimated.curves(Xhat[, c(2, 1)], hp.out) + 
+  ggrepel::geom_text_repel(aes(x = Xhat[, 2], y = Xhat[, 1], 
+                               label = characters.df$name[deg.A > 0],
+                               colour = factor(hp.out$z))) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  coord_fixed(.5) + 
+  labs(x = expression(x[2]), y = expression(x[1]))
+```
+
+In the final example,
 

### 2022-10-19 draft/draft.tex
almost done

@@
 
-\hypertarget{a-hierarchy-of-block-models}{%
-\subsection{A Hierarchy of Block Models}\label{a-hierarchy-of-block-models}}
+\hypertarget{sec:hierarchy}{%
+\subsection{A Hierarchy of Block Models}\label{sec:hierarchy}}
 
@@
 \hline
-Model & Parameters & Dimensions & Matrix Decomposition & Reduction to simpler model\\
+Model & Parameters & Dimension & Matrix Decomposition & Reduction to simpler model\\
 \hline
@@
 \subsection{Connecting the SBM and DCBM to the GRDPG}\label{sec:sbm-dcbm-grdpg}}
+
+In this section, we connect the SBM and DCBM to the GRDPG and show how this connection is useful for community detection and parameter estimation.
+We begin by showing that \emph{all} Bernoulli random graphs are GRDPGs.
+
+\begin{proposition}
+Let $A \sim \BG(P; \rho_n)$. 
+Then there exists a $d \leq n$ and $p + q = d$ such that for some $X \in \mathbb{R}^d$, $A \sim \GRDPG_{p, q}(X; \rho_n)$. 
+If $P$ is positive semidefinite, then $A \sim \RDPG(X; \rho_n)$. 
+\end{proposition}
+
+\begin{proof}
+$P$ is the edge probability matrix of a GRDPG if and only if there exists an $X \in \mathbb{R}^{n \times d}$ such that $P = X I_{p,q} X^\top$. 
+$P$ is a square and symmetric matrix, so it has an eigendecomposition $P = V \Lambda V^\top$ where $V \in \mathbb{R}^{n \times d}$, $\Lambda$ is a $d \times d$ diagonal matrix, and $d = \rank(P)$. 
+Let $p$ and $q$ respectively be the number of positive and negative eigenvalues of $P$. 
+Then we can equivalently write $P = V |\Lambda|^{1/2} I_{p,q} |\Lambda|^{1/2} V^\top$ where $|\cdot|$ is applied element-wise. 
+This is exactly the form of the GRDPG with latent vectors $V |\Lambda|^{1/2}$,
+i.e., if $A \sim \BG(P; \rho_n)$, then $A \sim \GRDPG_{p,q}(V |\Lambda|^{1/2}; \rho_n)$. 
+If $P$ is positive semidefinite, then $q = 0$ and we have $P = V \Lambda^{1/2} (V \Lambda^{1/2})^\top$, so $A \sim \RDPG(V \Lambda^{1/2}; \rho_n)$. 
+\end{proof}
+
+The SBM and DCBM are both Bernoulli random graphs, so they are also GRDPG.
+Recall that the SBM and DCBM are rank \(K\).
+Therefore, we can find \(X \in \mathbb{R}^{n \times K}\) that are the latent vectors of the SBM and DCBM.
+If we analyze the SBM and DCBM as GRDPGs, then we observe community-wise structures in the latent space that are useful for estimation.
+The latent vectors that generate the SBM consist of \(K\) point masses in \(\mathbb{R}^K\), each of which correspond to a community, and similarly, the latent vectors that generate the DCBM consist of \(K\) line segments emitting from the origin in \(\mathbb{R}^K\), each of which correspond to a community, and this property is observed in both assortative and disassortative models \citep{rubindelanchy2017statistical}.
+The ASE recovers these structures with noise, so clustering applied to the ASE can recover the original community labels.
+\citet{rubindelanchy2017statistical} showed that this approach to community detection for the SBM and DCBM is asymptotically consistent.
+Note that for a given \(P\), \(X\) is not unique, i.e., for any \(Q \in \mathbb{O}(p, q)\), \(X Q\) results in the same \(P\) (see remark \ref{rem:non_identifiable}).
+However, this does not change the general structure of the SBM and DCBM in the latent space.
+Multiplication by \(Q \in \mathbb{O}(p,q)\) does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, because \(Q\) is a linear transformation.
+
+\begin{example}
+Consider a two-community assortative SBM. 
+Recall that assortativity in SBMs is characterized by positive definite edge probability matrices. 
+Therefore, this model is also an RDPG. 
+To show this explicitly, suppose that $P$ is organized by community such that the $P^{(k \ell)}$ block corresponds to edges from community $k$ to $\ell$, and consider 
+$$X = \begin{bmatrix}
+\sqrt{\theta_{11}} & 0 \\ 
+\vdots & \vdots \\ 
+\sqrt{\theta_{11}} & 0 \\ 
+\sqrt{\theta_{12}^2 / \theta_{11}} & \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\vdots & \vdots \\
+\sqrt{\theta_{12}^2 / \theta_{11}} & \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\end{bmatrix}.$$
+This describes two point masses in $\mathbb{R}^2$, one at $(\sqrt{\theta_{11}}, 0)$ and another at $(\sqrt{\theta_{12}^2 / \theta_{11}}, \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}})$. 
+Taking the dot product of two vectors in the first point mass results in $\theta_{11}$, taking the dot product of two vectors in the second point mass results in $\theta_{22}$, and taking the dot product of a vector from the first point mass with a vector from the second point mass results in $\theta_{12}$, which correspond exactly to the edge probabilities between communities of the SBM.
+Therefore, $(X X^\top)_{ij} = \theta_{z_i, z_j}$. 
+Note that this latent configuration implicitly assumes that $\theta_{11} \theta_{22} > \theta_{12}^2$, which is the condition we saw for the assortative two-community SBM in example \ref{ex:sbm-assortativity}.
+
+If instead, we have a DCBM, consider the latent configuration $\Omega X$ where $\Omega = \diag(\omega_1, ..., \omega_n)$ is the diagonal matrix of degree parameters. 
+Then $P = \Omega X (\Omega X)^\top$ is the edge probability of a DCBM. 
+Writing out $\Omega X$ explicitly yields:
+$$X = \begin{bmatrix}
+\omega_1 \sqrt{\theta_{11}} & 0 \\ 
+\vdots & \vdots \\ 
+\omega_{n_1} \sqrt{\theta_{11}} & 0 \\ 
+\omega_{n_1 + 1} \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_{n_1 + 1} \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\vdots & \vdots \\
+\omega_n \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_n \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\end{bmatrix}.$$
+The resulting rows of $X$ are point masses that have been stretched out, turning into rays. 
+\end{example}
+
+\begin{example}
+Recall the UK faculty network in \ref{ex:uk-faculty}. 
+If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
+We will assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
+Because the fourth community only contains two members, we omit it from this analysis. 
+Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM. 
+Fitting three line segments to these data and clustering the embedding vectors by which line segment is closest results in a community detection accuracy of 83.5\%.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/ukfaculty-ase-1}
+
+}
+
+\caption{The ASE of the UK faculty network. The vertices are labeled by the school to which each faculty member belongs.}\label{fig:ukfaculty-ase}
+\end{figure}
+\end{example}
+
+The PABM is also a Bernoulli random graph, so it is also a GRDPG.
+We explore this in the next section (see section \ref{sec:pabm-grdpg}).
 
@@
 
+\hypertarget{previous-work-1}{%
+\subsubsection{Previous Work}\label{previous-work-1}}
+
 \hypertarget{sec:nonintersect}{%
@@
 Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting.
+The simulation results suggest that for large \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.
 
@@
 \end{figure}
+
+In the second example, we considered the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}.
+This graph does not come with ground truth labels.
+\citet{rubindelanchy2017statistical} and \citet{SannaPassino2022} analyzed this graph as a GRDPG, and the ASE suggests that an appropriate model for this graph is a disassortative DCBM with \(K = 2\) communities.
+Indeed, if we construct an ASE with \(p = 1\) and \(q = 1\) and label the points by character name, we see embedding vectors lie neatly along two rays emitting from the origin, with one ray populated by protagonists and another ray populated by antagonists (figure \ref{fig:hp-ase}).
+Nevertheless, we tried fitting an MBM on this graph via \(K\)-curves clustering.
+Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM.
+We initialized \(K\)-curves clustering by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known \emph{a priori}.
+The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/hp-1}
+
+}
+
+\caption{Harry Potter enmity graph. The vertices represent characters and the edges represent whether pairs of characters are enemies.}\label{fig:hp}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/hp-ase-1}
+
+}
+
+\caption{ASE of the Harry Potter emnity graph. The way in which the embedding vectors are organized suggest that a DCBM is an appropriate model for this graph.}\label{fig:hp-ase}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/hp-kcurves-1}
+
+}
+
+\caption{ASE of the Harry Potter emnity graph with quadratic Bezier curves fitted via $K$-curves clustering. The embedding vectors are colored by predicted label.}\label{fig:hp-kcurves}
+\end{figure}
+
+In the final example,
 

### 2022-10-19 draft/draft.Rmd
almost done

@@
 
+Manifolds have been observed in the ASEs of real observed graphs when they are studied as GRDPGs. 
+The *Drosophila* connectome graph \citep{Eichler141762} displays community-wise linear and quadratic structures in the ASE. 
+\citet{https://doi.org/10.48550/arxiv.1705.03297} used GMMs on the ASE for community detection and found that the curve-like structures in the ASE were incompatible with centroid-based clustering methods.
+\citet{SannaPassino2022} used Gaussian Processes to estimate curves, which resulted in better empirical fits. 
+\citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on the manifolds rather than the manifolds themselves.
+
+Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in $\mathbb{R}^d$ and identifying vectors belonging to different manifolds in $\mathbb{R}^d$.
+
 ## Algorithm for Nonintersecting Manifolds {#sec:nonintersect}
@@
 \label{eq:kcurves-loss}
-\sum_{k=1}^K \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2
+L(z_1, ..., z_n, g_1, ..., g_n; X) = \sum_{k=1}^K \sum_{i: z_i = k} \|x_i - g_k(t_i)\|^2
 \end{equation}
@@
 Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$ and its density function is such that $f(t) > a > 0$ for every $t \in [0, 1]$, and each $g_k$ is is a Bezier curve of order $R$.
-Suppose that for each community $k$, we have labels for $R + 1$ vertices. 
-Define the community detection error rate as $\frac{1}{n} \min_\pi \sum_i I(z_i = \pi(\hat{z}_i))$, where $\pi$ is a permutation function and $\hat{z}_i$ are the predicted labels from $K$-curves clustering.
-Then as $n \to \infty$, $K$-curves clustering produces zero community detection error rate, with high probability. 
+Suppose that for each community $k$, we have labels for at least $R + 1$ vertices. 
+Then as $n \to \infty$, the estimates outputted by $K$-curves clustering are such that 
+$$L(\hat{z}_1, ..., \hat{z}_n, \hat{g}_1, ..., \hat{g}_n; X) \stackrel{p}{\to} 0,$$
+where $L$ is defined as in equation \ref{eq:kcurves-loss}. 
 \end{theorem}
@@
 
-In the final example,
+In the final example, we examined the *Drosophila* connectome graph \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
+In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types, Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons. 
+In our analysis, we only considered neurons labeled as Kenyon Cells and Projection Neurons in the right hemisphere. 
+The resulting graph is illustrated in figure \ref{fig:mbconnectome-graph}.
+
+The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for this graph (figure \ref{fig:mbconnectome-ase}). 
+Fitting two quadratic Bezier curves to the ASE (figure \ref{fig:mbconnectome-kcurves}) results in 91.4\% accuracy.
+
+```{r mbconnectome-graph, fig.cap = 'Drosophila connectome graph. Each vertex represents a neuron, which are labeled by neuron type.', fig.height = 2, figh.width = 4}
+library(mbstructure)
+data(MBconnectome)
+
+graph <- generate.graph(newrdat, vdf.right)
+A <- as.matrix(igraph::as_adjacency_matrix(graph$g, type = 'both'))
+z <- graph$vdf$type
+ind.keep <- z %in% c('KC', 'PN')
+A <- A[ind.keep, ind.keep]
+z <- as.numeric(as.factor(as.numeric(z[ind.keep])))
+A <- sign(A + t(A))
+diag(A) <- 0
+
+ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
+
+```{r mbconnectome-ase, fig.cap = 'ASE of the Drosophila connectome graph. The embedding vectors are labeled by neuron type.', fig.height = 2, figh.width = 4}
+Xhat <- embedding(A, 2, 0)
+
+ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2],
+                 colour = factor(z))) + 
+  coord_fixed() +
+  theme_bw() + 
+  scale_colour_brewer(palette = 'Set1') + 
+  labs(x = expression(x[1]), y = expression(x[2]),
+       colour = NULL)
+```
+
+```{r mbconnectome-kcurves, fig.cap = 'ASE of the Drosophila connectome graph with quadratic Bezier curves fitted via $K$-curves clustering.
+The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 4, cache = TRUE}
+set.seed(314159)
+z.init <- sample.points(z, 3)
+
+mbconnectome.out <- manifold.clustering(Xhat, 2, 
+                              initialization = z.init,
+                              intercept = FALSE, 
+                              parallel = TRUE, 
+                              verbose = FALSE, 
+                              animate = FALSE, 
+                              curve.init = 'x')
+
+plot.estimated.curves(Xhat, mbconnectome.out) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  # coord_fixed(.5) + 
+  labs(x = expression(x[1]), y = expression(x[2]))
+```
 

### 2022-10-19 draft/draft.tex
almost done

@@
 
+Manifolds have been observed in the ASEs of real observed graphs when they are studied as GRDPGs.
+The \emph{Drosophila} connectome graph \citep{Eichler141762} displays community-wise linear and quadratic structures in the ASE.
+\citet{https://doi.org/10.48550/arxiv.1705.03297} used GMMs on the ASE for community detection and found that the curve-like structures in the ASE were incompatible with centroid-based clustering methods.
+\citet{SannaPassino2022} used Gaussian Processes to estimate curves, which resulted in better empirical fits.
+\citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on the manifolds rather than the manifolds themselves.
+
+Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in \(\mathbb{R}^d\) and identifying vectors belonging to different manifolds in \(\mathbb{R}^d\).
+
 \hypertarget{sec:nonintersect}{%
@@
 
-In the final example,
+In the final example, we examined the \emph{Drosophila} connectome graph \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
+In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types, Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons.
+In our analysis, we only considered neurons labeled as Kenyon Cells and Projection Neurons in the right hemisphere.
+The resulting graph is illustrated in figure \ref{fig:mbconnectome-graph}.
+
+The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for this graph (figure \ref{fig:mbconnectome-ase}).
+Fitting two quadratic Bezier curves to the ASE (figure \ref{fig:mbconnectome-kcurves}) results in 91.4\% accuracy.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/mbconnectome-graph-1}
+
+}
+
+\caption{Drosophila connectome graph. Each vertex represents a neuron, which are labeled by neuron type.}\label{fig:mbconnectome-graph}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/mbconnectome-ase-1}
+
+}
+
+\caption{ASE of the Drosophila connectome graph. The embedding vectors are labeled by neuron type.}\label{fig:mbconnectome-ase}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/mbconnectome-kcurves-1}
+
+}
+
+\caption{ASE of the Drosophila connectome graph with quadratic Bezier curves fitted via $K$-curves clustering. The embedding vectors are colored by predicted label.}\label{fig:mbconnectome-kcurves}
+\end{figure}
 

### 2022-10-19 draft/draft.Rmd
almost done

@@
 
-```{r mbconnectome-ase, fig.cap = 'ASE of the Drosophila connectome graph. The embedding vectors are labeled by neuron type.', fig.height = 2, figh.width = 4}
+```{r mbconnectome-ase, fig.cap = 'ASE of the Drosophila connectome graph. The embedding vectors are labeled by neuron type.', fig.height = 2, fig.width = 8}
 Xhat <- embedding(A, 2, 0)
@@
                  colour = factor(z))) + 
-  coord_fixed() +
+  coord_fixed(.5) +
   theme_bw() + 
+  theme(legend.position = 'none') +
   scale_colour_brewer(palette = 'Set1') + 
@@
 ```{r mbconnectome-kcurves, fig.cap = 'ASE of the Drosophila connectome graph with quadratic Bezier curves fitted via $K$-curves clustering.
-The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 4, cache = TRUE}
+The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 8, cache = TRUE}
 set.seed(314159)
@@
   scale_colour_brewer(palette = 'Set1') + 
-  # coord_fixed(.5) + 
+  coord_fixed(.5) +
   labs(x = expression(x[1]), y = expression(x[2]))
@@
 
+\begin{proof}[Proof of theorem \ref{thm:kcurves-semisupervised}]
+\end{proof}
+
 \newpage

### 2022-10-19 draft/draft.Rmd
almost done

@@
 \label{eq:kcurves-loss}
-L(z_1, ..., z_n, g_1, ..., g_n; X) = \sum_{k=1}^K \sum_{i: z_i = k} \|x_i - g_k(t_i)\|^2
+L(z_1, ..., z_n, g_1, ..., g_n; X) = \frac{1}{n} \sum_{k=1}^K \sum_{i: z_i = k} \|x_i - g_k(t_i)\|^2
 \end{equation}
@@
 \label{thm:kcurves-semisupervised}
-Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$ and its density function is such that $f(t) > a > 0$ for every $t \in [0, 1]$, and each $g_k$ is is a Bezier curve of order $R$.
+Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a Bezier curve of order $R$. 
 Suppose that for each community $k$, we have labels for at least $R + 1$ vertices. 
-Then as $n \to \infty$, the estimates outputted by $K$-curves clustering are such that 
+Then if $n \rho_n = \omega(\log^{4c} n)$, as $n \to \infty$, the estimates outputted by $K$-curves clustering are such that 
 $$L(\hat{z}_1, ..., \hat{z}_n, \hat{g}_1, ..., \hat{g}_n; X) \stackrel{p}{\to} 0,$$
-where $L$ is defined as in equation \ref{eq:kcurves-loss}. 
+where $L$ is the loss function defined in equation (\ref{eq:kcurves-loss}). 
 \end{theorem}
@@
 In the final example, we examined the *Drosophila* connectome graph \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
-In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types, Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons. 
+In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types (Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons). 
 In our analysis, we only considered neurons labeled as Kenyon Cells and Projection Neurons in the right hemisphere. 
@@
 \begin{proof}[Proof of theorem \ref{thm:kcurves-semisupervised}]
+Let $\hat{X}$ with rows $\hat{x}_i^\top$ be the ASE of $A$ drawn from an MBM in which each $g_k(t) = g(t; p_k)$ is a Bezier curve of order $R$. 
+Define the ASE error as $\epsilon_i = \|Q \hat{x}_i - x_i\|$ and $\epsilon = \max_i \epsilon_i$ where $Q$ is the appropriate indefinite orthogonal transformation for the ASE. 
+By \citet{rubindelanchy2017statistical}, $\epsilon = O_P \bigg(\frac{\log^c n}{\sqrt{n}} \bigg)$.
+
+Since $x_i = g(t_i; p_{z_i})$, $\|Q \hat{x}_i - g(t_i; p_{z_i})\| < \epsilon$.
+
+The statement of the theorem assumes that for each community $k$, there at least $R + 1$ points with known label $k$. 
+For simplicity, suppose we have exactly $R + 1$ (if there are more, then discard the rest). 
+Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly (see appendix B), yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial.
+Use the fitted Bezier curves to define $\delta_k(t) = \|Q g(t; \hat{p}_k) - g(t; p_k)\|$, $\delta_k = \max_t \delta_k(t)$, and $\delta = \max_k \delta_k$. 
+Since $P(\epsilon < M n^{-1/2} \log^c n) \to 1$ for constants $M, c > 0$, 
+$P(\delta_k < M n^{-1/2} \log^c n) \to 1$ for each $k$. 
+Since $\epsilon_i$ describes the error between each embedding point and its corresponding latent vertex, and $\delta_k(t_i)$ describes the error between the fitted curve and its corresponding true curve in the latent space,
+$L < \frac{1}{n} \sum_i \epsilon_i + \delta_{z_i}(t_i) < \frac{1}{n} n (\epsilon + \delta) = \epsilon + \delta$. 
+Then $P(L < 2 M n^{-1/2} \log^c n) > P(\epsilon + \delta < 2 M n^{-1/2} \log^c n) \to 1$ as $n \to \infty$.
+
 \end{proof}

### 2022-10-19 draft/draft.tex
almost done

@@
 In the final example, we examined the \emph{Drosophila} connectome graph \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
-In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types, Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons.
+In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types (Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons).
 In our analysis, we only considered neurons labeled as Kenyon Cells and Projection Neurons in the right hemisphere.
@@
 
+\begin{proof}[Proof of theorem \ref{thm:kcurves-semisupervised}]
+Let $\hat{X}$ with rows $\hat{x}_i^\top$ be the ASE of $A$ drawn from an MBM in which each $g_k(t) = g(t; p_k)$ is a Bezier curve of order $R$. 
+Define the ASE error as $\epsilon_i = \|Q \hat{x}_i - x_i\|$ and $\epsilon = \max_i \epsilon_i$ where $Q$ is the appropriate indefinite orthogonal transformation for the ASE. 
+By \citet{rubindelanchy2017statistical}, $\epsilon = O_P \bigg(\frac{\log^c n}{\sqrt{n}} \bigg)$.
+
+Since $x_i = g(t_i; p_{z_i})$, $\|Q \hat{x}_i - g(t_i; p_{z_i})\| < \epsilon$.
+
+The statement of the theorem assumes that for each community $k$, there at least $R + 1$ points with known label $k$. 
+For simplicity, suppose we have exactly $R + 1$ (if there are more, then discard the rest). 
+Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly (see appendix B), yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial.
+Use the fitted Bezier curves to define $\delta_k(t) = \|Q g(t; \hat{p}_k) - g(t; p_k)\|$, $\delta_k = \max_t \delta_k(t)$, and $\delta = \max_k \delta_k$. 
+Since $P(\epsilon < M n^{-1/2} \log^c n) \to 1$ for constants $M, c > 0$, 
+$P(\delta_k < M n^{-1/2} \log^c n) \to 1$ for each $k$. 
+Since $\epsilon_i$ describes the error between each embedding point and its corresponding latent vertex, and $\delta_k(t_i)$ describes the error between the fitted curve and its corresponding true curve in the latent space,
+$L < \frac{1}{n} \sum_i \epsilon_i + \delta_{z_i}(t_i) < \frac{1}{n} n (\epsilon + \delta) = \epsilon + \delta$. 
+Then $P(L < 2 M n^{-1/2} \log^c n) > P(\epsilon + \delta < 2 M n^{-1/2} \log^c n) \to 1$ as $n \to \infty$.
+
+\end{proof}
+
 \newpage
@@
 \section*{Appendix C: Supplementary Materials}
-
-\newpage
 

### 2022-10-19 draft/draft.Rmd
almost done

@@
 
-\section*{Appendix A: Proofs of Theorems}
+\section*{Appendix A: Proofs of Theorems from Section 3}
 
@@
 
+\section*{Appendix B: Proofs of Theorems from Section 4}
+
 \begin{lemma}
@@
 For simplicity, suppose we have exactly $R + 1$ (if there are more, then discard the rest). 
-Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly (see appendix B), yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial.
+Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly (see appendix C), yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial.
 Use the fitted Bezier curves to define $\delta_k(t) = \|Q g(t; \hat{p}_k) - g(t; p_k)\|$, $\delta_k = \max_t \delta_k(t)$, and $\delta = \max_k \delta_k$. 
@@
 Since $\epsilon_i$ describes the error between each embedding point and its corresponding latent vertex, and $\delta_k(t_i)$ describes the error between the fitted curve and its corresponding true curve in the latent space,
-$L < \frac{1}{n} \sum_i \epsilon_i + \delta_{z_i}(t_i) < \frac{1}{n} n (\epsilon + \delta) = \epsilon + \delta$. 
-Then $P(L < 2 M n^{-1/2} \log^c n) > P(\epsilon + \delta < 2 M n^{-1/2} \log^c n) \to 1$ as $n \to \infty$.
+$$L < \frac{1}{n} \sum_i \epsilon_i + \delta_{z_i}(t_i) < \frac{1}{n} n (\epsilon + \delta) = \epsilon + \delta.$$ 
+Then as $n \to \infty$, 
+$$P(L < 2 M n^{-1/2} \log^c n) > P(\epsilon + \delta < 2 M n^{-1/2} \log^c n) \to 1.$$
 
@@
 
-\section*{Appendix B: Details on Fitting Bezier Curves to Noisy Data}
+\section*{Appendix C: Details on Fitting Bezier Curves to Noisy Data}
 
@@
 
-\section*{Appendix C: Supplementary Materials}
-
 # References

### 2022-10-19 draft/draft.tex
almost done

@@
 
-\section*{Appendix B: Details on Fitting Bezier Curves to Noisy Data}
+\section*{Appendix C: Details on Fitting Bezier Curves to Noisy Data}
 
 \newpage
-
-\section*{Appendix C: Supplementary Materials}
 

### 2022-10-26 draft/draft.Rmd
minor tweak to proof of theorem 4.2
minor updates
adding in detaisls on bezier curve fitting

@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model. 
-Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements.
+Data are often represented as an $n \times p$ matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements.
 An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$.
@@
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector $x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top \in \mathbb{R}^p$, we observe $a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
+Instead of feature vector $x_i = \big[ x_{i1} \mid \cdots \mid x_{ip} \big]^\top \in \mathbb{R}^p$, we observe $a_i = \big[ a_{i1} \mid \cdots \mid a_{in} \big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
 For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
@@
 
-To estimate $\omega = \begin{bmatrix} \omega_1 & \cdots & \omega_n \end{bmatrix}^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
+To estimate $\omega = \big[ \omega_1, \cdots, \omega_n \big]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
 For simplicity, set $C = 1$. 
@@
 The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}_K$.
-Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \begin{bmatrix} \alpha_1 & \cdots & \alpha_K \end{bmatrix}^\top$ and $\{\theta_{k \ell}\}_K$.
+Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \big[ \alpha_1, \cdots, \alpha_K \big]^\top$ and $\{\theta_{k \ell}\}_K$.
 
@@
 Denote $d = p + q$ and let $\mathcal{X}$ be the subset of $\mathbb{R}^d$ such that, for any $x, y \in \mathcal{X}$, we have $x^\top I_{p,q} y \in [0, 1]$. 
-Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
+Let $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
 A graph $G$ with adjacency matrix $A$ is said to be a generalized random dot product graph with latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$ if $A \sim \BG(P)$ where the edge probability matrix $P$ is given by $P = \rho_n X I_{p,q} X^\top$, i.e., the entries of $P$ are of the form $P_{ij} = \rho_n x_i^\top I_{p,q} x_j$.
@@
 Suppose that latent vectors for an RDPG are drawn from the curve 
-$x(t) = \begin{bmatrix} t^2 & 2 t (t-1) \end{bmatrix}^\top$, $t \in [0, 1]$. 
+$x(t) = \big[ t^2, 2 t (t-1) \big]^\top$, $t \in [0, 1]$. 
 Let $x_i = x(t_i)$. 
@@
 In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
-This corresponds to a one-dimensional RDPG with a latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$. 
-In another view, the RDPG with latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
-
-Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \begin{bmatrix} \cos \frac{\pi}{2} t & \sin \frac{\pi}{2} t \end{bmatrix}^\top$ for $t \in [0, 1]$.
+This corresponds to a one-dimensional RDPG with a latent structure $X = \big[ \sqrt{p},  \cdots, \sqrt{p} \big]^\top$. 
+In another view, the RDPG with latent structure $X = \big[ \sqrt{p}, \cdots, \sqrt{p} \big]^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
+
+Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \big[ \cos \frac{\pi}{2} t, \sin \frac{\pi}{2} t \big]^\top$ for $t \in [0, 1]$.
 Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$.
@@
 \item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
-\item Compile the latent vectors into data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
+\item Compile the latent vectors into data matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
 \end{enumerate}
@@
 \label{ex:separated-curves}
-Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
-Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \big[ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) \big]^\top$ and $f_2(t) = \big[ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) \big]^\top$.
+Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
 Finally, let $A \sim \RDPG(X)$.
@@
 Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$.
-Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
+Let $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
 Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
@@
 2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
-3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
+3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
 4. Draw $A \sim \GRDPG_{p,q}(X; \rho_n)$.
@@
 \label{ex:intersect-curves}
-Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$.
+Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top$ and $g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top$.
 This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
@@
 Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
-In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \begin{bmatrix} 0 & 0 \end{bmatrix}^\top$, i.e., the curves meet at the origin. 
+In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \big[ 0, 0 \big]^\top$, i.e., the curves meet at the origin. 
 Here, an accuracy of 87.5\% is achieved.
@@
 
-For a simulation study, we again turn to examples.  \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}. 
-The exact simulation setup is as follows:
+We performed two simulation experiments for algorithm \ref{alg:kcurves}.
+In the first simulation setup, we again turn to examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}. 
+The setup is as follows:
 
@@
 2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
-3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$. 
-Collect the latent vectors into matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
+3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top$ and $g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top$. 
+Collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$. 
 4. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
@@
 Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing $p_1^{(0)} = p_2^{(0)} = (0, 0)$. 
-Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting. 
-The simulation results suggest that for large $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.
-
-```{r sim-curves, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates.', cache = TRUE}
+Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates for each $n$ and *a priori* known labels per community.
+The simulation results suggest that for sufficient $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.
+
+```{r sim-curves, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates in the simulation setting for $K = 2$ curves.', cache = TRUE}
 n.vec <- 2 ^ c(7, 8, 9, 10, 11)
@@
 
-clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced.csv')
+clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced-2/balanced.csv')
 
@@
 
+In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves. 
+The setup is as follows:
+
+1. Draw $z_1, ..., z_n \iid \Multinomial(1/3, 1/3, 1/3)$.
+2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
+3. Let each $x_i = g_{z_i}(t_i)$ where
+    i. $g_1(t) = \big[ 2 t (t-1), t^2, 0 \big]^\top$
+    ii. $g_2(t) = \big[ 0, t^2, 2 t (t-1) \big]^\top$
+    iii. $g_3(t) = \big[2 t (t-1), 0, t^2 \big]^\top$.
+4. Collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
+5. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
+
+As in the previous setup, we set $n = 128, 256, 512, 1024, 2048$, and we start with completely random initializations as well as initializations in which we know $4$ or $8$ labels from each community.
+For each setting, we simulate 50 replicates. 
+Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing $p_1^{(k)} = (0, 0, 0)$ for each $k$. 
+Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each $n$ and *a priori* known labels per community. 
+As in the previous $K = 2$ setup, the results suggest that for sufficient $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.
+Unlike in the previous setup, random initialization fails to improve with greater $n$.
+
+```{r sim-curves-3, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates in the simulation setup for $K = 3$ curves.', cache = TRUE}
+n.vec <- 2 ^ c(7, 8, 9, 10, 11)
+nsamp.vec <- c(0, 4, 8)
+
+clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced-3/balanced-3.csv')
+
+clust.summary.df <- clustering.df %>% 
+  dplyr::group_by(n, nsamp) %>% 
+  dplyr::summarise(
+    med.count = median(error.count),
+    first.q.count = quantile(error.count, .25),
+    third.q.count = quantile(error.count, .75),
+    med.rate = median(error.rate),
+    first.q.rate = quantile(error.rate, .25),
+    third.q.rate = quantile(error.rate, .75)
+  ) %>% 
+  dplyr::ungroup()
+
+ggplot(clust.summary.df) + 
+  theme_bw() + 
+  theme(text = element_text(size = 10)) + 
+  scale_y_log10() +
+  scale_x_log10(breaks = n.vec) +
+  labs(y = 'community detection error rate',
+       x = 'n',
+       colour = 'known labels\nper community', 
+       shape = NULL) + 
+  geom_line(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
+  geom_point(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
+  geom_errorbar(aes(x = n, 
+                    ymin = first.q.rate, ymax = third.q.rate, 
+                    colour = factor(nsamp)),
+                width = .125) + 
+  scale_colour_brewer(palette = 'Set1')
+```
+
 ## Applications
@@
   scale_colour_brewer(palette = 'Set1') + 
-  coord_fixed(.5) +
+  coord_fixed(.4) +
   labs(x = expression(x[1]), y = expression(x[2]))
@@
 
-Since $x_i = g(t_i; p_{z_i})$, $\|Q \hat{x}_i - g(t_i; p_{z_i})\| < \epsilon$.
-
 The statement of the theorem assumes that for each community $k$, there at least $R + 1$ points with known label $k$. 
@@
 $$P(L < 2 M n^{-1/2} \log^c n) > P(\epsilon + \delta < 2 M n^{-1/2} \log^c n) \to 1.$$
-
 \end{proof}
@@
 
+The curve fitting step in algorithm \ref{alg:kcurves} requires estimation of some $g: [0, 1] \to \mathbb{R}^d$ for each subset of points corresponding to each label. 
+One choice of curve is the Bezier polynomial. 
+The Bezier polynomial of order $R$ is defined by function $g$ with parameters $p$ such that
+$$g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}$$
+where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]$.
+
+The general idea behind Bezier curve fitting to noisy data is as follows: 
+We wish to minimize the squared loss function 
+\begin{equation}
+\label{eq:bezier-ls}
+L(p, t; X) = \sum_i \|x_i - g(t_i; p)\|^2.
+\end{equation}
+Suppose that we know the timepoints $t_1, ..., t_n$ which we wish to use to fit Bezier coefficients $p^{(0)}, ..., p^{(R)}$ to vectors $x_1, ..., x_n \in \mathbb{R}^d$ lying on or near a Bezier curve of order $R$.
+The squared loss then can be written as 
+$$L(p; t, X) = \|X - T p\|^2,$$
+where $p$ is defined as before, $T$ is an $n \times (R+1)$ matrix such that $T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}$, and $X = \big[ x_1 \mid \cdots \mid x_n \big]$. 
+Then the least squares estimates for the Bezier coefficients $p$ is simply the linear least squares estimate 
+\begin{equation}
+\label{eq:bezier-p-fit}
+\hat{p} = (T^\top T)^{-1} T^\top X.
+\end{equation}
+
+On the other hand, suppose that we know the Bezier coefficients $p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top$, and we wish to fit timepoints $t_1, ..., t_n$ to vectors $x_1, ..., x_n$ lying on or near a Bezier polynomial of order $R$.
+In this setting, we can fit each individual $t_i$ independently by minimizing $L(t_i; p, x_i) = \|x_i - g(t_i; p)\|^2$ for each $t_i$. 
+If $g(t_i; p, x_i)$ is a polynomial of order $R$, then $\dot{L}(t_i; p, x_i)$ is also a polynomial, of order $2 R - 1$. 
+Minimizing $L(t_i; p, x_i)$ then simply involves finding the roots of $\dot{L}(t_i; p, x_i)$ and choosing the root with the smallest objective value (in practice, the endpoints $t_i = 0$ and $t_i = 1$ are also checked).
+For the Bezier curve of order $R$, the full polynomial is given by:
+\begin{equation}
+\label{eq:dbezier}
+\dot{L}(t_i; p, x_i) = \bigg(\sum_{r=0}^R \binom{R}{r} (-1)^r c_r t^r \bigg) \bigg(\sum_{r=0}^{R-1} \binom{R-1}{r} (-1)^r c_{r+1} t^r \bigg),
+\end{equation}
+where $c_R = \sum_{r=0}^R (-1)^{R-r} \binom{R}{r} p^{(r)}$.
+The full Bezier polynomial fitting method is detailed in algorithm \ref{alg:bezier-fit}.
+
+\begin{algorithm}[h]
+\label{alg:bezier-fit}
+\DontPrintSemicolon
+\SetAlgoLined
+\setstretch{1.5}
+\KwData{$x_1, ..., x_n \in \mathbb{R}^d$ lying on or near a Bezier curve of order $R$, stopping criterion $\epsilon$.}
+\KwResult{Bezier coefficients $p$, timepoint values $t_1, ..., t_n$.}
+Initialize timepoints $t_1, ..., t_n$ (e.g., by sampling from a random distribution).\;
+\Repeat {the change in equation \ref{eq:bezier-ls} is less than $\epsilon$.} {
+  Fit $\hat{p}$ by equation \ref{eq:bezier-p-fit}.\;
+  \For {$i = 1, ..., n$} {
+    Fit $t_i$ by equation by finding the roots of equation \ref{eq:dbezier} and choosing the roots that minimize equation \ref{eq:bezier-ls}.\;
+  }
+}
+\caption{Procedure for fitting a Bezier polynomial to noisy data.}
+\end{algorithm}
+
+Algorithm \ref{alg:bezier-fit} is sensitive to initialization. 
+We propose two initialization strategies. 
+One involves using a one-dimensional Isomap \citep{Tenenbaum2000-ff} embedding normalized to the unit interval. 
+Since the vectors lie approximately on a one-dimensional manifold, a one-dimensional Isomap embedding should be able to approximate the manifold, and in particular the interpoint distances along the manifold, as long as the sample is dense along the manifold \citep{trosset2020rehabilitating}.
+
 \newpage
 
+\section*{Appendix D: Supplementary Materials}
+
+\newpage
+
 # References

### 2022-10-26 draft/draft.tex
minor tweak to proof of theorem 4.2
minor updates
adding in detaisls on bezier curve fitting

@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model.
-Data are often represented as an \(n \times p\) matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
+Data are often represented as an \(n \times p\) matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
 An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
@@
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector \(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top \in \mathbb{R}^p\), we observe \(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
+Instead of feature vector \(x_i = \big[ x_{i1} \mid \cdots \mid x_{ip} \big]^\top \in \mathbb{R}^p\), we observe \(a_i = \big[ a_{i1} \mid \cdots \mid a_{in} \big]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
 For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
@@
   The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}_K\).
-  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \begin{bmatrix} \alpha_1 & \cdots & \alpha_K \end{bmatrix}^\top\) and \(\{\theta_{k \ell}\}_K\).
+  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \big[ \alpha_1, \cdots, \alpha_K \big]^\top\) and \(\{\theta_{k \ell}\}_K\).
 \end{enumerate}
@@
 \item
-  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
+  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
 \item
@@
 
-For a simulation study, we again turn to examples. \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}.
-The exact simulation setup is as follows:
+We performed two simulation experiments for algorithm \ref{alg:kcurves}.
+In the first simulation setup, we again turn to examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}.
+The setup is as follows:
 
@@
 \item
-  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top\) and \(g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top\).
-  Collect the latent vectors into matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
+  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top\) and \(g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top\).
+  Collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
 \item
@@
 Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing \(p_1^{(0)} = p_2^{(0)} = (0, 0)\).
-Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting.
-The simulation results suggest that for large \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.
+Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates for each \(n\) and \emph{a priori} known labels per community.
+The simulation results suggest that for sufficient \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.
 
@@
 
-\caption{Median and IQR error rates over 50 replicates.}\label{fig:sim-curves}
+\caption{Median and IQR error rates over 50 replicates in the simulation setting for $K = 2$ curves.}\label{fig:sim-curves}
 \end{figure}
@@
 \caption{Fitted curves over 256 replicates. The curves are colored by the loss values.}\label{fig:repeat-curves}
+\end{figure}
+
+In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves.
+The setup is as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Draw \(z_1, ..., z_n \iid \Multinomial(1/3, 1/3, 1/3)\).
+\item
+  Draw \(t_1, ..., t_n \iid \Uniform(0, 1)\).
+\item
+  Let each \(x_i = g_{z_i}(t_i)\) where
+
+  \begin{enumerate}
+  \def\labelenumii{\roman{enumii}.}
+  \tightlist
+  \item
+    \(g_1(t) = \big[ 2 t (t-1), t^2, 0 \big]^\top\)
+  \item
+    \(g_2(t) = \big[ 0, t^2, 2 t (t-1) \big]^\top\)
+  \item
+    \(g_3(t) = \big[2 t (t-1), 0, t^2 \big]^\top\).
+  \end{enumerate}
+\item
+  Collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
+\item
+  Draw \(A \sim \RDPG(X; \rho_n)\) (for these simulations we fix \(\rho_n \equiv 1\)).
+\end{enumerate}
+
+As in the previous setup, we set \(n = 128, 256, 512, 1024, 2048\), and we start with completely random initializations as well as initializations in which we know \(4\) or \(8\) labels from each community.
+For each setting, we simulate 50 replicates.
+Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing \(p_1^{(k)} = (0, 0, 0)\) for each \(k\).
+Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each \(n\) and \emph{a priori} known labels per community.
+As in the previous \(K = 2\) setup, the results suggest that for sufficient \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.
+Unlike in the previous setup, random initialization fails to improve with greater \(n\).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/sim-curves-3-1}
+
+}
+
+\caption{Median and IQR error rates over 50 replicates in the simulation setup for $K = 3$ curves.}\label{fig:sim-curves-3}
 \end{figure}
@@
 \section*{Appendix C: Details on Fitting Bezier Curves to Noisy Data}
+
+The curve fitting step in algorithm \ref{alg:kcurves} requires estimation of some \(g: [0, 1] \to \mathbb{R}^d\) for each subset of points corresponding to each label.
+One choice of curve is the Bezier polynomial.
+The Bezier polynomial of order \(R\) is defined by function \(g\) with parameters \(p\) such that
+\[g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}\]
+where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]\).
+
+The general idea behind Bezier curve fitting to noisy data is as follows:
+We wish to minimize the squared loss function
+\begin{equation}
+\label{eq:bezier-ls}
+L(p, t; X) = \sum_i \|x_i - g(t_i; p)\|^2.
+\end{equation}
+Suppose that we know the timepoints \(t_1, ..., t_n\) which we wish to use to fit Bezier coefficients \(p^{(0)}, ..., p^{(R)}\) to vectors \(x_1, ..., x_n \in \mathbb{R}^d\) lying on or near a Bezier curve of order \(R\).
+The squared loss then can be written as
+\[L(p; t, X) = \|X - T p\|^2,\]
+where \(p\) is defined as before, \(T\) is an \(n \times (R+1)\) matrix such that \(T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}\), and \(X = \big[ x_1 \mid \cdots \mid x_n \big]\).
+Then the least squares estimates for the Bezier coefficients \(p\) is simply the linear least squares estimate
+\begin{equation}
+\label{eq:bezier-p-fit}
+\hat{p} = (T^\top T)^{-1} T^\top X.
+\end{equation}
+
+On the other hand, suppose that we know the Bezier coefficients \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top\), and we wish to fit timepoints \(t_1, ..., t_n\) to vectors \(x_1, ..., x_n\) lying on or near a Bezier polynomial of order \(R\).
+In this setting, we can fit each individual \(t_i\) independently by minimizing \(L(t_i; p, x_i) = \|x_i - g(t_i; p)\|^2\) for each \(t_i\).
+If \(g(t_i; p, x_i)\) is a polynomial of order \(R\), then \(\dot{L}(t_i; p, x_i)\) is also a polynomial, of order \(2 R - 1\).
+Minimizing \(L(t_i; p, x_i)\) then simply involves finding the roots of \(\dot{L}(t_i; p, x_i)\) and choosing the root with the smallest objective value (in practice, the endpoints \(t_i = 0\) and \(t_i = 1\) are also checked).
+For the Bezier curve of order \(R\), the full polynomial is given by:
+\begin{equation}
+\label{eq:dbezier}
+\dot{L}(t_i; p, x_i) = \bigg(\sum_{r=0}^R \binom{R}{r} (-1)^r c_r t^r \bigg) \bigg(\sum_{r=0}^{R-1} \binom{R-1}{r} (-1)^r c_{r+1} t^r \bigg),
+\end{equation}
+where \(c_R = \sum_{r=0}^R (-1)^{R-r} \binom{R}{r} p^{(r)}\).
+The full Bezier polynomial fitting method is detailed in algorithm \ref{alg:bezier-fit}.
+
+\begin{algorithm}[h]
+\label{alg:bezier-fit}
+\DontPrintSemicolon
+\SetAlgoLined
+\setstretch{1.5}
+\KwData{$x_1, ..., x_n \in \mathbb{R}^d$ lying on or near a Bezier curve of order $R$, stopping criterion $\epsilon$.}
+\KwResult{Bezier coefficients $p$, timepoint values $t_1, ..., t_n$.}
+Initialize timepoints $t_1, ..., t_n$ (e.g., by sampling from a random distribution).\;
+\Repeat {the change in equation \ref{eq:bezier-ls} is less than $\epsilon$.} {
+  Fit $\hat{p}$ by equation \ref{eq:bezier-p-fit}.\;
+  \For {$i = 1, ..., n$} {
+    Fit $t_i$ by equation by finding the roots of equation \ref{eq:dbezier} and choosing the roots that minimize equation \ref{eq:bezier-ls}.\;
+  }
+}
+\caption{Procedure for fitting a Bezier polynomial to noisy data.}
+\end{algorithm}
+
+Algorithm \ref{alg:bezier-fit} is sensitive to initialization.
+We propose two initialization strategies.
+One involves using a one-dimensional Isomap \citep{Tenenbaum2000-ff} embedding normalized to the unit interval.
+Since the vectors lie approximately on a one-dimensional manifold, a one-dimensional Isomap embedding should be able to approximate the manifold, and in particular the interpoint distances along the manifold, as long as the sample is dense along the manifold \citep{trosset2020rehabilitating}.
+
+\newpage
+
+\section*{Appendix D: Supplementary Materials}
 

### 2022-10-26 draft/draft.Rmd
added links to github repos

@@
 $$g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}$$
-where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]$.
+where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top$.
 
@@
 We propose two initialization strategies. 
-One involves using a one-dimensional Isomap \citep{Tenenbaum2000-ff} embedding normalized to the unit interval. 
-Since the vectors lie approximately on a one-dimensional manifold, a one-dimensional Isomap embedding should be able to approximate the manifold, and in particular the interpoint distances along the manifold, as long as the sample is dense along the manifold \citep{trosset2020rehabilitating}.
+The first involves using a one-dimensional Isomap \citep{Tenenbaum2000-ff} embedding normalized to the unit interval. 
+Since the vectors lie approximately on a one-dimensional manifold, a one-dimensional Isomap embedding should be able to approximate the manifold, and in particular the interpoint distances along the manifold, as long as the sample is sufficiently dense along the entire curve \citep{trosset2020rehabilitating}.
+The second initialization strategy involves choosing $R+1$ representative vectors. 
+Then it is possible to fit a Bezier polynomial of order $R$ exactly to these $R+1$ vectors. 
+If the representative vectors lie close to the true curve, then the fitted curve is also close to the true curve, providing a good initial guess. 
+In this strategy, we use the consistency of the ASE \citep{rubindelanchy2017statistical} to ensure that this is true asymptotically of any embedding vector in the ASE (see theorem \ref{thm:kcurves-semisupervised}).
 
@@
 
+The RMarkdown and \LaTeX files to compile this dissertation can be found at https://github.com/johneverettkoo/dissertation. 
+The code for simulations and data analyses in section \@ref(sec:pabm-grdpg) can be found at https://github.com/johneverettkoo/pabm-grdpg. 
+An R package implementing orthogonal spectral clustering (algorithm \ref(alg:osc)) is currently in progress and can be found at https://github.com/johneverettkoo/osc. 
+The code for simulations and data analyses in section \@ref(sec:mbm) can be found at https://github.com/johneverettkoo/manifold-block-models.
+
 \newpage

### 2022-10-26 draft/draft.tex
added links to github repos

@@
 \[g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}\]
-where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]\).
+where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top\).
 
@@
 We propose two initialization strategies.
-One involves using a one-dimensional Isomap \citep{Tenenbaum2000-ff} embedding normalized to the unit interval.
-Since the vectors lie approximately on a one-dimensional manifold, a one-dimensional Isomap embedding should be able to approximate the manifold, and in particular the interpoint distances along the manifold, as long as the sample is dense along the manifold \citep{trosset2020rehabilitating}.
+The first involves using a one-dimensional Isomap \citep{Tenenbaum2000-ff} embedding normalized to the unit interval.
+Since the vectors lie approximately on a one-dimensional manifold, a one-dimensional Isomap embedding should be able to approximate the manifold, and in particular the interpoint distances along the manifold, as long as the sample is sufficiently dense along the entire curve \citep{trosset2020rehabilitating}.
+The second initialization strategy involves choosing \(R+1\) representative vectors.
+Then it is possible to fit a Bezier polynomial of order \(R\) exactly to these \(R+1\) vectors.
+If the representative vectors lie close to the true curve, then the fitted curve is also close to the true curve, providing a good initial guess.
+In this strategy, we use the consistency of the ASE \citep{rubindelanchy2017statistical} to ensure that this is true asymptotically of any embedding vector in the ASE (see theorem \ref{thm:kcurves-semisupervised}).
 
@@
 \section*{Appendix D: Supplementary Materials}
+
+The RMarkdown and \LaTeX files to compile this dissertation can be found at \url{https://github.com/johneverettkoo/dissertation}.
+The code for simulations and data analyses in section \ref{sec:pabm-grdpg} can be found at \url{https://github.com/johneverettkoo/pabm-grdpg}.
+An R package implementing orthogonal spectral clustering (algorithm \ref(alg:osc)) is currently in progress and can be found at \url{https://github.com/johneverettkoo/osc}.
+The code for simulations and data analyses in section \ref{sec:mbm} can be found at \url{https://github.com/johneverettkoo/manifold-block-models}.
 

### 2022-10-27 draft/draft.Rmd
updated k=3 simulation with a more interesting setup

@@
 
-```{r repeat-curves, fig.height = 3, fig.width = 3, fig.cap = 'Fitted curves over 256 replicates. The curves are colored by the loss values.', cache = TRUE}
+```{r repeat-curves, fig.height = 2.5, fig.width = 12, fig.cap = 'Fitted curves over 256 replicates. The curves are colored by the loss values.', cache = TRUE}
 plot.repeat.study <- function(repeat.df) {
@@
   ggplot(curves.df) + 
-    geom_path(aes(x = x, y = y, group = interaction(z, iter),
+    geom_path(aes(x = y, y = x, group = interaction(z, iter),
                   colour = loss),
@@
     labs(colour = 'loss',
-         x = expression(x[1]),
-         y = expression(x[2])) + 
+         x = NULL,
+         y = NULL) + 
     viridis::scale_color_viridis()
@@
 In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves. 
-The setup is as follows:
+The mixture model is as follows (also see figure \ref{fig:three-curves}):
 
@@
     ii. $g_2(t) = \big[ 0, t^2, 2 t (t-1) \big]^\top$
-    iii. $g_3(t) = \big[2 t (t-1), 0, t^2 \big]^\top$.
+    iii. $g_3(t) = \big[2 t (t-1), t^2, 2 t (t-1) \big]^\top$.
 4. Collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
 5. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
+
+```{r three-curves, cache = TRUE, fig.height = 3, fig.width = 12, fig.cap = 'Latent vectors (left) and ASE (right) of one realization of the mixture model in the simulation setup with $K=3$ manifolds.'}
+par(mfrow = c(1, 3))
+
+n <- 512
+
+p.list <- list(matrix(c(0, 1, 0, 
+                        0, 0, 1,
+                        0, 0, 0),
+                      nrow = 3, ncol = 3),
+               matrix(c(0, 0, 0, 
+                        0, 0, 1,
+                        0, 1, 0),
+                      nrow = 3, ncol = 3),
+               matrix(c(0, 1, 0, 
+                        0, 0, 1,
+                        0, 1, 0),
+                      nrow = 3, ncol = 3))
+
+K <- length(p.list)
+
+z <- sample(seq(K), n, replace = TRUE)
+z <- sort(z)
+
+t <- runif(n)
+
+X <- lapply(seq(K), function(k) {
+  bezier.curve(t[z == k], 
+               p.list[[k]], 
+               intercept = TRUE)
+}) %>% 
+  do.call('rbind', .)
+
+p <- ncol(X)
+q <- 0
+
+P <- grdpg.edge.prob.matrix(X, p, q)
+A <- draw.graph(P)
+Xhat <- embedding(A, p, q)
+
+scatterplot3d::scatterplot3d(X[, 1], X[, 2], X[, 3],
+                             color = z, 
+                             xlab = NA, ylab = NA, zlab = NA,
+                             pch = '.', asp = 1,
+                             tick.marks = FALSE)
+
+scatterplot3d::scatterplot3d(Xhat[, 1], Xhat[, 2], Xhat[, 3], 
+                             color = z, 
+                             xlab = NA, ylab = NA, zlab = NA, 
+                             pch = '.', asp = 1, 
+                             tick.marks = FALSE,
+                             angle = 40)
+```
 
@@
 Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each $n$ and *a priori* known labels per community. 
-As in the previous $K = 2$ setup, the results suggest that for sufficient $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.
-Unlike in the previous setup, random initialization fails to improve with greater $n$.
+As in the previous $K = 2$ simulations, the results suggest that for sufficient $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.
+Unlike in the previous setup, random initialization results in low community detection error rates, close to the semi-supervised setups.
 
@@
 For simplicity, suppose we have exactly $R + 1$ (if there are more, then discard the rest). 
-Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly (see appendix C), yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial.
+Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly, yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial.
 Use the fitted Bezier curves to define $\delta_k(t) = \|Q g(t; \hat{p}_k) - g(t; p_k)\|$, $\delta_k = \max_t \delta_k(t)$, and $\delta = \max_k \delta_k$. 

### 2022-10-27 draft/draft.tex
updated k=3 simulation with a more interesting setup

@@
 In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves.
-The setup is as follows:
+The mixture model is as follows (also see figure \ref{fig:three-curves}):
 
@@
   \item
-    \(g_3(t) = \big[2 t (t-1), 0, t^2 \big]^\top\).
+    \(g_3(t) = \big[2 t (t-1), t^2, 2 t (t-1) \big]^\top\).
   \end{enumerate}
@@
 \end{enumerate}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/three-curves-1}
+
+}
+
+\caption{Latent vectors (left) and ASE (right) of one realization of the mixture model in the simulation setup with $K=3$ manifolds.}\label{fig:three-curves}
+\end{figure}
 
@@
 Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each \(n\) and \emph{a priori} known labels per community.
-As in the previous \(K = 2\) setup, the results suggest that for sufficient \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.
-Unlike in the previous setup, random initialization fails to improve with greater \(n\).
+As in the previous \(K = 2\) simulations, the results suggest that for sufficient \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.
+Unlike in the previous setup, random initialization results in low community detection error rates, close to the semi-supervised setups.
 

### 2022-10-27 draft/draft.Rmd
fixed typo in theorem 4.2

@@
 Then if $n \rho_n = \omega(\log^{4c} n)$, as $n \to \infty$, the estimates outputted by $K$-curves clustering are such that 
-$$L(\hat{z}_1, ..., \hat{z}_n, \hat{g}_1, ..., \hat{g}_n; X) \stackrel{p}{\to} 0,$$
+$$L(\hat{z}_1, ..., \hat{z}_n, \hat{g}_1, ..., \hat{g}_K; X) \stackrel{p}{\to} 0,$$
 where $L$ is the loss function defined in equation (\ref{eq:kcurves-loss}). 
@@
 
-As in the previous setup, we set $n = 128, 256, 512, 1024, 2048$, and we start with completely random initializations as well as initializations in which we know $4$ or $8$ labels from each community.
-For each setting, we simulate 50 replicates. 
-Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing $p_1^{(k)} = (0, 0, 0)$ for each $k$. 
+As in the previous setup, we set $n = 128, 256, 512, 1024, 2048$, and we started with completely random initializations as well as initializations in which we know $4$ or $8$ labels from each community.
+For each setting, we simulated 50 replicates. 
+Since the curves meet at the origin, we forced the fitted Bezier curves to meet at the origin as well by fixing $p_k^{(0)} = (0, 0, 0)$ for each $k$. 
 Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each $n$ and *a priori* known labels per community. 
 As in the previous $K = 2$ simulations, the results suggest that for sufficient $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.
-Unlike in the previous setup, random initialization results in low community detection error rates, close to the semi-supervised setups.
+Unlike in the previous setup, random initialization results in low community detection error rates, close to the semi-supervised setups for larger $n$.
 

### 2022-10-27 draft/draft.tex
fixed typo in theorem 4.2

@@
 
-As in the previous setup, we set \(n = 128, 256, 512, 1024, 2048\), and we start with completely random initializations as well as initializations in which we know \(4\) or \(8\) labels from each community.
-For each setting, we simulate 50 replicates.
-Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing \(p_1^{(k)} = (0, 0, 0)\) for each \(k\).
+As in the previous setup, we set \(n = 128, 256, 512, 1024, 2048\), and we started with completely random initializations as well as initializations in which we know \(4\) or \(8\) labels from each community.
+For each setting, we simulated 50 replicates.
+Since the curves meet at the origin, we forced the fitted Bezier curves to meet at the origin as well by fixing \(p_k^{(0)} = (0, 0, 0)\) for each \(k\).
 Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each \(n\) and \emph{a priori} known labels per community.
 As in the previous \(K = 2\) simulations, the results suggest that for sufficient \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.
-Unlike in the previous setup, random initialization results in low community detection error rates, close to the semi-supervised setups.
+Unlike in the previous setup, random initialization results in low community detection error rates, close to the semi-supervised setups for larger \(n\).
 

### 2022-10-28 draft/draft.Rmd
modified drosophila example to use all four labels

@@
 
-In the final example, we examined the *Drosophila* connectome graph \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
-In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types (Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons). 
-In our analysis, we only considered neurons labeled as Kenyon Cells and Projection Neurons in the right hemisphere. 
-The resulting graph is illustrated in figure \ref{fig:mbconnectome-graph}.
-
-The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for this graph (figure \ref{fig:mbconnectome-ase}). 
-Fitting two quadratic Bezier curves to the ASE (figure \ref{fig:mbconnectome-kcurves}) results in 91.4\% accuracy.
-
-```{r mbconnectome-graph, fig.cap = 'Drosophila connectome graph. Each vertex represents a neuron, which are labeled by neuron type.', fig.height = 2, figh.width = 4}
+In the final example, we examined the *Drosophila* connectome graphs \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
+This dataset consists of two graphs representing two networks of neurons, one for each hemisphere of the *Drosophila* brain. 
+In these graphs, each vertex is a neuron, and the labels correspond to one of four neuron types (Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons). 
+The resulting graphs are illustrated in figure \ref{fig:mbconnectome-graph}.
+
+The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}). 
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 69.9\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
+
+```{r mbconnectome-graph, fig.cap = 'Graphs of the Drosophila connectomes.
+The left and right are of the left and right hemispheres, respectively.
+Each vertex represents a neuron, which are labeled by neuron type.', fig.height = 2, figh.width = 12}
 library(mbstructure)
@@
 
-graph <- generate.graph(newrdat, vdf.right)
-A <- as.matrix(igraph::as_adjacency_matrix(graph$g, type = 'both'))
-z <- graph$vdf$type
-ind.keep <- z %in% c('KC', 'PN')
-A <- A[ind.keep, ind.keep]
-z <- as.numeric(as.factor(as.numeric(z[ind.keep])))
-A <- sign(A + t(A))
-diag(A) <- 0
-
-ggraph(A, layout = 'stress') + 
-  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
-  geom_node_point(aes(colour = factor(z)), 
+graph.r <- generate.graph(newrdat, vdf.right)
+A.r <- as.matrix(igraph::as_adjacency_matrix(graph.r$g, type = 'both'))
+z.r <- as.numeric(graph.r$vdf$type)
+n.r <- length(z.r)
+A.r <- sign(A.r + t(A.r))
+diag(A.r) <- 0
+graph.r <- ggraph(A.r, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n.r) / 2) + 
+  geom_node_point(aes(colour = factor(z.r)), 
                   show.legend = FALSE) + 
@@
   scale_colour_brewer(palette = 'Set1')
+
+graph.l <- generate.graph(newldat, vdf.left)
+A.l <- as.matrix(igraph::as_adjacency_matrix(graph.l$g, type = 'both'))
+z.l <- as.numeric(graph.l$vdf$type)
+n.l <- length(z.l)
+Al <- sign(A.l + t(A.l))
+diag(A.l) <- 0
+graph.l <- ggraph(A.l, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n.l) / 2) + 
+  geom_node_point(aes(colour = factor(z.l)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+
+gridExtra::grid.arrange(graph.l, graph.r, nrow = 1)
 ```
 
-```{r mbconnectome-ase, fig.cap = 'ASE of the Drosophila connectome graph. The embedding vectors are labeled by neuron type.', fig.height = 2, fig.width = 8}
-Xhat <- embedding(A, 2, 0)
-
-ggplot() + 
-  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2],
-                 colour = factor(z))) + 
+```{r mbconnectome-ase, fig.cap = 'ASE of the Drosophila connectome graphs. The embedding vectors are labeled by neuron type.', fig.height = 2, fig.width = 12}
+Xhat.l <- embedding(A.l, 2, 0)
+ase.l <- ggplot() + 
+  geom_point(aes(x = Xhat.l[, 1], y = Xhat.l[, 2],
+                 colour = factor(z.l))) + 
   coord_fixed(.5) +
@@
   scale_colour_brewer(palette = 'Set1') + 
-  labs(x = expression(x[1]), y = expression(x[2]),
-       colour = NULL)
+  labs(x = NULL, y = NULL, colour = NULL)
+
+Xhat.r <- embedding(A.r, 2, 0)
+ase.r <- ggplot() + 
+  geom_point(aes(x = Xhat.r[, 1], y = Xhat.r[, 2],
+                 colour = factor(z.r))) + 
+  coord_fixed(.5) +
+  theme_bw() + 
+  theme(legend.position = 'none') +
+  scale_colour_brewer(palette = 'Set1') + 
+  labs(x = NULL, y = NULL, colour = NULL)
+
+gridExtra::grid.arrange(ase.l, ase.r, nrow = 1)
 ```
 
-```{r mbconnectome-kcurves, fig.cap = 'ASE of the Drosophila connectome graph with quadratic Bezier curves fitted via $K$-curves clustering.
-The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 8, cache = TRUE}
-set.seed(314159)
-z.init <- sample.points(z, 3)
-
-mbconnectome.out <- manifold.clustering(Xhat, 2, 
-                                        initialization = z.init,
+```{r mbconnectome-kcurves, fig.cap = 'ASEs of the Drosophila connectome graphs with quadratic Bezier curves fitted via $K$-curves clustering.
+The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 12, cache = TRUE}
+set.seed(314159265)
+
+z.init.r <- sample.points(z.r, 10)
+z.init.l <- sample.points(z.l, 10)
+
+mbconnectome.out.r <- manifold.clustering(Xhat.r, 
+                                          length(unique(z.r)), 
+                                          initialization = z.init.r,
                                           intercept = FALSE, 
@@
                                           curve.init = 'x')
-
-plot.estimated.curves(Xhat, mbconnectome.out) + 
+fitted.r <- plot.estimated.curves(Xhat.r, mbconnectome.out.r) + 
   scale_colour_brewer(palette = 'Set1') + 
   coord_fixed(.4) +
-  labs(x = expression(x[1]), y = expression(x[2]))
+  labs(x = NULL, y = NULL)
+
+mbconnectome.out.l <- manifold.clustering(Xhat.l, 
+                                          length(unique(z.l)), 
+                                          initialization = z.init.l,
+                                          intercept = FALSE, 
+                                          parallel = TRUE, 
+                                          verbose = FALSE, 
+                                          animate = FALSE, 
+                                          curve.init = 'x')
+fitted.l <- plot.estimated.curves(Xhat.l, mbconnectome.out.l) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  coord_fixed(.4) +
+  labs(x = NULL, y = NULL)
+
+gridExtra::grid.arrange(fitted.l, fitted.r, nrow = 1)
 ```

### 2022-10-28 draft/draft.tex
modified drosophila example to use all four labels

@@
 
-In the final example, we examined the \emph{Drosophila} connectome graph \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
-In this graph, each vertex is a neuron, and the labels correspond to one of four neuron types (Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons).
-In our analysis, we only considered neurons labeled as Kenyon Cells and Projection Neurons in the right hemisphere.
-The resulting graph is illustrated in figure \ref{fig:mbconnectome-graph}.
-
-The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for this graph (figure \ref{fig:mbconnectome-ase}).
-Fitting two quadratic Bezier curves to the ASE (figure \ref{fig:mbconnectome-kcurves}) results in 91.4\% accuracy.
+In the final example, we examined the \emph{Drosophila} connectome graphs \citep{Eichler141762}, which has been studied as a GRDPG by \citet{athreya2017statistical}, \citet{https://doi.org/10.48550/arxiv.1705.03297}, and \citet{SannaPassino2022}.
+This dataset consists of two graphs representing two networks of neurons, one for each hemisphere of the \emph{Drosophila} brain.
+In these graphs, each vertex is a neuron, and the labels correspond to one of four neuron types (Kenyon Cells, Input Neurons, Output Neurons, and Projection Neurons).
+The resulting graphs are illustrated in figure \ref{fig:mbconnectome-graph}.
+
+The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}).
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 69.9\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
 
@@
 
-\caption{Drosophila connectome graph. Each vertex represents a neuron, which are labeled by neuron type.}\label{fig:mbconnectome-graph}
+\caption{Graphs of the Drosophila connectomes.
+The left and right are of the left and right hemispheres, respectively.
+Each vertex represents a neuron, which are labeled by neuron type.}\label{fig:mbconnectome-graph}
 \end{figure}
@@
 
-\caption{ASE of the Drosophila connectome graph. The embedding vectors are labeled by neuron type.}\label{fig:mbconnectome-ase}
+\caption{ASE of the Drosophila connectome graphs. The embedding vectors are labeled by neuron type.}\label{fig:mbconnectome-ase}
 \end{figure}
@@
 
-\caption{ASE of the Drosophila connectome graph with quadratic Bezier curves fitted via $K$-curves clustering. The embedding vectors are colored by predicted label.}\label{fig:mbconnectome-kcurves}
+\caption{ASEs of the Drosophila connectome graphs with quadratic Bezier curves fitted via $K$-curves clustering. The embedding vectors are colored by predicted label.}\label{fig:mbconnectome-kcurves}
 \end{figure}

### 2022-10-28 draft/draft.Rmd
adjusted a few plots

@@
 
-```{r mbconnectome-ase, fig.cap = 'ASE of the Drosophila connectome graphs. The embedding vectors are labeled by neuron type.', fig.height = 2, fig.width = 12}
+```{r mbconnectome-ase, fig.cap = 'ASEs of the Drosophila connectome graphs. The embedding vectors are labeled by neuron type.', fig.height = 2, fig.width = 8}
 Xhat.l <- embedding(A.l, 2, 0)
@@
 ```{r mbconnectome-kcurves, fig.cap = 'ASEs of the Drosophila connectome graphs with quadratic Bezier curves fitted via $K$-curves clustering.
-The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 12, cache = TRUE}
+The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 8, cache = TRUE}
 set.seed(314159265)
@@
   scale_colour_brewer(palette = 'Set1') + 
-  coord_fixed(.4) +
-  labs(x = NULL, y = NULL)
+  coord_fixed(.5) +
+  labs(x = NULL, y = NULL) + 
+  ylim(min(Xhat.r[, 2]), max(Xhat.r[, 2])) + 
+  xlim(min(Xhat.r[, 1]), max(Xhat.r[, 1]))
 
@@
   scale_colour_brewer(palette = 'Set1') + 
-  coord_fixed(.4) +
-  labs(x = NULL, y = NULL)
+  coord_fixed(.5) +
+  labs(x = NULL, y = NULL) + 
+  ylim(min(Xhat.l[, 2]), max(Xhat.l[, 2])) + 
+  xlim(min(Xhat.l[, 1]), max(Xhat.l[, 1]))
 

### 2022-10-28 draft/draft.Rmd
fixed typo that resuin mbconnectome data analysis that resulted in poorer performance

@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model. 
-Data are often represented as an $n \times p$ matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements.
+Data are often represented as an $n \times p$ matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements.
 An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$.
@@
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector $x_i = \big[ x_{i1} \mid \cdots \mid x_{ip} \big]^\top \in \mathbb{R}^p$, we observe $a_i = \big[ a_{i1} \mid \cdots \mid a_{in} \big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
+Instead of feature vector $x_i = \Big[ x_{i1} \mid \cdots \mid x_{ip} \Big]^\top \in \mathbb{R}^p$, we observe $a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
 For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
@@
 
-To estimate $\omega = \big[ \omega_1, \cdots, \omega_n \big]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
+To estimate $\omega = \Big[ \omega_1, \cdots, \omega_n \Big]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
 For simplicity, set $C = 1$. 
@@
 The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}_K$.
-Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \big[ \alpha_1, \cdots, \alpha_K \big]^\top$ and $\{\theta_{k \ell}\}_K$.
+Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \Big[ \alpha_1, \cdots, \alpha_K \Big]^\top$ and $\{\theta_{k \ell}\}_K$.
 
@@
 Denote $d = p + q$ and let $\mathcal{X}$ be the subset of $\mathbb{R}^d$ such that, for any $x, y \in \mathcal{X}$, we have $x^\top I_{p,q} y \in [0, 1]$. 
-Let $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
+Let $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
 A graph $G$ with adjacency matrix $A$ is said to be a generalized random dot product graph with latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$ if $A \sim \BG(P)$ where the edge probability matrix $P$ is given by $P = \rho_n X I_{p,q} X^\top$, i.e., the entries of $P$ are of the form $P_{ij} = \rho_n x_i^\top I_{p,q} x_j$.
@@
 Suppose that latent vectors for an RDPG are drawn from the curve 
-$x(t) = \big[ t^2, 2 t (t-1) \big]^\top$, $t \in [0, 1]$. 
+$x(t) = \Big[ t^2, 2 t (t-1) \Big]^\top$, $t \in [0, 1]$. 
 Let $x_i = x(t_i)$. 
@@
 In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
-This corresponds to a one-dimensional RDPG with a latent structure $X = \big[ \sqrt{p},  \cdots, \sqrt{p} \big]^\top$. 
-In another view, the RDPG with latent structure $X = \big[ \sqrt{p}, \cdots, \sqrt{p} \big]^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
-
-Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \big[ \cos \frac{\pi}{2} t, \sin \frac{\pi}{2} t \big]^\top$ for $t \in [0, 1]$.
+This corresponds to a one-dimensional RDPG with a latent structure $X = \Big[ \sqrt{p},  \cdots, \sqrt{p} \Big]^\top$. 
+In another view, the RDPG with latent structure $X = \Big[ \sqrt{p}, \cdots, \sqrt{p} \Big]^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
+
+Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \Big[ \cos \frac{\pi}{2} t, \sin \frac{\pi}{2} t \Big]^\top$ for $t \in [0, 1]$.
 Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$.
@@
 \item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
-\item Compile the latent vectors into data matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
+\item Compile the latent vectors into data matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
 \end{enumerate}
@@
 \label{ex:separated-curves}
-Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \big[ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) \big]^\top$ and $f_2(t) = \big[ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) \big]^\top$.
-Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
+Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \Big[ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) \Big]^\top$ and $f_2(t) = \Big[ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) \Big]^\top$.
+Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$.
 Finally, let $A \sim \RDPG(X)$.
@@
 Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$.
-Let $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
+Let $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
 Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
@@
 2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
-3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
+3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$.
 4. Draw $A \sim \GRDPG_{p,q}(X; \rho_n)$.
@@
 \label{ex:intersect-curves}
-Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top$ and $g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top$.
+Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$.
 This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
@@
 Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
-In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \big[ 0, 0 \big]^\top$, i.e., the curves meet at the origin. 
+In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \Big[ 0, 0 \Big]^\top$, i.e., the curves meet at the origin. 
 Here, an accuracy of 87.5\% is achieved.
@@
 2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
-3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top$ and $g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top$. 
-Collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$. 
+3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$. 
+Collect the latent vectors into matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$. 
 4. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
@@
 3. Let each $x_i = g_{z_i}(t_i)$ where
-    i. $g_1(t) = \big[ 2 t (t-1), t^2, 0 \big]^\top$
-    ii. $g_2(t) = \big[ 0, t^2, 2 t (t-1) \big]^\top$
-    iii. $g_3(t) = \big[2 t (t-1), t^2, 2 t (t-1) \big]^\top$.
-4. Collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
+    i. $g_1(t) = \Big[ 2 t (t-1), t^2, 0 \Big]^\top$
+    ii. $g_2(t) = \Big[ 0, t^2, 2 t (t-1) \Big]^\top$
+    iii. $g_3(t) = \Big[2 t (t-1), t^2, 2 t (t-1) \Big]^\top$.
+4. Collect the latent vectors into matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$.
 5. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}). 
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 69.9\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 73.3\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
 
@@
 The left and right are of the left and right hemispheres, respectively.
-Each vertex represents a neuron, which are labeled by neuron type.', fig.height = 2, figh.width = 12}
+Each vertex represents a neuron, which are labeled by neuron type.', fig.height = 2, figh.width = 12, cache = TRUE}
 library(mbstructure)
@@
 n.l <- length(z.l)
-Al <- sign(A.l + t(A.l))
+A.l <- sign(A.l + t(A.l))
 diag(A.l) <- 0
@@
 The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 8, cache = TRUE}
-set.seed(314159265)
+set.seed(1)
 
@@
 $$g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}$$
-where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top$.
+where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top$.
 
@@
 $$L(p; t, X) = \|X - T p\|^2,$$
-where $p$ is defined as before, $T$ is an $n \times (R+1)$ matrix such that $T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}$, and $X = \big[ x_1 \mid \cdots \mid x_n \big]$. 
+where $p$ is defined as before, $T$ is an $n \times (R+1)$ matrix such that $T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}$, and $X = \Big[ x_1 \mid \cdots \mid x_n \Big]$. 
 Then the least squares estimates for the Bezier coefficients $p$ is simply the linear least squares estimate 
@@
 
-On the other hand, suppose that we know the Bezier coefficients $p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top$, and we wish to fit timepoints $t_1, ..., t_n$ to vectors $x_1, ..., x_n$ lying on or near a Bezier polynomial of order $R$.
+On the other hand, suppose that we know the Bezier coefficients $p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top$, and we wish to fit timepoints $t_1, ..., t_n$ to vectors $x_1, ..., x_n$ lying on or near a Bezier polynomial of order $R$.
 In this setting, we can fit each individual $t_i$ independently by minimizing $L(t_i; p, x_i) = \|x_i - g(t_i; p)\|^2$ for each $t_i$. 

### 2022-10-28 draft/draft.tex
adjusted a few plots
fixed typo that resuin mbconnectome data analysis that resulted in poorer performance

@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model.
-Data are often represented as an \(n \times p\) matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
+Data are often represented as an \(n \times p\) matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
 An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
@@
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector \(x_i = \big[ x_{i1} \mid \cdots \mid x_{ip} \big]^\top \in \mathbb{R}^p\), we observe \(a_i = \big[ a_{i1} \mid \cdots \mid a_{in} \big]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
+Instead of feature vector \(x_i = \Big[ x_{i1} \mid \cdots \mid x_{ip} \Big]^\top \in \mathbb{R}^p\), we observe \(a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
 For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
@@
   The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}_K\).
-  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \big[ \alpha_1, \cdots, \alpha_K \big]^\top\) and \(\{\theta_{k \ell}\}_K\).
+  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \Big[ \alpha_1, \cdots, \alpha_K \Big]^\top\) and \(\{\theta_{k \ell}\}_K\).
 \end{enumerate}
@@
 \item
-  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
+  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\).
 \item
@@
 \item
-  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top\) and \(g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top\).
-  Collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
+  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top\) and \(g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top\).
+  Collect the latent vectors into matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\).
 \item
@@
   \item
-    \(g_1(t) = \big[ 2 t (t-1), t^2, 0 \big]^\top\)
-  \item
-    \(g_2(t) = \big[ 0, t^2, 2 t (t-1) \big]^\top\)
-  \item
-    \(g_3(t) = \big[2 t (t-1), t^2, 2 t (t-1) \big]^\top\).
+    \(g_1(t) = \Big[ 2 t (t-1), t^2, 0 \Big]^\top\)
+  \item
+    \(g_2(t) = \Big[ 0, t^2, 2 t (t-1) \Big]^\top\)
+  \item
+    \(g_3(t) = \Big[2 t (t-1), t^2, 2 t (t-1) \Big]^\top\).
   \end{enumerate}
 \item
-  Collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
+  Collect the latent vectors into matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\).
 \item
@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}).
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 69.9\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 73.3\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
 
@@
 
-\caption{ASE of the Drosophila connectome graphs. The embedding vectors are labeled by neuron type.}\label{fig:mbconnectome-ase}
+\caption{ASEs of the Drosophila connectome graphs. The embedding vectors are labeled by neuron type.}\label{fig:mbconnectome-ase}
 \end{figure}
@@
 \[g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}\]
-where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top\).
+where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top\).
 
@@
 \[L(p; t, X) = \|X - T p\|^2,\]
-where \(p\) is defined as before, \(T\) is an \(n \times (R+1)\) matrix such that \(T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}\), and \(X = \big[ x_1 \mid \cdots \mid x_n \big]\).
+where \(p\) is defined as before, \(T\) is an \(n \times (R+1)\) matrix such that \(T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}\), and \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]\).
 Then the least squares estimates for the Bezier coefficients \(p\) is simply the linear least squares estimate
@@
 
-On the other hand, suppose that we know the Bezier coefficients \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top\), and we wish to fit timepoints \(t_1, ..., t_n\) to vectors \(x_1, ..., x_n\) lying on or near a Bezier polynomial of order \(R\).
+On the other hand, suppose that we know the Bezier coefficients \(p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top\), and we wish to fit timepoints \(t_1, ..., t_n\) to vectors \(x_1, ..., x_n\) lying on or near a Bezier polynomial of order \(R\).
 In this setting, we can fit each individual \(t_i\) independently by minimizing \(L(t_i; p, x_i) = \|x_i - g(t_i; p)\|^2\) for each \(t_i\).

### 2022-10-29 draft/draft.Rmd
updating accuracy for mbconnectome example

@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}). 
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 73.3\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 78.5\% accuracy for the left hemisphere connectome and 74.6\% accuracy for the right hemisphere connectome.
 
@@
 The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 8, cache = TRUE}
-set.seed(1)
+set.seed(12345)
 

### 2022-10-29 draft/draft.tex
updating accuracy for mbconnectome example

@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}).
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 73.3\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 78.5\% accuracy for the left hemisphere connectome and 74.6\% accuracy for the right hemisphere connectome.
 

### 2022-11-02 draft/draft.Rmd
still writing

--- 2022-10-28 draft/draft.Rmd
@@
 # output: rticles::rss_article
-fontsize: 11pt
+fontsize: 12pt
 geometry: "left=1in,right=1in,top=1in,bottom=1in"
@@
 For example, the combinatorial Laplacian eigenmap is an embedding in which Euclidean distances between pairs of vectors in the embedding are equal to expected commute times between corresponding pairs of vertices in the graph \citep{vonLuxburg2007}.
+This embedding is constructed by computing the combinatorial Laplacian matrix $L = D - A$ in which $D$ is a diagonal matrix in which each $D_{ii}$ is the degree of vertex $i$. 
+Then the components of the embedding consist of $\big[ v_{n-1} / \sqrt{\lambda_{n - 1}} \mid \cdots \mid v_1 / \sqrt{\lambda_1} \big]$ in which $\lambda_i$ is the $i^{th}$ largest eigenvalue of $L$ and $v_i$ is its corresponding eigenvector ($\lambda_n = 0$ and is omitted).
+Then Lloyd's algorithm can be applied to this embedding. 
+More often in practice, the $d$ smallest eigenvalues are used to construct a lower dimensional embedding. 
 \end{example}
@@
 In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.
+
 The contributions of this work are as follows.
-First, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
-Using estimators with well-established consistency properties for the generalized random dot product graph, we develop consistent community detection and parameter estimation algorithms for the popularity adjusted block model.
+
+In section \@ref(sec:pabm-grdpg), we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
+More specifically, we show that the popularity adjusted block model is a generalized random dot product graph in which the latent vectors are organized by community and each community in the latent space is defined by a $K$-dimensional subspace, where $K$ is the number of communities in the model.
+We then exploit this rigid and linear structure for community detection by applying sparse subspace clustering on the adjacency spectral embedding of the graph to infer the community labels. 
+We also propose a novel algorithm called orthogonal spectral clustering, which constructs an embedding for the popularity adjusted block model which forces the communities to lie on *orthogonal* subspaces.
+Finally, we propose a spectral decomposition based parameter estimation algorithm for the popularity adjusted block model. 
+All three algorithms have consistency properties that result in zero error with increasing graph size.
+
+In section \@ref(sec:mbm), we extend the latent structures of the stochastic block model, degree corrected block model, and popularity adjusted block model to nonlinear latent structures. 
+Instead of points, line segments, and subspaces, we propose the manifold block model which is a class of generalized random dot product graph in which the latent vectors lie on one of $K$ manifolds. 
+In the case of nonintersecting manifolds, we show that a clustering algorithm based on a nearest algorithms graph approach on an embedding is consistent, resulting in zero clustering error for large graphs, as long as the latent vectors are sampled from the manifolds densely
+In the case of intersecting manifolds, we propose the $K$-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds.
 
@@
 
+The latent structures of the SBM and DCBM when viewed as GRDPGs lend themselves to convenient community detection algorithms based on existing clustering algorithms for Euclidean data. 
+The general strategy for these models is to construct the appropriate ASE, choosing the number of components based on the number of communities and whether to use positive or negative eigenvalues depending on assortativity or disassortativity, and then to apply the appropriate Euclidean-based clustering algorithm for the specific structure.
+In the case of the SBM, after embedding in $K$ dimensions, since the original latent structure consists of $K$ point masses, an appropriate clustering algorithm might be one based on centroids, such as $K$-means clustering or Gaussian mixture models.
+In the case of the DCBM, after embedding in $K$ dimensions, since the original latent structure consists of $K$ rays emanating from the origin, an appropriate clustering algorithm is one based on angles between points rather than distances, such as kernel $K$-means clustering using the cosine kernel.
+Another clustering algorithm for the ASE of a DCBM is to normalize the embedding vectors to norm 1 and then apply $K$-means to the normalized vectors.
+
 \begin{example}
@@
 If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
-We will assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
-Because the fourth community only contains two members, we omit it from this analysis. 
+We assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
+Because the fourth community only contains two members, we omitted it from this analysis. 
 Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM. 
-Fitting three line segments to these data and clustering the embedding vectors by which line segment is closest results in a community detection accuracy of 83.5\%.
-
-```{r ukfaculty-ase, fig.cap = 'The ASE of the UK faculty network. The vertices are labeled by the school to which each faculty member belongs.', fig.height = 3, fig.width = 4, cache = TRUE }
-data(UKfaculty)
+$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%.
+
+```{r ukfaculty-ase, fig.cap = 'A three-dimensional ASE of the UK faculty network using the three most positive eigenvalues and their corresponding eigenvectors.
+The vertices are labeled by the school to which each faculty member belongs.', fig.height = 3, fig.width = 5, cache = TRUE }
+data(UKfaculty, package = 'igraphdata')
 A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
@@
 Xhat <- embedding(A, 3, 0)
-pairs(Xhat, col = z, asp = 1)
+# pairs(Xhat, col = z, asp = 1)
+Xhat %>% 
+  as.data.frame %>% 
+  magrittr::set_colnames(c('comp 1', 'comp 2', 'comp 3')) %>% 
+  GGally::ggpairs(aes(colour = z)) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  theme_bw()
 ```
+
 \end{example}
@@
 The PABM is also a Bernoulli random graph, so it is also a GRDPG. 
-We explore this in the next section (see section \@ref(sec:pabm-grdpg)).
+The latent vectors of the PABM when viewed as a GRDPG form $K$ $K$-dimensional subspaces, which makes for an elegant parallel between the nested forms of the block models and the nested structures of the GRDPGs: The SBM consists of $K$ zero-dimensional subspaces, and DCBM consists of $K$ one-dimensional subspaces.
+We explore this connection further the next section (section \@ref(sec:pabm-grdpg)).
 
@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}). 
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 73.3\% accuracy for the left hemisphere connectome and 76.5\% accuracy for the right hemisphere connectome.
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 72.2\% accuracy for the left hemisphere connectome and 83.6\% accuracy for the right hemisphere connectome.
+The curves were fitted on three-dimensional ASEs with $p = 2$ and $q = 1$, which were chosen based on setting the threshold $|\lambda_i| > \sqrt{n}$ where $\lambda_i$ is the $i$^th^ greatest eigenvalue.
+Figure \ref{fig:mbconnectome-kcurves} shows the embedding and curves projected to the first two components.
 
@@
 ```{r mbconnectome-ase, fig.cap = 'ASEs of the Drosophila connectome graphs. The embedding vectors are labeled by neuron type.', fig.height = 2, fig.width = 8}
-Xhat.l <- embedding(A.l, 2, 0)
+Xhat.l <- embedding(A.l, 2, 1)
 ase.l <- ggplot() + 
@@
 
-Xhat.r <- embedding(A.r, 2, 0)
+Xhat.r <- embedding(A.r, 2, 1)
 ase.r <- ggplot() + 
@@
 The embedding vectors are colored by predicted label.', fig.height = 2, fig.width = 8, cache = TRUE}
-set.seed(1)
-
-z.init.r <- sample.points(z.r, 10)
-z.init.l <- sample.points(z.l, 10)
+set.seed(123456)
+
+z.init.r <- sample.points(z.r, 6)
+z.init.l <- sample.points(z.l, 6)
 

### 2022-11-02 draft/draft.tex
still writing

@@
 \documentclass[
-  11pt,
+  12pt,
 ]{article}
@@
 In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.
+
 The contributions of this work are as follows.
-First, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
-Using estimators with well-established consistency properties for the generalized random dot product graph, we develop consistent community detection and parameter estimation algorithms for the popularity adjusted block model.
+
+In section \ref{sec:pabm-grdpg}, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
+More specifically, we show that the popularity adjusted block model is a generalized random dot product graph in which the latent vectors are organized by community and each community in the latent space is defined by a \(K\)-dimensional subspace, where \(K\) is the number of communities in the model.
+We then exploit this rigid and linear structure for community detection by applying sparse subspace clustering on the adjacency spectral embedding of the graph to infer the community labels.
+We also propose a novel algorithm called orthogonal spectral clustering, which constructs an embedding for the popularity adjusted block model which forces the communities to lie on \emph{orthogonal} subspaces.
+Finally, we propose a spectral decomposition based parameter estimation algorithm for the popularity adjusted block model.
+All three algorithms have consistency properties that result in zero error with increasing graph size.
+
+In section \ref{sec:mbm}, we extend the latent structures of the stochastic block model, degree corrected block model, and popularity adjusted block model to nonlinear latent structures.
+Instead of points, line segments, and subspaces, we propose the manifold block model which is a class of generalized random dot product graph in which the latent vectors lie on one of \(K\) manifolds.
+In the case of nonintersecting manifolds, we show that a clustering algorithm based on a nearest algorithms graph approach on an embedding is consistent, resulting in zero clustering error for large graphs, as long as the latent vectors are sampled from the manifolds densely
+In the case of intersecting manifolds, we propose the \(K\)-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds.
 
@@
 
+The latent structures of the SBM and DCBM when viewed as GRDPGs lend themselves to convenient community detection algorithms based on existing clustering algorithms for Euclidean data.
+The general strategy for these models is to construct the appropriate ASE, choosing the number of components based on the number of communities and whether to use positive or negative eigenvalues depending on assortativity or disassortativity, and then to apply the appropriate Euclidean-based clustering algorithm for the specific structure.
+In the case of the SBM, after embedding in \(K\) dimensions, since the original latent structure consists of \(K\) point masses, an appropriate clustering algorithm might be one based on centroids, such as \(K\)-means clustering or Gaussian mixture models.
+In the case of the DCBM, after embedding in \(K\) dimensions, since the original latent structure consists of \(K\) rays emanating from the origin, an appropriate clustering algorithm is one based on angles between points rather than distances, such as kernel \(K\)-means clustering using the cosine kernel.
+Another clustering algorithm for the ASE of a DCBM is to normalize the embedding vectors to norm 1 and then apply \(K\)-means to the normalized vectors.
+
 \begin{example}
@@
 If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
-We will assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
-Because the fourth community only contains two members, we omit it from this analysis. 
+We assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
+Because the fourth community only contains two members, we omitted it from this analysis. 
 Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM. 
-Fitting three line segments to these data and clustering the embedding vectors by which line segment is closest results in a community detection accuracy of 83.5\%.
+$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%.
 
@@
 
-\caption{The ASE of the UK faculty network. The vertices are labeled by the school to which each faculty member belongs.}\label{fig:ukfaculty-ase}
-\end{figure}
+\caption{A three-dimensional ASE of the UK faculty network using the three most positive eigenvalues and their corresponding eigenvectors.
+The vertices are labeled by the school to which each faculty member belongs.}\label{fig:ukfaculty-ase}
+\end{figure}
+
 \end{example}
@@
 The PABM is also a Bernoulli random graph, so it is also a GRDPG.
-We explore this in the next section (see section \ref{sec:pabm-grdpg}).
+The latent vectors of the PABM when viewed as a GRDPG form \(K\) \(K\)-dimensional subspaces, which makes for an elegant parallel between the nested forms of the block models and the nested structures of the GRDPGs: The SBM consists of \(K\) zero-dimensional subspaces, and DCBM consists of \(K\) one-dimensional subspaces.
+We explore this connection further the next section (section \ref{sec:pabm-grdpg}).
 
@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}).
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 78.5\% accuracy for the left hemisphere connectome and 74.6\% accuracy for the right hemisphere connectome.
+Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 72.2\% accuracy for the left hemisphere connectome and 83.6\% accuracy for the right hemisphere connectome.
+The curves were fitted on three-dimensional ASEs with \(p = 2\) and \(q = 1\), which were chosen based on setting the threshold \(|\lambda_i| > \sqrt{n}\) where \(\lambda_i\) is the \(i\)\textsuperscript{th} greatest eigenvalue.
+Figure \ref{fig:mbconnectome-kcurves} shows the embedding and curves projected to the first two components.
 

### 2022-11-02 draft/draft.Rmd
still writing

@@
 \label{thm:kcurves-semisupervised}
-Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a Bezier curve of order $R$. 
+Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a Bezier curve of order $R$ that does not self-intersect (for any $s \neq t$, $g_k(s) \neq g_k(t)$).
 Suppose that for each community $k$, we have labels for at least $R + 1$ vertices. 

### 2022-11-02 draft/draft.Rmd
fixed typo

@@
   Parameters = c('$K (K-1) / 2$', '$K (K-1) / 2 + n$', '$n K$'),
-  Dimension = c('$K$', '$K$', '$K^2$'),
+  `$\\mathrm{rank}(P)$` = c('$K$', '$K$', '$K^2$'),
   `Matrix Decomposition` = c('$P = Z B Z^\\top$', '$P = (\\Omega Z) B (\\Omega Z)^\\top$', '$P = X \\Pi X^\\top$'),

### 2022-11-02 draft/draft.tex
fixed typo

@@
 \hline
-Model & Parameters & Dimension & Matrix Decomposition & Reduction to simpler model\\
+Model & Parameters & $\mathrm{rank}(P)$ & Matrix Decomposition & Reduction to simpler model\\
 \hline
@@
 Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM. 
-$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%.
+$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, compared to an EM-based method which achieves 84\%.
 

### 2022-11-05 draft/draft.Rmd
fixed typo
minor changes

@@
 This embedding is constructed by computing the combinatorial Laplacian matrix $L = D - A$ in which $D$ is a diagonal matrix in which each $D_{ii}$ is the degree of vertex $i$. 
-Then the components of the embedding consist of $\big[ v_{n-1} / \sqrt{\lambda_{n - 1}} \mid \cdots \mid v_1 / \sqrt{\lambda_1} \big]$ in which $\lambda_i$ is the $i^{th}$ largest eigenvalue of $L$ and $v_i$ is its corresponding eigenvector ($\lambda_n = 0$ and is omitted).
+Then the components of the embedding consist of $\big[ v_{n-1} / \sqrt{\lambda_{n - 1}} \mid \cdots \mid v_1 / \sqrt{\lambda_1} \big]^\top$ in which $\lambda_i$ is the $i^{th}$ largest eigenvalue of $L$ and $v_i$ is its corresponding eigenvector ($\lambda_n = 0$ and is omitted).
 Then Lloyd's algorithm can be applied to this embedding. 
@@
 
-To estimate $\omega = \Big[ \omega_1, \cdots, \omega_n \Big]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
+To estimate $\omega = \big[ \omega_1, \cdots, \omega_n \big]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
 For simplicity, set $C = 1$. 
@@
 
-As established in section \@ref(sec:grdpg), all Bernoulli random graphs are generalized random dot product graphs. 
+As established in section \@ref(sec:sbm-dcbm-grdpg), it is possible to represent any Bernoulli random graph as a generalized random dot product graph. 
 Whether this is useful for community detection in block models depends entirely on the configuration of the latent vectors. 
@@
 
+For the sake of completeness, we begin with the case of GRDPGs in which communities are represented by nonintersecting manifolds. 
 In this section, we consider the following setting: Suppose that each community $C_k$ is represented by a closed manifold $\mathcal{M}_k$ in the latent space of a RDPG or GRDPG. 

### 2022-11-05 draft/draft.tex
minor changes

@@
 
-As established in section \ref{sec:grdpg}, all Bernoulli random graphs are generalized random dot product graphs.
+As established in section \ref{sec:sbm-dcbm-grdpg}, it is possible to represent any Bernoulli random graph as a generalized random dot product graph.
 Whether this is useful for community detection in block models depends entirely on the configuration of the latent vectors.
@@
 
+For the sake of completeness, we begin with the case of GRDPGs in which communities are represented by nonintersecting manifolds.
 In this section, we consider the following setting: Suppose that each community \(C_k\) is represented by a closed manifold \(\mathcal{M}_k\) in the latent space of a RDPG or GRDPG.

### 2022-11-11 draft/draft.Rmd
still editing

@@
 \argmin_{z_1, ..., z_n} ~ \tilde{W}(z_1, ..., z_n; x_1, ..., x_n) =
-\argmin_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
+\argmin_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k}  d^2(x_i, x_j).
 $$
-Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
-If they are vectors in Euclidean space, then we can use $d(x_i, x_j) = \|x_i - x_j\|_2$, 
-or if they are vertices on a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
+Here, $d^2(x_i, x_j)$ is the squared distance between objects $x_i$ and $x_j$. 
+If they are vectors in Euclidean space, then we can use $d^2(x_i, x_j) = \|x_i - x_j\|_2^2$, 
+or if they are vertices on a graph, then $d^2(x_i, x_j)$ represents some notion of graph distance. 
 While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
-An alternative algorithm is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.
+An alternative algorithm for $K$-means clustering in this setting is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.
 
@@
 In this work, we focus primarily on undirected and unweighted graphs without self loops. 
+Oftentimes, graph data are paired with vertex attributes. 
+For instance, in a social network graph, in which vertices correspond to individuals and edges correspond to connections between pairs of individuals, each vertex (individual) might also have a set or vector of other observations (e.g., age, sex, location, etc.).
+This is often the setting for inference in the *exponential family of random graphs* (ERGM) \citep{kolaczyk2014statistical}.
+Another setting in which vertices have attributes is when the (oftentimes single) attribute is hidden or unobserved, and the inference task is to identify the attributes. 
+If those hidden attributes are discrete labels, then the inference task is called *community detection*, which is the main interest in this work.
 
@@
 \label{ex:uk-faculty}
-\citet{PhysRevE.77.016107} constructed a friendship network among $81$ faculty from various schools at a university in the UK. 
+\citet{PhysRevE.77.016107} constructed a friendship network among $81$ faculty from various schools at a university in the UK (available in the \texttt{igraphdata} R package \citep{igraph}). 
 These data are represented as a graph in which each vertex is a faculty member and edges between pairs of vertices indicate whether the two faculty are friends. 
@@
 ```
+
+In the setting in which the labels are known and the edges are assumed to be sampled from some distribution conditioned on the labels, then one inference task is to fit an ERGM to these data. 
+On the other hand, if the labels are unobserved and the task is to identify the labels (up to permutation) assuming that the labels are drawn from some distribution and the edges are also drawn from some distribution conditioned on the hidden labels, then the inference task is community detection.
+
 \end{example}
@@
 The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights. 
-The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction indicating to which mixture a new observation belongs.
+The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components.  
 In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.
@@
 One possible estimator for the lone parameter of an $\ER$ graph, $\theta$, is the maximum likelihood estimator, which is found by maximizing the log-likelihood function $\ell(\theta; A) = \log \theta \sum\limits_{i < j} A_{ij} + \log (1 - \theta) \sum\limits_{i < j} (1 - A_{ij})$.
-This can be solved directly by setting $\frac{d \ell}{d \theta} = 0$, which yields $\hat{\theta} = \frac{2 |E|}{n (n-1)}$.
+This can be solved directly by setting $\frac{d \ell}{d \theta} = 0$, which yields $\hat{\theta} = \frac{2 |E|}{n (n-1)}$, and $|E| = \sum_{i<j} A_{ij}$ is the number of edges in the graph.
 \end{example}
@@
 The contributions of this work are as follows. 
-
 In section \@ref(sec:pabm-grdpg), we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
@@
 
+We conclude with section \@ref(sec:comparison) in which we compare the four block models.
+
 \newpage
@@
 This type of model framework allows us to frame the goal of community detection as a statistical inference problem: 
-come up with estimators for the true community labels, up to permutation.
+construct estimators for the true community labels, up to permutation.
 
@@
 ```{r dating-sbm, fig.width = 4, , fig.height = 2, fig.cap = 'Stochastic block model of a dating network. This model is disassortative.', cache = TRUE}
+set.seed(123)
+
 n <- 2 ** 6
@@
 
-$$
+\begin{equation}
+\label{eq:sbm-eigenvalues}
 \lambda = \frac{n}{2} \bigg( 
@@
 \bigg).
-$$
+\end{equation}
 The first eigenvalue is always positive. 
@@
 \ell(z, \theta; A) = \sum_{i < j} \sum_{k, \ell} z_{ik} z_{j \ell} \big( 
-A_{ij} \log \theta_{k \ell} - (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big),
+A_{ij} \log \theta_{k \ell} + (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big),
 \end{equation}
@@
 \sum_{i < j} \sum_{k, \ell} z_{ik} z_{j \ell} \big( 
-A_{ij} \log \theta_{k \ell} - (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big).
+A_{ij} \log \theta_{k \ell} + (1 - A_{ij}) \log (1 - \theta_{k \ell}) \big).
 \end{equation}
@@
       E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \hat{\theta}_{k \ell} + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell})) \bigg)$.\;
+        \For {$\ell = 1, ..., K$} {
           M-step: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
+        }
       }
@@
 The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of $\pi_{ik} = 0.5$, $\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1$, and $\hat{\theta}_{12} = 0.9$, results in a community detection error rate of 0\% and parameter estimates $\hat{\theta}_{11} = 0.034$, $\hat{\theta}_{22} = 0.018$, and $\hat{\theta}_{12} = 0.201$, compared to the true parameters $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$.
+
+\begin{example}
+\label{ex:uk-faculty-sbm}
+Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
+\end{example}
 
@@
 Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
-Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM. 
+Let $K \geq 1$ be an integer that describes the number of communities and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM.
 Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
@@
 If the edge probability matrix $P$ for this graph is such that $P_{ij} = \rho_n \theta_{z_i, z_j} \omega_i \omega_j$ and $A \sim \BG(P)$, then $G$ is a degree corrected block model. 
-We use the notation $A \sim \DCBM(z, \{\theta_{k \ell}\}_K, \omega; \rho_n)$ to denote a random adjacency matrix drawn from the DCBM with labels $z$, base edge probabilities $\{\theta_{k \ell}\}_K$, and degree correction factors $\omega$.
+We use the notation $A \sim \DCBM(z, \{\theta_{k \ell}\}_K, \omega; \rho_n)$ to denote a random adjacency matrix drawn from the DCBM with labels $z$, base edge probabilities $\{\theta_{k \ell}\}_K$ (of which there are $K (K+1) / 2$), and degree correction factors $\omega \in \mathbb{R}^n$.
 Under the sampling scheme in which the labels are drawn from a multivariate distribution with class probabilities $\alpha$, we use the notation $A \sim \DCBM(\alpha, \{\theta_{k \ell}\}_K, \omega; \rho_n)$.
@@
 In example \ref{ex:sbm-dating}, we modeled an online dating network as an SBM in which the vertices correspond to individuals, the communities correspond to genders, and edges correspond to whether the online dating service successfully matched pairs of individuals.
-In this model, we assume that for each pair of genders, there is a constant probability of a match. 
+In this model, there is an implicit assumption that for each pair of genders, there is a constant probability of a match, e.g., each male member has a constant probability of matching with another male member as well as a (different) constant probability of matching with a female member.
 This does not take into account that some individuals are more likely to form connections than others, based on each individual's activity on the online service or popularity with other users. 
-Modeling this as a DCBM allows us to take this into account by adjusting each user's degree by $\omega_i$. 
-The following is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n \equiv 1$.
+Modeling this as a DCBM allows us to take this into account by adjusting each user's expected degree by a factor of $\omega_i$. 
+The following is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n = 1$.
 
@@
 However, it still assumes that each vertex within a community has the same relative propensity to connect to each community. 
-For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex.
+Referring again to the dating network example (examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}), the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's preferences.
 The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community. 
-In the dating network example, the PABM allows us to model each individual's preferences instead of using population level averages.
+In the dating network example, the PABM allows us to model each individual's sexual orientation instead of using population level averages.
 
@@
 \label{ex:pabm-dating}
-In examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, we modeled an online dating network as a SBM and a DCBM, in which the vertices correspond to individuals, the communities correspond to genders, and the edges represent whether the online dating service successfully matched pairs of individuals.
-In these models, it is assumed that all individuals have the same sexual orientation. 
-Using the parameters specified in these examples, we assume that each individual is 10 times more likely to match with a member of the opposite sex compared to a member of the same sex. 
+In examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, we modeled an online dating network as a SBM and a DCBM, in which the vertices correspond to individuals, the community labels correspond to genders, and the edges represent whether the online dating service successfully matched pairs of individuals.
+In these models, it is assumed that all individuals have the same sexual orientation, i.e., the odds of matching with a member of the same gender vs.
+matching with a member of the opposite gender is constant across all members of each gender.
+Using the parameters specified in these examples, we assume that each individual is 10 times more likely to match with a member of the opposite gender compared to a member of the same gender. 
 The PABM allows us to model each individual's sexual orientation. 
@@
 
+Like the SBM (and most mixture models), fitting a DCBM or PABM to a graph via direct likelihood maximization is NP-hard. 
+However, like the SBM (and most mixture models), the EM algorithm is a candidate for (indirect) likelihood maximization. 
+The EM algorithms for the DCBM and PABM are similar to algorithm \ref{algo:sbm-em}, and the same independence relaxation is required.
+
+\begin{algorithm}
+  \label{algo:dcbm-em}
+  \setstretch{1.5}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$, estimated degree factors $\hat{\omega}$.}
+  \caption{Approximate EM algorithm for the DCBM.}
+  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}_K$, $\hat{\omega}$.\;
+  \While{$\|\nabla \ell\| > \epsilon$} {
+    \For {$i = 1, ..., n$} {
+      \For {$k = 1, ..., K$} {
+      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j) + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j)) \bigg)$.\;
+        \For {$\ell = 1, ..., K$} {
+          M-step for $\theta_{k \ell}$: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell} \hat{\omega}_i \hat{\omega}_j}$.\;
+        }
+      }
+      M-step for $\omega_i$: $\hat{\omega}_i = \frac{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} \hat{\theta}_{k \ell} \hat{\omega}_j}$.\;
+    }
+  }
+\end{algorithm}
+
+\begin{algorithm}
+  \label{algo:pabm-em}
+  \setstretch{1.5}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated popularity parameters $\{\hat{\lambda}_{ik}\}_K$.}
+  \caption{Approximate EM algorithm for the PABM.}
+  Initialize $\{\pi_{ik}\}$, $\{\hat{\lambda}_{ik}\}_K$.\;
+  \While{$\|\nabla \ell\| > \epsilon$} {
+    \For {$i = 1, ..., n$} {
+      \For {$k = 1, ..., K$} {
+      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{ik} \hat{\lambda}_{j \ell}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{ik} \hat{\lambda}_{j \ell})) \bigg)$.\;
+      M-step: $\hat{\lambda}_{ik} = \frac{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j \ell}}$.\;
+      }
+    }
+  }
+\end{algorithm}
+
 \begin{remark}
@@
 \citet{noroozi2019estimation} suggest focusing estimation on the edge probability matrix $P$ instead of the individual parameters to avoid this ambiguity. 
-Applying an EM algorithm similar to algorithm \ref{algo:sbm-em} adapted for the DCBM to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying a similar EM algorithm adapted for the PABM to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
+Nevertheless, the naive EM algorithms appear to work well, at least for community detection (estimating the labels $z_i$, e.g., by setting $\hat{z}_i = \arg\max_k \pi_{ik}$), based on empirical results.
+Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
 \end{remark}
 
+\begin{example}
+\label{ex:uk-faculty-dcbm-pabm}
+Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
+In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}). 
+\end{example}
+
 ## A Hierarchy of Block Models {#sec:hierarchy}
 
-The PABM, DCBM, and SBM are nested models, that is, the SBM is a special case of the DCBM which in turn is a special case of the SBM. 
+The PABM, DCBM, and SBM are nested models, that is, the SBM is a special case of the DCBM which in turn is a special case of the PABM. 
 In this section, these connections are made explicit, and we explore the implications for modeling. 
@@
 The latent vectors that generate the SBM consist of $K$ point masses in $\mathbb{R}^K$, each of which correspond to a community, and similarly, the latent vectors that generate the DCBM consist of $K$ line segments emitting from the origin in $\mathbb{R}^K$, each of which correspond to a community, and this property is observed in both assortative and disassortative models \citep{rubindelanchy2017statistical}.
-The ASE recovers these structures with noise, so clustering applied to the ASE can recover the original community labels. 
+The ASE recovers these structures with noise (which vanishes almost surely as $n \to \infty$), so clustering applied to the ASE can recover the original community labels. 
 \citet{rubindelanchy2017statistical} showed that this approach to community detection for the SBM and DCBM is asymptotically consistent. 
+More specifically, under certain sparsity conditions, clustering on the ASE of the SBM and DCBM results in zero misclustered vertices. 
 Note that for a given $P$, $X$ is not unique, i.e., for any $Q \in \mathbb{O}(p, q)$, $X Q$ results in the same $P$ (see remark \ref{rem:non_identifiable}). 
 However, this does not change the general structure of the SBM and DCBM in the latent space. 
-Multiplication by $Q \in \mathbb{O}(p,q)$ does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, because $Q$ is a linear transformation.
+Multiplication by $Q \in \mathbb{O}(p,q)$ does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, since $Q$ is a linear transformation.
 
@@
 \begin{example}
-Recall the UK faculty network in \ref{ex:uk-faculty}. 
+Recall the UK faculty network in example \ref{ex:uk-faculty}. 
+In examples~\ref{ex:uk-faculty-sbm} and \ref{ex:uk-faculty-dcbm-pabm}, we fit an SBM, DCBM, and PABM to the graph using their corresponding EM algorithms. 
+While all three algorithms were able to correctly identify the labels for most of the vertices, it is not clear which model is the most appropriate for this graph. 
+Analyzing the graph as a GRDPG, however, provides a visual indication for which block model is the best fit. 
 If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
@@
 Because the fourth community only contains two members, we omitted it from this analysis. 
-Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM. 
-$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, compared to an EM-based method which achieves 84\%.
+Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM, since each community appears to be described by a line segment. 
+$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, which is an improvement over all three block model EM algorithms.
 
@@
 
+\end{example}
+
+\begin{example}
+Disassortative block models can also be characterized as GRDPGs. 
+Consider the dating network examples for the SBM (example \ref{ex:sbm-dating}) and DCBM (example \ref{ex:dcbm-dating}). 
+Since the edge probability matrices are not positive semidefinite (equation \ref{eq:sbm-eigenvalues}), these are disassortative models. 
+More specifically, the edge probability matrix for each has one positive eigenvalue and one negative eigenvalue. 
+Then the ASE for this model is $\hat{X} = \Big[ |\lambda_1|^{1/2} v_1 \mid |\lambda_n|^{1/2} v_n \Big]$. 
+A plot of the ASE of the graph in \ref{ex:sbm-dating} (figure \ref{fig:dating-ase}, left) suggests two (somewhat overlapping) ellipsoidal clusters, and fitting a GMM on the embedding results in a community detection accuracy of 96.9\%, which is comparable to algorithm \ref{algo:sbm-em} while being much simpler to implement.
+Likewise, a plot of the ASE of the graph in example~\ref{ex:dcbm-dating} (figure \ref{fig:dating-ase}, right) consists of points that lie along two lines that cross at the origin, and kernel $K$-means clustering using the cosine similarity kernel results in a community detection accuracy of 93.8\%, which is an improvement over algorithm \ref{algo:dcbm-em} while being simpler to implement.
+Also note that aside from the choice of the eigenvalues and eigenvectors (positive vs negative), the community detection algorithms for the SBM ($K$-means clustering or GMM on the ASE) and DCBM (kernel $K$-means clustering) are identical regardless of whether the model is assortative or disassortative.
+
+```{r dating-ase, cache = TRUE, fig.cap = 'ASEs of the graphs in examples \\ref{ex:sbm-dating}, which is a disassortative SBM (left), and \\ref{ex:dcbm-dating}, which is a disassortative DCBM (right).
+The colors correspond to vertex label.
+Note that the embedding vectors for the SBM cluster around two point masses while the embedding vectors for the DCBM cluster around two line segments.'}
+set.seed(123)
+
+n <- 2 ** 6
+z <- sample(c(1, 2), n, replace = TRUE)
+z <- sort(z)
+n1 <- sum(z == 1)
+n2 <- sum(z == 2)
+z[z == 1] <- 'F'
+z[z == 2] <- 'M'
+z.sbm <- z
+theta11 <- .02
+theta22 <- .02
+theta12 <- .2
+P <- matrix(theta12, nrow = n, ncol = n)
+P[seq(n1), seq(n1)] <- theta11
+P[seq(n1 + 1, n), seq(n1 + 1, n)] <- theta22
+A <- draw.graph(P)
+Xhat <- embedding(A, 1, 1)
+sbm.ase <- ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z.sbm))) + 
+  labs(x = NULL, y = NULL, colour = 'gender') + 
+  coord_fixed() + 
+  theme_bw() + 
+  scale_colour_brewer(palette = 'Set1') + 
+  theme(legend.position = 'none')
+
+set.seed(12345)
+
+n <- 2 ** 6
+z <- sample(c(1, 2), n, replace = TRUE)
+z <- sort(z)
+n1 <- sum(z == 1)
+n2 <- sum(z == 2)
+z[z == 1] <- 'F'
+z[z == 2] <- 'M'
+z.dcbm <- z
+theta11 <- .03
+theta22 <- .03
+theta12 <- .3
+omega <- runif(n, 1/3, 1)
+Omega <- omega %*% t(omega)
+P <- matrix(theta12, nrow = n, ncol = n)
+P[seq(n1), seq(n1)] <- theta11
+P[seq(n1 + 1, n), seq(n1 + 1, n)] <- theta22
+P <- P * Omega
+A <- draw.graph(P)
+Yhat <- embedding(A, 1, 1)
+dcbm.ase <- ggplot() + 
+  geom_point(aes(x = Yhat[, 1], y = Yhat[, 2], colour = factor(z.dcbm))) + 
+  labs(x = NULL, y = NULL, colour = 'gender') + 
+  coord_fixed() + 
+  theme_bw() + 
+  scale_colour_brewer(palette = 'Set1') + 
+  theme(legend.position = 'none')
+
+gridExtra::grid.arrange(sbm.ase, dcbm.ase, nrow = 1)
+```
 \end{example}
@@
 In this section, we explore GRDPGs with community-wise parameterized nonlinear latent structures. 
-
-To motivate this model, consider a generalization of the $\ER$ model in which each edge probability isn't constant but comes from a simple distribution. 
-As with all Bernoulli random graphs, this is also a GRDPG, but by using a specific type of distribution, it is possible to construct specific latent structures on which the latent vectors lie. 
-Or, in reverse, by constructing specific latent structures, the edge probabilities of the Bernoulli random graph follow specific distributions.
+We begin with a motivating example.
+
+\begin{example}[Macaque visuotactile brain areas and connections]
+\label{ex:macaque}
+\citet{https://doi.org/10.1111/j.1460-9568.2006.04678.x} constructed a graph based on observed 45 brain regions and 463 connections of the macaque monkey. 
+The regions were then labeled by whether the brain region corresponds to a visual function or a sensorimotor function. 
+This graph is visualized in figure \ref{fig:macaque-graph}. 
+The data are available in the \texttt{igraphdata} package \citep{igraph}.
+
+Treating this graph as an RDPG with a latent space dimension of 2, 
+we decomposed the adjacency matrix to construct its ASE (figure \ref{fig:macaque-ase}). 
+The embedding vectors, which approximate the latent vectors, appear to approximately lie along two curves that intersect at the origin, with each curve corresponding to a label (visual vs sensorimotor).
+While the ASE suggests a community-wise structure, the SBM, DCBM, and PABM are not suitable models for these data.
+
+```{r macaque-graph, fig.cap = 'Macaque visuotactile brain areas and connections network. The vertices represent brain areas.', fig.height = 2, fig.width = 3}
+data(macaque, package = 'igraphdata')
+A <- as.matrix(igraph::as_adjacency_matrix(macaque, type = 'both'))
+A <- sign(A + t(A))
+z <- as.numeric(factor(igraph::vertex_attr(macaque)$shape))
+
+ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+```
+
+```{r macaque-ase, fig.cap = 'ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.', fig.width = 4, fig.height = 2}
+Xhat <- embedding(A, 3, 0)
+ggplot() + 
+  geom_point(aes(x = Xhat[, 1],
+                 y = Xhat[, 2],
+                 colour = factor(z))) + 
+  labs(colour = NULL, y = expression(x[2]), x = expression(x[1])) + 
+  scale_colour_brewer(palette = 'Set1') + 
+  coord_fixed() + 
+  theme_bw()
+```
+\label{macaque}
+\end{example}
+
+One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that there are $K$ manifolds in the latent space on which the latent vectors lie. 
+Then each manifold can be thought of as a community. 
+The goal of community detection in this setting is to partition the graph that results from this latent configuration such that each partition maps onto each manifold. 
+Before we dive into the multi-manifold case, we can first consider the single manifold case as a generalization of the $\ER$ model.
 
@@
 \citet{https://doi.org/10.48550/arxiv.1705.03297} used GMMs on the ASE for community detection and found that the curve-like structures in the ASE were incompatible with centroid-based clustering methods.
-\citet{SannaPassino2022} used Gaussian Processes to estimate curves, which resulted in better empirical fits. 
-\citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on the manifolds rather than the manifolds themselves.
-
-Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in $\mathbb{R}^d$ and identifying vectors belonging to different manifolds in $\mathbb{R}^d$.
+\citet{SannaPassino2022} used Gaussian processes to estimate curves, which resulted in better empirical fits. 
+\citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on a single manifold rather than estimating the manifolds or identifying multiple manifolds.
+\citet{https://doi.org/10.48550/arxiv.2006.05168} and \citet{https://doi.org/10.48550/arxiv.2208.11665} noted that kernel GRDPGs (GRDPGs in which $P_{ij} = \kappa(x_i, x_j)$ for some kernel function $\kappa$) with community structure can have large dimensions and proposed using a lower dimension with manifold structure to approximate the kernel function.
+
+Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in $\mathbb{R}^d$ or identifying vectors belonging to different manifolds in $\mathbb{R}^d$. 
+Manifold learning itself is a large area of research with a long history \citep{Tenenbaum2000-ff, trosset2020rehabilitating, https://doi.org/10.48550/arxiv.2107.08089}, so we will focus primarily on the setting in which there are multiple manifolds and the goal is to identify which vector lies on which manifold (manifold clustering).
+The standard approach to manifold clustering is essentially a type of spectral clustering, involving spectral decomposition of some kernel matrix applied to the vectors lying on manifolds (or its graph Laplacian matrix).
+Unlike standard manifold clustering techniques such as ratio cut, it has been observed that while the leading eigenvalues discern subspaces or approximations of subspaces, some of the other eigenvalues or a mixture of other eigenvalues help discern nonlinear structures \citep{10945-32767, https://doi.org/10.48550/arxiv.2208.11665, https://doi.org/10.48550/arxiv.2201.10000}.
+Most of these methods also assume that the vectors lie exactly on the manifolds (i.e., the observed vector is $x_i \in \mathcal{M}_k$) instead of approximately on them (i.e., the observed vector is $x_i = y_i + \epsilon_i$ where $y_i \in \mathcal{M}_k$ and $\| \epsilon_i \|$ is small in some sense), which is the setting in which the clustering is done on the ASE.
 
@@
 
-In the first example, we applied $K$-curves clustering (algorithm \ref{alg:kcurves}) to the Macaque visuotactile brain areas and connections network \citep{https://doi.org/10.1111/j.1460-9568.2006.04678.x}, which is visualized in figure \ref{fig:macaque-graph}.
-In this graph, there are 45 vertices which each correspond to a brain area, and the communities represent each area's tactile function. 
-Applying an ASE with $p = 2$ and $q = 0$ to this graph results in an embedding that appears to fall approximately on two curves (figure \ref{fig:macaque-ase}) which intersect at the origin. 
+In the first example, we applied $K$-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}. 
+As in \ref{ex:macaque}, we used $d = 2$ to construct the ASE, then used $K$-curves clustering to fit two curves to these data and predict the labels. 
 Figure \ref{fig:macaque-fitted} shows the result of $K$-curves clustering on this graph, 
@@
 
-```{r macaque-graph, fig.cap = 'Macaque visuotactile brain areas and connections network. The vertices represent brain areas.', fig.height = 2, fig.width = 3}
-data(macaque, package = 'igraphdata')
+```{r macaque-fitted, fig.cap = '$K$-means clustering on the ASE of the Macaque brain areas network.
+The embedding vectors are colored by predicted label.', fig.width = 4, fig.height = 2, cache = TRUE}
 A <- as.matrix(igraph::as_adjacency_matrix(macaque, type = 'both'))
@@
 
-ggraph(A, layout = 'stress') + 
-  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
-  geom_node_point(aes(colour = factor(z)), 
-                  show.legend = FALSE) + 
-  labs(colour = NULL) + 
-  theme_void() + 
-  scale_colour_brewer(palette = 'Set1')
-```
-
-```{r macaque-ase, fig.cap = 'ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.', fig.width = 4, fig.height = 2}
 Xhat <- embedding(A, 2, 0)
-ggplot() + 
-  geom_point(aes(x = Xhat[, 1],
-                 y = Xhat[, 2],
-                 colour = factor(z))) + 
-  labs(colour = NULL, y = expression(x[2]), x = expression(x[1])) + 
-  scale_colour_brewer(palette = 'Set1') + 
-  coord_fixed() + 
-  theme_bw()
-```
-
-```{r macaque-fitted, fig.cap = '$K$-means clustering on the ASE of the Macaque brain areas network.
-The embedding vectors are colored by predicted label.', fig.width = 4, fig.height = 2, cache = TRUE}
+
 set.seed(123)
@@
 
-## Conclusions
+# Comparisons of Models and Conclusions {#sec:comparison}
 

### 2022-11-11 draft/draft.tex
still editing

@@
 \mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \tilde{W}(z_1, ..., z_n; x_1, ..., x_n) =
-\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} \big( d(x_i, x_j) \big)^2.
+\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k}  d^2(x_i, x_j).
 $$
-Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
-If they are vectors in Euclidean space, then we can use $d(x_i, x_j) = \|x_i - x_j\|_2$, 
-or if they are vertices on a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
+Here, $d^2(x_i, x_j)$ is the squared distance between objects $x_i$ and $x_j$. 
+If they are vectors in Euclidean space, then we can use $d^2(x_i, x_j) = \|x_i - x_j\|_2^2$, 
+or if they are vertices on a graph, then $d^2(x_i, x_j)$ represents some notion of graph distance. 
 While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
-An alternative algorithm is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.
+An alternative algorithm for $K$-means clustering in this setting is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.
 
@@
 In this work, we focus primarily on undirected and unweighted graphs without self loops.
+Oftentimes, graph data are paired with vertex attributes.
+For instance, in a social network graph, in which vertices correspond to individuals and edges correspond to connections between pairs of individuals, each vertex (individual) might also have a set or vector of other observations (e.g., age, sex, location, etc.).
+This is often the setting for inference in the \emph{exponential family of random graphs} (ERGM) \citep{kolaczyk2014statistical}.
+Another setting in which vertices have attributes is when the (oftentimes single) attribute is hidden or unobserved, and the inference task is to identify the attributes.
+If those hidden attributes are discrete labels, then the inference task is called \emph{community detection}, which is the main interest in this work.
 
@@
 \end{figure}
+
+In the setting in which the labels are known and the edges are assumed to be sampled from some distribution conditioned on the labels, then one inference task is to fit an ERGM to these data. 
+On the other hand, if the labels are unobserved and the task is to identify the labels (up to permutation) assuming that the labels are drawn from some distribution and the edges are also drawn from some distribution conditioned on the hidden labels, then the inference task is community detection.
+
 \end{example}
@@
 The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights.
-The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction indicating to which mixture a new observation belongs.
+The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components.\\
 In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.
@@
 The contributions of this work are as follows.
-
 In section \ref{sec:pabm-grdpg}, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
@@
 
+We conclude with section \ref{sec:comparison} in which we compare the four block models.
+
 \newpage
@@
 This type of model framework allows us to frame the goal of community detection as a statistical inference problem:
-come up with estimators for the true community labels, up to permutation.
+construct estimators for the true community labels, up to permutation.
 
@@
 The same algorithm applied to example \ref{ex:sbm-dating}, which is a disassortative model, using initial guesses of \(\pi_{ik} = 0.5\), \(\hat{\theta}_{11} = \hat{\theta}_{22} = 0.1\), and \(\hat{\theta}_{12} = 0.9\), results in a community detection error rate of 0\% and parameter estimates \(\hat{\theta}_{11} = 0.034\), \(\hat{\theta}_{22} = 0.018\), and \(\hat{\theta}_{12} = 0.201\), compared to the true parameters \(\theta_{11} = \theta_{22} = 0.02\) and \(\theta_{12} = 0.2\).
+
+\begin{example}
+\label{ex:uk-faculty-sbm}
+Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
+\end{example}
 
@@
 However, it still assumes that each vertex within a community has the same relative propensity to connect to each community.
-For example, going back to the dating network in examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the models assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex.
+Referring again to the dating network example (examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}), the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's preferences.
 The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community.
-In the dating network example, the PABM allows us to model each individual's preferences instead of using population level averages.
+In the dating network example, the PABM allows us to model each individual's sexual orientation instead of using population level averages.
 
@@
 \end{example}
+
+Like the SBM (and most mixture models), fitting a DCBM or PABM to a graph via direct likelihood maximization is NP-hard.
+However, like the SBM (and most mixture models), the EM algorithm is a candidate for (indirect) likelihood maximization.
+The EM algorithms for the DCBM and PABM are similar to algorithm \ref{algo:sbm-em}, and the same independence relaxation is required.
+
+\begin{algorithm}
+  \label{algo:dcbm-em}
+  \setstretch{1.5}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$, estimated degree factors $\hat{\omega}$.}
+  \caption{Approximate EM algorithm for the DCBM.}
+  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}_K$, $\hat{\omega}$.\;
+  \While{$\|\nabla \ell\| > \epsilon$} {
+    \For {$i = 1, ..., n$} {
+      \For {$k = 1, ..., K$} {
+      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j) + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j)) \bigg)$.\;
+        \For {$\ell = 1, ..., K$} {
+          M-step for $\theta_{k \ell}$: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell} \hat{\omega}_i \hat{\omega}_j}$.\;
+        }
+      }
+      M-step for $\omega_i$: $\hat{\omega}_i = \frac{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} \hat{\theta}_{k \ell} \hat{\omega}_j}$.\;
+    }
+  }
+\end{algorithm}
+
+\begin{algorithm}
+  \label{algo:pabm-em}
+  \setstretch{1.5}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \KwData{Adjacency matrix $A$, number of communities $K$}
+  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated popularity parameters $\{\hat{\lambda}_{ik}\}_K$.}
+  \caption{Approximate EM algorithm for the PABM.}
+  Initialize $\{\pi_{ik}\}$, $\{\hat{\lambda}_{ik}\}_K$.\;
+  \While{$\|\nabla \ell\| > \epsilon$} {
+    \For {$i = 1, ..., n$} {
+      \For {$k = 1, ..., K$} {
+      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{ik} \hat{\lambda}_{j \ell}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{ik} \hat{\lambda}_{j \ell})) \bigg)$.\;
+      M-step: $\hat{\lambda}_{ik} = \frac{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j \ell}}$.\;
+      }
+    }
+  }
+\end{algorithm}
 
@@
 \citet{noroozi2019estimation} suggest focusing estimation on the edge probability matrix $P$ instead of the individual parameters to avoid this ambiguity. 
-Applying an EM algorithm similar to algorithm \ref{algo:sbm-em} adapted for the DCBM to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying a similar EM algorithm adapted for the PABM to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
+Nevertheless, the naive EM algorithms appear to work well, at least for community detection (estimating the labels $z_i$, e.g., by setting $\hat{z}_i = \arg\max_k \pi_{ik}$), based on empirical results.
+Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
 \end{remark}
+
+\begin{example}
+\label{ex:uk-faculty-dcbm-pabm}
+Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
+In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}). 
+\end{example}
 
@@
 
-The PABM, DCBM, and SBM are nested models, that is, the SBM is a special case of the DCBM which in turn is a special case of the SBM.
+The PABM, DCBM, and SBM are nested models, that is, the SBM is a special case of the DCBM which in turn is a special case of the PABM.
 In this section, these connections are made explicit, and we explore the implications for modeling.
@@
 The latent vectors that generate the SBM consist of \(K\) point masses in \(\mathbb{R}^K\), each of which correspond to a community, and similarly, the latent vectors that generate the DCBM consist of \(K\) line segments emitting from the origin in \(\mathbb{R}^K\), each of which correspond to a community, and this property is observed in both assortative and disassortative models \citep{rubindelanchy2017statistical}.
-The ASE recovers these structures with noise, so clustering applied to the ASE can recover the original community labels.
+The ASE recovers these structures with noise (which vanishes almost surely as \(n \to \infty\)), so clustering applied to the ASE can recover the original community labels.
 \citet{rubindelanchy2017statistical} showed that this approach to community detection for the SBM and DCBM is asymptotically consistent.
+More specifically, under certain sparsity conditions, clustering on the ASE of the SBM and DCBM results in zero misclustered vertices.
 Note that for a given \(P\), \(X\) is not unique, i.e., for any \(Q \in \mathbb{O}(p, q)\), \(X Q\) results in the same \(P\) (see remark \ref{rem:non_identifiable}).
 However, this does not change the general structure of the SBM and DCBM in the latent space.
-Multiplication by \(Q \in \mathbb{O}(p,q)\) does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, because \(Q\) is a linear transformation.
+Multiplication by \(Q \in \mathbb{O}(p,q)\) does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, since \(Q\) is a linear transformation.
 
@@
 
+\end{example}
+
+\begin{example}
+Disassortative block models can also be characterized as GRDPGs. 
+Consider the dating network examples for the SBM (example \ref{ex:sbm-dating}) and DCBM (example \ref{ex:dcbm-dating}). 
+Since the edge probability matrices are not positive semidefinite (equation \ref{eq:sbm-eigenvalues}), these are disassortative models. 
+More specifically, the edge probability matrix for each has one positive eigenvalue and one negative eigenvalue. 
+Then the ASE for this model is $\hat{X} = \Big[ |\lambda_1|^{1/2} v_1 \mid |\lambda_n|^{1/2} v_n \Big]$. 
+A plot of the ASE of the graph in \ref{ex:sbm-dating} (figure \ref{fig:dating-ase}, left) suggests two (somewhat overlapping) ellipsoidal clusters, and fitting a GMM on the embedding results in a community detection accuracy of 96.9\%, which is comparable to algorithm \ref{algo:sbm-em} while being much simpler to implement.
+Likewise, a plot of the ASE of the graph in example~\ref{ex:dcbm-dating} (figure \ref{fig:dating-ase}, right) consists of points that lie along two lines that cross at the origin, and kernel $K$-means clustering using the cosine similarity kernel results in a community detection accuracy of 93.8\%, which is an improvement over algorithm \ref{algo:dcbm-em} while being simpler to implement.
+Also note that aside from the choice of the eigenvalues and eigenvectors (positive vs negative), the community detection algorithms for the SBM ($K$-means clustering or GMM on the ASE) and DCBM (kernel $K$-means clustering) are identical regardless of whether the model is assortative or disassortative.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/dating-ase-1}
+
+}
+
+\caption{ASEs of the graphs in examples \ref{ex:sbm-dating}, which is a disassortative SBM (left), and \ref{ex:dcbm-dating}, which is a disassortative DCBM (right).
+The colors correspond to vertex label.
+Note that the embedding vectors for the SBM cluster around two point masses while the embedding vectors for the DCBM cluster around two line segments.}\label{fig:dating-ase}
+\end{figure}
 \end{example}
@@
 In this section, we explore GRDPGs with community-wise parameterized nonlinear latent structures.
-
-To motivate this model, consider a generalization of the \(\ER\) model in which each edge probability isn't constant but comes from a simple distribution.
-As with all Bernoulli random graphs, this is also a GRDPG, but by using a specific type of distribution, it is possible to construct specific latent structures on which the latent vectors lie.
-Or, in reverse, by constructing specific latent structures, the edge probabilities of the Bernoulli random graph follow specific distributions.
+We begin with a motivating example.
+
+\begin{example}[Macaque visuotactile brain areas and connections]
+\label{ex:macaque}
+\citet{https://doi.org/10.1111/j.1460-9568.2006.04678.x} constructed a graph based on observed 45 brain regions and 463 connections of the macaque monkey. 
+The regions were then labeled by whether the brain region corresponds to a visual function or a sensorimotor function. 
+This graph is visualized in figure \ref{fig:macaque-graph}. 
+The data are available in the \texttt{igraphdata} package \citep{igraph}.
+
+Treating this graph as an RDPG with a latent space dimension of 2, 
+we decomposed the adjacency matrix to construct its ASE (figure \ref{fig:macaque-ase}). 
+The embedding vectors, which approximate the latent vectors, appear to approximately lie along two curves that intersect at the origin, with each curve corresponding to a label (visual vs sensorimotor).
+While the ASE suggests a community-wise structure, the SBM, DCBM, and PABM are not suitable models for these data.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-graph-1}
+
+}
+
+\caption{Macaque visuotactile brain areas and connections network. The vertices represent brain areas.}\label{fig:macaque-graph}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-ase-1}
+
+}
+
+\caption{ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.}\label{fig:macaque-ase}
+\end{figure}
+\label{macaque}
+\end{example}
+
+One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that there are \(K\) manifolds in the latent space on which the latent vectors lie.
+Then each manifold can be thought of as a community.
+The goal of community detection in this setting is to partition the graph that results from this latent configuration such that each partition maps onto each manifold.
+Before we dive into the multi-manifold case, we can first consider the single manifold case as a generalization of the \(\ER\) model.
 
@@
 \citet{https://doi.org/10.48550/arxiv.1705.03297} used GMMs on the ASE for community detection and found that the curve-like structures in the ASE were incompatible with centroid-based clustering methods.
-\citet{SannaPassino2022} used Gaussian Processes to estimate curves, which resulted in better empirical fits.
-\citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on the manifolds rather than the manifolds themselves.
-
-Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in \(\mathbb{R}^d\) and identifying vectors belonging to different manifolds in \(\mathbb{R}^d\).
+\citet{SannaPassino2022} used Gaussian processes to estimate curves, which resulted in better empirical fits.
+\citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on a single manifold rather than estimating the manifolds or identifying multiple manifolds.
+\citet{https://doi.org/10.48550/arxiv.2006.05168} and \citet{https://doi.org/10.48550/arxiv.2208.11665} noted that kernel GRDPGs (GRDPGs in which \(P_{ij} = \kappa(x_i, x_j)\) for some kernel function \(\kappa\)) with community structure can have large dimensions and proposed using a lower dimension with manifold structure to approximate the kernel function.
+
+Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in \(\mathbb{R}^d\) or identifying vectors belonging to different manifolds in \(\mathbb{R}^d\).
+Manifold learning itself is a large area of research with a long history \citep{Tenenbaum2000-ff, trosset2020rehabilitating, https://doi.org/10.48550/arxiv.2107.08089}, so we will focus primarily on the setting in which there are multiple manifolds and the goal is to identify which vector lies on which manifold (manifold clustering).
+The standard approach to manifold clustering is essentially a type of spectral clustering, involving spectral decomposition of some kernel matrix applied to the vectors lying on manifolds (or its graph Laplacian matrix).
+Unlike standard manifold clustering techniques such as ratio cut, it has been observed that while the leading eigenvalues discern subspaces or approximations of subspaces, some of the other eigenvalues or a mixture of other eigenvalues help discern nonlinear structures \citep{10945-32767, https://doi.org/10.48550/arxiv.2208.11665, https://doi.org/10.48550/arxiv.2201.10000}.
+Most of these methods also assume that the vectors lie exactly on the manifolds (i.e., the observed vector is \(x_i \in \mathcal{M}_k\)) instead of approximately on them (i.e., the observed vector is \(x_i = y_i + \epsilon_i\) where \(y_i \in \mathcal{M}_k\) and \(\| \epsilon_i \|\) is small in some sense), which is the setting in which the clustering is done on the ASE.
 
@@
 
-In the first example, we applied \(K\)-curves clustering (algorithm \ref{alg:kcurves}) to the Macaque visuotactile brain areas and connections network \citep{https://doi.org/10.1111/j.1460-9568.2006.04678.x}, which is visualized in figure \ref{fig:macaque-graph}.
-In this graph, there are 45 vertices which each correspond to a brain area, and the communities represent each area's tactile function.
-Applying an ASE with \(p = 2\) and \(q = 0\) to this graph results in an embedding that appears to fall approximately on two curves (figure \ref{fig:macaque-ase}) which intersect at the origin.
+In the first example, we applied \(K\)-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}.
+As in \ref{ex:macaque}, we used \(d = 2\) to construct the ASE, then used \(K\)-curves clustering to fit two curves to these data and predict the labels.
 Figure \ref{fig:macaque-fitted} shows the result of \(K\)-curves clustering on this graph,
 which yields a 4.4\% error rate.
-
-\begin{figure}[H]
-
-{\centering \includegraphics{draft_files/figure-latex/macaque-graph-1}
-
-}
-
-\caption{Macaque visuotactile brain areas and connections network. The vertices represent brain areas.}\label{fig:macaque-graph}
-\end{figure}
-
-\begin{figure}[H]
-
-{\centering \includegraphics{draft_files/figure-latex/macaque-ase-1}
-
-}
-
-\caption{ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.}\label{fig:macaque-ase}
-\end{figure}
 
@@
 
-\hypertarget{conclusions}{%
-\subsection{Conclusions}\label{conclusions}}
+\hypertarget{sec:comparison}{%
+\section{Comparisons of Models and Conclusions}\label{sec:comparison}}
 

### 2022-11-12 draft/draft.Rmd
still writing

@@
 source('~/dev/manifold-block-models/functions.R')
+source('~/dev/dissertation/em-algorithms.R')
 
@@
       E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \hat{\theta}_{k \ell} + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell})) \bigg)$.\;
+      }
+    }
+    \For {$k = 1, ..., K$} {
       \For {$\ell = 1, ..., K$} {
         M-step: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
-        }
       }
@@
 Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
+Suppose we assume that this graph is an SBM. 
+Then applying algorithm \ref{algo:sbm-em} to this network yields a community detection accuracy of 79.7\%. 
+For this analysis, the school with two faculty members is omitted (so $K=3$ and $n = 79$), and the EM algorithm was initialized with random $\{\pi_{ik}\}$, $\hat{\theta}_{k k} = 0.9$, and $\hat{\theta}_{k \ell} = 0.1$ for $k \neq \ell$.
+```{r cache = TRUE}
+data(UKfaculty, package = 'igraphdata')
+A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
+A <- sign(A + t(A))
+z <- factor(igraph::vertex_attr(UKfaculty)$Group)
+A <- A[z != 4, z != 4]
+z <- z[z != 4]
+K <- 3
+n <- length(z)
+
+set.seed(123)
+out <- em.sbm(A, 
+              t(apply(matrix(runif(n * K), nrow = n), 
+                      1, function(x) x / sum(x))), 
+              matrix(c(.9, .1, .1, .1, .9, .1, .1, .1, .9), nrow = 3))
+```
 \end{example}
@@
   \caption{Approximate EM algorithm for the DCBM.}
-  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}_K$, $\hat{\omega}$.\;
+  Initialize $\{\pi_{ik}\}$, $\{\hat{\theta}_{k \ell}\}_K$, $\hat{\omega}$.\;
   \While{$\|\nabla \ell\| > \epsilon$} {
@@
       E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j) + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j)) \bigg)$.\;
+      }
+    }
+    \For {$k = 1, ..., K$} {
       \For {$\ell = 1, ..., K$} {
@@
     }
+    \For {$i = 1, ..., n$} {
       M-step for $\omega_i$: $\hat{\omega}_i = \frac{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} \hat{\theta}_{k \ell} \hat{\omega}_j}$.\;
@@
         E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{ik} \hat{\lambda}_{j \ell}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{ik} \hat{\lambda}_{j \ell})) \bigg)$.\;
+      }
+    }
+    \For {$i = 1, ..., n$} {
+      \For {$k = 1, ..., K$} {
         M-step: $\hat{\lambda}_{ik} = \frac{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j \ell}}$.\;
@@
 Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
-In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}). 
+In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}), yielding a community detection accuracy of 79.7\%. 
+Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} yields a greater accuracy of 94.9\%. 
+This is expected, since the SBM is a special case of the DCBM, i.e., the SBM is the DCBM with some of the parameters restricted. 
+Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM. 
+Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a better fit (figure \ref{fig:uk-faculty-dcbm-omega}). 
+```{r uk-faculty-dcbm-omega, cache = TRUE, fig.width = 8, fig.height = 2, fig.cap = 'Histogram of the fitted degree factors for the UK faculty friendship network assuming that it is a DCBM.
+The DCBM was fit to this graph using algorithm \\ref{algo:dcbm-em}.
+The dispersion of the degree factors suggests that the DCBM is a better fit than the SBM.'}
+data(UKfaculty, package = 'igraphdata')
+A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
+A <- sign(A + t(A))
+z <- factor(igraph::vertex_attr(UKfaculty)$Group)
+A <- A[z != 4, z != 4]
+z <- z[z != 4]
+K <- 3
+n <- length(z)
+
+set.seed(123)
+out <- em.sbm(A, 
+              t(apply(matrix(runif(n * K), nrow = n), 1, function(x) x / sum(x))),
+              matrix(c(.9, .1, .1, .1, .9, .1, .1, .1, .9), nrow = K))
+out <- em.dcbm(A, 
+               out$pi,
+               out$theta,
+               runif(n))
+
+ggplot() + 
+  geom_histogram(aes(x = out$omega),
+                 colour = 'black', fill = 'white') + 
+  labs(x = 'Fitted degree factors for the UK faculty network') + 
+  theme_bw()
+```
 \end{example}
@@
 $$\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)$$
-where $x^\top$ and $\hat{x}^\top$ are the rows of $\hat{X}$ (from the ASE) and $X$, respectively.
+where $x_i^\top$ and $\hat{x}_i^\top$ are the rows of $\hat{X}$ (from the ASE) and $X$, respectively.
 
@@
 In examples~\ref{ex:uk-faculty-sbm} and \ref{ex:uk-faculty-dcbm-pabm}, we fit an SBM, DCBM, and PABM to the graph using their corresponding EM algorithms. 
-While all three algorithms were able to correctly identify the labels for most of the vertices, it is not clear which model is the most appropriate for this graph. 
+While all three algorithms were able to correctly identify the labels for most of the vertices, it is not immediately clear which model is the most appropriate for this graph. 
 Analyzing the graph as a GRDPG, however, provides a visual indication for which block model is the best fit. 
@@
 
+\newpage
+
 # Comparisons of Models and Conclusions {#sec:comparison}
@@
 
-\section*{Appendix A: Proofs of Theorems from Section 3}
+\section{Appendix A: Proofs of Theorems from Section 3}
 
@@
 
-\section*{Appendix B: Proofs of Theorems from Section 4}
+\section{Appendix B: Proofs of Theorems from Section 4}
 
@@
 
-\section*{Appendix C: Details on Fitting Bezier Curves to Noisy Data}
+\section{Appendix C: Details on Fitting Bezier Curves to Noisy Data}
 
@@
 
-\section*{Appendix D: Supplementary Materials}
+\section{Appendix D: Supplementary Materials}
 

### 2022-11-12 draft/draft.tex
still writing

@@
 \theoremstyle{remark}
-\newtheorem*{remark}{Remark}
+\newtheorem{remark}{Remark}
 \newtheorem*{solution}{Solution}
@@
 Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
+Suppose we assume that this graph is an SBM. 
+Then applying algorithm \ref{algo:sbm-em} to this network yields a community detection accuracy of 79.7\%. 
+For this analysis, the school with two faculty members is omitted (so $K=3$ and $n = 79$), and the EM algorithm was initialized with random $\{\pi_{ik}\}$, $\hat{\theta}_{k k} = 0.9$, and $\hat{\theta}_{k \ell} = 0.1$ for $k \neq \ell$.
+
 \end{example}
@@
 Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
-In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}). 
+In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}), yielding a community detection accuracy of 79.7\%. 
+Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} yields a greater accuracy of 94.9\%. 
+This is expected, since the SBM is a special case of the DCBM, i.e., the SBM is the DCBM with some of the parameters restricted. 
+Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM. 
+Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a better fit (figure \ref{fig:uk-faculty-dcbm-omega}). 
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/uk-faculty-dcbm-omega-1}
+
+}
+
+\caption{Histogram of the fitted degree factors for the UK faculty friendship network assuming that it is a DCBM.
+The DCBM was fit to this graph using algorithm \ref{algo:dcbm-em}.
+The dispersion of the degree factors suggests that the DCBM is a better fit than the SBM.}\label{fig:uk-faculty-dcbm-omega}
+\end{figure}
 \end{example}
@@
 \[\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)\]
-where \(x^\top\) and \(\hat{x}^\top\) are the rows of \(\hat{X}\) (from the ASE) and \(X\), respectively.
+where \(x_i^\top\) and \(\hat{x}_i^\top\) are the rows of \(\hat{X}\) (from the ASE) and \(X\), respectively.
 
@@
 
+\newpage
+
 \hypertarget{sec:comparison}{%

### 2022-11-16 draft/draft.Rmd
still need to finish conclusion

@@
 The contributions of this work are as follows. 
-In section \@ref(sec:pabm-grdpg), we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
+In section \@ref(sec:pabm-grdpg), we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific latent structure.
 More specifically, we show that the popularity adjusted block model is a generalized random dot product graph in which the latent vectors are organized by community and each community in the latent space is defined by a $K$-dimensional subspace, where $K$ is the number of communities in the model.
@@
 More formally, one assigns each vertex $v_i$ a community label $z_i \in \{1, ..., K\}$ and assumes a Bernoulli random graph in which $P_{ij} = g(z_i, z_j, \theta_i, \theta_j)$ for some function $g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]$ and $\theta_i \in \mathbb{R}^d$ is a vector of real-valued parameters associated with each vertex.
+While the graph is sampled conditioned on the labels $z_i$, it is assumed that the labels are unknown when the graph is observed. 
 Within the context of this work, we will call such models *block models*. 
 This type of model framework allows us to frame the goal of community detection as a statistical inference problem: 
-construct estimators for the true community labels, up to permutation.
+construct estimators for the unobserved true community labels, up to permutation.
 
@@
 Under this sampling scheme, there is no notion of increasing $n$.
+
 2. The vertices and edges are sampled in sequence, and we suppose that each vertex $v_i$ comes with a corresponding label, $z_i$, which determines its edge probabilities to the other vertices. 
 Under this sampling scheme, there is no assumed distribution on the labels, which are treated as fixed parameters rather than random variables, along with $\{\theta_{k \ell}\}_K$, but there is a notion of increasing sample size $n$, which allows for asymptotics.
+
 3. The vertices are sampled in sequence, and each label, $z_i$, is sampled as $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, $\sum_k^K \alpha_k = 1$. 
-The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}_K$.
-Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \Big[ \alpha_1, \cdots, \alpha_K \Big]^\top$ and $\{\theta_{k \ell}\}_K$.
+The edge probabilities are then determined by the sample $z = [z_1 \cdots z_n]^\top$ and the parameters $\{\theta_{k \ell}\}_K$.
+Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = [ \alpha_1, \cdots, \alpha_K ]^\top$ and $\{\theta_{k \ell}\}_K$.
 
@@
 An extension to these include Bayesian models in which there are prior distributions for the parameters, e.g., each $\theta_{k \ell}$ has a prior distribution with support $[0, 1]$, such as $\theta_{k \ell} \sim \Betadist(a_{k \ell}, b_{k \ell})$, and the mixture parameters $\alpha$ has a prior distribution on the $K-1$ simplex, such as $\alpha \sim \mathrm{Dirichlet}(\beta_1, ..., \beta_K)$.
+(See \citet{RePEc:eee:csdana:v:60:y:2013:i:c:p:12-31} and \citet{10.2307/43974715}.)
 
@@
 Modeling this as a DCBM allows us to take this into account by adjusting each user's expected degree by a factor of $\omega_i$. 
-The following is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n = 1$.
+Figure \ref{fig:dating-dcbm} is a visualization of the network from example \ref{ex:sbm-dating} modified as a DCBM with parameters $\theta_{11} = \theta_{22} = 0.03$, $\theta_{12} = 0.3$, each $\omega_i$ drawn uniformly between $1/3$ and $1$, and $\rho_n = 1$.
+The figure shows that some individuals (vertices) are a lot more popular than others in the form of having more matches with other individuals (edges).
 
@@
 Figure \ref{fig:dating-pabm} is a visualization of the online dating network modeled as a PABM in which the popularity parameters are drawn as $\lambda_{ik} \sim \Betadist(1, 6)$ if $z_i = k$ and $\lambda_{ik} \sim \Betadist(2, 2)$ if $z_i \neq k$.
-Like in the SBM and DCBM examples, the average edge probability between genders is approximately ten times that of the average edge probability within genders, but unlike in the SBM and DCBM examples, each individual has its own affinity toward each gender.
+Like in the SBM and DCBM examples, the average edge probability between genders is approximately ten times that of the average edge probability within genders, but unlike in the SBM and DCBM examples, we can see that some vertices have more connections to its own gender and some have more connections to the opposite gender.
 
@@
 The EM algorithms for the DCBM and PABM are similar to algorithm \ref{algo:sbm-em}, and the same independence relaxation is required. 
+Note that it is required for each $\theta_{k \ell} \omega_i \omega_j \in (0, 1)$ (for the DCBM) or $\lambda_{ik} \lambda_{j \ell} \in (0, 1)$ (for the PABM), and some attention is required to maintain this constraint when implementing these algorithms.
 
@@
 set.seed(123)
-out <- em.sbm(A, 
-              t(apply(matrix(runif(n * K), nrow = n), 1, function(x) x / sum(x))),
-              matrix(c(.9, .1, .1, .1, .9, .1, .1, .1, .9), nrow = K))
-out <- em.dcbm(A, 
-               out$pi,
-               out$theta,
+out.sbm <- em.sbm(A, 
+                  t(apply(matrix(runif(n * K), nrow = n), 
+                          1, 
+                          function(x) x / sum(x))),
+                  matrix(c(.9, .1, .1, .1, .9, .1, .1, .1, .9), 
+                         nrow = K))
+out.dcbm <- em.dcbm(A, 
+                    out.sbm$pi,
+                    out.sbm$theta,
                     runif(n))
 
+# zhat.dcbm <- factor(apply(out.dcbm$pi, 1, which.max))
+# zhat.dcbm <- model.matrix(~ zhat.dcbm)
+# zhat.dcbm[, 1] <- 1 - zhat.dcbm[, 2] * zhat.dcbm[, 3]
+# lambda.init <- matrix(runif(n * K), nrow = n, ncol = K)
+# lambda.init <- diag(out.dcbm$omega) %*% zhat.dcbm %*% out.dcbm$theta
+# lambda.init <- lambda.init / (max(lambda.init) + 1e-6)
+# out.pabm <- em.pabm(A, 
+#                     out.dcbm$pi,
+#                     lambda.init)
+# zhat.pabm <- apply(out.pabm$pi, 1, which.max)
+# table(z, zhat.pabm)
+
 ggplot() + 
-  geom_histogram(aes(x = out$omega),
+  geom_histogram(aes(x = out.dcbm$omega),
                  colour = 'black', fill = 'white') + 
@@
 Recall the UK faculty network in example \ref{ex:uk-faculty}. 
-In examples~\ref{ex:uk-faculty-sbm} and \ref{ex:uk-faculty-dcbm-pabm}, we fit an SBM, DCBM, and PABM to the graph using their corresponding EM algorithms. 
+In examples~\ref{ex:uk-faculty-sbm} and \ref{ex:uk-faculty-dcbm-pabm}, we fit an SBM and a DCBM to the graph using their corresponding EM algorithms (algorithms~\ref{algo:sbm-em} and \ref{algo:dcbm-em}).
 While all three algorithms were able to correctly identify the labels for most of the vertices, it is not immediately clear which model is the most appropriate for this graph. 
@@
 If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
-We assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
+Based on figure \ref{fig:friendship-network}, we assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
 Because the fourth community only contains two members, we omitted it from this analysis. 
 Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM, since each community appears to be described by a line segment. 
-$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, which is an improvement over all three block model EM algorithms.
+$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, which is an improvement over the EM algorithms while being substantially faster and easier to implement.
 
@@
 Then the ASE for this model is $\hat{X} = \Big[ |\lambda_1|^{1/2} v_1 \mid |\lambda_n|^{1/2} v_n \Big]$. 
-A plot of the ASE of the graph in \ref{ex:sbm-dating} (figure \ref{fig:dating-ase}, left) suggests two (somewhat overlapping) ellipsoidal clusters, and fitting a GMM on the embedding results in a community detection accuracy of 96.9\%, which is comparable to algorithm \ref{algo:sbm-em} while being much simpler to implement.
+A plot of the ASE of the graph in \ref{ex:sbm-dating} (figure \ref{fig:dating-ase}, left) suggests two (somewhat overlapping) ellipsoidal clusters, and fitting a GMM on the embedding results in a community detection accuracy of 96.9\%, which is comparable to algorithm \ref{algo:sbm-em} while being simpler to implement.
 Likewise, a plot of the ASE of the graph in example~\ref{ex:dcbm-dating} (figure \ref{fig:dating-ase}, right) consists of points that lie along two lines that cross at the origin, and kernel $K$-means clustering using the cosine similarity kernel results in a community detection accuracy of 93.8\%, which is an improvement over algorithm \ref{algo:dcbm-em} while being simpler to implement.
@@
 
-```{r dating-ase, cache = TRUE, fig.cap = 'ASEs of the graphs in examples \\ref{ex:sbm-dating}, which is a disassortative SBM (left), and \\ref{ex:dcbm-dating}, which is a disassortative DCBM (right).
+```{r dating-ase, cache = TRUE, fig.height = 2, fig.width = 5, fig.cap = 'ASEs of the graphs in examples \\ref{ex:sbm-dating}, which is a disassortative SBM (left), and \\ref{ex:dcbm-dating}, which is a disassortative DCBM (right).
 The colors correspond to vertex label.
@@
 In this section, we explore GRDPGs with community-wise parameterized nonlinear latent structures. 
-We begin with a motivating example.
+We begin with a motivating example (example \ref{ex:macaque}).
 
@@
 The embedding vectors, which approximate the latent vectors, appear to approximately lie along two curves that intersect at the origin, with each curve corresponding to a label (visual vs sensorimotor).
-While the ASE suggests a community-wise structure, the SBM, DCBM, and PABM are not suitable models for these data.
+While the ASE suggests a community-wise structure in the latent space, the SBM, DCBM, and PABM are not suitable models for these data.
 
@@
 
-One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that there are $K$ manifolds in the latent space on which the latent vectors lie. 
-Then each manifold can be thought of as a community. 
+One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that the latent vectors lie on one of $K$ manifolds, and the label of each latent vector's corresponding vertex is the label of its manifold.
 The goal of community detection in this setting is to partition the graph that results from this latent configuration such that each partition maps onto each manifold. 
-Before we dive into the multi-manifold case, we can first consider the single manifold case as a generalization of the $\ER$ model.
+Before diving into the multi-manifold case, we can first consider the single manifold case as a generalization of the $\ER$ model (example \ref{ex:er-curve}).
 
 \begin{example}
+\label{ex:er-curve}
 In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
@@
 Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the ASE of the graph.
-Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.
+Although there is no obvious community structure in the graph itself, the embedding shows a clear separation between the two communities. 
+More specifically, similar to the latent vectors, the embedding vectors lie approximately on one of two curve segments (one-dimensional manifolds), suggesting that the appropriate community detection approach in this setting is to determine to which (unobserved) curve to which each embedding vector is closest.
 
@@
 \citet{https://doi.org/10.48550/arxiv.1705.03297} used GMMs on the ASE for community detection and found that the curve-like structures in the ASE were incompatible with centroid-based clustering methods.
-\citet{SannaPassino2022} used Gaussian processes to estimate curves, which resulted in better empirical fits. 
+\citet{SannaPassino2022} used Gaussian processes to estimate curves in the embedding, which resulted in better empirical fits. 
 \citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on a single manifold rather than estimating the manifolds or identifying multiple manifolds.
-\citet{https://doi.org/10.48550/arxiv.2006.05168} and \citet{https://doi.org/10.48550/arxiv.2208.11665} noted that kernel GRDPGs (GRDPGs in which $P_{ij} = \kappa(x_i, x_j)$ for some kernel function $\kappa$) with community structure can have large dimensions and proposed using a lower dimension with manifold structure to approximate the kernel function.
+\citet{https://doi.org/10.48550/arxiv.2006.05168} and \citet{https://doi.org/10.48550/arxiv.2208.11665} noted that kernel GRDPGs (GRDPGs in which $P_{ij} = \kappa(x_i, x_j)$ for some kernel function $\kappa$) with community structure can have large dimensions and proposed using a lower dimension with manifold structure to approximate the nonlinearity of the kernel function.
 
 Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in $\mathbb{R}^d$ or identifying vectors belonging to different manifolds in $\mathbb{R}^d$. 
-Manifold learning itself is a large area of research with a long history \citep{Tenenbaum2000-ff, trosset2020rehabilitating, https://doi.org/10.48550/arxiv.2107.08089}, so we will focus primarily on the setting in which there are multiple manifolds and the goal is to identify which vector lies on which manifold (manifold clustering).
+Manifold learning itself is a large area of research with a long history \citep{Tenenbaum2000-ff, 10945-32767, trosset2020rehabilitating, https://doi.org/10.48550/arxiv.2107.08089}, so we will focus primarily on the setting in which there are multiple manifolds and the goal is to identify which vector lies on which manifold (manifold clustering).
 The standard approach to manifold clustering is essentially a type of spectral clustering, involving spectral decomposition of some kernel matrix applied to the vectors lying on manifolds (or its graph Laplacian matrix).
-Unlike standard manifold clustering techniques such as ratio cut, it has been observed that while the leading eigenvalues discern subspaces or approximations of subspaces, some of the other eigenvalues or a mixture of other eigenvalues help discern nonlinear structures \citep{10945-32767, https://doi.org/10.48550/arxiv.2208.11665, https://doi.org/10.48550/arxiv.2201.10000}.
+Unlike standard manifold clustering techniques such as ratio cut, it has been observed that while the leading eigenvalues discern subspaces or approximations of subspaces, some of the other eigenvalues or a mixture of other eigenvalues help discern nonlinear structures \citep{https://doi.org/10.48550/arxiv.2208.11665, https://doi.org/10.48550/arxiv.2201.10000}.
 Most of these methods also assume that the vectors lie exactly on the manifolds (i.e., the observed vector is $x_i \in \mathcal{M}_k$) instead of approximately on them (i.e., the observed vector is $x_i = y_i + \epsilon_i$ where $y_i \in \mathcal{M}_k$ and $\| \epsilon_i \|$ is small in some sense), which is the setting in which the clustering is done on the ASE.
@@
 
-For the sake of completeness, we begin with the case of GRDPGs in which communities are represented by nonintersecting manifolds. 
+We begin with the case of GRDPGs in which communities are represented by nonintersecting manifolds. 
 In this section, we consider the following setting: Suppose that each community $C_k$ is represented by a closed manifold $\mathcal{M}_k$ in the latent space of a RDPG or GRDPG. 
@@
 Furthermore, assume that $\delta > 0$, i.e., the manifolds do not intersect. 
+This model can be viewed as a generalization of the SBM, which, when viewed as a GRDPG, consists of a latent space configuration in which each community corresponds to a point mass in the latent space.
+In this model, we generalize the point mass to a manifold. 
+In the case of the DCBM in which each $\omega_i$ is bounded away from 0, i.e., $\omega_i \geq c > 0$, the nonintersecting MBM can be applied as well.
 
@@
 \end{theorem}
-
-If the manifolds are one-dimensional, then a more precise rate of convergence can be derived.
 
@@
 Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM. 
-We initialized $K$-curves clustering by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known *a priori*. 
+\citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example. 
+The curves were initialized by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known *a priori*. 
 The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.

### 2022-11-16 draft/draft.tex
still need to finish conclusion

@@
 The contributions of this work are as follows.
-In section \ref{sec:pabm-grdpg}, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
+In section \ref{sec:pabm-grdpg}, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific latent structure.
 More specifically, we show that the popularity adjusted block model is a generalized random dot product graph in which the latent vectors are organized by community and each community in the latent space is defined by a \(K\)-dimensional subspace, where \(K\) is the number of communities in the model.
@@
 More formally, one assigns each vertex \(v_i\) a community label \(z_i \in \{1, ..., K\}\) and assumes a Bernoulli random graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
+While the graph is sampled conditioned on the labels \(z_i\), it is assumed that the labels are unknown when the graph is observed.
 Within the context of this work, we will call such models \emph{block models}.
 This type of model framework allows us to frame the goal of community detection as a statistical inference problem:
-construct estimators for the true community labels, up to permutation.
+construct estimators for the unobserved true community labels, up to permutation.
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\tightlist
 \item
@@
   The vertices are sampled in sequence, and each label, \(z_i\), is sampled as \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), \(\sum_k^K \alpha_k = 1\).
-  The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}_K\).
-  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \Big[ \alpha_1, \cdots, \alpha_K \Big]^\top\) and \(\{\theta_{k \ell}\}_K\).
+  The edge probabilities are then determined by the sample \(z = [z_1 \cdots z_n]^\top\) and the parameters \(\{\theta_{k \ell}\}_K\).
+  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = [ \alpha_1, \cdots, \alpha_K ]^\top\) and \(\{\theta_{k \ell}\}_K\).
 \end{enumerate}
@@
 An extension to these include Bayesian models in which there are prior distributions for the parameters, e.g., each \(\theta_{k \ell}\) has a prior distribution with support \([0, 1]\), such as \(\theta_{k \ell} \sim \Betadist(a_{k \ell}, b_{k \ell})\), and the mixture parameters \(\alpha\) has a prior distribution on the \(K-1\) simplex, such as \(\alpha \sim \mathrm{Dirichlet}(\beta_1, ..., \beta_K)\).
+(See \citet{RePEc:eee:csdana:v:60:y:2013:i:c:p:12-31} and \citet{10.2307/43974715}.)
 
@@
 The EM algorithms for the DCBM and PABM are similar to algorithm \ref{algo:sbm-em}, and the same independence relaxation is required.
+Note that it is required for each \(\theta_{k \ell} \omega_i \omega_j \in (0, 1)\) (for the DCBM) or \(\lambda_{ik} \lambda_{j \ell} \in (0, 1)\) (for the PABM), and some attention is required to maintain this constraint when implementing these algorithms.
 
@@
 In this section, we explore GRDPGs with community-wise parameterized nonlinear latent structures.
-We begin with a motivating example.
+We begin with a motivating example (example \ref{ex:macaque}).
 
@@
 
-One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that there are \(K\) manifolds in the latent space on which the latent vectors lie.
-Then each manifold can be thought of as a community.
+One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that the latent vectors lie on one of \(K\) manifolds, and the label of each latent vector's corresponding vertex is the label of its manifold.
 The goal of community detection in this setting is to partition the graph that results from this latent configuration such that each partition maps onto each manifold.
-Before we dive into the multi-manifold case, we can first consider the single manifold case as a generalization of the \(\ER\) model.
+Before diving into the multi-manifold case, we can first consider the single manifold case as a generalization of the \(\ER\) model (example \ref{ex:er-curve}).
 
 \begin{example}
+\label{ex:er-curve}
 In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
@@
 \citet{https://doi.org/10.48550/arxiv.1705.03297} used GMMs on the ASE for community detection and found that the curve-like structures in the ASE were incompatible with centroid-based clustering methods.
-\citet{SannaPassino2022} used Gaussian processes to estimate curves, which resulted in better empirical fits.
+\citet{SannaPassino2022} used Gaussian processes to estimate curves in the embedding, which resulted in better empirical fits.
 \citet{athreya2020estimation} considered GRDPGs in which the vertices lie on curves or manifolds, but their focus was largely on estimating distributions on a single manifold rather than estimating the manifolds or identifying multiple manifolds.
-\citet{https://doi.org/10.48550/arxiv.2006.05168} and \citet{https://doi.org/10.48550/arxiv.2208.11665} noted that kernel GRDPGs (GRDPGs in which \(P_{ij} = \kappa(x_i, x_j)\) for some kernel function \(\kappa\)) with community structure can have large dimensions and proposed using a lower dimension with manifold structure to approximate the kernel function.
+\citet{https://doi.org/10.48550/arxiv.2006.05168} and \citet{https://doi.org/10.48550/arxiv.2208.11665} noted that kernel GRDPGs (GRDPGs in which \(P_{ij} = \kappa(x_i, x_j)\) for some kernel function \(\kappa\)) with community structure can have large dimensions and proposed using a lower dimension with manifold structure to approximate the nonlinearity of the kernel function.
 
 Outside of random graph inference, there is a larger body of work on fitting parametric curves to vectors in \(\mathbb{R}^d\) or identifying vectors belonging to different manifolds in \(\mathbb{R}^d\).
-Manifold learning itself is a large area of research with a long history \citep{Tenenbaum2000-ff, trosset2020rehabilitating, https://doi.org/10.48550/arxiv.2107.08089}, so we will focus primarily on the setting in which there are multiple manifolds and the goal is to identify which vector lies on which manifold (manifold clustering).
+Manifold learning itself is a large area of research with a long history \citep{Tenenbaum2000-ff, 10945-32767, trosset2020rehabilitating, https://doi.org/10.48550/arxiv.2107.08089}, so we will focus primarily on the setting in which there are multiple manifolds and the goal is to identify which vector lies on which manifold (manifold clustering).
 The standard approach to manifold clustering is essentially a type of spectral clustering, involving spectral decomposition of some kernel matrix applied to the vectors lying on manifolds (or its graph Laplacian matrix).
-Unlike standard manifold clustering techniques such as ratio cut, it has been observed that while the leading eigenvalues discern subspaces or approximations of subspaces, some of the other eigenvalues or a mixture of other eigenvalues help discern nonlinear structures \citep{10945-32767, https://doi.org/10.48550/arxiv.2208.11665, https://doi.org/10.48550/arxiv.2201.10000}.
+Unlike standard manifold clustering techniques such as ratio cut, it has been observed that while the leading eigenvalues discern subspaces or approximations of subspaces, some of the other eigenvalues or a mixture of other eigenvalues help discern nonlinear structures \citep{https://doi.org/10.48550/arxiv.2208.11665, https://doi.org/10.48550/arxiv.2201.10000}.
 Most of these methods also assume that the vectors lie exactly on the manifolds (i.e., the observed vector is \(x_i \in \mathcal{M}_k\)) instead of approximately on them (i.e., the observed vector is \(x_i = y_i + \epsilon_i\) where \(y_i \in \mathcal{M}_k\) and \(\| \epsilon_i \|\) is small in some sense), which is the setting in which the clustering is done on the ASE.
@@
 
-For the sake of completeness, we begin with the case of GRDPGs in which communities are represented by nonintersecting manifolds.
+We begin with the case of GRDPGs in which communities are represented by nonintersecting manifolds.
 In this section, we consider the following setting: Suppose that each community \(C_k\) is represented by a closed manifold \(\mathcal{M}_k\) in the latent space of a RDPG or GRDPG.
@@
 Furthermore, assume that \(\delta > 0\), i.e., the manifolds do not intersect.
+This model can be viewed as a generalization of the SBM, which, when viewed as a GRDPG, consists of a latent space configuration in which each community corresponds to a point mass in the latent space.
+In this model, we generalize the point mass to a manifold.
+In the case of the DCBM in which each \(\omega_i\) is bounded away from 0, i.e., \(\omega_i \geq c > 0\), the nonintersecting MBM can be applied as well.
 
@@
 Since the DCBM is a special case of the MBM, it should result in at least as good of a fit as the DCBM.
-We initialized \(K\)-curves clustering by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known \emph{a priori}.
+\citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example.
+The curves were initialized by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known \emph{a priori}.
 The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.

### 2022-11-16 em-algorithms.R
adding code for em algorithms

@@ -1,0 +1,314 @@
+em.sbm <- function(A, pi.start, theta.start, maxit = 100, eps = 1e-6) {
+  pi <- pi.start
+  theta <- theta.start
+  
+  n <- ncol(A)
+  K <- ncol(theta)
+  
+  iter <- 0
+  
+  while(TRUE) {
+    pi.old <- pi
+    theta.old <- theta
+    # e-step
+    pi <- update.pi.sbm(pi.old, A, theta)
+    
+    # m-step
+    theta <- update.theta.sbm(pi, A)
+    
+    if (Matrix::norm(pi.old - pi, 'F') < eps) break
+    
+    iter <- iter + 1
+    if (iter >= maxit) {
+      warning('failed to converge')
+      break
+    }
+  }
+  
+  return(list(pi = pi,
+              theta = theta, 
+              iter = iter))
+}
+
+update.pi.sbm.ik <- function(pi, A, theta, i, k) {
+  log.pi.ik <- 0
+  n <- ncol(A)
+  K <- ncol(theta)
+  for (j in seq(n)) {
+    for (l in seq(K)) {
+      if (j != i) {
+        log.pi.ik <- log.pi.ik + 
+          pi[j, l] * 
+          (A[i, j] * log(theta[k, l]) + 
+             (1 - A[i, j]) * log(1 - theta[k, l]))
+      }
+    }
+  }
+  return(log.pi.ik)
+}
+
+update.pi.sbm <- function(pi, A, theta) {
+  pi.new <- pi
+  n <- nrow(pi)
+  K <- ncol(pi)
+  for (i in seq(n)) {
+    for (k in seq(K)) {
+      pi.new[i, k] <- update.pi.sbm.ik(pi, A, theta, i, k)
+    }
+  }
+  pi.new <- pi.new - mean(pi.new)
+  pi.new <- exp(pi.new)
+  pi.new <- t(apply(pi.new, 1, function(x) x / sum(x)))
+  return(pi.new)
+}
+
+update.theta.sbm.kl <- function(pi, A, k, l) {
+  n <- nrow(pi)
+  numerator <- 0
+  denominator <- 0
+  for (i in seq(2, n)) {
+    for (j in seq(i - 1)) {
+      numerator <- numerator + A[i, j] * pi[i, k] * pi[j, l]
+      denominator <- denominator + pi[i, k] * pi[j, l]
+    }
+  }
+  return(numerator / denominator)
+}
+
[…]

### 2022-11-29 em-algorithms.R
fixed indexing n em algorithm for pabm

@@ -211,7 +211,8 @@ update.omega.dcbm.i <- function(pi, A, theta, omega, i) {
 }
 
 em.pabm <- function(A, pi.start, lambda.start, 
-                    maxit = 100, eps = 1e-9) {
+                    maxit = 100, eps = 1e-9,
+                    lambda.eps = 1e-9) {
   pi <- pi.start
   lambda <- lambda.start
   
@@ -224,11 +225,19 @@ em.pabm <- function(A, pi.start, lambda.start,
     pi.old <- pi
     lambda.old <- lambda
     
-    if (max(lambda) >= 1) {
-      lambda <- lambda / max(lambda + 1e-6)
+    if (min(lambda) <= lambda.eps) {
+      lambda <- lambda + lambda.eps
     }
+    if (max(lambda) >= 1 - lambda.eps) {
+      lambda <- lambda / max(lambda + lambda.eps)
+    }
+    
     pi <- update.pi.pabm(pi.old, A, lambda)
     lambda <- update.lambda.pabm(pi, A, lambda)
+    
+    print(Matrix::norm(pi.old - pi))
+    print(table(apply(pi, 1, which.max),
+                apply(pi.old, 1, which.max)))
     
     if (Matrix::norm(pi.old - pi, 'F') < eps) break
     
@@ -269,8 +278,8 @@ update.pi.pabm.ik <- function(pi, A, lambda, i, k) {
       if (j != i) {
         log.pi.ik <- log.pi.ik + 
           pi[j, l] * 
-          (A[i, j] * log(lambda[i, k] * lambda[j, l]) + 
-             (1 - A[i, j]) * log(1 - lambda[i, k] * lambda[j, l]))
+          (A[i, j] * log(lambda[i, l] * lambda[j, k]) + 
+             (1 - A[i, j]) * log(1 - lambda[i, l] * lambda[j, k]))
       }
     }
   }
@@ -283,15 +292,15 @@ update.lambda.pabm <- function(pi, A, lambda) {
   lambda.new <- matrix(NA, nrow = n, ncol = K)
   
   for (i in seq(n)) {
-    for (k in seq(K)) {
-      lambda.new[i, k] <- update.lambda.pabm.ik(pi, A, lambda, i, k)
+    for (l in seq(K)) {
+      lambda.new[i, l] <- update.lambda.pabm.il(pi, A, lambda, i, l)
     }
   }
   
   return(lambda.new)
 }
 
[…]

### 2022-11-29 em-algorithms.R
forgot to remove print statements in code

@@ -235,9 +235,9 @@ em.pabm <- function(A, pi.start, lambda.start,
     pi <- update.pi.pabm(pi.old, A, lambda)
     lambda <- update.lambda.pabm(pi, A, lambda)
     
-    print(Matrix::norm(pi.old - pi))
-    print(table(apply(pi, 1, which.max),
-                apply(pi.old, 1, which.max)))
+    # print(Matrix::norm(pi.old - pi))
+    # print(table(apply(pi, 1, which.max),
+    #             apply(pi.old, 1, which.max)))
     
     if (Matrix::norm(pi.old - pi, 'F') < eps) break
     

### 2022-11-30 draft/draft.Rmd
fixed typos in indexing, added more to the abstract

@@
       \For {$k = 1, ..., K$} {
-        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{ik} \hat{\lambda}_{j \ell}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{ik} \hat{\lambda}_{j \ell})) \bigg)$.\;
+        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{i \ell} \hat{\lambda}_{j k}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{i \ell} \hat{\lambda}_{j k})) \bigg)$.\;
       }
@@
     \For {$i = 1, ..., n$} {
-      \For {$k = 1, ..., K$} {
-        M-step: $\hat{\lambda}_{ik} = \frac{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j \ell}}$.\;
+      \For {$\ell = 1, ..., K$} {
+        M-step: $\hat{\lambda}_{i \ell} = \frac{\sum_{j \neq i} \sum_k \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_k \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j k}}$.\;
       }
@@
 Nevertheless, the naive EM algorithms appear to work well, at least for community detection (estimating the labels $z_i$, e.g., by setting $\hat{z}_i = \arg\max_k \pi_{ik}$), based on empirical results.
-Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 97\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
+Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 95\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
 \end{remark}
@@
 ```
-
-\newpage
 
@@
 In the first example, we applied $K$-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}. 
-As in \ref{ex:macaque}, we used $d = 2$ to construct the ASE, then used $K$-curves clustering to fit two curves to these data and predict the labels. 
+As in example \ref{ex:macaque}, we used $d = 2$ to construct the ASE, then used $K$-curves clustering to fit two curves to these data and predict the labels. 
 Figure \ref{fig:macaque-fitted} shows the result of $K$-curves clustering on this graph, 
@@
 
-In the second example, we considered the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}. 
+In the second example, we analyzed the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}. 
+Here, the vertices represent the various characters in the *Harry Potter* series, and edges represent whether pairs of characters are enemies. 
 This graph does not come with ground truth labels. 
@@
 \citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example. 
-The curves were initialized by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known *a priori*. 
+The curves were initialized by specifying that Lord Voldemort (the main antagonist) is in one community and Harry Potter (the titular protagonist) is in another, with the other vertices not known *a priori*.
 The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.
@@
 The code for simulations and data analyses in section \@ref(sec:pabm-grdpg) can be found at https://github.com/johneverettkoo/pabm-grdpg. 
-An R package implementing orthogonal spectral clustering (algorithm \ref(alg:osc)) is currently in progress and can be found at https://github.com/johneverettkoo/osc. 
+An R package implementing orthogonal spectral clustering (algorithm \ref{alg:osc}) is currently in progress and can be found at https://github.com/johneverettkoo/osc. 
 The code for simulations and data analyses in section \@ref(sec:mbm) can be found at https://github.com/johneverettkoo/manifold-block-models.

### 2022-11-30 draft/draft.tex
fixed typos in indexing, added more to the abstract

@@
 
-Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision.
+Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from analyzing data in the social sciences to training machine learning models for artificial intelligence tasks.
 One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels).
 Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs. 
-In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure), first by demonstrating how under the Generalized Random Dot Product Graph framework, all graphs generated by block models can be represented as feature vectors, then applying clustering methods for these feature vector representations, and finally deriving the asymptotic properties of these methods.
+In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure), first by demonstrating how under the generalized random dot product graph framework, all block models can be represented as a collection of feature vectors organized by community, applying clustering methods for these feature vector representations by exploiting the linear structures that the block models induce, and finally deriving the asymptotic properties of these methods.
+We further extend this connection between block models and community-organized generalized random dot product graphs to propose more flexible, nonlinear community structures, using real graphs with nonlinear structures as motivating examples.
 
@@
 \end{table}
-
-\newpage
 
@@
 In the first example, we applied \(K\)-curves clustering (algorithm \ref{alg:kcurves}) to the macaque visuotactile brain areas and connections network from example \ref{ex:macaque}.
-As in \ref{ex:macaque}, we used \(d = 2\) to construct the ASE, then used \(K\)-curves clustering to fit two curves to these data and predict the labels.
+As in example \ref{ex:macaque}, we used \(d = 2\) to construct the ASE, then used \(K\)-curves clustering to fit two curves to these data and predict the labels.
 Figure \ref{fig:macaque-fitted} shows the result of \(K\)-curves clustering on this graph,
@@
 
-In the second example, we considered the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}.
+In the second example, we analyzed the Harry Potter enmity graph \citep{harrypotter}, visualized in figure \ref{fig:hp}.
+Here, the vertices represent the various characters in the \emph{Harry Potter} series, and edges represent whether pairs of characters are enemies.
 This graph does not come with ground truth labels.
@@
 \citet{SannaPassino2022} fit quadratic curves on the embedding and suggested that there was some information to be gained by the nonlinearity, so we fit quadratic Bezier curves in this example.
-The curves were initialized by specifying that Lord Voldemort is in one community and Harry Potter is in another, with the other vertices not known \emph{a priori}.
+The curves were initialized by specifying that Lord Voldemort (the main antagonist) is in one community and Harry Potter (the titular protagonist) is in another, with the other vertices not known \emph{a priori}.
 The fitted quadratic curves are illustrated in figure \ref{fig:hp-kcurves}, which indeed appear to be very close to linear.
@@
 The code for simulations and data analyses in section \ref{sec:pabm-grdpg} can be found at \url{https://github.com/johneverettkoo/pabm-grdpg}.
-An R package implementing orthogonal spectral clustering (algorithm \ref(alg:osc)) is currently in progress and can be found at \url{https://github.com/johneverettkoo/osc}.
+An R package implementing orthogonal spectral clustering (algorithm \ref{alg:osc}) is currently in progress and can be found at \url{https://github.com/johneverettkoo/osc}.
 The code for simulations and data analyses in section \ref{sec:mbm} can be found at \url{https://github.com/johneverettkoo/manifold-block-models}.

### 2022-12-07 draft/draft.Rmd
added conclusion (still WIP)

@@
 Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form. 
+(The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.) 
 As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
@@
   labs(x = expression(x[2]), y = expression(x[1]))
+
 ```
@@
 # Comparisons of Models and Conclusions {#sec:comparison}
+
+```{r block-models, cache = TRUE, fig.width = 6, fig.height = 8, out.width = '50%', fig.cap = 'Graph visualizations (left) and ASEs (right) of the three main block models.
+Note the structures that the embedding vectors form for each model.'}
+set.seed(12345)
+
+n <- 64
+n1 <- n / 2
+n2 <- n / 2
+z.sbm <- c(rep(1, n1), rep(2, n2))
+p <- 1/2
+q <- 1/2
+r <- 1/8
+P <- matrix(r, nrow = n, ncol = n)
+P[seq(n1), seq(n1)] <- p
+P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
+X.sbm <- embedding(P, 2, 0)
+A.sbm <- draw.graph(P)
+Xhat.sbm <- embedding(A.sbm, 2, 0)
+Xhat.sbm[, 2] <- -Xhat.sbm[, 2]
+
+sbm.graph.plot <- ggraph(A.sbm, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z.sbm)),
+                  show.legend = FALSE) + 
+  labs(colour = NULL, 
+       title = 'SBM') + 
+  # theme_graph() + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+
+sbm.ase.plot <- ggplot() + 
+  geom_point(aes(x = Xhat.sbm[, 2], y = Xhat.sbm[, 1],
+                 colour = factor(z.sbm)),
+             shape = 4) + 
+  geom_point(aes(x = X.sbm[, 2], y = X.sbm[, 1], 
+                 colour = factor(z.sbm))) + 
+  # coord_fixed() +
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  labs(x = NULL, y = NULL, colour = NULL, 
+       title = 'Point Masses') + 
+  scale_colour_brewer(palette = 'Set1')
+
+n <- 128
+n1 <- n / 2
+n2 <- n / 2
+z.dcbm <- c(rep(1, n1), rep(2, n2))
+p <- 1/2
+q <- 1/2
+r <- 1/8
+P <- matrix(r, nrow = n, ncol = n)
+P[seq(n1), seq(n1)] <- p
+P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
+omega <- rbeta(n, 2, 1)
+P <- P * (omega %*% t(omega))
+X.dcbm <- embedding(P, 2, 0)
+A.dcbm <- draw.graph(P)
+Xhat.dcbm <- embedding(A.dcbm, 2, 0)
+Xhat.dcbm <- -Xhat.dcbm
+
+dcbm.graph.plot <- ggraph(A.dcbm, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z.dcbm)),
+                  show.legend = FALSE) + 
+  labs(colour = NULL, 
+       title = 'DCBM') + 
+  # theme_graph() + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+
+dcbm.ase.plot <- ggplot() + 
+  geom_point(aes(x = Xhat.dcbm[, 1], y = Xhat.dcbm[, 2],
+                 colour = factor(z.dcbm)),
+             shape = 4) + 
+  geom_point(aes(x = X.dcbm[, 1], y = X.dcbm[, 2], 
+                 colour = factor(z.dcbm))) + 
+  # coord_fixed() +
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  labs(x = NULL, y = NULL, colour = NULL, 
+       title = 'Rays') + 
+  scale_colour_brewer(palette = 'Set1')
+
+n <- 64
+Pz <- generate.P.beta(n)
+P <- Pz$P
+z.pabm <- Pz$clustering
+X.pabm <- embedding(P, 3, 1)
+A.pabm <- draw.graph(P)
+Xhat.pabm <- embedding(A.pabm, 3, 1)
+# Xhat.pabm[, 1] <- -Xhat.pabm[, 1]
+
+pabm.graph.plot <- ggraph(A.pabm, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z.pabm)),
+                  show.legend = FALSE) + 
+  labs(colour = NULL, 
+       title = 'PABM') + 
+  # theme_graph() + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+
+pabm.ase.plot <- ggplot() + 
+  geom_point(aes(x = Xhat.pabm[, 1], y = Xhat.pabm[, 2],
+                 colour = factor(z.pabm)),
+             shape = 4) + 
+  geom_point(aes(x = X.pabm[, 1], y = X.pabm[, 2], 
+                 colour = factor(z.pabm))) + 
+  # coord_fixed() +
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  labs(x = NULL, y = NULL, colour = NULL, 
+       title = 'Subspaces (Projected)') + 
+  scale_colour_brewer(palette = 'Set1')
+
+gridExtra::grid.arrange(sbm.graph.plot, sbm.ase.plot, 
+                        dcbm.graph.plot, dcbm.ase.plot,
+                        pabm.graph.plot, pabm.ase.plot,
+                        nrow = 3)
+```
+
+In this work, we showed the connection between block models and generalized random dot product graphs. 
+In particular, block models are a type of generalized random dot product graph in which the latent vectors are organized by community. 
+In the case of the stochastic block model, degree corrected block model, and popularity adjusted block model, the structures are linear. 
+This makes community detection via clustering in the latent space a straightforward task: Construct a graph embedding and apply a well-known clustering algorithm that is appropriate for the (linear) latent structure.
+For the SBM, the appropriate choice of clustering algorithm is $K$-means clustering or Gaussian mixture models. 
+For the DCBM, an appropriate choice is $K$-means clustering with the cosine kernel. 
+For the PABM, an appropriate choice is sparse subspace clustering, and we also developed orthogonal spectral clustering for this setting as well. 
+Furthermore, by looking at how the embedding vectors are organized, the embedding makes visibly clear which of the three block models is the best fit for the graph, and in the case in which the number of communities or the assortativity or disassortativity of the model is unknown, there are existing heuristics for determining these based on the size of the eigenvalues.
+This is in contrast to direct likelihood maximization (e.g., via expectation maximization), which can be more of a black box in determining the appropriate model, number of communities, graph structure, etc., and results in slower, less interpretable algorithms.
+In conclusion, the graph embedding approach, which relies on making the connection between the block models and the GRDPG, is a powerful tool in both exploratory data analysis and statistical inference on graphs with community structure.
+
+Moving beyond the linear-structure block models (SBM, DCBM, PABM), we showed that the GRDPG and ASE approach can recover nonlinear structures in the latent space. 
+This type of nonlinearity has been observed in various real networks, in particular brain connectivity networks. 
+While analyzing these graphs as GRDPGs reveals the (nonlinear) community structure, the linear block models are insufficient in modeling these graphs or performing statistical inference. 
+For this setting, we proposed the Manifold Block Model, which is a type of GRDPG in which the communities lie on nonlinear manifolds in the latent space.
 

### 2022-12-07 draft/draft.tex
added conclusion (still WIP)

@@
 \theoremstyle{remark}
-\newtheorem{remark}{Remark}
+\newtheorem*{remark}{Remark}
 \newtheorem*{solution}{Solution}
@@
 Nevertheless, making the conditional independence relaxation allows us to write the expectation and maximization steps in closed form.
+(The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.)
 As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
@@
 
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.5\linewidth]{draft_files/figure-latex/block-models-1}
+
+}
+
+\caption{Graph visualizations (left) and ASEs (right) of the three main block models. Note the structures that the embedding vectors form for each model.}\label{fig:block-models}
+\end{figure}
+
+In this work, we showed the connection between block models and generalized random dot product graphs.
+In particular, block models are a type of generalized random dot product graph in which the latent vectors are organized by community.
+In the case of the stochastic block model, degree corrected block model, and popularity adjusted block model, the structures are linear.
+This makes community detection via clustering in the latent space a straightforward task: Construct a graph embedding and apply a well-known clustering algorithm that is appropriate for the (linear) latent structure.
+For the SBM, the appropriate choice of clustering algorithm is \(K\)-means clustering or Gaussian mixture models.
+For the DCBM, an appropriate choice is \(K\)-means clustering with the cosine kernel.
+For the PABM, an appropriate choice is sparse subspace clustering, and we also developed orthogonal spectral clustering for this setting as well.
+Furthermore, by looking at how the embedding vectors are organized, the embedding makes visibly clear which of the three block models is the best fit for the graph, and in the case in which the number of communities or the assortativity or disassortativity of the model is unknown, there are existing heuristics for determining these based on the size of the eigenvalues.
+This is in contrast to direct likelihood maximization (e.g., via expectation maximization), which can be more of a black box in determining the appropriate model, number of communities, graph structure, etc., and results in slower, less interpretable algorithms.
+In conclusion, the graph embedding approach, which relies on making the connection between the block models and the GRDPG, is a powerful tool in both exploratory data analysis and statistical inference on graphs with community structure.
+
 \newpage

### 2022-12-14 draft/draft.Rmd
updated draft per meeting

@@
 
-One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that the latent vectors lie on one of $K$ manifolds, and the label of each latent vector's corresponding vertex is the label of its manifold.
+One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that the latent vectors lie on one of $K$ curves, and the label of each latent vector's corresponding vertex is the label of its curve.
+If we were to extend this to communities represented by higher dimensional structures, similar to the DCBM vs.
+PABM, a suitable modeling approach is to suppose that the latent vectors lie on one of $K$ manifolds, each of which represent one of the $K$ labels.
 The goal of community detection in this setting is to partition the graph that results from this latent configuration such that each partition maps onto each manifold. 
@@
 We saw that in the case of the DCBM and PABM, because the latent structures are not based on point masses, while the ASE does reveal the communities, the final clustering step requires other clustering algorithms that are tailored to the latent community structure.
-Likewise, in the following sections, we develop algorithms specific to the structure of the MBM, one in the case of disjoint manifolds (section \@ref(sec:nonintersect)) and one in the general case in which the latent manifolds can intersect (section \@ref(sec:intersect)).
+Likewise, in the following sections, we develop algorithms specific to the structure of the MBM, one in the case of disjoint manifolds and one in the general case in which the latent manifolds can intersect.
 We also derive theoretical consistency properties for both settings.
@@
 
-## Algorithm for Nonintersecting Manifolds {#sec:nonintersect}
+## Algorithms {#sec:nonintersect}
 
@@
 
-## Algorithm for Intersecting Manifolds {#sec:intersect}
-
-The consistency of the ASE provides asymptotic guarantees for perfect community label recovery for the nonintersecting MBM. 
-This relies on the fact that single-linkage clustering and $\eta$-neighborhood clustering graphs result in disconnected components if clusters of points are far away from each other. 
-However, many graphs do not behave so nicely. 
-For example, the DCBM and PABM can both be viewed as MBMs in which the latent vectors come from (linear) manifolds that intersect at the origin (see section \@ref(sec:sbm-dcbm-grdpg)). 
-Nonlinear intersecting manifolds have also been observed in ASEs of real graphs \citep{SannaPassino2022, https://doi.org/10.48550/arxiv.1709.05454}. 
-While the DCBM and PABM cases can be handled by algorithms specific to linear structures such as SSC and OSC, nonlinear intersecting manifolds require a more flexible approach. 
-For the intersecting case, we restrict the MBM to the case in which each $F_{z_i}$ is identical and univariate on the unit interval $[0, 1]$. 
+Algorithm \ref{alg:ase-singlelink} depends on the fact that the minimum distances between manifolds are nonzero and that the ASE is consistent. 
+Single linkage clustering assumes that the asymptotic result is true to some extent, i.e., the ASE error is small enough that the gap between embedding vectors from different manifolds is still nonzero.
+When the ASE error is insufficiently small, single linkage clustering on the ASE will not produce the correct result. 
+An alternative algorithm that may work when this condition is not fulfilled is a parametric approach in which we model the manifolds directly. 
+For ease of modeling, we restrict the MBM to the case in which each $F_{z_i}$ is identical and univariate on the unit interval $[0, 1]$. 
 The full mixture model is described as follows:
@@
 4. Draw $A \sim \GRDPG_{p,q}(X; \rho_n)$.
-
-\begin{example}
-\label{ex:intersect-curves}
-Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$.
-This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
-For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
-$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
-Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
-The community structure is also not clear from the adjacency matrix. 
-However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).
-
-```{r intersect-curves-example, fig.height = 2, fig.width = 8, fig.cap = 'Latent vectors on intersecting curves (left), along with an RDPG drawn from this configuration (center) and its ASE (right).', cache = TRUE}
-f1 <- function(t) {
-  x1 <- t ^ 2
-  x2 <- 2 * t - 2 * t ^ 2
-  return(cbind(x1, x2))
-}
-
-f2 <- function(t) {
-  x1 <- 2 * t - 2 * t ^ 2
-  x2 <- 1 - 2 * t + t ^ 2
-  return(cbind(x1, x2))
-}
-
-set.seed(314159)
-
-n1 <- 2 ** 7
-n2 <- n1
-n <- n1 + n2
-z <- c(rep(1, n1), rep(2, n2))
-
-a <- 1
-b <- 1
-t1 <- rbeta(n1, a, b)
-t2 <- rbeta(n2, a, b)
-T <- rbind(t1, t2)
-
-X1 <- f1(t1)
-X2 <- f2(t2)
-X <- rbind(X1, X2)
-
-latent.vecs <- ggplot() + 
-  geom_point(aes(x = X[, 1], y = X[, 2], colour = factor(z)), size = .5) + 
-  coord_fixed() + 
-  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
-  theme_bw() + 
-  theme(legend.position = 'none') + 
-  scale_colour_brewer(palette = 'Set1')
-
-P <- X %*% t(X)
-diag(P) <- 0
-A <- draw.graph(P)
-
-rdpg <- ggraph(A, layout = 'stress') + 
-  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
-  geom_node_point(aes(colour = factor(z)), 
-                  show.legend = FALSE) + 
-  labs(colour = NULL) + 
-  theme_void() + 
-  scale_colour_brewer(palette = 'Set1')
-
-Xhat <- embedding(A, 2, 0)
-
-ase <- ggplot() + 
-  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z)), size = .5) + 
-  coord_fixed() + 
-  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
-  theme_bw() + 
-  theme(legend.position = 'none') + 
-  scale_colour_brewer(palette = 'Set1')
-
-gridExtra::grid.arrange(latent.vecs, rdpg, ase, nrow = 1)
-```
-\end{example}
 
@@
 
+Similar to $K$-means clustering or the EM algorithm, $K$-curves clustering is sensitive to initialization. 
+One plausible initialization setting is the semi-supervised case, in which a few vertices from each community are labeled. 
+Such a setting allows for theoretical guarantees (theorem \ref{thm:kcurves-semisupervised}) and appears to work well in numerical examples.
+
 $K$-curves clustering either requires knowledge of the functional form of each $g_k$ or a function that is flexible enough to approximate it. 
@@
 Then we have $g_k(t) = g(t; p_k) = \sum_{r=0}^R p_k^{(r)} \binom{R}{r} (1-t)^{R-r} t^r$. 
+However, as we see in figure \ref{fig:sim-nonintersect}, an incorrect yet sufficiently flexible parametric curve can provide good results.
+
+\begin{example}
+\label{ex:sim-nonintersect}
+
+Consider the setting of example \ref{ex:separated-curves-2} but for various $n$, in this case, $n = 16, 32, 64, 128, 256, 512, 1024$. 
+For each $n$, we sampled 50 RDPGs with this latent configuration, and for each graph, we applied algorithm \ref{alg:ase-singlelink} as well as algorithm \ref{alg:kcurves} both with random initialization and with four known labels per community.
+Figure \ref{sim-nonintersect} shows that for sufficiently large graphs, single linkage clustering on the ASE is able to consistently produce perfect clustering, but $K$-curves clustering performs better for very small graphs.
+
+```{r sim-nonintersect, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates in the simulation setting for $K = 2$ curves.', cache = TRUE}
+n.vec <- 2 ^ c(4, 5, 6, 7, 8, 9, 10)
+
+clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/nonintersect/nonintersect.csv')
+
+clust.summary.df <- clustering.df %>% 
+  dplyr::filter(n %in% n.vec) %>% 
+  dplyr::filter(nsamp != '8') %>% 
+  dplyr::group_by(n, nsamp) %>% 
+  dplyr::summarise(
+    med.rate = median(error),
+    first.q.rate = quantile(error, .25),
+    third.q.rate = quantile(error, .75)
+  ) %>% 
+  dplyr::ungroup()
+
+ggplot(clust.summary.df) + 
+  theme_bw() + 
+  theme(text = element_text(size = 10)) + 
+  scale_y_log10() +
+  scale_x_log10(breaks = n.vec) +
+  labs(y = 'community detection error rate',
+       x = 'n',
+       colour = 'known labels\nper community', 
+       shape = NULL) + 
+  geom_line(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
+  geom_point(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
+  geom_errorbar(aes(x = n, 
+                    ymin = first.q.rate, ymax = third.q.rate, 
+                    colour = factor(nsamp)),
+                width = .125) + 
+  scale_colour_brewer(palette = 'Set1')
+```
+\end{example}
+
+Algorithm \ref{alg:ase-singlelink} also relies on the fact that the distances between manifolds is nonzero, i.e., they do not intersect. 
+However, many graphs do not behave so nicely. 
+For example, the DCBM and PABM can both be viewed as MBMs in which the latent vectors come from (linear) manifolds that intersect at the origin (see section \@ref(sec:sbm-dcbm-grdpg)). 
+Nonlinear intersecting manifolds have also been observed in ASEs of real graphs \citep{SannaPassino2022, https://doi.org/10.48550/arxiv.1709.05454}. 
+Example \ref{ex:macaque} also suggests that the MBM with intersecting curves is an appropriate approach to modeling the graph. 
+While the DCBM and PABM cases can be handled by algorithms specific to linear structures such as SSC and OSC, nonlinear intersecting manifolds require a more flexible approach.
+
+\begin{example}
+\label{ex:intersect-curves}
+Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$.
+This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
+For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
+$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
+Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
+The community structure is also not clear from the adjacency matrix. 
+However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).
+
+```{r intersect-curves-example, fig.height = 2, fig.width = 8, fig.cap = 'Latent vectors on intersecting curves (left), along with an RDPG drawn from this configuration (center) and its ASE (right).', cache = TRUE}
+f1 <- function(t) {
+  x1 <- t ^ 2
+  x2 <- 2 * t - 2 * t ^ 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- 2 * t - 2 * t ^ 2
+  x2 <- 1 - 2 * t + t ^ 2
+  return(cbind(x1, x2))
+}
+
+set.seed(314159)
+
+n1 <- 2 ** 7
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+T <- rbind(t1, t2)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+
+latent.vecs <- ggplot() + 
+  geom_point(aes(x = X[, 1], y = X[, 2], colour = factor(z)), size = .5) + 
+  coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  scale_colour_brewer(palette = 'Set1')
+
+P <- X %*% t(X)
+diag(P) <- 0
+A <- draw.graph(P)
+
+rdpg <- ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+
+Xhat <- embedding(A, 2, 0)
+
+ase <- ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z)), size = .5) + 
+  coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  scale_colour_brewer(palette = 'Set1')
+
+gridExtra::grid.arrange(latent.vecs, rdpg, ase, nrow = 1)
+```
+\end{example}
 
@@
 
-# Comparisons of Models and Conclusions {#sec:comparison}
+# Discussion {#sec:comparison}
 
 ```{r block-models, cache = TRUE, fig.width = 6, fig.height = 8, out.width = '50%', fig.cap = 'Graph visualizations (left) and ASEs (right) of the three main block models.
-Note the structures that the embedding vectors form for each model.'}
+The dots are the original Note the structures that the embedding vectors form for each model.'}
 set.seed(12345)
@@
 
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-graph-1}
+
+}
+
+\caption{Macaque visuotactile brain areas and connections network. The vertices represent brain areas.}\label{fig:macaque-graph-conclusion}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-ase-1}
+
+}
+
+\caption{ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.}\label{fig:macaque-ase}
+\end{figure}
+
 Moving beyond the linear-structure block models (SBM, DCBM, PABM), we showed that the GRDPG and ASE approach can recover nonlinear structures in the latent space. 
@@
 For this setting, we proposed the Manifold Block Model, which is a type of GRDPG in which the communities lie on nonlinear manifolds in the latent space. 
+We then proposed two algorithms, one in which the manifolds are disjoint and another in which the manifolds intersect
 

### 2022-12-14 draft/draft.tex
fixed issue where remare remarks are not numbered
updated draft per meeting

@@
 
-One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that the latent vectors lie on one of \(K\) manifolds, and the label of each latent vector's corresponding vertex is the label of its manifold.
+One way to model the data in example \ref{ex:macaque} as a GRDPG is to suppose that the latent vectors lie on one of \(K\) curves, and the label of each latent vector's corresponding vertex is the label of its curve.
+If we were to extend this to communities represented by higher dimensional structures, similar to the DCBM vs.~PABM, a suitable modeling approach is to suppose that the latent vectors lie on one of \(K\) manifolds, each of which represent one of the \(K\) labels.
 The goal of community detection in this setting is to partition the graph that results from this latent configuration such that each partition maps onto each manifold.
@@
 We saw that in the case of the DCBM and PABM, because the latent structures are not based on point masses, while the ASE does reveal the communities, the final clustering step requires other clustering algorithms that are tailored to the latent community structure.
-Likewise, in the following sections, we develop algorithms specific to the structure of the MBM, one in the case of disjoint manifolds (section \ref{sec:nonintersect}) and one in the general case in which the latent manifolds can intersect (section \ref{sec:intersect}).
+Likewise, in the following sections, we develop algorithms specific to the structure of the MBM, one in the case of disjoint manifolds and one in the general case in which the latent manifolds can intersect.
 We also derive theoretical consistency properties for both settings.
@@
 \hypertarget{sec:nonintersect}{%
-\subsection{Algorithm for Nonintersecting Manifolds}\label{sec:nonintersect}}
+\subsection{Algorithms}\label{sec:nonintersect}}
 
@@
 
-\hypertarget{sec:intersect}{%
-\subsection{Algorithm for Intersecting Manifolds}\label{sec:intersect}}
-
-The consistency of the ASE provides asymptotic guarantees for perfect community label recovery for the nonintersecting MBM.
-This relies on the fact that single-linkage clustering and \(\eta\)-neighborhood clustering graphs result in disconnected components if clusters of points are far away from each other.
-However, many graphs do not behave so nicely.
-For example, the DCBM and PABM can both be viewed as MBMs in which the latent vectors come from (linear) manifolds that intersect at the origin (see section \ref{sec:sbm-dcbm-grdpg}).
-Nonlinear intersecting manifolds have also been observed in ASEs of real graphs \citep{SannaPassino2022, https://doi.org/10.48550/arxiv.1709.05454}.
-While the DCBM and PABM cases can be handled by algorithms specific to linear structures such as SSC and OSC, nonlinear intersecting manifolds require a more flexible approach.
-For the intersecting case, we restrict the MBM to the case in which each \(F_{z_i}\) is identical and univariate on the unit interval \([0, 1]\).
+Algorithm \ref{alg:ase-singlelink} depends on the fact that the minimum distances between manifolds are nonzero and that the ASE is consistent.
+Single linkage clustering assumes that the asymptotic result is true to some extent, i.e., the ASE error is small enough that the gap between embedding vectors from different manifolds is still nonzero.
+When the ASE error is insufficiently small, single linkage clustering on the ASE will not produce the correct result.
+An alternative algorithm that may work when this condition is not fulfilled is a parametric approach in which we model the manifolds directly.
+For ease of modeling, we restrict the MBM to the case in which each \(F_{z_i}\) is identical and univariate on the unit interval \([0, 1]\).
 The full mixture model is described as follows:
@@
 \end{enumerate}
-
-\begin{example}
-\label{ex:intersect-curves}
-Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$.
-This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
-For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
-$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
-Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
-The community structure is also not clear from the adjacency matrix. 
-However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).
-
-\begin{figure}[H]
-
-{\centering \includegraphics{draft_files/figure-latex/intersect-curves-example-1}
-
-}
-
-\caption{Latent vectors on intersecting curves (left), along with an RDPG drawn from this configuration (center) and its ASE (right).}\label{fig:intersect-curves-example}
-\end{figure}
-\end{example}
 
@@
 
+Similar to \(K\)-means clustering or the EM algorithm, \(K\)-curves clustering is sensitive to initialization.
+One plausible initialization setting is the semi-supervised case, in which a few vertices from each community are labeled.
+Such a setting allows for theoretical guarantees (theorem \ref{thm:kcurves-semisupervised}) and appears to work well in numerical examples.
+
 \(K\)-curves clustering either requires knowledge of the functional form of each \(g_k\) or a function that is flexible enough to approximate it.
@@
 Then we have \(g_k(t) = g(t; p_k) = \sum_{r=0}^R p_k^{(r)} \binom{R}{r} (1-t)^{R-r} t^r\).
+However, as we see in figure \ref{fig:sim-nonintersect}, an incorrect yet sufficiently flexible parametric curve can provide good results.
+
+\begin{example}
+\label{ex:sim-nonintersect}
+
+Consider the setting of example \ref{ex:separated-curves-2} but for various $n$, in this case, $n = 16, 32, 64, 128, 256, 512, 1024$. 
+For each $n$, we sampled 50 RDPGs with this latent configuration, and for each graph, we applied algorithm \ref{alg:ase-singlelink} as well as algorithm \ref{alg:kcurves} both with random initialization and with four known labels per community.
+Figure \ref{sim-nonintersect} shows that for sufficiently large graphs, single linkage clustering on the ASE is able to consistently produce perfect clustering, but $K$-curves clustering performs better for very small graphs.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/sim-nonintersect-1}
+
+}
+
+\caption{Median and IQR error rates over 50 replicates in the simulation setting for $K = 2$ curves.}\label{fig:sim-nonintersect}
+\end{figure}
+\end{example}
+
+Algorithm \ref{alg:ase-singlelink} also relies on the fact that the distances between manifolds is nonzero, i.e., they do not intersect.
+However, many graphs do not behave so nicely.
+For example, the DCBM and PABM can both be viewed as MBMs in which the latent vectors come from (linear) manifolds that intersect at the origin (see section \ref{sec:sbm-dcbm-grdpg}).
+Nonlinear intersecting manifolds have also been observed in ASEs of real graphs \citep{SannaPassino2022, https://doi.org/10.48550/arxiv.1709.05454}.
+Example \ref{ex:macaque} also suggests that the MBM with intersecting curves is an appropriate approach to modeling the graph.
+While the DCBM and PABM cases can be handled by algorithms specific to linear structures such as SSC and OSC, nonlinear intersecting manifolds require a more flexible approach.
+
+\begin{example}
+\label{ex:intersect-curves}
+Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$.
+This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
+For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
+$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
+Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
+The community structure is also not clear from the adjacency matrix. 
+However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/intersect-curves-example-1}
+
+}
+
+\caption{Latent vectors on intersecting curves (left), along with an RDPG drawn from this configuration (center) and its ASE (right).}\label{fig:intersect-curves-example}
+\end{figure}
+\end{example}
 
@@
 \hypertarget{sec:comparison}{%
-\section{Comparisons of Models and Conclusions}\label{sec:comparison}}
+\section{Discussion}\label{sec:comparison}}
 
@@
 
-\caption{Graph visualizations (left) and ASEs (right) of the three main block models. Note the structures that the embedding vectors form for each model.}\label{fig:block-models}
+\caption{Graph visualizations (left) and ASEs (right) of the three main block models.
+The dots are the original Note the structures that the embedding vectors form for each model.}\label{fig:block-models}
 \end{figure}
@@
 In conclusion, the graph embedding approach, which relies on making the connection between the block models and the GRDPG, is a powerful tool in both exploratory data analysis and statistical inference on graphs with community structure.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-graph-1}
+
+}
+
+\caption{Macaque visuotactile brain areas and connections network. The vertices represent brain areas.}\label{fig:macaque-graph-conclusion}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/macaque-ase-1}
+
+}
+
+\caption{ASE on the Macaque brain areas network. The embedding suggests a community-wise manifold structure.}\label{fig:macaque-ase}
+\end{figure}
+
+Moving beyond the linear-structure block models (SBM, DCBM, PABM), we showed that the GRDPG and ASE approach can recover nonlinear structures in the latent space.
+This type of nonlinearity has been observed in various real networks, in particular brain connectivity networks.
+While analyzing these graphs as GRDPGs reveals the (nonlinear) community structure, the linear block models are insufficient in modeling these graphs or performing statistical inference.
+For this setting, we proposed the Manifold Block Model, which is a type of GRDPG in which the communities lie on nonlinear manifolds in the latent space.
+We then proposed two algorithms, one in which the manifolds are disjoint and another in which the manifolds intersect
 

### 2022-12-14 draft/draft.Rmd
redid one example with more interesting cases

@@
 $g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
-Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
+Clearly, single-linkage or $\eta$-neighborhood clustering does not work in this setting, both in the original latent space and in the ASE. 
 The community structure is also not clear from the adjacency matrix. 
@@
 ```
-\end{example}
-
-\begin{example}
-\label{ex:intersect-curves-fit}
+
 Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
 In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \Big[ 0, 0 \Big]^\top$, i.e., the curves meet at the origin. 
-Here, an accuracy of 87.5\% is achieved.
+This clustering achieves 87.5\% accuracy.
 
@@
 
-While example \ref{ex:intersect-curves-fit} achieved high accuracy, algorithm \ref{alg:kcurves} often results in poor performance when it is paired with poor initialization. 
-In example \ref{ex:intersect-curves-fit}, it just so happened that we randomly chose initial curves that iterated to good fits, but this is not guaranteed. 
-We also restricted the curves to meet at the origin, which appears to have helped guide the curves to good fits. 
-In figures \ref{fig:loss} and \ref{fig:init-strategies}, we show the results of various initializations on fitting curves to example \ref{ex:intersect-curves}, such as random initialization and spectral clustering.
-
-\begin{figure}[H]
-
-{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-7-1}
-
-}
-
-\caption{ASE labeled by estimated community labels for each initialization strategy.
-Upper right: random initialization assuming intersection at the origin.
-Upper middle: random initialization without assuming intersection at the origin.
-Upper right: initialization via spectral clustering assuming intersection at the origin.
-Lower left: initialization via spectral clustering without assuming intersection at the origin.
-Lower middle: initialization with ground truth labels, assuming intersection at the origin.
-Lower right: initialization with ground truth labels, without assuming intersection at the origin.}\label{fig:init-strategies}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-6-1}
-
-}
-
-\caption{Clustering loss vs. iteration for each run of $K$-curve clustering.}\label{fig:loss}
-\end{figure}
-
-Alternatively, if we turn this problem into a semi-supervised problem in which we know a few of the community labels, then the convergence of the ASE guarantees high accuracy.
+While the result of algorithm \ref{alg:kcurves} in example \ref{ex:intersect-curves} achieved high accuracy, this is not always guaranteed even for properly parameterized curves, especially when paired with poor initialization.
+Performance as a measure of clustering accuracy is also problematic when there are multiple plausible curve fits. 
+Figures \ref{fig:multifit} and \ref{fig:loss} show the results of various curve fits and initialization strategies to example \ref{ex:intersect-curves}. 
+While only one of them (quadratic fit with random initialization) results in high clustering accuracy, cubic fit with spectral clustering initialization achieves a lower loss value.
+
+```{r multifit, cache = TRUE, fig.height = 3, fig.width = 8, fig.cap = '$K$-curve clustering fits on the ASE using various approaches.
+Left: quadratic fit with random initialization.
+Middle: quadratic fit initialized via spectral clustering.
+Right: cubic fit initialized via spectral clustering.'}
+f1 <- function(t) {
+  x1 <- t ^ 2
+  x2 <- 2 * t - 2 * t ^ 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- 2 * t - 2 * t ^ 2
+  x2 <- 1 - 2 * t + t ^ 2
+  return(cbind(x1, x2))
+}
+
+set.seed(314159)
+
+n1 <- 2 ** 7
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+T <- rbind(t1, t2)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+
+P <- X %*% t(X)
+diag(P) <- 0
+A <- draw.graph(P)
+
+Xhat <- embedding(A, 2, 0)
+
+maxit <- 2 ** 4
+animation.dir <- '.'
+random.starts <- 2
+eps <- 1e-6
+normalize <- TRUE
+animate <- FALSE
+
+set.seed(1)
+random <- manifold.clustering(Xhat, 2, 
+                              A = A, 
+                              parallel = TRUE,
+                              intercept = FALSE,
+                              maxit = maxit, 
+                              normalize = normalize,
+                              eps = eps,
+                              verbose = FALSE,
+                              animate = animate,
+                              animation.dir = animation.dir,
+                              animation.title = 'random')
+random.plot <- plot.estimated.curves(Xhat, random) + 
+  labs(x = NULL, y = NULL, title = 'random initialization') + 
+  scale_colour_brewer(palette = 'Set1')
+
+set.seed(2)
+spectral <- manifold.clustering(
+  Xhat, 2, 
+  A = A, 
+  parallel = TRUE, 
+  intercept = FALSE, 
+  normalize = normalize, 
+  maxit = maxit, 
+  eps = eps,
+  initialization = 'spectral',
+  verbose = FALSE,
+  animate = animate,
+  animation.dir = animation.dir,
+  animation.title = 'spectral')
+spectral.plot <- plot.estimated.curves(Xhat, spectral) + 
+  labs(x = NULL, y = NULL, title = 'spectral initialization') + 
+  scale_colour_brewer(palette = 'Set1')
+
+set.seed(3)
+cubic <- manifold.clustering(
+  Xhat, 2, 
+  degree = 3, 
+  A = A, 
+  parallel = TRUE, 
+  intercept = FALSE, 
+  normalize = normalize, 
+  maxit = maxit, 
+  eps = eps,
+  initialization = 'spectral',
+  verbose = FALSE,
+  animate = animate,
+  animation.dir = animation.dir,
+  animation.title = 'cubic')
+cubic.plot <- plot.estimated.curves(Xhat, cubic) + 
+  labs(x = NULL, y = NULL, title = 'cubic fit') + 
+  scale_colour_brewer(palette = 'Set1')
+
+gridExtra::grid.arrange(random.plot, spectral.plot, cubic.plot, nrow = 1)
+```
+
+```{r loss, cache = TRUE, fig.cap = 'Clustering loss vs. iteration for each run of $K$-curve clustering.', fig.width = 8, fig.height = 2}
+loss.df <- dplyr::bind_rows(
+  dplyr::tibble(loss = random$loss, 
+                initialization = 'random, quadratic',
+                iteration = seq_along(random$loss)),
+  dplyr::tibble(loss = spectral$loss, 
+                initialization = 'spectral, quadratic',
+                iteration = seq_along(spectral$loss)),
+  dplyr::tibble(loss = cubic$loss, 
+                initialization = 'spectral, cubic',
+                iteration = seq_along(cubic$loss))
+) %>% 
+  dplyr::mutate(initialization = factor(initialization,
+                                        levels = c('random, quadratic',
+                                                   'spectral, quadratic',
+                                                   'spectral, cubic')))
+
+ggplot(loss.df) + 
+  geom_point(aes(x = iteration, y = loss, colour = initialization)) + 
+  geom_line(aes(x = iteration, y = loss, colour = initialization)) + 
+  # ylim(0, max.loss) +
+  scale_y_log10() +
+  labs(y = 'clustering loss') + 
+  theme_bw()
+```
 
@@
 We performed two simulation experiments for algorithm \ref{alg:kcurves}.
-In the first simulation setup, we again turn to examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}. 
+In the first simulation setup, we again turn to example \ref{ex:intersect-curves}. 
 The setup is as follows:
@@
 
-In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves. 
+In the second simulation, we extended the first simulation setup based on example \ref{ex:intersect-curves} into three dimensions and three curves. 
 The mixture model is as follows (also see figure \ref{fig:three-curves}):

### 2022-12-14 draft/draft.tex
redid one example with more interesting cases

@@
 \end{figure}
-\end{example}
-
-\begin{example}
-\label{ex:intersect-curves-fit}
+
 Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
 In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \Big[ 0, 0 \Big]^\top$, i.e., the curves meet at the origin. 
-Here, an accuracy of 87.5\% is achieved.
+This clustering achieves 87.5\% accuracy.
 
@@
 
-While example \ref{ex:intersect-curves-fit} achieved high accuracy, algorithm \ref{alg:kcurves} often results in poor performance when it is paired with poor initialization.
-In example \ref{ex:intersect-curves-fit}, it just so happened that we randomly chose initial curves that iterated to good fits, but this is not guaranteed.
-We also restricted the curves to meet at the origin, which appears to have helped guide the curves to good fits.
-In figures \ref{fig:loss} and \ref{fig:init-strategies}, we show the results of various initializations on fitting curves to example \ref{ex:intersect-curves}, such as random initialization and spectral clustering.
-
-\begin{figure}[H]
-
-{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-7-1}
-
-}
-
-\caption{ASE labeled by estimated community labels for each initialization strategy.
-Upper right: random initialization assuming intersection at the origin.
-Upper middle: random initialization without assuming intersection at the origin.
-Upper right: initialization via spectral clustering assuming intersection at the origin.
-Lower left: initialization via spectral clustering without assuming intersection at the origin.
-Lower middle: initialization with ground truth labels, assuming intersection at the origin.
-Lower right: initialization with ground truth labels, without assuming intersection at the origin.}\label{fig:init-strategies}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics{/home/johnkoo/dev/manifold-block-models/draft/draft_files/figure-latex/unnamed-chunk-6-1}
+While the result of algorithm \ref{alg:kcurves} in example \ref{ex:intersect-curves} achieved high accuracy, this is not always guaranteed even for properly parameterized curves, especially when paired with poor initialization.
+Performance as a measure of clustering accuracy is also problematic when there are multiple plausible curve fits.
+Figures \ref{fig:multifit} and \ref{fig:loss} show the results of various curve fits and initialization strategies to example \ref{ex:intersect-curves}.
+While only one of them (quadratic fit with random initialization) results in high clustering accuracy, cubic fit with spectral clustering initialization achieves a lower loss value.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/multifit-1}
+
+}
+
+\caption{$K$-curve clustering fits on the ASE using various approaches.
+Left: quadratic fit with random initialization.
+Middle: quadratic fit initialized via spectral clustering.
+Right: cubic fit initialized via spectral clustering.}\label{fig:multifit}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{draft_files/figure-latex/loss-1}
 
@@
 \end{figure}
-
-Alternatively, if we turn this problem into a semi-supervised problem in which we know a few of the community labels, then the convergence of the ASE guarantees high accuracy.
 
@@
 We performed two simulation experiments for algorithm \ref{alg:kcurves}.
-In the first simulation setup, we again turn to examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}.
+In the first simulation setup, we again turn to example \ref{ex:intersect-curves}.
 The setup is as follows:
@@
 
-In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves.
+In the second simulation, we extended the first simulation setup based on example \ref{ex:intersect-curves} into three dimensions and three curves.
 The mixture model is as follows (also see figure \ref{fig:three-curves}):

### 2022-12-24 draft/draft.Rmd
fixed a bunch of typos

@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model. 
-Data are often represented as an $n \times p$ matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements.
-An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$.
-For instance, $\phi(x)$ might first compute the distances from $x$ to $K$ points in $\mathbb{R}^p$ and then assign $x$ to the label of the nearest point. 
+Data are often represented as an $n \times d$ matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ in which each row $x_i \in \mathbb{R}^d$ is an observation of $d$ features and each column is a set of $n$ feature measurements.
+An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^d \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$.
+For instance, $\phi(x)$ might first compute the distances from $x$ to $K$ points in $\mathbb{R}^d$ and then assign $x$ to the label of the nearest point. 
 Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning).
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector $x_i = \Big[ x_{i1} \mid \cdots \mid x_{ip} \Big]^\top \in \mathbb{R}^p$, we observe $a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
+Instead of feature vector $x_i = \Big[ x_{i1} \mid \cdots \mid x_{id} \Big]^\top \in \mathbb{R}^d$, we observe $a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
 For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
@@
 \begin{example}[Graph clustering]
-$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
+$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that assigns labels $z_1, ..., z_n \in \{1, ..., K\}$ to feature vectors $x_1, ..., x_n \in \mathbb{R}^d$ by minimizing the objective function
 \begin{equation}
@@
 \argmin_{z_1, ..., z_n} ~ W(z_1, ..., z_n; x_1, ..., x_n) = 
-\argmin_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
+\argmin_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2.
 \end{equation}
-in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments which we wish to optimize, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
+$\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
 A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.
 
-If we have a graph instead of vectors in $\mathbb{R}^p$, it is not obvious how to translate $K$-means clustering to these data. 
+If we have a graph instead of vectors in $\mathbb{R}^d$, it is not obvious how to translate $K$-means clustering to these data. 
 In particular, there is no inherent notion of centroid or sample mean for graph data. 
@@
 Then Lloyd's algorithm can be applied to this embedding. 
-More often in practice, the $d$ smallest eigenvalues are used to construct a lower dimensional embedding. 
+More often in practice, the $d \ll n$ smallest eigenvalues are used to construct a lower dimensional embedding. 
 \end{example}
@@
 
-Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process. 
+One approach to analyzing graphs as network data is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process. 
 These estimated parameters are often then used for making predictions or deriving insights about the population. 
@@
 Additional structure must be introduced. 
-One such structured Bernoulli random graph model is the $\ER$ graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
+One such structured Bernoulli random graph model is the $\ER$ graph, first proposed by \cite{Gilbert:1959}, which restricts every edge probability to a constant value.
 
@@
 In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.
-
+In particular, this work explores the ways in which block models and latent space models are related and uses that connection to develop methods for analyzing and fitting models to graph data. 
 The contributions of this work are as follows. 
@@
 In the case of intersecting manifolds, we propose the $K$-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds.
-
-We conclude with section \@ref(sec:comparison) in which we compare the four block models.
+We conclude with section \@ref(sec:comparison) in which we compare the four block models and summarize the ways in which one might determine which block model is most appropriate for one's data or setting.
 
@@
 The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities $(k, \ell)$ has a constant edge probability $\theta_{k \ell}$.
-We now give the formal definition of the SBM within the context of Bernoulli random graphs:
+The formal definition of the SBM within the context of Bernoulli random graphs is as follows:
 
@@
 \label{ex:assort-sbm}
-We construct an SBM with two communities in which $\theta_{11} = 1/2$, $\theta_{12} = 1/8$, and $\theta_{22} = 1/4$. 
+In this SBM, there are two communities with parameters $\theta_{11} = 1/2$, $\theta_{12} = 1/8$, and $\theta_{22} = 1/4$. 
 Note that in this example, the within-community edge probabilities are greater than the between-community edge probability. 
@@
 \label{ex:sbm-dating}
-Consider an undirected graph in which each vertex $v_i$ represents users of an online dating service, each label $z_i$ represents the user's gender, and each edge represents a successful match between pairs of users.
-For simplicity, we will restrict the labels to female and male (denoted as $1$ and $2$). 
+Consider an undirected graph in which each vertex $v_i$ represents a user of an online dating service, each label $z_i$ represents the user's gender, and each edge represents a successful match between pairs of users.
+For simplicity, we will restrict the labels to female and male (denoted as $1$ and $2$, respectively). 
 If we model this as an SBM, then $\theta_{11}$ is the probability of a match between two female users, $\theta_{22}$ is the probability of a match between two male users, and $\theta_{12} = \theta_{21}$ is the probability of a match between a female user and a male user.
@@
 a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
-This corresponds to the high-level understanding of assortativity and disassortativity based on whether within or between community edge probabilities are greater. 
+This corresponds to the high-level understanding of assortativity and disassortativity based on whether within community edge probabilities are greater than the between community edge probabilities. 
 \end{example}
@@
 (The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.) 
-As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
+The EM algorithm for maximizing the likelihood function in equation \eqref{eq:sbm-loglik} is detailed Kin algorithm \ref{algo:sbm-em}.
 
@@
 In many real networks, the degree of each vertex often does not grow proportionally with the size of the network. 
-To account for this, a sparsity factor $\rho_n \in (0, 1]$ is introduced in the edge probabilities, i.e., $P_{ij} \leftarrow \rho_n P_{ij}$, for some sequence $\{\rho_n\}$ such that $\lim\limits_{n \to \infty} \rho_n = 0$.
+To account for this, a sparsity factor $\rho_n \in (0, 1]$ is introduced in the edge probabilities, i.e., $P_{ij} \leftarrow \rho_n P_{ij}$, for some sequence $\{\rho_n\}$. 
+Oftentimes the additional constraint of $\lim\limits_{n \to \infty} \rho_n = 0$ is included. 
 For example, a sparse SBM has edge probabilities $P_{ij} = \rho_n \theta_{z_i, z_j}$, for which we use the notation $A \sim \SBM(z, \{\theta_{k \ell}\}_K; \rho_n)$ or $A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K; \rho_n)$, depending on whether we treat the labels as random or fixed.
@@
 Finally, we also note that while $\rho_n$ limits the rate of growth of the expected degree, our theoretical results still require $n \rho_n$ to diverge to infinity, albeit at a slower rate than $O(n)$.
+For example, if $\rho_n \propto 1 / \sqrt{n}$, then the expected degree for each vertex of the SBM is $O(\sqrt{n})$. 
 In general, for consistency in estimation and inference, most results on Bernoulli random graphs require $n \rho_n = \omega \big( (\log n)^c \big)$ for some $c > 1$. 
@@
 The SBM assumes homogeneity within each community. 
-In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label. 
+More specifically, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label. 
 The degree corrected block model (DCBM) and popularity adjusted block model (PABM) are generalizations of the SBM that address these limitations. 
@@
 Let $G = (V, E)$ be a Bernoulli random graph with $n$ vertices, described by random adjacency matrix $A$. 
-Let $K \geq 1$ be an integer that describes the number of communities and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$), as in the SBM.
+Let $K \geq 1$ be an integer that describes the number of communities and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$) be the baseline community-wise edge probabilities, as in the SBM.
 Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
@@
 In example \ref{ex:sbm-dating}, we modeled an online dating network as an SBM in which the vertices correspond to individuals, the communities correspond to genders, and edges correspond to whether the online dating service successfully matched pairs of individuals.
-In this model, there is an implicit assumption that for each pair of genders, there is a constant probability of a match, e.g., each male member has a constant probability of matching with another male member as well as a (different) constant probability of matching with a female member.
+In this model, there is an implicit assumption that for each pair of genders, there is a constant probability of a match, e.g., each male member has a constant probability of matching with any other male member as well as a (different) constant probability of matching with any female member.
 This does not take into account that some individuals are more likely to form connections than others, based on each individual's activity on the online service or popularity with other users. 
@@
 However, it still assumes that each vertex within a community has the same relative propensity to connect to each community. 
-Referring again to the dating network example (examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}), the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's preferences.
+Referring again to the dating network examples  \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's sexual orientation.
 The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community. 
@@
 matching with a member of the opposite gender is constant across all members of each gender.
-Using the parameters specified in these examples, we assume that each individual is 10 times more likely to match with a member of the opposite gender compared to a member of the same gender. 
+Using the parameters specified in these examples, we assume that every individual is 10 times more likely to match with a member of the opposite gender compared to a member of the same gender. 
 The PABM allows us to model each individual's sexual orientation. 
+For example, if the $i^{th}$ individual is female ($z_i = 1$) and she is heterosexual, then a plausible set of popularity parameters for this individual is $\lambda_{i1} = 0$ and $\lambda_{i2} = 0.1$,
+if she is homosexual, then a plausible set of popularity parameters is $\lambda_{i1} = 0.1$ and $\lambda_{i2} = 0$, 
+and if she is somewhere in-between, then $\lambda_{i1} = \lambda_{i2} = 0.05$. 
 Figure \ref{fig:dating-pabm} is a visualization of the online dating network modeled as a PABM in which the popularity parameters are drawn as $\lambda_{ik} \sim \Betadist(1, 6)$ if $z_i = k$ and $\lambda_{ik} \sim \Betadist(2, 2)$ if $z_i \neq k$.
-Like in the SBM and DCBM examples, the average edge probability between genders is approximately ten times that of the average edge probability within genders, but unlike in the SBM and DCBM examples, we can see that some vertices have more connections to its own gender and some have more connections to the opposite gender.
+Like in the SBM and DCBM examples, the average edge probability between genders is approximately ten times that of the average edge probability within genders, but unlike in the SBM and DCBM examples, some vertices have more connections to its own gender and some have more connections to the opposite gender.
 
@@
 For example, in the case of the DCBM, doubling each $\theta_{k \ell}$ and dividing each $\omega_i$ by $\sqrt{2}$ results in the same edge probability matrix. 
-\citet{Karrer_2011} suggest imposing the constraint $\sum_{i : z_i = k} \omega_i = 1$ for each $k \in \{1, ..., K\}$, but other similar constraints can be used to enforce uniqueness, such as $\max_i \omega_i = 1$ and $\min_i \omega_i = 0$.
-Despite the lack of uniqueness without constraints, a naive EM type algorithm similar to the one in section \ref{sec:sbm} results in high community detection accuracy for most realizations of the DCBM and PABM, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees.
+\citet{Karrer_2011} suggested imposing the constraint $\sum_{i : z_i = k} \omega_i = 1$ for each $k \in \{1, ..., K\}$, but other similar constraints can be used to enforce uniqueness, such as $\max_i \omega_i = 1$ and $\min_i \omega_i = 0$.
+Despite the lack of uniqueness without constraints, naive EM algorithm result in high community detection accuracy in numerical experiments of the DCBM and PABM, although it is not clear what its asymptotic properties are or whether there are any theoretical guarantees.
 Likewise, it is not immediately obvious how to approach the estimation of the $\theta_{k \ell}$'s, $\omega_i$'s, or $\lambda_{ik}$'s due to their non-uniqueness. 
 \citet{noroozi2019estimation} suggest focusing estimation on the edge probability matrix $P$ instead of the individual parameters to avoid this ambiguity. 
-Nevertheless, the naive EM algorithms appear to work well, at least for community detection (estimating the labels $z_i$, e.g., by setting $\hat{z}_i = \arg\max_k \pi_{ik}$), based on empirical results.
-Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 95\% accuracy, but they fail to capture the original (which are not necessarily ``true'') parameters.
+Applying algorithm \ref{algo:dcbm-em} to example \ref{ex:dcbm-dating} results in 84\% accuracy, and applying algorithm \ref{algo:pabm-em} to example \ref{ex:pabm-dating} results in 95\% accuracy, but it is unclear how to quantify the quality of the parameter estimates outputted by these algorithms.
 \end{remark}
@@
 In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}), yielding a community detection accuracy of 79.7\%. 
-Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} yields a greater accuracy of 94.9\%. 
+Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} on the graph to fit a DCBM yields a greater accuracy of 94.9\%. 
 This is expected, since the SBM is a special case of the DCBM, i.e., the SBM is the DCBM with some of the parameters restricted. 
-Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM. 
-Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a better fit (figure \ref{fig:uk-faculty-dcbm-omega}). 
+Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM or it is overfitting. 
+Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a more appropriate model (figure \ref{fig:uk-faculty-dcbm-omega}). 
 ```{r uk-faculty-dcbm-omega, cache = TRUE, fig.width = 8, fig.height = 2, fig.cap = 'Histogram of the fitted degree factors for the UK faculty friendship network assuming that it is a DCBM.
@@
 
-In the Bernoulli random graph, the edge probability matrix $P$ is of size $n \times n$, but when $K \ll n$, the structure of $P$ is much simpler than its dimensions imply. 
+In block models, the edge probability matrix $P$ is of size $n \times n$, but when $K \ll n$, the structure of $P$ is much simpler than its dimensions imply. 
 In particular, $P$ is rank $K$ in the SBM and DCBM and rank $K^2$ in the PABM. 
@@
 We use the notation $A \sim \GRDPG_{p,q}(X; \rho_n)$ to denote a random adjacency matrix $A$ drawn from latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$.
-If the $n$ vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F; \rho_n)$.
+If the $n$ latent vectors in $X$ are drawn from probability distribution $F$ on support $\mathcal{X}$, we use the notation $A \sim \GRDPG_{p,q}(F; \rho_n)$.
 
@@
 $(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j = x_i^\top I_{p, q} x_j = P_{ij}$. 
-Unlike in the RDPG case, transforming the latent positions via multiplication by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or distances.
+Unless the graph is an RDPG ($q = 0$ and $p = d$) transforming the latent positions via multiplication by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or distances.
 \end{remark}
@@
 $$\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)$$
-where $x_i^\top$ and $\hat{x}_i^\top$ are the rows of $\hat{X}$ (from the ASE) and $X$, respectively.
+where $x_i^\top$ and $\hat{x}_i^\top$ are the rows of $X$ and $\hat{X}$, respectively.
 
@@
 $P$ is the edge probability matrix of a GRDPG if and only if there exists an $X \in \mathbb{R}^{n \times d}$ such that $P = X I_{p,q} X^\top$. 
-$P$ is a square and symmetric matrix, so it has an eigendecomposition $P = V \Lambda V^\top$ where $V \in \mathbb{R}^{n \times d}$, $\Lambda$ is a $d \times d$ diagonal matrix, and $d = \rank(P)$. 
+$P$ is a square and symmetric matrix, so it has an eigendecomposition $P = V \Lambda V^\top$ where $V \in \mathbb{R}^{n \times d}$, $\Lambda$ is a $d \times d$ diagonal matrix, and $d = \rank(P) \ll n$.
 Let $p$ and $q$ respectively be the number of positive and negative eigenvalues of $P$. 
@@
 \citet{rubindelanchy2017statistical} showed that this approach to community detection for the SBM and DCBM is asymptotically consistent. 
-More specifically, under certain sparsity conditions, clustering on the ASE of the SBM and DCBM results in zero misclustered vertices. 
-Note that for a given $P$, $X$ is not unique, i.e., for any $Q \in \mathbb{O}(p, q)$, $X Q$ results in the same $P$ (see remark \ref{rem:non_identifiable}). 
-However, this does not change the general structure of the SBM and DCBM in the latent space. 
+More specifically, under certain sparsity conditions, clustering on the ASE of the SBM and DCBM results in zero misclustered vertices if $n$ is sufficiently large. 
+Although for a given $P$, $X$ is not unique, i.e., for any $Q \in \mathbb{O}(p, q)$, $X Q$ results in the same $P$ (see remark \ref{rem:non_identifiable}), it does not change the general structure of the SBM and DCBM in the latent space.
 Multiplication by $Q \in \mathbb{O}(p,q)$ does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, since $Q$ is a linear transformation.
 
-\begin{example}
-Consider a two-community assortative SBM. 
+\begin{example}[Two community assortative SBM and DCBM]
 Recall that assortativity in SBMs is characterized by positive definite edge probability matrices. 
-Therefore, this model is also an RDPG. 
+Therefore, this model is also an RDPG ($p = d = 2$, $q = 0$). 
 To show this explicitly, suppose that $P$ is organized by community such that the $P^{(k \ell)}$ block corresponds to edges from community $k$ to $\ell$, and consider 
@@
 \end{bmatrix}.$$
-The resulting rows of $X$ are point masses that have been stretched out, turning into rays. 
+The resulting rows of $X$ are point masses that have been stretched out toward the origin, turning into rays. 
 \end{example}
@@
 Another clustering algorithm for the ASE of a DCBM is to normalize the embedding vectors to norm 1 and then apply $K$-means to the normalized vectors. 
+This suggests that the ASE is not only useful for clustering and inference but also for exploratory data analysis. 
+The structure of the ASE reveals the appropriate model for the graph.
 

### 2022-12-24 draft/draft.tex
fixed a bunch of typos

@@
 \LARGE{\textsc{Community Detection in the Setting of Generalized Random Dot Product Graphs}}\\
-\vspace*{6\baselineskip}
+\vspace*{4\baselineskip}
 \normalsize{John Koo}\\
-\vspace*{10\baselineskip}
+\vspace*{8\baselineskip}
 \singlespacing
@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model.
-Data are often represented as an \(n \times p\) matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
-An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
-For instance, \(\phi(x)\) might first compute the distances from \(x\) to \(K\) points in \(\mathbb{R}^p\) and then assign \(x\) to the label of the nearest point.
+Data are often represented as an \(n \times d\) matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\) in which each row \(x_i \in \mathbb{R}^d\) is an observation of \(d\) features and each column is a set of \(n\) feature measurements.
+An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^d \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
+For instance, \(\phi(x)\) might first compute the distances from \(x\) to \(K\) points in \(\mathbb{R}^d\) and then assign \(x\) to the label of the nearest point.
 Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning).
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector \(x_i = \Big[ x_{i1} \mid \cdots \mid x_{ip} \Big]^\top \in \mathbb{R}^p\), we observe \(a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
+Instead of feature vector \(x_i = \Big[ x_{i1} \mid \cdots \mid x_{id} \Big]^\top \in \mathbb{R}^d\), we observe \(a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
 For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
@@
 \mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ W(z_1, ..., z_n; x_1, ..., x_n) = 
-\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
+\mathop{\mathrm{arg\,min}}_{z_1, ..., z_n} ~ \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2.
 \end{equation}
-in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments which we wish to optimize, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters.
+$\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
 A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.
 
-If we have a graph instead of vectors in $\mathbb{R}^p$, it is not obvious how to translate $K$-means clustering to these data. 
+If we have a graph instead of vectors in $\mathbb{R}^d$, it is not obvious how to translate $K$-means clustering to these data. 
 In particular, there is no inherent notion of centroid or sample mean for graph data. 
@@
 
-Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process.
+One approach to analyzing graphs as network data is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process.
 These estimated parameters are often then used for making predictions or deriving insights about the population.
@@
 The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights.
-The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components.\\
+The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components.
 In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.
@@
 Additional structure must be introduced.
-One such structured Bernoulli random graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:
+One such structured Bernoulli random graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which restricts every edge probability to a constant value.
 
@@
 In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.
-
+In particular, this work explores the ways in which block models and latent space models are related and uses that connection to develop methods for analyzing and fitting models to graph data.
 The contributions of this work are as follows.
@@
 In the case of intersecting manifolds, we propose the \(K\)-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds.
-
-We conclude with section \ref{sec:comparison} in which we compare the four block models.
+We conclude with section \ref{sec:comparison} in which we compare the four block models and summarize the ways in which one might determine which block model is most appropriate for one's data or setting.
 
@@
 The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities \((k, \ell)\) has a constant edge probability \(\theta_{k \ell}\).
-We now give the formal definition of the SBM within the context of Bernoulli random graphs:
+The formal definition of the SBM within the context of Bernoulli random graphs is as follows:
 
@@
 (The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.)
-As an example, we write the EM algorithm for the likelihood function in equation \eqref{eq:sbm-loglik} in algorithm \ref{algo:sbm-em}.
+The EM algorithm for maximizing the likelihood function in equation \eqref{eq:sbm-loglik} is detailed Kin algorithm \ref{algo:sbm-em}.
 
@@
 In many real networks, the degree of each vertex often does not grow proportionally with the size of the network.
-To account for this, a sparsity factor \(\rho_n \in (0, 1]\) is introduced in the edge probabilities, i.e., \(P_{ij} \leftarrow \rho_n P_{ij}\), for some sequence \(\{\rho_n\}\) such that \(\lim\limits_{n \to \infty} \rho_n = 0\).\\
+To account for this, a sparsity factor \(\rho_n \in (0, 1]\) is introduced in the edge probabilities, i.e., \(P_{ij} \leftarrow \rho_n P_{ij}\), for some sequence \(\{\rho_n\}\).
+Oftentimes the additional constraint of \(\lim\limits_{n \to \infty} \rho_n = 0\) is included.
 For example, a sparse SBM has edge probabilities \(P_{ij} = \rho_n \theta_{z_i, z_j}\), for which we use the notation \(A \sim \SBM(z, \{\theta_{k \ell}\}_K; \rho_n)\) or \(A \sim \SBM(\alpha, \{\theta_{k \ell}\}_K; \rho_n)\), depending on whether we treat the labels as random or fixed.
@@
 Finally, we also note that while \(\rho_n\) limits the rate of growth of the expected degree, our theoretical results still require \(n \rho_n\) to diverge to infinity, albeit at a slower rate than \(O(n)\).
+For example, if \(\rho_n \propto 1 / \sqrt{n}\), then the expected degree for each vertex of the SBM is \(O(\sqrt{n})\).
 In general, for consistency in estimation and inference, most results on Bernoulli random graphs require \(n \rho_n = \omega \big( (\log n)^c \big)\) for some \(c > 1\).
@@
 The SBM assumes homogeneity within each community.
-In particular, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label.
+More specifically, in the SBM, each vertex's expected degree and affinity toward each community depends only on its community label.
 The degree corrected block model (DCBM) and popularity adjusted block model (PABM) are generalizations of the SBM that address these limitations.
@@
 However, it still assumes that each vertex within a community has the same relative propensity to connect to each community.
-Referring again to the dating network example (examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}), the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's preferences.
+Referring again to the dating network examples \ref{ex:sbm-dating} and \ref{ex:dcbm-dating}, the model assume that each individual is ten times more likely to connect with a member of the opposite sex than with a member of the same sex, without regard to each individual's sexual orientation.
 The popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1} allows each vertex to have its own affinity toward each community.
@@
 In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}), yielding a community detection accuracy of 79.7\%. 
-Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} yields a greater accuracy of 94.9\%. 
+Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} on the graph to fit a DCBM yields a greater accuracy of 94.9\%. 
 This is expected, since the SBM is a special case of the DCBM, i.e., the SBM is the DCBM with some of the parameters restricted. 
-Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM. 
-Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a better fit (figure \ref{fig:uk-faculty-dcbm-omega}). 
+Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM or it is overfitting. 
+Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a more appropriate model (figure \ref{fig:uk-faculty-dcbm-omega}). 
 \begin{figure}[H]
@@
 
-In the Bernoulli random graph, the edge probability matrix \(P\) is of size \(n \times n\), but when \(K \ll n\), the structure of \(P\) is much simpler than its dimensions imply.
+In block models, the edge probability matrix \(P\) is of size \(n \times n\), but when \(K \ll n\), the structure of \(P\) is much simpler than its dimensions imply.
 In particular, \(P\) is rank \(K\) in the SBM and DCBM and rank \(K^2\) in the PABM.
@@
 \[\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)\]
-where \(x_i^\top\) and \(\hat{x}_i^\top\) are the rows of \(\hat{X}\) (from the ASE) and \(X\), respectively.
+where \(x_i^\top\) and \(\hat{x}_i^\top\) are the rows of \(X\) and \(\hat{X}\), respectively.
 
@@
 \citet{rubindelanchy2017statistical} showed that this approach to community detection for the SBM and DCBM is asymptotically consistent.
-More specifically, under certain sparsity conditions, clustering on the ASE of the SBM and DCBM results in zero misclustered vertices.
-Note that for a given \(P\), \(X\) is not unique, i.e., for any \(Q \in \mathbb{O}(p, q)\), \(X Q\) results in the same \(P\) (see remark \ref{rem:non_identifiable}).
-However, this does not change the general structure of the SBM and DCBM in the latent space.
+More specifically, under certain sparsity conditions, clustering on the ASE of the SBM and DCBM results in zero misclustered vertices if \(n\) is sufficiently large.
+Although for a given \(P\), \(X\) is not unique, i.e., for any \(Q \in \mathbb{O}(p, q)\), \(X Q\) results in the same \(P\) (see remark \ref{rem:non_identifiable}), it does not change the general structure of the SBM and DCBM in the latent space.
 Multiplication by \(Q \in \mathbb{O}(p,q)\) does not change the fact that the latent vectors of an SBM are point masses or that the latent vectors of a DCBM are line segments, since \(Q\) is a linear transformation.
 
-\begin{example}
-Consider a two-community assortative SBM. 
+\begin{example}[Two community assortative SBM and DCBM]
 Recall that assortativity in SBMs is characterized by positive definite edge probability matrices. 
-Therefore, this model is also an RDPG. 
+Therefore, this model is also an RDPG ($p = d = 2$, $q = 0$). 
 To show this explicitly, suppose that $P$ is organized by community such that the $P^{(k \ell)}$ block corresponds to edges from community $k$ to $\ell$, and consider 

### 2022-12-24 draft/draft.Rmd
fixed more typos

@@
 Analyzing the graph as a GRDPG, however, provides a visual indication for which block model is the best fit. 
-If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
+If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can look at the ASE of the adjacency matrix and determine visually whether the embedding vectors lie around point masses or rays.
 Based on figure \ref{fig:friendship-network}, we assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
 Because the fourth community only contains two members, we omitted it from this analysis. 
-Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM, since each community appears to be described by a line segment. 
+Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM, since the embedding vectors for each community appear to lie on or near one of three line segments, which correspond to the three communities.
 $K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, which is an improvement over the EM algorithms while being substantially faster and easier to implement.
@@
 we decomposed the adjacency matrix to construct its ASE (figure \ref{fig:macaque-ase}). 
-The embedding vectors, which approximate the latent vectors, appear to approximately lie along two curves that intersect at the origin, with each curve corresponding to a label (visual vs sensorimotor).
+The embedding vectors, which approximate the latent vectors, appear to lie on or near two curves that intersect at the origin, with each curve corresponding to a label (visual vs. sensorimotor). 
 While the ASE suggests a community-wise structure in the latent space, the SBM, DCBM, and PABM are not suitable models for these data.

### 2022-12-24 draft/draft.tex
fixed more typos

@@
 Another clustering algorithm for the ASE of a DCBM is to normalize the embedding vectors to norm 1 and then apply \(K\)-means to the normalized vectors.
+This suggests that the ASE is not only useful for clustering and inference but also for exploratory data analysis.
+The structure of the ASE reveals the appropriate model for the graph.
 

### 2023-01-04 draft/draft.Rmd
responded to most comments

@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model. 
-Data are often represented as an $n \times d$ matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ in which each row $x_i \in \mathbb{R}^d$ is an observation of $d$ features and each column is a set of $n$ feature measurements.
+Data are often represented as an $n \times d$ matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$ in which each row $x_i \in \mathbb{R}^d$ is an observation of $d$ features and each column is a set of $n$ feature measurements.
 An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^d \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$.
@@
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector $x_i = \Big[ x_{i1} \mid \cdots \mid x_{id} \Big]^\top \in \mathbb{R}^d$, we observe $a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
-For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
+Instead of feature vector $x_i = [ x_{i1} \mid \cdots \mid x_{id} ]^\top \in \mathbb{R}^d$, we observe $a_i = [ a_{i1} \mid \cdots \mid a_{in} ]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$.
+For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding*, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
 
 \begin{example}[Graph clustering]
+We now provide an example of an example contrasting the difference between clustering $n$ objects represented as a collection of feature vectors vs.
+$n$ objects represented as vertices of a graph using the same clustering approach.
 $K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that assigns labels $z_1, ..., z_n \in \{1, ..., K\}$ to feature vectors $x_1, ..., x_n \in \mathbb{R}^d$ by minimizing the objective function
@@
 $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
-A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.
+A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of alternating coordinate descent algorithm in which at each iteration the cluster labels are updated according to the label of its nearest centroid and centroids are updated as the sample means of the points in each cluster.
 
@@
 This embedding is constructed by computing the combinatorial Laplacian matrix $L = D - A$ in which $D$ is a diagonal matrix in which each $D_{ii}$ is the degree of vertex $i$. 
-Then the components of the embedding consist of $\big[ v_{n-1} / \sqrt{\lambda_{n - 1}} \mid \cdots \mid v_1 / \sqrt{\lambda_1} \big]^\top$ in which $\lambda_i$ is the $i^{th}$ largest eigenvalue of $L$ and $v_i$ is its corresponding eigenvector ($\lambda_n = 0$ and is omitted).
+Then the components of the embedding consist of $[ v_{n-1} / \sqrt{\lambda_{n - 1}} \mid \cdots \mid v_1 / \sqrt{\lambda_1} ]^\top$ in which $\lambda_i$ is the $i^{th}$ largest eigenvalue of $L$ and $v_i$ is its corresponding eigenvector ($\lambda_n = 0$ and is omitted).
 Then Lloyd's algorithm can be applied to this embedding. 
@@
 For instance, in a social network graph, in which vertices correspond to individuals and edges correspond to connections between pairs of individuals, each vertex (individual) might also have a set or vector of other observations (e.g., age, sex, location, etc.).
-This is often the setting for inference in the *exponential family of random graphs* (ERGM) \citep{kolaczyk2014statistical}.
+This is often the setting for inference in the *exponential family of random graphs* (ERGM) \citep{10.2307/2287037}.
 Another setting in which vertices have attributes is when the (oftentimes single) attribute is hidden or unobserved, and the inference task is to identify the attributes. 
@@
 Then similar to how the edges are compiled into an adjacency matrix $A$, the edge probabilities can be compiled into an edge probability matrix $P \in [0, 1]^{n \times n}$. 
-This type of graph model is defined as a *Bernoulli random graph* (also called an inhomogeneous $\ER$ graph). 
+This type of graph model is defined as a *Bernoulli random graph* (also called an inhomogeneous $\ER$ graph \citep{10.5555/1276871.1276872}). 
 If $A$ is the adjacency matrix of a Bernoulli random graph with edge probability matrix $P$, we denote $A \sim \BG(P)$. 
-If the vertices and edge probabilities are sampled as a sequence of $n$ i.i.d. random variables from probability distribution $F$, then we denote $A \sim \BG(F, n)$.
+If the vertices and edge probabilities are sampled as a sequence of $n$ i.i.d. random variables from probability distribution $F$, then we denote $A \sim \BG(F)$.
 
@@
 Much work has been done in developing various Bernoulli random graph models that strike a balance between structure and flexibility. 
-Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called *block models*, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called *latent space models*.
+Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called *block models*, and edge probabilities based on positions in a latent space \citep{doi:10.1198/016214502388618906, 10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called *latent space models*.
 
@@
 
-To estimate $\omega = \big[ \omega_1, \cdots, \omega_n \big]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
+To estimate $\omega = [ \omega_1, \cdots, \omega_n ]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization.
 For simplicity, set $C = 1$. 
@@
 Instead of points, line segments, and subspaces, we propose the manifold block model which is a class of generalized random dot product graph in which the latent vectors lie on one of $K$ manifolds. 
-In the case of nonintersecting manifolds, we show that a clustering algorithm based on a nearest algorithms graph approach on an embedding is consistent, resulting in zero clustering error for large graphs, as long as the latent vectors are sampled from the manifolds densely
+In the case of nonintersecting manifolds, we propose an algorithm that first embeds the graph followed by nearest neighbors clustering, such as single linkage. 
+This algorithm is consistent, resulting in zero clustering error for large graphs, as long as the distribution of latent vectors on each manifold is dense. 
 In the case of intersecting manifolds, we propose the $K$-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds.
@@
 This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices $v_i$ and $v_j$ depend on the community assignments $z_i$ and $z_j$.
-More formally, one assigns each vertex $v_i$ a community label $z_i \in \{1, ..., K\}$ and assumes a Bernoulli random graph in which $P_{ij} = g(z_i, z_j, \theta_i, \theta_j)$ for some function $g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]$ and $\theta_i \in \mathbb{R}^d$ is a vector of real-valued parameters associated with each vertex.
+More formally, one assigns each vertex $v_i$ a community label $z_i \in \{1, ..., K\}$ and assumes a Bernoulli random graph in which $P_{ij} = g(z_i, z_j, \xi_i, \xi_j)$ for some function $g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]$ and $\xi_i \in \mathbb{R}^d$ is a vector of real-valued parameters associated with each vertex.
 While the graph is sampled conditioned on the labels $z_i$, it is assumed that the labels are unknown when the graph is observed. 
@@
 (The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.) 
-The EM algorithm for maximizing the likelihood function in equation \eqref{eq:sbm-loglik} is detailed Kin algorithm \ref{algo:sbm-em}.
+The EM algorithm for maximizing the likelihood function in equation \eqref{eq:sbm-loglik} is detailed in algorithm \ref{algo:sbm-em}.
 
@@
 \begin{split}
-E[d_i] & = \sum_j P_{ij} \\
-& = \sum_{j \neq i} \theta_{z_i, z_j} \\ 
-& = \Big( \sum_k n_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
-& = n \Big( \sum_k \alpha_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} \\
-& = O(n).
+E[d_i]  = \sum_j P_{ij} 
+ = \sum_{j \neq i} \theta_{z_i, z_j} \
+ = \Big( \sum_k n_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} 
+ = n \Big( \sum_k \alpha_k \theta_{z_i, k} \Big) - \theta_{z_i, z_i} 
+ = O(n).
 \end{split}
@@
 
-## Generalizations of the Stochastic Block Model: The Degree Corrected Block Model and the Popularity Adjusted Block Model {#sec:dcbm-pabm}
+## The Degree Corrected Block Model and the Popularity Adjusted Block Model {#sec:dcbm-pabm}
 
@@
 Denote $d = p + q$ and let $\mathcal{X}$ be the subset of $\mathbb{R}^d$ such that, for any $x, y \in \mathcal{X}$, we have $x^\top I_{p,q} y \in [0, 1]$. 
-Let $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
+Let $X = [ x_1 \mid \cdots \mid x_n ]^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
 A graph $G$ with adjacency matrix $A$ is said to be a generalized random dot product graph with latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$ if $A \sim \BG(P)$ where the edge probability matrix $P$ is given by $P = \rho_n X I_{p,q} X^\top$, i.e., the entries of $P$ are of the form $P_{ij} = \rho_n x_i^\top I_{p,q} x_j$.
@@
 Suppose that latent vectors for an RDPG are drawn from the curve 
-$x(t) = \Big[ t^2, 2 t (t-1) \Big]^\top$, $t \in [0, 1]$. 
+$x(t) = [ t^2, 2 t (t-1) ]^\top$, $t \in [0, 1]$. 
 Let $x_i = x(t_i)$. 
@@
 More specifically, the edge probability matrix for each has one positive eigenvalue and one negative eigenvalue. 
-Then the ASE for this model is $\hat{X} = \Big[ |\lambda_1|^{1/2} v_1 \mid |\lambda_n|^{1/2} v_n \Big]$. 
+Then the ASE for this model is $\hat{X} = [ |\lambda_1|^{1/2} v_1 \mid |\lambda_n|^{1/2} v_n ]$. 
 A plot of the ASE of the graph in \ref{ex:sbm-dating} (figure \ref{fig:dating-ase}, left) suggests two (somewhat overlapping) ellipsoidal clusters, and fitting a GMM on the embedding results in a community detection accuracy of 96.9\%, which is comparable to algorithm \ref{algo:sbm-em} while being simpler to implement.
@@
 For the general case, we can extend $\tilde{P} = X \Pi X^\top$ to larger $K$. 
-For a more concrete example of this, refer to Example~1. 
+For a more concrete example of this, refer to example~\ref{ex:pabm3}. 
 We once again consider $\tilde{P}$ as defined in
@@
 
-\begin{example} Let $A$ be a $3$ blocks PABM with latent vectors $\{\lambda^{(k \ell)} \colon 1 \leq k \leq 3, 1 \leq \ell \leq 3\}$. 
+\begin{example} 
+\label{ex:pabm3}
+Let $A$ be a $3$ blocks PABM with latent vectors $\{\lambda^{(k \ell)} \colon 1 \leq k \leq 3, 1 \leq \ell \leq 3\}$. 
 Using the same notation as in theorem~\ref{thm:pabm-grdpg}, we can define
@@
 to perform SSC on the rows of the spectral embedding of $A$, 
-since the embedding of $P$ consists of $K$ subspaces (theore \ref{thm:pabm-grdpg}), 
+since the embedding of $P$ consists of $K$ subspaces (theorem \ref{thm:pabm-grdpg}), 
 and so the embedding of $A$ will lie approximately on the $K$ subspaces. 
@@
 In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
-This corresponds to a one-dimensional RDPG with a latent structure $X = \Big[ \sqrt{p},  \cdots, \sqrt{p} \Big]^\top$. 
-In another view, the RDPG with latent structure $X = \Big[ \sqrt{p}, \cdots, \sqrt{p} \Big]^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
-
-Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \Big[ \cos \frac{\pi}{2} t, \sin \frac{\pi}{2} t \Big]^\top$ for $t \in [0, 1]$.
+This corresponds to a one-dimensional RDPG with a latent structure $X = [ \sqrt{p},  \cdots, \sqrt{p} ]^\top$. 
+In another view, the RDPG with latent structure $X = [ \sqrt{p}, \cdots, \sqrt{p} ]^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.
+
+Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = [ \cos \frac{\pi}{2} t, \sin \frac{\pi}{2} t ]^\top$ for $t \in [0, 1]$.
 Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$.
@@
 \item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
-\item Compile the latent vectors into data matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
+\item Compile the latent vectors into data matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
 \end{enumerate}
@@
 \label{ex:separated-curves}
-Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \Big[ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) \Big]^\top$ and $f_2(t) = \Big[ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) \Big]^\top$.
-Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$.
+Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = [ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) ]^\top$ and $f_2(t) = [ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) ]^\top$.
+Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$.
 Finally, let $A \sim \RDPG(X)$.
@@
 Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$.
-Let $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
+Let $X = [ x_1 \mid \cdots \mid x_n ]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
 Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
@@
 2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
-3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$.
+3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$.
 4. Draw $A \sim \GRDPG_{p,q}(X; \rho_n)$.
@@
 \label{ex:intersect-curves}
-Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$.
+Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = [ t^2, 2 t (1-t) ]^\top$ and $g_2(t) = [ 2 t (1-t), (1-t)^2 ]^\top$. 
 This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
@@
 Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
-In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \Big[ 0, 0 \Big]^\top$, i.e., the curves meet at the origin. 
+In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = [ 0, 0 ]^\top$, i.e., the curves meet at the origin. 
 This clustering achieves 87.5\% accuracy.
@@
 2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
-3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top$ and $g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top$. 
-Collect the latent vectors into matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$. 
+3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = [ t^2, 2 t (1-t) ]^\top$ and $g_2(t) = [ 2 t (1-t), (1-t)^2 ]^\top$. 
+Collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$. 
 4. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
@@
 3. Let each $x_i = g_{z_i}(t_i)$ where
-    i. $g_1(t) = \Big[ 2 t (t-1), t^2, 0 \Big]^\top$
-    ii. $g_2(t) = \Big[ 0, t^2, 2 t (t-1) \Big]^\top$
-    iii. $g_3(t) = \Big[2 t (t-1), t^2, 2 t (t-1) \Big]^\top$.
-4. Collect the latent vectors into matrix $X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top$.
+    i. $g_1(t) = [ 2 t (t-1), t^2, 0 ]^\top$
+    ii. $g_2(t) = [ 0, t^2, 2 t (t-1) ]^\top$
+    iii. $g_3(t) = [2 t (t-1), t^2, 2 t (t-1) ]^\top$.
+4. Collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$.
 5. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).
@@
 $$g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}$$
-where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top$.
+where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = [ p^{(0)} \mid \cdots \mid p^{(R)} ]^\top$.
 
@@
 $$L(p; t, X) = \|X - T p\|^2,$$
-where $p$ is defined as before, $T$ is an $n \times (R+1)$ matrix such that $T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}$, and $X = \Big[ x_1 \mid \cdots \mid x_n \Big]$. 
+where $p$ is defined as before, $T$ is an $n \times (R+1)$ matrix such that $T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}$, and $X = [ x_1 \mid \cdots \mid x_n ]$. 
 Then the least squares estimates for the Bezier coefficients $p$ is simply the linear least squares estimate 
@@
 
-On the other hand, suppose that we know the Bezier coefficients $p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top$, and we wish to fit timepoints $t_1, ..., t_n$ to vectors $x_1, ..., x_n$ lying on or near a Bezier polynomial of order $R$.
+On the other hand, suppose that we know the Bezier coefficients $p = [ p^{(0)} \mid \cdots \mid p^{(R)} ]^\top$, and we wish to fit timepoints $t_1, ..., t_n$ to vectors $x_1, ..., x_n$ lying on or near a Bezier polynomial of order $R$.
 In this setting, we can fit each individual $t_i$ independently by minimizing $L(t_i; p, x_i) = \|x_i - g(t_i; p)\|^2$ for each $t_i$. 

### 2023-01-04 draft/draft.tex
responded to most comments

@@
 \newtheorem{conjecture}{Conjecture}[section]
-\theoremstyle{definition}
+% \theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
+% \theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-\theoremstyle{definition}
+% \theoremstyle{definition}
 \newtheorem{exercise}{Exercise}[section]
-\theoremstyle{definition}
+% \theoremstyle{definition}
 \newtheorem{hypothesis}{Hypothesis}[section]
-\theoremstyle{remark}
+% \theoremstyle{remark}
 \newtheorem{remark}{Remark}
@@
 To illustrate this, consider the typical approach to building a statistical or machine learning model.
-Data are often represented as an \(n \times d\) matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\) in which each row \(x_i \in \mathbb{R}^d\) is an observation of \(d\) features and each column is a set of \(n\) feature measurements.
+Data are often represented as an \(n \times d\) matrix \(X = [ x_1 \mid \cdots \mid x_n ]^\top\) in which each row \(x_i \in \mathbb{R}^d\) is an observation of \(d\) features and each column is a set of \(n\) feature measurements.
 An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^d \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
@@
 However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
-Instead of feature vector \(x_i = \Big[ x_{i1} \mid \cdots \mid x_{id} \Big]^\top \in \mathbb{R}^d\), we observe \(a_i = \Big[ a_{i1} \mid \cdots \mid a_{in} \Big]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
-For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
+Instead of feature vector \(x_i = [ x_{i1} \mid \cdots \mid x_{id} ]^\top \in \mathbb{R}^d\), we observe \(a_i = [ a_{i1} \mid \cdots \mid a_{in} ]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
+For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.
 
 \begin{example}[Graph clustering]
+We now provide an example of an example contrasting the difference between clustering $n$ objects represented as a collection of feature vectors vs.
+$n$ objects represented as vertices of a graph using the same clustering approach.
 $K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that assigns labels $z_1, ..., z_n \in \{1, ..., K\}$ to feature vectors $x_1, ..., x_n \in \mathbb{R}^d$ by minimizing the objective function
@@
 For instance, in a social network graph, in which vertices correspond to individuals and edges correspond to connections between pairs of individuals, each vertex (individual) might also have a set or vector of other observations (e.g., age, sex, location, etc.).
-This is often the setting for inference in the \emph{exponential family of random graphs} (ERGM) \citep{kolaczyk2014statistical}.
+This is often the setting for inference in the \emph{exponential family of random graphs} (ERGM) \citep{10.2307/2287037}.
 Another setting in which vertices have attributes is when the (oftentimes single) attribute is hidden or unobserved, and the inference task is to identify the attributes.
@@
 Then similar to how the edges are compiled into an adjacency matrix \(A\), the edge probabilities can be compiled into an edge probability matrix \(P \in [0, 1]^{n \times n}\).
-This type of graph model is defined as a \emph{Bernoulli random graph} (also called an inhomogeneous \(\ER\) graph).
+This type of graph model is defined as a \emph{Bernoulli random graph} (also called an inhomogeneous \(\ER\) graph \citep{10.5555/1276871.1276872}).
 If \(A\) is the adjacency matrix of a Bernoulli random graph with edge probability matrix \(P\), we denote \(A \sim \BG(P)\).
-If the vertices and edge probabilities are sampled as a sequence of \(n\) i.i.d. random variables from probability distribution \(F\), then we denote \(A \sim \BG(F, n)\).
+If the vertices and edge probabilities are sampled as a sequence of \(n\) i.i.d. random variables from probability distribution \(F\), then we denote \(A \sim \BG(F)\).
 
@@
 Much work has been done in developing various Bernoulli random graph models that strike a balance between structure and flexibility.
-Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called \emph{block models}, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called \emph{latent space models}.
+Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called \emph{block models}, and edge probabilities based on positions in a latent space \citep{doi:10.1198/016214502388618906, 10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called \emph{latent space models}.
 
@@
 Instead of points, line segments, and subspaces, we propose the manifold block model which is a class of generalized random dot product graph in which the latent vectors lie on one of \(K\) manifolds.
-In the case of nonintersecting manifolds, we show that a clustering algorithm based on a nearest algorithms graph approach on an embedding is consistent, resulting in zero clustering error for large graphs, as long as the latent vectors are sampled from the manifolds densely
+In the case of nonintersecting manifolds, we propose an algorithm that first embeds the graph followed by nearest neighbors clustering, such as single linkage.
+This algorithm is consistent, resulting in zero clustering error for large graphs, as long as the distribution of latent vectors on each manifold is dense.
 In the case of intersecting manifolds, we propose the \(K\)-curves algorithm, which is a type of batch coordinate ascent algorithm which alternates between fitting manifolds to the embedding and reassigning clusters based on proximity to the manifolds.
@@
 This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices \(v_i\) and \(v_j\) depend on the community assignments \(z_i\) and \(z_j\).
-More formally, one assigns each vertex \(v_i\) a community label \(z_i \in \{1, ..., K\}\) and assumes a Bernoulli random graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
+More formally, one assigns each vertex \(v_i\) a community label \(z_i \in \{1, ..., K\}\) and assumes a Bernoulli random graph in which \(P_{ij} = g(z_i, z_j, \xi_i, \xi_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\xi_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
 While the graph is sampled conditioned on the labels \(z_i\), it is assumed that the labels are unknown when the graph is observed.
@@
 (The same type of approximation was used in the Bayesian context by \citet{10.2307/43974715}.)
-The EM algorithm for maximizing the likelihood function in equation \eqref{eq:sbm-loglik} is detailed Kin algorithm \ref{algo:sbm-em}.
+The EM algorithm for maximizing the likelihood function in equation \eqref{eq:sbm-loglik} is detailed in algorithm \ref{algo:sbm-em}.
 
@@
 \hypertarget{sec:dcbm-pabm}{%
-\subsection{Generalizations of the Stochastic Block Model: The Degree Corrected Block Model and the Popularity Adjusted Block Model}\label{sec:dcbm-pabm}}
+\subsection{The Degree Corrected Block Model and the Popularity Adjusted Block Model}\label{sec:dcbm-pabm}}
 
@@
 \item
-  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\).
+  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = [ x_1 \mid \cdots \mid x_n ]^\top\).
 \item
@@
 \item
-  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \Big[ t^2, 2 t (1-t) \Big]^\top\) and \(g_2(t) = \Big[ 2 t (1-t), (1-t)^2 \Big]^\top\).
-  Collect the latent vectors into matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\).
+  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = [ t^2, 2 t (1-t) ]^\top\) and \(g_2(t) = [ 2 t (1-t), (1-t)^2 ]^\top\).
+  Collect the latent vectors into matrix \(X = [ x_1 \mid \cdots \mid x_n ]^\top\).
 \item
@@
   \item
-    \(g_1(t) = \Big[ 2 t (t-1), t^2, 0 \Big]^\top\)
-  \item
-    \(g_2(t) = \Big[ 0, t^2, 2 t (t-1) \Big]^\top\)
-  \item
-    \(g_3(t) = \Big[2 t (t-1), t^2, 2 t (t-1) \Big]^\top\).
+    \(g_1(t) = [ 2 t (t-1), t^2, 0 ]^\top\)
+  \item
+    \(g_2(t) = [ 0, t^2, 2 t (t-1) ]^\top\)
+  \item
+    \(g_3(t) = [2 t (t-1), t^2, 2 t (t-1) ]^\top\).
   \end{enumerate}
 \item
-  Collect the latent vectors into matrix \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]^\top\).
+  Collect the latent vectors into matrix \(X = [ x_1 \mid \cdots \mid x_n ]^\top\).
 \item
@@
 \[g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}\]
-where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top\).
+where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = [ p^{(0)} \mid \cdots \mid p^{(R)} ]^\top\).
 
@@
 \[L(p; t, X) = \|X - T p\|^2,\]
-where \(p\) is defined as before, \(T\) is an \(n \times (R+1)\) matrix such that \(T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}\), and \(X = \Big[ x_1 \mid \cdots \mid x_n \Big]\).
+where \(p\) is defined as before, \(T\) is an \(n \times (R+1)\) matrix such that \(T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}\), and \(X = [ x_1 \mid \cdots \mid x_n ]\).
 Then the least squares estimates for the Bezier coefficients \(p\) is simply the linear least squares estimate
@@
 
-On the other hand, suppose that we know the Bezier coefficients \(p = \Big[ p^{(0)} \mid \cdots \mid p^{(R)} \Big]^\top\), and we wish to fit timepoints \(t_1, ..., t_n\) to vectors \(x_1, ..., x_n\) lying on or near a Bezier polynomial of order \(R\).
+On the other hand, suppose that we know the Bezier coefficients \(p = [ p^{(0)} \mid \cdots \mid p^{(R)} ]^\top\), and we wish to fit timepoints \(t_1, ..., t_n\) to vectors \(x_1, ..., x_n\) lying on or near a Bezier polynomial of order \(R\).
 In this setting, we can fit each individual \(t_i\) independently by minimizing \(L(t_i; p, x_i) = \|x_i - g(t_i; p)\|^2\) for each \(t_i\).

### 2023-01-06 draft/draft.Rmd
addressed almost all comments

@@
 
-In the setting in which the labels are known and the edges are assumed to be sampled from some distribution conditioned on the labels, then one inference task is to fit an ERGM to these data. 
+In the setting in which the labels are known and the edges are assumed to be sampled from some distribution conditioned on the labels, then one inference task is to fit a generative model (such as an ERGM) to these data.
 On the other hand, if the labels are unobserved and the task is to identify the labels (up to permutation) assuming that the labels are drawn from some distribution and the edges are also drawn from some distribution conditioned on the hidden labels, then the inference task is community detection.
@@
 \begin{example}
+In this example, we explore a graph for which likelihood maximization is a reasonable approach to estimation. 
 Consider a graph in which the vertices represent a collection of $n$ planets with intelligent life sending out radio waves, and the existence of an edge between vertices $i$ and $j$ represents whether planets $i$ and $j$ have established contact.
@@
 This is expected, since the SBM is a special case of the DCBM, i.e., the SBM is the DCBM with some of the parameters restricted. 
-Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM or it is overfitting. 
 Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a more appropriate model (figure \ref{fig:uk-faculty-dcbm-omega}). 
@@
 \label{ex:separated-curves}
-Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = [ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) ]^\top$ and $f_2(t) = [ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) ]^\top$.
+Define two nonintersecting one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = [ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) ]^\top$ and $f_2(t) = [ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) ]^\top$.
 Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$.
@@
 The remainder of this section explores under which conditions these criteria are met for the latent configuration, in which the latent vectors lie exactly on manifolds, as well as for the ASE of the GRDPG sampled from these latent configurations, in which the vectors lie approximately on manifolds with some noise that approaches zero asymptotically.
+More specifically, the following result in theorem \ref{theorem:nonintersect-grdpg} shows that single-linkage clustering on the ASE (algorithm \ref{alg:ase-singlelink}) results in perfect community detection for a sufficiently large yet finite $n$.
 
@@
 Apply single linkage clustering with $K$ communities on $\hat{X}$.\;
-\caption{ASE clustering for nonintersecting communities.}
+\caption{Graph manifold clustering via single-linkage.}
 \end{algorithm}
@@
 Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$.
-Let $X = [ x_1 \mid \cdots \mid x_n ]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$.
+Let $X = [ x_1 \mid \cdots \mid x_n ]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^{4c} n)$ for some $c > 1$.
 Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
@@
 For each $n$, we sampled 50 RDPGs with this latent configuration, and for each graph, we applied algorithm \ref{alg:ase-singlelink} as well as algorithm \ref{alg:kcurves} both with random initialization and with four known labels per community.
-Figure \ref{sim-nonintersect} shows that for sufficiently large graphs, single linkage clustering on the ASE is able to consistently produce perfect clustering, but $K$-curves clustering performs better for very small graphs.
+Figure \ref{fig:sim-nonintersect} shows that for sufficiently large graphs, single linkage clustering on the ASE is able to consistently produce perfect clustering, but $K$-curves clustering performs better for very small graphs if a few labels are known beforehand.
+$K$-curves clustering also has trouble with random initialization in this particular example. 
+This is because there are many plausible quadratic fits to random subsets of the embedding vectors.
 
@@
 
-```{r multifit, cache = TRUE, fig.height = 3, fig.width = 8, fig.cap = '$K$-curve clustering fits on the ASE using various approaches.
+```{r multifit, cache = TRUE, fig.height = 3, fig.width = 8, fig.cap = '$K$-curves clustering fits on the ASE using various approaches.
 Left: quadratic fit with random initialization.
@@
 
-```{r loss, cache = TRUE, fig.cap = 'Clustering loss vs. iteration for each run of $K$-curve clustering.', fig.width = 8, fig.height = 2}
+```{r loss, cache = TRUE, fig.cap = 'Clustering loss vs. iteration for each run of $K$-curves clustering.', fig.width = 8, fig.height = 2}
 loss.df <- dplyr::bind_rows(
   dplyr::tibble(loss = random$loss, 
-                initialization = 'random, quadratic',
+                `initialization and curve` = 'random, quadratic',
                 iteration = seq_along(random$loss)),
   dplyr::tibble(loss = spectral$loss, 
-                initialization = 'spectral, quadratic',
+                `initialization and curve` = 'spectral, quadratic',
                 iteration = seq_along(spectral$loss)),
   dplyr::tibble(loss = cubic$loss, 
-                initialization = 'spectral, cubic',
+                `initialization and curve` = 'spectral, cubic',
                 iteration = seq_along(cubic$loss))
 ) %>% 
-  dplyr::mutate(initialization = factor(initialization,
+  dplyr::mutate(`initialization and curve` = factor(`initialization and curve`,
                                         levels = c('random, quadratic',
@@
 ggplot(loss.df) + 
-  geom_point(aes(x = iteration, y = loss, colour = initialization)) + 
-  geom_line(aes(x = iteration, y = loss, colour = initialization)) + 
+  geom_point(aes(x = iteration, y = loss, colour = `initialization and curve`)) + 
+  geom_line(aes(x = iteration, y = loss, colour = `initialization and curve`)) + 
   # ylim(0, max.loss) +
@@
 \label{thm:kcurves-semisupervised}
-Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a Bezier curve of order $R$ that does not self-intersect (for any $s \neq t$, $g_k(s) \neq g_k(t)$).
+Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a curve of order $R$ that does not self-intersect (for any $s \neq t$, $g_k(s) \neq g_k(t)$).
 Suppose that for each community $k$, we have labels for at least $R + 1$ vertices. 
@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}). 
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 72.2\% accuracy for the left hemisphere connectome and 83.6\% accuracy for the right hemisphere connectome.
+Fitting four quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 72.2\% accuracy for the left hemisphere connectome and 83.6\% accuracy for the right hemisphere connectome.
 The curves were fitted on three-dimensional ASEs with $p = 2$ and $q = 1$, which were chosen based on setting the threshold $|\lambda_i| > \sqrt{n}$ where $\lambda_i$ is the $i$^th^ greatest eigenvalue.

### 2023-01-06 draft/draft.tex
addressed almost all comments

@@
 to perform SSC on the rows of the spectral embedding of \(A\),
-since the embedding of \(P\) consists of \(K\) subspaces (theore \ref{thm:pabm-grdpg}),
+since the embedding of \(P\) consists of \(K\) subspaces (theorem \ref{thm:pabm-grdpg}),
 and so the embedding of \(A\) will lie approximately on the \(K\) subspaces.
@@
 The remainder of this section explores under which conditions these criteria are met for the latent configuration, in which the latent vectors lie exactly on manifolds, as well as for the ASE of the GRDPG sampled from these latent configurations, in which the vectors lie approximately on manifolds with some noise that approaches zero asymptotically.
+More specifically, the following result in theorem \ref{theorem:nonintersect-grdpg} shows that single-linkage clustering on the ASE (algorithm \ref{alg:ase-singlelink}) results in perfect community detection for a sufficiently large yet finite \(n\).
 
@@
 
-\caption{$K$-curve clustering fits on the ASE using various approaches.
+\caption{$K$-curves clustering fits on the ASE using various approaches.
 Left: quadratic fit with random initialization.
@@
 
-\caption{Clustering loss vs. iteration for each run of $K$-curve clustering.}\label{fig:loss}
+\caption{Clustering loss vs. iteration for each run of $K$-curves clustering.}\label{fig:loss}
 \end{figure}
@@
 The ASE in two dimensions suggest that the MBM with one dimensional quadratic manifolds is an appropriate model for these graphs (figure \ref{fig:mbconnectome-ase}).
-Fitting two quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 72.2\% accuracy for the left hemisphere connectome and 83.6\% accuracy for the right hemisphere connectome.
+Fitting four quadratic Bezier curves to the ASEs (figure \ref{fig:mbconnectome-kcurves}) results in 72.2\% accuracy for the left hemisphere connectome and 83.6\% accuracy for the right hemisphere connectome.
 The curves were fitted on three-dimensional ASEs with \(p = 2\) and \(q = 1\), which were chosen based on setting the threshold \(|\lambda_i| > \sqrt{n}\) where \(\lambda_i\) is the \(i\)\textsuperscript{th} greatest eigenvalue.

### 2023-01-09 slides/slides.Rmd
finished first draft of slides

---
title: Community Detection in the Setting of Generalized Random Dot Product Graphs
author: |
  | John Koo
  |
  | Department of Statistics
  | Indiana University
date: 'January 2023'
output: 
  beamer_presentation:
    fig_crop: no
    theme: 'default'
    colortheme: 'beaver'
    includes:
      in_header: page_headers.tex
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \setbeamertemplate{itemize items}[circle]
- \usepackage{caption}
- \captionsetup[figure]{font=scriptsize}
- \captionsetup[figure]{labelformat=empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      fig.lp = '')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
library(ggplot2)
library(ggraph)
import::from(magrittr, `%>%`)
import::from(foreach, `%dopar%`)
theme_set(theme_bw())

source('~/dev/pabm-grdpg/functions.R')
source('~/dev/manifold-block-models/functions.R')
set.seed(314159)

doMC::registerDoMC(parallel::detectCores())
```

## Outline

1. Random Graph Models
    * Block Models
    * Generalized Random Dot Product Graphs

2. PABMs are GRDPGs

3. Manifold Block Models

# Random Graph Models

## Problem Setup

\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\Categorical}{\mathrm{Categorical}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
\newcommand{\SBM}{\mathrm{SBM}}
\newcommand{\DCBM}{\mathrm{DCBM}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\MBM}{\mathrm{MBM}}

```{r out.width = '50%', fig.height = 3, fig.width = 4, fig.cap = 'Friendship network of 81 faculty at a UK university (Nepusz et al., 2008). The vertices are labeled by school affiliation.'}
data('UKfaculty', package = 'igraphdata')
A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
A <- sign(A + t(A))
z <- factor(igraph::vertex_attr(UKfaculty)$Group)
n <- length(z)

ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = z), 
                  show.legend = FALSE) + 
  labs(colour = NULL) + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')
```
\begin{center}
How might we cluster the nodes of a network?
\end{center}

## Bernoulli Random Graphs

Let $G$ be an undirected and unweighted graph with $n$ vertices.

$G$ is described by adjacency matrix $A$ such that
$A_{ij} = \begin{cases} 
1 & \text{an edge connects vertices } i \text{ and } j \\
0 & \text{otherwise}
\end{cases}$

$A_{ji} = A_{ij}$ and $A_{ii} = 0$.

\vspace*{1\baselineskip}

$A \sim \BG(P)$ iff:

1. $P$ is a matrix of edge probabilities between pairs of vertices.
2. $A_{ij} \indep \Bernoulli(P_{ij})$ for each $i < j$.

## Block Models

Suppose each vertex $v_1, ..., v_n$ has labels $z_1, ..., z_n \in \{1, ..., K\}$,  
and each $P_{ij}$ depends on labels $z_i$ and $z_j$.  
Then $A \sim \BG(P)$ is a *block model*.

**Example 1**: Stochastic Block Model with $K=2$ communities.

::: columns

:::: column

$P_{ij} = \begin{cases} 
p & z_i = z_j = 1 \\
q & z_i = z_j = 2 \\
r & z_i \neq z_j
\end{cases}$

::::

:::: column

```{r, fig.height = 3, fig.width = 4, out.width = '100%'}
set.seed(123)

n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/4
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
A <- draw.graph(P)
qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
```

::::

:::

## Block Models

Erdos-Renyi Model (1959)

* $P_{ij} = \theta$ (not a block model)
* 1 parameter $\theta$

Stochastic Block Model (Lorrain and White, 1971)

* $P_{ij} = \theta_{z_i z_j}$
* $K (K + 1) / 2$ parameters $\theta_{kl}$

Degree Corrected Block Model (Karrer and Newman, 2011)

* $P_{ij} = \theta_{z_i z_j} \omega_i \omega_j$
* $K (K + 1) / 2 + n$ parameters $\theta_{kl}$, $\omega_i$

Popularity Adjusted Block Model (Sengupta and Chen, 2017)

* $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
* $K n$ parameters $\lambda_{ik}$

## Block Models

Dating network as an SBM: $P_{ij} = \theta_{z_i, z_j}$

* This model is disassortative; $\theta_{k \ell} > \theta_{kk}$ for $k \neq \ell$.
* Each individual has the same probability of matching with every other individual conditioned on gender.

```{r dating-sbm, fig.width = 4, , fig.height = 2, cache = TRUE}
set.seed(123)

n <- 2 ** 6
z <- sample(c(1, 2), n, replace = TRUE)
z <- sort(z)
n1 <- sum(z == 1)
n2 <- sum(z == 2)
z[z == 1] <- 'F'
z[z == 2] <- 'M'
theta11 <- .02
theta22 <- .02
theta12 <- .2
P <- matrix(theta12, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- theta11
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- theta22
A <- draw.graph(P)
# qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
ggraph(A, layout = 'auto') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)), 
                  show.legend = TRUE) + 
  labs(colour = 'Gender') + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')
```

## Block Models

Dating network as a DCBM: $P_{ij} = \omega_i \omega_j \theta_{z_i, z_j}$

* Models each individual's popularity---some individuals form more connections than others.
* Still assumes that all members of each gender have the same odds of matching between two genders.

```{r dating-dcbm, fig.width = 4, , fig.height = 2, cache = TRUE}
set.seed(12345)

n <- 2 ** 6
z <- sample(c(1, 2), n, replace = TRUE)
z <- sort(z)
n1 <- sum(z == 1)
n2 <- sum(z == 2)
z[z == 1] <- 'F'
z[z == 2] <- 'M'
theta11 <- .03
theta22 <- .03
theta12 <- .3
omega <- runif(n, 1/3, 1)
Omega <- omega %*% t(omega)
P <- matrix(theta12, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- theta11
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- theta22
P <- P * Omega
A <- draw.graph(P)
ggraph(A, layout = 'auto') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)), 
                  show.legend = TRUE) + 
  labs(colour = 'Gender') + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')
```

## Block Models

Dating network as a PABM: $P_{ij} = \lambda_{i, z_j} \lambda_{j, z_i}$

* Models each user's sexual orientation as well as popularity. 
* Some individuals are more likely to connect with members of the same gender while others are more likely to connect with members of the opposite gender.

```{r dating-pabm, fig.width = 4, , fig.height = 2, cache = TRUE}
set.seed(12345)
n <- 2 ** 6
Az <- draw.pabm.beta.2(n, 1, 6, 2, 2, .5)
A <- Az$A
z <- Az$z
z[z == 1] <- 'F'
z[z == 2] <- 'M'

ggraph(A, layout = 'auto') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)), 
                  show.legend = TRUE) + 
  labs(colour = 'Gender') + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')
```

## Hierarchy of Block Models

PABM $\to$ DCBM: $\lambda_{ik} = \sqrt{\theta_{z_i k}} \omega_i$

DCBM $\to$ SBM: $\omega_i = 1$

SBM $\to$ Erdos-Renyi: $\theta_{kl} = \theta$

\begin{center}
```{r, out.width = '125px'}
knitr::include_graphics('../proposal/noroozi-pensky-hierarchy.png')
```

\tiny{Noroozi and Pensky, 2021}
\end{center}

## Estimation for Block Models

Log-likelihood function for the SBM:

$$\ell(z, \theta; A) = \sum_{i<j} \sum_{k, \ell} z_{ij} z_{j \ell} \Big(A_{ij} \log_{k \ell} + (1 - A_{ij}) \log(1 - \theta_{k \ell}) \Big)$$

* Direct maximization is NP-hard.

* Expectation-Maximization requires an independence relaxation.

## Estimation for Block Models

\begin{algorithm}[H]
  \label{algo:sbm-em}
  \scriptsize
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$}
  \caption{Approximate EM algorithm for the SBM}
  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \hat{\theta}_{k \ell} + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell})) \bigg)$.\;
      }
    }
    \For {$k = 1, ..., K$} {
      \For {$\ell = 1, ..., K$} {
        M-step: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
      }
    }
  }
\end{algorithm}

## Generalized Random Dot Product Graphs

**Def** Generalized Random Dot Product Graph  
(Rubin-Delanchy, Cape, Tang, Priebe, 2022)

Let $I_{p,q} = \blockdiag(I_p, -I_q)$ and suppose that $x_1, \ldots, x_n \in \mathbb{R}^{p+q}$ are such that $x_i^\top I_{p,q} x_j \in [0,1]$.

Then $A \sim \GRDPG_{p, q}(X)$ iff $A \sim \BG(X I_{p,q} X^\top)$, where $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.

```{r rdpg-example, fig.length = 6, fig.height = 2, out.width = '100%', cache = TRUE, fig.cap = 'Latent vectors (left) and GRDPG (right).'}
set.seed(123)

n <- 200
t. <- runif(n)
x <- t. ^ 2
y <- 2 * t. * (1 - t.)

latent.vec.plot <- ggplot() + 
  geom_point(aes(x = x, y = y, colour = t.)) + 
  coord_fixed() + 
  theme_bw() + 
  labs(x = NULL, y = NULL) + 
  theme(legend.position = 'none') + 
  viridis::scale_colour_viridis()

X <- cbind(x, y)
P <- X %*% t(X)
A <- draw.graph(P)

rdpg.plot <- ggraph(A, layout = 'auto') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = t.), 
                  show.legend = FALSE) + 
  theme_void() + 
  viridis::scale_colour_viridis()

gridExtra::grid.arrange(latent.vec.plot, rdpg.plot, nrow = 1)
```

## Generalized Random Dot Product Graphs

Adjacency Spectral Embedding (Sussman et al., 2012) estimates $x_1, ..., x_n \in \mathbb{R}^{p+q}$ from $A$:

1. Let $\hat{\Lambda}$ be the diagonal matrix that contains the absolute values of the $p$ most positive and the $q$ most negative eigenvalues.
2. Let $\hat{V}$ be the matrix whose columns are the corresponding eigenvectors.
3. Compute $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$.

**Theorem**: $\max\limits_i \|\hat{X}_i - Q_n X_i \| = O_P \Big( \frac{(\log n)^c}{n^{1/2}} \Big)$ 
as $n \to \infty$

```{r ase-example, fig.length = 8, fig.height = 2, out.width = '100%', cache = TRUE, fig.cap = 'Latent vectors (left), GRDPG (center), and ASE (right).'}
set.seed(123)

n <- 200
t. <- runif(n)
x <- t. ^ 2
y <- 2 * t. * (1 - t.)

latent.vec.plot <- ggplot() + 
  geom_point(aes(x = x, y = y, colour = t.)) + 
  coord_fixed() + 
  theme_bw() + 
  labs(x = NULL, y = NULL) + 
  theme(legend.position = 'none') + 
  viridis::scale_colour_viridis()

X <- cbind(x, y)
P <- X %*% t(X)
A <- draw.graph(P)

rdpg.plot <- ggraph(A, layout = 'auto') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = t.), 
                  show.legend = FALSE) + 
  theme_void() + 
  viridis::scale_colour_viridis()

Y <- embedding(P, 2, 0)
Xhat <- embedding(A, 2, 0)
Xhat[, 1] <- -Xhat[, 1]
ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = t.)) + 
  geom_line(aes(x = Y[, 1], y = Y[, 2]), colour = 'red') + 
  coord_fixed() + 
  theme_bw() + 
  labs(x = NULL, y = NULL) + 
  theme(legend.position = 'none') + 
  viridis::scale_colour_viridis()

gridExtra::grid.arrange(latent.vec.plot, rdpg.plot, ase.plot, nrow = 1)
```

## SBMs are GRDPGs

All Bernoulli Graphs are GRDPGs.

**Example 1** (cont'd): SBM with $K = 2$.

::: columns

:::: column

$$P_{ij} = \begin{cases} 
p & z_i = z_j = 1 \\
q & z_i = z_j = 2 \\
r & z_i \neq z_j
\end{cases}$$

```{r fig.height = 3, fig.width = 6, out.width = '100%'}
set.seed(123)

n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/4
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
A <- draw.graph(P)
qgraph::qgraph(A, vsize = 4, groups = factor(z), legend = FALSE)
```

::::

:::: column

$$P = 
\begin{bmatrix} 
P^{(11)} & P^{(12)} \\
P^{(21)} & P^{(22)}
\end{bmatrix} =
X I_{2,0} X^\top$$

$$X = \begin{bmatrix} 
\sqrt{p} & 0 \\
\vdots & \vdots \\
\sqrt{p} & 0 \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p} \\
\vdots & \vdots \\
\sqrt{r^2 / p} & \sqrt{q - r^2 / p}
\end{bmatrix}$$

::::

:::

## SBMs are GRDPGs

**Example 1** (cont'd): To perform community detection,

1. Note that $A$ is a GRDPG because $P = X I_{2, 0} X^\top$.
2. Compute the ASE $A \approx \hat{X} \hat{X}^\top$ with 
$\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$.
3. Apply a clustering algorithm (e.g., $K$-means) to $\hat{X}$,  
noting that $\hat{X}$ approaches point masses as $n \to \infty$.

```{r, fig.height = 5, fig.width = 5, out.width = '50%', fig.cap = 'ASE of the adjacency matrix drawn from an SBM.'}
P.eigen = eigen(P)
X <- P.eigen$vectors[, 1:2] %*% diag(P.eigen$values[1:2] ** .5)

A.eigen <- eigen(A)
X.hat <- A.eigen$vectors[, 1:2] %*% diag(A.eigen$values[1:2] ** .5)
X.hat[, 2] <- -X.hat[, 2]
plot(X.hat, asp = 1, col = z * 2, xlab = NA, ylab = NA)
points(X, pch = 16, col = z * 2)
```

## DCBMs are GRDPGs

**Example 2**

* Convert example 1 to a DCBM by $P_{ij} \leftarrow \omega_i \omega_j P_{ij}$.
* This is equivalent to $P \leftarrow \Omega P \Omega$ where $\Omega = \diag(\omega_1, ..., \omega_n)$. 
* This is equivalent to left-multiplying the matrix of latent vectors $X$ by $\Omega$.
* The point masses in the latent structure of the SBM are "stretched out" toward the origin to form rays.

$$X = \begin{bmatrix}
\omega_1 \sqrt{\theta_{11}} & 0 \\ 
\vdots & \vdots \\ 
\omega_{n_1} \sqrt{\theta_{11}} & 0 \\ 
\omega_{n_1 + 1} \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_{n_1 + 1} \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
\vdots & \vdots \\
\omega_n \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_n \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
\end{bmatrix}$$

## DCBMs are GRDPGs

**Example 2** (cont'd): To perform community detection,

1. Note that $A$ is a GRDPG because $P = X I_{2, 0} X^\top$.
2. Compute the ASE $A \approx \hat{X} \hat{X}^\top$ with 
$\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$.
3. Apply a clustering algorithm (e.g., kernel $K$-means with the cosine kernel) to $\hat{X}$, noting that $\hat{X}$ approaches rays as $n \to \infty$.

```{r fig.height = 2, fig.width = 6, out.width = '100%', fig.cap = 'DCBM (left) and its ASE (right).'}
set.seed(123)

n1 <- 2 ** 7
n2 <- 2 ** 7
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/4
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q

omega <- runif(n)
P <- diag(omega) %*% P %*% diag(omega)
A <- draw.graph(P)

dcbm.plot <- ggraph(A, layout = 'auto') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)), 
                  show.legend = FALSE) + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

Xhat <- embedding(A, 2, 0)
Y <- embedding(P, 2, 0)

ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1], 
                 y = Xhat[, 2], 
                 colour = factor(z)),
             size = .25) + 
  scale_colour_brewer(palette = 'Set1') + 
  labs(x = NULL, y = NULL) + 
  theme(legend.position = 'none') + 
  coord_fixed()

gridExtra::grid.arrange(dcbm.plot, ase.plot, nrow = 1)
```

## Example: UK Faculty Network

* Treating graphs with community structure as a GRDPG is useful for both EDA and inference. 
* The ASE of the UK faculty network (Nepusz et al., 2008) reveals three communities and that a DCBM is an appropriate model. 
* Clustering on the ASE results in low community detection error.

```{r, fig.width = 6, fig.height = 2, out.width = '100%', fig.cap = 'UK faculty network (left) and its ASE (right).'}
data(UKfaculty, package = 'igraphdata')
A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
A <- sign(A + t(A))
z <- factor(igraph::vertex_attr(UKfaculty)$Group)
A <- A[z != 4, z != 4]
z <- z[z != 4]
Xhat <- embedding(A, 3, 0)
n <- length(z)

graph.plot <- ggraph(A, layout = 'auto') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)), 
                  show.legend = FALSE) + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 2], 
                 y = Xhat[, 3], 
                 colour = factor(z)),
             size = .5) + 
  scale_colour_brewer(palette = 'Set1') + 
  labs(x = 'comp 2', y = 'comp 3') + 
  theme(legend.position = 'none') + 
  coord_fixed()

gridExtra::grid.arrange(graph.plot, ase.plot, nrow = 1)
```

# PABMs are GRDPGs

## Popularity Vectors

**Lemma** (Noroozi, Rimal, and Pensky, 2020):

$A$ is sampled from a PABM if $P$ can be described as:

1. Let each $P^{(kl)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
between communities $k$ and $l$. 
2. Organize popularity parameters as vectors 
$\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
such that $\lambda^{(kl)}_i = \lambda_{k_i l}$ is the popularity parameter 
of the $i$^th^ vertex of community $k$ towards community $l$. 
3. Each block can be decomposed as 
$P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$.

This result reveals that $P$ is rank $K^2$ and has a linear latent structure.

## PABMs are GRDPGs

**Theorem** (KTT): $A \sim \PABM(\{\lambda_{ik}\}_K)$ is equivalent to 
$A \sim \GRDPG_{p, q}(X U)$ with

* $p = K (K + 1) / 2$, $q = K (K - 1) / 2$;
* $U$ is an orthogonal matrix;
* $X \in \mathbb{R}^{n \times K^2}$ is a block diagonal matrix 
composed of popularity vectors with each block corresponding to a community.

$$X = \begin{bmatrix}
\Lambda^{(1)} & \cdots & 0 \\
0 & \ddots & 0 \\
0 & \cdots & \Lambda^{(K)}
\end{bmatrix} 
\in \mathbb{R}^{n \times K^2}$$

$$\Lambda^{(k)} = \begin{bmatrix} 
\lambda^{(k1)} & \cdots & \lambda^{(kK)} 
\end{bmatrix} 
\in \mathbb{R}^{n_k \times K}$$

$$A \sim \PABM(\{\lambda_{ik}\}_K) \text{ iff } A \sim \GRDPG_{p, q}(X U)$$

## PABMs are GRDPGs

::: columns

:::: column

$$X = \begin{bmatrix}
\Lambda^{(1)} & \cdots & 0 \\
0 & \ddots & 0 \\
0 & \cdots & \Lambda^{(K)}
\end{bmatrix}$$

::::

:::: column

\vspace*{2\baselineskip}

$$U \in \mathbb{O}(K^2)$$

::::

:::

$$A \sim \PABM(\{\lambda^{(kl)}\}_K) \iff A \sim \GRDPG_{p, q}(X U)$$

**Remark 1** (orthogonality of subspaces):
If $y_i^\top$ and $y_j^\top$ are two rows of $X U$ corresponding 
to different communities, then $y_i^\top y_j = 0$.

**Remark 2** (non-uniqueness of the latent configuration):  
If $A \sim \GRDPG_{p, q}(Y)$, then $A \sim \GRDPG_{p, q}(Y Q)$ for any $Q$ in 
the indefinite orthogonal group with signature $p, q$.

**Remark 3**: Communities correspond to subspaces even with linear transformation $Q \in \mathbb{O}(p, q)$, but this may break the orthogonality property.

## Orthogonal Spectral Clustering

**Theorem** (KTT): 
If $P = V \Lambda V^\top$ and $B = n V V^\top$,  
then $B_{ij} = 0$ if $z_i \neq z_j$.

**Algorithm**: Orthogonal Spectral Clustering:

1. Let $V$ be the eigenvectors of $A$ corresponding to the $K (K+1)/2$ most 
positive and $K (K-1) / 2$ most negative eigenvalues.
2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
3. Construct graph $G$ using $B$ as its similarity matrix.
4. Partition $G$ into $K$ disconnected subgraphs.

**Theorem** (KTT): 
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix from OSC. 
Then $\forall$ pairs $(i, j)$ belonging to different communities 
and sparsity factor satisfying $n \rho_n = \omega\big((\log n)^{4c}\big)$,  
$\max_{i, j} \hat{B}_{ij} = 
O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)$ 
as $n \to \infty$.

**Corollary**: OSC results in zero clustering error as $n \to \infty$, with probability 1.

## Sparse Subspace Clustering

**Corollary**: The ASE of $A \sim \PABM(\{\lambda^{(kl)}\}_K)$ lies near a collection of $K$-dimensional subspaces in $K^2$ dimensions.

**Algorithm**: Sparse Subspace Clustering (Elhamifar & Vidal, 2009):

1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
subject to $x_i = X^\top c$ and $c^{(i)} = 0$.  
This is typically performed via LASSO: 
$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i}^\top c\|_2^2 + \lambda \|c\|_1$
2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
3. Construct affinity matrix $B = |C| + |C^\top|$.

## Sparse Subspace Clustering

Noroozi et al. observed that the rank of $P$ is $K^2$ 
and the columns of $P$ belonging to each community has rank $K$ 
to justify SSC for the PABM.

$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } A_{\cdot, i} = A c
\text{ and } c^{(i)} = 0$$

They were able to show that this obeys SDP if we replace $A$ with $P$.

GRDPG-based approach: Apply SSC to the ASE of $A$.

Stronger result: Apply SSC to the eigenvectors of $A$.

$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } 
\hat{v}_i = \hat{V}^\top c \text{ and } c^{(i)} = 0$$
$$A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$$

## Sparse Subspace Clustering

**Theorem** (KTT):

Let

* $P_n$ describe the edge probability matrix of the PABM with $n$ vertices, and 
$A_n \sim \BG(P_n)$;
* $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
$K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues.

Then

* For some $\lambda > 0$ and $N < \infty$, $\sqrt{n} \hat{V}_n$ obeys the Subspace Detection Property with probability 1 when $n > N$.

Remarks:

* For large $n$, we can identify $\lambda$ for SDP (Wang and Xu, 2016).
* SDP does not guarantee community detection.

## Simulation Results

1. $z_1, ..., z_n \iid \Categorical(1 / K, ..., 1 / K)$
2. $\lambda_{ik} \stackrel{iid}{\sim} \Betadist(a_{ik}, b_{ik})$  
$a_{ik} = \begin{cases} 2 & z_i = k \\ 1 & z_i \neq k \end{cases}$, 
$b_{ik} = \begin{cases} 1 & z_i = k \\ 2 & z_i \neq k \end{cases}$
3. $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
4. $A \sim \BG(P)$

```{r, fig.height = 2, fig.width = 8, out.width = '100%'}
clustering.df <- readr::read_csv('~/dev/pabm-grdpg/simulation-results/cluster-sim-balanced.csv')

clustering.df %>%
  dplyr::group_by(n, K) %>%
  dplyr::summarise(
    med.err = median(error.osc),
    first.q = quantile(error.osc, .25),
    third.q = quantile(error.osc, .75),
    med.err.ssc = median(error.ssc.ase),
    first.q.ssc = quantile(error.ssc.ase, .25),
    third.q.ssc = quantile(error.ssc.ase, .75),
    med.err.louvain = median(error.louvain),
    first.q.louvain = quantile(error.louvain, .25),
    third.q.louvain = quantile(error.louvain, .75),
    med.err.ssc.A = median(error.ssc.A),
    first.q.ssc.A = quantile(error.ssc.A, .25),
    third.q.ssc.A = quantile(error.ssc.A, .75)
  ) %>% 
  dplyr::ungroup() %>% 
  ggplot() +
  theme_bw() + 
  theme(legend.position = 'right',
        legend.direction = 'vertical',
        legend.justification = 0) + 
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096),
                labels = c(7, 8, 9, 10, 11, 12)) +
  scale_y_log10() + 
  labs(y = 'error count', 
       x = expression(log[2] (n)),
       colour = NULL, shape = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Orthogonal Spectral Clustering')) +
  geom_point(aes(x = n, y = med.err * n,
                 colour = 'Orthogonal Spectral Clustering', 
                 shape = 'Orthogonal Spectral Clustering')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Orthogonal Spectral Clustering'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'Sparse Subspace Clustering on ASE')) +
  geom_point(aes(x = n, y = med.err.ssc * n,
                 colour = 'Sparse Subspace Clustering on ASE', 
                 shape = 'Sparse Subspace Clustering on ASE')) +
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'Sparse Subspace Clustering on ASE'), width = .1) +
  geom_line(aes(x = n, y = med.err.ssc.A * n,
                colour = 'Sparse Subspace Clustering on Adj. Matrix')) +
  geom_point(aes(x = n, y = med.err.ssc.A * n,
                 colour = 'Sparse Subspace Clustering on Adj. Matrix', 
                 shape = 'Sparse Subspace Clustering on Adj. Matrix')) +
  geom_errorbar(aes(x = n, ymin = first.q.ssc.A * n, ymax = third.q.ssc.A * n,
                    colour = 'Sparse Subspace Clustering on Adj. Matrix'), 
                width = .1) +
  geom_line(aes(x = n, y = med.err.louvain * n,
                colour = 'Modularity Maximization')) + 
  geom_point(aes(x = n, y = med.err.louvain * n,
                 colour = 'Modularity Maximization', 
                 shape = 'Modularity Maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.louvain * n, ymax = third.q.louvain * n,
                    colour = 'Modularity Maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both')
```

## Simulation Results

1. $z_1, ..., z_n \iid \Categorical(\alpha_1, ..., \alpha_K)$ for $\alpha_k = \frac{k^{-1}}{\sum_{\ell=1}^K \ell^{-1}}$
2. $\lambda_{ik} \stackrel{iid}{\sim} \Betadist(a_{ik}, b_{ik})$  
$a_{ik} = \begin{cases} 2 & z_i = k \\ 1 & z_i \neq k \end{cases}$, 
$b_{ik} = \begin{cases} 1 & z_i = k \\ 2 & z_i \neq k \end{cases}$
3. $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
4. $A \sim \BG(P)$

```{r, fig.height = 2, fig.width = 8, out.width = '100%'}
clustering.df <- readr::read_csv('~/dev/pabm-grdpg/simulation-results/cluster-sim-imbalanced.csv')

clustering.df %>%
  dplyr::group_by(n, K) %>%
  dplyr::summarise(
    med.err = median(error.osc),
    first.q = quantile(error.osc, .25),
    third.q = quantile(error.osc, .75),
    med.err.ssc = median(error.ssc.ase),
    first.q.ssc = quantile(error.ssc.ase, .25),
    third.q.ssc = quantile(error.ssc.ase, .75),
    med.err.louvain = median(error.louvain),
    first.q.louvain = quantile(error.louvain, .25),
    third.q.louvain = quantile(error.louvain, .75),
    med.err.ssc.A = median(error.ssc.A),
    first.q.ssc.A = quantile(error.ssc.A, .25),
    third.q.ssc.A = quantile(error.ssc.A, .75)
  ) %>% 
  dplyr::ungroup() %>% 
  ggplot() +
  theme_bw() + 
  theme(legend.position = 'right',
        legend.direction = 'vertical',
        legend.justification = 0) + 
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096),
                labels = c(7, 8, 9, 10, 11, 12)) +
  scale_y_log10() + 
  labs(y = 'error count', 
       x = expression(log[2] (n)),
       colour = NULL, shape = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Orthogonal Spectral Clustering')) +
  geom_point(aes(x = n, y = med.err * n,
                 colour = 'Orthogonal Spectral Clustering', 
                 shape = 'Orthogonal Spectral Clustering')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Orthogonal Spectral Clustering'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'Sparse Subspace Clustering on ASE')) +
  geom_point(aes(x = n, y = med.err.ssc * n,
                 colour = 'Sparse Subspace Clustering on ASE', 
                 shape = 'Sparse Subspace Clustering on ASE')) +
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'Sparse Subspace Clustering on ASE'), width = .1) +
  geom_line(aes(x = n, y = med.err.ssc.A * n,
                colour = 'Sparse Subspace Clustering on Adj. Matrix')) +
  geom_point(aes(x = n, y = med.err.ssc.A * n,
                 colour = 'Sparse Subspace Clustering on Adj. Matrix', 
                 shape = 'Sparse Subspace Clustering on Adj. Matrix')) +
  geom_errorbar(aes(x = n, ymin = first.q.ssc.A * n, ymax = third.q.ssc.A * n,
                    colour = 'Sparse Subspace Clustering on Adj. Matrix'), 
                width = .1) +
  geom_line(aes(x = n, y = med.err.louvain * n,
                colour = 'Modularity Maximization')) + 
  geom_point(aes(x = n, y = med.err.louvain * n,
                 colour = 'Modularity Maximization', 
                 shape = 'Modularity Maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.louvain * n, ymax = third.q.louvain * n,
                    colour = 'Modularity Maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both')
```

## Simulation Results

1. $z_1, ..., z_n \iid \Categorical(1 / K, ..., 1 / K)$
2. $\lambda_{ik} \stackrel{iid}{\sim} \Betadist(a_{ik}, b_{ik})$  
$a_{ik} = \begin{cases} 1 & z_i = k \\ 2 & z_i \neq k \end{cases}$, 
$b_{ik} = \begin{cases} 2 & z_i = k \\ 1 & z_i \neq k \end{cases}$
3. $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
4. $A \sim \BG(P)$

```{r, fig.height = 2, fig.width = 8, out.width = '100%'}
clustering.df <- readr::read_csv('~/dev/pabm-grdpg/simulation-results/cluster-sim-disassortative.csv')

clustering.df %>%
  dplyr::group_by(n, K) %>%
  dplyr::summarise(
    med.err = median(error.osc),
    first.q = quantile(error.osc, .25),
    third.q = quantile(error.osc, .75),
    med.err.ssc = median(error.ssc.ase),
    first.q.ssc = quantile(error.ssc.ase, .25),
    third.q.ssc = quantile(error.ssc.ase, .75),
    med.err.louvain = median(error.louvain),
    first.q.louvain = quantile(error.louvain, .25),
    third.q.louvain = quantile(error.louvain, .75),
    med.err.ssc.A = median(error.ssc.A),
    first.q.ssc.A = quantile(error.ssc.A, .25),
    third.q.ssc.A = quantile(error.ssc.A, .75)
  ) %>% 
  dplyr::ungroup() %>% 
  ggplot() +
  theme_bw() + 
  theme(legend.position = 'right',
        legend.direction = 'vertical',
        legend.justification = 0) + 
  scale_x_log10(breaks = c(128, 256, 512, 1024, 2048, 4096),
                labels = c(7, 8, 9, 10, 11, 12)) +
  scale_y_log10() + 
  labs(y = 'error count', 
       x = expression(log[2] (n)),
       colour = NULL, shape = NULL) +
  geom_line(aes(x = n, y = med.err * n,
                colour = 'Orthogonal Spectral Clustering')) +
  geom_point(aes(x = n, y = med.err * n,
                 colour = 'Orthogonal Spectral Clustering', 
                 shape = 'Orthogonal Spectral Clustering')) +
  geom_errorbar(aes(x = n, ymin = first.q * n, ymax = third.q * n,
                    colour = 'Orthogonal Spectral Clustering'), width = .1) + 
  geom_line(aes(x = n, y = med.err.ssc * n,
                colour = 'Sparse Subspace Clustering on ASE')) +
  geom_point(aes(x = n, y = med.err.ssc * n,
                 colour = 'Sparse Subspace Clustering on ASE', 
                 shape = 'Sparse Subspace Clustering on ASE')) +
  geom_errorbar(aes(x = n, ymin = first.q.ssc * n, ymax = third.q.ssc * n,
                    colour = 'Sparse Subspace Clustering on ASE'), width = .1) +
  geom_line(aes(x = n, y = med.err.ssc.A * n,
                colour = 'Sparse Subspace Clustering on Adj. Matrix')) +
  geom_point(aes(x = n, y = med.err.ssc.A * n,
                 colour = 'Sparse Subspace Clustering on Adj. Matrix', 
                 shape = 'Sparse Subspace Clustering on Adj. Matrix')) +
  geom_errorbar(aes(x = n, ymin = first.q.ssc.A * n, ymax = third.q.ssc.A * n,
                    colour = 'Sparse Subspace Clustering on Adj. Matrix'), 
                width = .1) +
  geom_line(aes(x = n, y = med.err.louvain * n,
                colour = 'Modularity Maximization')) + 
  geom_point(aes(x = n, y = med.err.louvain * n,
                 colour = 'Modularity Maximization', 
                 shape = 'Modularity Maximization')) + 
  geom_errorbar(aes(x = n, ymin = first.q.louvain * n, ymax = third.q.louvain * n,
                    colour = 'Modularity Maximization'), width = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ K, labeller = 'label_both')
```

## Example

\begin{figure}[H]
{\centering \includegraphics[scale=.33]{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/mp-1}
}
\caption{Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by OSC.}\label{fig:mp}
\end{figure}

\begin{table}[H]
\centering
\begin{tabular}[t]{l|r|r|r}
\hline
Network & MM & SSC-ASE & OSC\\
\hline
British MPs & 0.003 & 0.012 & 0.006\\
\hline
Political Blogs & 0.050 & 0.187 & 0.062\\
\hline
DBLP & 0.028 & 0.072 & 0.059\\
\hline
\end{tabular}
\end{table}

## Example: Karantaka Villages

\begin{figure}[H]
{\centering \includegraphics[scale=.33]{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/unnamed-chunk-7-1}
}
\caption{Adjacency matrix of the Karnataka villages data, arranged by the clustering produced by OSC (left). The villages studied here are, from left to right, 12, 31, and 46.}\label{fig:households-figure}
\end{figure}

\begin{table}[H]
\centering
\begin{tabular}[t]{l|r|r|r}
\hline
Network & MM & SSC-ASE & OSC\\
\hline
Village 12 & 0.270 & 0.291 & 0.227\\
\hline
Village 31 & 0.125 & 0.059 & 0.051\\
\hline
Village 46 & 0.052 & 0.069 & 0.056\\
\hline
\end{tabular}
\end{table}

## Overview of Block Models as GRDPGs

::: columns

:::: {.column width='67%'}

```{r, cache = TRUE, fig.width = 6, fig.height = 2, out.width = '100%'}
set.seed(123)

n <- 64
n1 <- n / 2
n2 <- n / 2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/2
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
X <- embedding(P, 2, 0)
A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)
A.graph <- tidygraph::as_tbl_graph(A, directed = FALSE)

sbm.graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)),
                  show.legend = FALSE) + 
  labs(colour = NULL, 
       title = 'SBM') + 
  # theme_graph() + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

sbm.ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 2], y = Xhat[, 1],
                 colour = factor(z)),
             shape = 4) + 
  geom_point(aes(x = X[, 2], y = X[, 1], 
                 colour = factor(z))) + 
  # coord_fixed() +
  theme_bw() + 
  theme(legend.position = 'none') + 
  labs(x = NULL, y = NULL, colour = NULL, 
       title = 'Point Masses') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(sbm.graph.plot, sbm.ase.plot, nrow = 1)
```

```{r, cache = TRUE, fig.width = 6, fig.height = 2, out.width = '100%'}
set.seed(123)

n <- 128
n1 <- n / 2
n2 <- n / 2
z <- c(rep(1, n1), rep(2, n2))
p <- 1/2
q <- 1/2
r <- 1/8
P <- matrix(r, nrow = n, ncol = n)
P[seq(n1), seq(n1)] <- p
P[seq(n1 + 1, n), seq(n1 + 1, n)] <- q
omega <- rbeta(n, 2, 1)
P <- P * (omega %*% t(omega))
X <- embedding(P, 2, 0)
A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)
Xhat[, 2] <- -Xhat[, 2]
A.graph <- tidygraph::as_tbl_graph(A, directed = FALSE)

dcbm.graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)),
                  show.legend = FALSE) + 
  labs(colour = NULL, 
       title = 'DCBM') + 
  # theme_graph() + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

dcbm.ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2],
                 colour = factor(z)),
             shape = 4) + 
  geom_point(aes(x = X[, 1], y = X[, 2], 
                 colour = factor(z))) + 
  # coord_fixed() +
  theme_bw() + 
  theme(legend.position = 'none') + 
  labs(x = NULL, y = NULL, colour = NULL, 
       title = 'Rays') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(dcbm.graph.plot, dcbm.ase.plot, nrow = 1)
```

```{r, cache = TRUE, fig.width = 6, fig.height = 2, out.width = '100%'}
set.seed(123)

n <- 64
Pz <- generate.P.beta(n)
P <- Pz$P
z <- Pz$clustering
X <- embedding(P, 3, 1)
A <- draw.graph(P)
Xhat <- embedding(A, 3, 1)
# Xhat[, 1] <- -Xhat[, 1]
A.graph <- tidygraph::as_tbl_graph(A, directed = FALSE)

pabm.graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)),
                  show.legend = FALSE) + 
  labs(colour = NULL, 
       title = 'PABM') + 
  # theme_graph() + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

pabm.ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2],
                 colour = factor(z)),
             shape = 4) + 
  geom_point(aes(x = X[, 1], y = X[, 2], 
                 colour = factor(z))) + 
  # coord_fixed() +
  theme_bw() + 
  theme(legend.position = 'none') + 
  labs(x = NULL, y = NULL, colour = NULL, 
       title = 'Subspaces (Projected)') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(pabm.graph.plot, pabm.ase.plot, nrow = 1)
```

::::

:::: {.column width='33%'}

&nbsp;

&nbsp;

* K-means clustering
* Gaussian mixture models

&nbsp;

* K-means with cosine similarity
* GMM on angles

&nbsp;

&nbsp;

* Orthogonal spectral clustering
* Sparse subspace clustering

::::

:::

# Manifold Block Models

## Motivating Example

Macaque visuotactile brain areas and connections network  
(Negyessy et al., 2006)

* ASE suggests community structure.
* Structure is nonlinear---$K$-means, OSC, SSC, etc. are not appropriate for these data.

```{r macaque-graph, fig.cap = 'Macaque visuotactile brain areas and connections network (left) and ASE (right). The vertices represent brain areas.', fig.height = 2, fig.width = 6, out.width = '100%'}
data(macaque, package = 'igraphdata')
A <- as.matrix(igraph::as_adjacency_matrix(macaque, type = 'both'))
A <- sign(A + t(A))
z <- as.numeric(factor(igraph::vertex_attr(macaque)$shape))

graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)), 
                  show.legend = FALSE) + 
  labs(colour = NULL) + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

Xhat <- embedding(A, 3, 0)

ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1],
                 y = Xhat[, 2],
                 colour = factor(z))) + 
  labs(x = NULL, y = NULL) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(graph.plot, ase.plot, nrow = 1)
```

## Manifold Block Model

Let $p, q \geq 0$, $d = p + q \geq 1$, $1 \leq r < d$, $K \geq 2$, and $n > K$ be integers. 
Define manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \in \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1]\}$ each by continuous function $g_k : [0, 1]^r \to \mathcal{X}$. 
Define probability distributions $F_1, ..., F_K$ each with support $[0, 1]^r$. 
Then the following mixture model is a *manifold block model*:

\begin{enumerate}
\item Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$.
\item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
\item Compile the latent vectors into data matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X)$. 
\end{enumerate}

## $K$-Curves Clustering

\begin{algorithm}[H]
\label{alg:kcurves}
\scriptsize
\DontPrintSemicolon
\SetAlgoLined
\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$, $q$, stopping criterion $\epsilon$}
\KwResult{Community assignments $1, ..., K$, curves $g_1, ..., g_K$}
Compute $X$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
Initialize community labels $z_1, ..., z_n$.\;
\Repeat {the change in $\sum_k \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2$ is less than $\epsilon$} {
\For {$k = 1, ..., K$} {
Define $X_k$ as the rows of $X$ for which $z_i = k$.\;
Fit curve $g_k$ and positions $t_{k_i}$ to $X_k$ by minimizing $\sum_{k_i} \|x_{k_i} - g_k(t_{k_i})\|^2$.\;
}
\For {$k = 1, ..., K$} {
Assign $z_i \leftarrow \arg\min_\ell \|x_i - g_\ell(t_i)\|^2$.\;
}
}
\caption{$K$-curves clustering.}
\end{algorithm}

## $K$-Curves Clustering

```{r multifit, cache = TRUE, fig.height = 3, fig.width = 8, fig.cap = '$K$-curves clustering fits on the ASE using various approaches. Left: quadratic fit with random initialization. Middle: quadratic fit initialized via spectral clustering. Right: cubic fit initialized via spectral clustering.', out.width = '100%'}
f1 <- function(t) {
  x1 <- t ^ 2
  x2 <- 2 * t - 2 * t ^ 2
  return(cbind(x1, x2))
}

f2 <- function(t) {
  x1 <- 2 * t - 2 * t ^ 2
  x2 <- 1 - 2 * t + t ^ 2
  return(cbind(x1, x2))
}

set.seed(314159)

n1 <- 2 ** 7
n2 <- n1
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))

a <- 1
b <- 1
t1 <- rbeta(n1, a, b)
t2 <- rbeta(n2, a, b)
T <- rbind(t1, t2)

X1 <- f1(t1)
X2 <- f2(t2)
X <- rbind(X1, X2)

P <- X %*% t(X)
diag(P) <- 0
A <- draw.graph(P)

Xhat <- embedding(A, 2, 0)

maxit <- 2 ** 4
animation.dir <- '.'
random.starts <- 2
eps <- 1e-6
normalize <- TRUE
animate <- FALSE

set.seed(1)
random <- manifold.clustering(Xhat, 2, 
                              A = A, 
                              parallel = TRUE,
                              intercept = FALSE,
                              maxit = maxit, 
                              normalize = normalize,
                              eps = eps,
                              verbose = FALSE,
                              animate = animate,
                              animation.dir = animation.dir,
                              animation.title = 'random')
random.plot <- plot.estimated.curves(Xhat, random) + 
  labs(x = NULL, y = NULL, title = 'random initialization') + 
  scale_colour_brewer(palette = 'Set1')

set.seed(2)
spectral <- manifold.clustering(
  Xhat, 2, 
  A = A, 
  parallel = TRUE, 
  intercept = FALSE, 
  normalize = normalize, 
  maxit = maxit, 
  eps = eps,
  initialization = 'spectral',
  verbose = FALSE,
  animate = animate,
  animation.dir = animation.dir,
  animation.title = 'spectral')
spectral.plot <- plot.estimated.curves(Xhat, spectral) + 
  labs(x = NULL, y = NULL, title = 'spectral initialization') + 
  scale_colour_brewer(palette = 'Set1')

set.seed(3)
cubic <- manifold.clustering(
  Xhat, 2, 
  degree = 3, 
  A = A, 
  parallel = TRUE, 
  intercept = FALSE, 
  normalize = normalize, 
  maxit = maxit, 
  eps = eps,
  initialization = 'spectral',
  verbose = FALSE,
  animate = animate,
  animation.dir = animation.dir,
  animation.title = 'cubic')
cubic.plot <- plot.estimated.curves(Xhat, cubic) + 
  labs(x = NULL, y = NULL, title = 'cubic fit') + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(random.plot, spectral.plot, cubic.plot, nrow = 1)
```

```{r loss, cache = TRUE, fig.cap = 'Clustering loss vs. iteration for each run of $K$-curves clustering.', fig.width = 8, fig.height = 2, out.width = '100%'}
loss.df <- dplyr::bind_rows(
  dplyr::tibble(loss = random$loss, 
                `initialization and curve` = 'random, quadratic',
                iteration = seq_along(random$loss)),
  dplyr::tibble(loss = spectral$loss, 
                `initialization and curve` = 'spectral, quadratic',
                iteration = seq_along(spectral$loss)),
  dplyr::tibble(loss = cubic$loss, 
                `initialization and curve` = 'spectral, cubic',
                iteration = seq_along(cubic$loss))
) %>% 
  dplyr::mutate(`initialization and curve` = factor(`initialization and curve`,
                                        levels = c('random, quadratic',
                                                   'spectral, quadratic',
                                                   'spectral, cubic')))

ggplot(loss.df) + 
  geom_point(aes(x = iteration, y = loss, colour = `initialization and curve`)) + 
  geom_line(aes(x = iteration, y = loss, colour = `initialization and curve`)) + 
  # ylim(0, max.loss) +
  scale_y_log10() +
  labs(y = 'clustering loss') + 
  theme_bw()
```

## $K$-Curves Clustering

$$L(z_1, ..., z_n, g_1, ..., g_K; X) = \frac{1}{n} \sum_{k=1}^K \sum_{i: z_i = k} \|x_i - g_k(t_i)\|^2$$

**Theorem** (KKT): Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a curve of order $R$ that does not self-intersect (for any $s \neq t$, $g_k(s) \neq g_k(t)$). 
Suppose that for each community $k$, we have labels for at least $R + 1$ vertices. 
Then if $n \rho_n = \omega(\log^{4c} n)$, as $n \to \infty$, the estimates outputted by $K$-curves clustering are such that 
$$L(\hat{z}_1, ..., \hat{z}_n, \hat{g}_1, ..., \hat{g}_K; X) \stackrel{p}{\to} 0.$$

## Simulation Results

1. Draw $z_1, ..., z_n \iid \Multinomial(1/2, 1/2)$.
2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = [ t^2, 2 t (1-t) ]^\top$ and $g_2(t) = [ 2 t (1-t), (1-t)^2 ]^\top$. 
Collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$. 
4. Draw $A \sim \RDPG(X)$ .

```{r sim-curves, fig.height = 3, fig.width = 8, cache = TRUE, out.width = '100%'}
n.vec <- 2 ^ c(7, 8, 9, 10, 11)
nsamp.vec <- c(0, 4, 8)

clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced-2/balanced.csv')

clust.summary.df <- clustering.df %>% 
  dplyr::group_by(n, nsamp) %>% 
  dplyr::summarise(
    med.count = median(error.count),
    first.q.count = quantile(error.count, .25),
    third.q.count = quantile(error.count, .75),
    med.rate = median(error.rate),
    first.q.rate = quantile(error.rate, .25),
    third.q.rate = quantile(error.rate, .75)
  ) %>% 
  dplyr::ungroup()

ggplot(clust.summary.df) + 
  theme_bw() + 
  theme(text = element_text(size = 10)) + 
  scale_y_log10() +
  scale_x_log10(breaks = n.vec) +
  labs(y = 'community detection error rate',
       x = 'n',
       colour = 'known labels\nper community', 
       shape = NULL) + 
  geom_line(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
  geom_point(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
  geom_errorbar(aes(x = n, 
                    ymin = first.q.rate, ymax = third.q.rate, 
                    colour = factor(nsamp)),
                width = .125) + 
  scale_colour_brewer(palette = 'Set1')
```

## Simulation Results

* Same setup as before but fix $n = 512$ and only random initializations.
* Repeat simulation 256 times and look at the distribution of fitted curves and loss values.

```{r repeat-curves, fig.height = 2, fig.width = 8, cache = TRUE, out.width = '100%'}
plot.repeat.study <- function(repeat.df) {
  repeat.df <- na.omit(repeat.df)
  min.t <- 0
  max.t <- 1
  length.out <- 1e2
  t. <- seq(min.t, max.t, length.out = length.out)
  niter <- nrow(repeat.df)
  p.df <- repeat.df[c('p121', 'p122', 'p131', 'p132', 
                      'p221', 'p222', 'p231', 'p232')]
  p.clust <- kmeans(p.df, 4)$cluster
  loss.clust <- kmeans(repeat.df$loss, 2)$cluster
  curves.df <- foreach::foreach(i = seq(niter), .combine = dplyr::bind_rows) %dopar% {
    p1 <- matrix(c(repeat.df$p121[i], repeat.df$p122[i], 
                   repeat.df$p131[i], repeat.df$p132[i]),
                 nrow = 2, byrow = TRUE)
    p1 <- rbind(0, p1)
    p2 <- matrix(c(repeat.df$p221[i], repeat.df$p222[i],
                   repeat.df$p231[i], repeat.df$p232[i]),
                 nrow = 2, byrow = TRUE)
    p2 <- rbind(0, p2)
    p <- list(p1, p2)
    plyr::ldply(seq_along(p), function(k) {
      curve <- bezier.curve(t., p[[k]])
      if (mean(curve[, 1]) < 0) {
        curve[, 1] <- -curve[, 1]
      }
      dplyr::tibble(t = t.,
                    x = curve[, 1],
                    y = curve[, 2],
                    z = k,
                    iter = i,
                    p.clust = p.clust[i],
                    loss.clust = loss.clust[i],
                    loss = repeat.df$loss[i])
    })
  }
  ggplot(curves.df) + 
    geom_path(aes(x = y, y = x, group = interaction(z, iter),
                  colour = loss),
              alpha = 1 / log(niter)) + 
    theme_bw() + 
    coord_fixed() + 
    labs(colour = 'loss',
         x = NULL,
         y = NULL) + 
    viridis::scale_color_viridis()
}

plot.repeat.study(repeat.df)
```

```{r repeat-losses, fig.height = 2, fig.width = 5, cache = TRUE, out.width = '67%'}
repeat.df <- readr::read_csv('~/dev/manifold-block-models/simulations/repeat-study/repeat-study.csv')
ggplot(repeat.df) + 
  geom_histogram(aes(x = loss),
                 colour = 'black', fill = 'white') + 
  labs(x = 'loss values') + 
  theme_bw()
```

## Simulation Results

1. Draw $z_1, ..., z_n \iid \Multinomial(1/3, 1/3, 1/3)$.
2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
3. Let each $x_i = g_{z_i}(t_i)$ where
    i. $g_1(t) = [ 2 t (t-1), t^2, 0 ]^\top$
    ii. $g_2(t) = [ 0, t^2, 2 t (t-1) ]^\top$
    iii. $g_3(t) = [2 t (t-1), t^2, 2 t (t-1) ]^\top$.
4. Collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$.
5. Draw $A \sim \RDPG(X)$.

```{r three-curves, cache = TRUE, fig.height = 4, fig.width = 15, fig.cap = 'Latent vectors (left) and ASE (right) of one realization of the mixture model in the simulation setup with $K=3$ manifolds.', out.width = '100%'}
par(mfrow = c(1, 2))

n <- 512

p.list <- list(matrix(c(0, 1, 0, 
                        0, 0, 1,
                        0, 0, 0),
                      nrow = 3, ncol = 3),
               matrix(c(0, 0, 0, 
                        0, 0, 1,
                        0, 1, 0),
                      nrow = 3, ncol = 3),
               matrix(c(0, 1, 0, 
                        0, 0, 1,
                        0, 1, 0),
                      nrow = 3, ncol = 3))

K <- length(p.list)

z <- sample(seq(K), n, replace = TRUE)
z <- sort(z)

t <- runif(n)

X <- lapply(seq(K), function(k) {
  bezier.curve(t[z == k], 
               p.list[[k]], 
               intercept = TRUE)
}) %>% 
  do.call('rbind', .)

p <- ncol(X)
q <- 0

P <- grdpg.edge.prob.matrix(X, p, q)
A <- draw.graph(P)
Xhat <- embedding(A, p, q)

scatterplot3d::scatterplot3d(X[, 1], X[, 2], X[, 3],
                             color = z, 
                             xlab = NA, ylab = NA, zlab = NA,
                             pch = '.', asp = 1,
                             tick.marks = FALSE)

scatterplot3d::scatterplot3d(Xhat[, 1], Xhat[, 2], Xhat[, 3], 
                             color = z, 
                             xlab = NA, ylab = NA, zlab = NA, 
                             pch = '.', asp = 1, 
                             tick.marks = FALSE,
                             angle = 40)
```

## Simulation Results

1. Draw $z_1, ..., z_n \iid \Multinomial(1/3, 1/3, 1/3)$.
2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
3. Let each $x_i = g_{z_i}(t_i)$ where
    i. $g_1(t) = [ 2 t (t-1), t^2, 0 ]^\top$
    ii. $g_2(t) = [ 0, t^2, 2 t (t-1) ]^\top$
    iii. $g_3(t) = [2 t (t-1), t^2, 2 t (t-1) ]^\top$.
4. Collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$.
5. Draw $A \sim \RDPG(X)$.

```{r sim-curves-3, fig.height = 3, fig.width = 8, out.width = '100%', cache = TRUE}
n.vec <- 2 ^ c(7, 8, 9, 10, 11)
nsamp.vec <- c(0, 4, 8)

clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced-3/balanced-3.csv')

clust.summary.df <- clustering.df %>% 
  dplyr::group_by(n, nsamp) %>% 
  dplyr::summarise(
    med.count = median(error.count),
    first.q.count = quantile(error.count, .25),
    third.q.count = quantile(error.count, .75),
    med.rate = median(error.rate),
    first.q.rate = quantile(error.rate, .25),
    third.q.rate = quantile(error.rate, .75)
  ) %>% 
  dplyr::ungroup()

ggplot(clust.summary.df) + 
  theme_bw() + 
  theme(text = element_text(size = 10)) + 
  scale_y_log10() +
  scale_x_log10(breaks = n.vec) +
  labs(y = 'community detection error rate',
       x = 'n',
       colour = 'known labels\nper community', 
       shape = NULL) + 
  geom_line(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
  geom_point(aes(x = n, y = med.rate, colour = factor(nsamp))) + 
  geom_errorbar(aes(x = n, 
                    ymin = first.q.rate, ymax = third.q.rate, 
                    colour = factor(nsamp)),
                width = .125) + 
  scale_colour_brewer(palette = 'Set1')
```

## Example: Macaque Brain Areas Network

```{r macaque-graph-fit, fig.cap = 'Macaque visuotactile brain areas and connections network (left), ASE (middle), and fitted curves (right). The vertices represent brain areas.', fig.height = 2, fig.width = 8, out.width = '100%', cache = TRUE}
data(macaque, package = 'igraphdata')
A <- as.matrix(igraph::as_adjacency_matrix(macaque, type = 'both'))
A <- sign(A + t(A))
z <- as.numeric(factor(igraph::vertex_attr(macaque)$shape))

graph.plot <- ggraph(A, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
  geom_node_point(aes(colour = factor(z)), 
                  show.legend = FALSE) + 
  labs(colour = NULL) + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

Xhat <- embedding(A, 2, 0)

ase.plot <- ggplot() + 
  geom_point(aes(x = Xhat[, 1],
                 y = Xhat[, 2],
                 colour = factor(z))) + 
  labs(x = NULL, y = NULL) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  scale_colour_brewer(palette = 'Set1') + 
  coord_fixed()

set.seed(123)
z.spectral <- spectral.clustering(A)
macaque.out <- manifold.clustering(Xhat, 2, 
                                   # initialization = 'random',
                                   initialization = z.spectral,
                                   # initialization = z,
                                   intercept = FALSE, 
                                   parallel = TRUE, 
                                   verbose = FALSE, 
                                   animate = FALSE, 
                                   curve.init = 'x')

fit.plot <- plot.estimated.curves(Xhat, macaque.out) + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(graph.plot, ase.plot, fit.plot, nrow = 1)
```

## Example: Drosophila connectome (Eichler et al., 2017)

```{r mbconnectome-graph, fig.height = 2, figh.width = 6, cache = TRUE, out.width = '67%'}
library(mbstructure)
data(MBconnectome)

graph.r <- generate.graph(newrdat, vdf.right)
A.r <- as.matrix(igraph::as_adjacency_matrix(graph.r$g, type = 'both'))
z.r <- as.numeric(graph.r$vdf$type)
n.r <- length(z.r)
A.r <- sign(A.r + t(A.r))
diag(A.r) <- 0
graph.r <- ggraph(A.r, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n.r) / 2) + 
  geom_node_point(aes(colour = factor(z.r)), 
                  show.legend = FALSE) + 
  labs(colour = NULL) + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

graph.l <- generate.graph(newldat, vdf.left)
A.l <- as.matrix(igraph::as_adjacency_matrix(graph.l$g, type = 'both'))
z.l <- as.numeric(graph.l$vdf$type)
n.l <- length(z.l)
A.l <- sign(A.l + t(A.l))
diag(A.l) <- 0
graph.l <- ggraph(A.l, layout = 'stress') + 
  geom_edge_link(alpha = 1 / sqrt(n.l) / 2) + 
  geom_node_point(aes(colour = factor(z.l)), 
                  show.legend = FALSE) + 
  labs(colour = NULL) + 
  theme_void() + 
  scale_colour_brewer(palette = 'Set1')

gridExtra::grid.arrange(graph.l, graph.r, nrow = 1)
```

```{r mbconnectome-ase, fig.height = 2, fig.width = 8, out.width = '100%'}
Xhat.l <- embedding(A.l, 2, 1)
ase.l <- ggplot() + 
  geom_point(aes(x = Xhat.l[, 1], y = Xhat.l[, 2],
                 colour = factor(z.l))) + 
  coord_fixed(.5) +
  theme_bw() + 
  theme(legend.position = 'none') +
  scale_colour_brewer(palette = 'Set1') + 
  labs(x = NULL, y = NULL, colour = NULL)

Xhat.r <- embedding(A.r, 2, 1)
ase.r <- ggplot() + 
  geom_point(aes(x = Xhat.r[, 1], y = Xhat.r[, 2],
                 colour = factor(z.r))) + 
  coord_fixed(.5) +
  theme_bw() + 
  theme(legend.position = 'none') +
  scale_colour_brewer(palette = 'Set1') + 
  labs(x = NULL, y = NULL, colour = NULL)

gridExtra::grid.arrange(ase.l, ase.r, nrow = 1)
```

```{r mbconnectome-kcurves, fig.height = 2, fig.width = 8, cache = TRUE, out.width = '100%'}
set.seed(123456)

z.init.r <- sample.points(z.r, 6)
z.init.l <- sample.points(z.l, 6)

mbconnectome.out.r <- manifold.clustering(Xhat.r, 
                                          length(unique(z.r)), 
                                          initialization = z.init.r,
                                          intercept = FALSE, 
                                          parallel = TRUE, 
                                          verbose = FALSE, 
                                          animate = FALSE, 
                                          curve.init = 'x')
fitted.r <- plot.estimated.curves(Xhat.r, mbconnectome.out.r) + 
  scale_colour_brewer(palette = 'Set1') + 
  coord_fixed(.5) +
  labs(x = NULL, y = NULL) + 
  ylim(min(Xhat.r[, 2]), max(Xhat.r[, 2])) + 
  xlim(min(Xhat.r[, 1]), max(Xhat.r[, 1]))

mbconnectome.out.l <- manifold.clustering(Xhat.l, 
                                          length(unique(z.l)), 
                                          initialization = z.init.l,
                                          intercept = FALSE, 
                                          parallel = TRUE, 
                                          verbose = FALSE, 
                                          animate = FALSE, 
                                          curve.init = 'x')
fitted.l <- plot.estimated.curves(Xhat.l, mbconnectome.out.l) + 
  scale_colour_brewer(palette = 'Set1') + 
  coord_fixed(.5) +
  labs(x = NULL, y = NULL) + 
  ylim(min(Xhat.l[, 2]), max(Xhat.l[, 2])) + 
  xlim(min(Xhat.l[, 1]), max(Xhat.l[, 1]))

gridExtra::grid.arrange(fitted.l, fitted.r, nrow = 1)
```

### 2023-01-09 slides/slides.Rmd
updated slides

@@
 $$X = \begin{bmatrix}
-\omega_1 \sqrt{\theta_{11}} & 0 \\ 
+\omega_1 \sqrt{p} & 0 \\ 
 \vdots & \vdots \\ 
-\omega_{n_1} \sqrt{\theta_{11}} & 0 \\ 
-\omega_{n_1 + 1} \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_{n_1 + 1} \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\omega_{n_1} \sqrt{p} & 0 \\ 
+\omega_{n_1 + 1} \sqrt{r^2 / p} & \omega_{n_1 + 1} \sqrt{q - r^2 / p} \\ 
 \vdots & \vdots \\
-\omega_n \sqrt{\theta_{12}^2 / \theta_{11}} & \omega_n \sqrt{\theta_{22} - \theta_{12}^2 / \theta_{11}} \\ 
+\omega_n \sqrt{r^2 / p} & \omega_n \sqrt{q - r^2 / p} \\ 
 \end{bmatrix}$$
@@
 4. Partition $G$ into $K$ disconnected subgraphs.
+
+## Orthogonal Spectral Clustering
 
@@
   theme(legend.position = 'none') + 
+  coord_fixed() + 
   scale_colour_brewer(palette = 'Set1')

### 2023-01-09 draft/draft.Rmd
finished first draft of slides
added acknowledgements

@@
 \label{eq:kcurves-loss}
-L(z_1, ..., z_n, g_1, ..., g_n; X) = \frac{1}{n} \sum_{k=1}^K \sum_{i: z_i = k} \|x_i - g_k(t_i)\|^2
+L(z_1, ..., z_n, g_1, ..., g_K; X) = \frac{1}{n} \sum_{k=1}^K \sum_{i: z_i = k} \|x_i - g_k(t_i)\|^2
 \end{equation}
@@
 This makes community detection via clustering in the latent space a straightforward task: Construct a graph embedding and apply a well-known clustering algorithm that is appropriate for the (linear) latent structure.
-For the SBM, the appropriate choice of clustering algorithm is $K$-means clustering or Gaussian mixture models. 
+For the SBM, some appropriate choices of clustering algorithm are $K$-means clustering and Gaussian mixture models. 
 For the DCBM, an appropriate choice is $K$-means clustering with the cosine kernel. 
-For the PABM, an appropriate choice is sparse subspace clustering, and we also developed orthogonal spectral clustering for this setting as well. 
-Furthermore, by looking at how the embedding vectors are organized, the embedding makes visibly clear which of the three block models is the best fit for the graph, and in the case in which the number of communities or the assortativity or disassortativity of the model is unknown, there are existing heuristics for determining these based on the size of the eigenvalues.
+For the PABM, an appropriate choice is sparse subspace clustering, and we also developed orthogonal spectral clustering for this setting. 
+These all have asymptotic consistency properties, and we proved that sparse subspace clustering and orthogonal spectral clustering result in perfect community detection for the PABM for a sufficiently large yet finite graph size.
+Furthermore, by looking at how the embedding vectors are organized, the embedding makes visibly clear which of the three block models is the best fit for the graph (figure \ref{fig:block-models}), and in the case in which the number of communities or the assortativity or disassortativity of the model is unknown, there are existing heuristics for determining these based on the size of the eigenvalues.
 This is in contrast to direct likelihood maximization (e.g., via expectation maximization), which can be more of a black box in determining the appropriate model, number of communities, graph structure, etc., and results in slower, less interpretable algorithms.
@@
 
+Moving beyond the linear-structure block models (SBM, DCBM, PABM), we showed that the GRDPG and ASE approach can recover nonlinear structures in the latent space. 
+This type of nonlinearity has been observed in various real networks, in particular brain connectivity networks (figures \ref{fig:macaque-graph-conclusion} and \ref{fig:macaque-ase}). 
+While analyzing these graphs as GRDPGs reveals the (nonlinear) community structure, the linear block models are insufficient in modeling these graphs or performing statistical inference. 
+For this setting, we proposed the manifold block model, which is a type of GRDPG in which each community corresponds to a on nonlinear manifold in the latent space. 
+This led to two algorithms for manifold clustering on the latent space, the first which assumes that the manifolds are disjoint and involves performing single linkage clustering on the embedding, and the second which assumes that the manifolds are described by parametric curves ($K$-curves clustering).
+Both algorithms have provable asymptotic consistency properties, which we were able to demonstrate via simulation studies. 
+A natural extension of this work is modifying $K$-curves clustering to higher dimensional manifolds, developing more robust initialization strategies, and connecting MBMs to more interpretable block models outside the GRDPG framework.
+
 \begin{figure}[H]
@@
 \end{figure}
-
-Moving beyond the linear-structure block models (SBM, DCBM, PABM), we showed that the GRDPG and ASE approach can recover nonlinear structures in the latent space. 
-This type of nonlinearity has been observed in various real networks, in particular brain connectivity networks. 
-While analyzing these graphs as GRDPGs reveals the (nonlinear) community structure, the linear block models are insufficient in modeling these graphs or performing statistical inference. 
-For this setting, we proposed the Manifold Block Model, which is a type of GRDPG in which the communities lie on nonlinear manifolds in the latent space. 
-We then proposed two algorithms, one in which the manifolds are disjoint and another in which the manifolds intersect
 

### 2023-01-09 draft/draft.tex
added acknowledgements

@@
 
-December 1, 2022
+January 13, 2023
 
@@
 \LARGE{\bf{Acknowledgements}}
+\vspace*{1\baselineskip}
 \end{center}
 
+I would like to first thank my advisors Michael W. Trosset and Minh Tang, whose leadership and insight have shaped and guided my research journey. 
+I met Michael as a first year MSSS student taking his class on high dimensional data analysis, where he challenged us to investigate an open problem. 
+From the way in which I went about my solution, he was able to identify my strengths and weaknesses as a novice researcher and provide me with additional problems suited to my abilities and interests.
+It was through Michael that I met Minh, who became my co-advisor. 
+Much of my work is owed to following the breadcrumbs of intuition that Minh scattered for me in strategic locations. 
+Both Michael and Minh have been exceedingly patient with me as I sniffed out the paths that they set forth and slowly arrived at my own solutions. 
+I am indebted to their kindness, persistence, and wisdom.
+
+I would like to extend my gratitude toward my research committee members, Julia Fukuyama, Roni Khardon, and Fangzheng Xie. 
+I spent much of my time at IU as a TA in Julia's computational statistics class, and I learned much from watching her teach and work with students. 
+Roni's machine learning classes provided me with insight on using probabilistic and statistical methods outside the classical statistics setting, and PAC learning in particular has helped me understand my research objectives.
+Fangzheng always asks insightful questions and challenges me with perspectives I hadn't considered.
+
+None of this would have been possible without Dana Fielding and Kelly Hanna, whose work behind the scenes has been invaluable in making my time at IU smooth and pleasant. 
+I do not think I would have been able to complete anything as a graduate student without their help and encouragement.
+
 \hypersetup{linkcolor = black}
@@
 \begin{center}
-\LARGE{Abstract}
+\LARGE{\bf{Abstract}}
 \end{center}
@@
 This makes community detection via clustering in the latent space a straightforward task: Construct a graph embedding and apply a well-known clustering algorithm that is appropriate for the (linear) latent structure.
-For the SBM, the appropriate choice of clustering algorithm is \(K\)-means clustering or Gaussian mixture models.
+For the SBM, some appropriate choices of clustering algorithm are \(K\)-means clustering and Gaussian mixture models.
 For the DCBM, an appropriate choice is \(K\)-means clustering with the cosine kernel.
-For the PABM, an appropriate choice is sparse subspace clustering, and we also developed orthogonal spectral clustering for this setting as well.
-Furthermore, by looking at how the embedding vectors are organized, the embedding makes visibly clear which of the three block models is the best fit for the graph, and in the case in which the number of communities or the assortativity or disassortativity of the model is unknown, there are existing heuristics for determining these based on the size of the eigenvalues.
+For the PABM, an appropriate choice is sparse subspace clustering, and we also developed orthogonal spectral clustering for this setting.
+These all have asymptotic consistency properties, and we proved that sparse subspace clustering and orthogonal spectral clustering result in perfect community detection for the PABM for a sufficiently large yet finite graph size.
+Furthermore, by looking at how the embedding vectors are organized, the embedding makes visibly clear which of the three block models is the best fit for the graph (figure \ref{fig:block-models}), and in the case in which the number of communities or the assortativity or disassortativity of the model is unknown, there are existing heuristics for determining these based on the size of the eigenvalues.
 This is in contrast to direct likelihood maximization (e.g., via expectation maximization), which can be more of a black box in determining the appropriate model, number of communities, graph structure, etc., and results in slower, less interpretable algorithms.
@@
 
+Moving beyond the linear-structure block models (SBM, DCBM, PABM), we showed that the GRDPG and ASE approach can recover nonlinear structures in the latent space.
+This type of nonlinearity has been observed in various real networks, in particular brain connectivity networks (figures \ref{fig:macaque-graph-conclusion} and \ref{fig:macaque-ase}).
+While analyzing these graphs as GRDPGs reveals the (nonlinear) community structure, the linear block models are insufficient in modeling these graphs or performing statistical inference.
+For this setting, we proposed the manifold block model, which is a type of GRDPG in which each community corresponds to a on nonlinear manifold in the latent space.
+This led to two algorithms for manifold clustering on the latent space, the first which assumes that the manifolds are disjoint and involves performing single linkage clustering on the embedding, and the second which assumes that the manifolds are described by parametric curves (\(K\)-curves clustering).
+Both algorithms have provable asymptotic consistency properties, which we were able to demonstrate via simulation studies.
+A natural extension of this work is modifying \(K\)-curves clustering to higher dimensional manifolds, developing more robust initialization strategies, and connecting MBMs to more interpretable block models outside the GRDPG framework.
+
 \begin{figure}[H]
@@
 \end{figure}
-
-Moving beyond the linear-structure block models (SBM, DCBM, PABM), we showed that the GRDPG and ASE approach can recover nonlinear structures in the latent space.
-This type of nonlinearity has been observed in various real networks, in particular brain connectivity networks.
-While analyzing these graphs as GRDPGs reveals the (nonlinear) community structure, the linear block models are insufficient in modeling these graphs or performing statistical inference.
-For this setting, we proposed the Manifold Block Model, which is a type of GRDPG in which the communities lie on nonlinear manifolds in the latent space.
-We then proposed two algorithms, one in which the manifolds are disjoint and another in which the manifolds intersect
 

### 2023-01-10 draft/draft.Rmd
responded to more comments

@@
 Let $P$ be the $n \times n$ probability matrix of a two-community SBM. 
-Then $rank(P) = 2$, and its two nonzero eigenvalues are:
-
+Then $P$ can be written as $P = Z \Theta Z^\top$ where $\Theta$ is the $2 \times 2$ matrix of the community edge probabilities $\{\theta_{k \ell}\}$, and $Z$ is an $n \times 2$ matrix such that $z_{ik} = 1$ if vertex $i$ is in community $k$ and $0$ otherwise.
+The rank of $\Theta$ is 2, and  its two nonzero eigenvalues are:
 \begin{equation}
 \label{eq:sbm-eigenvalues}
-\lambda = \frac{n}{2} \bigg( 
+\lambda = \bigg( 
 \theta_{11} + \theta_{22} \pm 
@@
 The second eigenvalue is positive if $\theta_{11} \theta_{22} > \theta_{12}^2$. 
-Thus, $P$ describes an assortative SBM if $\theta_{11} \theta_{22} > \theta_{12}^2$ or 
-a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
+Since the rank and positive semidefiniteness of $P$ is given by $\Theta$, $P$ is positive semidefiniteif $\theta_{11} \theta_{22} > \theta_{12}^2$, which correspond to assortative and disassortative SBMs.
 This corresponds to the high-level understanding of assortativity and disassortativity based on whether within community edge probabilities are greater than the between community edge probabilities. 

### 2023-01-10 draft/draft.tex
responded to more comments

@@
 Indiana University \\
-December 2022}
+January 2023}
 \vspace*{3\baselineskip}

### 2023-01-10 draft/draft.Rmd
responded to more comments

@@
 
-The PABM is also a Bernoulli random graph, so it is also a GRDPG. 
-The latent vectors of the PABM when viewed as a GRDPG form $K$ $K$-dimensional subspaces, which makes for an elegant parallel between the nested forms of the block models and the nested structures of the GRDPGs: The SBM consists of $K$ zero-dimensional subspaces, and DCBM consists of $K$ one-dimensional subspaces.
-We explore this connection further the next section (section \@ref(sec:pabm-grdpg)).
+The PABM is also a Bernoulli random graph, so it is also a GRDPG, and it has community-wise structure in the latent space like the SBM and DCBM. 
+We will show in section \@ref(sec:pabm-grdpg) that this structure is comprised of $K$ $K$-dimensional subspaces in $\mathbb{R}^{K^2}$. 
+This relationship leads to an elegant parallel between the nested form of block models \citep{Noroozi2022} and the nested structures of the GRDPG. 
+More specifically, the latent positions of an SBM form $K$ point masses (zero-dimensional subspaces), and the latent positions of a DCBM form $K$ rays (one-dimensional subspaces).
 
@@
 \label{theorem:nonintersect-grdpg}
-Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$.
-Let $X = [ x_1 \mid \cdots \mid x_n ]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^{4c} n)$ for some $c > 1$.
+Let $A$ be a graph on $n$ vertices from the MBM model in
+definition \ref{def:mbm} wherein $n \rho_n = \omega(\log^{c}{n})$ for some $c > 1$ and the manifolds $\mathcal{M}_1, \dots, \mathcal{M}_{K}$ satisfy $\min_{k \neq \ell} \min_{x \in \mathcal{M}_k, y \in \mathcal{M}_{\ell}} \|x  - y\| \geq \delta$ for some $\delta > 0$ not depending on $n$ or $K$.
+Let $\hat{X}$ be the ASE of $A$ onto $\mathbb{R}^{d}$. 
 Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 

### 2023-01-10 draft/draft.tex
responded to more comments

@@
 
-The PABM is also a Bernoulli random graph, so it is also a GRDPG.
-The latent vectors of the PABM when viewed as a GRDPG form \(K\) \(K\)-dimensional subspaces, which makes for an elegant parallel between the nested forms of the block models and the nested structures of the GRDPGs: The SBM consists of \(K\) zero-dimensional subspaces, and DCBM consists of \(K\) one-dimensional subspaces.
-We explore this connection further the next section (section \ref{sec:pabm-grdpg}).
+The PABM is also a Bernoulli random graph, so it is also a GRDPG, and it has community-wise structure in the latent space like the SBM and DCBM.
+We will show in section \ref{sec:pabm-grdpg} that this structure is comprised of \(K\) \(K\)-dimensional subspaces in \(\mathbb{R}^{K^2}\).
+This relationship leads to an elegant parallel between the nested form of block models \citep{Noroozi2022} and the nested structures of the GRDPG.
+More specifically, the latent positions of an SBM form \(K\) point masses (zero-dimensional subspaces), and the latent positions of a DCBM form \(K\) rays (one-dimensional subspaces).
 

### 2023-01-10 slides/slides.Rmd
fixed a couple typos
fixed a typo
script ells

@@
 * $P_{ij} = \theta_{z_i z_j}$
-* $K (K + 1) / 2$ parameters $\theta_{kl}$
+* $K (K + 1) / 2$ parameters $\theta_{k \ell}$
 
@@
 * $P_{ij} = \theta_{z_i z_j} \omega_i \omega_j$
-* $K (K + 1) / 2 + n$ parameters $\theta_{kl}$, $\omega_i$
+* $K (K + 1) / 2 + n$ parameters $\theta_{k \ell}$, $\omega_i$
 
@@
 
-SBM $\to$ Erdos-Renyi: $\theta_{kl} = \theta$
+SBM $\to$ Erdos-Renyi: $\theta_{k \ell} = \theta$
 
@@
 
-$$\ell(z, \theta; A) = \sum_{i<j} \sum_{k, \ell} z_{ij} z_{j \ell} \Big(A_{ij} \log_{k \ell} + (1 - A_{ij}) \log(1 - \theta_{k \ell}) \Big)$$
+$$\ell(z, \theta; A) = \sum_{i<j} \sum_{k, \ell} z_{ik} z_{j \ell} \Big(A_{ij} \log_{k \ell} + (1 - A_{ij}) \log(1 - \theta_{k \ell}) \Big)$$
 
@@
 
-1. Let each $P^{(kl)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
+1. Let each $P^{(k \ell)}$ denote the $n_k \times n_l$ matrix of edge probabilities 
 between communities $k$ and $l$. 
 2. Organize popularity parameters as vectors 
-$\lambda^{(kl)} \in \mathbb{R}^{n_k}$ 
-such that $\lambda^{(kl)}_i = \lambda_{k_i l}$ is the popularity parameter 
+$\lambda^{(k \ell)} \in \mathbb{R}^{n_k}$ 
+such that $\lambda^{(k \ell)}_i = \lambda_{k_i l}$ is the popularity parameter 
 of the $i$^th^ vertex of community $k$ towards community $l$. 
 3. Each block can be decomposed as 
-$P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$.
+$P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(lk)})^\top$.
 
@@
 
-$$A \sim \PABM(\{\lambda^{(kl)}\}_K) \iff A \sim \GRDPG_{p, q}(X U)$$
+$$A \sim \PABM(\{\lambda^{(k \ell)}\}_K) \iff A \sim \GRDPG_{p, q}(X U)$$
 
@@
 
-**Corollary**: The ASE of $A \sim \PABM(\{\lambda^{(kl)}\}_K)$ lies near a collection of $K$-dimensional subspaces in $K^2$ dimensions.
+**Corollary**: The ASE of $A \sim \PABM(\{\lambda^{(k \ell)}\}_K)$ lies near a collection of $K$-dimensional subspaces in $K^2$ dimensions.
 
@@
 Noroozi et al. observed that the rank of $P$ is $K^2$ 
-and the columns of $P$ belonging to each community has rank $K$ 
+and the columns of $P$ belonging to each community have rank $K$ 
 to justify SSC for the PABM.
@@
 
-**Theorem** (KKT): Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a curve of order $R$ that does not self-intersect (for any $s \neq t$, $g_k(s) \neq g_k(t)$).
+**Theorem** (KKT):  
+Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a curve of order $R$ that does not self-intersect (for any $s \neq t$, $g_k(s) \neq g_k(t)$).
 Suppose that for each community $k$, we have labels for at least $R + 1$ vertices. 

### 2023-01-11 slides/slides.Rmd
fixed typo

@@
 
-$$\ell(z, \theta; A) = \sum_{i<j} \sum_{k, \ell} z_{ik} z_{j \ell} \Big(A_{ij} \log_{k \ell} + (1 - A_{ij}) \log(1 - \theta_{k \ell}) \Big)$$
+$$\ell(z, \theta; A) = \sum_{i<j} \sum_{k, \ell} z_{ik} z_{j \ell} \Big(A_{ij} \log \theta_{k \ell} + (1 - A_{ij}) \log(1 - \theta_{k \ell}) \Big)$$
 
@@
 
-## Example: Karantaka Villages
+## Example: Karantaka Villages (Banerjee et al., 2013)
 
@@
 
-**Theorem** (KKT):  
+**Theorem** (KTT):  
 Let $A \sim \MBM(\{\alpha_1, ..., \alpha_K\}, \{F, ..., F\}, \{g_1, ..., g_K\}; \rho_n)$ such that $F$ has support $[0, 1]$, and each $g_k(t) = g(t; p_k)$ is is a curve of order $R$ that does not self-intersect (for any $s \neq t$, $g_k(s) \neq g_k(t)$).

### 2023-01-12 slides/slides.Rmd
reference to jcgs paper

@@
 
-Erdos-Renyi Model (1959)
-
-* $P_{ij} = \theta$ (not a block model)
-* 1 parameter $\theta$
-
 Stochastic Block Model (Lorrain and White, 1971)
@@
 
-Popularity Adjusted Block Model (Sengupta and Chen, 2017)
+Popularity Adjusted Block Model (Sengupta and Chen, 2018)
 
@@
 composed of popularity vectors with each block corresponding to a community. 
+* $P = (X U) I_{pq} (X U)^\top$.
 
@@
 $$A \sim \PABM(\{\lambda_{ik}\}_K) \text{ iff } A \sim \GRDPG_{p, q}(X U)$$
-
-## PABMs are GRDPGs
-
-::: columns
-
-:::: column
-
-$$X = \begin{bmatrix}
-\Lambda^{(1)} & \cdots & 0 \\
-0 & \ddots & 0 \\
-0 & \cdots & \Lambda^{(K)}
-\end{bmatrix}$$
-
-::::
-
-:::: column
-
-\vspace*{2\baselineskip}
-
-$$U \in \mathbb{O}(K^2)$$
-
-::::
-
-:::
-
-$$A \sim \PABM(\{\lambda^{(k \ell)}\}_K) \iff A \sim \GRDPG_{p, q}(X U)$$
-
-**Remark 1** (orthogonality of subspaces):
-If $y_i^\top$ and $y_j^\top$ are two rows of $X U$ corresponding 
-to different communities, then $y_i^\top y_j = 0$.
-
-**Remark 2** (non-uniqueness of the latent configuration):  
-If $A \sim \GRDPG_{p, q}(Y)$, then $A \sim \GRDPG_{p, q}(Y Q)$ for any $Q$ in 
-the indefinite orthogonal group with signature $p, q$.
-
-**Remark 3**: Communities correspond to subspaces even with linear transformation $Q \in \mathbb{O}(p, q)$, but this may break the orthogonality property.
 
@@
 
+## PABMs are GRDPGs
+
+* Published in JCGS.
+* Code and preprint drafts available at https://github.com/johneverettkoo/pabm-grdpg
+* R package WIP at https://github.com/johneverettkoo/osc
+
 ## Overview of Block Models as GRDPGs

### 2023-01-12 slides/slides.Rmd
starting to add references

@@
 
-```{r mbconnectome-kcurves, fig.height = 2, fig.width = 8, cache = TRUE, out.width = '100%'}
+```{r mbconnectome-kcurves, fig.height = 2, fig.width = 8, cache = TRUE, out.width = '100%', fig.cap = 'Left and right hemispheres of the Drosophila connectome (top), ASEs (middle), and fitted curves (bottom).'}
 set.seed(123456)
@@
 ```
+
+## Expected Degree and Sparsity Factor
+
+Let $d_1, ..., d_n$ be the degrees of vertices $v_1, ..., v_n$. 
+For $P$ defined by a block model or GRDPG, $E[d_i] = O(n)$. 
+In most real networks, the degree does not scale linearly with the size of the network.
+
+Introduce a sparsity factor $\rho_n$ such that $A \sim \BG(\rho_n P)$.
+
+* Sparse regime: $\lim\limits_{n \to \infty} \rho_n \to 0$
+* Dense regime: $\rho_n = 1$
+
+Most results still require $n \rho_n = \omega (\log^c n)$, i.e., 
+expected degree still diverges, but at a logarithmic rate.
+
+## References

### 2023-01-13 slides/slides.Rmd
finished adding references

@@
 }
-\caption{Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by OSC.}\label{fig:mp}
+\caption{Adjacency matrices of (from left to right) the British MPs Twitter network (Greene and Cunningham, 2013), Political Blogs network (Adamic and Glance, 2005), and DBLP network (Gao et al., 2009) after sorting by the clustering outputted by OSC.}\label{fig:mp}
 \end{figure}
@@
 * Published in JCGS.
-* Code and preprint drafts available at https://github.com/johneverettkoo/pabm-grdpg
+* Code, data, and preprint drafts available at https://github.com/johneverettkoo/pabm-grdpg
 * R package WIP at https://github.com/johneverettkoo/osc
@@
 Define manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \in \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1]\}$ each by continuous function $g_k : [0, 1]^r \to \mathcal{X}$.
-Define probability distributions $F_1, ..., F_K$ each with support $[0, 1]^r$. 
+Define probability distribution $F$ with support $[0, 1]^r$. 
 Then the following mixture model is a *manifold block model*:
@@
 \begin{enumerate}
-\item Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$.
-\item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
+\item Draw labels $z_1, ..., z_n \iid \Categorical(\alpha_1, ..., \alpha_K)$.
+\item Draw latent vectors by first taking $t_1,..., t_n \iid F$ and then computing each $x_i = g_{z_i}(t_i)$. 
 \item Compile the latent vectors into data matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X)$. 
 \end{enumerate}
+
+## Manifold Block Model
+
+```{r intersect-curves-example, fig.height = 2, fig.width = 8, fig.cap = 'Latent vectors on intersecting curves (left), along with an RDPG drawn from this configuration (center) and its ASE (right).', cache = TRUE, out.width='100%'}
+f1 <- function(t) {
+  x1 <- t ^ 2
+  x2 <- 2 * t - 2 * t ^ 2
+  return(cbind(x1, x2))
+}
+
+f2 <- function(t) {
+  x1 <- 2 * t - 2 * t ^ 2
+  x2 <- 1 - 2 * t + t ^ 2
+  return(cbind(x1, x2))
+}
+
+set.seed(314159)
+
+n1 <- 2 ** 7
+n2 <- n1
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+
+a <- 1
+b <- 1
+t1 <- rbeta(n1, a, b)
+t2 <- rbeta(n2, a, b)
+T <- rbind(t1, t2)
+
+X1 <- f1(t1)
+X2 <- f2(t2)
+X <- rbind(X1, X2)
+
+latent.vecs <- ggplot() + 
+  geom_point(aes(x = X[, 1], y = X[, 2], colour = factor(z)), size = .5) + 
+  # coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  scale_colour_brewer(palette = 'Set1')
+
+P <- X %*% t(X)
+diag(P) <- 0
+A <- draw.graph(P)
+
+rdpg <- ggraph(A, layout = 'stress') + 
+  geom_edge_link(alpha = 1 / sqrt(n) / 2) + 
+  geom_node_point(aes(colour = factor(z)), 
+                  show.legend = FALSE) + 
+  labs(colour = NULL) + 
+  theme_void() + 
+  scale_colour_brewer(palette = 'Set1')
+
+Xhat <- embedding(A, 2, 0)
+
+ase <- ggplot() + 
+  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z)), size = .5) + 
+  # coord_fixed() + 
+  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
+  theme_bw() + 
+  theme(legend.position = 'none') + 
+  scale_colour_brewer(palette = 'Set1')
+
+gridExtra::grid.arrange(latent.vecs, rdpg, ase, nrow = 1)
+```
 
@@
 Collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$. 
-4. Draw $A \sim \RDPG(X)$ .
+4. Draw $A \sim \GRDPG_{2,0}(X)$ .
 
@@
 4. Collect the latent vectors into matrix $X = [ x_1 \mid \cdots \mid x_n ]^\top$.
-5. Draw $A \sim \RDPG(X)$.
+5. Draw $A \sim \GRDPG_{3,0}(X)$.
 
@@
 
-n <- 512
+n <- 1024
 
@@
 
+# Additional Slides
+
 ## Expected Degree and Sparsity Factor
@@
 
-Introduce a sparsity factor $\rho_n$ such that $A \sim \BG(\rho_n P)$.
+Introduce a sparsity factor $\rho_n$ and draw $A \sim \BG(\rho_n P)$.
 
@@
 Most results still require $n \rho_n = \omega (\log^c n)$, i.e., 
-expected degree still diverges, but at a logarithmic rate.
+expected degree still diverges, but at a logarithmic rate instead of linearly.
+
+## Subspace Detection Property
+
+If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
+$B_{ij} = 0$ for each $x_i$ and $x_j$ belonging to different subspaces.
+
+::: columns
+
+:::: column
+
+```{r, cache = TRUE, fig.height = 3, fig.width = 4, out.width = '50%'}
+set.seed(314159)
+
+n <- 64
+
+par(mar = rep(1, 4))
+
+x1 <- runif(n, 1 / sqrt(50), 1)
+y1 <- rep(0, n)
+x2 <- rep(0, n)
+y2 <- x1
+z <- c(rep(1, n), rep(2, n))
+X <- cbind(c(x1, x2), c(y1, y2))
+X <- X + cbind(runif(n * 2, 0, .025),
+               runif(n * 2, 0, .025))
+X <- X %*% rbind(c(cos(pi / 9), sin(pi / 9)),
+                 c(-sin(pi/ 9), cos(pi / 9)))
+plot(X, col = z * 2, asp = 1, xlab = NA, ylab = NA)
+```
+
+::::
+
+:::: column
+
+```{r, fig.height = 3, fig.width = 5, out.width = '75%'}
+Y <- t(X)
+Y <- sweep(Y, 2, apply(Y, 2, function(y) sqrt(sum(y ** 2))), `/`)
+Y[is.nan(Y)] <- 0
+lambda <- .05
+N <- ncol(Y)
+B <- plyr::aaply(seq(N), 1, function(i) {
+  y <- Y[, i]
+  X <- Y[, -i]
+  betahat <- glmnet::glmnet(X, y, lambda = lambda, intercept = FALSE) %>% 
+    coef() %>% 
+    as.numeric()
+  if (i != N) {
+    betahat <- c(betahat[seq(i)], 0, betahat[seq(i + 1, N)])
+  } else {
+    betahat <- c(betahat, 0)
+  }
+  betahat <- betahat[-1]
+  return(betahat)
+}, .parallel = TRUE) %>% 
+  abs()
+B <- sweep(B, 2, apply(B, 2, max), `/`)
+B[is.nan(B)] <- 0
+W <- B + t(B)
+
+qgraph::qgraph(sign(W), layout = 'spring', groups = factor(z), legend = FALSE)
+```
+
+::::
+
+:::
+
+**Theorem** (Wang and Xu, 2016): 
+If data matrix $X$ consists of points lying close to 
+low-dimensional subspaces such that:
+
+1. Each point's distance to its subspace is sufficiently small.
+2. The points corresponding to each subspace cover a sufficient amount of 
+that subspace.
+3. The cosine of the angles between pairs of subspaces is sufficiently small.
+
+Then $X$ obeys SDP.
+
+## Parameter Estimation
+
+1. Let $A^{(kl)}$ be the block of edges between communities $k$ and $l$. 
+2. For each $k, l \in [K]$, do:
+    i. Compute the SVD $A^{(kl)} = U \Sigma V^\top$.
+    ii. Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
+    Assign $\sigma^{(kl)} \leftarrow \Sigma_{11}^{1/2}$. 
+    iii. Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
+    $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
+
+**Theorem** (KTT):
+
+$$\max_{k, l} \|\hat{\lambda}^{(kl)} - \lambda^{(kl)}\| =
+O_P \Big( \frac{(\log n_k)^c}{\sqrt{n_k}} \Big)$$
+
+## Indefinite Orthogonal Group
+
+$$\mathbb{O}(p, q) = \{Q : Q I_{p, q} Q^\top = I_{p, q}\}$$
+
+* $Q^\top Q \neq I$
+* If $A \sim \GRDPG_{p, q}(X)$, then $A \sim \GRDPG_{p, q}(X Q)$
+* $(Q x)^\top (Q y) = x^\top Q^\top Q y \neq x^\top y$
+* $\|Q\| \neq 1$ $\implies$ $\|Q x - Q y\| \neq \|x - y\|$
+
+## Fitting Bezier Curves to Noisy Data
+
+Let $g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}$ be a Bezier curve of order $R$ with coefficients $p = [p^{(0)} \mid \cdots \mid p^{(R)} ]^\top$.
+
+Objective: minimize squared loss: $L(p, T; X) = \|X - T p\|^2$.
+
+$T$ is the $n \times (R+1)$ matrix of index points, i.e., $T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}$.
+
+If we know $T$, we can fit $p$: $\hat{p} = (T^\top T)^{-1} T^\top X$.
+
+If we know $p$, we can fit each $t_1, ..., t_n$ by finding the roots of $\dot{L}(t_i; p, x_i) = \bigg(\sum_{r=0}^R \binom{R}{r} (-1)^r c_r t^r \bigg) \bigg(\sum_{r=0}^{R-1} \binom{R-1}{r} (-1)^r c_{r+1} t^r \bigg)$, where $c_R = \sum_{r=0}^R (-1)^{R-r} \binom{R}{r} p^{(r)}$.
+
+## ASE Clustering for Nonintersecting Communities
+
+\begin{algorithm}[H]
+\label{alg:ase-singlelink}
+\scriptsize
+\DontPrintSemicolon
+\SetAlgoLined
+\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $(p, q)$.}
+\KwResult{Community assignments $z_1, ..., z_n \in \{1, ..., K\}$.}
+Compute $\hat{X}$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
+Apply single linkage clustering with $K$ communities on $\hat{X}$.\;
+\caption{Graph manifold clustering via single-linkage.}
+\end{algorithm}
+
+```{r two-curves, fig.cap = 'Manifold block model from two disjoint curves.
+The latent configuration is on the left, a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.', fig.height = 2, fig.width = 8, out.width='100%'}
+set.seed(123456)
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+u1 <- runif(n1)
+u2 <- runif(n2)
+x1 <- cos(pi / 3 * u1)
+y1 <- sin(pi / 3 * u1)
+x2 <- 1 + cos(pi / 3 * u2 + pi)
+y2 <- 1 + sin(pi / 3 * u2 + pi)
+data.matrix <- cbind(c(x1, x2), c(y1, y2))
+A <- draw.graph(data.matrix %*% t(data.matrix))
+Xhat <- embedding(A, 2, 0)
+
+par(mfrow = c(1, 3))
+# par(fig = c(0, .7, 0, 1))
+plot(data.matrix, col = z * 2, asp = 1,
+     xlab = NA, ylab = NA, 
+     # xaxt = 'n', yaxt = 'n',
+     xlim = c(-.1, 1.1), ylim = c(0, 1))
+# ggplot() +
+#   geom_point(aes(x = data.matrix[, 1], y = data.matrix[, 2], colour = factor(z))) +
+#   labs(x = NULL, y = NULL) +
+#   scale_colour_brewer(palette = 'Set1') +
+#   theme(legend.position = 'none') +
+#   coord_fixed()
+# par(fig = c(.3, 1, 0, 1))
+qgraph::qgraph(A, groups = factor(z), legend = FALSE)
+plot(Xhat, col = z * 2, asp = 1, 
+     # xaxt = 'n', yaxt = 'n', 
+     xlab = NA, ylab = NA)
+```
+
+## ASE Clustering for Nonintersecting Communities
+
+```{r fig.height = 2, fig.width = 8, out.width='100%'}
+set.seed(123456)
+n1 <- 2 ** 5
+n2 <- 2 ** 5
+n <- n1 + n2
+z <- c(rep(1, n1), rep(2, n2))
+u1 <- runif(n1)
+u2 <- runif(n2)
+x1 <- cos(pi / 3 * u1)
+y1 <- sin(pi / 3 * u1)
+x2 <- 1 + cos(pi / 3 * u2 + pi)
+y2 <- 1 + sin(pi / 3 * u2 + pi)
+data.matrix <- cbind(c(x1, x2), c(y1, y2))
+A <- draw.graph(data.matrix %*% t(data.matrix))
+Xhat <- embedding(A, 2, 0)
+
+par(mfrow = c(1, 3))
+# par(fig = c(0, .7, 0, 1))
+plot(data.matrix, col = z * 2, asp = 1,
+     xlab = NA, ylab = NA, 
+     # xaxt = 'n', yaxt = 'n',
+     xlim = c(-.1, 1.1), ylim = c(0, 1))
+# ggplot() +
+#   geom_point(aes(x = data.matrix[, 1], y = data.matrix[, 2], colour = factor(z))) +
+#   labs(x = NULL, y = NULL) +
+#   scale_colour_brewer(palette = 'Set1') +
+#   theme(legend.position = 'none') +
+#   coord_fixed()
+# par(fig = c(.3, 1, 0, 1))
+qgraph::qgraph(A, groups = factor(z), legend = FALSE)
+plot(Xhat, col = z * 2, asp = 1, 
+     # xaxt = 'n', yaxt = 'n', 
+     xlab = NA, ylab = NA)
+```
+
+**Theorem** (KTT): Let $A$ be a graph on $n$ vertices from the MBM model in wherein $n \rho_n = \omega(\log^{c}{n})$ for some $c > 1$ and the manifolds $\mathcal{M}_1, \dots, \mathcal{M}_{K}$ satisfy $\min_{k \neq \ell} \min_{x \in \mathcal{M}_k, y \in \mathcal{M}_{\ell}} \|x  - y\| \geq \delta$ for some $\delta > 0$ not depending on $n$ or $K$.
+Let $\hat{X}$ be the ASE of $A$ onto $\mathbb{R}^{d}$. 
+Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
+Then for some $C > 0$ and any $\eta \in (0, C \delta)$, $\lim\limits_{n \to \infty} P(A_n(\eta)) = 1$.
 
 ## References
+
+* Adamic, L. A. and Glance, N. (2005). The political blogosphere and the 2004 U.S. election: Divided they blog. In *Proceedings of the 3rd International Workshop on Link Discovery*, LinkKDD '05. 
+* Banerjee, A., Chandrasekhar, A. G., Duflo, E., and Jackson, M. O. (2013). The diffusion of Microfinance. Harvard Dataverse. 
+* Eichler, K., Li, F., Litwin-Kumar, A., Park, Y., Andrade, I., Schneider-Mizell, C.
+M., Saumweber, T., Huser, A., Eschbach, C., Gerber, B., Fetter, R.
+D., Truman, J.
+W., Priebe, C.
+E., Abbot, L.
+F., Thum, A.
+S., Zlatic, M., and Cardona, A.
+(2017).
+The complete connectome of a learning and memory center in an insect brain.
+*bioRxiv*.
+
+## References
+
+* Elhamifar, E. and Vidal, R. (2009). Sparse Subspace Clustering. *2009 IEEE Conference on Computer Vision and Pattern Recognition*. 
+* Gao, J., Liang, F., Fan, W., Sun, Y., and Han, J.
+(2009).
+Graph-based consensus maximization among multiple supervised and unsupervised models.
+*Advances in Neural Information Processing Systems 22*.
+* Greene, D. and Cunningham, P. (2013). Producing a unified graph representation from multiple social network views. *CoRR*. 
+* Karrer, B. and Newman, M. E. J. (2011). Stochastic blockmodels and community structure in networks. *Physical Review E*. 
+* Koo, J., Tang, M., and Trosset, M. W. (2022). Popularity adjusted block models are generalized random dot product graphs. *Journal of Computational and Graphical Statistics*.
+
+## References
+
+* Lorrain, F., and White, H. C. (1971). Structural equivalence of individuals in social networks. *The Journal of Mathematical Society*. 
+* Nepusz, T., Petroczi, A., Negyessy, L, and Bazso, F. (2008). Fuzzy communities and the concept of bridgeness in complex networks. *Phys. Rev. E*. 
+* Noroozi, M. and Pensky, M. (2022). The hierarchy of block models. *Sankhya A*.
+* Noroozi, M., Rimal, R., and Pensky, M. (2019). Estimation and clustering in the popularity adjusted block model. *Journal of the Royal Statistical Society: Series B (Statistical Methodology)*.
+
+## References
+
+* Negyessy, L., Nepusz, T., Kocsis, L., and Bazso, F.
+(2006).
+Prediction of the main cortical areas and connections involved in the tactile function of the visual cortex by network analysis.
+*European Journal of Neuroscience*.
+* Rubin-Delanchy, P., Cape, J., Tang, M., and Priebe, C.
+E.
+(2022).
+A statistical interpretation of spectral embedding: The generalised random dot product graph.
+*Journal of the Royal Statistical Society: Series B (Statistical Methodology)*.
+* Sengupta, S. and Chen, Y. (2018). A block model for node popularity in networks with community structure. *Journal of the Royal Statistical Society, Series B (Statistical Methodology)*. 
+* Sussman, D. L., Tang, M., Fishkind, D. E., and Priebe, C. E. (2012). A consistent adjacency spectral embedding for stochastic blockmodel graphs. *Journal of the American Statistical Association*.
+
+## References
+
+* Wang, Y.-X. and Xu, H. (2016). Noisy sparse subspace clustering. *Journal of Machine Learning Research*.

### 2023-01-13 slides/slides.Rmd
tried to make a plot prettier

@@
 
-1. Draw $z_1, ..., z_n \iid \Multinomial(1/2, 1/2)$.
+1. Draw $z_1, ..., z_n \iid \Categorical(1/2, 1/2)$.
 2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
@@
 
-1. Draw $z_1, ..., z_n \iid \Multinomial(1/3, 1/3, 1/3)$.
+1. Draw $z_1, ..., z_n \iid \Categorical(1/3, 1/3, 1/3)$.
 2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
@@
 ```{r three-curves, cache = TRUE, fig.height = 4, fig.width = 15, fig.cap = 'Latent vectors (left) and ASE (right) of one realization of the mixture model in the simulation setup with $K=3$ manifolds.', out.width = '100%'}
-par(mfrow = c(1, 2))
+par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
 

### 2023-01-13 slides/slides.Rmd
made a plot prettier

@@
                              xlab = NA, ylab = NA, zlab = NA,
-                             pch = '.', asp = 1,
-                             tick.marks = FALSE)
+                             # pch = '.', 
+                             asp = 1,
+                             tick.marks = FALSE,
+                             mar = c(0, 0, 0, 0))
 
@@
                              xlab = NA, ylab = NA, zlab = NA, 
-                             pch = '.', asp = 1, 
+                             # pch = '.', 
+                             asp = 1, 
                              tick.marks = FALSE,
-                             angle = 40)
+                             angle = 40,
+                             mar = c(0, 0, 0, 0))
 ```

### 2023-03-27 draft/draft.Rmd
revisions

@@
 - \usepackage{comment}
+- \usepackage{sectsty}
+- \usepackage{tocloft}
+- \usepackage{titletoc}
 - \DeclareMathOperator*{\argmax}{arg\,max}
@@
 ```{=tex}
+\sectionfont{\fontsize{12}{12}\scshape}
+\subsectionfont{\fontsize{12}{12}\selectfont}
 \setlength{\parindent}{20.0pt}
@@
 \newpage
-
-# References

### 2023-03-27 draft/draft.tex
revisions

@@
 \usepackage{comment}
+\usepackage{sectsty}
+\usepackage{tocloft}
+\usepackage{titletoc,titlesec}
 \DeclareMathOperator*{\argmax}{arg\,max}
 \DeclareMathOperator*{\argmin}{arg\,min}
+
 \ifLuaTeX
@@
 \newtheorem{conjecture}{Conjecture}[section]
-% \theoremstyle{definition}
+\theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-% \theoremstyle{definition}
+\theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-% \theoremstyle{definition}
+\theoremstyle{definition}
 \newtheorem{exercise}{Exercise}[section]
-% \theoremstyle{definition}
+\theoremstyle{definition}
 \newtheorem{hypothesis}{Hypothesis}[section]
-% \theoremstyle{remark}
+\theoremstyle{remark}
 \newtheorem{remark}{Remark}
@@
 \begin{center}
-\LARGE{\textsc{Community Detection in the Setting of Generalized Random Dot Product Graphs}}\\
-\vspace*{4\baselineskip}
+\Large{\textsc{Community Detection in the Setting of Generalized Random Dot Product Graphs}}\\
+\vspace*{6\baselineskip}
 \normalsize{John Koo}\\
@@
 Indiana University \\
-January 2023}
+March 2023}
 \vspace*{3\baselineskip}
@@
 
-\begin{tabular}{@{}p{1in}p{4in}@{}}
-Approved: & \hrulefill \\
+\begin{tabular}
+{>{\raggedright\arraybackslash}p{1.5in}
+ >{\raggedleft\arraybackslash}p{4.75in}}
+Doctoral Committee: & \hrulefill \\
 & Michael W. Trosset, Ph.D. \\
@@
 \begin{center}
-\LARGE{\bf{Acknowledgements}}
+\bf{Acknowledgements}
 \vspace*{1\baselineskip}
@@
 \begin{center}
-\LARGE{\bf{Abstract}}
+John Koo\\
+\textsc{Community Detection in the Setting of Generalized Random Dot Product Graphs}\\
 \end{center}
 
-\vspace*{2\baselineskip}
+\vspace*{1\baselineskip}
 
@@
 
-\addcontentsline{toc}{section}{\contentsname}
+% \addcontentsline{toc}{section}{Curriculum Vitae}
 
 \newpage
-\pagenumbering{arabic}
 \hypersetup{linkcolor = blue}
@@
 \setcounter{tocdepth}{2}
+
+\renewcommand{\cfttoctitlefont}{\bf}
+
 \tableofcontents
 }
+\sectionfont{\fontsize{12}{12}\scshape}
+\subsectionfont{\fontsize{12}{12}\selectfont}
 \setlength{\parindent}{20.0pt}
@@
 \newpage
+
+\pagenumbering{arabic}
 
@@
 
-\renewcommand\refname{References}
 \bibliography{bibliography.bib}
 
+\newpage
+
+\pagenumbering{gobble}
+
+\section*{Curriculum Vitae}
+
+\begin{singlespace}
+
+\subsection*{Education}
+
+\begin{itemize}
+  \item Ph.D., Statistical Science, March 2023 \\
+        Indiana University, Bloomington, IN \\
+        Dissertation: \textit{Community Detection in the Setting of Generalized Random Dot Product Graphs} \\
+        Advisors: Michael W. Trosset and Minh Tang
+  \item M.S., Mathematics, August 2015 \\
+        Tarleton State University, Stephenville, TX \\
+        Thesis: \textit{Statistical Methods for Crop Yield Monitoring via Remote Sensing} \\
+        Advisors: Bert Little and Jesse Crawford
+  \item B.A., Mathematics, Physics, August 2012 \\
+        New York University, New York, NY
+\end{itemize}
+
+\subsection*{Research Interests}
+
+Statistical Inference on graphs and networks, computational statistics, dimensionality reduction, statistical pattern recognition, manifold learning
+
+\subsection*{Publications}
+
+\begin{itemize}
+  \item John Koo, Minh Tang, Michael W.
+Trosset (2023) Popularity Adjusted Block Models are Generalized Random Dot Product Graphs, \textit{Journal of Computational and Graphical Statistics}, 32:1, 131-144, DOI: 10.1080/10618600.2022.2081576
+\end{itemize}
+
+\subsection*{Talks and Presentations}
+
+\begin{itemize}
+  \item \textit{Popularity Adjusted Block Models are Generalized Random Dot Product Graphs}. Joint Statistical Meetings, Washington, DC. August 2023.
+  \item \textit{Popularity Adjusted Block Models are Generalized Random Dot Product Graphs}. MIDAS Future Leaders Summit, Ann Arbor, MI. April 2023.
+  \item \textit{Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph}. Symposium on Data Science and Statistics, Virtual. June 2021. 
+  \item \textit{Introduction to Git and GitHub}. Machine Learning for Research Club at IU, Bloomington, IN. February 2021. 
+\end{itemize}
+
+\subsection*{Software}
+
+\begin{itemize}
+  \item R package \verb|osc|: An implementation of orthogonal spectral clustering for community detection in linear latent structure block models.
+Available at \verb|https://github.com/johneverettkoo/osc|.
+\end{itemize}
+
+\subsection*{Courses Taught}
+
+\begin{itemize}
+  \item Spring 2023. STAT-S 301: Applied Statistical Methods for Business. 
+  \item Spring 2018. STAT-S 301: Applied Statistical Methods for Business. 
+\end{itemize}
+
+\subsection*{Industry Experience}
+
+\begin{itemize}
+  \item Data Science Consultant \\
+        Freelance, Bloomington, IN \\
+        2020 - present
+  \item Data Science Intern \\
+        The Climate Corporation, Seattle, WA \\
+        May 2019 - August 2019
+  \item Data Science Intern \\
+        The Bee Corp, Bloomington, IN \\
+        May 2018 - August 2018
+  \item Data Analyst \\
+        The Climate Corporation, San Francisco, CA \\
+        September 2015 - July 2017
+  \item Research Intern \\
+        Center for Agribusiness Excellence, Stephenville, TX \\
+        June 2014 - August 2015
+  \item Software Test Engineer \\
+        DEKA Research, Manchester, NH \\
+        January 2014 - May 2014
+  \item Mathematics Teaching Fellow \\
+        Lawrence High School, Lawrence, MA \\
+        July 2012 - June 2013
+\end{itemize}
+
+\end{singlespace}
+
+\cftaddtitleline{toc}{section}{References}{100}
+
+\cftaddtitleline{toc}{section}{Curriculum Vitae}{}
+
 \end{document}
